### 2020-04-30 description/gazeHMM_description.Rmd
upload algorithm first version

---
title: "gazeHMM Description"
author: "Malte Lüken"
date: "27.04.2020"
output: pdf_document
---

# General

The algorithm is divided in three parts: Preprocessing, classification, and postprocessing.

Input: x-coordinate, y-coordinate, timestamp, unit of coordinates (px vs. va), screen resolution, screen dimensions, 
distance to screen, sampling rate, blink indicator, number of states in the HMM, start values for the HMM, downsampling factor,
control options for fitting the HMM.

# Preprocessing

1. The algorithm checks whether samples are valid: Samples for which the x- or y-coordinates are NA, infinity, or (0,0) are marked (internally) as invalid.

2. Samples that match the blink indicator are labeled as blinks.

3. If the unit is px, the x- and y-coordinates are transformed into va with: 
    i. The coordinates are centered ([0,0] is center)
    ii. $\theta_{deg} = atan(\frac{x*res_x}{d_{screen}*dim_x})\frac{180}{\pi}$

4. Samples that are outside the screen are marked as invalid.

5. A Butterworth filter (3rd order, critical frequency = 0.5 of the Nyquist frequency) is applied to the valid samples (this step is optional).

6. Velocities and accelerations for the x- and y-coordiantes are calculated using a Savitzky-Golay filter (3rd order, length = 5).

7. Combined velocities and accelerations are calculated as: $v = f \sqrt{v_x^2 + v_y^2}$ and $a = f \sqrt{a_x^2 + a_y^2}$ with $f$ as the sampling rate.

8. Velocities and accelerations of value zero are nudged to be 0.01.

9. Samples with velocity above 1000 deg/s or acceleration above 100000 deg/$s^2$ are marked as invalid.

10. The sample-to-sample angle is calculated as: 
$$\theta_t = atan2(\frac{y_{t+1} - y_t}{x_{t+1} - x_t}) - atan2(\frac{y_t - y_{t-1}}{x_t - x_{t-1}}).$$ 
Angles are normalized to $[0,2\pi]$ by adding $2\pi$ to negative values.

11. Samples that have an angle with value NA are marked as invalid.

12. The velocity, acceleration, and angle of invalid samples are set to NA.

Output: Data frame with x-coordinate, y-coordinate, timestamp, velocity, acceleration, angle, and initial label variable with only blinks.

# Classifcation

Input: Data frame from preprocessing and HMM related general input variables.

1. Sets default start values if none are provided.

2. Downsamples velocity and acceleration.

3. Fits a HMM with velocity, acceleration, and angle as dependent variables. 
Velocity and acceleration are modeled by a two-parameter gamma distribution (shape and scale). 
Angle is modeled by a uniform distribution for the first component and von-Mises distributions (mean and kappa) for further components.

Output: Fitted model.

# Postprocessing

Input: Data frame from preprocessing and posterior state sequence of the fitted model.

1. Relabels posterior states of samples with recursive algorithm:
    i. If a fixation or smooth pursuit sample has no previous or following sample with the same label (single sample event), 
it is relabeled as the label from the previous sample.
    ii. If a PSO sample is followed by a saccade sample or following a non-saccade sample,
it is relabeled as the last different label.
    iii. Saccade events (batch of consecutive saccade samples) below 10 ms are relabeled as the label from the previous event.
    iv. These steps are applied recursively until all conditions hold.

2. For each event (batch of subsequent samples belonging to the same label), the following metrics are computed:
    i. Fixations:
        a. Duration: Max timestamp - min timestamp
        b. Position: Trimmed mean (20%) of x- and y-coordinates
    ii. Saccades, PSOs, and smooth pursuits:
        a. Duration: Max timestamp - min timestamp
        b. Start: First x- and y- coordinates
        c. End: Last x- and y- coordinates
        d. Amplitude: Euclidian distance from start to end
        e. Max. velocity
        f. Avg. velocity
        g. Max. acceleration
        h. Avg. acceleration
        i. Direction: $\theta = atan2(\frac{end_y - start_y}{end_x - start_x})$

For duration all samples are used, for the rest only valid samples (from preprocessing).

General output: List with a data frame containing the labeled samples, a list with data frames for the different events,
and the fitted model.

### 2020-04-30 algorithm/gazeHMM.R

@@ -1,0 +1,36 @@
+### gazeHMM - full algorithm ###
+
+# Author: Malte Lüken
+# Date: 24.04.2020
+
+
+classify_gaze_data <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
+                               nstates, respstart, trstart, instart, sf = c(10, 10),
+                               fit.control = em.control(maxit = 5000, random.start = F)) {
+  
+  source("algorithm/preprocessing.R")
+  source("algorithm/model.R")
+  source("algorithm/postprocessing.R")
+  
+  
+  # Preprocessing
+  
+  prep <- preprocess(x, y, t, unit, res, dim, dist, fr, blink)
+  
+  
+  # Model classification
+  
+  model.fit <- HMM_classify(prep, nstates, respstart, trstart, instart, sf, fit.control)
+  
+  
+  # Postprocessing
+  
+  post <- postprocess(prep, model.fit@posterior$state)
+  
+  
+  # Output: Data frame with samples, data frames with events, fitted model
+  
+  output <- list(samples = post$samples, events = post$events, model = model.fit)
+  
+  return(output)
+}

### 2020-04-30 algorithm/model.R

@@ -1,0 +1,118 @@
+### Classification by hidden Markov model ###
+
+# Author: Malte Lüken
+# Date: 25.05.2020
+
+
+HMM_classify <- function(data, nstates, respstart, trstart, instart, 
+                         sf = c(10, 10),
+                         fit.control = em.control(maxit = 5000, random.start = F)) {
+  
+  require(depmixS4)
+  source("algorithm/model_helper_functions.R")
+  
+  
+  # Check if starting values for response model supplied, set to default if not
+  
+  if(missing(respstart)) {
+    if(nstates == 1) {
+      
+      respstart <- list(fix = list(vel = c(1, 1), acc = c(1, 1)))
+      
+    } else if(nstates == 2) {
+      
+      respstart <- list(fix = list(vel = c(1, 1), acc = c(1, 1)), 
+                        sac = list(vel = c(5, 5), acc = c(5, 5), angle = c(0, 10)))
+      
+    } else if(nstates == 3) {
+      
+      respstart <- list(fix = list(vel = c(1, 1), acc = c(1, 1)), 
+                        sac = list(vel = c(5, 5), acc = c(5, 5), angle = c(0, 10)),
+                        pso = list(vel = c(5, 5), acc = c(5, 5), angle = c(pi, 10)))
+      
+    } else if(nstates == 4) {
+      
+      respstart <- list(fix = list(vel = c(1, 1), acc = c(1, 1)), 
+                        sac = list(vel = c(5, 5), acc = c(5, 5), angle = c(0, 10)),
+                        pso = list(vel = c(5, 5), acc = c(5, 5), angle = c(pi, 10)),
+                        sp = list(vel = c(2, 2), acc = c(2, 2), angle = c(0, 2)))
+      
+    } else if(nstates == 5) {
+      
+      respstart <- list(fix = list(vel = c(1, 1), acc = c(1, 1)), 
+                        sac = list(vel = c(5, 5), acc = c(5, 5), angle = c(0, 10)),
+                        pso = list(vel = c(5, 5), acc = c(5, 5), angle = c(pi, 10)),
+                        sp = list(vel = c(2, 2), acc = c(2, 2), angle = c(0, 2)),
+                        mic = list(vel = c(2, 2), acc = c(5, 5), angle = c(0, 10)))
+      
+    } 
+  }
+  
+  
+  # Check if starting values for transition model are supplied, set to default if not
+  
[…]

### 2020-04-30 algorithm/model_helper_functions.R

@@ -1,0 +1,473 @@
+### Model functions ###
+
+# Author: Malte Lüken
+# Date: 01.04.2020
+
+library(depmixS4)
+library(CircStats)
+library(gamlss)
+library(gamlss.dist)
+
+
+# Create response model for uniform distribution --------------------------
+
+setClass("unif", contains = "response")
+
+setGeneric("unif", function(y, ...) standardGeneric("unif"))
+
+
+# Define the method that creates the response class
+
+setMethod("unif", 
+          signature(y = "ANY"), 
+          function(y, fixed = NULL, ...) {
+            
+            y <- matrix(y, length(y))
+            x <- matrix(1)
+            parameters <- list()
+            npar <- 2 # min and max
+            
+            if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
+            
+            parameters$min <- min(y)
+            parameters$max <- max(y)
+            
+            mod <- new("unif", parameters = parameters, fixed = fixed, x = x, y = y, npar = npar)
+            
+            return(mod)
+          }
+)
+
+
+# Define method that prints parameters
+
+setMethod("show", "unif",
+          function(object) {
+            
+            cat("Model of type uniform\n")
+            cat("Parameters: \n")
+            cat("min: ", object@parameters$min, "\n")
+            cat("max: ", object@parameters$max, "\n")
+          }
+)
+
+
+# Define method that evaluates the density
+
+setMethod("dens", "unif",
+          function(object, log = FALSE) {
+            
+            dens <- ifelse(is.na(object@y), 1, dunif(object@y, min = object@parameters$min, max = object@parameters$max, log = log))
+            
+          }
+)
+
+
+# Define method that retrieves parameters
+
+setMethod("getpars", "response",
+          function(object, which = "pars", ...) {
+            
+            switch(which,
+                   "pars" = {
+                     parameters <- numeric()
+                     parameters <- unlist(object@parameters)
+                     pars <- parameters
+                   },
+                   "fixed" = {
+                     pars <- object@fixed
+                   }
+            )
+            return(pars)
+          }
+)
+
+
+# Define method that sets parameters
+
+setMethod("setpars", "unif",
+          function(object, values, which = "pars", ...) {
+            
+            npar <- npar(object)
+            
+            if(length(values) != npar) stop("length of 'values' must be", npar)
+            
+            nms <- names(object@parameters)
+            
+            switch(which,
+                   "pars"= {
+                     object@parameters$min <- values[1]
[…]

### 2020-04-30 algorithm/plotting_functions.R

@@ -1,0 +1,74 @@
+### Plotting functions ###
+
+# Author: Malte Lüken
+# Date: 26.04.2020
+
+
+plot_samples_time <- function(object, from = 0, to = Inf, display = c("x", "y", "vel", "acc", "angle"), add.event = F) {
+  
+  require(ggplot2)
+  require(dplyr)
+  
+  data_long <- object$samples %>% 
+    gather(all_of(display), key = metric, value = value, factor_key = T) %>% 
+    mutate_at(vars(label, state), as.factor) %>%
+    dplyr::filter(t > from, t <= to)
+
+  p <- ggplot(data_long , aes(x = t, y = value)) + facet_wrap(vars(metric), scales = "free") +
+    geom_path() + xlab("time in s") + ylab("metric") +
+    geom_point(aes(color = label, shape = state))
+  
+  if(add.event) p <- p + geom_text(aes(label = event, hjust = 1, vjust = 1, size = 3))
+    
+  return(p)
+}
+
+
+plot_samples_xy <- function(object, from = 0, to = Inf, add.event = F, add.fixpos = F) {
+  
+  require(ggplot2)
+  require(dplyr)
+  
+  data <- object$samples %>% 
+    dplyr::filter(t > from, t <= to, label != 0) %>%
+    mutate_at(vars(label, state), as.factor) %>% 
+    group_by(event) %>%
+    mutate(fix.x = ifelse(label == 1, mean(x[!is.na(vel)], trim = 0.2), NA),
+           fix.y = ifelse(label == 1, mean(y[!is.na(vel)], trim = 0.2), NA),)
+  
+  p <- ggplot(data, aes(x, y)) + 
+    geom_point(aes(color = label, shape = state)) +
+    geom_path()
+  
+  if(add.event) p <- p + geom_text(aes(label = event), hjust = 1, vjust = 1, size = 3)
+  
+  if(add.fixpos) p <- p + geom_point(aes(fix.x, fix.y), color = "green1")
+  
+  return(p)
+}
+
+
+plot_posterior_time <- function(object, from = 0, to = Inf, prob = T) {
+  
+  require(ggplot2)
+  require(dplyr)
+  
+  data <- cbind(object$samples, object$model@posterior[,-1]) %>%
+    gather(contains("S"), key = "state", value = "post.prob") %>%
+    dplyr::filter(state != "state", t > from, t <= to)
+  
+  entropy <- data %>% 
[…]

### 2020-04-30 algorithm/postprocessing.R

@@ -1,0 +1,127 @@
+### Postprocessing algorithm ###
+
+# Author: Malte Lüken
+# Date: 20.04.2020
+
+
+postprocess <- function(data, state) { # data frame with dependent variables, vector with state for each observation
+  
+  source("algorithm/postprocessing_helper_functions.R")
+  
+  
+  # Assign labels to events
+  
+  FIX <- 1
+  SAC <- 2
+  PSO <- 3
+  SP <- 4
+  
+  
+  # Mark invalid samples
+  
+  valid <- !is.na(data$vel) & !is.na(data$acc) & !is.na(data$angle)
+  
+  
+  # Relabel samples
+  
+  df <- reclassify(state, data$t)
+  
+  label <- df$label
+  number <- df$number
+  
+  warning(paste(sum(label != state), " samples were relabeled during postprocessing!"))
+  
+  
+  # Initialize output event metrics 
+  
+  dur <- numeric(max(number))
+  event <- integer(max(number))
+  
+  x <- numeric(max(number))
+  y <- numeric(max(number))
+  
+  start.x <- numeric(max(number))
+  end.x <- numeric(max(number))
+  start.y <- numeric(max(number))
+  end.y <- numeric(max(number))
+  
+  amp <- numeric(max(number))
+  
+  max.vel <- numeric(max(number))
+  avg.vel <- numeric(max(number))
+  max.acc <- numeric(max(number))
+  avg.acc <- numeric(max(number))
+  
+  dir <- numeric(max(number))
+  
+  
+  # Calculate output event metrics
+  
+  for (n in unique(number)) {
+    
+    # Duration
+    
+    if(n > 1) {
+      dur[n] <- max(data$t[number == n], na.rm = T) - max(data$t[number == (n-1)], na.rm = T)
+    } else {
+      dur[n] <- max(data$t[number == n], na.rm = T) - min(data$t[number == n], na.rm = T)
+    }
+    event[n] <- max(label[number == n])
+    
+    if(event[n] == FIX) {
+      
+      # Fixation position
+
+      x[n] <- ifelse(sum(valid[number == n]) == 0, NA, mean(data$x[number == n & valid], trim = 0.2, na.rm = T))
+      y[n] <- ifelse(sum(valid[number == n]) == 0, NA, mean(data$y[number == n & valid], trim = 0.2, na.rm = T))
+      
+    } else {
+      
+      # Start and end position
+      
[…]

### 2020-04-30 algorithm/postprocessing_helper_functions.R

@@ -1,0 +1,159 @@
+### Postprocessing helper functions ###
+
+# Author: Malte Lüken
+# Date: 23.04.2020
+
+
+# Function to retrieve last (different) state
+
+last_state <- function(state, s) {
+  if(state[s-1] != state[s] || (s-1) == 1) {
+    
+    return(state[s-1])
+  } else {
+    
+    return(last_state(state, s-1))
+  }
+}
+
+
+# Function to relabel samples
+
+reclassify <- function(state, t) {
+  
+  # Assign labels to events
+  
+  FIX <- 1
+  SAC <- 2
+  PSO <- 3
+  SP <- 4
+  
+  
+  # Relabel single sample fixations and smooth pursuit
+  
+  number <- integer(length(state))
+  
+  number[1] <- 1
+  
+  counter <- 1
+  
+  N <- length(state)
+  
+  for (s in 2:(N-1)) {
+    
+    if(state[s] %in% c(FIX, SP) && state[s-1] != state[s] && state[s+1] != state[s]) {
+      
+      state[s] <- state[s-1]
+      
+    } 
+    
+    
+    # Count consecutive samples with same label (events)
+    
+    if(state[s] != state[s-1]) counter <- counter + 1
+    
+    number[s] <- counter
+  }
+  
+  number[length(number)] <- max(number)
+  
+  
+  # Relabel saccades with duration below threshold
+  
+  dur <- numeric(max(number))
+  event <- integer(max(number))
+  
+  for (n in unique(number)) {
+    
+    if(n > 1) {
+      dur[n] <- max(t[number == n], na.rm = T) - max(t[number == (n-1)], na.rm = T)
+    } else {
+      dur[n] <- max(t[number == n], na.rm = T) - min(t[number == n], na.rm = T)
+    }
+    event[n] <- max(state[number == n])
+    
+    if(event[n] == SAC && dur[n] < 0.01 && n > 1) {
+      
+      state[number == n] <- last_state(event, n)
+      
+    }
+  }
+  
+  
+  # Relabel PSOs that do not occur immediately after saccade
+  
+  for (s in 2:(N-1)) {
+    if(state[s] == PSO && state[s+1] == SAC) {
+      
+      state[s] <- state[s+1]
+      
+    } else if (state[s] == PSO && state[s-1] %in% c(FIX, SP) && state[s+1] != SAC) {
+      
+      state[s] <- state[s-1]
+      
+    }
+  }
+  
+  
[…]

### 2020-04-30 algorithm/preprocessing.R

@@ -1,0 +1,130 @@
+### Preprocessing ###
+
+# Author: Malte Lüken
+# Date: 07.04.2020
+
+
+preprocess <- function(x, y, t, unit = "px",
+                       res, dim, dist,
+                       fr, blink = NULL) {
+  
+  require(signal)
+  source("algorithm/preprocessing_helper_functions.R")
+  
+  
+  # Initialize sample metrics
+  
+  vel <- rep(NA, length(x))
+  acc <- rep(NA, length(x))
+  angle <- rep(NA, length(x))
+  label <- rep(NA, length(x)) # for blinks
+  
+  
+  # Check for NA and Inf and (0,0)
+  
+  valid <- ifelse(is.na(x) | is.na(y) | 
+                    is.infinite(x) | is.infinite(y) | 
+                    (x == 0 & y == 0), F, T)
+  
+  
+  # Label blinks
+  
+  if(is.numeric(blink) && length(blink) == 2) {
+    
+    label[x == blink[1] & y == blink[2]] <- 0
+    
+  } else if(is.logical(blink) && length(blink) == length(t)) {
+    
+    label[blink] <- 0
+    
+  }
+  
+  
+  # Convert pixels to degrees of visual angle
+  
+  res.va <- c(px2va(res[1], res = res[1], dim = dim[1], dist = dist),
+              px2va(res[2], res = res[2], dim = dim[2], dist = dist))
+  
+  if(unit == "px") {
+    
+    x.va <- px2va(x, res = res[1], dim = dim[1], dist = dist)
+    y.va <- px2va(y, res = res[2], dim = dim[2], dist = dist)
+    
+  } else {
+    
+    x.va <- x
+    y.va <- y
+    
+  }
+  
+  
+  # Check if samples are on screen
+  
+  valid[valid] <- ifelse(x.va[valid] < -res.va[1] | x.va[valid] > res.va[1] | 
+                    y.va[valid] < -res.va[2] | y.va[valid] > res.va[2], F, valid[valid])
+  
+  
+  # Apply Butterworth filter
+
+  # x.va[valid] <- as.numeric(signal::filter(butter(3, W = 0.5), x = x.va[valid]))
+  # y.va[valid] <- as.numeric(signal::filter(butter(3, W = 0.5), x = y.va[valid]))
+  
+  
[…]

### 2020-04-30 algorithm/preprocessing_helper_functions.R

@@ -1,0 +1,35 @@
+### Preprocessing helper functions ###
+
+# Author: Malte Lüken
+# Date: 01.04.2020
+
+
+# Function to convert gaze coordinates from pixels into degrees of visual angle
+
+px2va <- function(x, dim, res, dist) { # x- or y-pos in px, dim in mm, res in px, dist in mm
+  
+  x_centered <- x - res/2 # transform to centered coordinate system (0,0)
+  
+  radian <- atan(x_centered/(dist*(res/dim))) # visual angle in radians
+  
+  degree <- radian*(180/pi) # visual angle in degrees
+  
+  return(degree) 
+}
+  
+
+# Function to compute the relative angle between subsequent samples 
+
+calc_theta <- function(x, y) { # x-pos, y-pos
+  
+  require(dplyr)
+  
+  diff_x <- x - lag(x) # x_t - x_t-1
+  diff_y <- y - lag(y) # y_t - y_t-1
+  
+  angle <- atan2(diff_y, diff_x) # absolute angle of vector xy_t - xy_t-1
+  theta <- lead(angle) - angle # relative angle of two vectors at sample t
+  theta.mirrored <- ifelse(theta < 0, theta + 2*pi, theta) # mirror negative relative angles to positive range
+  
+  return(theta.mirrored)
+}

### 2020-04-30 test/test.R

@@ -1,0 +1,107 @@
+### gazeHMM test ###
+
+# Author: Malte Lüken
+# Date: 27.04.2020
+
+
+library(tidyverse)
+library(depmixS4)
+source("algorithm/gazeHMM.R")
+source("algorithm/plotting_functions.R")
+
+
+# Test on static data -----------------------------------------------------
+
+# Eprep data set
+
+load("data/Eprep")
+
+dim <- c(473, 296) # screen dimensions
+res <- c(1680, 1050) # screen resolution
+dist <- 600 # distance to screen
+fr <- 500 # frame rate
+
+set.seed(123)
+
+Eprep.fit.3 <- Eprep %>%
+  dplyr::filter(block == sample(unique(Eprep$block), 1)) %>%
+  group_by(subject) %>%
+  summarise(fit = list(try(classify_gaze_data(x = xp, y = yp, t = (time - time[1])/1e3, 
+                                              unit = "px", res = res, dim = dim, fr = fr, dist = dist, 
+                                              nstates = 3)))) %>%
+  pull(fit)
+
+
+# Inspect parameters
+
+lapply(Eprep.fit.3, function(x) {summary(x$model)})
+
+
+# Plot classification result
+
+for (i in 1:length(Eprep.fit.3)) {
+  
+  set.seed(i*25)
+  
+  tp <- sample(Eprep.fit.3[[i]]$samples$t[Eprep.fit.3[[i]]$samples$t > 0.5], 1)
+  
+  int <- c(tp - 0.5, tp) # 0.5 second interval
+  
+  plot_samples_time(Eprep.fit.3[[i]], from = int[1], to = int[2]) + 
+    ggsave(filename = paste("Eprep_3comp_subj_", i, "_overtime.png", sep = ""), 
+           width = 12, height = 6)
+  
+  plot_samples_xy(Eprep.fit.3[[i]], from = int[1], to = int[2]) + 
+    ggsave(filename = paste("Eprep_3comp_subj_", i, "_xy.png", sep = ""), 
+           width = 12, height = 6)
+  
+}
+
+
+# Test on dynamic data ----------------------------------------------------
+
+# D2010 data set
+
+load("data/D2010")
+
+dim <- c(400, 300) # screen dimension
+res <- c(1280, 960) # screen resolution
+dist <- 450 # distance to screen
+fr <- 250 # frame rate
+
+set.seed(123)
+
+D2010.fit.4 <- D2010 %>%
[…]

### 2020-05-01 algorithm/gazeHMM.R
include minimum saccade duration argument

@@ -6,7 +6,8 @@
 
 classify_gaze_data <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
                                nstates, respstart, trstart, instart, sf = c(10, 10),
-                               fit.control = em.control(maxit = 5000, random.start = F)) {
+                               fit.control = em.control(maxit = 5000, random.start = F),
+                               min.sac = 0.01) {
   
   source("algorithm/preprocessing.R")
   source("algorithm/model.R")
@@ -25,7 +26,7 @@ classify_gaze_data <- function(x, y, t, unit = "px", res, dim, dist, fr, blink =
   
   # Postprocessing
   
-  post <- postprocess(prep, model.fit@posterior$state)
+  post <- postprocess(prep, model.fit@posterior$state, min.sac)
   
   
   # Output: Data frame with samples, data frames with events, fitted model

### 2020-05-01 algorithm/postprocessing.R

@@ -4,7 +4,7 @@
 # Date: 20.04.2020
 
 
-postprocess <- function(data, state) { # data frame with dependent variables, vector with state for each observation
+postprocess <- function(data, state, min.sac) { # data frame with dependent variables, vector with sample states, minimum saccade duration
   
   source("algorithm/postprocessing_helper_functions.R")
   
@@ -24,7 +24,7 @@ postprocess <- function(data, state) { # data frame with dependent variables, ve
   
   # Relabel samples
   
-  df <- reclassify(state, data$t)
+  df <- reclassify(state, data$t, min.sac)
   
   label <- df$label
   number <- df$number

### 2020-05-01 algorithm/postprocessing_helper_functions.R

@@ -6,7 +6,7 @@
 
 # Function to retrieve last (different) state
 
-last_state <- function(state, s) {
+last_state <- function(state, s) { # vector with event states, index of current event
   if(state[s-1] != state[s] || (s-1) == 1) {
     
     return(state[s-1])
@@ -19,7 +19,7 @@ last_state <- function(state, s) {
 
 # Function to relabel samples
 
-reclassify <- function(state, t) {
+reclassify <- function(state, t, min.sac) { # vector with samples states, vector with sample timestamps, minimum saccade duration
   
   # Assign labels to events
   
@@ -72,7 +72,7 @@ reclassify <- function(state, t) {
     }
     event[n] <- max(state[number == n])
     
-    if(event[n] == SAC && dur[n] < 0.01 && n > 1) {
+    if(event[n] == SAC && dur[n] < min.sac && n > 1) {
       
       state[number == n] <- last_state(event, n)
       
@@ -135,7 +135,7 @@ reclassify <- function(state, t) {
     }
     event[n] <- max(state[number == n])
     
-    if(event[n] == SAC && dur[n] < 0.01 && n > 1) {
+    if(event[n] == SAC && dur[n] < min.sac && n > 1) {
       
       minDuration <- F
       
@@ -153,7 +153,7 @@ reclassify <- function(state, t) {
     
   } else {
     
-    return(reclassify(state, t))
+    return(reclassify(state, t, min.sac))
     
   }
 }

### 2020-05-01 algorithm/gazeHMM.R
add savitzky-golay filter arguments

@@ -5,6 +5,7 @@
 
 
 classify_gaze_data <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
+                               sg.order = 3, sg.length = 5, 
                                nstates, respstart, trstart, instart, sf = c(10, 10),
                                fit.control = em.control(maxit = 5000, random.start = F),
                                min.sac = 0.01) {
@@ -16,7 +17,7 @@ classify_gaze_data <- function(x, y, t, unit = "px", res, dim, dist, fr, blink =
   
   # Preprocessing
   
-  prep <- preprocess(x, y, t, unit, res, dim, dist, fr, blink)
+  prep <- preprocess(x, y, t, unit, res, dim, dist, fr, blink, sg.order, sg.length)
   
   
   # Model classification

### 2020-05-01 algorithm/preprocessing.R

@@ -4,9 +4,8 @@
 # Date: 07.04.2020
 
 
-preprocess <- function(x, y, t, unit = "px",
-                       res, dim, dist,
-                       fr, blink = NULL) {
+preprocess <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
+                       sg.order, sg.length) {
   
   require(signal)
   source("algorithm/preprocessing_helper_functions.R")
@@ -72,16 +71,12 @@ preprocess <- function(x, y, t, unit = "px",
   
   # Calculate velocity and acceleration
   
-  x.vel <- sgolayfilt(x.va[valid], m = 1)
-  y.vel <- sgolayfilt(y.va[valid], m = 1)
-  # x.vel <- lead(x.va[valid]) - lag(x.va[valid])
-  # y.vel <- lead(y.va[valid]) - lag(y.va[valid])
+  x.vel <- sgolayfilt(x.va[valid], m = 1, p = sg.order, n = sg.length)
+  y.vel <- sgolayfilt(y.va[valid], m = 1, p = sg.order, n = sg.length)
   vel[valid] <- sqrt(x.vel^2 + y.vel^2) * fr
   
-  x.acc <- sgolayfilt(x.va[valid], m = 2)
-  y.acc <- sgolayfilt(y.va[valid], m = 2)
-  # x.acc <- (lead(x.va[valid]) - x.va[valid]) - (lag(x.va[valid] - x.va[valid]))
-  # y.acc <- (lead(y.va[valid]) - y.va[valid]) - (lag(y.va[valid] - y.va[valid]))
+  x.acc <- sgolayfilt(x.va[valid], m = 2, p = sg.order, n = sg.length)
+  y.acc <- sgolayfilt(y.va[valid], m = 2, p = sg.order, n = sg.length)
   acc[valid] <- sqrt(x.acc^2 + y.acc^2) * fr
   
   valid[valid] <- ifelse(is.na(vel[valid]) | is.na(acc[valid]), F, valid[valid])

### 2020-05-04 algorithm/model_helper_functions.R
maximize positive parameters on log-scale

@@ -179,7 +179,7 @@ setMethod("vMF",
               if(length(pstart) != npar) stop("length of 'pstart' must be ", npar)
               
               parameters$mu <- pstart[1]
-              parameters$kappa <- pstart[2]
+              parameters$kappa <- log(pstart[2])
             }
             
             mod <- new("vMF", parameters = parameters, fixed = fixed, x = x, y = y, npar = npar)
@@ -206,7 +206,7 @@ setMethod("show", "vMF",
 setMethod("dens","vMF",
           function(object) {
             
-            dens <- ifelse(is.na(object@y), 1, dvm(object@y, mu = object@parameters$mu, kappa = object@parameters$kappa))
+            dens <- ifelse(is.na(object@y), 1, dvm(object@y, mu = object@parameters$mu, kappa = exp(object@parameters$kappa)))
             
           }
 )
@@ -279,7 +279,7 @@ setMethod("fit","vMF",
               
               miss <- is.na(y)
               
-              dens <- ifelse(miss, 1, dvm(y, mu = par[1], kappa = par[2]))
+              dens <- ifelse(miss, 1, dvm(y, mu = par[1], exp(kappa = par[2])))
               
               return(-sum(w*log(dens)))
             }
@@ -290,10 +290,9 @@ setMethod("fit","vMF",
             # Optimize weighted ll for von-Mises distribution
             
             fit <- BB::BBoptim(par = init, fn = ll_vMF, y = y, w = w,
-                               lower = c(0, 1e-5), upper = c(pi, Inf), quiet = T)
+                               lower = c(-Inf, -Inf), upper = c(Inf, Inf), quiet = T)
             
             pars <- fit$par
-            names(pars) <- c("mu", "kappa")
             object <- setpars(object,pars)
             
             return(object)
@@ -348,8 +347,8 @@ setMethod("altGamma",
             if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
             if(!is.null(pstart)) {
               if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
-              parameters$shape <- pstart[1]
-              parameters$scale <- pstart[2]
+              parameters$shape <- log(pstart[1])
+              parameters$scale <- log(pstart[2])
             }
             mod <- new("altGamma",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
             mod
@@ -368,8 +367,8 @@ setMethod("show","altGamma",
 setMethod("dens","altGamma",
           function(object,log=FALSE) {
             
[…]

### 2020-05-04 test/test.R
add bivariate plots

@@ -58,6 +58,28 @@ for (i in 1:length(Eprep.fit.3)) {
 }
 
 
+# Plot bivariate classification 
+
+for (i in 1:length(Eprep.fit.3)) {
+  
+  ggplot(Eprep.fit.3[[i]]$samples, aes(vel, acc, color = as.factor(state))) + 
+    geom_point() + scale_color_discrete(name = "state", labels = c("Fix", "Sac", "PSO")) +
+    ggsave(filename = paste("Eprep_3comp_subj_", i, "_bivariate_vel_acc.png", sep = ""), 
+           width = 6, height = 6)
+  
+  ggplot(Eprep.fit.3[[i]]$samples, aes(vel, angle, color = as.factor(state))) + 
+    geom_point() + scale_color_discrete(name = "state", labels = c("Fix", "Sac", "PSO")) +
+    ggsave(filename = paste("Eprep_3comp_subj_", i, "_bivariate_vel_angle.png", sep = ""), 
+           width = 6, height = 6)
+  
+  ggplot(Eprep.fit.3[[i]]$samples, aes(acc, angle, color = as.factor(state))) + 
+    geom_point() + scale_color_discrete(name = "state", labels = c("Fix", "Sac", "PSO")) +
+    ggsave(filename = paste("Eprep_3comp_subj_", i, "_bivariate_acc_angle.png", sep = ""), 
+           width = 6, height = 6)
+  
+}
+
+
 # Test on dynamic data ----------------------------------------------------
 
 # D2010 data set
@@ -104,4 +126,26 @@ for (i in 1:length(D2010.fit.4)) {
   plot_samples_xy(D2010.fit.4[[i]], from = int[1], to = int[2]) + 
     ggsave(filename = paste("D2010_4comp_subj_", i, "_xy.png", sep = ""), 
            width = 12, height = 6)
+}
+
+
+# Plot bivariate classification
+
+for (i in 1:length(D2010.fit.4)) {
+  
+  ggplot(D2010.fit.4[[i]]$samples, aes(vel, acc, color = as.factor(state))) + 
+    geom_point() + scale_color_discrete(name = "state", labels = c("Fix", "Sac", "PSO")) +
+    ggsave(filename = paste("D2010_4comp_subj_", i, "_bivariate_vel_acc.png", sep = ""), 
+           width = 6, height = 6)
+  
+  ggplot(D2010.fit.4[[i]]$samples, aes(vel, angle, color = as.factor(state))) + 
[…]

### 2020-05-05 simulation/preregistration_parameter_recovery.Rmd
upload preregistration draft

---
title: "gazeHMM: Parameter recovery simulation"
author: "Malte Lüken, Simon Kucharsky, & Ingmar Visser"
date: "5 5 2020"
output: pdf_document
---

In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM. The purpose of this algorithm is to classify eye movements into distinct events (e.g., fixations). The full algorithm will be available on Github upon publication.

In general, gazeHMM consists of three steps: First, the raw gaze data is preprocessed. Missing samples are identified, the data are filtered, and for each sample the velocity, acceleration, and difference in angle to the next sample is calculated. Second, the HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events. Moreover, the HMM computes the posterior probability of each sample for belonging to the events. Each sample is labeled as the event with the highest posterior probability. Third, a postprocessing routine relabels samples that are violating theoretical and practical constraints. Since the HMM might not model the data perfectly, this step improves classification performance. This simulation study will only consider the parameter recovery of the HMM but not the performance of the preprocessing and postprocessing routines of the algorithm.

# Model description

The HMM estimates unobservable states that have generated the data. In the context of eye-tracking, each state corresponds to an eye movement event. In gazeHMM, the model can have between two and four states. State one always corresponds to fixations, state two always to saccades, state three to postsaccadic oscillations (PSOs), and state four to smooth pursuits. Moreover, the HMM is multivariate but conditionally independent on the states. Thus, it has three independent response models for every state describing velocity, acceleration, and angle. The velocity and acceleration response models for all four states use two-parameter gamma distributions (shape and scale parametrization). The angle response model for the fixation state uses a uniform distribution, while the other three states use a two-parameter von-Mises distribution (mean and concentration parametrization) to model angle. Both the transition model and the initial state model use a multinomial distribution (with the logit and the identity link function, respectively). Note that no covariates are included in the model, only intercepts for the parameters are estimated.

Depending on the number of states, the HMM has a different number of parameters. The response models of the HMM contain six parameters per state. The transition models have $k^k$ parameters in total with k being the number of states. The initial state models have one paramter per state. In sum, HMMs with two, three, and four states have X, X, and X paramters. The denotation of each parameter is shown in Table 1.

*Optimization*

Table 1

|Parameter|Description|
|:--------|:----------|
|$\pi$   |Initial probabilities for state *i*|
|$a_{ij}$ |Transition probability from state *i* to *j*|
|$

# Model parameter recovery

The parameter recovery study for the HMM is divided into two parts: In the first part, we will vary the parameters of the model and estimate the recovery. In the second part, the parameters of the model will be fixed. Instead, we will manipulate the size of the sample generated by the model level, the noise added to the generated data, the starting values for the HMM, and the amount of missing data.

## Variable parameters

First, we will repeatedly generate data with the HMM and the parameters displayed in Table 2. Second, the HMM estimates the parameters from the generated data. Third, we will compare the estimated parameters with the true parameters. We will treat the parameter recovery as satisfactory when the 95% confidence bands include the true parameter values.

Table 2

|State|Parameter|Min.|Max.|Fixed|
|-----|---------|----|----|-----|
|1-4|$\pi$|-|-|$1/k$|
|1-4|$a_{i=j}$|0.01|0.99|0.9|
|1-4|$a_{i \neq j}$|0.01|0.99|0.1|
|1|$\alpha_{vel}$|1|5|3|
|1|$\beta_{vel}$|0.5|1|0.75|
|1|$\alpha_{acc}$|1|5|3|
|1|$\beta_{acc}$|0.05|0.5|0.15|
|2|$\alpha_{vel}$|1|5|3|
|2|$\beta_{vel}$|5|15|10|
|2|$\alpha_{acc}$|1|5|3|
|2|$\beta_{acc}$|1|5|3|
|2|$\mu$|-|-|0|
|2|$\kappa$|0.1|10|1|
|3|$\alpha_{vel}$|1|5|3|
|3|$\beta_{vel}$|1|5|3|
|3|$\alpha_{acc}$|1|3|2|
|3|$\beta_{acc}$|1|3|2|
|3|$\mu$|-|-|$\pi$|
|3|$\kappa$|0.1|10|1|
|4|$\alpha_{vel}$|1|5|3|
|4|$\beta_{vel}$|1|2|1.5|
|4|$\alpha_{acc}$|1|5|3|
|4|$\beta_{acc}$|0.05|0.25|0.15|
|4|$\mu$|-|-|0|
|4|$\kappa$|0.1|10|1|

## Scenario 2: Sample size and noise

In this scenario, the parameters of the HMM are fixed to the values shown in Table 2. For HMMs with two, three, and four states, we will repeatedly generate data with the model for 500, 2500, and 10.000 samples. In addition, gaussian white noise with paramters $\sigma_{vel; acc} \in \{1, 5, 25\}$ and $\kappa_{angle} \in \{0.1, 1, 10\}$ will be added to the data. This will results in 27 ($5 \times 3 \times 3$) conditions with *D* repetitions each.

##

### 2020-05-06 simulation/preregistration_parameter_recovery.Rmd
add simulation method for gamma distribution

@@
 
-The parameter recovery study for the HMM is divided into two parts: In the first part, we will vary the parameters of the model and estimate the recovery.
-In the second part, the parameters of the model will be fixed.
-Instead, we will manipulate the size of the sample generated by the model level, the noise added to the generated data, the starting values for the HMM, and the amount of missing data.
+To estimate parameter recovery of the HMM, we repeatedly generate data with the model under different conditions and a different number of states.
+For every condition and number of states $k \in \{2, 3, 4\}$, the model will generate $D = 100$ data sets with parameters evenly distributed on a given interval.
+The same model is then applied to estimate the parameters from the generated data.
+Afterwards, the estimated parameters are compared with the "true" parameters that generated the data.
+We treat a parameter as successfully recovered when the 95% confidence bands of the estimated parameters include the true parameters.
+We further require a 90% median rate of correct classifications across the data sets for the model to be satisfactorily accurate.
 
-## Variable parameters
+The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilites.
+To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$.
+Mean parameters of the von-Mises distribution will always start at their true value.
 
-First, we will repeatedly generate data with the HMM and the parameters displayed in Table 2.
-Second, the HMM estimates the parameters from the generated data.
-Third, we will compare the estimated parameters with the true parameters.
-We will treat the parameter recovery as satisfactory when the 95% confidence bands include the true parameter values.
+In the first part of the simulation study, we will estimate parameter recovery under ideal condititions.
+These imply an average sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples.
+The parameters that we will use to generate the data are displayed in Table 2.
+We will vary one parameter at a time on the given interval and set the other parameters to their fixed value.
+This simulation will result in $D_{total} = 100 \times 3 \times 20 = 6000$ recoveries.
+
+In the second part, we will vary the sample size of the generated data and the amount of noise added to it.
+For sample sizes of $N \in \{500, 2500, 10000\}$, we will generate data sets and add white noise.
+To velocity and acceleration, we will add gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$.
+Noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ will be added to angle.
+This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.
+
+In the third part, we will increase the variation in the starting values used for parameter estimation.
+For the shape, scale, and concentration parameters, we will increase the scale parameters of the gamma distributions to $\beta \in \{2, 4, 8\}$ all at the same time.
+The sample size will be $N = 2500$ and no noise will be added to the data.
+This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.
+
+In the last part, we will set intervals of the generated data to be missing.
+The size of the missing data interval will be $m \in [1,100]$ and the number of intervals will be $n_{miss} \in \{1, 3, 5\}$.
+The sample size will be $N = 2500$ and no noise will be added to the data.
+This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.
 
@@
 
-|State|Parameter|Min.|Max.|Fixed|
-|-----|---------|----|----|-----|
-|1-4|$\pi$|-|-|$1/k$|
-|1-4|$a_{i=j}$|0.01|0.99|0.9|
-|1-4|$a_{i \neq j}$|0.01|0.99|0.1|
-|1|$\alpha_{vel}$|1|5|3|
-|1|$\beta_{vel}$|0.5|1|0.75|
-|1|$\alpha_{acc}$|1|5|3|
-|1|$\beta_{acc}$|0.05|0.5|0.15|
-|2|$\alpha_{vel}$|1|5|3|
-|2|$\beta_{vel}$|5|15|10|
-|2|$\alpha_{acc}$|1|5|3|
-|2|$\beta_{acc}$|1|5|3|
-|2|$\mu$|-|-|0|
-|2|$\kappa$|0.1|10|1|
-|3|$\alpha_{vel}$|1|5|3|
-|3|$\beta_{vel}$|1|5|3|
-|3|$\alpha_{acc}$|1|3|2|
-|3|$\beta_{acc}$|1|3|2|
-|3|$\mu$|-|-|$\pi$|
-|3|$\kappa$|0.1|10|1|
-|4|$\alpha_{vel}$|1|5|3|
-|4|$\beta_{vel}$|1|2|1.5|
-|4|$\alpha_{acc}$|1|5|3|
-|4|$\beta_{acc}$|0.05|0.25|0.15|
-|4|$\mu$|-|-|0|
-|4|$\kappa$|0.1|10|1|
+|State|Parameter|Interval|Fixed|
+|-----|---------|--------|-----|
+|1-4|$p$*|-|$1/k$|
+|1-4|$a_{i=j}$|[.01,.99]|0.9|
+|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|
+|1|$\alpha_{vel}$|[1,5]|3|
+|1|$\beta_{vel}$|[0.5,1]|0.75|
+|1|$\alpha_{acc}$|[1,5]|3|
+|1|$\beta_{acc}$|[0.05,0.5]|0.15|
+|1|$a$*|-|-|
+|1|$b$*|-|-|
+|2|$\alpha_{vel}$|[1,5]|3|
+|2|$\beta_{vel}$|[5,15]|10|
+|2|$\alpha_{acc}$|[1,5]|3|
+|2|$\beta_{acc}$|[1,5]|3|
+|2|$\mu$*|-|0|
+|2|$\kappa$|$1/[0.1,10]$|1|
+|3|$\alpha_{vel}$|[1,5]|3|
+|3|$\beta_{vel}$|[1,5]|3|
+|3|$\alpha_{acc}$|[1,3]|2|
+|3|$\beta_{acc}$|[1,3]|2|
+|3|$\mu$*|-|$\pi$|
+|3|$\kappa$|$1/[0.1,10]$|1|
+|4|$\alpha_{vel}$|[1,5]|3|
+|4|$\beta_{vel}$|[1,2]|1.5|
+|4|$\alpha_{acc}$|[1,5]|3|
+|4|$\beta_{acc}$|[0.05,0.25]|0.15|
+|4|$\mu$*|-|0|
+|4|$\kappa$|$1/[0.1,10]$|1|
 
-## Scenario 2: Sample size and noise
-
-In this scenario, the parameters of the HMM are fixed to the values shown in Table 2.
-For HMMs with two, three, and four states, we will repeatedly generate data with the model for 500, 2500, and 10.000 samples.
-In addition, gaussian white noise with paramters $\sigma_{vel; acc} \in \{1, 5, 25\}$ and $\kappa_{angle} \in \{0.1, 1, 10\}$ will be added to the data.
-This will results in 27 ($5 \times 3 \times 3$) conditions with *D* repetitions each.
-
-##
+*Note*.
+Parameters marked with * will not be varied but always set to their fixed values.
+Parameters *a* and *b* (the minimum and maximum of the uniform distribution) are always obtained from the data.

### 2020-05-06 algorithm/model_helper_functions.R

@@ -5,8 +5,6 @@
 
 library(depmixS4)
 library(CircStats)
-library(gamlss)
-library(gamlss.dist)
 
 
 # Create response model for uniform distribution --------------------------
@@ -452,5 +450,21 @@ setMethod("predict","altGamma",
           function(object) {
             ret <- object@parameters$shape*object@parameters$scale
             return(ret)
+          }
+)
+
+
+# Define method to simulate
+
+setMethod("simulate", signature(object = "altGamma"),
+          function(object, nsim = 1, seed) {
+            
+            if(!is.null(seed)) set.seed(seed)
+            
+            nt <- nrow(object@y)
+            
+            sim <- rgamma(nt*nsim, shape = object@parameters$shape, scale = object@parameters$scale)
+            
+            return(as.matrix(sim))
           }
 )

### 2020-05-06 simulation/preregistration_parameter_recovery.Rmd
add parameter recovery script

@@
 title: "gazeHMM: Parameter recovery simulation"
-author: "Malte Lüken, Simon Kucharsky, & Ingmar Visser"
-date: "5 5 2020"
+author: "Malte Lüken, Simon Kucharský, & Ingmar Visser"
+affiliation: "University of Amsterdam"
+date: "05.05.2020"
 output: pdf_document
@@
 In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM.
-The purpose of this algorithm is to classify eye movements into distinct events (e.g., fixations).
-The full algorithm will be available on Github upon publication.
+The purpose of this algorithm is to classify eye movements into distinct events (e.g., fixations and saccades).
+The full algorithm will be available on Git hub upon publication.
 
@@
 Third, a postprocessing routine relabels samples that are violating theoretical and practical constraints.
-Since the HMM might not model the data perfectly, this step improves classification performance.
 This simulation study will only consider the parameter recovery of the HMM but not the performance of the preprocessing and postprocessing routines of the algorithm.
 
-# Model description
+# Model Description
 
-The HMM estimates unobservable states that have generated the data.
+The HMM estimates non-observable states that have generated the data.
 In the context of eye-tracking, each state corresponds to an eye movement event.
@@
 
-Depending on the number of states, the HMM has a different number of parameters.
-The response models of the HMM contain six parameters per state.
-The transition models have $k^k$ parameters in total with k being the number of states.
-The initial state models have one paramter per state.
-In sum, HMMs with two, three, and four states have X, X, and X paramters.
-The denotation of each parameter is shown in Table 1.
+The HMM uses an expectation-maximization algorithm to estimate parameters under the hidden states.
+The parameters of each response model (except for the uniform distribution) are estimated with a spectral projected gradient method and Barzilai-Borwein step lengths.
 
-*Optimization*
-
-Table 1
-
-|Parameter|Description|
-|:--------|:----------|
-|$\pi$   |Initial probabilities for state *i*|
-|$a_{ij}$ |Transition probability from state *i* to *j*|
-|$
-
-# Model parameter recovery
+# Model Parameter Recovery
 
@@
 
-The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilites.
+The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilities.
 To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$.
@@
 
-In the first part of the simulation study, we will estimate parameter recovery under ideal condititions.
+In the first part of the simulation study, we will estimate parameter recovery under ideal conditions.
 These imply an average sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples.
-The parameters that we will use to generate the data are displayed in Table 2.
+The parameters that we will use to generate the data are displayed in Table 1.
 We will vary one parameter at a time on the given interval and set the other parameters to their fixed value.
-This simulation will result in $D_{total} = 100 \times 3 \times 20 = 6000$ recoveries.
+For two states, 10 parameters will be varied one at a time, for three states 15, and for four states 20 parameters.
+This simulation will result in $D_{total} = 100 \times (10+15+20) = 4500$ recoveries.
 
 In the second part, we will vary the sample size of the generated data and the amount of noise added to it.
+The model parameters will be set to the fixed values from Table 1.
 For sample sizes of $N \in \{500, 2500, 10000\}$, we will generate data sets and add white noise.
-To velocity and acceleration, we will add gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$.
+To velocity and acceleration, we will add Gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$.
 Noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ will be added to angle.
@@
 In the third part, we will increase the variation in the starting values used for parameter estimation.
+The model parameters will be set to the fixed values from Table 1.
 For the shape, scale, and concentration parameters, we will increase the scale parameters of the gamma distributions to $\beta \in \{2, 4, 8\}$ all at the same time.
@@
 In the last part, we will set intervals of the generated data to be missing.
+The model parameters will be set to the fixed values from Table 1.
 The size of the missing data interval will be $m \in [1,100]$ and the number of intervals will be $n_{miss} \in \{1, 3, 5\}$.
@@
 
-Table 2
+Table 1
 
-|State|Parameter|Interval|Fixed|
-|-----|---------|--------|-----|
-|1-4|$p$*|-|$1/k$|
-|1-4|$a_{i=j}$|[.01,.99]|0.9|
-|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|
-|1|$\alpha_{vel}$|[1,5]|3|
-|1|$\beta_{vel}$|[0.5,1]|0.75|
-|1|$\alpha_{acc}$|[1,5]|3|
-|1|$\beta_{acc}$|[0.05,0.5]|0.15|
-|1|$a$*|-|-|
-|1|$b$*|-|-|
-|2|$\alpha_{vel}$|[1,5]|3|
-|2|$\beta_{vel}$|[5,15]|10|
-|2|$\alpha_{acc}$|[1,5]|3|
-|2|$\beta_{acc}$|[1,5]|3|
-|2|$\mu$*|-|0|
-|2|$\kappa$|$1/[0.1,10]$|1|
-|3|$\alpha_{vel}$|[1,5]|3|
-|3|$\beta_{vel}$|[1,5]|3|
-|3|$\alpha_{acc}$|[1,3]|2|
-|3|$\beta_{acc}$|[1,3]|2|
-|3|$\mu$*|-|$\pi$|
-|3|$\kappa$|$1/[0.1,10]$|1|
-|4|$\alpha_{vel}$|[1,5]|3|
-|4|$\beta_{vel}$|[1,2]|1.5|
-|4|$\alpha_{acc}$|[1,5]|3|
-|4|$\beta_{acc}$|[0.05,0.25]|0.15|
-|4|$\mu$*|-|0|
-|4|$\kappa$|$1/[0.1,10]$|1|
+|State|Parameter|Interval|Fixed|Description|
+|-----|---------|--------|-----|-----------|
+|1-4|$p_i^+$|-|$1/k$|Initial state probability for starting in state *i*|
+|1-4|$a_{i=j}$|[.01,.99]|0.9|Transition probability for staying in the previous state *i*|
+|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|Transition probability for switching to from state *i* to a different state *j*|
+|1|$\alpha_{vel}$|[1,5]|3|Shape parameter of the velocity gamma distribution|
+|1|$\beta_{vel}$|[0.1,0.6]|0.35|Scale parameter of the velocity gamma distribution|
+|1|$\alpha_{acc}$|[1,5]|3|Shape parameter of the acceleration gamma distribution|
+|1|$\beta_{acc}$|[0.05,0.25]|0.15|Scale parameter of the acceleration gamma distribution|
+|1|$a^+$|-|0|Minimum of the uniform distribution|
+|1|$b^+$|-|$2\pi$|Maximum of the uniform distribution|
+|2|$\alpha_{vel}$|[1,5]|3||
+|2|$\beta_{vel}$|[5,25]|15||
+|2|$\alpha_{acc}$|[1,5]|3||
+|2|$\beta_{acc}$|[1,5]|3||
+|2|$\mu^+$|-|0|Mean parameter of the von-Mises distribution|
+|2|$\kappa$|$1/[0.1,10]$|1|Concentration parameter of the von Mises distribution
+|3|$\alpha_{vel}$|[1,5]|3||
+|3|$\beta_{vel}$|[1,5]|3||
+|3|$\alpha_{acc}$|[1,3]|2||
+|3|$\beta_{acc}$|[1,3]|2||
+|3|$\mu^+$|-|$\pi$||
+|3|$\kappa$|$1/[0.1,10]$|1||
+|4|$\alpha_{vel}$|[1,5]|3||
+|4|$\beta_{vel}$|[1,2]|1.5||
+|4|$\alpha_{acc}$|[1,5]|3||
+|4|$\beta_{acc}$|[0.05,0.25]|0.15||
+|4|$\mu^+$|-|0||
+|4|$\kappa$|$1/[0.1,10]$|1||
 
-*Note*.
-Parameters marked with * will not be varied but always set to their fixed values.
-Parameters *a* and *b* (the minimum and maximum of the uniform distribution) are always obtained from the data.
+*Note*. Parameters marked with $^+$ will not be varied but always set to their fixed value. *k* is the number of states in the model.

### 2020-05-06 algorithm/model_helper_functions.R

@@ -319,7 +319,7 @@ setMethod("simulate", signature(object = "vMF"),
             
             nt <- nrow(object@y)
             
-            sim <- rvm(nt*nsim, mean = object@parameters$mu, k = object@parameters$kappa)
+            sim <- rvm(nt*nsim, mean = object@parameters$mu, k = exp(object@parameters$kappa))
             
             return(as.matrix(sim))
           }
@@ -463,7 +463,7 @@ setMethod("simulate", signature(object = "altGamma"),
             
             nt <- nrow(object@y)
             
-            sim <- rgamma(nt*nsim, shape = object@parameters$shape, scale = object@parameters$scale)
+            sim <- rgamma(nt*nsim, shape = exp(object@parameters$shape), scale = exp(object@parameters$scale))
             
             return(as.matrix(sim))
           }

### 2020-05-06 simulation/model_simulation.R

@@ -1,0 +1,53 @@
+### Simulation from hidden Markov model ###
+
+# Author: Malte Lüken
+# Date: 06.05.2020
+
+HMM_simulate <- function(n, nstates, respstart, trstart, instart) {
+  
+  require(depmixS4)
+  source("algorithm/model_helper_functions.R")
+  
+  
+  data <- data.frame(vel = numeric(n),
+                     acc = numeric(n),
+                     angle = seq(0, 2*pi, length.out = n))
+  
+  
+  # Create response model
+  
+  resp <- list(list(altGamma(data$vel, pstart = respstart[["fix"]][["vel"]]),
+                    altGamma(data$acc, pstart = respstart[["fix"]][["acc"]]),
+                    unif(data$angle)))
+  
+  for (s in 2:nstates) {
+    
+    resp[[s]] <- list(altGamma(data$vel, pstart = respstart[[s]][["vel"]]),
+                      altGamma(data$acc, pstart = respstart[[s]][["acc"]]),
+                      vMF(data$angle, pstart = respstart[[s]][["angle"]]))
+    
+  }
+  
+  
+  # Create state transition model
+  
+  trans <- list()
+  
+  for (s in 1:nstates) {
+    
+    trans[[s]] <- transInit(~ 1, nstates = nstates, data = data, pstart = c(trstart[s,]))
+    
+  }
+  
+  
+  # Create initial state model
+  
+  init <- transInit(~ 1, ns = nstates, ps = instart, family = multinomial("identity"), data = data.frame(1))
+  
+  
+  # Combine models
+  
+  model <- makeDepmix(response = resp, transition = trans, prior = init, homogeneous = F)
+  
+  return(model)
+}

### 2020-05-06 simulation/parameter_recovery.R

@@ -1,0 +1,106 @@
+### Hidden Markov model parameter recovery ###
+
+# Author: Malte Lüken
+# Date: 06.05.2020
+
+library(depmixS4)
+library(parallel)
+source("simulation/model_simulation.R")
+
+
+# Part 1 ------------------------------------------------------------------
+
+D <- 100
+N <- 2500
+K <- 2:4
+
+aij <- seq(0.01, 0.99, length.out = D)
+
+alpha.vel <- matrix(NA, D, max(K))
+beta.vel <- matrix(NA, D, max(K))
+alpha.acc <- matrix(NA, D, max(K))
+beta.acc <- matrix(NA, D, max(K))
+
+alpha.vel[,1] <- seq(1, 5, length.out = D)
+beta.vel[,1] <- seq(0.1, 0.6, length.out = D)
+alpha.acc[,1] <- seq(1, 5, length.out = D)
+beta.acc[,1] <- seq(0.05, 0.25, length.out = D)
+
+alpha.vel[,2] <- seq(1, 5, length.out = D)
+beta.vel[,2] <- seq(5, 25, length.out = D)
+alpha.acc[,2] <- seq(1, 5, length.out = D)
+beta.acc[,2] <- seq(1, 5, length.out = D)
+
+alpha.vel[,3] <- seq(1, 5, length.out = D)
+beta.vel[,3] <- seq(1, 5, length.out = D)
+alpha.acc[,3] <- seq(1, 5, length.out = D)
+beta.acc[,3] <- seq(1, 3, length.out = D)
+
+alpha.vel[,4] <- seq(1, 5, length.out = D)
+beta.vel[,4] <- seq(1, 5, length.out = D)
+alpha.acc[,4] <- seq(1, 3, length.out = D)
+beta.acc[,4] <- seq(1, 3, length.out = D)
+
+kappa <- 1/seq(0.1, 10, length.out = D)
+
+alpha.vel.fix <- c(3, 3, 3, 3)
+beta.vel.fix <- c(0.35, 15, 3, 1.5)
+alpha.acc.fix <- c(3, 3, 3, 3)
+beta.acc.fix <- c(0.15, 3, 2, 0.15)
+
+mu.fix <- c(0, pi, 0)
+kappa.fix <- 1
+
+par.true <- cbind(seq(1, 5, length.out = D),
+                  seq(0.1, 0.6, length.out = D),
+                  seq(1, 5, length.out = D),
[…]

### 2020-05-07 simulation/preregistration_parameter_recovery.Rmd
update parameter recovery

@@
 title: "gazeHMM: Parameter recovery simulation"
-author: "Malte Lüken, Simon Kucharský, & Ingmar Visser"
+author: "Malte Lüken, Simon Kucharsky, & Ingmar Visser"
 affiliation: "University of Amsterdam"
-date: "05.05.2020"
+date: "`r format(Sys.Date(), '%d.%m.%Y')`"
 output: pdf_document
@@
 In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM.
-The purpose of this algorithm is to classify eye movements into distinct events (e.g., fixations and saccades).
-The full algorithm will be available on Git hub upon publication.
+The purpose of this algorithm is to classify eye movements into distinct events.
+These events include fixations, saccades, and optionally postsaccadic oscillations (PSOs) and smooth pursuits.
+The full algorithm will be available on GitHub (www.github.com/maltelueken/gazeHmm) upon publication.
 
@@
 Second, the HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events.
-Moreover, the HMM computes the posterior probability of each sample for belonging to the events.
-Each sample is labeled as the event with the highest posterior probability.
+Moreover, the HMM computes the posterior probability of each sample for belonging to each event.
+The samples are labeled as the event with the highest posterior probability.
 Third, a postprocessing routine relabels samples that are violating theoretical and practical constraints.
-This simulation study will only consider the parameter recovery of the HMM but not the performance of the preprocessing and postprocessing routines of the algorithm.
+This simulation study will only consider the parameter recovery of the HMM but not the performance of the preprocessing and postprocessing routines in the algorithm.
 
@@
 In gazeHMM, the model can have between two and four states.
-State one always corresponds to fixations, state two always to saccades, state three to postsaccadic oscillations (PSOs), and state four to smooth pursuits.
+State one always corresponds to fixations, state two always to saccades, state three to PSOs, and state four to smooth pursuits.
 Moreover, the HMM is multivariate but conditionally independent on the states.
@@
 The velocity and acceleration response models for all four states use two-parameter gamma distributions (shape and scale parametrization).
-The angle response model for the fixation state uses a uniform distribution, while the other three states use a two-parameter von-Mises distribution (mean and concentration parametrization) to model angle.
+The angle response model for the fixation state uses a uniform distribution, while the other three states use a two-parameter von-Mises distribution (mean and concentration parametrization) to describe angle.
 Both the transition model and the initial state model use a multinomial distribution (with the logit and the identity link function, respectively).
@@
 
-The HMM uses an expectation-maximization algorithm to estimate parameters under the hidden states.
-The parameters of each response model (except for the uniform distribution) are estimated with a spectral projected gradient method and Barzilai-Borwein step lengths.
+The HMM is implemented in R (CITE) using the package depmixS4 (CITE).
+An expectation-maximization algorithm estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE).
+The parameters of each response model (except for the uniform distribution) are estimated through maximum likelihood with a spectral projected gradient method (CITE) and Barzilai-Borwein step lengths (CITE) as implemented in the package BB (CITE).
 
-# Model Parameter Recovery
+# Parameter Recovery
 
-To estimate parameter recovery of the HMM, we repeatedly generate data with the model under different conditions and a different number of states.
-For every condition and number of states $k \in \{2, 3, 4\}$, the model will generate $D = 100$ data sets with parameters evenly distributed on a given interval.
+To estimate parameter recovery of the HMM, we repeatedly generate data with the model under different conditions and different numbers of states.
+For every condition and number of states $k \in \{2, 3, 4\}$, the model will generate $D = 100$ data sets with parameters varied in a specified interval in equidistant steps.
 The same model is then applied to estimate the parameters from the generated data.
-Afterwards, the estimated parameters are compared with the "true" parameters that generated the data.
-We treat a parameter as successfully recovered when the 95% confidence bands of the estimated parameters include the true parameters.
-We further require a 90% median rate of correct classifications across the data sets for the model to be satisfactorily accurate.
+Then, the estimated parameters are compared with the "true" parameters that generated the data.
+We will assess parameter recovery descriptively by computing the mean squared error between the estimated and true parameter values separately for each parameter.
+Additionally, we will apply a bivariate linear regression for each parameter with the true parameter values as the independent and the estimated parameter values as the dependent variable.
+For a parameter to be recovered, we expect the regression intercepts to be close to zero and the regression slopes to be close to one.
+Next to parameter recovery, we will also examine the accuracy of the classficiation through the model.
+Therefore, we will compare the true states with the estimated states for each data set by using Cohen's Kappa.
+We will treat Kappa values above 0.8 as satisfactorily accurate.
 
@@
 In the first part of the simulation study, we will estimate parameter recovery under ideal conditions.
-These imply an average sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples.
+These imply a medium sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples.
 The parameters that we will use to generate the data are displayed in Table 1.
-We will vary one parameter at a time on the given interval and set the other parameters to their fixed value.
-For two states, 10 parameters will be varied one at a time, for three states 15, and for four states 20 parameters.
+We will vary one parameter at a time on the given interval while setting the other parameters to their fixed value.
+The initial state probabilities will always stay at their fixed value.
+We will only vary the transition probabilities for staying in the same state.
+The remaining probability mass will be evenly distributed across the probabilities for switching to a different state (see Table 1).
+For two states 10, for three states 15, and for four states 20 parameters will be varied one at a time.
 This simulation will result in $D_{total} = 100 \times (10+15+20) = 4500$ recoveries.
@@
 Table 1
+
+*HMM parameter values for generating the data*
 

### 2020-05-07 simulation/parameter_recovery.R

@@ -6,90 +6,60 @@
 library(depmixS4)
 library(parallel)
 source("simulation/model_simulation.R")
+source("algorithm/model_helper_functions.R")
 
 
 # Part 1 ------------------------------------------------------------------
 
-D <- 100
+D <- 5
 N <- 2500
 K <- 2:4
 
 aij <- seq(0.01, 0.99, length.out = D)
 
-alpha.vel <- matrix(NA, D, max(K))
-beta.vel <- matrix(NA, D, max(K))
-alpha.acc <- matrix(NA, D, max(K))
-beta.acc <- matrix(NA, D, max(K))
+par.int.true <- cbind(seq(1, 5, length.out = D),
+                      seq(0.1, 0.6, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(0.05, 0.25, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(5, 25, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(1, 5, length.out = D),
+                      1/seq(0.1, 10, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(1, 3, length.out = D),
+                      1/seq(0.1, 10, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(1, 5, length.out = D),
+                      seq(1, 3, length.out = D),
+                      seq(1, 3, length.out = D),
+                      1/seq(0.1, 10, length.out = D))
 
-alpha.vel[,1] <- seq(1, 5, length.out = D)
-beta.vel[,1] <- seq(0.1, 0.6, length.out = D)
-alpha.acc[,1] <- seq(1, 5, length.out = D)
-beta.acc[,1] <- seq(0.05, 0.25, length.out = D)
 
-alpha.vel[,2] <- seq(1, 5, length.out = D)
-beta.vel[,2] <- seq(5, 25, length.out = D)
-alpha.acc[,2] <- seq(1, 5, length.out = D)
-beta.acc[,2] <- seq(1, 5, length.out = D)
+par.fix.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)), 
+                     sac = list(vel = c(3, 15), acc = c(3, 3), angle = c(0, 1)),
[…]

### 2020-05-08 simulation/preregistration_parameter_recovery.Rmd
update parameter recovery preregistration

@@
 These events include fixations, saccades, and optionally postsaccadic oscillations (PSOs) and smooth pursuits.
-The full algorithm will be available on GitHub (www.github.com/maltelueken/gazeHmm) upon publication.
+The full algorithm will be available on GitHub (www.github.com/maltelueken/gazeHMM) upon publication.
 
@@
 
-The HMM estimates non-observable states that have generated the data.
+The HMM estimates non-observable states that have generated the data (CITE).
 In the context of eye-tracking, each state corresponds to an eye movement event.
@@
 
-To estimate parameter recovery of the HMM, we repeatedly generate data with the model under different conditions and different numbers of states.
-For every condition and number of states $k \in \{2, 3, 4\}$, the model will generate $D = 100$ data sets with parameters varied in a specified interval in equidistant steps.
-The same model is then applied to estimate the parameters from the generated data.
-Then, the estimated parameters are compared with the "true" parameters that generated the data.
-We will assess parameter recovery descriptively by computing the mean squared error between the estimated and true parameter values separately for each parameter.
-Additionally, we will apply a bivariate linear regression for each parameter with the true parameter values as the independent and the estimated parameter values as the dependent variable.
-For a parameter to be recovered, we expect the regression intercepts to be close to zero and the regression slopes to be close to one.
-Next to parameter recovery, we will also examine the accuracy of the classficiation through the model.
-Therefore, we will compare the true states with the estimated states for each data set by using Cohen's Kappa.
-We will treat Kappa values above 0.8 as satisfactorily accurate.
+To estimate parameter recovery of the HMM, it will repeatedly generate data with a set of parameters (true parameter values).
+The same model will then be applied to estimate the parameters from the generated data (estimated parameter values).
+We will compare the true with the estimated parameter values to assess whether a parameter was recovered by the model.
+Additionally, we will compare the true states of the HMM with the estimated states to judge how accurate the model recovers the states that generated the data.
+
+## Starting values
 
@@
 To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$.
+The gamma distributions ensure that the starting values will be positive.
 Mean parameters of the von-Mises distribution will always start at their true value.
 
-In the first part of the simulation study, we will estimate parameter recovery under ideal conditions.
-These imply a medium sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples.
-The parameters that we will use to generate the data are displayed in Table 1.
-We will vary one parameter at a time on the given interval while setting the other parameters to their fixed value.
-The initial state probabilities will always stay at their fixed value.
-We will only vary the transition probabilities for staying in the same state.
-The remaining probability mass will be evenly distributed across the probabilities for switching to a different state (see Table 1).
-For two states 10, for three states 15, and for four states 20 parameters will be varied one at a time.
-This simulation will result in $D_{total} = 100 \times (10+15+20) = 4500$ recoveries.
+## Design
+### Parameter Variation
+
+The simulation study will be divided into four parts.
+In the first part, we will vary the parameters of the HMM.
+For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters will be varied respectively.
+For each parameter, the HMM will generate 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps.
+This will result in $100 \times (10+15+20) = 4500$ recoveries.
+Only one parameter will be varied at once, the other parameters will be set to their fixed values.
+We will not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
+For the transition probabilities, we will only vary the probabilities for staying in the same state (diagonals of the tranistion matrix) to reduce the complexity of the simulation.
+The left probability mass will be split evenly between the probabilites for switching to a different state (per row of the transition matrix).
+Moreover, we will not modify the mean parameters of the von-Mises distribution: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
+
+Where it was possible, we chose the intervals and fixed values for each parameter based on reported ranges for eye movement events in the literature.
+Otherwise, intervals and fixed values were chosen according to theoretical reasons and based on insights from data used to develop the model.
+Table 1 shows the theoretical ranges for each event and Table 2 shows the intervals and fixed values for each parameter in the simulation.
+Parameters were divided by 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
+We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above 5 approximate a symmetric distribution).
+The scale parameters will be set so that the respective distribution will match the ranges reported in the literature.
+Since the concentration parameters of the von-Mises distribution are the inverse of standard deviations, they will be varied on the inverse scale.
+
+### Sample Size and Noise Variation
 
 In the second part, we will vary the sample size of the generated data and the amount of noise added to it.
-The model parameters will be set to the fixed values from Table 1.
-For sample sizes of $N \in \{500, 2500, 10000\}$, we will generate data sets and add white noise.
-To velocity and acceleration, we will add Gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$.
-Noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ will be added to angle.
-This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.
+The model parameters will be set to their fixed values.
+For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we will generate 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
+We chose samples sizes roughly corresponding to small, medium, and large eye-tracking data sets of a single participant and trial.
+To simulate measurement error, we will add Gaussian white noise with $\mu = 0$ and $\sigma \in [1,5]$ varying between data sets to velocity and acceleration.
+Values that decrease to zero or below will be set to 0.001.
+We will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying between data sets to angle.
+
+### Variation of Starting Values
 
 In the third part, we will increase the variation in the starting values used for parameter estimation.
-The model parameters will be set to the fixed values from Table 1.
-For the shape, scale, and concentration parameters, we will increase the scale parameters of the gamma distributions to $\beta \in \{2, 4, 8\}$ all at the same time.
-The sample size will be $N = 2500$ and no noise will be added to the data.
-This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.
+The model parameters will be set to the fixed values.
+For the shape, scale, and concentration parameters, we will simultaneously increase the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta \in \{2, 4, 8\}$, 100 data sets with $N = 2500$ samples will be generated each ($100 \times 3 \times 3 = 900$ recoveries).
+
+### Missing data
 
 In the last part, we will set intervals of the generated data to be missing.
-The model parameters will be set to the fixed values from Table 1.
-The size of the missing data interval will be $m \in [1,100]$ and the number of intervals will be $n_{miss} \in \{1, 3, 5\}$.
-The sample size will be $N = 2500$ and no noise will be added to the data.
-This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.
+The model parameters will be set to the fixed values.
+For $k \in \{2, 3, 4\}$ states and $n_{miss} \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries).
+The size of the missing data interval $m \in [1,250]$ samples will vary between the data sets.
+
+## Recovery Measures
+
+For each parameter, we will calculate the mean squared error (MSE) between the true and estimated parameter values.
+Additonally, we will apply a bivariate linear regression  with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval.
+The MSE and the regression coefficients will serve as descriptive measures for parameter recovery.
+Lower MSE values, regression intercepts closer to zero, and regression slopes closer to one will be interpreted as better recovery.
+
+To assess state recovery, we will compute the proportion of states (for all states taken together, not for each state seperately) that have been classified correctly by the model for each generated data set.
+Higher proportions will be interpreted as better model accuracy.
+Overall, we require a median accuracy of 0.9 for the model to be satisfactorily accurate.
 
 Table 1
+
+*Theoretical ranges of response variables used to generate parameter values*
+
+|Event|Resp. variable|Range|Reference|
+|-----|--------------|-----|---------|
+|Fixation|Velocity|<30|Larsson et al. (2013)|
+|Fixation|Acceleration|<15|-|
+|Fixation|Angle|uniform|-|
+|Saccade|Velocity|30-500|Larsson et al. (2013)|
+|Saccade|Acceleration|10-250|-|
+|Saccade|Angle|~0|Pekkanen & Lappi (2017)|
+|PSO|Velocity|20-100|Larsson et al. (2013)|
+|PSO|Acceleration|10-90|-|
+|PSO|Angle|~$\pi$|Pekkanen & Lappi (2017)|
+|Smooth pursuit|Velocity|<100|Larsson et al. (2013)|
+|Smooth pursuit|Acceleration|<15|-|
+|Smooth pursuit|Angle|~0|Larsson et al. (2013)|
+
+*Note*. Units are °/s (velocity), °/s^2^ (acceleration), and radians (angle). ~ indicates that the distribution has a peak at this value.
+
+Table 2
 

### 2020-05-09 simulation/preregistration_parameter_recovery.Rmd
update parameter recovery

@@
 Values that decrease to zero or below will be set to 0.001.
-We will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying between data sets to angle.
+To angle, we will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying in equidistant steps between data sets.
 
@@
 For $k \in \{2, 3, 4\}$ states and $n_{miss} \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries).
-The size of the missing data interval $m \in [1,250]$ samples will vary between the data sets.
+The size of the missing data interval $m \in [1,250]$ samples will vary in equidistant steps between the data sets.
 
@@
 Overall, we require a median accuracy of 0.9 for the model to be satisfactorily accurate.
+
+When a model can not be fitted to a data set, it will be excluded from the recovery. We expect at least 95% of all models to be fitted.
+
+***
 
@@
 *Note*. Units are °/s (velocity), °/s^2^ (acceleration), and radians (angle). ~ indicates that the distribution has a peak at this value.
+
+***
 

### 2020-05-09 simulation/model_simulation.R

@@ -3,7 +3,7 @@
 # Author: Malte Lüken
 # Date: 06.05.2020
 
-HMM_simulate <- function(n, nstates, respstart, trstart, instart) {
+HMM_simulate <- function(n, nstates, trueresp, truetr, truein) {
   
   require(depmixS4)
   source("algorithm/model_helper_functions.R")
@@ -16,15 +16,15 @@ HMM_simulate <- function(n, nstates, respstart, trstart, instart) {
   
   # Create response model
   
-  resp <- list(list(altGamma(data$vel, pstart = respstart[["fix"]][["vel"]]),
-                    altGamma(data$acc, pstart = respstart[["fix"]][["acc"]]),
+  resp <- list(list(altGamma(data$vel, pstart = trueresp[["fix"]][["vel"]]),
+                    altGamma(data$acc, pstart = trueresp[["fix"]][["acc"]]),
                     unif(data$angle)))
   
   for (s in 2:nstates) {
     
-    resp[[s]] <- list(altGamma(data$vel, pstart = respstart[[s]][["vel"]]),
-                      altGamma(data$acc, pstart = respstart[[s]][["acc"]]),
-                      vMF(data$angle, pstart = respstart[[s]][["angle"]]))
+    resp[[s]] <- list(altGamma(data$vel, pstart = trueresp[[s]][["vel"]]),
+                      altGamma(data$acc, pstart = trueresp[[s]][["acc"]]),
+                      vMF(data$angle, pstart = trueresp[[s]][["angle"]]))
     
   }
   
@@ -35,14 +35,14 @@ HMM_simulate <- function(n, nstates, respstart, trstart, instart) {
   
   for (s in 1:nstates) {
     
-    trans[[s]] <- transInit(~ 1, nstates = nstates, data = data, pstart = c(trstart[s,]))
+    trans[[s]] <- transInit(~ 1, nstates = nstates, data = data, pstart = c(truetr[s,]))
     
   }
   
   
   # Create initial state model
   
-  init <- transInit(~ 1, ns = nstates, ps = instart, family = multinomial("identity"), data = data.frame(1))
+  init <- transInit(~ 1, ns = nstates, ps = truein, family = multinomial("identity"), data = data.frame(1))
   
   
   # Combine models

### 2020-05-09 simulation/parameter_recovery.R

@@ -11,11 +11,18 @@ source("algorithm/model_helper_functions.R")
 
 # Part 1 ------------------------------------------------------------------
 
+# Global parameters
+
 D <- 5
 N <- 2500
-K <- 2:4
+
+
+# Transition probability interval
 
 aij <- seq(0.01, 0.99, length.out = D)
+
+
+# Response parameter intervals
 
 par.int.true <- cbind(seq(1, 5, length.out = D),
                       seq(0.1, 0.6, length.out = D),
@@ -38,39 +45,97 @@ par.int.true <- cbind(seq(1, 5, length.out = D),
                       1/seq(0.1, 10, length.out = D))
 
 
+# Fixed response parameter values
+
 par.fix.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)), 
                      sac = list(vel = c(3, 15), acc = c(3, 3), angle = c(0, 1)),
                      pso = list(vel = c(3, 3), acc = c(2, 2), angle = c(pi, 1)),
                      sp = list(vel = c(3, 1.5), acc = c(3, 0.15), angle = c(0, 1)))
 
+
+# Parameter indicators
+
 par.states <- c(rep(1, 4), rep(2:4, each = 5))
-par.names <- c("alpha", "beta", "alpha", "beta", rep(c("alpha", "beta", "alpha", "beta", "kappa"), 3))
-par.resp <- c(rep(1:2, each = 2), rep(c(1, 1, 2, 2, 3),  3))
-
-ncores <- detectCores()
-clust <- makeCluster(ncores)
-clusterExport(clust, list("HMM_simulate", "N", "par.fix.true"))
-                               
-estimates <- parLapply(clust, par.int.true[,1], function(x, N = 500, K = 2, k = 2, par = "alpha", met = 1, true = par.fix.true) {
-  
-  par.fix.true[[k]][[met]][1] <- 3
-  
-  instart <- rep(1/K, K)
-  
-  trstart <- matrix(1/K, nrow = K, ncol = K)
-  
-  model <- HMM_simulate(n = N, nstates = k, respstart = true, trstart = trstart, instart = instart)
-  
-  sim <- simulate(model)
-  
-  class(sim) <- "depmix"
-  
-  out <- fit(sim, emc = em.control(random.start = F))
-  
-  return(out)
-})
[…]

### 2020-05-10 simulation/preregistration_parameter_recovery.Rmd
update parameter recovery

@@
 The HMM is implemented in R (CITE) using the package depmixS4 (CITE).
-An expectation-maximization algorithm estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE).
+An expectation-maximization algorithm (with a maximum of 5000 iterations) estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE).
 The parameters of each response model (except for the uniform distribution) are estimated through maximum likelihood with a spectral projected gradient method (CITE) and Barzilai-Borwein step lengths (CITE) as implemented in the package BB (CITE).
@@
 We will not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
-For the transition probabilities, we will only vary the probabilities for staying in the same state (diagonals of the tranistion matrix) to reduce the complexity of the simulation.
+For the transition probabilities, we will only vary the probabilities for staying in the same state (diagonals of the tranistion matrix) simultaneously to reduce the complexity of the simulation.
 The left probability mass will be split evenly between the probabilites for switching to a different state (per row of the transition matrix).
@@
 Values that decrease to zero or below will be set to 0.001.
-To angle, we will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying in equidistant steps between data sets.
+To angle, we will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying between data sets.
+$\sigma$ and $\kappa$ will vary simultaneously in equidistant steps in their intervals.
 

### 2020-05-10 simulation/parameter_recovery.R

@@ -17,14 +17,10 @@ D <- 5
 N <- 2500
 
 
-# Transition probability interval
+# Parameter intervals
 
-aij <- seq(0.01, 0.99, length.out = D)
-
-
-# Response parameter intervals
-
-par.int.true <- cbind(seq(1, 5, length.out = D),
+par.int.true <- cbind(seq(0.01, 0.99, length.out = D),
+                      seq(1, 5, length.out = D),
                       seq(0.1, 0.6, length.out = D),
                       seq(1, 5, length.out = D),
                       seq(0.05, 0.25, length.out = D),
@@ -55,87 +51,382 @@ par.fix.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)),
 
 # Parameter indicators
 
-par.states <- c(rep(1, 4), rep(2:4, each = 5))
-par.names <- c(1, 2, 1, 2, rep(c(1, 2, 1, 2, 2), 3))
-par.resps <- c(rep(1:2, each = 2), rep(c(1, 1, 2, 2, 3),  3))
+par.states <- c(0, rep(1, 4), rep(2:4, each = 5))
+par.names <- c(0, 1, 2, 1, 2, rep(c(1, 2, 1, 2, 2), 3))
+par.resps <- c(0, rep(1:2, each = 2), rep(c(1, 1, 2, 2, 3),  3))
 
 
-# Prepare parallel processing
+# Parameter recovery simulation
 
-estimates <- list()
+estimates.1 <- list()
 
-for (p in 1:sum(par.states <= 2)) {
-
-  ncores <- detectCores()
-  clust <- makeCluster(ncores)
-  clusterExport(clust, list("HMM_simulate", "N", "par.fix.true", "par.states", "par.names", "par.resps", "p"))
-  
-  estimates[[p]] <- parLapply(clust, par.int.true[,p], function(x, samples = N, k = 2, par.state = par.states[p], 
-                                                           par.name = par.names[p], par.resp = par.resps[p], 
-                                                           trueResp = par.fix.true) {
+for (ns in 2:4) {
+  for (p in 1:sum(par.states <= ns)) {
     
-    # Set true parameter values
+    # Prepare parallel processing
     
-    trueResp[[par.state]][[par.resp]][par.name] <- x
-    
-    trueIn <- rep(1/k, k)
-    
-    trueTr <- matrix(c(0.9, 0.1, 
-                       0.1, 0.9), nrow = k, ncol = k)
[…]

### 2020-05-11 simulation/preregistration_parameter_recovery.Rmd
update parameter recovery

@@
 ---
-title: "gazeHMM: Parameter recovery simulation"
-author: "Malte Lüken, Simon Kucharsky, & Ingmar Visser"
+title: "Preregistration: Estimating the Parameter Recovery of the Hidden Markov Model in gazeHMM"
+author: |
+ | Malte Lüken, Simon Kucharsky, & Ingmar Visser
+ |
+ | University of Amsterdam, Department of Psychology
 affiliation: "University of Amsterdam"
@@
 The HMM is implemented in R (CITE) using the package depmixS4 (CITE).
-An expectation-maximization algorithm (with a maximum of 5000 iterations) estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE).
+An expectation-maximization algorithm (maximum iterations: 5000, relative tolerance: 1e-08) estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE).
 The parameters of each response model (except for the uniform distribution) are estimated through maximum likelihood with a spectral projected gradient method (CITE) and Barzilai-Borwein step lengths (CITE) as implemented in the package BB (CITE).
@@
 The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilities.
-To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$.
-The gamma distributions ensure that the starting values will be positive.
+To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated.
+This setup ensures that the starting values are positive, their distribution is moderately skewed, and the mode of the their distributions equals the true parameter value.
 Mean parameters of the von-Mises distribution will always start at their true value.
@@
 This will result in $100 \times (10+15+20) = 4500$ recoveries.
-Only one parameter will be varied at once, the other parameters will be set to their fixed values.
+Only one parameter will be varied at once, the other parameters will be set to their default values.
 We will not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
@@
 
-Where it was possible, we chose the intervals and fixed values for each parameter based on reported ranges for eye movement events in the literature.
-Otherwise, intervals and fixed values were chosen according to theoretical reasons and based on insights from data used to develop the model.
-Table 1 shows the theoretical ranges for each event and Table 2 shows the intervals and fixed values for each parameter in the simulation.
+Where it was possible, we chose the intervals and default values for each parameter based on reported ranges for eye movement events in the literature.
+Otherwise, intervals and default values were chosen according to theoretical reasons and based on insights from data used to develop the model.
+Table 1 shows the assumed ranges for each event and Table 2 shows the intervals and default values for each parameter in the simulation.
 Parameters were divided by 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
 We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above 5 approximate a symmetric distribution).
-The scale parameters will be set so that the respective distribution will match the ranges reported in the literature.
+The scale parameters will be set so that the respective distribution will match the assumed ranges.
 Since the concentration parameters of the von-Mises distribution are the inverse of standard deviations, they will be varied on the inverse scale.
@@
 In the second part, we will vary the sample size of the generated data and the amount of noise added to it.
-The model parameters will be set to their fixed values.
+The model parameters will be set to their default values.
 For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we will generate 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
 We chose samples sizes roughly corresponding to small, medium, and large eye-tracking data sets of a single participant and trial.
-To simulate measurement error, we will add Gaussian white noise with $\mu = 0$ and $\sigma \in [1,5]$ varying between data sets to velocity and acceleration.
-Values that decrease to zero or below will be set to 0.001.
-To angle, we will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying between data sets.
-$\sigma$ and $\kappa$ will vary simultaneously in equidistant steps in their intervals.
+To simulate noise, we will replace response values $y$ per sample with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets to velocity and acceleration.
+This procedure ensures that velocity and acceleration values remain positive and are drawn from moderately skewed distributions with modes equal to the original values.
+To angle, we will add white noise from a von-Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
+$\tau_{noise}$ and $\kappa_noise$ will vary simultaneously in equidistant steps in their intervals.
 
@@
 In the third part, we will increase the variation in the starting values used for parameter estimation.
-The model parameters will be set to the fixed values.
-For the shape, scale, and concentration parameters, we will simultaneously increase the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta \in \{2, 4, 8\}$, 100 data sets with $N = 2500$ samples will be generated each ($100 \times 3 \times 3 = 900$ recoveries).
+The model parameters will be set to their default values.
+For the shape, scale, and concentration parameters, we will simultaneously increase the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta_{start} = (\psi_{true}/2)\tau_{start}$ with $\tau_{start} \in \{1, 2, 3\}$, each 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries).
 
@@
 In the last part, we will set intervals of the generated data to be missing.
-The model parameters will be set to the fixed values.
-For $k \in \{2, 3, 4\}$ states and $n_{miss} \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries).
-The size of the missing data interval $m \in [1,250]$ samples will vary in equidistant steps between the data sets.
+The model parameters will be set to their default values.
+For $k \in \{2, 3, 4\}$ states and $m \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries).
+The size of the missing data interval $l \in [1,200]$ samples will vary in equidistant steps between the data sets.
 
-## Recovery Measures
+## Recovery Analysis
 
@@
 Higher proportions will be interpreted as better model accuracy.
-Overall, we require a median accuracy of 0.9 for the model to be satisfactorily accurate.
 
-When a model can not be fitted to a data set, it will be excluded from the recovery. We expect at least 95% of all models to be fitted.
+When a model can not be fitted to a data set, it will be excluded from the recovery.
 
@@
 
-*Theoretical ranges of response variables used to generate parameter values*
+*Assumed ranges of response variables used to generate parameter values*
 
-|Event|Resp. variable|Range|Reference|
-|-----|--------------|-----|---------|
-|Fixation|Velocity|<30|Larsson et al. (2013)|
-|Fixation|Acceleration|<15|-|
-|Fixation|Angle|uniform|-|
-|Saccade|Velocity|30-500|Larsson et al. (2013)|
-|Saccade|Acceleration|10-250|-|
-|Saccade|Angle|~0|Pekkanen & Lappi (2017)|
-|PSO|Velocity|20-100|Larsson et al. (2013)|
-|PSO|Acceleration|10-90|-|
-|PSO|Angle|~$\pi$|Pekkanen & Lappi (2017)|
-|Smooth pursuit|Velocity|<100|Larsson et al. (2013)|
-|Smooth pursuit|Acceleration|<15|-|
-|Smooth pursuit|Angle|~0|Larsson et al. (2013)|
+|Event|Resp. variable|Range|
+|-----|--------------|-----|
+|Fixation|Velocity|<30|
+|Fixation|Acceleration|<10|
+|Fixation|Angle|uniform|
+|Saccade|Velocity|50-1000|
+|Saccade|Acceleration|10-300|
+|Saccade|Angle|~0|
+|PSO|Velocity|20-100|
+|PSO|Acceleration|10-90|
+|PSO|Angle|~$\pi$|
+|Smooth pursuit|Velocity|<100|
+|Smooth pursuit|Acceleration|<15|
+|Smooth pursuit|Angle|~0|
 
@@
 |-----|---------|--------|-----|-----------|
-|1-4|$p_i^+$|-|$1/k$|Initial state probability for starting in state *i*|
+|1-4|$\rho_i^*$|-|$1/k$|Initial state probability for starting in state *i*|
 |1-4|$a_{i=j}$|[.01,.99]|0.9|Transition probability for staying in the previous state *i*|
@@
 |1|$\beta_{acc}$|[0.05,0.25]|0.15|Scale parameter of the acceleration gamma distribution|
-|1|$a^+$|-|0|Minimum of the uniform distribution|
-|1|$b^+$|-|$2\pi$|Maximum of the uniform distribution|
+|1|$min^*$|-|0|Minimum of the uniform distribution|
+|1|$max^*$|-|$2\pi$|Maximum of the uniform distribution|
 |2|$\alpha_{vel}$|[1,5]|3||
@@
 |2|$\beta_{acc}$|[1,5]|3||
-|2|$\mu^+$|-|0|Mean parameter of the von-Mises distribution|
+|2|$\mu^*$|-|0|Mean parameter of the von-Mises distribution|
 |2|$\kappa$|$1/[0.1,10]$|1|Concentration parameter of the von Mises distribution
@@
 |3|$\beta_{acc}$|[1,3]|2||
-|3|$\mu^+$|-|$\pi$||
+|3|$\mu^*$|-|$\pi$||
 |3|$\kappa$|$1/[0.1,10]$|1||
@@
 |4|$\beta_{acc}$|[0.05,0.25]|0.15||
-|4|$\mu^+$|-|0||
+|4|$\mu^*$|-|0||
 |4|$\kappa$|$1/[0.1,10]$|1||
 
-*Note*. Parameters marked with $^+$ will not be varied but always set to their fixed value. *k* is the number of states in the model.
+*Note*. Parameters marked with $^*$ will not be varied but always set to their default values. *k* is the number of states in the model.

### 2020-05-11 simulation/parameter_recovery.R

@@ -3,17 +3,32 @@
 # Author: Malte Lüken
 # Date: 06.05.2020
 
+
+# Global parameters -------------------------------------------------------
+
 library(depmixS4)
 library(parallel)
 source("simulation/model_simulation.R")
 source("algorithm/model_helper_functions.R")
 
 
-# Part 1 ------------------------------------------------------------------
-
-# Global parameters
+# Number of data sets to be generated
 
 D <- 5
+
+
+# Default response parameter values
+
+par.fix.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)), 
+                     sac = list(vel = c(3, 15), acc = c(3, 3), angle = c(0, 1)),
+                     pso = list(vel = c(3, 3), acc = c(2, 2), angle = c(pi, 1)),
+                     sp = list(vel = c(3, 1.5), acc = c(3, 0.15), angle = c(0, 1)))
+
+
+# Part 1 ------------------------------------------------------------------
+
+# Sample size
+
 N <- 2500
 
 
@@ -39,14 +54,6 @@ par.int.true <- cbind(seq(0.01, 0.99, length.out = D),
                       seq(1, 3, length.out = D),
                       seq(1, 3, length.out = D),
                       1/seq(0.1, 10, length.out = D))
-
-
-# Fixed response parameter values
-
-par.fix.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)), 
-                     sac = list(vel = c(3, 15), acc = c(3, 3), angle = c(0, 1)),
-                     pso = list(vel = c(3, 3), acc = c(2, 2), angle = c(pi, 1)),
-                     sp = list(vel = c(3, 1.5), acc = c(3, 0.15), angle = c(0, 1)))
 
 
 # Parameter indicators
@@ -76,8 +83,7 @@ for (ns in 2:4) {
                                                                                         par.state = par.states[p], 
                                                                                         par.name = par.names[p], 
                                                                                         par.resp = par.resps[p], 
-                                                                                        trueResp = par.fix.true, 
-                                                                                        seed = 123) {
+                                                                                        trueResp = par.fix.true) {
       
[…]

### 2020-05-11 simulation/parameter_recovery_analysis.R

@@ -1,0 +1,160 @@
+### Parameter recovery analysis ###
+
+# Author: Malte Lüken
+# Date: 11.05.2020
+
+
+library(tidyverse)
+library(depmixS4)
+
+
+# Load data
+
+for (part in 1:4) {
+  
+  load(file = paste("simulation/part", part, ".Rdata", sep = ""))
+  
+}
+
+linkinv <- function(eta,base) {
+  linv <- function(eta,base) {
+    pp <- numeric(length(eta))
+    if(any(is.infinite(eta)) || any(eta > log(.Machine$double.xmax)) || any(eta < log(.Machine$double.xmin))) {
+      pp[which(is.infinite(eta))] <- 1
+      pp[which(eta > log(.Machine$double.xmax))] <- 1 # change this to something better!
+    } else {
+      expb <- exp(eta)
+      sumb <- sum(expb)
+      pp[base] <- 1/sumb
+      pp[-base] <- expb[-base]/sumb
+    }
+    return(pp)
+  }
+  if(is.matrix(eta)) {
+    if(ncol(eta)==1) {
+      pp <- as.matrix(apply(eta,1,linv,base=base)) # fixes problem with column matrix eta
+    } else pp <- t(apply(eta,1,linv,base=base)) 	
+  } else {
+    pp <- linv(eta,base)
+  }
+  return(pp)
+}
+
+
+# Function to transform parameters to normal scale
+
+backtrans <- function(x) {
+  
+  out <- x
+  
+  nms <- names(x)
+  
+  out[str_detect(nms, "(Intercept)")] <- linkinv(x[str_detect(nms, "(Intercept)")], base = 1)
+  out[nms %in% c("shape", "scale", "kappa")] <- exp(x[nms %in% c("shape", "scale", "kappa")])
+  
+  return(out)
+}
+
+
+# Calculate MSE
+
+mse <- list()
+
+for (part in 1:4) {
+  
+  mse[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
+    lapply(x, function(y) {
+      sqerr <- lapply(y, function(z) {
+        
+        err <- try((backtrans(z$pars.true) - backtrans(z$pars.est))^2)
+        
+        if(is.numeric(err)) {
+          
+          out <- err
+          
+        } else {
+          
+          out <- rep(NA, length(z$pars.true))
+          
+        }
+        
+        names(out) <- names(z$pars.true)
+        
+        return(out)
+      })
+      
[…]

### 2020-05-12 algorithm/model_helper_functions.R
back-transform log-scale parameters in show method

@@ -194,7 +194,7 @@ setMethod("show", "vMF",
             cat("Model of type vMF\n")
             cat("Parameters: \n")
             cat("mu: ", object@parameters$mu, "\n")
-            cat("kappa: ", object@parameters$kappa, "\n")
+            cat("kappa: ", exp(object@parameters$kappa), "\n")
           }
 )
 
@@ -357,8 +357,8 @@ setMethod("show","altGamma",
           function(object) {
             cat("Model of type altGamma (see ?altGammamlss for details) \n")
             cat("Parameters: \n")
-            cat("shape: ", object@parameters$shape, "\n")
-            cat("scale: ", object@parameters$scale, "\n")
+            cat("shape: ", exp(object@parameters$shape), "\n")
+            cat("scale: ", exp(object@parameters$scale), "\n")
           }
 )
 

### 2020-05-12 simulation/parameter_recovery_analysis.R
update parameter recovery

@@ -39,7 +39,23 @@ linkinv <- function(eta,base) {
   }
   return(pp)
 }
-
+linkfun <- function(p, base) {
+  lfun <- function(p, base) {
+    p <- p/sum(p)
+    beta <- numeric(length(p))
+    if (any(p == 1)) 
+      beta[which(p == 1)] = Inf
+    else beta[-base] <- log(p[-base]/p[base])
+    return(beta)
+  }
+  if (is.matrix(p)) {
+    beta <- t(apply(p, 1, lfun, base = base))
+  }
+  else {
+    beta <- lfun(p, base)
+  }
+  return(beta)
+}
 
 # Function to transform parameters to normal scale
 
@@ -49,11 +65,17 @@ backtrans <- function(x) {
   
   nms <- names(x)
   
-  out[str_detect(nms, "(Intercept)")] <- linkinv(x[str_detect(nms, "(Intercept)")], base = 1)
+  out[str_detect(nms, "(Intercept)")] <- as.vector(apply(matrix(x[str_detect(nms, "(Intercept)")],
+                                                        ncol = sqrt(length(x[str_detect(nms, "(Intercept)")])),
+                                                        byrow = T), 1, linkinv, base = 1))
+  
   out[nms %in% c("shape", "scale", "kappa")] <- exp(x[nms %in% c("shape", "scale", "kappa")])
   
   return(out)
 }
+mlogit
+
+#backtrans(estimates.1$`3`[[1]][[3]]$pars.est)
 
 
 # Calculate MSE
@@ -105,56 +127,111 @@ mse.data <- lapply(mse, function(x) lapply(x, as.data.frame))
 mse.data <- lapply(mse.data, function(x) lapply(x, function(y) {as.data.frame(t(as.matrix(y)))}))
 
 
-# Calculate linear regression weights
+# Calculate regression weights for transition probabilities
 
-regw <- list()
+regw.tr <- list()
 
-for (part in 1:1) {
+regw.tr <- lapply(get("estimates.1"), function(x) {
   
-  regw[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
+  varpar <- lapply(x[1], function(y) {
     
-    index <- 1:length(x)
-    
-    #print(index)
-    
-    lapply(index, function(i, y) {
+    lapply(y, function(z) {
       
-      print(i)
+      nms <- names(z$pars.true)
       
-      lapply(y[[i]], function(z) {
-        
[…]

### 2020-05-12 simulation/parameter_recovery_simulation.R

@@ -8,6 +8,7 @@
 
 library(depmixS4)
 library(parallel)
+library(tidyverse)
 source("simulation/model_simulation.R")
 source("algorithm/model_helper_functions.R")
 
@@ -74,7 +75,7 @@ for (ns in 2:4) {
     
     ncores <- detectCores()
     clust <- makeCluster(ncores)
-    clusterExport(clust, list("HMM_simulate", "N", "par.fix.true", "par.states", "par.names", "par.resps", "p", "ns"))
+    clusterExport(clust, list("HMM_simulate", "str_detect", "N", "par.fix.true", "par.states", "par.names", "par.resps", "p", "ns"))
     
     
     # Iterate over intervals
@@ -132,7 +133,8 @@ for (ns in 2:4) {
       respStart[["sp"]][["angle"]][1] <- 0
       
       inStart <- trueIn
-      trStart <- logit(matrix(1/k, nrow = k, ncol = k))
+      trStart <- matrix(1/k, nrow = k, ncol = k)
+      trStart <- apply(trStart, 1, model.sim@transition[[1]]@family$linkfun, base = model.sim@transition[[1]]@family$base)
       
       start <- c(inStart, trStart, unlist(respStart)[1:(6*k)])
       
@@ -149,6 +151,9 @@ for (ns in 2:4) {
       # Calculate accuracy
       
       acc <- try(mean(model.sim@states == model.fit@posterior$state))
+      
+      
+      # Backtransform parameters
       
       output <- list(pars.true = getpars(model), pars.start = getpars(model.start), pars.est = try(getpars(model.fit)), accuracy = acc)
       
@@ -254,7 +259,8 @@ for (ns in 2:4) {
       respStart[["sp"]][["angle"]][1] <- 0
       
       inStart <- trueIn
-      trStart <- logit(matrix(1/k, nrow = k, ncol = k))
+      trStart <- matrix(1/k, nrow = k, ncol = k)
+      trStart <- apply(trStart, 1, model.sim@transition[[1]]@family$linkfun, base = model.sim@transition[[1]]@family$base)
       
       start <- c(inStart, trStart, unlist(respStart)[1:(6*k)])
       
@@ -353,7 +359,8 @@ for (ns in 2:4) {
       respStart[["sp"]][["angle"]][1] <- 0
       
       inStart <- trueIn
-      trStart <- logit(matrix(1/k, nrow = k, ncol = k))
[…]

### 2020-05-13 simulation/preregistration/preregistration_parameter_recovery.Rmd
update parameter recovery

--- 2020-05-11 simulation/preregistration_parameter_recovery.Rmd
@@
 date: "`r format(Sys.Date(), '%d.%m.%Y')`"
-output: pdf_document
+csl: "apa.csl"
+bibliography: "gazeHMM_prereg.bib"
+output: 
+  pdf_document
 ---
@@
 Missing samples are identified, the data are filtered, and for each sample the velocity, acceleration, and difference in angle to the next sample is calculated.
-Second, the HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events.
+Second, a HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events.
 Moreover, the HMM computes the posterior probability of each sample for belonging to each event.
@@
 
-The HMM estimates non-observable states that have generated the data (CITE).
+The HMM estimates non-observable states that have generated the data [@Visser2019].
 In the context of eye-tracking, each state corresponds to an eye movement event.
@@
 
-The HMM is implemented in R (CITE) using the package depmixS4 (CITE).
-An expectation-maximization algorithm (maximum iterations: 5000, relative tolerance: 1e-08) estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE).
-The parameters of each response model (except for the uniform distribution) are estimated through maximum likelihood with a spectral projected gradient method (CITE) and Barzilai-Borwein step lengths (CITE) as implemented in the package BB (CITE).
+The HMM is implemented in R [version: 3.6.3; @RCoreTeam2020] using the package depmixS4 [@Visser2010].
+An expectation-maximization algorithm [maximum iterations: 5000, relative tolerance: 1e-08; @Dempster1977; @McLachlan1997] estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973].
+The parameters of each response model (except for the uniform distribution) are estimated through maximum likelihood with a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988] as implemented in the package BB [@Varadhan2009].
 
@@
 To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated.
-This setup ensures that the starting values are positive, their distribution is moderately skewed, and the mode of the their distributions equals the true parameter value.
-Mean parameters of the von-Mises distribution will always start at their true value.
+This setup ensures that the starting values are positive, their distributions are moderately skewed, and the modes of their distributions equal the true parameter values.
+Mean parameters of the von-Mises distribution will always start at their true values.
 
@@
 Only one parameter will be varied at once, the other parameters will be set to their default values.
+All parameters of the HMM will be estimated freely (i.e., there are no fixed parameters in the model).
 We will not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
-For the transition probabilities, we will only vary the probabilities for staying in the same state (diagonals of the tranistion matrix) simultaneously to reduce the complexity of the simulation.
-The left probability mass will be split evenly between the probabilites for switching to a different state (per row of the transition matrix).
-Moreover, we will not modify the mean parameters of the von-Mises distribution: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
+For the transition probabilities, we will only simultaneously vary the probabilities for staying in the same state (diagonals of the tranistion matrix) to reduce the complexity of the simulation.
+The left over probability mass will be split evenly between the probabilites for switching to a different state (per row of the transition matrix).
+Moreover, we will not modify the mean parameters of the von-Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
 
@@
 Parameters were divided by 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
-We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above 5 approximate a symmetric distribution).
+We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above 5 approach a symmetric distribution).
 The scale parameters will be set so that the respective distribution will match the assumed ranges.
@@
 For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we will generate 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
-We chose samples sizes roughly corresponding to small, medium, and large eye-tracking data sets of a single participant and trial.
-To simulate noise, we will replace response values $y$ per sample with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets to velocity and acceleration.
+These samples sizes roughly correspond to small, medium, and large eye-tracking data sets for a single participant and trial.
+To simulate noise, we will replace velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
 This procedure ensures that velocity and acceleration values remain positive and are drawn from moderately skewed distributions with modes equal to the original values.
 To angle, we will add white noise from a von-Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
-$\tau_{noise}$ and $\kappa_noise$ will vary simultaneously in equidistant steps in their intervals.
+$\tau_{noise}$ and $\kappa_{noise}$ will vary simultaneously in equidistant steps in their intervals.
 
@@
 *Note*. Parameters marked with $^*$ will not be varied but always set to their default values. *k* is the number of states in the model.
+
+# References

### 2020-05-13 simulation/preregistration/preregistration_parameter_recovery.Rmd
update parameter recovery

@@
 bibliography: "gazeHMM_prereg.bib"
-output: 
-  pdf_document
+output: pdf_document
 ---
@@
 Parameters were divided by 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
-We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above 5 approach a symmetric distribution).
+We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above five approach a symmetric distribution).
 The scale parameters will be set so that the respective distribution will match the assumed ranges.
@@
 
-For each parameter, we will calculate the mean squared error (MSE) between the true and estimated parameter values.
-Additonally, we will apply a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval.
-The MSE and the regression coefficients will serve as descriptive measures for parameter recovery.
-Lower MSE values, regression intercepts closer to zero, and regression slopes closer to one will be interpreted as better recovery.
+For each parameter, we will calculate the root median square proportion deviation [RMSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$RMSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
+We will treat $RMSPD < 0.1$ as good, $0.1 \le RMSPD < 0.5$ as moderate, and $RMSPD \ge 5$ as bad recovery of a parameter.
+By taking the median, we will reduce the influence of potential outliers in the estimation and using proportions will enable us to compare RMSPD values across parameters and data sets.
 
-To assess state recovery, we will compute the proportion of states (for all states taken together, not for each state seperately) that have been classified correctly by the model for each generated data set.
-Higher proportions will be interpreted as better model accuracy.
+Additionally, we will apply a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval.
+Regression slopes closer to one will indicate that the model better captures parameter change.
+Regression intercepts different from zero will reflect bias in parameter estimation.
+
+To assess state recovery, we will compute Cohen's kappa (for all states taken together, not for each state seperately) as a measure of agreement between true and estimated states for each generated data set.
+Higher kappa values will be interpreted as better model accuracy.
+We will adopt the ranges proposed by @Landis1977 to interpret kappa values.
 
@@
 
-*Assumed ranges of response variables used to generate parameter values*
+*Assumed Ranges of Response Variables Used to Generate Parameter Values*
 
@@
 
-*HMM parameter values for generating the data*
+*HMM Parameter Values for Generating the Data*
 
-|State|Parameter|Interval|Fixed|Description|
+|State|Parameter|Interval|Default|Description|
 |-----|---------|--------|-----|-----------|

### 2020-05-13 simulation/parameter_recovery_analysis.R

@@ -78,17 +78,18 @@ mlogit
 #backtrans(estimates.1$`3`[[1]][[3]]$pars.est)
 
 
-# Calculate MSE
+# Calculate normalized root mean square deviation
 
-mse <- list()
+rmsd <- list()
 
 for (part in 1:4) {
   
-  mse[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
+  rmsd[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
     lapply(x, function(y) {
       sqerr <- lapply(y, function(z) {
         
-        err <- try((backtrans(z$pars.true) - backtrans(z$pars.est))^2)
+        err <- try(((backtrans(z$pars.est) - backtrans(z$pars.true))/
+                      ifelse(backtrans(z$pars.true) == 0, 2*pi, backtrans(z$pars.true)))^2)
         
         if(is.numeric(err)) {
           
@@ -110,12 +111,24 @@ for (part in 1:4) {
       nms <- names(sqerr[[1]])
 
       pars <- matrix(unlist(sqerr), nrow = rows, byrow = T)
-
-      msqerr <- apply(pars, 2, mean, na.rm = T)
-
+      
+      msqerr <- apply(pars, 2, median, na.rm = T)
+      
       names(msqerr) <- nms
-
-      return(msqerr)
+      
+      rmsqerr <- sqrt(msqerr)
+      
+      # parstrue <- lapply(y, function(z) {backtrans(z$pars.true)})
+      # 
+      # parstruemat <- matrix(unlist(parstrue), nrow = rows, byrow = T)
+      # 
+      # meantrue <- apply(parstruemat, 2, mean, na.rm = T)
+      # 
+      # meantrue <- ifelse(meantrue == 0, 2*pi, meantrue)
+      # 
+      # print(rmsqerr/meantrue)
+      
+      return(rmsqerr)
     })
   })
 }
@@ -123,8 +136,8 @@ for (part in 1:4) {
 
 # Display MSE in data frame
 
-mse.data <- lapply(mse, function(x) lapply(x, as.data.frame))
-mse.data <- lapply(mse.data, function(x) lapply(x, function(y) {as.data.frame(t(as.matrix(y)))}))
+rmsd.data <- lapply(rmsd, function(x) lapply(x, as.data.frame))
[…]

### 2020-05-14 simulation/preregistration/preregistration_parameter_recovery.Rmd
change parameter ranges and default values

@@
 
-*Assumed Ranges of Response Variables Used to Generate Parameter Values*
+*Assumed ranges of response variables used to generate parameter values*
 
@@
 |-----|--------------|-----|
-|Fixation|Velocity|<30|
-|Fixation|Acceleration|<10|
+|Fixation|Velocity|0-50|
+|Fixation|Acceleration|0-50|
 |Fixation|Angle|uniform|
 |Saccade|Velocity|50-1000|
-|Saccade|Acceleration|10-300|
+|Saccade|Acceleration|50-500|
 |Saccade|Angle|~0|
@@
 |PSO|Angle|~$\pi$|
-|Smooth pursuit|Velocity|<100|
-|Smooth pursuit|Acceleration|<15|
+|Smooth pursuit|Velocity|20-100|
+|Smooth pursuit|Acceleration|0-30|
 |Smooth pursuit|Angle|~0|
@@
 
-*HMM Parameter Values for Generating the Data*
+*HMM parameter values for generating the data*
 
@@
 |2|$\alpha_{vel}$|[1,5]|3||
-|2|$\beta_{vel}$|[5,25]|15||
+|2|$\beta_{vel}$|[5,15]|10||
 |2|$\alpha_{acc}$|[1,5]|3||
@@
 |3|$\alpha_{vel}$|[1,5]|3||
-|3|$\beta_{vel}$|[1,5]|3||
-|3|$\alpha_{acc}$|[1,3]|2||
-|3|$\beta_{acc}$|[1,3]|2||
+|3|$\beta_{vel}$|[0.5,1.5]|1||
+|3|$\alpha_{acc}$|[1,5]|3||
+|3|$\beta_{acc}$|[1,5]|3||
 |3|$\mu^*$|-|$\pi$||
@@
 |4|$\alpha_{vel}$|[1,5]|3||
-|4|$\beta_{vel}$|[1,2]|1.5||
+|4|$\beta_{vel}$|[0.5,1.5]|1||
 |4|$\alpha_{acc}$|[1,5]|3||

### 2020-05-14 simulation/parameter_recovery_simulation.R

@@ -21,9 +21,9 @@ D <- 5
 # Default response parameter values
 
 par.fix.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)), 
-                     sac = list(vel = c(3, 15), acc = c(3, 3), angle = c(0, 1)),
-                     pso = list(vel = c(3, 3), acc = c(2, 2), angle = c(pi, 1)),
-                     sp = list(vel = c(3, 1.5), acc = c(3, 0.15), angle = c(0, 1)))
+                     sac = list(vel = c(3, 10), acc = c(3, 3), angle = c(0, 1)),
+                     pso = list(vel = c(3, 1), acc = c(3, 3), angle = c(pi, 1)),
+                     sp = list(vel = c(3, 1), acc = c(3, 0.15), angle = c(0, 1)))
 
 
 # Part 1 ------------------------------------------------------------------
@@ -41,19 +41,19 @@ par.int.true <- cbind(seq(0.01, 0.99, length.out = D),
                       seq(1, 5, length.out = D),
                       seq(0.05, 0.25, length.out = D),
                       seq(1, 5, length.out = D),
-                      seq(5, 25, length.out = D),
+                      seq(5, 15, length.out = D),
                       seq(1, 5, length.out = D),
                       seq(1, 5, length.out = D),
                       1/seq(0.1, 10, length.out = D),
                       seq(1, 5, length.out = D),
+                      seq(0.5, 1.5, length.out = D),
                       seq(1, 5, length.out = D),
                       seq(1, 5, length.out = D),
-                      seq(1, 3, length.out = D),
                       1/seq(0.1, 10, length.out = D),
                       seq(1, 5, length.out = D),
+                      seq(0.5, 1.5, length.out = D),
                       seq(1, 5, length.out = D),
-                      seq(1, 3, length.out = D),
-                      seq(1, 3, length.out = D),
+                      seq(0.05, 0.25, length.out = D),
                       1/seq(0.1, 10, length.out = D))
 
 

### 2020-05-14 simulation/parameter_recovery_simulation.R
change accuracy to cohen's kappa

@@ -9,6 +9,7 @@
 library(depmixS4)
 library(parallel)
 library(tidyverse)
+library(psych)
 source("simulation/model_simulation.R")
 source("algorithm/model_helper_functions.R")
 
@@ -75,7 +76,7 @@ for (ns in 2:4) {
     
     ncores <- detectCores()
     clust <- makeCluster(ncores)
-    clusterExport(clust, list("HMM_simulate", "str_detect", "N", "par.fix.true", "par.states", "par.names", "par.resps", "p", "ns"))
+    clusterExport(clust, list("HMM_simulate", "cohen.kappa", "N", "par.fix.true", "par.states", "par.names", "par.resps", "p", "ns"))
     
     
     # Iterate over intervals
@@ -150,10 +151,7 @@ for (ns in 2:4) {
       
       # Calculate accuracy
       
-      acc <- try(mean(model.sim@states == model.fit@posterior$state))
-      
-      
-      # Backtransform parameters
+      acc <- try(cohen.kappa(data.frame(x = model.sim@states, y = model.fit@posterior$state))$kappa)
       
       output <- list(pars.true = getpars(model), pars.start = getpars(model.start), pars.est = try(getpars(model.fit)), accuracy = acc)
       
@@ -194,7 +192,7 @@ for (ns in 2:4) {
     
     ncores <- detectCores()
     clust <- makeCluster(ncores)
-    clusterExport(clust, list("HMM_simulate", "D", "par.fix.true", "noise.sigma.int", "noise.kappa.int", "ss", "ns"))
+    clusterExport(clust, list("HMM_simulate", "cohen.kappa", "D", "par.fix.true", "noise.sigma.int", "noise.kappa.int", "ss", "ns"))
     
     
     # Iterate over intervals
@@ -276,7 +274,7 @@ for (ns in 2:4) {
       
       # Calculate accuracy
       
-      acc <- try(mean(model.sim@states == model.fit@posterior$state))
+      acc <- try(cohen.kappa(data.frame(x = model.sim@states, y = model.fit@posterior$state))$kappa)
       
       output <- list(pars.true = getpars(model), pars.start = getpars(model.start), pars.est = try(getpars(model.fit)), accuracy = acc)
       
@@ -316,7 +314,7 @@ for (ns in 2:4) {
     
     ncores <- detectCores()
     clust <- makeCluster(ncores)
[…]

### 2020-05-14 simulation/preregistration/preregistration_parameter_recovery.Rmd
update parameter recovery preregistration

@@
 Moreover, the HMM is multivariate but conditionally independent on the states.
-Thus, it has three independent response models for every state describing velocity, acceleration, and angle.
+Thus, every state has three independent response models describing velocity, acceleration, and angle.
 The velocity and acceleration response models for all four states use two-parameter gamma distributions (shape and scale parametrization).
@@
 
-Where it was possible, we chose the intervals and default values for each parameter based on reported ranges for eye movement events in the literature.
-Otherwise, intervals and default values were chosen according to theoretical reasons and based on insights from data used to develop the model.
+We define approximate ranges for each response variable and will choose true parameter intervals and default values so that they will produces samples that roughly correspond to these ranges.
 Table 1 shows the assumed ranges for each event and Table 2 shows the intervals and default values for each parameter in the simulation.
-Parameters were divided by 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
+Parameters will be scaled down by factor 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
 We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above five approach a symmetric distribution).
-The scale parameters will be set so that the respective distribution will match the assumed ranges.
+The scale parameters will be set so that the respective distribution will approximately match the assumed ranges.
 Since the concentration parameters of the von-Mises distribution are the inverse of standard deviations, they will be varied on the inverse scale.
@@
 
-For each parameter, we will calculate the root median square proportion deviation [RMSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$RMSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
-We will treat $RMSPD < 0.1$ as good, $0.1 \le RMSPD < 0.5$ as moderate, and $RMSPD \ge 5$ as bad recovery of a parameter.
-By taking the median, we will reduce the influence of potential outliers in the estimation and using proportions will enable us to compare RMSPD values across parameters and data sets.
+For each parameter, we will calculate the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$RMdSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
+We will treat $RMdSPD < 0.1$ as good, $0.1 \le RMdSPD < 0.5$ as moderate, and $RMdSPD \ge 5$ as bad recovery of a parameter.
+By taking the median, we will reduce the influence of potential outliers in the estimation and using proportions will enable us to compare RMdSPD values across parameters and data sets.
 
-Additionally, we will apply a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval.
+Additionally, we will apply a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in Part 1.
 Regression slopes closer to one will indicate that the model better captures parameter change.
-Regression intercepts different from zero will reflect bias in parameter estimation.
+Regression intercepts different from zero will reflect a bias in parameter estimation.
 
-To assess state recovery, we will compute Cohen's kappa (for all states taken together, not for each state seperately) as a measure of agreement between true and estimated states for each generated data set.
+To assess state recovery, we will compute Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set.
 Higher kappa values will be interpreted as better model accuracy.
@@
 
-*Assumed ranges of response variables used to generate parameter values*
+*Approximate ranges of response variables used to generate parameter values*
 
@@
 
-*Note*. Units are °/s (velocity), °/s^2^ (acceleration), and radians (angle). ~ indicates that the distribution has a peak at this value.
+*Note*.
+Units are °/s (velocity), °/s^2^ (acceleration), and radians (angle).
+~ indicates that the distribution has a peak at this value.
+Velocity ranges are based on event velocities reported in @Larsson2013.
+Since we would also like to test the algorithm on extreme data distributions, we extended the ranges beyond those found in typical eye movement data.
 
@@
 
-*HMM parameter values for generating the data*
+*HMM parameter values for generating data*
 

### 2020-05-15 simulation/preregistration/preregistration_parameter_recovery.Rmd
Fix typos

@@
 
-In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM.
+In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM), which is part of our recently developed algorithm gazeHMM.
 The purpose of this algorithm is to classify eye movements into distinct events.
@@
 In general, gazeHMM consists of three steps: First, the raw gaze data is preprocessed.
-Missing samples are identified, the data are filtered, and for each sample the velocity, acceleration, and difference in angle to the next sample is calculated.
-Second, a HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events.
+Missing samples are identified, the data are filtered, and for each sample, the velocity, acceleration, and difference in angle to the next sample is calculated.
+Second, an HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events.
 Moreover, the HMM computes the posterior probability of each sample for belonging to each event.
@@
 State one always corresponds to fixations, state two always to saccades, state three to PSOs, and state four to smooth pursuits.
-Moreover, the HMM is multivariate but conditionally independent on the states.
+The HMM is multivariate but conditionally independent on the states.
 Thus, every state has three independent response models describing velocity, acceleration, and angle.
@@
 We will compare the true with the estimated parameter values to assess whether a parameter was recovered by the model.
-Additionally, we will compare the true states of the HMM with the estimated states to judge how accurate the model recovers the states that generated the data.
+Additionally, we will compare the true states of the HMM with the estimated states to judge how accurately the model recovers the states that generated the data.
 
@@
 
-We define approximate ranges for each response variable and will choose true parameter intervals and default values so that they will produces samples that roughly correspond to these ranges.
+We define approximate ranges for each response variable and will choose true parameter intervals and default values so that they will produce samples that roughly correspond to these ranges.
 Table 1 shows the assumed ranges for each event and Table 2 shows the intervals and default values for each parameter in the simulation.
 Parameters will be scaled down by factor 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
-We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above five approach a symmetric distribution).
+We will set the intervals for shape parameters of the gamma distributions for all events to [1,5] to examine how skewness influences the recovery (shape values above five approach a symmetric distribution).
 The scale parameters will be set so that the respective distribution will approximately match the assumed ranges.
@@
 The model parameters will be set to their default values.
-For the shape, scale, and concentration parameters, we will simultaneously increase the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta_{start} = (\psi_{true}/2)\tau_{start}$ with $\tau_{start} \in \{1, 2, 3\}$, each 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries).
+For the shape, scale, and concentration parameters, we will simultaneously increase the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta_{start} = (\psi_{true}/2)\tau_{start}$ with $\tau_{start} \in \{1, 2, 3\}$, 100 data sets with $N = 2500$ samples will be generated each ($100 \times 3 \times 3 = 900$ recoveries).
 
@@
 For $k \in \{2, 3, 4\}$ states and $m \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries).
-The size of the missing data interval $l \in [1,200]$ samples will vary in equidistant steps between the data sets.
+The length of the missing data interval $l \in [1,200]$ samples will vary in equidistant steps between the data sets.
 
@@
 
-When a model can not be fitted to a data set, it will be excluded from the recovery.
+Models that can not be fitted will be excluded from the recovery.
 

### 2020-05-17 simulation/parameter_recovery_analysis.R
add plots for simulation results

@@ -138,6 +138,9 @@ for (part in 1:4) {
 
 rmsd.data <- lapply(rmsd, function(x) lapply(x, as.data.frame))
 rmsd.data <- lapply(rmsd.data, function(x) lapply(x, function(y) {as.data.frame(t(as.matrix(y)))}))
+rmsd.data <- lapply(rmsd.data, function(x) lapply(1:length(x), function(y, data) {
+  names(data[[y]]) <- names(rmsd[[1]][[y]][[1]])
+  return(data[[y]])}, data = x))
 
 
 # Calculate regression weights for transition probabilities
@@ -191,13 +194,6 @@ regw.tr <- lapply(get("estimates.1"), function(x) {
     
     return(out)
   })
-
-  reg <- lapply(df, function(z) {
-
-    lmfit <- lm(est ~ true, data = z)
-
-    return(lmfit)
-  })
 })
 
 
@@ -240,20 +236,173 @@ regw.resp <- lapply(get("estimates.1"), function(x) {
 
     return(out)
   })
-
-  reg <- lapply(df, function(z) {
-
-    lmfit <- lm(est ~ true, data = z)
-
-    return(lmfit)
-  })
 })
 
 
-# Plot MSE
+# Plot RMdSPD part 1
 
-lapply(mse.data, function(x) { lapply(x, function(y) {
+plots.rmsd.1 <- lapply(rmsd.data[[1]], function(x) {
   
-  ggplot(data = y)
+  names.pars.varied <- list(bquote(a["i=j"]), bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
+                            bquote(beta["acc"]), bquote(kappa))
   
-})})
+  if(ncol(x) == 18) {
+    
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]))
+    
+  } else if(ncol(x) == 30) {
+    
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]))
+    
+  } else {
+    
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]), bquote(a["i4"]))
+    
+  }
+  
+  names.pars.est <- append(trnames, list(bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
+                         bquote(beta["acc"]), bquote(mu), bquote(kappa)))
+  
[…]

### 2020-05-17 validation/validation_Andersson2017.R
create validation analysis for Andersson et al. (2017) data set

@@ -1,0 +1,83 @@
+### Validation Andersson et al. (2017) ###
+
+# Author: Malte Lüken
+# Date:17.05.2020
+
+library(R.matlab)
+library(tidyverse)
+library(depmixS4)
+library(parallel)
+
+source("algorithm/gazeHMM.R")
+source("algorithm/model_helper_functions.R")
+source("algorithm/plotting_functions.R")
+
+
+# Load data
+
+dir <- "~/Uni/Psychologie Master/Internship/Data/Test data/Andersson et al. (2017)/annotated_data/data used in the article"
+
+A2017 <- list()
+
+for(i in c("dots", "img", "video")) {
+  
+  filenames.MN <- list.files(path = paste(dir, "/", i, "/", sep = ""), pattern = "_MN")
+  filenames.RA <- list.files(path = paste(dir, "/", i, "/", sep = ""), pattern = "_RA")
+  
+  data.MN <- lapply(paste(dir, "/", i, "/", filenames.MN, sep = ""), function(x) readMat(x)$ETdata[[1]])
+  data.RA <- lapply(paste(dir, "/", i, "/", filenames.RA, sep = ""), function(x) readMat(x)$ETdata[[1]])
+  
+  A2017[[i]] <- lapply(1:length(data.MN), function(i, x, y) {
+
+    print(length(x[[i]]) - length(y[[i]]))
+
+    comb <- as.data.frame(x[[i]]) %>% left_join(as.data.frame(y[[i]]), by = names(.)[1:5])
+
+    names(comb) <- c("t", "h_pupil", "v_pupil", "x", "y", "label_MN", "label_RA")
+
+    return(comb)
+  }, x = data.MN, y = data.RA)
+}
+
+
+# Classify data
+
+res <- c(1024, 768)
+dim <- c(380, 300)
+dist <- 670
+fr <- 500
+
+ncores <- detectCores()
+clust <- makeCluster(ncores)
+clusterExport(clust, list("classify_gaze_data", "res", "dim", "dist", "fr", "A2017"))
+
+A2017.fit <- lapply(A2017, function(stim) parLapply(clust, stim, function(df) {
+  
+  if(all(is.nan(df$t) || df$t == 0)) {
+    
+    df$t <- seq(0, length(df$t)-1)/fr
+      
+  } else {
+    
+    df$t <- (df$t - df$t[1])/1e6
+    
+    df <- df[df$t >= 0,]
+    
+  }
+  
+  fit <- list()
+  
+  for (k in 2:5) {
+    
[…]

### 2020-05-18 algorithm/model_helper_functions.R
fix bug with exponentiated kappa parameter

@@ -277,7 +277,7 @@ setMethod("fit","vMF",
               
               miss <- is.na(y)
               
-              dens <- ifelse(miss, 1, dvm(y, mu = par[1], exp(kappa = par[2])))
+              dens <- ifelse(miss, 1, dvm(y, mu = par[1], kappa = exp(par[2])))
               
               return(-sum(w*log(dens)))
             }

### 2020-05-18 validation/validation_Andersson2017.R
add 1-state HMM

@@ -3,7 +3,6 @@
 # Author: Malte Lüken
 # Date:17.05.2020
 
-library(R.matlab)
 library(tidyverse)
 library(depmixS4)
 library(parallel)
@@ -11,6 +10,38 @@ library(parallel)
 source("algorithm/gazeHMM.R")
 source("algorithm/model_helper_functions.R")
 source("algorithm/plotting_functions.R")
+source("algorithm/preprocessing.R")
+
+
+# Function to fit 1-state HMM
+
+onestate_HMM <- function(data, respstart, sf = c(10, 10)) {
+  
+  source("algorithm/model_helper_functions.R")
+  
+  if(missing(respstart)) {
+    
+    respstart <- list(fix = list(vel = c(1, 1), acc = c(1, 1)))
+    
+  }
+  
+  if(all(sf > 0)) {
+    
+    data$vel <- data$vel/sf[1]
+    data$acc <- data$acc/sf[2]
+    
+  }
+  
+  velresp <- fit(altGamma(data$vel, pstart = respstart[["fix"]][["vel"]]))
+  accresp <- fit(altGamma(data$acc, pstart = respstart[["fix"]][["acc"]]))
+  angresp <- fit(unif(data$angle))
+  
+  ll <- sum(dens(velresp, log = T) + dens(accresp, log = T) + dens(angresp, log = T))
+  
+  n <- nrow(data)
+  
+  return(list(ll, n, response = list(velresp, accresp, angresp)))
+}
 
 
 # Load data
@@ -49,7 +80,7 @@ fr <- 500
 
 ncores <- detectCores()
 clust <- makeCluster(ncores)
-clusterExport(clust, list("classify_gaze_data", "res", "dim", "dist", "fr", "A2017"))
+clusterExport(clust, list("classify_gaze_data", "onestate_HMM", "preprocess", "res", "dim", "dist", "fr", "A2017"))
 
 A2017.fit <- lapply(A2017, function(stim) parLapply(clust, stim, function(df) {
   
@@ -67,14 +98,18 @@ A2017.fit <- lapply(A2017, function(stim) parLapply(clust, stim, function(df) {
   
   fit <- list()
   
+  fit[["1"]] <- try(onestate_HMM(preprocess(x = df$x, y = df$y, t = df$t, unit = "px",
+                                            res = res, dim = dim, dist = dist, fr = fr, blink = c(0, 0),
+                                            sg.order = 3, sg.length = 5)))
+  
   for (k in 2:5) {
     
[…]

### 2020-05-18 validation/validation_Ehinger2019.R
create validation analysis for Ehinger et al. (2019) data set

@@ -1,0 +1,127 @@
+### Validation Ehinger et al. (2019) ###
+
+# Author: Malte Lüken
+# Date: 18.05.2020
+
+library(tidyverse)
+library(depmixS4)
+library(parallel)
+library(FDBeye)
+library(eyelinker)
+
+source("algorithm/gazeHMM.R")
+source("algorithm/model_helper_functions.R")
+source("algorithm/plotting_functions.R")
+source("algorithm/preprocessing.R")
+
+
+# Function to fit 1-state HMM
+
+onestate_HMM <- function(data, respstart, sf = c(10, 10)) {
+  
+  source("algorithm/model_helper_functions.R")
+  
+  if(missing(respstart)) {
+    
+    respstart <- list(fix = list(vel = c(1, 1), acc = c(1, 1)))
+    
+  }
+  
+  if(all(sf > 0)) {
+    
+    data$vel <- data$vel/sf[1]
+    data$acc <- data$acc/sf[2]
+    
+  }
+  
+  velresp <- fit(altGamma(data$vel, pstart = respstart[["fix"]][["vel"]]))
+  accresp <- fit(altGamma(data$acc, pstart = respstart[["fix"]][["acc"]]))
+  angresp <- fit(unif(data$angle))
+  
+  ll <- sum(dens(velresp, log = T) + dens(accresp, log = T) + dens(angresp, log = T))
+  
+  n <- nrow(data)
+  
+  return(list(ll, n, response = list(velresp, accresp, angresp)))
+}
+
+
+# Load Eyelink data
+
+dir <- "~/Uni/Psychologie Master/Internship/Data/Test data/Ehinger et al. (2019)/"
+
+
+# # Convert .EDF files into .asc files
+# 
+# setwd("~/Uni/Psychologie Master/Internship/Data/Test data/Ehinger et al. (2019)/")
+# 
+# Sys.getenv("C:/Programme (x86)/SR Research/EyeLink/EDF_Access_API/Example", names = T)
+# 
+# edf.filenames <- list.files(path = getwd(), pattern = ".EDF")
+# 
+# converted <- edf2asc(edf.filenames)
+
+  
+filenames.el <- list.files(path = dir, pattern = ".asc")
+  
+E2019 <- lapply(paste(dir, "/", filenames.el, sep = ""), function(x) { 
+  
+  data <- read.asc(x) 
+  
+  task4 <- list()
+  task5 <- list()
+  
+  for (b in 1:6) {
+    
[…]

### 2020-05-19 simulation/parameter_recovery_simulation.R
add seeds for reproducibility

@@ -14,6 +14,11 @@ source("simulation/model_simulation.R")
 source("algorithm/model_helper_functions.R")
 
 
+# Record starting time
+
+time.begin <- Sys.time()
+
+
 # Number of data sets to be generated
 
 D <- 5
@@ -76,22 +81,28 @@ for (ns in 2:4) {
     
     ncores <- detectCores()
     clust <- makeCluster(ncores)
-    clusterExport(clust, list("HMM_simulate", "cohen.kappa", "N", "par.fix.true", "par.states", "par.names", "par.resps", "p", "ns"))
+    clusterExport(clust, list("HMM_simulate", "cohen.kappa", "D", "N", "par.int.true", "par.fix.true", "par.states", "par.names", "par.resps", "p", "ns"))
     
     
     # Iterate over intervals
     
-    estimates.1[[as.character(ns)]][[p]] <- parLapply(clust, par.int.true[,p], function(x, samples = N, k = ns, 
-                                                                                        par.state = par.states[p], 
-                                                                                        par.name = par.names[p], 
-                                                                                        par.resp = par.resps[p], 
-                                                                                        trueResp = par.fix.true) {
+    estimates.1[[as.character(ns)]][[p]] <- parLapply(clust, 1:D, function(x, samples = N, k = ns,
+                                                                        par.int = par.int.true[,p],
+                                                                        par.state = par.states[p], 
+                                                                        par.name = par.names[p], 
+                                                                        par.resp = par.resps[p], 
+                                                                        trueResp = par.fix.true) {
+      
+      # Generate data seed
+      
+      seed.data <- k*1e6 + p*1e4 + x*1e2
+      
       
       # Set true parameter values
       
       if(par.name > 0 && par.resp > 0) {
         
-        trueResp[[par.state]][[par.resp]][par.name] <- x
+        trueResp[[par.state]][[par.resp]][par.name] <- par.int[x]
         
[…]

### 2020-05-19 simulation/preregistration/preregistration_parameter_recovery.Rmd
fix typos

@@
 For each parameter, we will calculate the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$RMdSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
-We will treat $RMdSPD < 0.1$ as good, $0.1 \le RMdSPD < 0.5$ as moderate, and $RMdSPD \ge 5$ as bad recovery of a parameter.
+We will treat $RMdSPD < 0.1$ as good, $0.1 \le RMdSPD < 0.5$ as moderate, and $RMdSPD \ge 0.5$ as bad recovery of a parameter.
 By taking the median, we will reduce the influence of potential outliers in the estimation and using proportions will enable us to compare RMdSPD values across parameters and data sets.
@@
 
-***
+\newpage
 

### 2020-05-19 simulation/parameter_recovery_simulation.R

@@ -570,7 +570,7 @@ save("estimates.4", file = "simulation/part4.Rdata")
 
 # Record ending time
 
-time.end <- Sys.time
+time.end <- Sys.time()
 
 
 # Record session info 

### 2020-05-21 simulation/parameter_recovery_simulation.R
final simulation (100 data sets)

@@ -21,7 +21,7 @@ time.begin <- Sys.time()
 
 # Number of data sets to be generated
 
-D <- 5
+D <- 100
 
 
 # Default response parameter values

### 2020-05-21 simulation/parameter_recovery_analysis.R
add accuracy plots

@@ -57,6 +57,7 @@ linkfun <- function(p, base) {
   return(beta)
 }
 
+
 # Function to transform parameters to normal scale
 
 backtrans <- function(x) {
@@ -239,6 +240,33 @@ regw.resp <- lapply(get("estimates.1"), function(x) {
 })
 
 
+# Summarise accuracy
+
+acc.data <- list()
+
+for (part in 1:4) {
+  
+  acc.data[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
+    out <- lapply(x, function(y) { 
+      out <- lapply(y, function(z) {
+        
+        if(is.numeric(z$accuracy)) {
+          acc <- z$accuracy
+        } else {
+          acc <- NA
+        }
+        
+        return(acc)
+      })
+      
+      return(as.vector(reduce(out, cbind)))
+    })
+    
+    return(as.data.frame(t(reduce(out, cbind))))
+  })
+}
+
+
 # Plot RMdSPD part 1
 
 plots.rmsd.1 <- lapply(rmsd.data[[1]], function(x) {
@@ -351,12 +379,12 @@ plots.rmsd.234 <- lapply(2:4, function(y, data) lapply(data[[y]], function(x) {
   return(p)
 }), data = rmsd.data)
 
-print(plots.rmsd.234[[3]][[1]])
+print(plots.rmsd.234[[3]][[3]])
 
 
 # Plot linear regressions for transition probabilities
 
-D <- 5
+D <- 100
 
 regw.tr.data <- lapply(regw.tr, function(x) lapply(x, function(y) rbind(y)))
 regw.tr.data <- lapply(regw.tr.data, function(x) reduce(x, rbind))
@@ -376,7 +404,7 @@ plots.lm.tr <- lapply(1:length(regw.tr.data), function(x, y) {
   return(p)
 }, y = regw.tr.data)
 
-print(plots.lm.tr[[2]])
+print(plots.lm.tr[[3]])
 
 
 # Plot linear regressions for response parameters
@@ -406,3 +434,64 @@ plots.lm.resp <- lapply(regw.resp.data, function(x) {
 })
 
 print(plots.lm.resp[[3]])
+
+
+# Plot accuracy part 1
+
+plots.acc.1 <- lapply(acc.data[[1]], function(x) {
+  
+  names.pars.varied <- list(bquote(a["i=j"]), bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
[…]

### 2020-05-22 simulation/parameter_recovery_analysis.R
add erroneous model plots

@@ -74,9 +74,6 @@ backtrans <- function(x) {
   
   return(out)
 }
-mlogit
-
-#backtrans(estimates.1$`3`[[1]][[3]]$pars.est)
 
 
 # Calculate normalized root mean square deviation
@@ -368,7 +365,7 @@ plots.rmsd.234 <- lapply(2:4, function(y, data) lapply(data[[y]], function(x) {
     label.cond <- "N"
   } else if (y == 3) {
     names.cond <- c("1", "2", "3")
-    label.cond <- bquote(beta["start"])
+    label.cond <- bquote(tau["start"])
   } else {
     names.cond <- c("1", "3", "5")
     label.cond <- "m"
@@ -379,7 +376,7 @@ plots.rmsd.234 <- lapply(2:4, function(y, data) lapply(data[[y]], function(x) {
   return(p)
 }), data = rmsd.data)
 
-print(plots.rmsd.234[[3]][[3]])
+print(plots.rmsd.234[[1]][[1]])
 
 
 # Plot linear regressions for transition probabilities
@@ -483,7 +480,7 @@ plots.acc.234 <- lapply(1:3, function(y, data) {
     label.cond <- "N"
   } else if (y == 2) {
     names.cond <- c("1", "2", "3")
-    label.cond <- bquote(beta["start"])
+    label.cond <- bquote(tau["start"])
   } else {
     names.cond <- c("1", "3", "5")
     label.cond <- "m"
@@ -494,4 +491,114 @@ plots.acc.234 <- lapply(1:3, function(y, data) {
   return(p)
 }, data = acc.data.234)
 
-print(plots.acc.234[[3]])
+print(plots.acc.234[[2]])
+
+
+# Plot accuracy over interval parts 2 and 4
+
+plots.acc.int.24 <- lapply(c(1, 3), function(y, data) {
+  
+  x <- as_tibble(data[[y]], .name_repair = "unique")
+  
+  if(y == 1) {
+  
+    int <- rep(seq(1, 5, length.out = D), 9) 
+  
+  } else {
+    
+    int <- rep(floor(seq(1, 200, length.out = D)), 9)
+    
+  }
+  
+  data.long <- x %>%
+    mutate(cond = as.factor(rep(1:3, 3)),
+           k = as.factor(rep(2:4, each = 3))) %>%
+    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
+    mutate(int = int)
+  
[…]

### 2020-06-03 simulation/parameter_recovery_analysis.R
add exploratory analysis label switching

@@ -602,3 +602,71 @@ plots.err.234 <- lapply(1:3, function(y, data) {
 }, data = acc.data.234)
 
 print(plots.err.234[[1]])
+
+
+# Exploratory analysis label switching ------------------------------------
+
+library(psych)
+
+load("simulation/part3_expl.Rdata")
+
+labsw <- lapply(1:length(estimates.3), function(x) {
+  lapply(estimates.3[[x]], function(y) {
+    lapply(y, function(z) {
+      
+      kappa <- try(cohen.kappa(z$states)$kappa)
+      
+      states <- z$states
+      
+      if(x == 2 & is.numeric(kappa) & kappa < 0.95) {
+          
+        states$y <- ifelse(z$states$y == 3, 2, ifelse(z$states$y == 2, 3, z$states$y))
+          
+        kappa <- cohen.kappa(states)$kappa
+        
+        if(is.numeric(kappa) & kappa < 0.95) {
+          
+          states$y <- ifelse(z$states$y == 3, 1, ifelse(z$states$y == 1, 3, z$states$y))
+          
+          kappa <- cohen.kappa(states)$kappa
+        }
+        
+        if(is.numeric(kappa) & kappa < 0.95) {
+          
+          states$y <- ifelse(z$states$y == 3, 2, ifelse(z$states$y == 1, 3, 1))
+          
+          kappa <- cohen.kappa(states)$kappa
+        }
+      }
+      
+      if(x == 3 & is.numeric(kappa) & kappa < 0.5) {
+        
+        states$y <- ifelse(z$states$y == 3, 2, ifelse(z$states$y == 2, 3, z$states$y))
+        
+        kappa <- cohen.kappa(states)$kappa
+        
+        if(is.numeric(kappa) & kappa < 0.5) {
+
+          states$y <- ifelse(z$states$y == 3, 4, ifelse(z$states$y == 4, 3, z$states$y))
+
+          kappa <- cohen.kappa(states)$kappa
+        }
+
+        if(is.numeric(kappa) & kappa < 0.5) {
+
+          states$y <- ifelse(z$states$y == 3, 1, ifelse(z$states$y == 1, 3, z$states$y))
+
+          kappa <- cohen.kappa(states)$kappa
+        }
+        
+        if(is.numeric(kappa) & kappa < 0.5) {
+          
+          kappa <- cohen.kappa(z$states)$kappa
+        }
+      }
+      
+      return(kappa)
+    })
+  })
+})
+

### 2020-06-03 simulation/results.Rmd
upload results simulation and validation

---
title: "Results parameter recovery"
author: "Malte Lüken"
date: "22.05.2020"
output: 
  ioslides_presentation:
    widescreen: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r load data and prepare functions, include=FALSE}

library(tidyverse)
library(depmixS4)

for (part in 1:4) {

  load(file = paste("~/Uni/Psychologie Master/Internship/GazeHMM/simulation/part", part, ".Rdata", sep = ""))

}

linkinv <- function(eta,base) {
  linv <- function(eta,base) {
    pp <- numeric(length(eta))
    if(any(is.infinite(eta)) || any(eta > log(.Machine$double.xmax)) || any(eta < log(.Machine$double.xmin))) {
      pp[which(is.infinite(eta))] <- 1
      pp[which(eta > log(.Machine$double.xmax))] <- 1 # change this to something better!
    } else {
      expb <- exp(eta)
      sumb <- sum(expb)
      pp[base] <- 1/sumb
      pp[-base] <- expb[-base]/sumb
    }
    return(pp)
  }
  if(is.matrix(eta)) {
    if(ncol(eta)==1) {
      pp <- as.matrix(apply(eta,1,linv,base=base)) # fixes problem with column matrix eta
    } else pp <- t(apply(eta,1,linv,base=base)) 	
  } else {
    pp <- linv(eta,base)
  }
  return(pp)
}
linkfun <- function(p, base) {
  lfun <- function(p, base) {
    p <- p/sum(p)
    beta <- numeric(length(p))
    if (any(p == 1)) 
      beta[which(p == 1)] = Inf
    else beta[-base] <- log(p[-base]/p[base])
    return(beta)
  }
  if (is.matrix(p)) {
    beta <- t(apply(p, 1, lfun, base = base))
  }
  else {
    beta <- lfun(p, base)
  }
  return(beta)
}

# Function to transform parameters to normal scale

backtrans <- function(x) {

  out <- x

  nms <- names(x)

  out[str_detect(nms, "(Intercept)")] <- as.vector(apply(matrix(x[str_detect(nms, "(Intercept)")],
                                                        ncol = sqrt(length(x[str_detect(nms, "(Intercept)")])),
                                                        byrow = T), 1, linkinv, base = 1))

  out[nms %in% c("shape", "scale", "kappa")] <- exp(x[nms %in% c("shape", "scale", "kappa")])

  return(out)
}

```

# Root Median Square Proportion Deviation (RMdSPD)

```{r calculate RMdSPD, include=FALSE}

rmsd <- list()

for (part in 1:4) {

  rmsd[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
    lapply(x, function(y) {
      sqerr <- lapply(y, function(z) {

        err <- try(((backtrans(z$pars.est) - backtrans(z$pars.true))/
                      ifelse(backtrans(z$pars.true) == 0, 2*pi, backtrans(z$pars.true)))^2)

        if(is.numeric(err)) {

          out <- err

        } else {

          out <- rep(NA, length(z$pars.true))

        }

        names(out) <- names(z$pars.true)

        return(out)
      })

      rows <- length(sqerr)

      nms <- names(sqerr[[1]])

      pars <- matrix(unlist(sqerr), nrow = rows, byrow = T)

      msqerr <- apply(pars, 2, median, na.rm = T)

      names(msqerr) <- nms

      rmsqerr <- sqrt(msqerr)

      return(rmsqerr)
    })
  })
}

# Display MSE in data frame

rmsd.data <- lapply(rmsd, function(x) lapply(x, as.data.frame))
rmsd.data <- lapply(rmsd.data, function(x) lapply(x, function(y) {as.data.frame(t(as.matrix(y)))}))
rmsd.data <- lapply(rmsd.data, function(x) lapply(1:length(x), function(y, data) {
  names(data[[y]]) <- names(rmsd[[1]][[y]][[1]])
  return(data[[y]])}, data = x))

plots.rmsd.1 <- lapply(rmsd.data[[1]], function(x) {

  names.pars.varied <- list(bquote(a["i=j"]), bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
                            bquote(beta["acc"]), bquote(kappa))

  if(ncol(x) == 18) {

    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]))

  } else if(ncol(x) == 30) {

    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]))

  } else {

    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]), bquote(a["i4"]))

  }

  names.pars.est <- append(trnames, list(bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
                         bquote(beta["acc"]), bquote(mu), bquote(kappa)))

  x <- as_tibble(x, .name_repair = "unique")

  data.long <- x %>%
    mutate(par.varied = c(0, 1, 2, 3, 4, rep(c(1, 2, 3, 4, 5), (nrow(x) %/% 5)-1)),
           state.varied = c(1, rep(1, 4), rep(2:((nrow(x) %/% 5)), each = 5))) %>%
    pivot_longer(names(x), names_to = "par.est", values_to = "RMdSPD", names_repair = "unique") %>%
    mutate(RMdSPD = cut(RMdSPD, breaks = c(0, 0.1, 0.5, Inf), c("good", "moderate", "bad")),
           state.est = rep(c(rep(1:max(state.varied), max(state.varied)+1),
                         rep(1:max(state.varied), each = 6)), nrow(x)),
           par.est = rep(c(rep(1, max(state.varied)), 
                           rep(2:(max(state.varied)+1), each = max(state.varied)), 
                           rep((max(state.varied)+2):(max(state.varied)+7), max(state.varied))), nrow(x))) %>%
    mutate_at(vars(par.varied, state.varied, par.est, state.est), as.factor)

  p <- ggplot(data = data.long, aes(x = par.varied, y = par.est, fill = RMdSPD)) +
    geom_tile() + facet_grid(cols = vars(state.varied), rows = vars(state.est)) +
    scale_x_discrete(labels = names.pars.varied) +
    scale_y_discrete(labels = names.pars.est)
    theme(axis.text = element_text(size = 14))

  return(p)
})

plots.rmsd.234 <- lapply(2:4, function(y, data) lapply(data[[y]], function(x) {

  if(ncol(x) == 18) {

    k <- 2

    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]))

  } else if(ncol(x) == 30) {

    k <- 3

    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]))

  } else {

    k <-4

    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]), bquote(a["i4"]))

  }

  names.pars.est <- append(trnames, list(bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
                                         bquote(beta["acc"]), bquote(mu), bquote(kappa)))

  x <- as_tibble(x, .name_repair = "unique")

  data.long <- x %>%
    mutate(cond = 1:3) %>%
    pivot_longer(names(x), names_to = "par.est", values_to = "RMdSPD", names_repair = "unique") %>%
    mutate(state.est = rep(c(rep(1:k, k+1),
                             rep(1:k, each = 6)), nrow(x)),
           par.est = rep(c(rep(1, k), 
                           rep(2:(k+1), each = k), 
                           rep((k+2):(k+7), k)), nrow(x))) %>%
    mutate_at(vars(cond, par.est, state.est), as.factor)

  p <- ggplot(data = data.long, aes(x = par.est, y = RMdSPD, color = cond)) +
    geom_point(position = position_dodge(0.25)) + facet_grid(cols = vars(state.est)) +
    scale_x_discrete(labels = names.pars.est) +
    scale_y_continuous(breaks = c(0.1, 0.5, 1, 5, 10)) +
    geom_hline(yintercept = 0.1, linetype = "dashed") +
    geom_hline(yintercept = 0.5, linetype = "dashed")

  if(y == 2) {
    names.cond <- c("500", "2500", "10000")
    label.cond <- "N"
  } else if (y == 3) {
    names.cond <- c("1", "2", "3")
    label.cond <- bquote(tau["start"])
  } else {
    names.cond <- c("1", "3", "5")
    label.cond <- "m"
  }

  p <- p + scale_color_discrete(name = label.cond, labels = names.cond)

  return(p)
}), data = rmsd.data)

```

## Part 1
### 2 States

```{r plot part 1 2 states}

print(plots.rmsd.1[[1]])

```

## Part 1
### 3 States

```{r plot part 1 3 states}

print(plots.rmsd.1[[2]])

```

## Part 1
### 4 States

```{r plot part 1 4 states}

print(plots.rmsd.1[[3]])

```

## Part 2
### 2 States

```{r plot part 2 2 states}

print(plots.rmsd.234[[1]][[1]])

```

## Part 2
### 3 States

```{r plot part 2 3 states}

print(plots.rmsd.234[[1]][[2]])

```

## Part 2
### 4 States

```{r plot part 2 4 states}

print(plots.rmsd.234[[1]][[3]])

```

## Part 3
### 2 States

```{r plot part 3 2 states}

print(plots.rmsd.234[[2]][[1]])

```

## Part 3
### 3 States

```{r plot part 3 3 states}

print(plots.rmsd.234[[2]][[2]])

```

## Part 3
### 4 States

```{r plot part 3 4 states}

print(plots.rmsd.234[[2]][[3]])

```

## Part 4
### 2 States

```{r plot part 4 2 states}

print(plots.rmsd.234[[3]][[1]])

```

## Part 4
### 3 States

```{r plot part 4 3 states}

print(plots.rmsd.234[[3]][[2]])

```

## Part 4
### 4 States

```{r plot part 4 4 states}

print(plots.rmsd.234[[3]][[3]])

```

# Linear Regression Between Estimated and True Parameters

```{r calculate linear regressions, include=F}

# Calculate regression weights for transition probabilities

regw.tr <- list()

regw.tr <- lapply(get("estimates.1"), function(x) {

  varpar <- lapply(x[1], function(y) {

    lapply(y, function(z) {

      nms <- names(z$pars.true)

      pars.tr <- logical(length(z$pars.true))

      pars.tr[str_detect(nms, "(Intercept)")] <- T

      intpar <- try(cbind(backtrans(z$pars.true[pars.tr]), backtrans(z$pars.est[pars.tr])))

      if(is.numeric(intpar)) {
        out <- intpar
      } else {
        out <- matrix(NA, nrow = length(z$pars.true[pars.tr]), ncol = 2)
      }

      out <- apply(out, 1, list)

      return(out)
    })
  })

  df <- list()

  for (i in 1:length(varpar[[1]][[1]])) {

    df[[i]] <- lapply(varpar, function(y, index) {

      lapply(y, function(z) {z[[index]]})

    }, index = i)
  }

  df <- lapply(df, as.data.frame)

  df <- lapply(df, function(z) {

    out <- as.data.frame(t(as.matrix(z)))

    names(out) <- c("true", "est")

    return(out)
  })
})

# Calculate linear regression weights for response parameters

regw.resp <- list()

regw.resp <- lapply(get("estimates.1"), function(x) {

  index <- 1:length(x[-1])

  varpar <- lapply(index, function(i, y) {

    lapply(y[[i]], function(z) {

      nms <- names(z$pars.true)

      #pars.tr <- logical(length(z$pars.true))
      pars.resp <- logical(length(z$pars.true))

      #pars.tr[str_detect(nms, "(Intercept)")] <- T
      pars.resp[nms %in% c("shape", "scale", "kappa")] <- T

      intpar <- try(c(backtrans(z$pars.true[pars.resp][i]), backtrans(z$pars.est[pars.resp][i])))

      if(is.numeric(intpar)) {
        return(intpar)
      } else {
        return(rep(NA, 2))
      }
    })
  }, y = x[-1])

  df <- lapply(varpar, as.data.frame)
  df <- lapply(df, function(z) {

    out <- as.data.frame(t(as.matrix(z)))

    names(out) <- c("true", "est")

    return(out)
  })
})

# Plot linear regressions for transition probabilities

D <- 100

regw.tr.data <- lapply(regw.tr, function(x) lapply(x, function(y) rbind(y)))
regw.tr.data <- lapply(regw.tr.data, function(x) reduce(x, rbind))

plots.lm.tr <- lapply(1:length(regw.tr.data), function(x, y) {

  data <- y[[x]] %>% 
    mutate(par = rep(1:(x+1)^2, each = D),
           from = rep(rep(1:(x+1), each = D), x+1),
           to = rep(1:(x+1), each = D*(x+1)))

  p <- ggplot(data, aes(x = true, y = est)) + 
    facet_grid(rows = vars(from), cols = vars(to), labeller = label_both) +
    geom_point() + geom_smooth(method = "lm") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed")

  return(p)
}, y = regw.tr.data)

# Plot linear regressions for response parameters

regw.resp.data <- lapply(regw.resp, function(x) lapply(x, function(y) rbind(y)))
regw.resp.data <- lapply(regw.resp.data, function(x) reduce(x, rbind))

parnames <- list(bquote(alpha["vel;1"]), bquote(beta["vel;1"]), bquote(alpha["acc;1"]), bquote(beta["acc;1"]),
                 bquote(alpha["vel;2"]), bquote(beta["vel;2"]), bquote(alpha["acc;2"]), bquote(beta["acc;2"]), bquote(kappa["2"]),
                 bquote(alpha["vel;3"]), bquote(beta["vel;3"]), bquote(alpha["acc;3"]), bquote(beta["acc;3"]), bquote(kappa["3"]),
                 bquote(alpha["vel;4"]), bquote(beta["vel;4"]), bquote(alpha["acc;4"]), bquote(beta["acc;4"]), bquote(kappa["4"]))

plots.lm.resp <- lapply(regw.resp.data, function(x) {

  npar <- nrow(x) %/% D

  data <- x %>% 
    mutate(par = rep(1:npar, each = D),
           type = rep(1, 2, 1, 2))

  p <- ggplot(data, aes(x = true, y = est)) + 
    facet_wrap(vars(par), scales = "free", labeller = label_bquote(.(parnames[[par]]))) +
    geom_point() + geom_smooth(method = "lm") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed")

  return(p)
})

```

## Transition probabilities
### 2 States

```{r}

print(plots.lm.tr[[1]])

```

## Transition probabilities
### 3 States

```{r}

print(plots.lm.tr[[2]])

```

## Transition probabilities
### 4 States

```{r}

print(plots.lm.tr[[3]])

```

## Response parameters
### 2 States

```{r}

print(plots.lm.resp[[1]])

```

## Response parameters
### 3 States

```{r}

print(plots.lm.resp[[2]])

```

## Response parameters
### 4 States

```{r}

print(plots.lm.resp[[3]])

```

# Accuracy (Cohen's Kappa)

```{r calculate accuracy, include=FALSE}

# Summarise accuracy

acc.data <- list()

for (part in 1:4) {

  acc.data[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
    out <- lapply(x, function(y) { 
      out <- lapply(y, function(z) {

        if(is.numeric(z$accuracy)) {
          acc <- z$accuracy
        } else {
          acc <- NA
        }

        return(acc)
      })

      return(as.vector(reduce(out, cbind)))
    })

    return(as.data.frame(t(reduce(out, cbind))))
  })
}

# Plot accuracy part 1

plots.acc.1 <- lapply(acc.data[[1]], function(x) {

  names.pars.varied <- list(bquote(a["i=j"]), bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
                            bquote(beta["acc"]), bquote(kappa))

  x <- as_tibble(x, .name_repair = "unique")

  data.long <- x %>%
    mutate(par.varied = c(0, 1, 2, 3, 4, rep(c(1, 2, 3, 4, 5), (nrow(x) %/% 5)-1)),
           state.varied = c(1, rep(1, 4), rep(2:((nrow(x) %/% 5)), each = 5))) %>%
    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
    mutate_at(vars(par.varied, state.varied), as.factor)

  p <- ggplot(data = data.long, aes(x = par.varied, y = accuracy)) +
    geom_boxplot() + facet_grid(cols = vars(state.varied)) +
    scale_x_discrete(labels = names.pars.varied)

  return(p)
})

# Plot accuracy parts 2, 3, and 4

acc.data.234 <- lapply(acc.data[2:4], function(x) reduce(x, rbind))

plots.acc.234 <- lapply(1:3, function(y, data) {

  x <- as_tibble(data[[y]], .name_repair = "unique")

  data.long <- x %>%
    mutate(cond = as.factor(rep(1:3, 3)),
           k = as.factor(rep(2:4, each = 3))) %>%
    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy")

  p <- ggplot(data = data.long, aes(x = k, y = accuracy, color = cond)) +
    geom_boxplot() + 
    scale_x_discrete(name = "k (number of states)")

  if(y == 1) {
    names.cond <- c("500", "2500", "10000")
    label.cond <- "N"
  } else if (y == 2) {
    names.cond <- c("1", "2", "3")
    label.cond <- bquote(tau["start"])
  } else {
    names.cond <- c("1", "3", "5")
    label.cond <- "m"
  }

  p <- p + scale_color_discrete(name = label.cond, labels = names.cond)

  return(p)
}, data = acc.data.234)

# Plot accuracy over interval parts 2 and 4

plots.acc.int.24 <- lapply(c(1, 3), function(y, data) {

  x <- as_tibble(data[[y]], .name_repair = "unique")

  if(y == 1) {

    int <- rep(seq(1, 5, length.out = D), 9)

  } else {

    int <- rep(floor(seq(1, 200, length.out = D)), 9)

  }

  data.long <- x %>%
    mutate(cond = as.factor(rep(1:3, 3)),
           k = as.factor(rep(2:4, each = 3))) %>%
    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
    mutate(int = int)

  p <- ggplot(data = data.long, aes(x = int, y = accuracy, color = cond)) +
    facet_grid(cols = vars(k)) + 
    geom_point() # + geom_smooth(method = "lm")

  if(y == 1) {
    names.cond <- c("500", "2500", "10000")
    label.cond <- "N"
    x.name <- bquote(tau["noise"])
  } else {
    names.cond <- c("1", "3", "5")
    label.cond <- "m"
    x.name <- "l"
  }

  p <- p + scale_color_discrete(name = label.cond, labels = names.cond) +
    scale_x_continuous(name = x.name)

  return(p)
}, data = acc.data.234)

```

## Part 1
### 2 States

```{r}

print(plots.acc.1[[1]])

```

## Part 1
### 3 States

```{r}

print(plots.acc.1[[2]])

```

## Part 1
### 4 States

```{r}

print(plots.acc.1[[3]])

```

## Part 2

```{r}

print(plots.acc.int.24[[1]])

```

## Part 3

```{r}

print(plots.acc.234[[2]])

```

## Part 4

```{r}

print(plots.acc.int.24[[2]])

```

# Erroneous Models

```{r calculate proportion of erroneous models, include=FALSE}

# Plot proportion of erroneous models part 1

plots.err.1 <- lapply(acc.data[[1]], function(x) {

  names.pars.varied <- list(bquote(a["i=j"]), bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
                            bquote(beta["acc"]), bquote(kappa))

  x <- as_tibble(x, .name_repair = "unique")

  data.long <- x %>%
    mutate(par.varied = c(0, 1, 2, 3, 4, rep(c(1, 2, 3, 4, 5), (nrow(x) %/% 5)-1)),
           state.varied = c(1, rep(1, 4), rep(2:((nrow(x) %/% 5)), each = 5))) %>%
    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
    mutate_at(vars(par.varied, state.varied), as.factor) %>%
    group_by(par.varied, state.varied) %>%
    summarise(error = mean(is.na(accuracy)))

  p <- ggplot(data = data.long, aes(x = par.varied, y = error)) +
    geom_point() + facet_grid(cols = vars(state.varied)) +
    scale_x_discrete(labels = names.pars.varied) + 
    scale_y_continuous(name = "proportion erroneous models", limits = c(0, 1))

  return(p)
})

# Plot proportion of erroneous models parts 2, 3, 4

plots.err.234 <- lapply(1:3, function(y, data) {

  x <- as_tibble(data[[y]], .name_repair = "unique")

  data.long <- x %>%
    mutate(cond = as.factor(rep(1:3, 3)),
           k = as.factor(rep(2:4, each = 3))) %>%
    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
    group_by(cond, k) %>%
    summarise(error = mean(is.na(accuracy)))

  p <- ggplot(data = data.long, aes(x = k, y = error, color = cond)) +
    geom_point(position = position_dodge(width = 0.25)) + 
    scale_x_discrete(name = "k (number of states)") +
    scale_y_continuous(name = "proportion erroneous models", limits = c(0, 1))

  if(y == 1) {
    names.cond <- c("500", "2500", "10000")
    label.cond <- "N"
  } else if (y == 2) {
    names.cond <- c("1", "2", "3")
    label.cond <- bquote(tau["start"])
  } else {
    names.cond <- c("1", "3", "5")
    label.cond <- "m"
  }

  p <- p + scale_color_discrete(name = label.cond, labels = names.cond)

  return(p)
}, data = acc.data.234)

print(plots.err.234[[1]])

```

## Part 1
### 2 States

```{r}

print(plots.err.1[[1]])

```

## Part 1
### 3 States

```{r}

print(plots.err.1[[2]])

```

## Part 1
### 4 States

```{r}

print(plots.err.1[[3]])

```

## Part 2

```{r}

print(plots.err.234[[1]])

```

## Part 3

```{r}

print(plots.err.234[[2]])

```

## Part 4

```{r}

print(plots.err.234[[3]])

```

### 2020-06-03 validation/results.Rmd
upload results simulation and validation

---
title: "Results validation"
author: "Malte Lüken"
date: "31.05.2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(psych)
library(depmixS4)
library(knitr)
library(kableExtra)

source("~/Uni/Psychologie Master/Internship/GazeHMM/algorithm/model_helper_functions.R")

```

# Data set Andersson et al. (2017)

```{r load data and compute metrics, include=FALSE}

# Load data

load("~/Uni/Psychologie Master/Internship/GazeHMM/validation/Andersson2017_fitted.Rdata")
load("~/Uni/Psychologie Master/Internship/GazeHMM/validation/Andersson2017_raw.Rdata")

# Compute Schwarz weights

schwarz.weights <- function(bic, na.rm = T) {

  d.bic <- bic - min(bic, na.rm = na.rm) # eq 2

  exp(-0.5 * d.bic)/sum(exp(-0.5 * d.bic), na.rm = na.rm) # eq 4

}

A2017.bic <- lapply(A2017.fit, function(stim) {
  out <- lapply(stim, function(subj) {
    out <- lapply(1:length(subj), function(mod) {

      if(mod == 1) {

        bic <- try(-2*subj[[mod]][["LL"]] + 6*log(subj[[mod]][["N"]]))

      } else {

        bic <- try(BIC(subj[[mod]]$model))

      }

      return(ifelse(is.numeric(bic), bic, NA))
    })

    return(schwarz.weights(unlist(out)))
  })

  df <- as.data.frame(reduce(out, rbind))

  names(df) <- paste("model_", 1:length(stim[[1]]), sep = "")

  return(df)
})

# Create Schwarz weight plots

bic.plot <- lapply(A2017.bic, function(x) {

  data.long <- x %>% 
    mutate(subject = 1:nrow(x)) %>%
    pivot_longer(names(x), names_to = "model", values_to = "weight") %>%
    mutate_at(c("subject", "model"), as.factor)

  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
    scale_fill_continuous(name = "Schwarz\nweight") +
    scale_x_discrete(name = "Number of states in model", labels = as.character(1:5))

  return(p)
})

# Calculate event descriptives for human coders

fr <- 500

A2017.events <- lapply(A2017, function(stim) {
  lapply(1:4, function(e) {
    out <- lapply(stim, function(df) {

      if(all(is.nan(df$t) || df$t == 0)) {

        df$t <- seq(0, length(df$t)-1)/fr

      } else {

        df$t <- (df$t - df$t[1])/1e6

        df <- df[df$t >= 0,]

      }

      counter_MN <- 1
      counter_RA <- 1

      number_MN <- numeric(nrow(df))
      number_RA <- numeric(nrow(df))

      number_MN[1] <- 1
      number_RA[1] <- 1

      for (i in 2:nrow(df)) {
        if(df$label_MN[i] != df$label_MN[i-1]) counter_MN <- counter_MN + 1
        if(df$label_RA[i] != df$label_RA[i-1]) counter_RA <- counter_RA + 1

        number_MN[i] <- counter_MN
        number_RA[i] <- counter_RA

      }

      dur_MN <- numeric(max(number_MN))
      event_MN <- numeric(max(number_MN))

      for (n in unique(number_MN)) {
        if(n > 1) {
          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - max(df$t[number_MN == (n-1)], na.rm = T)
        } else {
          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - min(df$t[number_MN == n], na.rm = T)
        }

        event_MN[n] <- max(df$label_MN[number_MN == n], na.rm = T)
      }

      dur_RA <- numeric(max(number_RA))
      event_RA <- numeric(max(number_RA))

      for (n in unique(number_RA)) {
        if(n > 1) {
          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - max(df$t[number_RA == (n-1)], na.rm = T)
        } else {
          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - min(df$t[number_RA == n], na.rm = T)
        }

        event_RA[n] <- max(df$label_RA[number_RA == n], na.rm = T)
      }

      return(list(dur_MN[event_MN == e], dur_RA[event_RA == e]))
    })

    df <- reduce(out, cbind)

    dur_MN <- unlist(df[1,])
    dur_RA <- unlist(df[2,])

    return(data.frame(MN = c(mean(dur_MN), sd(dur_MN), length(dur_MN)),
                      RA = c(mean(dur_RA), sd(dur_RA), length(dur_RA))))
  })
})

# Calculate RMSD between event distribution descriptives

ref <- list(dot = list(fix = list(CDT = c(60, 127, 165),
                                  EM = c(NA, NA, NA),
                                  IDT = c(323, 146, 8),
                                  IKF = c(217, 184, 72),
                                  IMST = c(268, 140, 12),
                                  IHMM = c(214, 286, 67),
                                  IVT = c(203, 282, 71),
                                  NH = c(380, 333, 30),
                                  BIT = c(189, 113, 67),
                                  LNS = c(NA, NA, NA)),
                       sac = list(CDT = c(NA, NA, NA),
                                  EM = c(17, 14, 93),
                                  IDT = c(32, 14, 10),
                                  IKF = c(60, 26, 29),
                                  IMST = c(13, 5, 18),
                                  IHMM = c(41, 17, 27),
                                  IVT = c(36, 14, 28),
                                  NH = c(43, 16, 42),
                                  BIT = c(NA, NA, NA),
                                  LNS = c(26, 11, 53)),
                       pso = list(NH = c(24, 12, 17),
                                  LNS = c(20, 9, 31))),
            img = list(fix = list(CDT = c(397, 559, 251),
                                  EM = c(NA, NA, NA),
                                  IDT = c(399, 328, 242),
                                  IKF = c(174, 239, 513),
                                  IMST = c(304, 293, 333),
                                  IHMM = c(133, 216, 701),
                                  IVT = c(114, 204, 827),
                                  NH = c(258, 299, 292),
                                  BIT = c(209, 136, 423),
                                  LNS = c(NA, NA, NA)),
                       sac = list(CDT = c(NA, NA, NA),
                                  EM = c(25, 22, 787),
                                  IDT = c(25, 15, 258),
                                  IKF = c(62, 37, 353),
                                  IMST = c(17, 10, 335),
                                  IHMM = c(48, 26, 368),
                                  IVT = c(41, 22, 373),
                                  NH = c(50, 20, 344),
                                  BIT = c(NA, NA, NA),
                                  LNS = c(29, 12, 390)),
                       pso = list(NH = c(28, 13, 237),
                                  LNS = c(25, 9, 319))),
            vid = list(fix = list(CDT = c(213, 297, 211),
                                  EM = c(NA, NA, NA),
                                  IDT = c(554, 454, 48),
                                  IKF = c(258, 296, 169),
                                  IMST = c(526, 825, 71),
                                  IHMM = c(234, 319, 194),
                                  IVT = c(202, 306, 227),
                                  NH = c(429, 336, 83),
                                  BIT = c(248, 215, 170),
                                  LNS = c(NA, NA, NA)),
                       sac = list(CDT = c(NA, NA, NA),
                                  EM = c(20, 16, 252),
                                  IDT = c(24, 53, 41),
                                  IKF = c(55, 20, 107),
                                  IMST = c(18, 10, 76),
                                  IHMM = c(42, 18, 109),
                                  IVT = c(36, 16, 112),
                                  NH = c(44, 18, 104),
                                  BIT = c(NA, NA, NA),
                                  LNS = c(28, 12, 122)),
                       pso = list(NH = c(28, 13, 78),
                                  LNS = c(24, 10, 87))))

A2017.rmsd <- lapply(2:4, function(k) {
  lapply(1:length(A2017.fit), function(x) { 
    out <- lapply(1:k, function(y) {
      out <- lapply(A2017.fit[[x]], function(z) {

        if(class(z[[k]]) == "gazeHMM") z[[k]]$events[[y]]

      })

      df <- try(reduce(out, rbind))

      alg <- try(c(mean(df$dur, na.rm = T), sd(df$dur, na.rm = T), nrow(df)))

      if(y == 4) {
        M <- try(cbind(A2017.events[[x]][[y]], alg))
      } else {
        M <- try(cbind(A2017.events[[x]][[y]], alg, as.data.frame(ref[[x]][[y]])))
        M[1:2,4:ncol(M)] <- M[1:2,4:ncol(M)]/1e3
      }

      M.norm <- try(apply(M, 1, function(x) {(x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))}))

      if(y == 4) {
        rmsd <- try(sum(sqrt((M.norm[3,] - colSums(M.norm[1:2,]/2))^2)))
      } else {
        rmsd <- try(apply(M.norm[3:nrow(M.norm),], 1, function(r) {sum(sqrt((r - colSums(M.norm[1:2,]/2))^2))}))
      }

      coder <- try(sum(sqrt((M.norm[1,] - M.norm[2,])^2)))

      rmsd <- c(coder, coder, rmsd)

      return(rbind(M, rmsd))
    })
  })
})

# Create tables for RMSDs

rmsd.table <- lapply(A2017.rmsd, function(x) {
  lapply(x, function(y) {
    lapply(y, function(z) {

      tabmat <- t(apply(z, 2, function(d) {as.character(round(d, 3))}))

      tabmat[3,] <- paste("**", tabmat[3,], "**", sep = "")

      colnames(tabmat) <- c("Mean dur", "SD dur", "# events", "RMSD")
      rownames(tabmat)[1:3] <- c("coderMN", "coderRA", "gazeHMM")

      return(kable(tabmat, align = "r") %>% kable_styling(font_size = 14))
    })
  })
})

# Compute Cohen's kappa and confusion matrices

A2017.acc <- lapply(2:4, function(k) {
  lapply(1:length(A2017.fit), function(x) { 
    out <- lapply(1:length(A2017.fit[[x]]), function(y) {

      if(class(A2017.fit[[x]][[y]][[as.character(k)]]) == "gazeHMM") {
        cod.MN <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_MN))
        cod.RA <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_RA))

        return(rbind(cod.MN, cod.RA))
      }
    })

    df <- reduce(out, rbind)

    df[df[,1] == 0,1] <- 5

    events <- lapply(1:k, function(e) {

      alg <- df[,1] == e
      cod <- df[,2] == e

      mat <- matrix(c(sum(alg & cod, na.rm = T), sum(alg & !cod, na.rm = T), 
                      sum(!alg & cod, na.rm = T), sum(!alg & !cod, na.rm = T)), 2, 2)

      over <- mat[2,1]/sum(mat[,1])
      under <- mat[1,2]/sum(mat[,2])

      return(list(kappa = cohen.kappa(mat)$kappa, 
                  over = over,
                  under = under))
    })

    ratio <- mean(df[,1] != df[,2], na.rm = T)

    conmat <- table(df[,1], df[,2])

    return(list(events = events, ratio =  ratio, conmat = conmat))
  })
})

# Create tables for Cohen's kappa and confusion matrices

kappa.table <- lapply(A2017.acc, function(x) {
  out <- lapply(x, function(y){
    out <- lapply(y$events, function(z){
      return(round(z$kappa, 3))
    })

    return(reduce(out, cbind))
  })

  tabmat <- reduce(out, rbind)

  tabmat <- apply(tabmat, 2, as.character)

  colnames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Smooth pursuits")[1:ncol(tabmat)]
  rownames(tabmat) <- c("Dots", "Image", "Video")

  return(kable(tabmat, align = "r") %>% kable_styling(font_size = 14))
})

conf.table <- lapply(A2017.acc, function(x) {
  out <- lapply(x, function(y){

    tabmat <- apply(y$conmat, 2, as.character)

    diag(tabmat)[1:(nrow(tabmat)-1)] <- paste("**", diag(tabmat)[1:(nrow(tabmat)-1)], "**", sep = "")

    tabmat[nrow(tabmat), 5] <- paste("**", tabmat[nrow(tabmat), 5], "**", sep = "")

    colnames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Smooth pursuits", "Blinks", "Other")
    rownames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Smooth pursuits", "Blinks", "Other")[1:nrow(tabmat)]
    rownames(tabmat)[nrow(tabmat)] <- "Blinks"

    return(kable(as.data.frame(tabmat), align = "r") %>% kable_styling(font_size = 14))
  })
})

```

## Model Comparison (Schwarz weight)
### Image

```{r}
print(bic.plot[[2]])
```

## Model Comparison (Schwarz weight)
### Moving dot

```{r}
print(bic.plot[[1]])
```

## Model Comparison (Schwarz weight)
### Video

```{r}
print(bic.plot[[3]])
```

## Event Durations
### 2-state Model
#### Image - Fixations

```{r, results="asis"}
print(rmsd.table[[1]][[2]][[1]])
```

## Event Durations
### 2-state Model
#### Image - Saccades

```{r, results="asis"}
print(rmsd.table[[1]][[2]][[2]])
```

## Event Durations
### 2-state Model
#### Dots - Fixations

```{r, results="asis"}
print(rmsd.table[[1]][[1]][[1]])
```

## Event Durations
### 2-state Model
#### Dots - Saccades

```{r, results="asis"}
print(rmsd.table[[1]][[1]][[2]])
```

## Event Durations
### 2-state Model
#### Video - Fixations

```{r, results="asis"}
print(rmsd.table[[1]][[3]][[1]])
```

## Event Durations
### 2-state Model
#### Video - Saccades

```{r, results="asis"}
print(rmsd.table[[1]][[3]][[2]])
```

## Event Durations
### 3-state Model
#### Image - Fixations

```{r, results="asis"}
print(rmsd.table[[2]][[2]][[1]])
```

## Event Durations
### 3-state Model
#### Image - Saccades

```{r, results="asis"}
print(rmsd.table[[2]][[2]][[2]])
```

## Event Durations
### 3-state Model
#### Image - PSOs

```{r, results="asis"}
print(rmsd.table[[2]][[2]][[3]])
```

## Event Durations
### 3-state Model
#### Dots - Fixations

```{r, results="asis"}
print(rmsd.table[[2]][[1]][[1]])
```

## Event Durations
### 3-state Model
#### Dots - Saccades

```{r, results="asis"}
print(rmsd.table[[2]][[1]][[2]])
```

## Event Durations
### 3-state Model
#### Dots - PSOs

```{r, results="asis"}
print(rmsd.table[[2]][[1]][[3]])
```

## Event Durations
### 3-state Model
#### Video - Fixations

```{r, results="asis"}
print(rmsd.table[[2]][[3]][[1]])
```

## Event Durations
### 3-state Model
#### Video - Saccades

```{r, results="asis"}
print(rmsd.table[[2]][[3]][[2]])
```

## Event Durations
### 3-state Model
#### Video - PSOs

```{r, results="asis"}
print(rmsd.table[[2]][[3]][[3]])
```

## Event Durations
### 4-state Model
#### Image - Fixations

```{r, results="asis"}
print(rmsd.table[[3]][[2]][[1]])
```

## Event Durations
### 4-state Model
#### Image - Saccades

```{r, results="asis"}
print(rmsd.table[[3]][[2]][[2]])
```

## Event Durations
### 4-state Model
#### Image - PSOs

```{r, results="asis"}
print(rmsd.table[[3]][[2]][[3]])
```

## Event Durations
### 4-state Model
#### Image - Smooth pursuits

```{r, results="asis"}
print(rmsd.table[[3]][[2]][[4]])
```

## Event Durations
### 4-state Model
#### Dots - Fixations

```{r, results="asis"}
print(rmsd.table[[3]][[1]][[1]])
```

## Event Durations
### 4-state Model
#### Dots - Saccades

```{r, results="asis"}
print(rmsd.table[[3]][[1]][[2]])
```

## Event Durations
### 4-state Model
#### Dots - PSOs

```{r, results="asis"}
print(rmsd.table[[3]][[1]][[3]])
```

## Event Durations
### 4-state Model
#### Dots - Smooth pursuits

```{r, results="asis"}
print(rmsd.table[[3]][[1]][[4]])
```

## Event Durations
### 4-state Model
#### Video - Fixations

```{r, results="asis"}
print(rmsd.table[[3]][[3]][[1]])
```

## Event Durations
### 4-state Model
#### Video - Saccades

```{r, results="asis"}
print(rmsd.table[[3]][[3]][[2]])
```

## Event Durations
### 4-state Model
#### Video - PSOs

```{r, results="asis"}
print(rmsd.table[[3]][[3]][[3]])
```

## Event Durations
### 4-state Model
#### Video - Smooth pursuits

```{r, results="asis"}
print(rmsd.table[[3]][[3]][[4]])
```

## Agreement with Human Coders (Cohen's kappa)
### 2-state Model

```{r, results='asis'}
print(kappa.table[[1]])
```

## Agreement with Human Coders (Cohen's kappa)
### 3-state Model

```{r, results='asis'}
print(kappa.table[[2]])
```

## Agreement with Human Coders (Cohen's kappa)
### 4-state Model

```{r, results='asis'}
print(kappa.table[[3]])
```

## Agreement with Human Coders (Confusion Matrix)
### 2-state Model
#### Dots

```{r, results='asis'}
print(conf.table[[1]][[1]])
```

## Agreement with Human Coders (Confusion Matrix)
### 2-state Model
#### Image

```{r, results='asis'}
print(conf.table[[1]][[2]])
```

## Agreement with Human Coders (Confusion Matrix)
### 2-state Model
#### Video

```{r, results='asis'}
print(conf.table[[1]][[3]])
```

## Agreement with Human Coders (Confusion Matrix)
### 3-state Model
#### Dots

```{r, results='asis'}
print(conf.table[[2]][[1]])
```

## Agreement with Human Coders (Confusion Matrix)
### 3-state Model
#### Image

```{r, results='asis'}
print(conf.table[[2]][[2]])
```

## Agreement with Human Coders (Confusion Matrix)
### 3-state Model
#### Video

```{r, results='asis'}
print(conf.table[[2]][[3]])
```

## Agreement with Human Coders (Confusion Matrix)
### 4-state Model
#### Dots

```{r, results='asis'}
print(conf.table[[3]][[1]])
```

## Agreement with Human Coders (Confusion Matrix)
### 4-state Model
#### Image

```{r, results='asis'}
print(conf.table[[3]][[2]])
```

## Agreement with Human Coders (Confusion Matrix)
### 4-state Model
#### Video

```{r, results='asis'}
print(conf.table[[3]][[3]])

### 2020-06-04 manuscript/internship_report.Rmd
upload first draft simulation results

---
title             : "Keeping an Eye on Hidden Markov Models in Gaze Data Classification"
shorttitle        : "HMMs in Gaze Classification"

author: 
  - name          : "Malte Valentin Lueken"
    affiliation   : "1"
    address       : "Speelmanstraat 7-3 1063ZC Amsterdam"
    email         : "malte.luken@student.uva.nl"

affiliation:
  - id            : "1"
    institution   : "University of Amsterdam"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.

  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.

  One sentence clearly stating the **general problem** being addressed by  this particular study.

  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).

  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.

  One or two sentences to put the results into a more **general context**.

  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "keywords"
wordcount         : "X"

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

dir <- "~/Uni/Psychologie Master/Internship/GazeHMM"

library(tidyverse)
require(signal)
library(flexcircmix)
source(paste(dir, "algorithm/preprocessing_helper_functions.R", sep = "/"))

```

```{r load data, include=FALSE}

load(paste(dir, "validation/Andersson2017_raw.Rdata", sep = "/"))

res <- c(1024, 768)
dim <- c(380, 300)
dist <- 670
fr <- 500

```

# Algorithm Development

As part of answering my research questions, I developed an algorithm named *gazeHMM* to classify gaze data into discrete eye movement events. The following section describes the development of gazeHMM and its final version that has been used to obtain results. Technical details can be found in the Technical Appendix.

## Eye movement metrics

Many different metrics can be used to describe gaze data and eye movement events (Zemblys et al.). For classifcation, the goal is to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user (Zemblys et al.). Since not relying on parameter settings by the user was one criterion for choosing metrics, I excluded them from the decision process. Finally, I chose to use three metrics that have been used in previous algorithms: velocity, acceleration, and sample-to-sample angle (e.g., Larsson et al., Pekkanen and Lappi). The reasons to chose these three are mainly theoretical. Figure X displays the theoretically expected distributions of samples belonging to different eye movement events. Fixations typically inherit samples with low velocity and acceleration (CITE). Due to tremor, the angle between samples should not follow any direction but a random walk (CITE). In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction (CITE). PSO samples tend to have moderate velocity and high acceleration since they occur between high velocity saccades and low velocity fixations. They are specifically distinguished by their change in direction around 180° (CITE). Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction (CITE). In sum, velocity, acceleration, and sample-to-sample angle should theoretically separate eye movement events well.

```{r, theoretical event sample distributions, results='asis'}

```

## Fitering and smoothing

Like metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics. Their purpose is to remove noise or artifacts from the gaze data that could distort the classification (CITE). Filtering (smoothing) methods that are applied before computing the eye movement metrics target the recorded gaze position. Unfortunately, this might also erase differences in the metrics between samples. For instance, tremor movements during fixations would be filtered out and in consequence, the angle of the samples would stay the same. This would make it harder for the algorithm to separate fixations from smooth pursuits. The same phenomenon would apply to PSOs and saccades (see Figure X).

```{r, filtering example, results='asis'}

filter.ex <- A2017$video[[3]] %>%
  mutate(x = px2va(x, dim[1], res[1], dist),
         y = px2va(y, dim[2], res[2], dist),
         angle = calc_theta(x, y),
         x.filt = as.numeric(signal::filter(signal::butter(3, 0.3), x)),
         y.filt = as.numeric(signal::filter(signal::butter(3, 0.3), y)),
         angle.filt = calc_theta(x.filt, y.filt),
         t = (t-t[1])/1e6) %>%
  dplyr::filter(t > 0.5, t < 0.75)

ggplot(filter.ex) + 
  geom_path(aes(x = t, y = force_neg_pi_pi(angle)), color = "blue") +
  geom_path(aes(x = t, y = force_neg_pi_pi(angle.filt)), color = "red") +
  scale_y_circular(units = "radians", limits = c(-pi, pi))

```

Instead of filtering before computing the metrics, I decided to combine both in one step. Previous algorithms have used two methods that both filter (smooth) and compute derivatives of the gaze position (i.e., velocity and acceleration). (CITE) compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively, CITE) of the gaze position to be used by their algorithm. Similarly, (CITE) use a Savitzky-Golay (SG) filter (CITE) to estimate velocity and acceleration from a gaze signal. Figure X displays velocity and acceleration signals obtained by both methods. They follow a similar trend, but the SG velocity signal is less noisy and consequently, the SG acceleration signal is lower than the Laplace acceleration signal. I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events. To preserve motor noise in the change in angle signal, gazeHMM uses a Sobel filter to compute the first discrete derivative of sample angle.

```{r, filter comparison, results='asis'}

filter.comp <- A2017$video[[3]] %>%
  mutate(x = px2va(x, dim[1], res[1], dist),
         y = px2va(y, dim[2], res[2], dist),
         vel.x.sg = signal::sgolayfilt(x, m = 1),
         vel.y.sg = signal::sgolayfilt(y, m = 1),
         vel.sg = sqrt(vel.x.sg^2 + vel.y.sg^2)*fr,
         acc.x.sg = signal::sgolayfilt(x, m = 2),
         acc.y.sg = signal::sgolayfilt(y, m = 2),
         acc.sg = sqrt(acc.x.sg^2 + acc.y.sg^2)*fr^2,
         vel.x.sb = lead(x) - lag(x),
         vel.y.sb = lead(y) - lag(y),
         vel.sb = sqrt(vel.x.sb^2 + vel.y.sb^2)*fr,
         acc.x.lp = lead(x) - 2*x + lag(x),
         acc.y.lp = lead(y) - 2*y + lag(y),
         acc.lp = sqrt(acc.x.lp^2 + acc.y.lp^2)*fr^2,
         t = (t-t[1])/1e6) %>%
  pivot_longer(c("vel.sg", "vel.sb", "acc.sg", "acc.lp"), names_to = c("metric", "filter"), names_sep = "[:punct:]") %>%
  dplyr::filter(t > 0.5, t < 1.5)

ggplot(filter.comp, aes(x = t, y = value, color = filter)) + 
  geom_path() + facet_grid(rows = vars(metric), scales = "free_y")

```

## The generative model

The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM). It can have between two and four states that correspond to different eye movement events (see Table X). Thus, users can choose wether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth puruits.

In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model (CITE). In gazeHMM, the response model consists of three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal obtained by the Sobel filter. The response variables are treated as independent conditional on the states. I acknowledge that conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated). This step was taken merely to keep the HMM simple and identifyable.

Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them). However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation) and due to noise above zero. Second, the distributions of both signals appear to be positively skewed conditionally on the states and, third, have variances increasing with their mean. Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties. In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose. However, it has to be noted the gamma was choosen because of convenience and the best fitting distribution might be different between subjects and tasks.

To model the sample-to-sample angle, CITE described the Fisher-transformed cosine of sample angle with a Gaussian distribution. This approach can be problematic because the cosine of sample angle only ranges between -1 and 1 (corresponding to angles of $\pi$ and 0). Thus, it only operates on half of the unit circle and $\cos(\frac{1}{2}\pi) = \cos(\frac{3}{2}\pi)$. This property can be desirable if one is only interested in expression angles between 0 and $\pi$ (which is in event classification). However, the sample-to-sample angle distributions for the different events will also be skewed having their modes at 0 (for saccades and smooth pursuits) or $\pi$ (for PSOs). This consequence makes the Fisher-transformation and Gaussian not the best choices. Instead, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution to model the derivative of sample angle. Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$) which should lead to rather symmetric distributions. Moreover, the uniform distribution can distinguish fixations from the other events.

```{r angle distributions, results="asis"}

angle.dist <- A2017$video[[3]] %>%
  mutate(x = px2va(x, dim[1], res[1], dist),
         y = px2va(y, dim[2], res[2], dist),
         angle.full = force_neg_pi_pi(calc_theta(x, y)),
         #angle.full = force_neg_pi_pi(atan2(y - lag(y), x - lag(x))),
         cos.angle = cos(angle.full),
         z.trans = atanh(cos.angle),
         t = (t-t[1])/1e6) %>%
  pivot_longer(c("angle.full", "cos.angle", "z.trans"), names_to = "method") %>%
  dplyr::filter(t > 0.5, t < 0.75)

ggplot(angle.dist, aes(x = t, y = value)) + geom_path() + facet_grid(rows = vars(method), scales = "free")

```

To describe the initial state model and the transition model I used multinomial distributions (with a category for each state). For all three submodels of the HMM, only parameter intercepts were estimated.

## Optimization and classification

The paramters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm (EM CITE). The EM is generally suitable to estimate likelihoods with missing variables. For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data (velocity, acceleration, and sample-to-sample angle) and the expected hidden states (eye movement events; Visser & Speekenbrink). The sequence of hidden states is estimated through the Viterbi algorithm (CITE) by maximizing the posterior state probability (Visser & Speekenbrink). Parameters of the response distributions (except for the uniform distribution) were optimized on the log-scale (except for the mean parameter of the von-Mises distribution) using a spectral projected gradient method (CITE) and Barzilai-Borwein steplengths (CITE).

## Postprocessing

Eye movement events usually obey certain theoretical (e.g., PSOs only follow saccades) and practical constraints (e.g., eye-tracking resolutions) (CITE). However, constraining the parameter estimation in gazeHMM led mostly to unreasonable results or did not improve the classification. For example, constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the classification output at all. Moreover, I could not explicitly control the duration of events in the HMM which ocassionally led to unreasonable short events. In consequence, I implemented a postprocessing routine in gazeHMM to heuristically compensate for such violations. This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events as the previous event.

## Implementation

The algorithm is implemented in R (version 3.6.3; CITE) and uses the packages signal (CITE) to compute velocity and acceleration signals, dplyr (CITE) to calculate sample-to-sample angles, depmixS4 (CITE) for the HMM, and BB (CITE) for Barzilai-Borwein spectral projected gradient optimization. The algorithm will be available on Github (LINK).

# Simulation Study Design

# Results
## Simulation study

With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences. The most critical decrease in recovery occured when noise was added to the data generated by models including smooth pursuits. In the noise condition, scale parameters of gamma distributions were often badly recovered. Higher sample sizes slightly improved parameter recovery, but neither variation in starting values nor missing data affected it. In contrast, the accuracy of the model was linearly descreasing with more data missing but not influenced by manipulating parameters, noise, or starting value variation. Adding more states to the HMM generally decreased the parameter recovery and the accuracy. The full results are described in the following sections.

### Parameter variation

In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation between the true and estimated parameters and the accuracy between the true and estimated state sequence. Figure X displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM with two states. The RMdSPD was below 0.1 for all estimated and manipulated parameters (except the irrelevant initial state probabilities). Thus, no matter which relevant parameter changed in the HMM, the median deviation did not exceed 10% of the true parameter value. The regressions between manipulated true and estimated parameters for two states are shown in Figures X and X. With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely: Changing the true parameters led to linear changes in the estimated parameters. However, the deviation between true and estimated response parameters seemed to increase slighlty with parameter magnitude. Considering accuracy, Figure X displays Cohen's kappa between true and estimated hidden state sequences for two states. With two exceptions, kappa values were almost one, indicating nearly perfect agreement. In sum, the HMM with two states recovered parameters and hidden states very well with outliers only occuring rarely.

For three states, the RMdSPD is shown in Figure X. When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently below 0.5. Varying $\kappa$ in states two and three led to RMdSPDs below 0.5 in the respective states. Otherwise, RMdSPDs were consistently lower than 0.1, indicating a median deviation between true and estimated parameters below 10% of the true parameter. Inspecting the regressions between manipulated true and estimated parameters (see Figures X and X) revealed a strong and unbiased linear relationship (intercepts close to zero and slopes close to one). Again, the deviation seemed to increase with true parameter magnitude (reverse for kappas). In contrast to the two-state HMM, larger deviatons and more outliers were observed. Cohen's kappa values for the three-state HMM are presented in Figure X. For most estimated models, the kappas between true and estimated state sequences were above 0.95, indicating almost perfect agreement. However, for some models I observed kappas clustered around zero or -0.33, suggesting that state labels switched (CITE). To summarize, the three-state HMM displayed a slightly worse parameter recover with more outliers than the two-state model but shows an almost equally good accuracy.

The RMdSPDs for the four-state HMM is illustrated in Figure X. For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were below 0.5. Estimated kappa parameters in smooth pursuits were also often below 0.5 when paramteres in states two, three, and four were varied. Otherwise, RMdSPDs were below 0.1, indicating a median deviation below 10% of the true parameter. Looking at the regressions between true and estimated paramters, Figure X shows strong and unbiased relationships but larger deviations and more outliers than in the previous models, especially for states one and four. Accuracy indicated by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences. Here, some kappa values clustered around 0.25 and zero, which again can be interpreted as the results of label switching. Summarizing, the parameter recovery for the four-state HMM was slightly worse with even more outliers compared to the two- and three-state models. Especially the recovery of transition parameters in states corresponding to fixations and smooth pursuits decreased.

Overall, simulations in part one demonstrated that the HMM recovered parameters at least moderately well when parameters were manipulated (all RMdSPDs below 0.5). The HMM estimated state sequences very accurately. Adding states to the model decreased the accuracy and recovery slightly, leading to more outliers, especially when smooth pursuits were added.

### Sample size and noise variation

In the second part, I varied the sample size of the HMM and added noise to the generated data. For the two-state HMM, the RMdSPDs were above 0.5 for $\beta_{vel}$ and $\beta_{acc}$ in both states (see Figure X). The other estimated parameters showed RMdSPDs close to or below 0.1. Increasing the sample size seemed to improve RMdSPDs for most parameters slightly. For $\beta_{vel}$ and $\beta_{acc}$ in both states, models with 2500 samples had the lowest RMdSPDs. Accuracy measured by Cohen's kappa was almost perfect with kappa values very close to one (see Figure X, left plot). To conclude, adding noise did not affect the parameter recover or accuracy of the two-state HMM, whereas increasing the sample size improved the recovery slightly.

The RMdSPDs for the three-state HMM and $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states. Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5. For most parameters in all three states, higher sample sizes had lower RMdSPDs. The accuracy of the estimated models was almost perfect with most kappa values above 0.95 (see Figure X, middle plot). Several outliers clustered around kappas of zero and -0.33, signaling label switching. For the three-state HMM, adding noise led to a slightly worse recovery and accuracy overall. However, scale parameters of gamma distributions in all three states were only badly recovered.

RMdSPDs regarding the four-state HMM are displayed in Figure X. For states one and four, values for most parameters (including all transition probabilites) were above 0.5. Similarly, RMdSPDs for $\beta_{vel}$ and $\beta_{acc}$ in states two and three were above 0.5. For states two and three, higher samples sizes showed slightly lower RMdSPDs. As in the previous part, most Cohen's kappa values ranged between 0.6 and 0.9, meaning substantial to almost perfect agreement between true and estimated states (Figure X, right plot). Multiple kappa values clustered around 0.25 or zero, which can be explained by label switching. In summary, adding noise to data generated by the four-state HMM heavily decreased the recovery for most parameters in states corresponding to fixations and smooth pursuits. For saccade and PSO states, only scale parameters of gamma distributions were badly recovered, but increasing the sample size slightly improved the recovery.

In general, the HMM recovered parameters well despite noise being added to the data. However, when smooth pursuits were added to the model, the parameter recovery for fixation and smooth pursuit states substantially decreased. When PSOs or smooth pursuits were included, scale parameters of gamma distributions were badly recovered. Increasing the samples in the HMM slightly improved the recovery of most parameters. The accuracy of the model was slightly lowered when including more states, but it was neither affected by the range of noise $\tau_{noise}$ nor the sample size.

### Variation of starting values

The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM. For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure X). Cohen's kappa values were slightly lower than 1, indicating almost perfect accuracy (see Figure X). For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5. As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33. Regarding the four state-HMM, RMdSPS of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1. The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero. In conclusion, the parameter recovery of the model worked very well. State sequences were accurately estimated and more states being included in the model slightly decreased the accuracy. Increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor accuracy.

Since the manipulation in this part was not effective, I decided to explore post-hoc whether clusters of low Cohen's kappa values were due to label switching. According to Visser and Speekenbrink (CITE), label switching occurs when exchanging the order of the states leads to equally likely models. Thus, the model is accurate but has oppositely estimated states compared to the true states. In this case, an HMM with three states that is perfectly accurate but has one label swtiched would have a Cohen's kappa of 0. To test if label switching occured, I manually switched one or two labels post-hoc (see Figure X). It can be seen that this approach resolved low accuracy clusters for three and four states.

### Missing data

In the last part, intervals of varying length in the data were set to be missing. Regarding parameter recovery, RMdSPDs were almost exactly mirroring those in the previous part for two, three, and four states. Looking at accuracies, Figure X illustrates an interaction between the length of and amount of missing data intervals. Cohen's kappa values linearly decreased with the length of missing data intervals and the decrease became linearly steeper when more intervals were included. For two and three state models, kappas started at almost one and decreased to 0.6 for five missing intervals, indicating substantial to almost perfect accuracy. For the four-state model, kappas started around 0.85 and decreased to 0.5 for five missing intervals, suggesting moderate to substantial accuracy. As in previous parts, a few kappa values clustered around zero and -0.33 for the three-state model and around 0.6, 0.25, and zero for the four-state model. In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model up to a moderate extent.

# Technical Appendix

### 2020-06-05 algorithm/postprocessing.R
change duration computation (exclude blinks)

@@ -21,6 +21,8 @@ postprocess <- function(data, state, min.sac) { # data frame with dependent vari
   
   valid <- !is.na(data$vel) & !is.na(data$acc) & !is.na(data$angle)
   
+  state[!valid] <- 0
+  
   
   # Relabel samples
   
@@ -109,7 +111,7 @@ postprocess <- function(data, state, min.sac) { # data frame with dependent vari
   }
   
   data$state <- state
-  data$label[is.na(data$label)] <- label[is.na(data$label)] # only labels for valid samples
+  data$label <- label
   data$event <- number
   
   

### 2020-06-08 manuscript/internship_report.Rmd
update report

@@
 
-floatsintext      : no
+floatsintext      : yes
 figurelist        : no
@@
 
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = "asis")
 
@@
 
+library(papaja)
 library(tidyverse)
-require(signal)
+library(signal)
 library(flexcircmix)
+library(CircStats)
+library(depmixS4)
 source(paste(dir, "algorithm/preprocessing_helper_functions.R", sep = "/"))
+
+```
+
+```{r set graphics theme, include=FALSE}
+
+theme_set(theme_apa())
 
@@
 Many different metrics can be used to describe gaze data and eye movement events (Zemblys et al.).
-For classifcation, the goal is to find those metrics that separate the gaze data samples belonging to different events the best.
-However, many metrics rely on thresholds or window ranges that have to be set by the user (Zemblys et al.).
-Since not relying on parameter settings by the user was one criterion for choosing metrics, I excluded them from the decision process.
-Finally, I chose to use three metrics that have been used in previous algorithms: velocity, acceleration, and sample-to-sample angle (e.g., Larsson et al., Pekkanen and Lappi).
-The reasons to chose these three are mainly theoretical.
-Figure X displays the theoretically expected distributions of samples belonging to different eye movement events.
+Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best.
+However, many metrics rely on thresholds or window ranges that have to be set by the user (e.g., the distance between the mean position in a 100 ms window before and after each sample, see Olsson, 2007, and Zemblys et al.).
+This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics.
+Since I proposed to choose metrics that do not rely on such user settings, I used velocity, acceleration, and sample-to-sample angle as metrics in gazeHMM.
+Theoretically, these three metrics should separate eye movement events clearly.
 Fixations typically inherit samples with low velocity and acceleration (CITE).
@@
 PSO samples tend to have moderate velocity and high acceleration since they occur between high velocity saccades and low velocity fixations.
-They are specifically distinguished by their change in direction around 180° (CITE).
+They are specifically distinguished by their change in direction clustered around 180° (CITE).
 Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction (CITE).
-In sum, velocity, acceleration, and sample-to-sample angle should theoretically separate eye movement events well.
-
-```{r, theoretical event sample distributions, results='asis'}
-
-```
 
@@
 
-Like metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics.
+As with metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics.
 Their purpose is to remove noise or artifacts from the gaze data that could distort the classification (CITE).
@@
 Unfortunately, this might also erase differences in the metrics between samples.
-For instance, tremor movements during fixations would be filtered out and in consequence, the angle of the samples would stay the same.
+For instance, tremor movements during fixations would be filtered out and, in consequence, the samples would follow the same direction.
 This would make it harder for the algorithm to separate fixations from smooth pursuits.
-The same phenomenon would apply to PSOs and saccades (see Figure X).
+The same problem applies to PSOs and saccades where filtering could erase the oscillations (see Figure \@ref(fig:filter-angle)).
 
-```{r, filtering example, results='asis'}
+(ref:filter-angle) Example data from Andersson et al.
+(2017) displayed as sample-to-sample angle (in rad) over time (in s).
+Line color indicates whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue).
+For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3.
+
+```{r, filter-angle, results='asis', fig.cap="(ref:filter-angle)"}
 
@@
   geom_path(aes(x = t, y = force_neg_pi_pi(angle.filt)), color = "red") +
-  scale_y_circular(units = "radians", limits = c(-pi, pi))
+  scale_y_circular(name = "Sample-to-sample angle (in rad)", units = "radians", limits = c(-pi, pi)) +
+  scale_x_continuous(name = "Time (in s)")
 
@@
 Instead of filtering before computing the metrics, I decided to combine both in one step.
-Previous algorithms have used two methods that both filter (smooth) and compute derivatives of the gaze position (i.e., velocity and acceleration).
+Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration).
 (CITE) compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively, CITE) of the gaze position to be used by their algorithm.
 Similarly, (CITE) use a Savitzky-Golay (SG) filter (CITE) to estimate velocity and acceleration from a gaze signal.
-Figure X displays velocity and acceleration signals obtained by both methods.
-They follow a similar trend, but the SG velocity signal is less noisy and consequently, the SG acceleration signal is lower than the Laplace acceleration signal.
+Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods.
+They follow a similar trend, but the SG velocity signal is less noisy and, consequently, the SG acceleration signal is lower than the Laplace acceleration signal.
 I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
-To preserve motor noise in the change in angle signal, gazeHMM uses a Sobel filter to compute the first discrete derivative of sample angle.
+To preserve motor noise in the change in angle signal, gazeHMM computes the first discrete derivative of sample angle directly (see Technical Appendix).
 
-```{r, filter comparison, results='asis'}
+(ref:filter-comparison) Example data from Andersson et al.
+(2017) displayed as velocity (deg/seconds) and acceleration (in deg/s^2) over time (in s).
+Line colors indicate which filter has been used to derive the signal.
+Savitzky-Golay filters were applied with order three and length five (corresponding to 10 ms).
+
+```{r, filter-comparison, results='asis', fig.cap="(ref:filter-comparison)"}
 
@@
          t = (t-t[1])/1e6) %>%
-  pivot_longer(c("vel.sg", "vel.sb", "acc.sg", "acc.lp"), names_to = c("metric", "filter"), names_sep = "[:punct:]") %>%
+  pivot_longer(c("vel.sg", "vel.sb", "acc.sg", "acc.lp"), names_to = c("metric", "filter"), names_sep = "[:punct:]", ) %>%
+  mutate_at(c("metric", "filter"), as.factor) %>%
   dplyr::filter(t > 0.5, t < 1.5)
 
+filter.comp$metric <- factor(filter.comp$metric, labels = c("Acceleration", "Velocity"))
+
 ggplot(filter.comp, aes(x = t, y = value, color = filter)) + 
-  geom_path() + facet_grid(rows = vars(metric), scales = "free_y")
+  geom_path() + facet_wrap(vars(metric), nrow = 2, scales = "free_y", strip.position = "left") + 
+  scale_x_continuous(name = "Time (in s)") +
+  scale_y_continuous(name = "") +
+  scale_color_discrete(name = "Filter", labels = c("Laplace", "Sobel", "Savitzky-Golay")) +
+  theme(strip.placement = "outside")
 
@@
 The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM).
-It can have between two and four states that correspond to different eye movement events (see Table X).
+It can have between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
 Thus, users can choose wether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth puruits.
-
 In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model (CITE).
-In gazeHMM, the response model consists of three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal obtained by the Sobel filter.
-The response variables are treated as independent conditional on the states.
+In gazeHMM, the response model consists of three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal.
+The response variables are treated as conditionally independent on the states.
 I acknowledge that conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
 This step was taken merely to keep the HMM simple and identifyable.
-
 Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them).
 However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation) and due to noise above zero.
-Second, the distributions of both signals appear to be positively skewed conditionally on the states and, third, have variances increasing with their mean.
+Second, the distributions of both signals appear to be positively skewed conditionally on the states and, third, to have variances increasing with their mean.
 Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties.
 In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose.
-However, it has to be noted the gamma was choosen because of convenience and the best fitting distribution might be different between subjects and tasks.
-
-To model the sample-to-sample angle, CITE described the Fisher-transformed cosine of sample angle with a Gaussian distribution.
-This approach can be problematic because the cosine of sample angle only ranges between -1 and 1 (corresponding to angles of $\pi$ and 0).
-Thus, it only operates on half of the unit circle and $\cos(\frac{1}{2}\pi) = \cos(\frac{3}{2}\pi)$.
-This property can be desirable if one is only interested in expression angles between 0 and $\pi$ (which is in event classification).
-However, the sample-to-sample angle distributions for the different events will also be skewed having their modes at 0 (for saccades and smooth pursuits) or $\pi$ (for PSOs).
-This consequence makes the Fisher-transformation and Gaussian not the best choices.
-Instead, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution to model the derivative of sample angle.
+It has to be noted that the gamma was choosen because of convenience and the best fitting distribution might be different between subjects and tasks.
+To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution to model the derivative of sample angle.
 Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$) which should lead to rather symmetric distributions.
 Moreover, the uniform distribution can distinguish fixations from the other events.
+The initial state model and the transition model used multinomial distributions (with a category for each state). For all three submodels of the HMM, only parameter intercepts were estimated.
 
-```{r angle distributions, results="asis"}
+## Missing data and blinks
 
-angle.dist <- A2017$video[[3]] %>%
-  mutate(x = px2va(x, dim[1], res[1], dist),
-         y = px2va(y, dim[2], res[2], dist),
-         angle.full = force_neg_pi_pi(calc_theta(x, y)),
-         #angle.full = force_neg_pi_pi(atan2(y - lag(y), x - lag(x))),
-         cos.angle = cos(angle.full),
-         z.trans = atanh(cos.angle),
-         t = (t-t[1])/1e6) %>%
-  pivot_longer(c("angle.full", "cos.angle", "z.trans"), names_to = "method") %>%
-  dplyr::filter(t > 0.5, t < 0.75)
-
-ggplot(angle.dist, aes(x = t, y = value)) + geom_path() + facet_grid(rows = vars(method), scales = "free")
-
-```
-
-To describe the initial state model and the transition model I used multinomial distributions (with a category for each state).
-For all three submodels of the HMM, only parameter intercepts were estimated.
+HMMs can estimate parameters and hidden states despite missing data.
+They either treat data as missing at random or as missing dependent on states (Visser & Speekenbrink).
+In gazeHMM, I assumed data to be missing at random.
+Most of the missing data in eye movement classification are due to blinks.
+The user can indicate in gazHMM which samples should be labeled as blinks (other missing samples are treated as noise).
+Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs.
+Since these samples could distort the classification of saccades in the HMM, I decided to remove them heuristically.
+Before classifying the samples, gazeHMM sets all samples within 50 ms before and after blink samples as missing.
+The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.
+When evaluating the likelihood of the HMM given parameters and the joint density of response distributions for missing data, the model integrates over all possible values.
+The resulting density for missing samples is therefore set to one (Visser & Speekenbrink).
 
@@
 
-The paramters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm (EM CITE).
+The paramters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm (CITE).
 The EM is generally suitable to estimate likelihoods with missing variables.
@@
 
-Eye movement events usually obey certain theoretical (e.g., PSOs only follow saccades) and practical constraints (e.g., eye-tracking resolutions) (CITE).
-However, constraining the parameter estimation in gazeHMM led mostly to unreasonable results or did not improve the classification.
-For example, constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the classification output at all.
-Moreover, I could not explicitly control the duration of events in the HMM which ocassionally led to unreasonable short events.
-In consequence, I implemented a postprocessing routine in gazeHMM to heuristically compensate for such violations.
-This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events as the previous event.
+After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence.
+I implemented this rountine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
+Moreover, gazeHMM did explicitly control the duration of events in the HMM which ocassionally led to unreasonably short events.
+Thus, the postprocessing routine heuristically compensates for such violations.
+This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events.
+Samples were relabeled as the state of the previous event.
+Finally, samples initially indicated as missing were labeled as noise (including blinks) and event metrics were computed (e.g., fixation durations).
 
@@
 
-# Simulation Study Design
+# Simulation Study
+
+To assess how well the HMM recovers parameters and state sequences, I conducted a simulation study.
+The design and analysis of the study were preregistered on OSF (LINK).
+The major part of this section is a direct copy of that preregistration.
+When appropriate, additional explanations were added and tenses adapted to make the design easier to read and understand.
+In the study, the HMM repeatedly generated data with a set of parameters (true parameter values).
+The same model was then be applied to estimate the parameters from the generated data (estimated parameter values).
+I compared the true with the estimated parameter values to assess whether a parameter was recovered by the model.
+Additionally, I compared the true states of the HMM with the estimated states to judge how accurately the model recovered the states that generated the data.
+
+## Design
+### Parameter Variation
+
+The simulation study was divided into four parts.
+In the first part, I varied the parameters of the HMM.
+For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were varied respectively.
+For each parameter, the HMM generated 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps.
+This resulted in $100 \times (10+15+20) = 4500$ recoveries.
+Only one parameter was varied at once, the other parameters were set to their default values.
+All parameters of the HMM were estimated freely (i.e., there were no fixed parameters in the model).
+I did not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
+For the transition probabilities, I only simultaneously varied the probabilities for staying in the same state (diagonals of the tranistion matrix) to reduce the complexity of the simulation.
+The left over probability mass was split evenly between the probabilites for switching to a different state (per row of the transition matrix).
+Moreover, I did not modify the mean parameters of the von-Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
+I defined approximate ranges for each response variable and chose true parameter intervals and default values so that they produced samples that roughly corresponded to these ranges.
+Table \@ref(tab:sim-ranges) shows the assumed ranges for each event and Table X shows the intervals and default values for each parameter in the simulation.
+Parameters were scaled down by factor 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
+I set the intervals for shape parameters of the gamma distributions for all events to [1,5] to examine how skewness influenced the recovery (shape values above five approach a symmetric distribution).
+The scale parameters were set so that the respective distribution approximately matched the assumed ranges.
+Since the concentration parameters of the von-Mises distribution are the inverse of standard deviations, they were varied on the inverse scale.
+
+### Sample Size and Noise Variation
+
+In the second part, I varied the sample size of the generated data and the amount of noise added to it.
+The model parameters were set to their default values.
+For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, I generated 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
+These samples sizes roughly corresponded to small, medium, and large eye-tracking data sets for a single participant and trial.
+To simulate noise, I replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
+This procedure ensured that velocity and acceleration values remained positive and were drawn from moderately skewed distributions with modes equal to the original values.
+To angle, I added white noise from a von-Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
+$\tau_{noise}$ and $\kappa_{noise}$ were varied simultaneously in equidistant steps in their intervals.
+
+### Variation of Starting Values
+
+In the third part, I increased the variation in the starting values used for parameter estimation.
+The model parameters were set to their default values.
+For the shape, scale, and concentration parameters, I simultaneously increased the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta_{start} = (\psi_{true}/2)\tau_{start}$ with $\tau_{start} \in \{1, 2, 3\}$, 100 data sets with $N = 2500$ samples were generated each ($100 \times 3 \times 3 = 900$ recoveries).
+
+### Missing data
+
+In the last part, I set intervals of the generated data to be missing.
+The model parameters were set to their default values.
+For $k \in \{2, 3, 4\}$ states and $m \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples were generated ($100 \times 3 \times 3 = 900$ recoveries).
+The length of the missing data interval $l \in [1,200]$ samples varied in equidistant steps between the data sets.
+
+## Starting values
+
+The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
+To generate random starting values for the estimation of shape, scale, and concentration parameters, I used gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated.
+This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
+Mean parameters of the von-Mises distribution always started at their true values.
+
+## Data analysis
+
+For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values[^1]: $$RMdSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
+I treated $RMdSPD < 0.1$ as good, $0.1 \le RMdSPD < 0.5$ as moderate, and $RMdSPD \ge 0.5$ as bad recovery of a parameter.
+By taking the median, I reduced the influence of potential outliers in the estimation and using proportions enabled me to compare RMdSPD values across parameters and data sets.
+Additionally, I applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in Part 1.
+Regression slopes closer to one indicated that the model better captured parameter change.
+Regression intercepts different from zero reflected a bias in parameter estimation.
+To assess state recovery, I computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set.
+Higher kappa values were interpreted as better model accuracy.
+I adopted the ranges proposed by @Landis1977 to interpret kappa values.
+Models that can not be fitted were excluded from the recovery.
+
+[^1]: This measure is only appropriate when $\psi_{true} \ne 0$. This was not the case for some mean parameters of the von-Mises distributions. In those cases, I used $\psi_{true} = 2\pi$ instead.
+
+(ref:sim-ranges) Approximate ranges of response variables used to generate parameter values
+(ref:sim-ranges-note) Units are °/s (velocity), °/s^2^ (acceleration), and radians (angle).
+~ indicates that the distribution has a peak at this value.
+Velocity ranges are based on event velocities reported in @Larsson2013.
+Since I would also like to test the algorithm on extreme data distributions, I extended the ranges beyond those found in typical eye movement data.
+
+```{r, sim-ranges}
+
+apa_table(data.frame("Event" = rep(c("Fixation", "Saccade", "PSO", "Smooth pursuit"), each = 3),
+                     "Resp. variable" = rep(c("Velocity", "Acceleration", "Angle"), 4),
+                     "Range" = c("0-50", "0-50", "uniform", "50-1000", "50-500", "~0", "20-100", "10-90", paste("~", expression(pi), sep = ""), "20-100", "0-30", "~0")), 
+          caption = "(ref:sim-ranges)",
+          note = "(ref:sim-ranges-note)")
+
+```
+
+Table X
+
+*HMM parameter values used to generate data*
+
+|State|Parameter|Interval|Default|Description|
+|-----|---------|--------|-----|-----------|
+|1-4|$\rho_i^*$|-|$1/k$|Initial state probability for starting in state *i*|
+|1-4|$a_{i=j}$|[.01,.99]|0.9|Transition probability for staying in the previous state *i*|
+|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|Transition probability for switching to from state *i* to a different state *j*|
+|1|$\alpha_{vel}$|[1,5]|3|Shape parameter of the velocity gamma distribution|
+|1|$\beta_{vel}$|[0.1,0.6]|0.35|Scale parameter of the velocity gamma distribution|
+|1|$\alpha_{acc}$|[1,5]|3|Shape parameter of the acceleration gamma distribution|
+|1|$\beta_{acc}$|[0.05,0.25]|0.15|Scale parameter of the acceleration gamma distribution|
+|1|$min^*$|-|0|Minimum of the uniform distribution|
+|1|$max^*$|-|$2\pi$|Maximum of the uniform distribution|
+|2|$\alpha_{vel}$|[1,5]|3||
+|2|$\beta_{vel}$|[5,15]|10||
+|2|$\alpha_{acc}$|[1,5]|3||
+|2|$\beta_{acc}$|[1,5]|3||
+|2|$\mu^*$|-|0|Mean parameter of the von-Mises distribution|
+|2|$\kappa$|$1/[0.1,10]$|1|Concentration parameter of the von Mises distribution
+|3|$\alpha_{vel}$|[1,5]|3||
+|3|$\beta_{vel}$|[0.5,1.5]|1||
+|3|$\alpha_{acc}$|[1,5]|3||
+|3|$\beta_{acc}$|[1,5]|3||
+|3|$\mu^*$|-|$\pi$||
+|3|$\kappa$|$1/[0.1,10]$|1||
+|4|$\alpha_{vel}$|[1,5]|3||
+|4|$\beta_{vel}$|[0.5,1.5]|1||
+|4|$\alpha_{acc}$|[1,5]|3||
+|4|$\beta_{acc}$|[0.05,0.25]|0.15||
+|4|$\mu^*$|-|0||
+|4|$\kappa$|$1/[0.1,10]$|1||
+
+*Note*. Parameters marked with $^*$ will not be varied but always set to their default values. *k* is the number of states in the model.
+
+```{r sim results and helper functions, include=FALSE}
+
+# Load simulation results
+
+for (part in 1:4) {
+
+  load(file = paste("~/Uni/Psychologie Master/Internship/GazeHMM/simulation/part", part, ".Rdata", sep = ""))
+
+}
+
+# Create functions to apply and invert mlogit link function (from depmixS4, Visser & Speekenbrink, 2019)
+
+linkfun <- function(p, base) {
+  lfun <- function(p, base) {
+    p <- p/sum(p)
+    beta <- numeric(length(p))
+    if (any(p == 1)) 
+      beta[which(p == 1)] = Inf
+    else beta[-base] <- log(p[-base]/p[base])
+    return(beta)
+  }
+  if (is.matrix(p)) {
+    beta <- t(apply(p, 1, lfun, base = base))
+  }
+  else {
+    beta <- lfun(p, base)
+  }
+  return(beta)
+}
+
+linkinv <- function(eta,base) {
+  linv <- function(eta,base) {
+    pp <- numeric(length(eta))
+    if(any(is.infinite(eta)) || any(eta > log(.Machine$double.xmax)) || any(eta < log(.Machine$double.xmin))) {
+      pp[which(is.infinite(eta))] <- 1
+      pp[which(eta > log(.Machine$double.xmax))] <- 1 # change this to something better!
+    } else {
+      expb <- exp(eta)
+      sumb <- sum(expb)
+      pp[base] <- 1/sumb
+      pp[-base] <- expb[-base]/sumb
+    }
+    return(pp)
+  }
+  if(is.matrix(eta)) {
+    if(ncol(eta)==1) {
+      pp <- as.matrix(apply(eta,1,linv,base=base)) # fixes problem with column matrix eta
+    } else pp <- t(apply(eta,1,linv,base=base)) 	
+  } else {
+    pp <- linv(eta,base)
+  }
+  return(pp)
+}
+
+# Create function to transform parameters to normal scale
+
+backtrans <- function(x) {
+
+  out <- x
+
+  nms <- names(x)
+
+  out[str_detect(nms, "(Intercept)")] <- as.vector(apply(matrix(x[str_detect(nms, "(Intercept)")],
+                                                        ncol = sqrt(length(x[str_detect(nms, "(Intercept)")])),
+                                                        byrow = T), 1, linkinv, base = 1))
+
+  out[nms %in% c("shape", "scale", "kappa")] <- exp(x[nms %in% c("shape", "scale", "kappa")])
+
+  return(out)
+}
+
+```
+
+```{r calculate RMdSPD, include=FALSE}
+
+rmsd <- list()
+
+for (part in 1:4) {
+
+  rmsd[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
+    lapply(x, function(y) {
+      sqerr <- lapply(y, function(z) {
+
+        err <- try(((backtrans(z$pars.est) - backtrans(z$pars.true))/
+                      ifelse(backtrans(z$pars.true) == 0, 2*pi, backtrans(z$pars.true)))^2)
+
+        if(is.numeric(err)) {
+
+          out <- err
+
+        } else {
+
+          out <- rep(NA, length(z$pars.true))
+
+        }
+
+        names(out) <- names(z$pars.true)
+
+        return(out)
+      })
+
+      rows <- length(sqerr)
+
+      nms <- names(sqerr[[1]])
+
+      pars <- matrix(unlist(sqerr), nrow = rows, byrow = T)
+
+      msqerr <- apply(pars, 2, median, na.rm = T)
+
+      names(msqerr) <- nms
+
+      rmsqerr <- sqrt(msqerr)
+
+      return(rmsqerr)
+    })
+  })
+}
+
+# Display RMdSPD in data frame
+
+rmsd.data <- lapply(rmsd, function(x) lapply(x, as.data.frame))
+rmsd.data <- lapply(rmsd.data, function(x) lapply(x, function(y) {as.data.frame(t(as.matrix(y)))}))
+rmsd.data <- lapply(rmsd.data, function(x) lapply(1:length(x), function(y, data) {
+  names(data[[y]]) <- names(rmsd[[1]][[y]][[1]])
+  return(data[[y]])}, data = x))
+
+# Create plots for part 1
+
+plots.rmsd.1 <- lapply(rmsd.data[[1]], function(x) {
+
+  names.pars.varied <- list(bquote(a["i=j"]), bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
+                            bquote(beta["acc"]), bquote(kappa))
+
+  if(ncol(x) == 18) {
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]))
+
+  } else if(ncol(x) == 30) {
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]))
+
+  } else {
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]), bquote(a["i4"]))
+
+  }
+
+  names.pars.est <- append(trnames, list(bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
+                         bquote(beta["acc"]), bquote(mu), bquote(kappa)))
+
+  x <- as_tibble(x, .name_repair = "unique")
+
+  data.long <- x %>%
+    mutate(par.varied = c(0, 1, 2, 3, 4, rep(c(1, 2, 3, 4, 5), (nrow(x) %/% 5)-1)),
+           state.varied = c(1, rep(1, 4), rep(2:((nrow(x) %/% 5)), each = 5))) %>%
+    pivot_longer(names(x), names_to = "par.est", values_to = "RMdSPD", names_repair = "unique") %>%
+    mutate(state.est = rep(c(rep(1:max(state.varied), max(state.varied)+1),
+                         rep(1:max(state.varied), each = 6)), nrow(x)),
+           par.est = rep(c(rep(1, max(state.varied)), 
+                           rep(2:(max(state.varied)+1), each = max(state.varied)), 
+                           rep((max(state.varied)+2):(max(state.varied)+7), max(state.varied))), nrow(x))) %>%
+    mutate_at(vars(par.varied, state.varied, par.est, state.est), as.factor)
+
+  p <- ggplot(data = data.long, aes(x = par.varied, y = par.est, fill = RMdSPD)) +
+    geom_tile() + facet_grid(cols = vars(state.varied), rows = vars(state.est)) +
+    scale_x_discrete(name = "Varied parameter", labels = names.pars.varied) +
+    scale_y_discrete(name = "Estimated parameter", labels = names.pars.est) +
+    scale_fill_distiller(breaks = c(0, 0.1, 0.5, 1), palette = "Spectral")
+
+  return(p)
+})
+
+# Create plots for parts 2,3, and 4
+
+plots.rmsd.234 <- lapply(2:4, function(y, data) lapply(data[[y]], function(x) {
+
+  if(ncol(x) == 18) {
+
+    k <- 2
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]))
+
+  } else if(ncol(x) == 30) {
+
+    k <- 3
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]))
+
+  } else {
+
+    k <-4
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]), bquote(a["i4"]))
+
+  }
+
+  names.pars.est <- append(trnames, list(bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
+                                         bquote(beta["acc"]), bquote(mu), bquote(kappa)))
+
+  x <- as_tibble(x, .name_repair = "unique")
+
+  data.long <- x %>%
+    mutate(cond = 1:3) %>%
+    pivot_longer(names(x), names_to = "par.est", values_to = "RMdSPD", names_repair = "unique") %>%
+    mutate(state.est = rep(c(rep(1:k, k+1),
+                             rep(1:k, each = 6)), nrow(x)),
+           par.est = rep(c(rep(1, k), 
+                           rep(2:(k+1), each = k), 
+                           rep((k+2):(k+7), k)), nrow(x))) %>%
+    mutate_at(vars(cond, par.est, state.est), as.factor)
+
+  p <- ggplot(data = data.long, aes(x = par.est, y = RMdSPD, color = cond)) +
+    geom_point(position = position_dodge(0.25)) + facet_grid(cols = vars(state.est)) +
+    scale_x_discrete(name = "Estimated parameter", labels = names.pars.est) +
+    scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2)) +
+    geom_hline(yintercept = 0.1, linetype = "dashed") +
+    geom_hline(yintercept = 0.5, linetype = "dashed")
+
+  if(y == 2) {
+    names.cond <- c("500", "2500", "10000")
+    label.cond <- "N"
+  } else if (y == 3) {
+    names.cond <- c("1", "2", "3")
+    label.cond <- bquote(tau["start"])
+  } else {
+    names.cond <- c("1", "3", "5")
+    label.cond <- "m"
+  }
+
+  p <- p + scale_color_discrete(name = label.cond, labels = names.cond)
+
+  return(p)
+}), data = rmsd.data)
+
+```
+
+```{r calculate linear regressions, include=FALSE}
+
+# Calculate regression weights for transition probabilities
+
+regw.tr <- list()
+
+regw.tr <- lapply(get("estimates.1"), function(x) {
+
+  varpar <- lapply(x[1], function(y) {
+
+    lapply(y, function(z) {
+
+      nms <- names(z$pars.true)
+
+      pars.tr <- logical(length(z$pars.true))
+
+      pars.tr[str_detect(nms, "(Intercept)")] <- T
+
+      intpar <- try(cbind(backtrans(z$pars.true[pars.tr]), backtrans(z$pars.est[pars.tr])))
+
+      if(is.numeric(intpar)) {
+        out <- intpar
+      } else {
+        out <- matrix(NA, nrow = length(z$pars.true[pars.tr]), ncol = 2)
+      }
+
+      out <- apply(out, 1, list)
+
+      return(out)
+    })
+  })
+
+  df <- list()
+
+  for (i in 1:length(varpar[[1]][[1]])) {
+
+    df[[i]] <- lapply(varpar, function(y, index) {
+
+      lapply(y, function(z) {z[[index]]})
+
+    }, index = i)
+  }
+
+  df <- lapply(df, as.data.frame)
+
+  df <- lapply(df, function(z) {
+
+    out <- as.data.frame(t(as.matrix(z)))
+
+    names(out) <- c("true", "est")
+
+    return(out)
+  })
+})
+
+# Calculate linear regression weights for response parameters
+
+regw.resp <- list()
+
+regw.resp <- lapply(get("estimates.1"), function(x) {
+
+  index <- 1:length(x[-1])
+
+  varpar <- lapply(index, function(i, y) {
+
+    lapply(y[[i]], function(z) {
+
+      nms <- names(z$pars.true)
+
+      #pars.tr <- logical(length(z$pars.true))
+      pars.resp <- logical(length(z$pars.true))
+
+      #pars.tr[str_detect(nms, "(Intercept)")] <- T
+      pars.resp[nms %in% c("shape", "scale", "kappa")] <- T
+
+      intpar <- try(c(backtrans(z$pars.true[pars.resp][i]), backtrans(z$pars.est[pars.resp][i])))
+
+      if(is.numeric(intpar)) {
+        return(intpar)
+      } else {
+        return(rep(NA, 2))
+      }
+    })
+  }, y = x[-1])
+
+  df <- lapply(varpar, as.data.frame)
+  df <- lapply(df, function(z) {
+
+    out <- as.data.frame(t(as.matrix(z)))
+
+    names(out) <- c("true", "est")
+
+    return(out)
+  })
+})
+
+# Plot linear regressions for transition probabilities
+
+D <- 100
+
+regw.tr.data <- lapply(regw.tr, function(x) lapply(x, function(y) rbind(y)))
+regw.tr.data <- lapply(regw.tr.data, function(x) reduce(x, rbind))
+
+plots.lm.tr <- lapply(1:length(regw.tr.data), function(x, y) {
+
+  data <- y[[x]] %>% 
+    mutate(par = rep(1:(x+1)^2, each = D),
+           from = rep(rep(1:(x+1), each = D), x+1),
+           to = rep(1:(x+1), each = D*(x+1)))
+
+  p <- ggplot(data, aes(x = true, y = est)) + 
+    facet_grid(rows = vars(from), cols = vars(to), labeller = label_both) +
+    geom_point() + geom_smooth(method = "lm") +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    scale_x_continuous(name = "True transition probability") +
+    scale_y_continuous(name = "Estimated transition probability")
+
+  return(p)
+}, y = regw.tr.data)
+
+# Plot linear regressions for response parameters
+
+regw.resp.data <- lapply(regw.resp, function(x) lapply(x, function(y) rbind(y)))
+regw.resp.data <- lapply(regw.resp.data, function(x) reduce(x, rbind))
+
+parnames <- list(bquote(alpha["vel;1"]), bquote(beta["vel;1"]), bquote(alpha["acc;1"]), bquote(beta["acc;1"]),
+                 bquote(alpha["vel;2"]), bquote(beta["vel;2"]), bquote(alpha["acc;2"]), bquote(beta["acc;2"]), bquote(kappa["2"]),
+                 bquote(alpha["vel;3"]), bquote(beta["vel;3"]), bquote(alpha["acc;3"]), bquote(beta["acc;3"]), bquote(kappa["3"]),
+                 bquote(alpha["vel;4"]), bquote(beta["vel;4"]), bquote(alpha["acc;4"]), bquote(beta["acc;4"]), bquote(kappa["4"]))
+
+plots.lm.resp <- lapply(regw.resp.data, function(x) {
+
+  npar <- nrow(x) %/% D
+
+  data <- x %>% 
+    mutate(par = rep(1:npar, each = D),
+           type = rep(1, 2, 1, 2))
+
+  p <- ggplot(data, aes(x = true, y = est)) + 
+    facet_wrap(vars(par), scales = "free", labeller = label_bquote(.(parnames[[par]]))) +
+    geom_point() + geom_smooth(method = "lm") +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    scale_x_continuous(name = "True transition probability") +
+    scale_y_continuous(name = "Estimated transition probability")
+
+  return(p)
+})
+
+```
+
+```{r calculate accuracy, include=FALSE}
+
+# Summarise accuracy
+
+acc.data <- list()
+
+for (part in 1:4) {
+
+  acc.data[[part]] <- lapply(get(paste("estimates.", part, sep = "")), function(x) {
+    out <- lapply(x, function(y) { 
+      out <- lapply(y, function(z) {
+
+        if(is.numeric(z$accuracy)) {
+          acc <- z$accuracy
+        } else {
+          acc <- NA
+        }
+
+        return(acc)
+      })
+
+      return(as.vector(reduce(out, cbind)))
+    })
+
+    return(as.data.frame(t(reduce(out, cbind))))
+  })
+}
+
+# Plot accuracy part 1
+
+plots.acc.1 <- lapply(acc.data[[1]], function(x) {
+
+  names.pars.varied <- list(bquote(a["i=j"]), bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
+                            bquote(beta["acc"]), bquote(kappa))
+
+  x <- as_tibble(x, .name_repair = "unique")
+
+  data.long <- x %>%
+    mutate(par.varied = c(0, 1, 2, 3, 4, rep(c(1, 2, 3, 4, 5), (nrow(x) %/% 5)-1)),
+           state.varied = c(1, rep(1, 4), rep(2:((nrow(x) %/% 5)), each = 5))) %>%
+    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
+    mutate_at(vars(par.varied, state.varied), as.factor)
+
+  p <- ggplot(data = data.long, aes(x = par.varied, y = accuracy)) +
+    geom_boxplot(outlier.shape = 4) + facet_grid(cols = vars(state.varied)) +
+    scale_x_discrete(name = "Varied parameter", labels = names.pars.varied) + 
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-0.33, 0, 0.25, 0.5, 0.75, 1))
+
+  return(p)
+})
+
+# Plot accuracy parts 2, 3, and 4
+
+acc.data.234 <- lapply(acc.data[2:4], function(x) reduce(x, rbind))
+
+plots.acc.234 <- lapply(1:3, function(y, data) {
+
+  x <- as_tibble(data[[y]], .name_repair = "unique")
+
+  data.long <- x %>%
+    mutate(cond = as.factor(rep(1:3, 3)),
+           k = as.factor(rep(2:4, each = 3))) %>%
+    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy")
+
+  p <- ggplot(data = data.long, aes(x = k, y = accuracy, color = cond)) +
+    geom_boxplot(outlier.shape = 4) + 
+    scale_x_discrete(name = "k (number of states)")
+
+  if(y == 1) {
+    names.cond <- c("500", "2500", "10000")
+    label.cond <- "N"
+  } else if (y == 2) {
+    names.cond <- c("1", "2", "3")
+    label.cond <- bquote(tau["start"])
+  } else {
+    names.cond <- c("1", "3", "5")
+    label.cond <- "m"
+  }
+
+  p <- p + scale_color_discrete(name = label.cond, labels = names.cond)
+
+  return(p)
+}, data = acc.data.234)
+
+# Plot accuracy over interval parts 2 and 4
+
+plots.acc.int.24 <- lapply(c(1, 3), function(y, data) {
+
+  x <- as_tibble(data[[y]], .name_repair = "unique")
+
+  if(y == 1) {
+
+    int <- rep(seq(1, 5, length.out = D), 9)
+
+  } else {
+
+    int <- rep(floor(seq(1, 200, length.out = D)), 9)
+
+  }
+
+  data.long <- x %>%
+    mutate(cond = as.factor(rep(1:3, 3)),
+           k = as.factor(rep(2:4, each = 3))) %>%
+    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
+    mutate(int = int)
+
+  p <- ggplot(data = data.long, aes(x = int, y = accuracy, color = cond)) +
+    facet_grid(cols = vars(k), labeller = label_both) + 
+    geom_point()
+
+  if(y == 1) {
+    names.cond <- c("500", "2500", "10000")
+    label.cond <- "N"
+    x.name <- bquote(tau["noise"])
+  } else {
+    names.cond <- c("1", "3", "5")
+    label.cond <- "m"
+    x.name <- "l"
+  }
+
+  p <- p + scale_color_discrete(name = label.cond, labels = names.cond) +
+    scale_x_continuous(name = x.name) + 
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-0.33, 0, 0.25, 0.5, 0.75, 1))
+
+  return(p)
+}, data = acc.data.234)
+
+```
 
@@
 In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation between the true and estimated parameters and the accuracy between the true and estimated state sequence.
-Figure X displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM with two states.
+Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM with two states.
 The RMdSPD was below 0.1 for all estimated and manipulated parameters (except the irrelevant initial state probabilities).
 Thus, no matter which relevant parameter changed in the HMM, the median deviation did not exceed 10% of the true parameter value.
-The regressions between manipulated true and estimated parameters for two states are shown in Figures X and X.
+The regressions between manipulated true and estimated parameters for two states are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
 With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely: Changing the true parameters led to linear changes in the estimated parameters.
 However, the deviation between true and estimated response parameters seemed to increase slighlty with parameter magnitude.
-Considering accuracy, Figure X displays Cohen's kappa between true and estimated hidden state sequences for two states.
+Considering accuracy, Figure \@ref(fig:plot-acc-1-2) displays Cohen's kappa between true and estimated hidden state sequences for two states.
 With two exceptions, kappa values were almost one, indicating nearly perfect agreement.
@@
 
-For three states, the RMdSPD is shown in Figure X.
+(ref:plot-rmdspd-1-2) RMdSPD between true and estimated parameters of the two-state HMM in part one of the simulation.
+Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
+Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-1-2, fig.cap="(ref:plot-rmdspd-1-2)"}
+
+print(plots.rmsd.1[[1]])
+
+```
+
+(ref:plot-regtr-1-2) Regression lines between true and estimated transition probabilities for the two-state HMM in part one.
+Top facet labels show to and right facet labels show from which state the HMM is moving.
+
+```{r plot-regtr-1-2, fig.cap="(ref:plot-regtr-1-2)"}
+
+print(plots.lm.tr[[1]])
+
+```
+
+(ref:plot-regresp-1-2) Regression lines between true and estimated transition response parameters of the two-state HMM in part one. Top facet labels indicate response parameters.
+
+```{r plot-regresp-1-2, fig.cap="(ref:plot-regresp-1-2)"}
+
+print(plots.lm.resp[[1]])
+
+```
+
+(ref:plot-acc-1-2) Boxplots displaying Cohen's kappa depending on which parameter of the two-state HMM has been manipulated in part one.
+Top facet labels indicate for which state parameters have been manipulated.
+Black solid lines symbolize medians.
+Crosses represent outliers (distance to first/third quartile higher than 1.5 times the inter-quartile range [IQR]).
+
+```{r, plot-acc-1-2, fig.cap="(ref:plot-acc-1-2)"}
+
+print(plots.acc.1[[1]])
+
+```
+
+For three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
 When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently below 0.5.
@@
 Otherwise, RMdSPDs were consistently lower than 0.1, indicating a median deviation between true and estimated parameters below 10% of the true parameter.
-Inspecting the regressions between manipulated true and estimated parameters (see Figures X and X) revealed a strong and unbiased linear relationship (intercepts close to zero and slopes close to one).
+Inspecting the regressions between manipulated true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed a strong and unbiased linear relationship (intercepts close to zero and slopes close to one).
 Again, the deviation seemed to increase with true parameter magnitude (reverse for kappas).
 In contrast to the two-state HMM, larger deviatons and more outliers were observed.
-Cohen's kappa values for the three-state HMM are presented in Figure X.
+Cohen's kappa values for the three-state HMM are presented in Figure \@ref(fig:plot-acc-1-3).
 For most estimated models, the kappas between true and estimated state sequences were above 0.95, indicating almost perfect agreement.
@@
 
-The RMdSPDs for the four-state HMM is illustrated in Figure X.
+(ref:plot-rmdspd-1-3) RMdSPD between true and estimated parameters of the three-state HMM in part one of the simulation.
+Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
+Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-1-3, fig.cap="(ref:plot-rmdspd-1-3)"}
+
+print(plots.rmsd.1[[2]])
+
+```
+
+(ref:plot-regtr-1-3) Regression lines between true and estimated transition probabilities for the three-state HMM in part one.
+Top facet labels show to and right facet labels show from which state the HMM is moving.
+
+```{r plot-regtr-1-3, fig.cap="(ref:plot-regtr-1-3)"}
+
+print(plots.lm.tr[[2]])
+
+```
+
+(ref:plot-regresp-1-3) Regression lines between true and estimated transition response parameters of the three-state HMM in part one. Top facet labels indicate response parameters.
+
+```{r plot-regresp-1-3, fig.cap="(ref:plot-regresp-1-3)"}
+
+print(plots.lm.resp[[2]])
+
+```
+
+(ref:plot-acc-1-3) Boxplots displaying Cohen's kappa depending on which parameter of the three-state HMM has been manipulated in part one.
+Top facet labels indicate for which state parameters have been manipulated.
+Black solid lines symbolize medians and hinges the first and third quartile.
+Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
+Crosses represent outliers.
+
+```{r, plot-acc-1-3, fig.cap="(ref:plot-acc-1-3)"}
+
+print(plots.acc.1[[2]])
+
+```
+
+The RMdSPDs for the four-state HMM is illustrated in Figure \@ref(fig:plot-rmdspd-1-4).
 For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were below 0.5.
@@
 Otherwise, RMdSPDs were below 0.1, indicating a median deviation below 10% of the true parameter.
-Looking at the regressions between true and estimated paramters, Figure X shows strong and unbiased relationships but larger deviations and more outliers than in the previous models, especially for states one and four.
-Accuracy indicated by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences.
+Looking at the regressions between true and estimated paramters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) show strong and unbiased relationships but larger deviations and more outliers than in the previous models, especially for states one and four.
+Accuracy indicated by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
 Here, some kappa values clustered around 0.25 and zero, which again can be interpreted as the results of label switching.
@@
 Especially the recovery of transition parameters in states corresponding to fixations and smooth pursuits decreased.
-
 Overall, simulations in part one demonstrated that the HMM recovered parameters at least moderately well when parameters were manipulated (all RMdSPDs below 0.5).
@@
 
+(ref:plot-rmdspd-1-4) RMdSPD between true and estimated parameters of the four-state HMM in part one of the simulation.
+Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
+Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-1-4, fig.cap="(ref:plot-rmdspd-1-4)"}
+
+print(plots.rmsd.1[[3]] + theme(axis.text = element_text(size = 6)))
+
+```
+
+(ref:plot-regtr-1-4) Regression lines between true and estimated transition probabilities for the four-state HMM in part one.
+Top facet labels show to and right facet labels show from which state the HMM is moving.
+
+```{r plot-regtr-1-4, fig.cap="(ref:plot-regtr-1-4)", fig.height=8}
+
+print(plots.lm.tr[[3]] + theme(axis.text = element_text(size = 8)))
+
+```
+
+(ref:plot-regresp-1-4) Regression lines between true and estimated transition response parameters of the four-state HMM in part one. Top facet labels indicate response parameters.
+
+```{r plot-regresp-1-4, fig.cap="(ref:plot-regresp-1-4)", fig.height=8}
+
+print(plots.lm.resp[[3]] + theme(axis.text = element_text(size = 6)))
+
+```
+
+(ref:plot-acc-1-4) Boxplots displaying Cohen's kappa depending on which parameter of the four-state HMM has been manipulated in part one.
+Top facet labels indicate for which state parameters have been manipulated.
+Black solid lines symbolize medians and hinges the first and third quartile.
+Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
+Crosses represent outliers.
+
+```{r, plot-acc-1-4, fig.cap="(ref:plot-acc-1-4)"}
+
+print(plots.acc.1[[3]] + theme(axis.text = element_text(size = 8)))
+
+```
+
 ### Sample size and noise variation
@@
 In the second part, I varied the sample size of the HMM and added noise to the generated data.
-For the two-state HMM, the RMdSPDs were above 0.5 for $\beta_{vel}$ and $\beta_{acc}$ in both states (see Figure X).
+For the two-state HMM, the RMdSPDs were above 0.5 for $\beta_{vel}$ and $\beta_{acc}$ in both states (see Figure \@ref(fig:plot-rmdspd-2-2)).
 The other estimated parameters showed RMdSPDs close to or below 0.1.
@@
 For $\beta_{vel}$ and $\beta_{acc}$ in both states, models with 2500 samples had the lowest RMdSPDs.
-Accuracy measured by Cohen's kappa was almost perfect with kappa values very close to one (see Figure X, left plot).
+Accuracy measured by Cohen's kappa was almost perfect with kappa values very close to one (see Figure \@ref(fig:plot-acc-2), left plot).
 To conclude, adding noise did not affect the parameter recover or accuracy of the two-state HMM, whereas increasing the sample size improved the recovery slightly.
 
-The RMdSPDs for the three-state HMM and $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states.
+(ref:plot-rmdspd-2-2) RMdSPD between true and estimated parameters of the two-state HMM in part two of the simulation.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-2-2, fig.cap="(ref:plot-rmdspd-2-2)"}
+
+print(plots.rmsd.234[[1]][[1]])
+
+```
+
+(ref:plot-acc-2) Cohen's kappa depending on the variation of noise added to the data generated by the HMM.
+Colours indicate different sizes of generated data.
+Top facet labels indicate the number of states in the HMM.
+
+```{r, plot-acc-2, fig.cap="(ref:plot-acc-2)"}
+
+print(plots.acc.int.24[[1]])
+
+```
+
+The RMdSPDs for the three-state HMM and $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)).
 Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5.
 For most parameters in all three states, higher sample sizes had lower RMdSPDs.
-The accuracy of the estimated models was almost perfect with most kappa values above 0.95 (see Figure X, middle plot).
+The accuracy of the estimated models was almost perfect with most kappa values above 0.95 (see Figure \@ref(fig:plot-acc-2), middle plot).
 Several outliers clustered around kappas of zero and -0.33, signaling label switching.
@@
 
-RMdSPDs regarding the four-state HMM are displayed in Figure X.
+(ref:plot-rmdspd-2-3) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-2-3, fig.cap="(ref:plot-rmdspd-2-3)"}
+
+print(plots.rmsd.234[[1]][[2]])
+
+```
+
+RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
 For states one and four, values for most parameters (including all transition probabilites) were above 0.5.
@@
 For states two and three, higher samples sizes showed slightly lower RMdSPDs.
-As in the previous part, most Cohen's kappa values ranged between 0.6 and 0.9, meaning substantial to almost perfect agreement between true and estimated states (Figure X, right plot).
+As in the previous part, most Cohen's kappa values ranged between 0.6 and 0.9, meaning substantial to almost perfect agreement between true and estimated states (Figure \@ref(fig:plot-acc-2), right plot).
 Multiple kappa values clustered around 0.25 or zero, which can be explained by label switching.
@@
 For saccade and PSO states, only scale parameters of gamma distributions were badly recovered, but increasing the sample size slightly improved the recovery.
-
 In general, the HMM recovered parameters well despite noise being added to the data.
@@
 The accuracy of the model was slightly lowered when including more states, but it was neither affected by the range of noise $\tau_{noise}$ nor the sample size.
+
+(ref:plot-rmdspd-2-4) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-2-4, fig.cap="(ref:plot-rmdspd-2-4)"}
+
+print(plots.rmsd.234[[1]][[3]])
+
+```
 
@@
 Increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor accuracy.
-
 Since the manipulation in this part was not effective, I decided to explore post-hoc whether clusters of low Cohen's kappa values were due to label switching.
@@
 
+## Validation
+### Model comparison
+
+To test whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states to two benchmark data sets and compared their Schwarz weights.
+For the Andersson et al.
+(2017) data set, I expected the highest weights for the three-state models in the image condition.
+For the moving dots and video condition, I predicted the highest weights for the four-state models.
+Figure X shows the Schwarz weights for different subjects and models in the image condition.
+For all subjects, the five-state model displays the highest weight, suggesting that it is the most likely to have generated the data.
+In the moving dots condition, the five-state model has the highest weights for most subjects, but the one-, three, and four-state models exhibit the highest weights for two subjects each (see Figure X).
+The video condition has the same pattern as the image condition, as the five-state model consistently has the highest weights (see Figure X).
+In contrast to my hypothesis, applying different HMMs to the Andersson et al.
+data suggests that the five-state model has most likely generated the data.
+Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the likelihoods of the models.
+
+Ehinger data set
+
+### Comparison against other algorithms
+
+The hypothesis that HMMs improve the classification performance of eye movement events was investigated by comparing gazeHMM against other algorithms.
+Therefore, I applied gazeHMM with four events to the Andersson et al.
+(2017) data set.
+First, I calculated the RMSD [^2] as described in the original article by Andersson et al.
+for all algorithms included in the study plus gazeHMM.
+Table X shows that in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations due to a large amount of very short fixations.
+For saccades, gazeHMM has the fourth lowest RMSD of the algorithms in the image and video conditions but the second highest in the moving dots condition (see Table X).
+Here, gazeHMM overestimated the number and duration of saccades (either slightly or substantially).
+Moreover, gazeHMM underestimated the number and duration of PSOs in the image and moving dots conditions, leading to the highest and intermediate RMSD among the competitors, repsectively (see Table X).
+In the video condition the duration was slightly overestimated but the duration was underestimated with the highest RMSD.
+No other algorithm was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than among the human coders (see Figure X).
+In sum, except for saccades and PSOs, gazeHMM performed worse with regard to the duration of classified events compared to the other algorithms.
+
+[^2]: In my proposal, I declared to compute the RMSE instead of RMSD.
+However, since I conducted an in sample comparison the term deviation is more conventional.
+Andersson et al.
+(2017) also referred to it as RMSD.
+
+Second, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of agreement between algorithms and human coders (see Table X).
+For fixations, absolute kappa values in all three conditions indicated a slight to fair agreement between gazeHMM and human coders.
+Compared to the other algorithms, the agreement of gazeHMM was the lowest in the image condition but the hightest in the moving dots and video conditions.
+In contrast, gazeHMM showed absolute kappa values for saccades that correspond to a moderate to substantial agreement.
+In all three conditions, the agreement was intermediate in the moving dots condition and higher than for most other algorithms in the image and video conditions.
+For PSOs, absolute kappa values showed slight to fair agreement agreement to human coders.
+It was lower in the moving dots condition but intermediate in the image and video conditions compared to the other algorithms.
+The agreement for smooth puruits was moderate in the moving dots condition but slight in the image and video conditions.
+No other algorithm in the study was designed to detect smooth pursuits.
+In sum, gazeHMM classified samples as saccades more similar to human coders than most other algorithms.
+For the other events, the classification differed from human coding more than for most other algorithms.
+
+Confusion rate and matrix
+
+Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms with regard to RMSD, sample-to-sample agreement, or disagreement ratio.
+It rather fared better or worse than most algorithms for different events and conditions.
+Especially for saccades, the agreement was relatively high and improved upon the IHMM algorithm included in the study.
+Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data might be more appropriate and yield better validation results.
+Table X shows the RMSD between gazeHMM and human coders only for the image condition.
+For fixations, gazeHMM shows a comparably low RMSD and only two other algorithms have lower values.
+Oppositely, saccades had the highest RMSD among the compared algorithms.
+The same result was found for PSOs.
+The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table X (only for the image condition).
+For fixations, the absolute kappa indicated substantial agreement and was high compared to the other algorithms.
+The absolute kappa for saccades corresponded to moderate agreement.
+Compared to the other algorithms, it was rather low.
+For PSOs, the absolute agreement was fair, but intermediate with regard to the other algorithms.
+Summing up, using gazeHMM with three states increased the agreement to human coding with regard to fixation durations and samples classified as fixations.
+However, for saccades and PSOs the agreement decreased.
+
 # Technical Appendix

### 2020-06-10 manuscript/internship_report.Rmd
update report

@@
 library(depmixS4)
+library(psych)
 source(paste(dir, "algorithm/preprocessing_helper_functions.R", sep = "/"))
@@
 # Algorithm Development
-
-As part of answering my research questions, I developed an algorithm named *gazeHMM* to classify gaze data into discrete eye movement events.
+As part of answering my research questions, I developed an algorithm named gazeHMM to classify gaze data into discrete eye movement events.
 The following section describes the development of gazeHMM and its final version that has been used to obtain results.
@@
 ## Eye movement metrics
-
-Many different metrics can be used to describe gaze data and eye movement events (Zemblys et al.).
+Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018].
 Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best.
@@
 This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics.
-Since I proposed to choose metrics that do not rely on such user settings, I used velocity, acceleration, and sample-to-sample angle as metrics in gazeHMM.
+Belonging to the most basic metrics, which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [similar to relative or change in angle; @Larsson2013] in gazeHMM.
 Theoretically, these three metrics should separate eye movement events clearly.
-Fixations typically inherit samples with low velocity and acceleration (CITE).
-Due to tremor, the angle between samples should not follow any direction but a random walk (CITE).
-In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction (CITE).
-PSO samples tend to have moderate velocity and high acceleration since they occur between high velocity saccades and low velocity fixations.
-They are specifically distinguished by their change in direction clustered around 180° (CITE).
-Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction (CITE).
+Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
+Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017].
+In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction.
+PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013].
+They are specifically distinguished by their change in direction clustered around 180° [@Pekkanen2017].
+Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
 
 ## Fitering and smoothing
-
 As with metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics.
-Their purpose is to remove noise or artifacts from the gaze data that could distort the classification (CITE).
+Their purpose is to remove noise or artifacts from the gaze data that could distort the classification [@Duchowski2017].
 Filtering (smoothing) methods that are applied before computing the eye movement metrics target the recorded gaze position.
-Unfortunately, this might also erase differences in the metrics between samples.
-For instance, tremor movements during fixations would be filtered out and, in consequence, the samples would follow the same direction.
-This would make it harder for the algorithm to separate fixations from smooth pursuits.
-The same problem applies to PSOs and saccades where filtering could erase the oscillations (see Figure \@ref(fig:filter-angle)).
-
-(ref:filter-angle) Example data from Andersson et al.
-(2017) displayed as sample-to-sample angle (in rad) over time (in s).
+While removing noise, filters might also erase differences in the metrics between samples.
+For instance, tremor movements during fixations could be filtered out and in consequence, the samples would follow the same direction.
+This could complicate separating fixations from smooth pursuits.
+The same problem applies to PSOs and saccades where filtering could erase oscillations (see Figure \@ref(fig:filter-angle)).
+
+(ref:filter-angle) Example data from @Andersson2017 displayed as sample-to-sample angle (in rad) over time (in s).
 Line color indicates whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue).
 For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3.
-
-```{r, filter-angle, results='asis', fig.cap="(ref:filter-angle)"}
+Full line segment marks where potential oscillations are filtered out.
+Dashed line segment illustrates where potentially random change in angle is filtered out.
+
+```{r, filter-angle, fig.cap="(ref:filter-angle)"}
 
@@
   geom_path(aes(x = t, y = force_neg_pi_pi(angle.filt)), color = "red") +
+  annotate(geom = "segment", x = 0.525, xend = 0.54, y = -3, yend = -3, color = "black", size = 0.8) +
+  annotate(geom = "segment", x = 0.635, xend = 0.67, y = -3, yend = -3, color = "black", size = 0.8, linetype = 6) +
   scale_y_circular(name = "Sample-to-sample angle (in rad)", units = "radians", limits = c(-pi, pi)) +
@@
 Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration).
-(CITE) compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively, CITE) of the gaze position to be used by their algorithm.
-Similarly, (CITE) use a Savitzky-Golay (SG) filter (CITE) to estimate velocity and acceleration from a gaze signal.
+@Houpt2018 compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm.
+Similarly, @Holmqvist2010 use a Savitzky-Golay (SG) filter @Savitzky1964 to estimate velocity and acceleration from a gaze signal.
 Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods.
-They follow a similar trend, but the SG velocity signal is less noisy and, consequently, the SG acceleration signal is lower than the Laplace acceleration signal.
-I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
-To preserve motor noise in the change in angle signal, gazeHMM computes the first discrete derivative of sample angle directly (see Technical Appendix).
+They follow a similar trend, but the SG velocity signal is less noisy and acceleration signal is consequently lower than the Laplace acceleration signal.
+Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
+To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the backward difference (see Technical Appendix).
 
 (ref:filter-comparison) Example data from Andersson et al.
-(2017) displayed as velocity (deg/seconds) and acceleration (in deg/s^2) over time (in s).
+(2017) displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s).
 Line colors indicate which filter has been used to derive the signal.
@@
 
-```{r, filter-comparison, results='asis', fig.cap="(ref:filter-comparison)"}
+```{r, filter-comparison, fig.cap="(ref:filter-comparison)"}
 
@@
 ## The generative model
-
 The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM).
-It can have between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
+It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
 Thus, users can choose wether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth puruits.
-In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model (CITE).
+In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model [@Visser2019].
 In gazeHMM, the response model consists of three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal.
@@
 I acknowledge that conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
-This step was taken merely to keep the HMM simple and identifyable.
+This step was merely taken to keep the HMM simple and identifyable.
 Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them).
 However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation) and due to noise above zero.
-Second, the distributions of both signals appear to be positively skewed conditionally on the states and, third, to have variances increasing with their mean.
+Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean.
 Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties.
 In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose.
-It has to be noted that the gamma was choosen because of convenience and the best fitting distribution might be different between subjects and tasks.
-To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution to model the derivative of sample angle.
+It has to be noted that the gamma was choosen because of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
+To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution.
 Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$) which should lead to rather symmetric distributions.
@@
 ## Missing data and blinks
-
 HMMs can estimate parameters and hidden states despite missing data.
-They either treat data as missing at random or as missing dependent on states (Visser & Speekenbrink).
-In gazeHMM, I assumed data to be missing at random.
+They either treat data as missing at random or as missing dependent on states [@Visser2019].
+In gazeHMM, I assume data to be missing at random.
 Most of the missing data in eye movement classification are due to blinks.
@@
 When evaluating the likelihood of the HMM given parameters and the joint density of response distributions for missing data, the model integrates over all possible values.
-The resulting density for missing samples is therefore set to one (Visser & Speekenbrink).
+The resulting density for missing samples is therefore set to one [@Visser2019].
 
 ## Optimization and classification
-
-The paramters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm (CITE).
+The paramters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997].
 The EM is generally suitable to estimate likelihoods with missing variables.
-For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data (velocity, acceleration, and sample-to-sample angle) and the expected hidden states (eye movement events; Visser & Speekenbrink).
-The sequence of hidden states is estimated through the Viterbi algorithm (CITE) by maximizing the posterior state probability (Visser & Speekenbrink).
-Parameters of the response distributions (except for the uniform distribution) were optimized on the log-scale (except for the mean parameter of the von-Mises distribution) using a spectral projected gradient method (CITE) and Barzilai-Borwein steplengths (CITE).
+For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data (velocity, acceleration, and sample-to-sample angle) and the expected hidden states [eye movement events; @Visser2019].
+The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability.
+Parameters of the response distributions (except for the uniform distribution) were optimized on the log-scale (except for the mean parameter of the von-Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein steplengths [@Barzilai1988].
 
 ## Postprocessing
-
 After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence.
 I implemented this rountine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
-Moreover, gazeHMM did explicitly control the duration of events in the HMM which ocassionally led to unreasonably short events.
+Moreover, gazeHMM did not explicitly control the duration of events in the HMM which ocassionally led to unreasonably short events.
 Thus, the postprocessing routine heuristically compensates for such violations.
@@
 ## Implementation
-
-The algorithm is implemented in R (version 3.6.3; CITE) and uses the packages signal (CITE) to compute velocity and acceleration signals, dplyr (CITE) to calculate sample-to-sample angles, depmixS4 (CITE) for the HMM, and BB (CITE) for Barzilai-Borwein spectral projected gradient optimization.
-The algorithm will be available on Github (LINK).
+The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal (CITE) to compute velocity and acceleration signals, dplyr (CITE) to calculate sample-to-sample angles, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
+The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
 
 # Simulation Study
-
 To assess how well the HMM recovers parameters and state sequences, I conducted a simulation study.
-The design and analysis of the study were preregistered on OSF (LINK).
-The major part of this section is a direct copy of that preregistration.
+The design and analysis of the study were preregistered on the Open Science Framework (https://doi.org/10.17605/OSF.IO/VDJGP).
+The major part of this section is a direct copy of the preregistration.
 When appropriate, additional explanations were added and tenses adapted to make the design easier to read and understand.
@@
 ### Parameter Variation
-
 The simulation study was divided into four parts.
@@
 ### Sample Size and Noise Variation
-
 In the second part, I varied the sample size of the generated data and the amount of noise added to it.
@@
 ### Variation of Starting Values
-
 In the third part, I increased the variation in the starting values used for parameter estimation.
@@
 ### Missing data
-
 In the last part, I set intervals of the generated data to be missing.
@@
 ## Starting values
-
 The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
@@
 ## Data analysis
-
 For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values[^1]: $$RMdSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
@@
 apa_table(data.frame("Event" = rep(c("Fixation", "Saccade", "PSO", "Smooth pursuit"), each = 3),
-                     "Resp. variable" = rep(c("Velocity", "Acceleration", "Angle"), 4),
+                     "Resp variable" = rep(c("Velocity", "Acceleration", "Angle"), 4),
                      "Range" = c("0-50", "0-50", "uniform", "50-1000", "50-500", "~0", "20-100", "10-90", paste("~", expression(pi), sep = ""), "20-100", "0-30", "~0")), 
           caption = "(ref:sim-ranges)",
-          note = "(ref:sim-ranges-note)")
+          note = "(ref:sim-ranges-note)",
+          escape = F)
+
+```
+
+```{r, sim-parameters}
+
+sim.pars.df <- data.frame(State = c(rep("1-4", 3), as.character(rep(1:4, each = 6))),
+                          #Parameter = c(""),
+                          Interval = c("-", "[.01,.99]", "1-a(i=j)/(k-1)", 
+                                       "[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
+                                       "[1,5]", "[5,15]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
+                                       "[1,5]", "[0.5,1.5]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
+                                       "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]"),
+                          Default = c("1/k", "0.9", "0.1/(k-1)", 
+                                      "3", "0.35", "3", "0.25", "0", paste(expression(pi), sep = ""),
+                                      "3", "10", "3", "3", "0", "1",
+                                      "3", "1", "3", "3", paste(expression(pi), sep = ""), "1",
+                                      "3", "1", "3", "0.15", "0", "1"),
+                          Description = c("Initial state probability for starting in state *i*",
+                                          "Transition probability for staying in the previous state *i*",
+                                          "Transition probability for switching to from state *i* to a different state *j*",
+                                          "Shape parameter of the velocity gamma distribution",
+                                          "Scale parameter of the velocity gamma distribution",
+                                          "Shape parameter of the acceleration gamma distribution",
+                                          "Scale parameter of the acceleration gamma distribution",
+                                          "Minimum of the uniform distribution",
+                                          "Maximum of the uniform distribution",
+                                          "", "", "", "",
+                                          "Mean parameter of the von-Mises distribution",
+                                          "Concentration parameter of the von Mises distribution",
+                                          rep("", 12)))
+
+sim.pars.names <- c(expression(rho[i]), expression(a[i==j]), expression(a[i!=j]),
+                   expression(alpha[vel]))
+
+rownames(sim.pars.df)[1:4] <- sim.pars.names
+
+apa_table(sim.pars.df, row.names = T)
 
@@
     names.cond <- c("1", "3", "5")
-    label.cond <- "m"
+    label.cond <- "Number of\nintervals"
   }
@@
     geom_boxplot(outlier.shape = 4) + 
-    scale_x_discrete(name = "k (number of states)")
+    scale_x_discrete(name = "k (number of states)")  + 
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-0.33, 0, 0.25, 0.5, 0.75, 1))
 
@@
     names.cond <- c("1", "3", "5")
-    label.cond <- "m"
-    x.name <- "l"
+    label.cond <- "Number of\nintervals"
+    x.name <- "Interval length (in samples)"
   }
@@
 
+# Exploratory analysis for label switching
+
+load(paste(dir, "simulation/part3_expl.Rdata", sep = "/"))
+
+labsw <- lapply(1:length(estimates.3), function(x) {
+  out <- lapply(estimates.3[[x]], function(y) {
+    out <- lapply(y, function(z) {
+
+      kappa <- try(cohen.kappa(z$states)$kappa)
+
+      states <- z$states
+
+      if(x == 2 & is.numeric(kappa) & kappa < 0.95) {
+
+        states$y <- ifelse(z$states$y == 3, 2, ifelse(z$states$y == 2, 3, z$states$y))
+
+        kappa <- cohen.kappa(states)$kappa
+
+        if(is.numeric(kappa) & kappa < 0.95) {
+
+          states$y <- ifelse(z$states$y == 3, 1, ifelse(z$states$y == 1, 3, z$states$y))
+
+          kappa <- cohen.kappa(states)$kappa
+        }
+
+        if(is.numeric(kappa) & kappa < 0.95) {
+
+          states$y <- ifelse(z$states$y == 3, 2, ifelse(z$states$y == 1, 3, 1))
+
+          kappa <- cohen.kappa(states)$kappa
+        }
+      }
+
+      if(x == 3 & is.numeric(kappa) & kappa < 0.5) {
+
+        states$y <- ifelse(z$states$y == 3, 2, ifelse(z$states$y == 2, 3, z$states$y))
+
+        kappa <- cohen.kappa(states)$kappa
+
+        if(is.numeric(kappa) & kappa < 0.5) {
+
+          states$y <- ifelse(z$states$y == 3, 4, ifelse(z$states$y == 4, 3, z$states$y))
+
+          kappa <- cohen.kappa(states)$kappa
+        }
+
+        if(is.numeric(kappa) & kappa < 0.5) {
+
+          states$y <- ifelse(z$states$y == 3, 1, ifelse(z$states$y == 1, 3, z$states$y))
+
+          kappa <- cohen.kappa(states)$kappa
+        }
+
+        if(is.numeric(kappa) & kappa < 0.5) {
+
+          kappa <- cohen.kappa(z$states)$kappa
+        }
+      }
+
+      if(is.numeric(kappa)) {return(kappa)
+        } else {return(NA)}
+    })
+
+    return(as.vector(reduce(out, cbind)))
+  })
+
+  return(t(reduce(out, cbind)))
+})
+
+labsw.df <- as.data.frame(reduce(labsw, rbind))
+
+# Plot exploratory analysis results
+
+plots.acc.expl <- lapply(1, function(y, data) {
+
+  x <- as_tibble(data, .name_repair = "unique")
+
+  data.long <- x %>%
+    mutate(cond = as.factor(rep(1:3, 3)),
+           k = as.factor(rep(2:4, each = 3))) %>%
+    pivot_longer(names(x), names_to = "par.est", values_to = "accuracy")
+
+  p <- ggplot(data = data.long, aes(x = k, y = accuracy, color = cond)) +
+    geom_boxplot(outlier.shape = 4) + 
+    scale_x_discrete(name = "k (number of states)")  + 
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-0.33, 0, 0.25, 0.5, 0.75, 1))
+
+  names.cond <- c("1", "2", "3")
+  label.cond <- bquote(tau["start"])
+
+  p <- p + scale_color_discrete(name = label.cond, labels = names.cond)
+
+  return(p)
+}, data = labsw.df)
+
 ```
@@
 ## Simulation study
-
-With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences.
-The most critical decrease in recovery occured when noise was added to the data generated by models including smooth pursuits.
-In the noise condition, scale parameters of gamma distributions were often badly recovered.
-Higher sample sizes slightly improved parameter recovery, but neither variation in starting values nor missing data affected it.
-In contrast, the accuracy of the model was linearly descreasing with more data missing but not influenced by manipulating parameters, noise, or starting value variation.
-Adding more states to the HMM generally decreased the parameter recovery and the accuracy.
-The full results are described in the following sections.
-
 ### Parameter variation
-
-In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation between the true and estimated parameters and the accuracy between the true and estimated state sequence.
-Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM with two states.
-The RMdSPD was below 0.1 for all estimated and manipulated parameters (except the irrelevant initial state probabilities).
-Thus, no matter which relevant parameter changed in the HMM, the median deviation did not exceed 10% of the true parameter value.
-The regressions between manipulated true and estimated parameters for two states are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
-With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely: Changing the true parameters led to linear changes in the estimated parameters.
-However, the deviation between true and estimated response parameters seemed to increase slighlty with parameter magnitude.
-Considering accuracy, Figure \@ref(fig:plot-acc-1-2) displays Cohen's kappa between true and estimated hidden state sequences for two states.
+#### Two States
+In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation and accuracy between the true and estimated parameters and state sequence.
+Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM.
+The RMdSPD was below 0.1 for all estimated and manipulated parameters, that is, the median deviation never exceeded 10% of the true parameters [^1].
+The regressions between manipulated true and estimated parameters are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
+With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely.
+The deviation seemed to increase slighlty with parameter magnitude.
+Considering accuracy, Figure \@ref(fig:plot-acc-1-2) displays Cohen's kappa between true and estimated hidden state sequences.
 With two exceptions, kappa values were almost one, indicating nearly perfect agreement.
-In sum, the HMM with two states recovered parameters and hidden states very well with outliers only occuring rarely.
+In sum, the HMM with two states recovered parameters and hidden states very well and outliers only occured rarely.
+
+[^1]: Note that the intial state probability $\rho_i$ has also been estimated.
+Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to an RMdSPD of one).
+Therefore, I decided not to include it in the analysis.
 
@@
 
-For three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
+#### Three States
+For the simulation with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
 When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently below 0.5.
@@
 Otherwise, RMdSPDs were consistently lower than 0.1, indicating a median deviation between true and estimated parameters below 10% of the true parameter.
-Inspecting the regressions between manipulated true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed a strong and unbiased linear relationship (intercepts close to zero and slopes close to one).
+Inspecting the regressions between manipulated true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one).
 Again, the deviation seemed to increase with true parameter magnitude (reverse for kappas).
@@
 For most estimated models, the kappas between true and estimated state sequences were above 0.95, indicating almost perfect agreement.
-However, for some models I observed kappas clustered around zero or -0.33, suggesting that state labels switched (CITE).
-To summarize, the three-state HMM displayed a slightly worse parameter recover with more outliers than the two-state model but shows an almost equally good accuracy.
+However, for some models, I observed kappas clustered around zero or -0.33, suggesting that state labels switched.
+To summarize, the three-state HMM had a slightly worse parameter recovery and more outliers than the two-state model but shows a similar good accuracy.
 
@@
 
+#### Four States
 The RMdSPDs for the four-state HMM is illustrated in Figure \@ref(fig:plot-rmdspd-1-4).
@@
 Otherwise, RMdSPDs were below 0.1, indicating a median deviation below 10% of the true parameter.
-Looking at the regressions between true and estimated paramters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) show strong and unbiased relationships but larger deviations and more outliers than in the previous models, especially for states one and four.
-Accuracy indicated by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
-Here, some kappa values clustered around 0.25 and zero, which again can be interpreted as the results of label switching.
-Summarizing, the parameter recovery for the four-state HMM was slightly worse with even more outliers compared to the two- and three-state models.
+Looking at the regressions between true and estimated paramters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) show strong and unbiased relationships.
+However, there were larger deviations and more outliers than in the previous models, especially for states one and four.
+Accuracy measured by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
+Here, some kappa values clustered around 0.25 and zero, which again can be interpreted as the result of label switching.
+Summarizing, the parameter recovery for the four-state HMM was slightly worse with even more outliers compared to the three-state model.
 Especially the recovery of transition parameters in states corresponding to fixations and smooth pursuits decreased.
@@
 ### Sample size and noise variation
-
+#### Two States
 In the second part, I varied the sample size of the HMM and added noise to the generated data.
@@
 (ref:plot-rmdspd-2-2) RMdSPD between true and estimated parameters of the two-state HMM in part two of the simulation.
+Colours indicate different sizes of generated data.
 Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
@@
 
-The RMdSPDs for the three-state HMM and $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)).
+#### Three States
+The RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)).
 Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5.
-For most parameters in all three states, higher sample sizes had lower RMdSPDs.
+For most parameters across all three states, higher sample sizes had lower RMdSPDs.
 The accuracy of the estimated models was almost perfect with most kappa values above 0.95 (see Figure \@ref(fig:plot-acc-2), middle plot).
@@
 
+#### Four States
 RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
@@
 Increasing the samples in the HMM slightly improved the recovery of most parameters.
-The accuracy of the model was slightly lowered when including more states, but it was neither affected by the range of noise $\tau_{noise}$ nor the sample size.
+The accuracy of the model was slightly lowered when including more states, but it was neither affected by the noise variability $\tau_{noise}$ nor the sample size.
 
@@
 ### Variation of starting values
-
 The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM.
-For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure X).
-Cohen's kappa values were slightly lower than 1, indicating almost perfect accuracy (see Figure X).
-For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5.
-As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33.
-Regarding the four state-HMM, RMdSPS of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1.
-The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero.
+For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure \@ref(fig:plot-rmdspd-3-2)).
+Cohen's kappa values were slightly lower than 1, indicating almost perfect accuracy (see Figure \@ref(fig:plot-acc-3)).
+For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5 (see Figure \@ref(fig:plot-rmdspd-3-3)).
+As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33 (see Figure \@ref(fig:plot-acc-3)).
+Regarding the four state-HMM, RMdSPS of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1 (see Figure \@ref(fig:plot-rmdspd-3-4)).
+The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero (see Figure \@ref(fig:plot-acc-3)).
 In conclusion, the parameter recovery of the model worked very well.
@@
 Increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor accuracy.
-Since the manipulation in this part was not effective, I decided to explore post-hoc whether clusters of low Cohen's kappa values were due to label switching.
-According to Visser and Speekenbrink (CITE), label switching occurs when exchanging the order of the states leads to equally likely models.
+
+(ref:plot-rmdspd-3-2) RMdSPD between true and estimated parameters of the two-state HMM in part three of the simulation.
+Colours indicate the variation in starting values used to estimate parameters.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-3-2, fig.cap="(ref:plot-rmdspd-3-2)"}
+
+print(plots.rmsd.234[[2]][[1]])
+
+```
+
+(ref:plot-acc-3) Boxplots displaying Cohen's kappa depending on the number of states in the HMM in part three.
+Colours indicate the variation in starting values used to estimate parameters.
+Solid vertical lines symbolize medians and hinges the first and third quartile.
+Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
+Crosses represent outliers.
+
+```{r plot-acc-3, fig.cap="(ref:plot-acc-3)"}
+
+print(plots.acc.234[[2]])
+
+```
+
+(ref:plot-rmdspd-3-3) RMdSPD between true and estimated parameters of the three-state HMM in part three of the simulation.
+Colours indicate the variation in starting values used to estimate parameters.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-3-3, fig.cap="(ref:plot-rmdspd-3-3)"}
+
+print(plots.rmsd.234[[2]][[2]])
+
+```
+
+(ref:plot-rmdspd-3-4) RMdSPD between true and estimated parameters of the four-state HMM in part three of the simulation.
+Colours indicate the variation in starting values used to estimate parameters.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-3-4, fig.cap="(ref:plot-rmdspd-3-4)"}
+
+print(plots.rmsd.234[[2]][[3]])
+
+```
+
+In this part, I explored post-hoc whether clusters of low Cohen's kappa values were due to label switching.
+According to Visser and Speekenbrink [-@Visser2019], label switching occurs when exchanging the order of the states leads to equally likely models.
 Thus, the model is accurate but has oppositely estimated states compared to the true states.
 In this case, an HMM with three states that is perfectly accurate but has one label swtiched would have a Cohen's kappa of 0.
-To test if label switching occured, I manually switched one or two labels post-hoc (see Figure X).
+To test if label switching occured, I manually switched one or two labels post-hoc (see Figure \@ref(fig:plot-acc-3-expl)).
 It can be seen that this approach resolved low accuracy clusters for three and four states.
 
+(ref:plot-acc-3-expl) Boxplots displaying Cohen's kappa depending on the number of states in the HMM in part three after state labels were switched post-hoc (exploratory analysis).
+Colours indicate the variation in starting values used to estimate parameters.
+Solid vertical lines symbolize medians and hinges the first and third quartile.
+Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
+Crosses represent outliers.
+
+```{r plot-acc-3-expl, fig.cap="(ref:plot-acc-3-expl)"}
+
+print(plots.acc.expl[[1]])
+
+```
+
 ### Missing data
-
-In the last part, intervals of varying length in the data were set to be missing.
-Regarding parameter recovery, RMdSPDs were almost exactly mirroring those in the previous part for two, three, and four states.
-Looking at accuracies, Figure X illustrates an interaction between the length of and amount of missing data intervals.
+In the last part, data intervals of varying length were set to be missing.
+Regarding parameter recovery, RMdSPDs were almost exactly mirroring those in the previous part for two, three, and four states (see Figures \@ref(fig:plot-rmdspd-4-2), \@ref(fig:plot-rmdspd-4-3), and \@ref(fig:plot-rmdspd-4-4), respectively).
+Looking at accuracies, Figure \@ref(fig:plot-acc-4) illustrates an interaction between the length of and amount of missing data intervals.
 Cohen's kappa values linearly decreased with the length of missing data intervals and the decrease became linearly steeper when more intervals were included.
@@
 For the four-state model, kappas started around 0.85 and decreased to 0.5 for five missing intervals, suggesting moderate to substantial accuracy.
-As in previous parts, a few kappa values clustered around zero and -0.33 for the three-state model and around 0.6, 0.25, and zero for the four-state model.
-In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model up to a moderate extent.
+As in previous parts, a several kappa values clustered around zero and -0.33 for the three-state model and around 0.6, 0.25, and zero for the four-state model.
+In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model from an nearly perfect to a moderate extent.
+With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences.
+The most critical decrease in recovery occured when noise was added to the data generated by models including smooth pursuit.
+In the noise condition, scale parameters of gamma distributions were often badly recovered.
+Higher sample sizes slightly improved parameter recovery, but neither variation in starting values nor missing data affected it.
+In contrast, the accuracy of the model was linearly descreasing with more data missing but not influenced by manipulating parameters, noise, or starting value variation.
+Adding more states to the HMM generally decreased the parameter recovery and the accuracy.
+
+(ref:plot-rmdspd-4-2) RMdSPD between true and estimated parameters of the two-state HMM in part four of the simulation.
+Colours indicate the number of missing data intervals.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-4-2, fig.cap="(ref:plot-rmdspd-4-2)"}
+
+print(plots.rmsd.234[[3]][[1]])
+
+```
+
+(ref:plot-rmdspd-4-3) RMdSPD between true and estimated parameters of the three-state HMM in part four of the simulation.
+Colours indicate the number of missing data intervals.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-4-3, fig.cap="(ref:plot-rmdspd-4-3)"}
+
+print(plots.rmsd.234[[3]][[2]])
+
+```
+
+(ref:plot-rmdspd-4-4) RMdSPD between true and estimated parameters of the four-state HMM in part four of the simulation.
+Colours indicate the number of missing data intervals.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+
+```{r plot-rmdspd-4-4, fig.cap="(ref:plot-rmdspd-4-4)"}
+
+print(plots.rmsd.234[[3]][[3]])
+
+```
+
+(ref:plot-acc-4) Cohen's kappa depending on the length of missing data intervals. Colours indicate the number of missing data intervals. Top facet labels indicate the number of states in the HMM.
+
+```{r, plot-acc-4, fig.cap="(ref:plot-acc-4)"}
+
+print(plots.acc.int.24[[2]])
+
+```
 
@@
 
-To test whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states to two benchmark data sets and compared their Schwarz weights.
+```{r load validation results, include=FALSE}
+
+load("~/Uni/Psychologie Master/Internship/GazeHMM/validation/Andersson2017_fitted.Rdata")
+load("~/Uni/Psychologie Master/Internship/GazeHMM/validation/Andersson2017_raw.Rdata")
+
+```
+
+```{r calculate Schwarz weights, include=FALSE}
+
+# Compute Schwarz weights
+
+schwarz.weights <- function(bic, na.rm = T) {
+
+  d.bic <- bic - min(bic, na.rm = na.rm) # eq 2
+
+  exp(-0.5 * d.bic)/sum(exp(-0.5 * d.bic), na.rm = na.rm) # eq 4
+
+}
+
+A2017.bic <- lapply(A2017.fit, function(stim) {
+  out <- lapply(stim, function(subj) {
+    out <- lapply(1:length(subj), function(mod) {
+
+      if(mod == 1) {
+
+        bic <- try(-2*subj[[mod]][["LL"]] + 6*log(subj[[mod]][["N"]]))
+
+      } else {
+
+        bic <- try(BIC(subj[[mod]]$model))
+
+      }
+
+      return(ifelse(is.numeric(bic), bic, NA))
+    })
+
+    return(schwarz.weights(unlist(out)))
+  })
+
+  df <- as.data.frame(reduce(out, rbind))
+
+  names(df) <- paste("model_", 1:length(stim[[1]]), sep = "")
+
+  return(df)
+})
+
+# Create Schwarz weight plots
+
+bic.plot <- lapply(A2017.bic, function(x) {
+
+  data.long <- x %>% 
+    mutate(subject = 1:nrow(x)) %>%
+    pivot_longer(names(x), names_to = "model", values_to = "weight") %>%
+    mutate_at(c("subject", "model"), as.factor)
+
+  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
+    scale_fill_continuous(name = "Schwarz\nweight") +
+    scale_x_discrete(name = "Number of states in model", labels = as.character(1:5))
+
+  return(p)
+})
+
+```
+
+To test whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states on two benchmark data sets and compared their Schwarz weights.
 For the Andersson et al.
@@
 data suggests that the five-state model has most likely generated the data.
-Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the likelihoods of the models.
+Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
 
@@
 
-### Comparison against other algorithms
+### Comparison to other algorithms
+
+```{r calculate duration RMSD, include=FALSE}
+
+# Calculate event descriptives for human coders
+
+fr <- 500
+
+A2017.events <- lapply(A2017, function(stim) {
+  lapply(1:4, function(e) {
+    out <- lapply(stim, function(df) {
+
+      if(all(is.nan(df$t) || df$t == 0)) {
+
+        df$t <- seq(0, length(df$t)-1)/fr
+
+      } else {
+
+        df$t <- (df$t - df$t[1])/1e6
+
+        df <- df[df$t >= 0,]
+
+      }
+
+      counter_MN <- 1
+      counter_RA <- 1
+
+      number_MN <- numeric(nrow(df))
+      number_RA <- numeric(nrow(df))
+
+      number_MN[1] <- 1
+      number_RA[1] <- 1
+
+      for (i in 2:nrow(df)) {
+        if(df$label_MN[i] != df$label_MN[i-1]) counter_MN <- counter_MN + 1
+        if(df$label_RA[i] != df$label_RA[i-1]) counter_RA <- counter_RA + 1
+
+        number_MN[i] <- counter_MN
+        number_RA[i] <- counter_RA
+
+      }
+
+      dur_MN <- numeric(max(number_MN))
+      event_MN <- numeric(max(number_MN))
+
+      for (n in unique(number_MN)) {
+        if(n > 1) {
+          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - max(df$t[number_MN == (n-1)], na.rm = T)
+        } else {
+          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - min(df$t[number_MN == n], na.rm = T)
+        }
+
+        event_MN[n] <- max(df$label_MN[number_MN == n], na.rm = T)
+      }
+
+      dur_RA <- numeric(max(number_RA))
+      event_RA <- numeric(max(number_RA))
+
+      for (n in unique(number_RA)) {
+        if(n > 1) {
+          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - max(df$t[number_RA == (n-1)], na.rm = T)
+        } else {
+          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - min(df$t[number_RA == n], na.rm = T)
+        }
+
+        event_RA[n] <- max(df$label_RA[number_RA == n], na.rm = T)
+      }
+
+      return(list(dur_MN[event_MN == e], dur_RA[event_RA == e]))
+    })
+
+    df <- reduce(out, cbind)
+
+    dur_MN <- unlist(df[1,])
+    dur_RA <- unlist(df[2,])
+
+    return(data.frame(MN = c(mean(dur_MN), sd(dur_MN), length(dur_MN)),
+                      RA = c(mean(dur_RA), sd(dur_RA), length(dur_RA))))
+  })
+})
+
+# Calculate RMSD between event distribution descriptives
+
+ref <- list(dot = list(fix = list(CDT = c(60, 127, 165),
+                                  EM = c(NA, NA, NA),
+                                  IDT = c(323, 146, 8),
+                                  IKF = c(217, 184, 72),
+                                  IMST = c(268, 140, 12),
+                                  IHMM = c(214, 286, 67),
+                                  IVT = c(203, 282, 71),
+                                  NH = c(380, 333, 30),
+                                  BIT = c(189, 113, 67),
+                                  LNS = c(NA, NA, NA)),
+                       sac = list(CDT = c(NA, NA, NA),
+                                  EM = c(17, 14, 93),
+                                  IDT = c(32, 14, 10),
+                                  IKF = c(60, 26, 29),
+                                  IMST = c(13, 5, 18),
+                                  IHMM = c(41, 17, 27),
+                                  IVT = c(36, 14, 28),
+                                  NH = c(43, 16, 42),
+                                  BIT = c(NA, NA, NA),
+                                  LNS = c(26, 11, 53)),
+                       pso = list(NH = c(24, 12, 17),
+                                  LNS = c(20, 9, 31))),
+            img = list(fix = list(CDT = c(397, 559, 251),
+                                  EM = c(NA, NA, NA),
+                                  IDT = c(399, 328, 242),
+                                  IKF = c(174, 239, 513),
+                                  IMST = c(304, 293, 333),
+                                  IHMM = c(133, 216, 701),
+                                  IVT = c(114, 204, 827),
+                                  NH = c(258, 299, 292),
+                                  BIT = c(209, 136, 423),
+                                  LNS = c(NA, NA, NA)),
+                       sac = list(CDT = c(NA, NA, NA),
+                                  EM = c(25, 22, 787),
+                                  IDT = c(25, 15, 258),
+                                  IKF = c(62, 37, 353),
+                                  IMST = c(17, 10, 335),
+                                  IHMM = c(48, 26, 368),
+                                  IVT = c(41, 22, 373),
+                                  NH = c(50, 20, 344),
+                                  BIT = c(NA, NA, NA),
+                                  LNS = c(29, 12, 390)),
+                       pso = list(NH = c(28, 13, 237),
+                                  LNS = c(25, 9, 319))),
+            vid = list(fix = list(CDT = c(213, 297, 211),
+                                  EM = c(NA, NA, NA),
+                                  IDT = c(554, 454, 48),
+                                  IKF = c(258, 296, 169),
+                                  IMST = c(526, 825, 71),
+                                  IHMM = c(234, 319, 194),
+                                  IVT = c(202, 306, 227),
+                                  NH = c(429, 336, 83),
+                                  BIT = c(248, 215, 170),
+                                  LNS = c(NA, NA, NA)),
+                       sac = list(CDT = c(NA, NA, NA),
+                                  EM = c(20, 16, 252),
+                                  IDT = c(24, 53, 41),
+                                  IKF = c(55, 20, 107),
+                                  IMST = c(18, 10, 76),
+                                  IHMM = c(42, 18, 109),
+                                  IVT = c(36, 16, 112),
+                                  NH = c(44, 18, 104),
+                                  BIT = c(NA, NA, NA),
+                                  LNS = c(28, 12, 122)),
+                       pso = list(NH = c(28, 13, 78),
+                                  LNS = c(24, 10, 87))))
+
+A2017.rmsd <- lapply(2:4, function(k) {
+  lapply(1:length(A2017.fit), function(x) { 
+    out <- lapply(1:k, function(y) {
+      out <- lapply(A2017.fit[[x]], function(z) {
+
+        if(class(z[[k]]) == "gazeHMM") z[[k]]$events[[y]]
+
+      })
+
+      df <- try(reduce(out, rbind))
+
+      alg <- try(c(mean(df$dur, na.rm = T), sd(df$dur, na.rm = T), nrow(df)))
+
+      if(y == 4) {
+        M <- try(cbind(A2017.events[[x]][[y]], alg))
+      } else {
+        M <- try(cbind(A2017.events[[x]][[y]], alg, as.data.frame(ref[[x]][[y]])))
+        M[1:2,4:ncol(M)] <- M[1:2,4:ncol(M)]/1e3
+      }
+
+      M.norm <- try(apply(M, 1, function(x) {(x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))}))
+
+      if(y == 4) {
+        rmsd <- try(sum(sqrt((M.norm[3,] - colSums(M.norm[1:2,]/2))^2)))
+      } else {
+        rmsd <- try(apply(M.norm[3:nrow(M.norm),], 1, function(r) {sum(sqrt((r - colSums(M.norm[1:2,]/2))^2))}))
+      }
+
+      coder <- try(sum(sqrt((M.norm[1,] - M.norm[2,])^2)))
+
+      rmsd <- c(coder, coder, rmsd)
+
+      return(rbind(M, rmsd))
+    })
+  })
+})
+
+# Create tables for RMSDs
+
+rmsd.table <- lapply(1:length(A2017.rmsd), function(x) {
+  out <- lapply(1:length(A2017.rmsd[[x]]), function(y) {
+    out <- lapply(1:length(A2017.rmsd[[x]][[y]]), function(z) {
+
+      tabmat <- t(apply(A2017.rmsd[[x]][[y]][[z]], 2, function(d) {as.character(round(d, 3))}))
+
+      # tabmat[3,] <- paste("**", tabmat[3,], "**", sep = "")
+
+      colnames(tabmat) <- c("Mean", "SD", "Events", "RMSD")
+      rownames(tabmat)[1:3] <- c("coderMN", "coderRA", "gazeHMM")
+
+      return(cbind(z, rownames(tabmat), tabmat))
+    })
+
+    cbind(y, reduce(out, rbind))
+  })
+
+  df <- as.data.frame(reduce(out, rbind), stringsAsFactors = F) %>%
+    mutate(Condition = factor(y, labels = c("moving dots", "image", "video")))
+
+  lapply(1:(x+1), function(x) {
+
+    df <- df %>% dplyr::filter(z == x) %>%
+      dplyr::select(-c("z", "y")) %>%
+      rename(Algorithm = V3) %>%
+      pivot_wider(names_from = "Condition", names_glue = "{Condition}_{.value}", names_sort = T, values_from = c("Mean", "SD", "Events", "RMSD"))
+
+    df %>% dplyr::select(c("Algorithm", "image_Mean", "image_SD", "image_Events", "image_RMSD",
+                           "moving dots_Mean", "moving dots_SD", "moving dots_Events", "moving dots_RMSD",
+                           "video_Mean", "video_SD", "video_Events", "video_RMSD"))
+  })
+})
+
+```
+
+```{r calculate Cohen s kappa and confusion matrices, include= FALSE}
+
+# Compute Cohen's kappa and confusion matrices
+
+ref.acc <- matrix(c(0.92, 0.81, 0.83, 0.95, 0.91, 0.94, 0.88, 0.82, 0.83, 0.00, 0.00, 0.00,
+                    0.92, 0.84, 0.82, 0.95, 0.91, 0.94, 0.88, 0.80, 0.81, 0.00, 0.00, 0.00,
+                    0.38, 0.06, 0.11, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.00, 0.00, 0.00, 0.64, 0.66, 0.67, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.36, 0.00, 0.03, 0.45, 0.26, 0.38, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.63, 0.03, 0.14, 0.58, 0.46, 0.59, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.38, 0.00, 0.03, 0.54, 0.30, 0.52, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.67, 0.03, 0.13, 0.69, 0.60, 0.71, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.67, 0.03, 0.13, 0.75, 0.63, 0.76, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.52, 0.00, 0.01, 0.67, 0.60, 0.68, 0.24, 0.20, 0.25, 0.00, 0.00, 0.00,
+                    0.67, 0.03, 0.14, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
+                    0.00, 0.00, 0.00, 0.81, 0.75, 0.81, 0.64, 0.59, 0.63, 0.00, 0.00, 0.00), ncol = 12, byrow = T)
+
+A2017.acc <- lapply(2:4, function(k) {
+  lapply(1:length(A2017.fit), function(x) { 
+    out <- lapply(1:length(A2017.fit[[x]]), function(y) {
+
+      if(class(A2017.fit[[x]][[y]][[as.character(k)]]) == "gazeHMM") {
+        cod.MN <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_MN))
+        cod.RA <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_RA))
+
+        return(rbind(cod.MN, cod.RA))
+      }
+    })
+
+    df <- reduce(out, rbind)
+
+    df[df[,1] == 0,1] <- 5
+
+    events <- lapply(1:k, function(e) {
+
+      alg <- df[,1] == e
+      cod <- df[,2] == e
+
+      mat <- matrix(c(sum(alg & cod, na.rm = T), sum(alg & !cod, na.rm = T), 
+                      sum(!alg & cod, na.rm = T), sum(!alg & !cod, na.rm = T)), 2, 2)
+
+      over <- mat[2,1]/sum(mat[,1])
+      under <- mat[1,2]/sum(mat[,2])
+
+      return(list(kappa = cohen.kappa(mat)$kappa, 
+                  over = over,
+                  under = under))
+    })
+
+    ratio <- mean(df[,1] != df[,2], na.rm = T)
+
+    conmat <- table(df[,1], df[,2])
+
+    return(list(events = events, ratio =  ratio, conmat = conmat))
+  })
+})
+
+# Create tables for Cohen's kappa and confusion matrices
+
+kappa.table <- lapply(A2017.acc, function(x) {
+  out <- lapply(x, function(y){
+    out <- lapply(y$events, function(z){
+      return(round(z$kappa, 3))
+    })
+
+    return(reduce(out, cbind))
+  })
+
+  tabmat <- reduce(out, rbind)
+
+  tabmat <- rbind(tabmat[2,], tabmat[1,], tabmat[3,])
+
+  return(as.vector(tabmat))
+})
+
+conf.table <- lapply(A2017.acc, function(x) {
+  out <- lapply(x, function(y){
+
+    tabmat <- apply(y$conmat, 2, as.character)
+
+    # diag(tabmat)[1:(nrow(tabmat)-1)] <- paste("**", diag(tabmat)[1:(nrow(tabmat)-1)], "**", sep = "")
+
+    # tabmat[nrow(tabmat), 5] <- paste("**", tabmat[nrow(tabmat), 5], "**", sep = "")
+
+    colnames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks", "Other")
+    rownames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks", "Other")[1:nrow(tabmat)]
+    rownames(tabmat)[nrow(tabmat)] <- "Blinks"
+
+    return(as.data.frame(tabmat))
+  })
+})
+
+# Create figure for disagreement ratio
+
+ref.disag <- list(c(7, 7, A2017.acc[[3]][[2]]$ratio*100, A2017.acc[[2]][[2]]$ratio*100, 23, 92, 20, 24, 20, 20, 19, 32, 31, 84),
+                  c(11, 11, A2017.acc[[3]][[1]]$ratio*100, A2017.acc[[2]][[1]]$ratio*100, 89, 96, 86, 85, 86, 84, 84, 93, 89, 93),
+                  c(19, 19, A2017.acc[[3]][[3]]$ratio*100, A2017.acc[[2]][[3]]$ratio*100, 64, 95, 61, 62, 61, 59, 59, 70, 67, 92))
+
+disag.df <- data.frame(Image = ref.disag[[1]], Dots = ref.disag[[2]], Video = ref.disag[[3]],
+                       Algorithm = c("coderMN", "coderRA", "gazeHMM-4", "gazeHMM-3", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
+                       stringsAsFactors = F)
+
+disag.df$Algorithm <- factor(disag.df$Algorithm, levels = unique(disag.df$Algorithm))
+
+disag.df <- disag.df %>% pivot_longer(c("Image", "Dots", "Video"), names_to = "Condition", values_to = "Ratio")
+
+disag.plot <- ggplot(disag.df, aes(x = Algorithm, y = Ratio, fill = Condition)) + 
+  geom_col(position = "dodge") + 
+  scale_y_continuous(name = "Disagreement ratio (in %)", limits = c(0, 100), breaks = c(0, 0.25, 0.5, 0.75, 1)*100) +
+  scale_x_discrete(name = "") + coord_flip()
+
+```
 
@@
 for all algorithms included in the study plus gazeHMM.
-Table X shows that in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations due to a large amount of very short fixations.
-For saccades, gazeHMM has the fourth lowest RMSD of the algorithms in the image and video conditions but the second highest in the moving dots condition (see Table X).
+Table \@ref(tab:tab-rmsd-4-fix) shows that in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations due to a large amount of very short fixations.
+For saccades, gazeHMM has the fourth lowest RMSD of the algorithms in the image and video conditions but the second highest in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)).
 Here, gazeHMM overestimated the number and duration of saccades (either slightly or substantially).
-Moreover, gazeHMM underestimated the number and duration of PSOs in the image and moving dots conditions, leading to the highest and intermediate RMSD among the competitors, repsectively (see Table X).
+Moreover, gazeHMM underestimated the number and duration of PSOs in the image and moving dots conditions, leading to the highest and intermediate RMSD among the competitors, repsectively (see Table \@ref(tab:tab-rmsd-4-pso)).
 In the video condition the duration was slightly overestimated but the duration was underestimated with the highest RMSD.
-No other algorithm was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than among the human coders (see Figure X).
+No other algorithm was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than among the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)).
 In sum, except for saccades and PSOs, gazeHMM performed worse with regard to the duration of classified events compared to the other algorithms.
 
+(ref:tab-rmsd-4-fix) Mean and standard deviations of fixation durations, number of fixations, and RMSD between fixation durations classified by algorithms and human coders
+(ref:tab-rmsd-4-fix-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+
+```{r tab-rmsd-4-fix}
+
+apa_table(rmsd.table[[3]][[1]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          caption = "(ref:tab-rmsd-4-fix)",
+          note = "(ref:tab-rmsd-4-fix-note)")
+
+```
+
+(ref:tab-rmsd-4-sac) Mean and standard deviations of saccade durations, number of saccades, and RMSD between saccade durations classified by algorithms and human coders
+(ref:tab-rmsd-4-sac-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+
+```{r tab-rmsd-4-sac}
+
+apa_table(rmsd.table[[3]][[1]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          caption = "(ref:tab-rmsd-4-sac)",
+          note = "(ref:tab-rmsd-4-sac-note)")
+
+```
+
+(ref:tab-rmsd-4-pso) Mean and standard deviations of PSO durations, number of PSOs, and RMSD between PSO durations classified by algorithms and human coders
+(ref:tab-rmsd-4-pso-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+
+```{r tab-rmsd-4-pso}
+
+apa_table(rmsd.table[[3]][[1]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          caption = "(ref:tab-rmsd-4-pso)",
+          note = "(ref:tab-rmsd-4-pso-note)")
+
+```
+
+(ref:tab-rmsd-4-sp) Mean and standard deviations of smooth pursuit durations, number of smooth pursuits, and RMSD between smooth pursuit durations classified by algorithms and human coders
+(ref:tab-rmsd-4-sp-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+
+```{r tab-rmsd-4-sp}
+
+apa_table(rmsd.table[[3]][[1]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          caption = "(ref:tab-rmsd-4-sp)",
+          note = "(ref:tab-rmsd-4-sp-note)")
+
+```
+
 [^2]: In my proposal, I declared to compute the RMSE instead of RMSD.
-However, since I conducted an in sample comparison the term deviation is more conventional.
+However, since I conducted an in sample comparison, the term deviation is more appropriate.
 Andersson et al.
@@
 
-Second, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of agreement between algorithms and human coders (see Table X).
+Second, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
 For fixations, absolute kappa values in all three conditions indicated a slight to fair agreement between gazeHMM and human coders.
@@
 
-Confusion rate and matrix
+(ref:tab-kappa-4) Cohen's kappa between human coders and algorithms for different conditions and events
+(ref:tab-kappa-4-note) Negative values were set to zero. gazeHMM classified four events. Table adapted from @Andersson2017.
+
+```{r tab-kappa-4}
+
+tab.kappa.4 <- cbind(c("coderMN", "coderRA", "gazeHMM", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
+                     as.data.frame(rbind(ref.acc[1:2,], kappa.table[[3]], ref.acc[3:nrow(ref.acc),])))
+
+apa_table(tab.kappa.4,
+          col.names = c("Algorithm", rep(c("Image", "Dots", "Video"), 4)),
+          col_spanners = list("Fixations" = c(2, 4), "Saccades" = c(5, 7), "PSOs" = c(8, 10), "Smooth pursuit" = c(11, 13)),
+          font_size = "tiny",
+          caption = "(ref:tab-kappa-4)",
+          note = "(ref:tab-kappa-4-note)")
+
+```
+
+Third, I computed the disagreement ratio as the percentage of samples that were classified differently by gazeHMM compared to the human coders.
+Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events has a higher disagreement ratio than all but two of the other algorithms.
+In the moving dots and video conditions, the disagreement ratios are considerably lower than for all the other algorithms.
+Notably, gazeHMM's disagreement ratios were similar across conditions (around 45%).
+
+(ref:plot.disag) Disagreement ratio between algorithms and human coders for different conditions. gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.
+
+```{r plot-disag, fig.cap="(ref:plot.disag)"}
+
+print(disag.plot)
+
+```
+
+(ref:tab-conf-4) Confusion matrix between gazeHMM (rows) and human coders (columns) for different events and conditions
+(ref:tab-conf-4-note) gazeHMM classified four events and blinks.
+
+```{r tab-conf-4}
+
+tab.conf.4 <- cbind(Event = rep(c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks"), 3), reduce(conf.table[[3]], rbind)) %>%
+  mutate(Condition = rep(c("Moving dots", "Image", "Video"), each = 5)) %>%
+  arrange(Condition) %>%
+  dplyr::select(-Condition)
+
+apa_table(tab.conf.4,
+          stub_indents = list("Image" = c(1:5), "Moving dots" = c(6:10), "Video" = c(11:15)),
+          midrules = c(6, 12),
+          caption = "(ref:tab-conf-4)",
+          note = "(ref:tab-conf-4-note)")
+
+```
 
@@
 The same result was found for PSOs.
-The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table X (only for the image condition).
+The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kapp-3) (only for the image condition).
 For fixations, the absolute kappa indicated substantial agreement and was high compared to the other algorithms.
@@
 For PSOs, the absolute agreement was fair, but intermediate with regard to the other algorithms.
+Figure \@ref(fig:plot-disag) illustrates the disagreement ratio between gazeHMM with three events and the human coders compared to the other algorithms.
+In the image condition, the disagreement ratio was slightly lower for gazeHMM than for all the other algorithms.
 Summing up, using gazeHMM with three states increased the agreement to human coding with regard to fixation durations and samples classified as fixations.
@@
 
+(ref:tab-kappa-3) Cohen's kappa between human coders and algorithms for different events in the image condition
+(ref:tab-kappa-3-note) Negative values were set to zero. gazeHMM-3 classified three and gazeHMM-4 classified four events. Table adapted from @Andersson2017.
+
+```{r tab-kappa-3}
+
+tab.kappa.3 <- cbind(c("coderMN", "coderRA", "gazeHMM-4", "gazeHMM-3", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
+                     as.data.frame(rbind(ref.acc[1:2, 1:9], kappa.table[[3]][1:9], kappa.table[[2]], ref.acc[3:nrow(ref.acc), 1:9]))) %>%
+  dplyr::select(c(1, seq(2, 10, 3)))
+
+apa_table(tab.kappa.3,
+          col.names = c("Algorithm", "Fixations", "Saccades", "PSOs"),
+          caption = "(ref:tab-kappa-3)",
+          note = "(ref:tab-kappa-3-note)")
+
+```
+
 # Technical Appendix

### 2020-06-12 manuscript/internship_report.Rmd
update report

@@
 
-floatsintext      : yes
-figurelist        : no
-tablelist         : no
+floatsintext      : no
+figurelist        : yes
+tablelist         : yes
 footnotelist      : no
@@
 Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best.
-However, many metrics rely on thresholds or window ranges that have to be set by the user (e.g., the distance between the mean position in a 100 ms window before and after each sample, see Olsson, 2007, and Zemblys et al.).
+However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007, and @Zemblys2018].
 This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics.
@@
 (ref:filter-angle) Example data from @Andersson2017 displayed as sample-to-sample angle (in rad) over time (in s).
-Line color indicates whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue).
+Line colors indicate whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue).
 For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3.
-Full line segment marks where potential oscillations are filtered out.
-Dashed line segment illustrates where potentially random change in angle is filtered out.
+The full line segment marks where potential oscillations are filtered out.
+The dashed line segment illustrates where potentially random change in angle is filtered out.
 
@@
 Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods.
-They follow a similar trend, but the SG velocity signal is less noisy and acceleration signal is consequently lower than the Laplace acceleration signal.
+They follow a similar trend, but the SG velocity signal is less noisy and the acceleration signal is consequently lower than the Laplace acceleration signal.
 Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
-To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the backward difference (see Technical Appendix).
-
-(ref:filter-comparison) Example data from Andersson et al.
-(2017) displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s).
+To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the forward difference (see Technical Appendix).
+
+(ref:filter-comparison) Example data from @Andersson2017 displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s).
 Line colors indicate which filter has been used to derive the signal.
@@
 In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose.
-It has to be noted that the gamma was choosen because of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
+It has to be noted that the gamma was choosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
 To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution.
@@
 Moreover, the uniform distribution can distinguish fixations from the other events.
-The initial state model and the transition model used multinomial distributions (with a category for each state). For all three submodels of the HMM, only parameter intercepts were estimated.
+The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, only parameter intercepts were estimated.
 
@@
 HMMs can estimate parameters and hidden states despite missing data.
-They either treat data as missing at random or as missing dependent on states [@Visser2019].
+They either treat data as missing at random or as missing depending on states [@Visser2019].
 In gazeHMM, I assume data to be missing at random.
@@
           caption = "(ref:sim-ranges)",
-          note = "(ref:sim-ranges-note)",
-          escape = F)
-
-```
-
-```{r, sim-parameters}
-
-sim.pars.df <- data.frame(State = c(rep("1-4", 3), as.character(rep(1:4, each = 6))),
-                          #Parameter = c(""),
-                          Interval = c("-", "[.01,.99]", "1-a(i=j)/(k-1)", 
-                                       "[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
+          note = "(ref:sim-ranges-note)")
+
+```
+
+(ref:sim-parameters-trans) Intervals and default parameter values for the response models of the HMM in the simulation
+(ref:sim-parameters-trans-note) The transition probability for staying in the same state is denoted by $a_{i=j}$ and the probability for switching to a different state by $a_{i\neq j}$.
+The number of states in the model is denoted by *k*.
+
+```{r, sim-parameters-trans}
+
+sim.pars.trans <- as.data.frame(matrix(c("Interval", "-", "[.01,.99]", "1-a(i=j)/(k-1)",
+                                           "Default", "1/k", "0.9", "0.1/(k-1)"), nrow = 2, byrow = T))
+
+apa_table(sim.pars.trans,
+          col.names = c("", "Initial\nstate\nprob.", "Trans. prob.\nsame state", "Trans. prob.\nother state"),
+          caption = "(ref:sim-parameters-trans)",
+          note = "(ref:sim-parameters-trans-note)")
+
+```
+
+(ref:sim-parameters-resp) Intervals and default parameter values for the response models of the HMM in the simulation
+(ref:sim-parameters-resp-note) Shape parameters are denoted by $\alpha$, scale parameters by $\beta$, mean parameters by $\mu$, and concentration parameters by $\kappa$.
+The default values for the uniform distribution in state one were min = 0 and max = $2\pi$.
+
+```{r, sim-parameters-resp}
+
+sim.pars.resp <- as.data.frame(t(matrix(c("[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
+                                  "3", "0.35", "3", "0.25", "-", "-",
                                   "[1,5]", "[5,15]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
+                                  "3", "10", "3", "3", "0", "1",
                                   "[1,5]", "[0.5,1.5]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
-                                       "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]"),
-                          Default = c("1/k", "0.9", "0.1/(k-1)", 
-                                      "3", "0.35", "3", "0.25", "0", paste(expression(pi), sep = ""),
-                                      "3", "10", "3", "3", "0", "1",
                                   "3", "1", "3", "3", paste(expression(pi), sep = ""), "1",
-                                      "3", "1", "3", "0.15", "0", "1"),
-                          Description = c("Initial state probability for starting in state *i*",
-                                          "Transition probability for staying in the previous state *i*",
-                                          "Transition probability for switching to from state *i* to a different state *j*",
-                                          "Shape parameter of the velocity gamma distribution",
-                                          "Scale parameter of the velocity gamma distribution",
-                                          "Shape parameter of the acceleration gamma distribution",
-                                          "Scale parameter of the acceleration gamma distribution",
-                                          "Minimum of the uniform distribution",
-                                          "Maximum of the uniform distribution",
-                                          "", "", "", "",
-                                          "Mean parameter of the von-Mises distribution",
-                                          "Concentration parameter of the von Mises distribution",
-                                          rep("", 12)))
-
-sim.pars.names <- c(expression(rho[i]), expression(a[i==j]), expression(a[i!=j]),
-                   expression(alpha[vel]))
-
-rownames(sim.pars.df)[1:4] <- sim.pars.names
-
-apa_table(sim.pars.df, row.names = T)
+                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]",
+                                  "3", "1", "3", "0.15", "0", "1"), nrow = 6, byrow = F)))
+
+apa_table(cbind(rep(c("Interval", "Default"), 4), sim.pars.resp),
+          col.names = c("", "Shape", "Scale", "Shape", "Scale", "Mean", "Concentration"),
+          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
+          stub_indents = list("State 1" = c(1, 2), "State 2" = c(3, 4), "State 3" = c(5, 6), "State 4" = c(7, 8)),
+          caption = "(ref:sim-parameters-resp)",
+          note = "(ref:sim-parameters-resp-note)")
 
@@
 
-```{r sim results and helper functions, include=FALSE}
+```{r load sim results and helper functions, include=FALSE}
 
@@
 ## Validation
+### Starting values and model fitting
+I applied gazeHMM on two benchmark data sets to test my hypotheses.
+For both data sets, I used the same starting values (see Table X) to estimate the parameters of the HMM.
+To improve the fitting of the gamma distributions, velocity and acceleration signals were scale down by factor 100 (and so were the starting values for their gamma distributions).
+The algorithm was applied separately for every participant and every condition.
+For the data set by Andersson et al.
+[-@Andersson2017], the algorithm was successfully fitted for every participant and condition.
+
+(ref:tab-starting-values) Starting values for fitting gazeHMM to benchmark data sets
+(ref:tab-starting-values-note) Starting values for velocity and acceleration signals are shown before scaling down by factor 100.
+Values for event 5 were chosen so that they approximately match plausible distributions for microsaccades.
+
+```{r tab-starting-values}
+
+tab.start <- as.data.frame(matrix(c(10, 10, 10, 10, 0, 2*pi,
+                      50, 50, 50, 50, 0, 10,
+                      50, 50, 50, 50, pi, 10,
+                      20, 20, 20, 20, 0, 10,
+                      20, 20, 50, 50, 0, 10), nrow = 5, byrow = T))
+
+colnames(tab.start) <- c("Shape", "Scale", "Shape", "Scale", "Mean", "Concentration")
+rownames(tab.start) <- c("Fixation", "Saccade", "PSO", "Pursuit", "Event 5")
+
+apa_table(as.data.frame(tab.start),
+          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
+          caption = "(ref:tab-starting-values)",
+          note = "(ref:tab-starting-values-note)")
+
+```
+
 ### Model comparison
@@
 
-To test whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states on two benchmark data sets and compared their Schwarz weights.
+To examine whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states on two benchmark data sets and compared their Schwarz weights.
 For the Andersson et al.
-(2017) data set, I expected the highest weights for the three-state models in the image condition.
+[-@Andersson2017] data set, I expected the highest weights for the three-state models in the image condition.
 For the moving dots and video condition, I predicted the highest weights for the four-state models.
@@
 Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
-
-Ehinger data set
+The model comparison for the Ehinger et al.
+[-@Ehinger2019] data set yielded similar results (see Appendix).
 
@@
 (2017) data set.
-First, I calculated the RMSD [^2] as described in the original article by Andersson et al.
+First, I calculated the RMSD [^1] as described in the original article by Andersson et al.
 for all algorithms included in the study plus gazeHMM.
@@
 
-[^2]: In my proposal, I declared to compute the RMSE instead of RMSD.
+[^1]: In my proposal, I declared to compute the RMSE instead of RMSD.
 However, since I conducted an in sample comparison, the term deviation is more appropriate.
@@
 
-(ref:plot.disag) Disagreement ratio between algorithms and human coders for different conditions. gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.
-
-```{r plot-disag, fig.cap="(ref:plot.disag)"}
+(ref:plot-disag) Disagreement ratio between algorithms and human coders for different conditions. gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.
+
+```{r plot-disag, fig.cap="(ref:plot-disag)"}
 
@@
 Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data might be more appropriate and yield better validation results.
-Table X shows the RMSD between gazeHMM and human coders only for the image condition.
+Table \@ref(tab:tab-rmsd-3) shows the RMSD between gazeHMM and human coders only for the image condition.
 For fixations, gazeHMM shows a comparably low RMSD and only two other algorithms have lower values.
@@
 The same result was found for PSOs.
-The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kapp-3) (only for the image condition).
+The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kappa-3) (only for the image condition).
 For fixations, the absolute kappa indicated substantial agreement and was high compared to the other algorithms.
@@
 
+(ref:tab-rmsd-3) Mean and standard deviation of event durations, number of events, and RMSD between event durations classified by algorithms and human coders for the image condition
+(ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+
+```{r tab-rmsd-3}
+
+tab.rmsd.3 <- lapply(A2017.rmsd[[2]][[2]], function(x) {
+
+  df <- as.data.frame(t(as.matrix(x)))
+
+  rownames(df)[3] <- "gazeHMM"
+
+  out <- cbind(algorithm = rownames(df), df)
+
+  return(out)
+}) %>% reduce(left_join, by = "algorithm")
+
+apa_table(tab.rmsd.3,
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Fixations" = c(2, 5), "Saccades" = c(6, 9), "PSOs" = c(10, 13)),
+          font_size = "tiny",
+          caption = "(ref:tab-rmsd-3)",
+          note = "(ref:tab-rmsd-3-note)")
+
+```
+
 (ref:tab-kappa-3) Cohen's kappa between human coders and algorithms for different events in the image condition

### 2020-06-14 manuscript/internship_report.Rmd
update report

@@
 wordcount         : "X"
+
+bibliography      : "report.bib"
 
@@
 library(psych)
-source(paste(dir, "algorithm/preprocessing_helper_functions.R", sep = "/"))
+library(here)
+source(here("algorithm/preprocessing_helper_functions.R"))
+source(here("algorithm/model_helper_functions.R"))
 
@@
 
-load(paste(dir, "validation/Andersson2017_raw.Rdata", sep = "/"))
+load(here("validation/Andersson2017_raw.Rdata"))
 
@@
 
-## Eye movement metrics
+## Eye Movement Metrics
 Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018].
@@
 
-## Fitering and smoothing
+## Fitering and Smoothing
 As with metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics.
@@
 @Houpt2018 compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm.
-Similarly, @Holmqvist2010 use a Savitzky-Golay (SG) filter @Savitzky1964 to estimate velocity and acceleration from a gaze signal.
+Similarly, @Nystrom2010 use a Savitzky-Golay (SG) filter @Savitzky1964 to estimate velocity and acceleration from a gaze signal.
 Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods.
@@
 
-## The generative model
+## The Generative Model
 The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM).
@@
 
-## Missing data and blinks
+## Missing Data and Blinks
 HMMs can estimate parameters and hidden states despite missing data.
@@
 
-## Optimization and classification
+## Optimization and Classification
 The paramters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997].
@@
 ## Implementation
-The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal (CITE) to compute velocity and acceleration signals, dplyr (CITE) to calculate sample-to-sample angles, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
+The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
 The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
@@
 I defined approximate ranges for each response variable and chose true parameter intervals and default values so that they produced samples that roughly corresponded to these ranges.
-Table \@ref(tab:sim-ranges) shows the assumed ranges for each event and Table X shows the intervals and default values for each parameter in the simulation.
+Table \@ref(tab:sim-ranges) shows the assumed ranges for each event and Tables \@ref(tab:sim-parameters-trans) and \@ref(tab:sim-parameters-resp) show the intervals and default values for each parameter in the simulation.
 Parameters were scaled down by factor 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
@@
 
-### Missing data
+### Missing Data
 In the last part, I set intervals of the generated data to be missing.
@@
 
-## Starting values
+## Starting Values
 The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
@@
 
-## Data analysis
+## Data Analysis
 For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values[^1]: $$RMdSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
@@
 By taking the median, I reduced the influence of potential outliers in the estimation and using proportions enabled me to compare RMdSPD values across parameters and data sets.
-Additionally, I applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in Part 1.
+Additionally, I applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in part one.
 Regression slopes closer to one indicated that the model better captured parameter change.
@@
 I adopted the ranges proposed by @Landis1977 to interpret kappa values.
-Models that can not be fitted were excluded from the recovery.
+Models that could not be fitted were excluded from the recovery.
 
@@
 
-Table X
-
-*HMM parameter values used to generate data*
-
-|State|Parameter|Interval|Default|Description|
-|-----|---------|--------|-----|-----------|
-|1-4|$\rho_i^*$|-|$1/k$|Initial state probability for starting in state *i*|
-|1-4|$a_{i=j}$|[.01,.99]|0.9|Transition probability for staying in the previous state *i*|
-|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|Transition probability for switching to from state *i* to a different state *j*|
-|1|$\alpha_{vel}$|[1,5]|3|Shape parameter of the velocity gamma distribution|
-|1|$\beta_{vel}$|[0.1,0.6]|0.35|Scale parameter of the velocity gamma distribution|
-|1|$\alpha_{acc}$|[1,5]|3|Shape parameter of the acceleration gamma distribution|
-|1|$\beta_{acc}$|[0.05,0.25]|0.15|Scale parameter of the acceleration gamma distribution|
-|1|$min^*$|-|0|Minimum of the uniform distribution|
-|1|$max^*$|-|$2\pi$|Maximum of the uniform distribution|
-|2|$\alpha_{vel}$|[1,5]|3||
-|2|$\beta_{vel}$|[5,15]|10||
-|2|$\alpha_{acc}$|[1,5]|3||
-|2|$\beta_{acc}$|[1,5]|3||
-|2|$\mu^*$|-|0|Mean parameter of the von-Mises distribution|
-|2|$\kappa$|$1/[0.1,10]$|1|Concentration parameter of the von Mises distribution
-|3|$\alpha_{vel}$|[1,5]|3||
-|3|$\beta_{vel}$|[0.5,1.5]|1||
-|3|$\alpha_{acc}$|[1,5]|3||
-|3|$\beta_{acc}$|[1,5]|3||
-|3|$\mu^*$|-|$\pi$||
-|3|$\kappa$|$1/[0.1,10]$|1||
-|4|$\alpha_{vel}$|[1,5]|3||
-|4|$\beta_{vel}$|[0.5,1.5]|1||
-|4|$\alpha_{acc}$|[1,5]|3||
-|4|$\beta_{acc}$|[0.05,0.25]|0.15||
-|4|$\mu^*$|-|0||
-|4|$\kappa$|$1/[0.1,10]$|1||
-
-*Note*. Parameters marked with $^*$ will not be varied but always set to their default values. *k* is the number of states in the model.
-
 ```{r load sim results and helper functions, include=FALSE}
@@
 
-  load(file = paste("~/Uni/Psychologie Master/Internship/GazeHMM/simulation/part", part, ".Rdata", sep = ""))
+  load(file = paste(here("simulation/part"), part, ".Rdata", sep = ""))
 
@@
     scale_x_discrete(name = "Varied parameter", labels = names.pars.varied) + 
-    scale_y_continuous(name = "Cohen's kappa", breaks = c(-0.33, 0, 0.25, 0.5, 0.75, 1))
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1))
 
@@
     scale_x_discrete(name = "k (number of states)")  + 
-    scale_y_continuous(name = "Cohen's kappa", breaks = c(-0.33, 0, 0.25, 0.5, 0.75, 1))
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1))
 
@@
     scale_x_continuous(name = x.name) + 
-    scale_y_continuous(name = "Cohen's kappa", breaks = c(-0.33, 0, 0.25, 0.5, 0.75, 1))
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1))
 
@@
 
-# Results
-## Simulation study
-### Parameter variation
+# Exploratory Results
+## Simulation Study
+### Parameter Variation
 #### Two States
@@
 Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM.
-The RMdSPD was below 0.1 for all estimated and manipulated parameters, that is, the median deviation never exceeded 10% of the true parameters [^1].
+The RMdSPD was below 0.1 for all estimated and manipulated parameters, that is, the median deviation never exceeded 10% of the true parameters [^2].
 The regressions between manipulated true and estimated parameters are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
@@
 
-[^1]: Note that the intial state probability $\rho_i$ has also been estimated.
+[^2]: Note that the intial state probability $\rho_i$ has also been estimated.
 Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to an RMdSPD of one).
@@
 
-print(plots.lm.resp[[1]])
+print(plots.lm.resp[[1]] + theme(axis.text = element_text(size = 10)))
 
@@
 
-print(plots.rmsd.1[[2]])
+print(plots.rmsd.1[[2]]+ theme(axis.text = element_text(size = 8)))
 
@@
 
-```{r plot-regresp-1-3, fig.cap="(ref:plot-regresp-1-3)"}
-
-print(plots.lm.resp[[2]])
+```{r plot-regresp-1-3, fig.cap="(ref:plot-regresp-1-3)", fig.height=7}
+
+print(plots.lm.resp[[2]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-### Sample size and noise variation
+### Sample Size and Noise Variation
 #### Two States
@@
 
-print(plots.rmsd.234[[1]][[1]])
+print(plots.rmsd.234[[1]][[1]] + theme(axis.text = element_text(size = 10)))
 
@@
 
-print(plots.rmsd.234[[1]][[2]])
+print(plots.rmsd.234[[1]][[2]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-print(plots.rmsd.234[[1]][[3]])
-
-```
-
-### Variation of starting values
+print(plots.rmsd.234[[1]][[3]] + theme(axis.text = element_text(size = 6)))
+
+```
+
+### Variation of Starting Values
 The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM.
@@
 
-print(plots.rmsd.234[[2]][[1]])
+print(plots.rmsd.234[[2]][[1]] + theme(axis.text = element_text(size = 10)))
 
@@
 
-print(plots.acc.234[[2]])
+print(plots.acc.234[[2]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-print(plots.rmsd.234[[2]][[2]])
+print(plots.rmsd.234[[2]][[2]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-print(plots.rmsd.234[[2]][[3]])
+print(plots.rmsd.234[[2]][[3]] + theme(axis.text = element_text(size = 6)))
 
@@
 
-print(plots.acc.expl[[1]])
-
-```
-
-### Missing data
+print(plots.acc.expl[[1]] + theme(axis.text = element_text(size = 8)))
+
+```
+
+### Missing Data
 In the last part, data intervals of varying length were set to be missing.
@@
 
-print(plots.rmsd.234[[3]][[1]])
+print(plots.rmsd.234[[3]][[1]] + theme(axis.text = element_text(size = 10)))
 
@@
 
-print(plots.rmsd.234[[3]][[2]])
+print(plots.rmsd.234[[3]][[2]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-print(plots.rmsd.234[[3]][[3]])
+print(plots.rmsd.234[[3]][[3]] + theme(axis.text = element_text(size = 6)))
 
@@
 ## Validation
-### Starting values and model fitting
+### Starting Values and Model Fitting
 I applied gazeHMM on two benchmark data sets to test my hypotheses.
-For both data sets, I used the same starting values (see Table X) to estimate the parameters of the HMM.
-To improve the fitting of the gamma distributions, velocity and acceleration signals were scale down by factor 100 (and so were the starting values for their gamma distributions).
+For both data sets, I used the same set of starting values (initial state model: $\rho=1/k$; transition model: $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$; response model: see Table \@ref(tab:tab-starting-values)) to estimate the parameters of the HMM.
+In contrast to the simulation study, generating random starting value often led to bad model fits and label switching between states.
+To improve the fitting of the gamma distributions, velocity and acceleration signals were scaled down by factor 100 (so were the starting values for their gamma distributions).
 The algorithm was applied separately for every participant and every condition.
-For the data set by Andersson et al.
-[-@Andersson2017], the algorithm was successfully fitted for every participant and condition.
-
-(ref:tab-starting-values) Starting values for fitting gazeHMM to benchmark data sets
+For the data set by @Andersson2017, the algorithm was successfully fitted for every participant and condition.
+
+(ref:tab-starting-values) Starting values for response models in fitting gazeHMM to benchmark data sets
 (ref:tab-starting-values-note) Starting values for velocity and acceleration signals are shown before scaling down by factor 100.
@@
 
-tab.start <- as.data.frame(matrix(c(10, 10, 10, 10, 0, 2*pi,
+tab.start <- as.data.frame(matrix(c(10, 10, 10, 10, NA, NA,
                       50, 50, 50, 50, 0, 10,
@@
           col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-starting-values)",
@@
 
-### Model comparison
+### Model Comparison
 
@@
 
-load("~/Uni/Psychologie Master/Internship/GazeHMM/validation/Andersson2017_fitted.Rdata")
-load("~/Uni/Psychologie Master/Internship/GazeHMM/validation/Andersson2017_raw.Rdata")
+load(here("validation/Andersson2017_fitted.Rdata"))
+load(here("validation/Andersson2017_raw.Rdata"))
 
@@
   p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
-    scale_fill_continuous(name = "Schwarz\nweight") +
-    scale_x_discrete(name = "Number of states in model", labels = as.character(1:5))
+    scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
+    scale_y_discrete(name = "Subject") +
+    scale_fill_distiller(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")
 
@@
 
-To examine whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states on two benchmark data sets and compared their Schwarz weights.
-For the Andersson et al.
-[-@Andersson2017] data set, I expected the highest weights for the three-state models in the image condition.
-For the moving dots and video condition, I predicted the highest weights for the four-state models.
-Figure X shows the Schwarz weights for different subjects and models in the image condition.
-For all subjects, the five-state model displays the highest weight, suggesting that it is the most likely to have generated the data.
-In the moving dots condition, the five-state model has the highest weights for most subjects, but the one-, three, and four-state models exhibit the highest weights for two subjects each (see Figure X).
-The video condition has the same pattern as the image condition, as the five-state model consistently has the highest weights (see Figure X).
-In contrast to my hypothesis, applying different HMMs to the Andersson et al.
-data suggests that the five-state model has most likely generated the data.
+To examine whether HMMs are accurately describing eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
+Schwarz weights are the result of transforming a set of BIC values.
+They can be interpreted as the probability for a model to have generated the data it was fitted to.
+For the @Andersson2017 data set, I expected the highest weights for the three-state models in the image condition.
+For the moving dots and video conditions, I predicted the highest weights for the four-state model.
+Figure \@ref(fig:plot-schwarz-weights-img) shows the Schwarz weights for different subjects and models in the image condition.
+For all subjects, the five-state model displays the highest weight, suggesting that it hast most likely generated the data.
+In the moving dots condition, the five-state model has the highest weights for most subjects, but the one-, three, and four-state models exhibit the highest weights for two subjects each (see Figure \@ref(fig:plot-schwarz-weights-dots)).
+The video condition has the same pattern as the image condition, as the five-state model consistently has the highest weights (see Figure \@ref(fig:plot-schwarz-weights-vid)).
+In contrast to my hypothesis, applying different HMMs to the @Andersson2017 data suggests that the five-state model has most likely generated the data.
 Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
-The model comparison for the Ehinger et al.
-[-@Ehinger2019] data set yielded similar results (see Appendix).
-
-### Comparison to other algorithms
+The model comparison for the @Ehinger2019 data set yielded similar results (see Appendix).
+
+(ref:plot-schwarz-weights-img) Schwarz weights displayed for each subject and HMMs with different numbers of states.
+Models were applied to the image condition of the @Andersson2017 data set.
+Higher weights indicate better model fit.
+
+```{r plot-schwarz-weights-img, fig.cap="(ref:plot-schwarz-weights-img)"}
+
+print(bic.plot[[2]])
+
+```
+
+(ref:plot-schwarz-weights-dots) Schwarz weights displayed for each subject and HMMs with different numbers of states.
+Models were applied to the moving dots condition of the @Andersson2017 data set.
+Higher weights indicate better model fit.
+
+```{r plot-schwarz-weights-dots, fig.cap="(ref:plot-schwarz-weights-dots)"}
+
+print(bic.plot[[1]])
+
+```
+
+(ref:plot-schwarz-weights-vid) Schwarz weights displayed for each subject and HMMs with different numbers of states.
+Models were applied to the video condition of the @Andersson2017 data set.
+Higher weights indicate better model fit.
+
+```{r plot-schwarz-weights-vid, fig.cap="(ref:plot-schwarz-weights-vid)"}
+
+print(bic.plot[[3]])
+
+```
+
+### Comparison to Other Algorithms
 
@@
   geom_col(position = "dodge") + 
-  scale_y_continuous(name = "Disagreement ratio (in %)", limits = c(0, 100), breaks = c(0, 0.25, 0.5, 0.75, 1)*100) +
+  scale_y_continuous(name = "Disagreement (in %)", limits = c(0, 100), breaks = c(0, 0.25, 0.5, 0.75, 1)*100) +
   scale_x_discrete(name = "") + coord_flip()
@@
 
+#### Event Durations
 The hypothesis that HMMs improve the classification performance of eye movement events was investigated by comparing gazeHMM against other algorithms.
-Therefore, I applied gazeHMM with four events to the Andersson et al.
-(2017) data set.
-First, I calculated the RMSD [^1] as described in the original article by Andersson et al.
-for all algorithms included in the study plus gazeHMM.
-Table \@ref(tab:tab-rmsd-4-fix) shows that in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations due to a large amount of very short fixations.
-For saccades, gazeHMM has the fourth lowest RMSD of the algorithms in the image and video conditions but the second highest in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)).
-Here, gazeHMM overestimated the number and duration of saccades (either slightly or substantially).
-Moreover, gazeHMM underestimated the number and duration of PSOs in the image and moving dots conditions, leading to the highest and intermediate RMSD among the competitors, repsectively (see Table \@ref(tab:tab-rmsd-4-pso)).
-In the video condition the duration was slightly overestimated but the duration was underestimated with the highest RMSD.
-No other algorithm was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than among the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)).
-In sum, except for saccades and PSOs, gazeHMM performed worse with regard to the duration of classified events compared to the other algorithms.
+Therefore, I applied gazeHMM with four events to the @Andersson2017 data set.
+Therefore, I calculated the RMSD [^3] of event durations  as described in the original article by Andersson et al.
+for all algorithms included in the study plus gazeHMM: First, the mean and standard deviation of event durations as well as the number of events were normalized to [0,1] for each coder and algorithm.
+Second, the sum of the root square deviation between each algorithm and the average of the two human coders is computed.
+This metric can only be relatively interpreted [@Andersson2017].
+Table \@ref(tab:tab-rmsd-4-fix) shows that, in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations.
+The high RMSD values were mostly due to a large amount of very short fixations.
+For saccades, gazeHMM has a relatively low RMSD of the algorithms in the image and video condition and a relatively high RMSD in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)).
+Here, gazeHMM overestimated the duration of saccades, as indicated by high mean and SD durations.
+For PSOs, gazeHMM had the highest RMSD in the image and video condition (see Table \@ref(tab:tab-rmsd-4-pso)).
+In the moving dots condition, its RMSD was lower than for NH [@Nystrom2010] but higher than for LNS [@Larsson2013].
+Here, gazeHMM underestimated the number of PSOs and their duration compared to the human coders.
+No other algorithm in the study was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)).
+In sum, gazeHMM did not outperform all other algorithms regarding the RMSD of event durations compared to human coders.
+The duration of fixations and smooth pursuits was largely underestimated and the number of these two events largely overestimated.
+For saccades and PSOs, gazeHMM estimated relatively accurate durations and event numbers, but didn't reach the best algorithms.
 
@@
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
           font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-rmsd-4-fix)",
@@
 
-apa_table(rmsd.table[[3]][[1]],
+apa_table(rmsd.table[[3]][[2]],
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
           font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-rmsd-4-sac)",
@@
 
-apa_table(rmsd.table[[3]][[1]],
+apa_table(rmsd.table[[3]][[3]],
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
           font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-rmsd-4-pso)",
@@
 
-apa_table(rmsd.table[[3]][[1]],
+apa_table(rmsd.table[[3]][[4]],
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Moving dots" = c(2, 5), "Image" = c(6, 9), "Video" = c(10, 13)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
           font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-rmsd-4-sp)",
@@
 
-[^1]: In my proposal, I declared to compute the RMSE instead of RMSD.
+[^3]: In my proposal, I declared to compute the RMSE instead of RMSD.
 However, since I conducted an in sample comparison, the term deviation is more appropriate.
@@
 
-Second, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
+#### Sample-to-sample Agreement
+Moreover, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of sample-to-sample agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
+Cohen's kappa indicates the agreement between to classifiers accounting for the agreement due to chance.
+Absolute values were interpreted according to @Landis1977.
 For fixations, absolute kappa values in all three conditions indicated a slight to fair agreement between gazeHMM and human coders.
-Compared to the other algorithms, the agreement of gazeHMM was the lowest in the image condition but the hightest in the moving dots and video conditions.
-In contrast, gazeHMM showed absolute kappa values for saccades that correspond to a moderate to substantial agreement.
-In all three conditions, the agreement was intermediate in the moving dots condition and higher than for most other algorithms in the image and video conditions.
+Compared to the other algorithms, the agreement of gazeHMM was the lowest in the image condition but the highest in the moving dots and video condition.
+For saccades, gazeHMM showed kappa values that correspond to a moderate to substantial agreement were relatively high compared to the other algorithms.
+However, gazeHMM never reached the highest agreement.
 For PSOs, absolute kappa values showed slight to fair agreement agreement to human coders.
-It was lower in the moving dots condition but intermediate in the image and video conditions compared to the other algorithms.
-The agreement for smooth puruits was moderate in the moving dots condition but slight in the image and video conditions.
-No other algorithm in the study was designed to detect smooth pursuits.
-In sum, gazeHMM classified samples as saccades more similar to human coders than most other algorithms.
-For the other events, the classification differed from human coding more than for most other algorithms.
+It was relatively low in the moving dots condition and mediocre in the image and video conditions compared to NH and LNS.
+The agreement for smooth puruits was moderate in the moving dots condition but slight in the image and video conditions. No other algorithm in the study was designed to detect smooth pursuits.  
+In sum, gazeHMM revealed a relatively high sample-to-sample agreement to human coders for saccades.
+For PSOs and smooth pursuits, I found mediocre agreement.
+The performance for fixations was absolutely and relatively low.
 
@@
 
-Third, I computed the disagreement ratio as the percentage of samples that were classified differently by gazeHMM compared to the human coders.
-Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events has a higher disagreement ratio than all but two of the other algorithms.
-In the moving dots and video conditions, the disagreement ratios are considerably lower than for all the other algorithms.
-Notably, gazeHMM's disagreement ratios were similar across conditions (around 45%).
-
-(ref:plot-disag) Disagreement ratio between algorithms and human coders for different conditions. gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.
+#### Disagreement and Confusion
+At last, I computed the overall disagreement [^4] as the percentage of all samples taken together that were classified differently by gazeHMM than the human coders.
+Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events had a higher overall disagreement than all but two of the other algorithms.
+In the moving dots and video conditions, the disagreement was considerably lower than for all the other algorithms.
+Notably, gazeHMM's disagreement was similar across conditions (around 45%).
+In summary, gazeHMM had a lower overall disagreement than all other algorithms in the moving dots and video conditions but higher disagreement in the image condition.
+
+[^4]: Even though I proposed to compute the *disagreement ratio* and @Andersson2017 also used that term, the comparison technically involves percentages and not ratios.
+Thus, I decdided to use only the term *disagreement* instead.
+
+(ref:plot-disag) Disagreement between algorithms and human coders for different conditions (in %). gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.
 
@@
 
-Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms with regard to RMSD, sample-to-sample agreement, or disagreement ratio.
-It rather fared better or worse than most algorithms for different events and conditions.
-Especially for saccades, the agreement was relatively high and improved upon the IHMM algorithm included in the study.
-Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data might be more appropriate and yield better validation results.
+Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms in terms of RMSD and sample-to-sample agreement.
+The two criteria rather indicated a mediocre performance of gazeHMM.
+However, the overall disagreement was lower for gazeHMM than for all other algorithms in the dynamic conditions but the highest in the static image condition.
+Looking at the confusion matrix in Table \@ref(tab:tab-conf-4), it can be seen that gazeHMM confused fixations and smooth pursuit to a large extend.
+In the image condition, it overclassified smooth pursuit events, while they were underclassified in the dynamic conditions.
+
+### gazeHMM With Three States
+Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data seems theoretically more appropriate.
+Including only three states could prevent gazeHMM from overclassifying smooth pursuits in the image data and yield better validation results.
+Therefore, I decided to apply gazeHMM with three states to the image dat and repeat the comparison to other algorithms exploratorily.
 Table \@ref(tab:tab-rmsd-3) shows the RMSD between gazeHMM and human coders only for the image condition.
-For fixations, gazeHMM shows a comparably low RMSD and only two other algorithms have lower values.
-Oppositely, saccades had the highest RMSD among the compared algorithms.
-The same result was found for PSOs.
-The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kappa-3) (only for the image condition).
-For fixations, the absolute kappa indicated substantial agreement and was high compared to the other algorithms.
+For fixations, gazeHMM shows a comparably low RMSD but it was relatively high for saccades and PSOs.
+The sample-to-sample agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kappa-3) (only for the image condition).
+For fixations, the absolute kappa indicated substantial agreement and was relatively high compared to the other algorithms.
 The absolute kappa for saccades corresponded to moderate agreement.
 Compared to the other algorithms, it was rather low.
-For PSOs, the absolute agreement was fair, but intermediate with regard to the other algorithms.
-Figure \@ref(fig:plot-disag) illustrates the disagreement ratio between gazeHMM with three events and the human coders compared to the other algorithms.
-In the image condition, the disagreement ratio was slightly lower for gazeHMM than for all the other algorithms.
-Summing up, using gazeHMM with three states increased the agreement to human coding with regard to fixation durations and samples classified as fixations.
-However, for saccades and PSOs the agreement decreased.
+For PSOs, the absolute agreement was fair, but relatively low compared to the other algorithms.
+Figure \@ref(fig:plot-disag) illustrates the overall disagreement between gazeHMM with three events and the human coders compared to the other algorithms.
+In all three conditions, the disagreement was slightly lower for gazeHMM than for all the other algorithms.
+Compared to gazeHMM with four states, the RMSD decreased while the sample-to-sample agreement increased for fixations.
+For saccades and PSOs, the opposite pattern was found.
+The overall disagreement substantially decreased.
+Overall, applying gazeHMM with three states to static data seems more appropriate, since the increase in fixation classification outweighs the decrease in classification of saccades and PSOs.
+There was even some support for my second hypothesis since gazeHMM had lower overall disagreement to human coders than all other algorithms.
 
@@
           font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-rmsd-3)",
@@
 
+# Discussion
+In this report, I presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events.
+The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von-Mises and a uniform distribution.
+An HMM classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
+A simulation study revealed that the generative model of gazeHMM recovered parameters and hidden state sequence well with a few exceptions.
+Adding smooth pursuits to the model and noise to the generated data were the most critical factors for decreasing recovery.
+In contrast, higher sample sizes yielded a better parameter recovery.
+The variation in starting values did not affect the recovery but the amount of missing data decreased the recovery of hidden states.
+I applied gazeHMM with different numbers of states to benchmark data and compared the model fit, showing that a five-state HMM had consistently most likely generated the data.
+This result speaks against my first hypothesis that HMMs are useful generative models for eye movements.
+When comparing gazeHMM with four events against other algorithms and human coders, it did not show the best performance for the duration and number of classified events nor the sample-to-sample agreement across all conditions.
+However, it had a lower overall disagreement for dynamic data but higher disagreement for static data.
+I argued that four states might not be appropriate for static data and additionally applied gazeHMM with three states to them.
+That led to substantially better classifications for fixations, but slightly worse performance for saccades and PSOs.
+With three states, gazeHMM had a lower overall disagreement than all other algorithms.
+I treat this pattern as moderate evidence for my second hypothesis that HMMs improve the classification of eye movements.
+Taken together, I only found some evidence for the one part of my research question: HMMs improve classifying gaze data but not describing them.
+
+\newpage
+
 # Technical Appendix
+# References
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+<div id = "refs"></div>
+\endgroup

### 2020-06-15 manuscript/internship_report.Rmd
add discussion

@@
 
+```{r calculate integrated complete likelihood criterion, include=FALSE}
+
+# Compute ICL
+
+ICL <- function(mod) {
+
+  -2*logLik(mod, method = "classification") + npar(mod)*log(nobs(mod))
+
+}
+
+A2017.icl <- lapply(A2017.fit, function(stim) {
+  out <- lapply(stim, function(subj) {
+    out <- lapply(1:length(subj), function(mod) {
+
+      if(mod == 1) {
+
+        icl <- try(-2*subj[[mod]][["LL"]] + 6*log(subj[[mod]][["N"]]))
+
+      } else {
+
+        icl <- try(ICL(subj[[mod]]$model))
+
+      }
+
+      return(ifelse(is.numeric(icl), icl, NA))
+    })
+
+    return(schwarz.weights(unlist(out)))
+  })
+
+  df <- as.data.frame(reduce(out, rbind))
+
+  names(df) <- paste("model_", 1:length(stim[[1]]), sep = "")
+
+  return(df)
+})
+
+# Create ICL weight plots
+
+icl.plot <- lapply(A2017.icl, function(x) {
+
+  data.long <- x %>% 
+    mutate(subject = 1:nrow(x)) %>%
+    pivot_longer(names(x), names_to = "model", values_to = "weight") %>%
+    mutate_at(c("subject", "model"), as.factor)
+
+  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
+    scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
+    scale_y_discrete(name = "Subject") +
+    scale_fill_distiller(name = "ICL weight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")
+
+  return(p)
+})
+
+```
+
 To examine whether HMMs are accurately describing eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
@@
 The model comparison for the @Ehinger2019 data set yielded similar results (see Appendix).
+A recent model recovery study showed that the BIC tended to prefer overly complex HMMs when they were misspecified [e.g., the conditional independence assumption was violated; @Pohle2017].
+Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2013] performed better in choosing the correct data generating model.
+Therefore, I computed the weighted ICL criterion for the HMMs fitted to the benchmark data sets.
+Using the ICL as the model selection criterion yielded very similar results to the BIC (see Appendix).
+The preference for the five-state model was even more consistent across conditions and subjects.
 
@@
 I treat this pattern as moderate evidence for my second hypothesis that HMMs improve the classification of eye movements.
-Taken together, I only found some evidence for the one part of my research question: HMMs improve classifying gaze data but not describing them.
+Taken together, I only found some evidence for one part of my research question: HMMs improve classifying gaze data but not describing them.
+Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle.
+The overlap can cause the HMM to confuse parameters and hidden states.
+The decrease in parameter recovery (especially for shape parameters) due to noise shows that the overlap is enhanced by more dispersion in the data.
+The shape parameters might be particularly vulnerable to extreme data points.
+The decrease in state recovery for long and many intervals of missing data can be explained by the tendency of HMMs to stay in the same state when no information is available from the data.
+However, the decrease was rather small, leading to the conclusion that the HMM can deal well with missing data.
+Despite these drawbacks, the recovery of the generative model in gazeHMM seems very promising.
+The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).
+The model comparison on the benchmark data clearly provided evidence against HMMs being useful generative models for eye movement data.
+Even using a model selection criterion that identified correct models succesfully in previous model comparison studies [@Pohle2017], did not yield the expected pattern.
+There are several explanations for this result:
+(1) *There were more eye movement events present in the data than I expected*.
+Eye movement events can be divided into subevents.
+For example, fixations consist of drift and tremor movements [@Duchowski2017] and PSOs encompass dynamic, static, and glissadic over- and undershoots [@Larsson2013].
+A study on a recently developed HMM algorithm supports this explanation: @Houpt2018 applied the unsupervised BP-AR-HMM algorithm to the @Andersson2017 data set and classified more distinct states than the human coders.
+Some of the states classified by BP-AR-HMM matched the same event coded by humans.
+Since the subevents are usually not interesting for users of classification algorithms, the ability of HMMs to classify might limit their ability to describe eye movements here.
+(2) *Model selection criteria are generally not appropriate for comparing HMMs with differnt numbers of states*.
+This argument has been discussed in the field of ecology [see @Li2017], where studies found that selection criteria preferred models with more states than expected [similar to the result of this study; e.g., @Langrock2015].
+@Li2017 explain this bias with the simplicity of the submodels in HMMs: Initial state, transition, and response models for each state are usually relatively simple.
+When they do not describe the processes in the respective states accurately, the selection criteria compensate for that by preferring a model with more states.
+Thus, there are not more latent states present in the data, but the submodels of the HMM are misspecified or too simple.
+Correcting for model misspecifications, led to better model recovery in studies on animal movements [@Langrock2015; @Li2017].
+However, @Pohle2017 showed in simulations that the ICL identified the correct model despite several misspecifications.
+It has to be noted that the study by @Pohle2017 only used data generating models with two states.
+(3) *The submodels of gazeHMM were misspecified*.
+@Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models.
+Both situations could apply to gazeHMM and eye movement data.
+Outliers occur frequently in eye-tracking data due to measurement error.
+Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task the data is obtained from [@Langrock2015].
+Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf.
+@Li2017].
+This aspect could indeed oversimplify the complex nature of eye movement data.
+Comparing gazeHMM to other algorithms on benchmark data showed that gazeHMM moderately improved the event classification in agreement with human coders.
+However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yieled different results.
+The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events.
+Of course, that makes it still a useful tool for classification.
+Nevertheless, Cohen's kappa values of 0.72 (saccades - video - four states) or 0.68 (fixations - image - three states) indicate substantial agreement to human coders, especially in light of the maximum reference of ~0.8 from the simulation study.
+At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018 observed substantial differences between coders and within coders over time.
+Despite these differences, they recommend comparisons with human coding to demonstrate the performance of new algorithms and finding errors in their design.
+
+## Advantages of gazeHMM
+In view of the four proposed goals that gazeHMM should fulfill, I can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they have only minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
+Their default values should be appriate for most applications.
+A major advantage of gazeHMM is that it does not require human labeled data as input.
+Instead, it estimates all parameters and hidden states from the data.
+Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
+This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data.
+Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit.
+Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2017], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
+However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events.
+Therefore, I recommend using gazeHMM with four events onyl for exploratory purposes.
+Without smooth pursuits, I consider gazeHMM's classification as appropriate for application.
+Lastly, its implementation in R using depmixS4 [@Visser2010] should make gazeHMM a tool that is easy to use and customize for individual needs.
+
+## Future Directions
+Despite its advantages, there several aspects in which gazeHMM can be improved:
+
+Multivariate gamma - conditional independence
+Blinks and noise
+Non-parametric response distributions
+Time varying covariates
+Semi-HMM
+Metric separating fixations and sp
+Constraining parameters
 
@@
 
+# Appendix
 # Technical Appendix

### 2020-06-16 manuscript/internship_report.Rmd
add conclusion

@@
 Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best.
-However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007, and @Zemblys2018].
+However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018].
 This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics.
@@
 
+\newpage
+
 # Simulation Study
@@
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
-    scale_x_continuous(name = "True transition probability") +
-    scale_y_continuous(name = "Estimated transition probability")
+    scale_x_continuous(name = "True response parameter") +
+    scale_y_continuous(name = "Estimated response parameter")
 
@@
 ```
+
+\newpage
 
@@
 
-(ref:plot-regresp-1-2) Regression lines between true and estimated transition response parameters of the two-state HMM in part one. Top facet labels indicate response parameters.
+(ref:plot-regresp-1-2) Regression lines between true and estimated response parameters of the two-state HMM in part one. Top facet labels indicate response parameters.
 
@@
 
-(ref:plot-regresp-1-3) Regression lines between true and estimated transition response parameters of the three-state HMM in part one. Top facet labels indicate response parameters.
+(ref:plot-regresp-1-3) Regression lines between true and estimated response parameters of the three-state HMM in part one. Top facet labels indicate response parameters.
 
@@
 
-(ref:plot-regresp-1-4) Regression lines between true and estimated transition response parameters of the four-state HMM in part one. Top facet labels indicate response parameters.
+(ref:plot-regresp-1-4) Regression lines between true and estimated response parameters of the four-state HMM in part one. Top facet labels indicate response parameters.
 
@@
 
+\newpage
+
 # Discussion
@@
 The overlap can cause the HMM to confuse parameters and hidden states.
-The decrease in parameter recovery (especially for shape parameters) due to noise shows that the overlap is enhanced by more dispersion in the data.
-The shape parameters might be particularly vulnerable to extreme data points.
+The decrease in parameter recovery (especially for scale parameters) due to noise shows that the overlap is enhanced by more dispersion in the data.
+The scale parameters might be particularly vulnerable to extreme data points.
 The decrease in state recovery for long and many intervals of missing data can be explained by the tendency of HMMs to stay in the same state when no information is available from the data.
@@
 However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events.
-Therefore, I recommend using gazeHMM with four events onyl for exploratory purposes.
+Therefore, I recommend using gazeHMM with four events only for exploratory purposes.
 Without smooth pursuits, I consider gazeHMM's classification as appropriate for application.
@@
 ## Future Directions
-Despite its advantages, there several aspects in which gazeHMM can be improved:
-
-Multivariate gamma - conditional independence
-Blinks and noise
-Non-parametric response distributions
-Time varying covariates
-Semi-HMM
-Metric separating fixations and sp
-Constraining parameters
+Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for an example, see @Balakrishnan2009].
+Potential problem of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters).
+Another option to model the correlation could be to include one of the response variables as a covariate of the other.
+Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred.
+This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.
+Third, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data.
+As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements.
+As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
+Fourth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonable short saccades, gazeHMM could explicitly model the duration of events.
+This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2015].
+Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs.
+This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades.
+A drawback of the extension would be higher computational costs.
+Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM.
+This could potentially be achieved by using different response distributions or parameter optimization methods.
+Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.
+
+## Conclusions
+Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events.
+For smooth pursuits, the classification is not optimal and thus not recommended, yet.
+On one hand, the algorithm has some advantages over concurrent event classfication algorithms, not relying on human-labeled input being the most important one.
+On the other hand, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events.
+These difficulties highlight general challenges for HMMs in general, but also specifically for eye movements.
+Revisting my research question, I conclude that HMMs are suitable models for classifying eye movement events but still lack the ability to accurately describe them.
+
+*Acknowledgment*. I would like to thank my supervisors Ingmar Visser and Simon Kucharsky for their helpful advice on the development and implementation of the study.
 

### 2020-06-17 algorithm/gazeHMM.R
add sampling rate in postprocessing

@@ -166,7 +166,7 @@ gazeHMM <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
   
   # Postprocessing
   
-  post <- postprocess(prep, model.fit@posterior$state, min.sac)
+  post <- postprocess(prep, model.fit@posterior$state, fr, min.sac)
   
   
   # Output: Data frame with samples, data frames with events, fitted model

### 2020-06-17 algorithm/postprocessing.R

@@ -4,7 +4,7 @@
 # Date: 20.04.2020
 
 
-postprocess <- function(data, state, min.sac) { # data frame with dependent variables, vector with sample states, minimum saccade duration
+postprocess <- function(data, state, fr, min.sac) { # data frame with dependent variables, vector with sample states, minimum saccade duration
   
   source("algorithm/postprocessing_helper_functions.R")
   
@@ -32,6 +32,11 @@ postprocess <- function(data, state, min.sac) { # data frame with dependent vari
   number <- df$number
   
   warning(paste(sum(label != state), " samples were relabeled during postprocessing!"))
+  
+  
+  # Multiply acceleration by sampling rate
+  
+  data$acc <- data$acc * fr
   
   
   # Initialize output event metrics 

### 2020-06-17 algorithm/preprocessing.R

@@ -107,7 +107,7 @@ preprocess <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
   # Remove outliers
   
   max.vel <- 1e3
-  max.acc <- 1e5
+  max.acc <- 1e2
   
   valid[valid] <- ifelse(vel[valid] > max.vel | acc[valid] > max.acc, F, valid[valid])
   

### 2020-06-17 simulation/model_simulation.R
remove source command

@@ -6,8 +6,6 @@
 HMM_simulate <- function(n, nstates, trueresp, truetr, truein) {
   
   require(depmixS4)
-  source("algorithm/model_helper_functions.R")
-  
   
   data <- data.frame(vel = numeric(n),
                      acc = numeric(n),

### 2020-06-17 manuscript/internship_report.Rmd
add appendices

@@
 bibliography      : "report.bib"
+appendix          : 
+ - "report_appendix.Rmd"
+ - "report_technical_appendix.Rmd"
 
@@
     scale_fill_distiller(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")
-
-  return(p)
-})
-
-```
-
-```{r calculate integrated complete likelihood criterion, include=FALSE}
-
-# Compute ICL
-
-ICL <- function(mod) {
-
-  -2*logLik(mod, method = "classification") + npar(mod)*log(nobs(mod))
-
-}
-
-A2017.icl <- lapply(A2017.fit, function(stim) {
-  out <- lapply(stim, function(subj) {
-    out <- lapply(1:length(subj), function(mod) {
-
-      if(mod == 1) {
-
-        icl <- try(-2*subj[[mod]][["LL"]] + 6*log(subj[[mod]][["N"]]))
-
-      } else {
-
-        icl <- try(ICL(subj[[mod]]$model))
-
-      }
-
-      return(ifelse(is.numeric(icl), icl, NA))
-    })
-
-    return(schwarz.weights(unlist(out)))
-  })
-
-  df <- as.data.frame(reduce(out, rbind))
-
-  names(df) <- paste("model_", 1:length(stim[[1]]), sep = "")
-
-  return(df)
-})
-
-# Create ICL weight plots
-
-icl.plot <- lapply(A2017.icl, function(x) {
-
-  data.long <- x %>% 
-    mutate(subject = 1:nrow(x)) %>%
-    pivot_longer(names(x), names_to = "model", values_to = "weight") %>%
-    mutate_at(c("subject", "model"), as.factor)
-
-  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
-    scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
-    scale_y_discrete(name = "Subject") +
-    scale_fill_distiller(name = "ICL weight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")
 
@@
 
-\newpage
-
-# Appendix
-# Technical Appendix
 # References

### 2020-06-17 manuscript/report_appendix.Rmd
upload appendix

---
suppress-bibliography: true
---

(ref:plot-sim-exp) Simulated data from the HMM with the default parameters of the simulation study. Displayed are acceleration (in deg/s$^2$, multiplied by 10000), sample-to-sample (relative) angle (in rad), and velocity (in deg/s, multiplied by 10) over time (in s). Acceleration and velocity were multiplied back to their natural scale. Note that PSOs (State 3, blue dots) do not only occur after saccades because the HMM parameters were unconstrained.

```{r plot-sim-exp, results='hold', fig.cap="(ref:plot-sim-exp)"}

source(here("simulation/model_simulation.R"))

resp.sim.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)), 
                     sac = list(vel = c(3, 10), acc = c(3, 3), angle = c(0, 1)),
                     pso = list(vel = c(3, 1), acc = c(3, 3), angle = c(pi, 1)),
                     sp = list(vel = c(3, 1), acc = c(3, 0.15), angle = c(0, 1)))

tr.sim.true <- matrix(0.1/3, 4, 4)

diag(tr.sim.true) <- 0.9

sim.model <- simulate(HMM_simulate(n = 500, nstates = 4, trueresp = resp.sim.true, truetr = tr.sim.true,
                          truein = rep(1/4, 4)), seed = 123)

data.frame(State = as.factor(sim.model@states), Velocity = sim.model@response[[1]][[1]]@y*10,
                           Acceleration = sim.model@response[[1]][[2]]@y*1e4, Angle = sim.model@response[[1]][[3]]@y) %>% 
  mutate(Time = 1:500*0.002) %>%
  pivot_longer(c("Velocity", "Acceleration", "Angle"), names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = ifelse(Metric == "Angle", "Rel. angle", Metric)) %>%
  ggplot(aes(x = Time, y = Value)) + 
  facet_grid(rows = vars(Metric), scales = "free_y", switch =  "y") + 
  geom_path() + geom_point(aes(color = State), size = 0.8) + 
  scale_y_continuous(name = "") + scale_x_continuous(name = "Time (in s)") +
  theme(strip.placement = "outside")

```

```{r calculate integrated complete likelihood criterion, include=FALSE}

# Compute ICL

ICL <- function(mod) {

  -2*logLik(mod, method = "classification") + npar(mod)*log(nobs(mod))

}

A2017.icl <- lapply(A2017.fit, function(stim) {
  out <- lapply(stim, function(subj) {
    out <- lapply(1:length(subj), function(mod) {

      if(mod == 1) {

        icl <- try(-2*subj[[mod]][["LL"]] + 6*log(subj[[mod]][["N"]]))

      } else {

        icl <- try(ICL(subj[[mod]]$model))

      }

      return(ifelse(is.numeric(icl), icl, NA))
    })

    return(schwarz.weights(unlist(out)))
  })

  df <- as.data.frame(reduce(out, rbind))

  names(df) <- paste("model_", 1:length(stim[[1]]), sep = "")

  return(df)
})

# Create ICL weight plots

icl.plot <- lapply(A2017.icl, function(x) {

  data.long <- x %>% 
    mutate(subject = 1:nrow(x)) %>%
    pivot_longer(names(x), names_to = "model", values_to = "weight") %>%
    mutate_at(c("subject", "model"), as.factor)

  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
    scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
    scale_y_discrete(name = "Subject") +
    scale_fill_distiller(name = "ICL weight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")

  return(p)
})

```

(ref:plot-icl-weights-img) ICL weights displayed for each subject and HMMs with different numbers of states. Models were applied to the image condition of the @Andersson2017 data set. Higher weights indicate better model fit.

```{r plot-icl-weights-img, results='hold', fig.cap="(ref:plot-icl-weights-img)"}

print(icl.plot[[2]])

```

(ref:plot-icl-weights-dots) ICL weights displayed for each subject and HMMs with different numbers of states. Models were applied to the moving dots condition of the @Andersson2017 data set. Higher weights indicate better model fit.

```{r plot-icl-weights-dots, results='hold', fig.cap="(ref:plot-icl-weights-dots)"}

print(icl.plot[[1]])

```

(ref:plot-icl-weights-vid) ICL weights displayed for each subject and HMMs with different numbers of states. Models were applied to the video condition of the @Andersson2017 data set. Higher weights indicate better model fit.

```{r plot-icl-weights-vid, results='hold', fig.cap="(ref:plot-icl-weights-vid)"}

print(icl.plot[[3]])

```

(ref:plot-example-classification) Example data from @Andersson2017 displayed as x-, and y-coordinates, velocity (in deg/s), acceleration (in deg/s$^2$), and sample-to-sample (relative) angle (in rad). Colors indicate classification by gazeHMM and symbols by human coder MN. Note that the human coder classified all fixations from gazeHMM as smooth pursuits and delayed the separation between the saccade and PSOs.

```{r plot-example-classification, results='hold', fig.cap="(ref:plot-example-classification)", fig.height=8}

cbind(A2017.fit[[3]][[3]][[4]]$samples, label_MN = as.factor(A2017[[3]][[3]]$label_MN)) %>% 
  gather(c("x", "y", "vel", "acc", "angle"), key = metric, value = value, factor_key = T) %>% 
  mutate_at(vars(label, state), as.factor) %>%
  dplyr::filter(t > 4.0, t <= 4.5) %>% 
  rename(coderMN = label_MN, gazeHMM = label, Time = t) %>% 
  mutate(metric = fct_recode(metric, "x-coordinate" = "x", "y-coordinate" = "y", "Velocity" = "vel",
                             "Acceleration" = "acc", "Rel. angle" = "angle"),
         coderMN = fct_recode(coderMN, "Fix" = "1", "Sacc" = "2", "PSO" = "3", "SP" = "4", "Blink" = "5", "Other" = "6"),
         gazeHMM = fct_recode(gazeHMM, "Noise" = "0", "Fix" = "1", "Sacc" = "2", "PSO" = "3", "SP" = "4")) %>%
  ggplot(aes(x = Time, y = value)) + 
  facet_wrap(vars(metric), ncol = 1, scales = "free", strip.position = "left") +
  geom_path() + xlab("Time in s") + ylab("") +
  geom_point(aes(color = gazeHMM, shape = coderMN), size = 0.8) + 
  theme(strip.placement = "outside",
        axis.text = element_text(size = 8))

```

(ref:tab-example-trans) Transition probabability estimates for example data from @Andersson2017
(ref:tab-example-trans-note) Rows indicate from which state and columns into which state the model transitions.

```{r tab-example-trans, results='hold'}

tab.exp.tr <- matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[5:20], 2), 4, 4, byrow = T)

apa_table(tab.exp.tr,
          col.names = as.character(1:4),
          col_spanners = list("To state" = c(1,4)),
          row.names = TRUE,
          added_stub_head = "From\nstate",
          caption = "(ref:tab-example-trans)",
          note = "(ref:tab-example-trans-note)")

```

(ref:tab-example-resp) Response parameter estimates for example data from @Andersson2017
(ref:tab-example-resp-note) Con = concentration parameter of the von-Mises distribution.

```{r tab-example-resp, results='hold'}

tab.exp.resp <- matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[21:44], 2), 4, 6, byrow = T)

tab.exp.resp[1, c(5,6)] <- NA

apa_table(tab.exp.resp,
          col.names = c("Shape (vel)", "Scale (vel)", "Shape (acc)", "Scale (acc)", "Mean", "Con"),
          row.names = TRUE,
          format.args = list(na_string = "-"),
          added_stub_head = "State",
          caption = "(ref:tab-example-resp)",
          note = "(ref:tab-example-resp-note)")

```

### 2020-06-17 manuscript/report_technical_appendix.Rmd
upload technical appendix

---
suppress-bibliography: true
---

# Technical Details
The algorithm is divided in three parts: Preprocessing, classification, and postprocessing.

Input: x-coordinate, y-coordinate, timestamp, unit of coordinates (px vs. deg), screen resolution, screen dimensions, distance to screen, sampling rate, blink indicator, order of SG filter, length of SG filter, number of states in the HMM, start values for the HMM, downsampling factor, random response start values, seed for random response start values, control options for fitting the HMM, minimum saccade duration.

## Preprocessing

(@) The algorithm checks whether samples are valid: Samples for which the x- or y-coordinates are NA, infinity, or (0,0) are marked (internally) as invalid.
(@) Samples that match the blink indicator are labeled as blinks.
(@) If the unit is px, the x- and y-coordinates are transformed into va with: 
    i. The coordinates are centered ([0,0] is center)
    ii. $\theta_{deg} = atan(\frac{x*res_x}{d_{screen}*dim_x})\frac{180}{\pi}$
(@) Samples that are outside the screen are marked as invalid.
(@) Velocities and accelerations for the x- and y-coordiantes are calculated using a Savitzky-Golay filter (3rd order, length = 5).
(@) Combined velocities and accelerations are calculated as: $v = f \sqrt{v_x^2 + v_y^2}$ and $a = f \sqrt{a_x^2 + a_y^2}$ with $f$ as the sampling rate.
(@) Velocities and accelerations of value zero are nudged to be 0.01.
(@) Samples with velocity above 1000 deg/s or acceleration above 100000 deg/s$^2$ are marked as invalid.
(@) The sample-to-sample angle is calculated as: 
$$\theta_t = atan2(\frac{y_{t+1} - y_t}{x_{t+1} - x_t}) - atan2(\frac{y_t - y_{t-1}}{x_t - x_{t-1}}).$$ 
Angles are normalized to $[0,2\pi]$ by adding $2\pi$ to negative values.
(@) Samples that have an angle with value NA are marked as invalid.
(@) The velocity, acceleration, and angle of invalid samples are set to NA.

Output: Data frame with x-coordinate, y-coordinate, timestamp, velocity, acceleration, angle, and initial label variable with blinks.

## Classifcation
Input: Data frame from preprocessing and HMM related general input variables.

(@) Sets default start values if none are provided. 
(@) Downsamples velocity and acceleration.
(@) Fits a HMM with velocity, acceleration, and angle as dependent variables.  
Velocity and acceleration are modeled by a two-parameter gamma distribution (shape and scale). Angle is modeled by a uniform distribution for the first component and von-Mises distributions (mean and kappa) for further components. Initial state probabilities are optimized on the identity scale and transition probabilities on the multinomial logit scale. All response parameters except for the mean parameters of von-Mises distributions are optimized on the log-scale.

Output: Fitted model.

## Postprocessing
Input: Data frame from preprocessing and posterior state sequence of the fitted model.

(@) Relabels posterior states of samples with recursive algorithm:
    i. If a fixation or smooth pursuit sample has no previous or following sample with the same label (single sample event), 
it is relabeled as the label from the previous sample.
    ii. If a PSO sample is followed by a saccade sample or following a non-saccade sample,
it is relabeled as the last different label.
    iii. Saccade events (batch of consecutive saccade samples) below the minimum saccade duration (in s) are relabeled as the label from the previous event.
    iv. These steps are applied recursively until all conditions are met.
(@) For each event (batch of subsequent valid samples belonging to the same label), the following metrics are computed:
    i. Fixations:
        a. Duration: Max. timestamp - min. timestamp
        b. Position: Trimmed mean (20%) of x- and y-coordinates
    ii. Saccades, PSOs, and smooth pursuits:
        a. Duration: Max. timestamp - min. timestamp
        b. Start: First x- and y- coordinates
        c. End: Last x- and y- coordinates
        d. Amplitude: Euclidian distance from start to end
        e. Max. velocity
        f. Avg. velocity
        g. Max. acceleration $\cdot 1000$
        h. Avg. acceleration $\cdot 1000$
        i. Direction: $\theta = atan2(\frac{end_y - start_y}{end_x - start_x})$

General output: List with a data frame containing the labeled samples, a list with data frames for the different events,
and the fitted model.

### 2020-06-17 manuscript/internship_report.Rmd
add Ehinger et al (2019) results

@@
 The following section describes the development of gazeHMM and its final version that has been used to obtain results.
-Technical details can be found in the Technical Appendix.
+Technical details can be found in Appendix B.
 
@@
 PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013].
-They are specifically distinguished by their change in direction clustered around 180° [@Pekkanen2017].
+They are specifically distinguished by their change in direction clustered around 180Ã‚Â° [@Pekkanen2017].
 Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
@@
 
-```{r, filter-angle, fig.cap="(ref:filter-angle)"}
+```{r filter-angle, fig.cap="(ref:filter-angle)"}
 
@@
 Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
-To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the forward difference (see Technical Appendix).
+To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the forward difference (see Appendix B).
 
@@
 
-```{r, filter-comparison, fig.cap="(ref:filter-comparison)"}
+```{r filter-comparison, fig.cap="(ref:filter-comparison)"}
 
@@
 (ref:sim-ranges) Approximate ranges of response variables used to generate parameter values
-(ref:sim-ranges-note) Units are °/s (velocity), °/s^2^ (acceleration), and radians (angle).
+(ref:sim-ranges-note) Units are Ã‚Â°/s (velocity), Ã‚Â°/s^2^ (acceleration), and radians (angle).
 ~ indicates that the distribution has a peak at this value.
@@
 
-```{r, sim-ranges}
+```{r sim-ranges}
 
@@
 
-```{r, sim-parameters-trans}
+```{r sim-parameters-trans}
 
@@
 
-```{r, sim-parameters-resp}
+```{r sim-parameters-resp}
 
@@
 
-```{r, plot-acc-1-2, fig.cap="(ref:plot-acc-1-2)"}
+```{r plot-acc-1-2, fig.cap="(ref:plot-acc-1-2)"}
 
@@
 
-```{r, plot-acc-1-3, fig.cap="(ref:plot-acc-1-3)"}
+```{r plot-acc-1-3, fig.cap="(ref:plot-acc-1-3)"}
 
@@
 
-```{r, plot-acc-1-4, fig.cap="(ref:plot-acc-1-4)"}
+```{r plot-acc-1-4, fig.cap="(ref:plot-acc-1-4)"}
 
@@
 
-```{r, plot-acc-2, fig.cap="(ref:plot-acc-2)"}
+```{r plot-acc-2, fig.cap="(ref:plot-acc-2)"}
 
@@
 
-```{r, plot-acc-4, fig.cap="(ref:plot-acc-4)"}
+```{r plot-acc-4, fig.cap="(ref:plot-acc-4)"}
 
@@
 I applied gazeHMM on two benchmark data sets to test my hypotheses.
+Note that, for the @Ehinger2019 data set, I only fitted models to the Eyelink data to reduce the complexity of the analyses.
 For both data sets, I used the same set of starting values (initial state model: $\rho=1/k$; transition model: $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$; response model: see Table \@ref(tab:tab-starting-values)) to estimate the parameters of the HMM.
@@
 To improve the fitting of the gamma distributions, velocity and acceleration signals were scaled down by factor 100 (so were the starting values for their gamma distributions).
-The algorithm was applied separately for every participant and every condition.
+The algorithm was applied separately for every participant and every condition, task, or block.
 For the data set by @Andersson2017, the algorithm was successfully fitted for every participant and condition.
@@
 Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
-The model comparison for the @Ehinger2019 data set yielded similar results (see Appendix).
+The model comparison for the @Ehinger2019 data set yielded similar results (see Figures \@ref(fig:plot-schwarz-weights-task4) and \@ref(fig:plot-schwarz-weights-task5)).
 A recent model recovery study showed that the BIC tended to prefer overly complex HMMs when they were misspecified [e.g., the conditional independence assumption was violated; @Pohle2017].
-Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2013] performed better in choosing the correct data generating model.
-Therefore, I computed the weighted ICL criterion for the HMMs fitted to the benchmark data sets.
-Using the ICL as the model selection criterion yielded very similar results to the BIC (see Appendix).
+Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2000] performed better in choosing the correct data generating model.
+Therefore, I computed the weighted ICL criterion for the HMMs fitted to the @Andersson2017 data set.
+Using the ICL as the model selection criterion yielded very similar results to the BIC (see Figures \@ref(fig:plot-icl-weights-img), \@ref(fig:plot-icl-weights-dots), and \@ref(fig:plot-icl-weights-vid)).
 The preference for the five-state model was even more consistent across conditions and subjects.
@@
 There are several explanations for this result:
-(1) *There were more eye movement events present in the data than I expected*.
+*There were more eye movement events present in the data than I expected*.
 Eye movement events can be divided into subevents.
@@
 Since the subevents are usually not interesting for users of classification algorithms, the ability of HMMs to classify might limit their ability to describe eye movements here.
-(2) *Model selection criteria are generally not appropriate for comparing HMMs with differnt numbers of states*.
+*Model selection criteria are generally not appropriate for comparing HMMs with differnt numbers of states*.
 This argument has been discussed in the field of ecology [see @Li2017], where studies found that selection criteria preferred models with more states than expected [similar to the result of this study; e.g., @Langrock2015].
@@
 It has to be noted that the study by @Pohle2017 only used data generating models with two states.
-(3) *The submodels of gazeHMM were misspecified*.
+*The submodels of gazeHMM were misspecified*.
 @Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models.
@@
 Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit.
-Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2017], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
+Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2018], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
 However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events.
@@
 Fourth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonable short saccades, gazeHMM could explicitly model the duration of events.
-This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2015].
+This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2006].
 Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs.
@@
 
-*Acknowledgment*. I would like to thank my supervisors Ingmar Visser and Simon Kucharsky for their helpful advice on the development and implementation of the study.
-
 # References

### 2020-06-17 manuscript/report_appendix.Rmd
add Ehinger et al (2019) results

@@
 
-(ref:plot-sim-exp) Simulated data from the HMM with the default parameters of the simulation study.
+# Additional Figures and Tables
+
+(ref:plot-sim-ex) Simulated data from the HMM with the default parameters of the simulation study.
 Displayed are acceleration (in deg/s$^2$, multiplied by 10000), sample-to-sample (relative) angle (in rad), and velocity (in deg/s, multiplied by 10) over time (in s).
@@
 
-```{r plot-sim-exp, results='hold', fig.cap="(ref:plot-sim-exp)"}
+```{r plot-sim-ex, fig.cap="(ref:plot-sim-ex)"}
 
-source(here("simulation/model_simulation.R"))
+source(here::here("simulation/model_simulation.R"))
 
@@
   theme(strip.placement = "outside")
+
+```
+
+```{r calculate Schwarz weights for Ehinger et al., include=FALSE}
+
+load(here("validation/Ehinger2019_fitted.Rdata"))
+
+E2019.bic <- lapply(E2019.fit, function(subj) {
+  out <- lapply(subj, function(task) {
+    out <- lapply(task, function(block) {
+      out <- lapply(1:length(block), function(mod) {
+
+        if(mod == 1) {
+
+          bic <- try(-2*block[[mod]][[1]] + 6*log(block[[mod]][[2]]))
+
+        } else {
+
+          bic <- try(BIC(block[[mod]]$model))
+
+        }
+
+        return(ifelse(is.numeric(bic), bic, NA))
+      })
+
+      return(schwarz.weights(unlist(out)))
+    })
+
+    df <- as.data.frame(reduce(out, rbind))
+
+    names(df) <- paste("model_", 1:ncol(df), sep = "")
+
+    #blockm <- apply(df, 2, median, na.rm = T)
+
+    return(df)
+  })
+
+  return(out)
+})
+
+# Create Schwarz weight plots
+
+bic.plot <- list()
+
+bic.plot[[4]] <- reduce(reduce(E2019.bic, rbind)[,1], rbind) %>%
+  mutate(Subject = rep(1:15, each = 6),
+         Block = rep(1:6, 15)) %>%
+  pivot_longer(contains("model"), names_to = "model", values_to = "weight") %>%
+  mutate_at(c("Subject", "Block", "model"), as.factor) %>%
+  ggplot(aes(x = model, y = Subject)) + geom_tile(aes(fill = weight)) +
+  facet_wrap(vars(Block), ncol = 2, labeller = label_both) +
+  scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
+  scale_y_discrete(name = "Subject") +
+  scale_fill_distiller(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), 
+                       palette = "Spectral")
+
+bic.plot[[5]] <- reduce(reduce(E2019.bic, rbind)[,2], rbind) %>%
+  mutate(Subject = rep(1:15, each = 6),
+         Block = rep(1:6, 15)) %>%
+  pivot_longer(contains("model"), names_to = "model", values_to = "weight") %>%
+  mutate_at(c("Subject", "Block", "model"), as.factor) %>%
+  ggplot(aes(x = model, y = Subject)) + geom_tile(aes(fill = weight)) +
+  facet_wrap(vars(Block), ncol = 2, labeller = label_both) +
+  scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
+  scale_y_discrete(name = "Subject") +
+  scale_fill_distiller(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), 
+                       palette = "Spectral")
+
+```
+
+(ref:plot-schwarz-weights-task4) Schwarz weights displayed for each subject and HMMs with different numbers of states.
+Models were applied to task 4 of the Ehinger et al.
+(2019) data set.
+Higher weights indicate better model fit.
+Grey tiles indicate erroneous model fits.
+
+```{r plot-schwarz-weights-task4, fig.cap="(ref:plot-schwarz-weights-task4)", fig.height=8}
+
+print(bic.plot[[4]])
+
+```
+
+(ref:plot-schwarz-weights-task5) Schwarz weights displayed for each subject and HMMs with different numbers of states.
+Models were applied to task 5 of the Ehinger et al.
+(2019) data set.
+Higher weights indicate better model fit.
+Grey tiles indicate erroneous model fits.
+
+```{r plot-schwarz-weights-task5, fig.cap="(ref:plot-schwarz-weights-task5)", fig.height=8}
+
+print(bic.plot[[5]])
 
@@
 (ref:plot-icl-weights-img) ICL weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the image condition of the @Andersson2017 data set.
+Models were applied to the image condition of the Andersson et al.
+(2017) data set.
 Higher weights indicate better model fit.
 
-```{r plot-icl-weights-img, results='hold', fig.cap="(ref:plot-icl-weights-img)"}
+```{r plot-icl-weights-img, fig.cap="(ref:plot-icl-weights-img)"}
 
@@
 (ref:plot-icl-weights-dots) ICL weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the moving dots condition of the @Andersson2017 data set.
+Models were applied to the moving dots condition of the Andersson et al.
+(2017) data set.
 Higher weights indicate better model fit.
 
-```{r plot-icl-weights-dots, results='hold', fig.cap="(ref:plot-icl-weights-dots)"}
+```{r plot-icl-weights-dots, fig.cap="(ref:plot-icl-weights-dots)"}
 
@@
 (ref:plot-icl-weights-vid) ICL weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the video condition of the @Andersson2017 data set.
+Models were applied to the video condition of the Andersson et al.
+(2017) data set.
 Higher weights indicate better model fit.
 
-```{r plot-icl-weights-vid, results='hold', fig.cap="(ref:plot-icl-weights-vid)"}
+```{r plot-icl-weights-vid,  fig.cap="(ref:plot-icl-weights-vid)"}
 
@@
 
-(ref:plot-example-classification) Example data from @Andersson2017 displayed as x-, and y-coordinates, velocity (in deg/s), acceleration (in deg/s$^2$), and sample-to-sample (relative) angle (in rad).
+(ref:plot-example-classification) Example data from Andersson et al.
+(2017) displayed as x-, and y-coordinates, velocity (in deg/s), acceleration (in deg/s), and sample-to-sample (relative) angle (in rad).
 Colors indicate classification by gazeHMM and symbols by human coder MN.
@@
 
-```{r plot-example-classification, results='hold', fig.cap="(ref:plot-example-classification)", fig.height=8}
+```{r plot-example-classification, fig.cap="(ref:plot-example-classification)", fig.height=7.5}
 
@@
 
-(ref:tab-example-trans) Transition probabability estimates for example data from @Andersson2017
+(ref:tab-example-trans) Transition probabability estimates for example data from Andersson et al. (2017)
 (ref:tab-example-trans-note) Rows indicate from which state and columns into which state the model transitions.
 
-```{r tab-example-trans, results='hold'}
+```{r tab-example-trans}
 
@@
 
-(ref:tab-example-resp) Response parameter estimates for example data from @Andersson2017
+(ref:tab-example-resp) Response parameter estimates for example data from Andersson et al. (2017)
 (ref:tab-example-resp-note) Con = concentration parameter of the von-Mises distribution.
 
-```{r tab-example-resp, results='hold'}
+```{r tab-example-resp}
 

### 2020-06-17 manuscript/report_technical_appendix.Rmd
add Ehinger et al (2019) results

@@
     i. The coordinates are centered ([0,0] is center)
-    ii. $\theta_{deg} = atan(\frac{x*res_x}{d_{screen}*dim_x})\frac{180}{\pi}$
+    ii. $\theta_{deg} = atan(\frac{x \cdot res_x}{d_{screen} \cdot dim_x})\frac{180}{\pi}$
 (@) Samples that are outside the screen are marked as invalid.
-(@) Velocities and accelerations for the x- and y-coordiantes are calculated using a Savitzky-Golay filter (3rd order, length = 5).
-(@) Combined velocities and accelerations are calculated as: $v = f \sqrt{v_x^2 + v_y^2}$ and $a = f \sqrt{a_x^2 + a_y^2}$ with $f$ as the sampling rate.
-(@) Velocities and accelerations of value zero are nudged to be 0.01.
-(@) Samples with velocity above 1000 deg/s or acceleration above 100000 deg/s$^2$ are marked as invalid.
+(@) Velocities and accelerations for the x- and y-coordiantes are calculated using an SG filter (3rd order, length = 5).
+(@) Combined velocitiy and acceleration are calculated as: $v = f \sqrt{v_x^2 + v_y^2}$ and $a = f \sqrt{a_x^2 + a_y^2}$, with $f$ as the sampling rate.
+Note that acceleration is only multiplied by the sampling rate once.
+The second multiplication is done during postprocessing.
+(@) Velocity and acceleration of value zero are nudged to be 0.01.
+(@) Samples with velocity above 1000 deg/s or acceleration above 100 deg/s are marked as invalid.
 (@) The sample-to-sample angle is calculated as: 
@@
         f. Avg. velocity
-        g. Max. acceleration $\cdot 1000$
-        h. Avg. acceleration $\cdot 1000$
+        g. Max. acceleration $\times f$
+        h. Avg. acceleration $\times f$
         i. Direction: $\theta = atan2(\frac{end_y - start_y}{end_x - start_x})$

### 2020-06-17 validation/validation_Andersson2017.R

@@ -8,6 +8,7 @@ library(depmixS4)
 library(parallel)
 library(R.matlab)
 library(psych)
+library(signal)
 
 source("algorithm/gazeHMM.R")
 source("algorithm/model_helper_functions.R")
@@ -104,7 +105,7 @@ fr <- 500
 
 ncores <- detectCores()
 clust <- makeCluster(ncores)
-clusterExport(clust, list("gazeHMM", "onestate_HMM", "preprocess", "res", "dim", "dist", "fr", "A2017"))
+clusterExport(clust, list("gazeHMM", "onestate_HMM", "preprocess", "butter", "filter", "res", "dim", "dist", "fr", "A2017"))
 
 A2017.fit <- lapply(1:length(A2017), function(stim) parLapply(clust, 1:length(A2017[[stim]]), function(subj) {
   

### 2020-06-17 validation/validation_Ehinger2019.R

@@ -98,24 +98,27 @@ fr <- 500
 
 ncores <- detectCores()
 clust <- makeCluster(ncores)
-clusterExport(clust, list("classify_gaze_data", "onestate_HMM", "preprocess", "res", "dim", "dist", "fr", "E2019"))
+clusterExport(clust, list("gazeHMM", "onestate_HMM", "preprocess", "res", "dim", "dist", "fr", "E2019"))
 
 E2019.fit <- parLapply(clust, E2019, function(subj) lapply(subj, function(task) lapply(task, function(df) {
   
   df$time <- (df$time - df$time[1])/1e3
-
+  
+  trstart <- matrix(0.1/(k-1), k, k)
+  diag(trstart) <- 0.9
   
   fit <- list()
   
   fit[["1"]] <- try(onestate_HMM(preprocess(x = df$xp, y = df$yp, t = df$time, unit = "px",
                                             res = res, dim = dim, dist = dist, fr = fr, blink = is.na(df$ps),
-                                            sg.order = 3, sg.length = 5)))
+                                            sg.order = 3, sg.length = 5, sf = c(100, 100))))
   
   for (k in 2:5) {
 
-    fit[[as.character(k)]] <- try(classify_gaze_data(x = df$xp, y = df$yp, t = df$time, unit = "px",
-                                                 res = res, dim = dim, dist = dist, fr = fr, blink = is.na(df$ps),
-                                                 nstates = k))
+    fit[[as.character(k)]] <- try(gazeHMM(x = df$xp, y = df$yp, t = df$time, unit = "px",
+                                          res = res, dim = dim, dist = dist, fr = fr, blink = is.na(df$ps),
+                                          nstates = k, trstart = trstart, 
+                                          random.respstart = F))
 
   }
   
@@ -125,3 +128,15 @@ E2019.fit <- parLapply(clust, E2019, function(subj) lapply(subj, function(task) 
 stopCluster(clust)
 
 save("E2019.fit", file = "validation/Ehinger2019_fitted.Rdata")
+
+
+# Compare models with different states
+
+schwarz.weights <- function(bic, na.rm = T) {
+  
+  d.bic <- bic - min(bic, na.rm = na.rm) # eq 2
+  
+  exp(-0.5 * d.bic)/sum(exp(-0.5 * d.bic), na.rm = na.rm) # eq 4
+  
+}
+

### 2020-06-21 manuscript/internship_report.Rmd
incorporate feedback

@@
   - name          : "Malte Valentin Lueken"
-    affiliation   : "1"
-    address       : "Speelmanstraat 7-3 1063ZC Amsterdam"
-    email         : "malte.luken@student.uva.nl"
-
-affiliation:
-  - id            : "1"
-    institution   : "University of Amsterdam"
 
 authornote: |
-  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-
-  Enter author note here.
+  Affiliation: University of Amsterdam, Graduate School of Psychology, Department of Methodology and Statistics.
+
+  This manuscript constitutes the internship report as part of the Research Master's Psychology.
+
+  Student number: 12750166. Email address: malte.luken@student.uva.nl.
+
+  Specialization: Methodology and Statistics. Supervisor: dhr. dr. Ingmar Visser.
+
+  I would like to thank Simon Kucharsky and Ingmar Visser for their helpful advice on developing and implementing this research project.
 
 abstract: |
-  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
-
-  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
-
-  One sentence clearly stating the **general problem** being addressed by  this particular study.
-
-  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
-
-  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
-
-  One or two sentences to put the results into a more **general context**.
-
-  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
-
-  <!-- https://tinyurl.com/ybremelq -->
+  Eye-tracking allows researchers to infer cognitive processes from eye movements that are classified into distinct events.
+Parsing the events is typically done by algorithms that transform, filter, classify, and merge raw data samples.
+Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere weaknesses.
+I developed gazeHMM, an algorithm that improves on previous HMM algorithms.
+The development was guided by the question of whether HMMs are useful at describing eye movements and whether they improve the event classification.
+A simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences with only few exceptions.
+When applied to benchmark data, model comparisons for gazeHMM with different events did not match theoretical expectations.
+Compared to other algorithms, gazeHMM improved the event classification which was assessed by the similarity to human coding.
+The classification of smooth pursuits was found to be inadequate and is thus only recommended for exploration.
+Otherwise, I consider gazeHMM as appropriate for practical applications.
+In sum, gazeHMM improves classification but is not accurately describing eye movements, yet.
+I identify several explanations and discuss potential remedies for this result.
 
@@
 ---
+
+# Introduction
+For how long did humans seek to open a window into each other's minds, revealing their hidden cognitive processes?
+While many methods aiming to achieve this have been developed in the last decades, eye-tracking as one of them has become more and more reliable, accurate, and widely applied in cognitive psychology [@Duchowski2017].
+Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017].
+Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015].
+Commonly distinguished events are fixations, saccades, smooth pursuit, and post-saccadic oscillations (PSOs).
+Classifying eye-tracking data reduces their complexity and is the first step towards cognitive interpretation [@Andersson2017; @Salvucci2000].
+The classification is normally done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017].
+@Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
+As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most classification algorithms consist of four steps: First, the raw gaze position is transformed into a metric of position and/or time.
+Second, a filtering or smoothing procedure is applied to the data to separate the gaze signal from noise and artifacts [@Spakov2012].
+Third, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
+Fourth, the algorithm decides which candidates to accept, relabel, or merge [@Hessels2017; @Komogortsev2010].
+A frequently implemented method uses probabilistic models to classify events.
+Instead of discrete classification, these models assign probabilities for belonging to an event to each sample.
+Probabilistic models have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
+324].
+One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs).
+Figure \@ref(fig:plot-hmm-structure) illustrates the structure of HMMs: Parallel to the gaze signal evolves a sequence of distinct states that cannot be directly observed.
+Each gaze sample depends on its corresponding state.
+Each state depends on the previous but not on earlier states of the sequence [@Cappe2005; @Visser2019].
+In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as events and gaze data are dependent time series.
+On this basis, several classification algorithms using HMMs have been developed.
+One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named identification by HMM [I-HMM]).
+Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM.
+Recently, @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR).
+The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and angle (named NSLR-HMM).
+Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named Bayesian microsaccade detection [BMD]).
+Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).
+Several studies have tested the performance of HMM algorithms against other classification methods: I-HMM has been deemed as robust against noise, behaviorally accurate, and showing a high sample-to-sample agreement with a human coder [@Andersson2017; @Komogortsev2010; @Salvucci2000].
+However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012].
+NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019, @Pekkanen2017].
+In sum, HMMs seem to be a promising method for classifying eye movements.
+Still, the existing HMM algorithms each have their weaknesses that could be improved upon (and that could explain their inferior performance in comparison to other methods).
+First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
+Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability.
+It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals.
+Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017].
+The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events.
+Therefore, the authors suggest using it as an exploratory tool followed by further event classification.
+HMMs can also be used as a generative model to make inferences about the underlying processes of eye movements, a property that has been rarely used in the context of classification [cf.
+@Mihali2017].
+For example, an HMM with four events could be compared to one with three events to see whether an event is present in the data.
+I propose to develop an algorithm that uses HMMs to classify eye movement data. It will address the weaknesses of previous HMM algorithms and utilize the generative property of HMMs.
 
@@
 library(here)
+library(grid)
+library(gridExtra)
+library(DiagrammeR)
+library(DiagrammeRsvg)
+library(magrittr)
+library(rsvg)
 source(here("algorithm/preprocessing_helper_functions.R"))
@@
 ```
+
+(ref:plot-algorithm-workflow) Example workflow for eye movement event classification algorithms: (a) the raw gaze signal is transformed into a velocity signal (in deg/s); (b) the velocity signal is smoothed and filtered; (c) samples are labeled as events (indicated by colors), and (d) relabeled.
+Sequences of samples belonging to the same event are merged (indicated by black segments).
+
+```{r plot-algorithm-workflow, fig.cap="(ref:plot-algorithm-workflow)", fig.height = 8}
+
+set.seed(126)
+
+velRaw <- sample(c(rgamma(25, 3, 1/6), rgamma(7, 25, 1/5), rgamma(8, 15, 1/5)), 40)
+
+sample <- 1:length(velRaw)
+
+velSmooth <- as.numeric(smooth(velRaw, kind = "3R"))
+
+eventPre <- ifelse(velSmooth > 40, "s", "f")
+
+eventPost <- ifelse(velSmooth < 60, "f", eventPre)
+
+df <- data.frame(sample, velRaw, velSmooth, eventPre, eventPost)
+
+# Create plot
+
+p1 <- ggplot(df, aes(sample, velRaw)) + geom_point()
+
+p2 <- ggplot(df, aes(sample, velSmooth)) + geom_point()
+
+p3 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPre))
+
+p4 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPost))
+
+plotList <- list(p1, p2, p3, p4)
+
+for(p in 1:length(plotList)) {
+
+  plotList[[p]] <- plotList[[p]] + scale_y_continuous(name = "Velocity\n(deg/s)", 
+                                                      limits = c(0, 200), breaks = c(0, 100, 200)) + 
+    xlab("") + geom_line(color = "grey") + ggtitle(letters[p]) + 
+    theme(legend.position = "none")
+
+}
+
+plotList[[4]] <- plotList[[4]] + xlab("Sample") +
+  geom_segment(aes(x = 1, y = 180, xend = 4, yend = 180), size = 1.2) +
+  geom_segment(aes(x = 5, y = 180, xend = 9, yend = 180), size = 1.2) +
+  geom_segment(aes(x = 10, y = 180, xend = 27, yend = 180), size = 1.2) +
+  geom_segment(aes(x = 28, y = 180, xend = 29, yend = 180), size = 1.2) +
+  geom_segment(aes(x = 30, y = 180, xend = 40, yend = 180), size = 1.2)
+
+grid.arrange(plotList[[1]], plotList[[2]], plotList[[3]], plotList[[4]], ncol = 1)
+
+```
+
+(ref:plot-hmm-structure) Example structure of HMMs: The observed velocity signal *V* depends on hidden states *S* (displayed for six samples).
+Each state depends on the previous state.
+Colors indicate different states.
+Figure inspired by Visser and Speekenbrink [-@Visser2019, p.
+133].
+
+```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)"}
+
+grViz("
+  digraph basic_hmm {
+
+    graph [overlap = true, fontsize = 10, fontname = 'Times-Roman']
+
+    node [shape = square, fixedsize = true, width = 0.5, penwidth = 1.5]
+    Y1 [label = '<I>V@_{@@1-1}</I>']
+    Y2 [label = '<I>V@_{@@1-2}</I>']
+    Y3 [label = '<I>V@_{@@1-3}</I>']
+    Y4 [label = '<I>V@_{@@1-4}</I>']
+    Y5 [label = '<I>V@_{@@1-5}</I>']
+    Y6 [label = '<I>V@_{@@1-6}</I>']
+
+    node [shape = circle, fixedsize = true, width = 0.5]
+    S1 [label = '<I>S@_{@@1-1}</I>', color = '@@2-1']
+    S2 [label = '<I>S@_{@@1-2}</I>', color = '@@2-2']
+    S3 [label = '<I>S@_{@@1-3}</I>', color = '@@2-3']
+    S4 [label = '<I>S@_{@@1-4}</I>', color = '@@2-4']
+    S5 [label = '<I>S@_{@@1-5}</I>', color = '@@2-5']
+    S6 [label = '<I>S@_{@@1-6}</I>', color = '@@2-6']
+
+    edge [minlen = 2, arrowsize = 0.7]
+    S1->{S2 Y1}; S2->{S3 Y2}; S3->{S4 Y3}; S4->{S5 Y4}; S5->{S6 Y5}; S6->Y6;
+
+    {rank = same; S1 S2 S3 S4 S5 S6}
+    {rank = same; Y1 Y2 Y3 Y4 Y5 Y6}
+
+  }
+
+  [1]: 1:6
+  [2]: rep(c('#F8766D','#00BFC4'), each = 3)
+", width = 300, height = 300)
+
+```
+
+## Research Question
+Can HMMs improve describing and classifying eye-tracking data?
+This is the general research question of my proposal and can be specified as following: Are HMMs useful generative models for recorded eye movements that improve classifying them into events?
+This research question will guide the development of the algorithm and leads to two hypotheses: First, I predict that HMMs are useful generative models for eye movements.
+This hypothesis will be supported if the algorithm yields events that match theoretical expectations for benchmark data.
+Other results will be treated as evidence against the hypothesis.
+Second, I hypothesize that HMMs will improve the classification performance of eye movements.
+I will treat it as support for this hypothesis if the algorithm outperforms algorithms using other methods when applied to benchmark data.
+I will treat other results as evidence against the hypothesis.
+
+# Operationalization
+The proposed algorithm aims to improve the drawbacks of previous HMM algorithms.
+Therefore, its classification method is oriented to four major goals: (a) it does not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it covers the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it confirms rather than explores the presence of events in the data.
+Besides implementing the algorithm in the statistical programming language R @RCoreTeam2020, I will develop its concrete technical design exploratorily.
+The first step will be choosing an appropriate metric that allows separating the gaze data into events.
+@Zemblys2018 provide an overview of different gaze data metrics and compare how well they predict human-labeled events.
+The next step will choosing a filtering/smoothing method to reduce the noise in the gaze data [for a comparison of different gaze data filters, see @Spakov2012].
+The criteria for choosing a metric and a filter will be its accuracy and not relying on user parameter settings.
+The generative part of the algorithm will allow users to build an HMM which matches their theoretical assumptions about the gaze data.
+The algorithm will take a preprocessed gaze data metric (e.g., filtered velocity) as input and optimize the model parameters.
+The output of the model will be probabilities for hidden states.
+The states can be interpreted as event labels for each sample.
+For instance, the model could assign to a sample a probability of 0.6 for belonging to a fixation and 0.4 for belonging to a saccade.
+These probabilities can be used to compute the hidden state sequence (e.g., by choosing events with the highest probability).
+Moreover, the model will output the parameters of the distribution of labeled samples for each event.
+It will allow generating samples from these distributions to simulate gaze data.
+Before testing the hypotheses, I will conduct a simulation study to estimate the parameter recovery of the generative model: Artificial eye movement data will be generated by the model.
+The parameters of the model will be varied across sensible ranges.
+Then, the model will be applied to the data.
+Finally, the estimated results of the model will be compared to the true parameters that have been used to generate the data.
+The first hypothesis will be tested by comparing different generative models on benchmark data.
+The models will differ in their assumed number of events in the data.
+The model comparison will result in a preferred model which scores best on the evaluation criterium.
+The preferred model is expected to match the theoretically assumed number of present events in the benchmark data.
+To test the second hypothesis, the proposed algorithm will be compared against other algorithms that have also been applied to benchmark data.
+The performance of the proposed algorithm will be assessed by metrics that are commonly used in classification (e.g., Cohen's Kappa).
+
+## Sample Characteristics
+The hypotheses will be tested on two benchmark data sets that have been published in previous studies.
+The first set was published in @Andersson2017 and contains eye-tracking data from 17 subjects.
+Three types of stimuli were presented on a screen ($1024 \times 768$ pixels) to each subject: A static image, a linearly moving dot, and a video.
+Subjects were instructed to look freely at the images and follow the movements of the dots and the objects in the videos.
+Their eye movements were recorded with a Hi-Speed 1250 tower-mounted eye-tracker (SensoMotoric Instruments) and a sampling rate of 500 Hz.
+Only data from the right eye were used.
+The data were labeled by two human experts in eye movement research as fixations, saccades, smooth pursuits, PSOs, or blinks [@Andersson2017, p.
+621].
+The second data set, published by @Ehinger2019, consists of eye-tracking data from 15 subjects who performed six blocks of 10 different experimental tasks.
+Here, I will only consider data from Tasks 4 and 5 because they are qualitatively different from the first data set.
+In Task 4, subjects were instructed to fixate a central target on the screen ($1920 \times 1080$ pixels) for 20 s.
+In Task 5, subjects also fixated a central target but were instructed to blink each time they heard a beep.
+Seven beeps with a duration of 100 ms and a 1.5 s interval in between were presented.
+For both tasks, eye-movements were recorded by two eye-trackers simultaneously (EyeLink 1000, SR Research; and Pupil Labs glasses, Pupil Labs) with a sampling rate of 250-500 Hz and 240 Hz (interpolated), respectively [@Ehinger2019, pp.
+3-11].
+
+## Data Analysis
+I will treat the parameter recovery of the generative model as satisfactory when it yields a 95% CI covering the true parameters.
+The median rate of correct classifications should be a minimum of 90%.
+The first hypothesis will be tested using a model comparison approach: For both data sets, models with one, two, three, four, and five eye movement events plus a noise event will be compared.
+The noise event will represent blinks.
+For the first data set and static images, I expect the model with three events (excl.
+noise) to be the preferred model since smooth pursuit movements are not assumed to be present.
+For moving dots and video stimuli, I expect the model with four events (excl.
+noise) to be preferred, since smooth pursuits are expected in the data.
+For the second data set, I expect a model with one event (excl.
+noise) to be preferred because only fixations should be present in the data from both tasks.
+For all comparisons, the preferred model will be chosen according to the highest Schwarz weight [@Wagenmakers2004].
+I will test the second hypothesis using the metrics that have been used in the original study of the first data set: @Andersson2017 compared ten algorithms to human-labeled data by the root mean squared error (RMSE) of event distribution descriptives, the sample-to-sample agreement indicated by Cohen's Kappa, and the disagreement ratio across all samples indicated by the confusion matrix.
+I expect the algorithm with a generative model assuming four events (excl.
+noise) to achieve either a lower RMSE, higher Cohen's Kappa, or lower disagreement ratio than all the other algorithms which are evaluated in the study.
+
+## Intended Results
+Preferred generative models that match the theoretical assumptions on the number of events in the data would support the notion that HMMs describe gaze data appropriately.
+If a generative model with fewer events than expected is preferred, this would mean that the model is not able to distinguish the events well enough.
+If more events than expected are preferred, this would indicate that the model is describing noise or an unexpected event is present in the data.
+Outperforming other algorithms in agreement to human-coded data would demonstrate that the proposed algorithm classifies events similar to human coders.
+However, @Hooge2018a observed considerable differences in how humans code gaze data.
+They concluded that human coders cannot be seen as a gold standard for event classification.
+The proposed algorithm would be the first allowing the user to conduct model comparisons as part of the classification.
+This comparison could be used to investigate the data more thoroughly on a qualitative (e.g., assumed present events) or quantitative (e.g., event sample distributions) level.
+In contrast to many other classification algorithms, the proposed algorithm would not only classify gaze data but also explain why it is classified this way by using a generative model.
+This outstanding property of the algorithm would guarantee its competitiveness against other popular algorithms, especially the recently developed machine learning approaches which do not make generative assumptions [e.g., @Bellet2019].
+It could spark a shift from pure classification output towards understanding eye movements on a more fundamental level.
 
@@
 PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013].
-They are specifically distinguished by their change in direction clustered around 180Ã‚Â° [@Pekkanen2017].
+They are specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
 Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
@@
 It has to be noted that the gamma was choosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
-To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution.
-Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$) which should lead to rather symmetric distributions.
-Moreover, the uniform distribution can distinguish fixations from the other events.
-The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, only parameter intercepts were estimated.
+To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution.
+Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions.
+Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution.
+Thus, the uniform distribution should distinguish fixations from the other events.
+The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, no time-varying covariates were included.
 
@@
 In gazeHMM, I assume data to be missing at random.
-Most of the missing data in eye movement classification are due to blinks.
-The user can indicate in gazHMM which samples should be labeled as blinks (other missing samples are treated as noise).
+However, most of the missing data in eye movement classification are due to blinks.
+Therefore, the user can indicate in gazHMM which samples should be labeled as blinks (other missing samples are treated as noise).
 Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs.
@@
 The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability.
-Parameters of the response distributions (except for the uniform distribution) were optimized on the log-scale (except for the mean parameter of the von-Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein steplengths [@Barzilai1988].
+Parameters of the response distributions (except for the uniform distribution) were optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein steplengths [@Barzilai1988].
 
@@
 The left over probability mass was split evenly between the probabilites for switching to a different state (per row of the transition matrix).
-Moreover, I did not modify the mean parameters of the von-Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
+Moreover, I did not modify the mean parameters of the von Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
 I defined approximate ranges for each response variable and chose true parameter intervals and default values so that they produced samples that roughly corresponded to these ranges.
@@
 The scale parameters were set so that the respective distribution approximately matched the assumed ranges.
-Since the concentration parameters of the von-Mises distribution are the inverse of standard deviations, they were varied on the inverse scale.
+Since the concentration parameters of the von Mises distribution are the inverse of standard deviations, they were varied on the inverse scale.
+An example of simulated data from the HMM with default parameters is visualized in Figure \@ref(fig:plot-sim-ex).
 
@@
 This procedure ensured that velocity and acceleration values remained positive and were drawn from moderately skewed distributions with modes equal to the original values.
-To angle, I added white noise from a von-Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
+To angle, I added white noise from a von Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
 $\tau_{noise}$ and $\kappa_{noise}$ were varied simultaneously in equidistant steps in their intervals.
@@
 This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
-Mean parameters of the von-Mises distribution always started at their true values.
-
-## Data Analysis
-For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values[^1]: $$RMdSPD = \sqrt{Med((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2)}.$$
-I treated $RMdSPD < 0.1$ as good, $0.1 \le RMdSPD < 0.5$ as moderate, and $RMdSPD \ge 0.5$ as bad recovery of a parameter.
+Mean parameters of the von Mises distribution always started at their true values.
+
+## Data Analysis [^6]
+For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values[^1]: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
+I treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter.
 By taking the median, I reduced the influence of potential outliers in the estimation and using proportions enabled me to compare RMdSPD values across parameters and data sets.
@@
 
-[^1]: This measure is only appropriate when $\psi_{true} \ne 0$. This was not the case for some mean parameters of the von-Mises distributions. In those cases, I used $\psi_{true} = 2\pi$ instead.
+[^6]: Note that this section deviates from my proposed analysis plan. However, since I preregistered the changed analysis plan, I treated the results of the simulation as confirmatory.
+[^1]: This measure is only appropriate when $\psi_{true} \ne 0$. This was not the case for some mean parameters of the von Mises distributions. In those cases, I used $\psi_{true} = 2\pi$ instead.
 
 (ref:sim-ranges) Approximate ranges of response variables used to generate parameter values
-(ref:sim-ranges-note) Units are Ã‚Â°/s (velocity), Ã‚Â°/s^2^ (acceleration), and radians (angle).
+(ref:sim-ranges-note) Units are deg/s (velocity), deg/s^2^ (acceleration), and radians (angle).
 ~ indicates that the distribution has a peak at this value.
@@
 
-# Exploratory Results
+# Results [^5]
 ## Simulation Study
@@
 In sum, the HMM with two states recovered parameters and hidden states very well and outliers only occured rarely.
+
+[^5]: The design and analysis plan of the simulation study have not been included in my proposal.
+However, since I preregistered both the design and analysis, I treated the results as confirmatory.
+In general, the evaluation of an algorithm can hardly be proposed in detail before the algorithm has been developed.
+Therefore, I chose to report a general results section.
+Nevertheless, I indicated whenever results were obtained through exploratory analyses that were not planned beforehand.
 
@@
 For the data set by @Andersson2017, the algorithm was successfully fitted for every participant and condition.
+An example for event classification by gazeHMM is displayed in Figure \@(fig:plot-example-classification) and examples for parameter estimates are presented in Tables \@ref(tab:tab-example-trans) and \@ref(tab:tab-example-resp).
 
@@
 [^4]: Even though I proposed to compute the *disagreement ratio* and @Andersson2017 also used that term, the comparison technically involves percentages and not ratios.
-Thus, I decdided to use only the term *disagreement* instead.
+Thus, I decided to use only the term *disagreement* instead.
 
@@
 In this report, I presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events.
-The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von-Mises and a uniform distribution.
+The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von Mises and a uniform distribution.
 An HMM classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
@@
 I applied gazeHMM with different numbers of states to benchmark data and compared the model fit, showing that a five-state HMM had consistently most likely generated the data.
-This result speaks against my first hypothesis that HMMs are useful generative models for eye movements.
+Thus, the model comparison provided evidence against my first hypothesis that the events in gazeHMM will match theoretical expectations.
 When comparing gazeHMM with four events against other algorithms and human coders, it did not show the best performance for the duration and number of classified events nor the sample-to-sample agreement across all conditions.
@@
 With three states, gazeHMM had a lower overall disagreement than all other algorithms.
-I treat this pattern as moderate evidence for my second hypothesis that HMMs improve the classification of eye movements.
-Taken together, I only found some evidence for one part of my research question: HMMs improve classifying gaze data but not describing them.
+I treat this pattern as moderate evidence for my second hypothesis that gazeHMM improves the classification of eye movements.
+Taken together, I only found some evidence for one part of my research question, that gazeHMM improves classifying gaze data but not describing them.
 Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle.
@@
 The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).
-The model comparison on the benchmark data clearly provided evidence against HMMs being useful generative models for eye movement data.
-Even using a model selection criterion that identified correct models succesfully in previous model comparison studies [@Pohle2017], did not yield the expected pattern.
+The model comparison on the benchmark data clearly provided evidence against gazeHMM being an accurate generative model for eye movement data.
+Even using a model selection criterion that identified correct models succesfully in previous studies [@Pohle2017], did not yield the expected pattern.
 There are several explanations for this result:
@@
 The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events.
-Of course, that makes it still a useful tool for classification.
 Nevertheless, Cohen's kappa values of 0.72 (saccades - video - four states) or 0.68 (fixations - image - three states) indicate substantial agreement to human coders, especially in light of the maximum reference of ~0.8 from the simulation study.
-At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018 observed substantial differences between coders and within coders over time.
+At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
 Despite these differences, they recommend comparisons with human coding to demonstrate the performance of new algorithms and finding errors in their design.
@@
 ## Advantages of gazeHMM
-In view of the four proposed goals that gazeHMM should fulfill, I can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they have only minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
-Their default values should be appriate for most applications.
+In view of the four proposed goals that gazeHMM should fulfill, I can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they only have minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
+Their default values should be appropriate for most applications.
 A major advantage of gazeHMM is that it does not require human labeled data as input.
 Instead, it estimates all parameters and hidden states from the data.
-Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
+Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018a], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
 This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data.
@@
 This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.
-Third, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data.
+Third, one solution to diverging results when comparing gazeHMM with different events could be model averaging: Instead of using the maximum posterior state probability of each sample from the preferred model, the probabilities could be weighted according to a model selection criterion (e.g., Schwarz weight) and averaged.
+Then, the maximum averaged probability could be used to classify the samples into events.
+This approach could lead to a more robust classification because averaging might reduce the inaccuracy of each competing model.
+However, the model comparison for gazeHMM often showed extreme weights for a five-state model, which would lead to a very limited influence of the other models in the averaged probabilities.
+Fourth, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data.
 As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements.
 As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
-Fourth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonable short saccades, gazeHMM could explicitly model the duration of events.
+Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and auto-correlation functions [see @Zucchini2016].
+Fifth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonably short saccades, gazeHMM could explicitly model the duration of events.
 This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2006].
@@
 
-## Conclusions
+## Conclusion
 Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events.
@@
 On the other hand, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events.
-These difficulties highlight general challenges for HMMs in general, but also specifically for eye movements.
-Revisting my research question, I conclude that HMMs are suitable models for classifying eye movement events but still lack the ability to accurately describe them.
+These difficulties highlight challenges for HMMs in general, but also specifically for eye movements.
+Revisting my research question, I conclude that gazeHMM is suitable for classifying eye movement events but still lacks the ability to accurately describe them.
+
+\newpage
 

### 2020-06-22 manuscript/report_technical_appendix.Rmd
add blink window parameter

@@
 Input: x-coordinate, y-coordinate, timestamp, unit of coordinates (px vs.
-deg), screen resolution, screen dimensions, distance to screen, sampling rate, blink indicator, order of SG filter, length of SG filter, number of states in the HMM, start values for the HMM, downsampling factor, random response start values, seed for random response start values, control options for fitting the HMM, minimum saccade duration.
+deg), screen resolution, screen dimensions, distance to screen, sampling rate, blink indicator, blink window, order of SG filter, length of SG filter, number of states in the HMM, start values for the HMM, downsampling factor, random response start values, seed for random response start values, control options for fitting the HMM, minimum saccade duration.
 
@@
 (@) Samples that match the blink indicator are labeled as blinks.
+(@) All samples within +/- *blinkwindow* around blink indicated samples are marked as invalid and labeled as blinks.
 (@) If the unit is px, the x- and y-coordinates are transformed into va with: 

### 2020-06-22 algorithm/gazeHMM.R

@@ -4,7 +4,7 @@
 # Date: 24.04.2020
 
 
-gazeHMM <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
+gazeHMM <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL, b.win,
                     sg.order = 3, sg.length = 5, 
                     nstates, respstart, trstart, instart, sf = c(10, 10), 
                     random.respstart = T, start.seed = NULL,
@@ -102,6 +102,8 @@ gazeHMM <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
        "'blink' must be either a numeric vector of length 2 or a logical vector with the same length as 't'"
        )
   
+  if(!(is.numeric(b.win) && length(b.win) == 1 && b.win >= 0)) stop("'b.win' must be a single non-negative value")
+  
   if(!is.integer(nstates)) stop("'nstates must be integer'")
   if(length(nstates) != 1) stop("'nstates' must be a single value")
   if(nstates < 2 || nstates > 5) stop("'nstates' must have a value between 2 and 5")
@@ -156,7 +158,7 @@ gazeHMM <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
   
   # Preprocessing
   
-  prep <- preprocess(x, y, t, unit, res, dim, dist, fr, blink, sg.order, sg.length)
+  prep <- preprocess(x, y, t, unit, res, dim, dist, fr, blink, b.win, sg.order, sg.length)
   
   
   # Model classification

### 2020-06-22 algorithm/preprocessing.R

@@ -4,7 +4,7 @@
 # Date: 07.04.2020
 
 
-preprocess <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
+preprocess <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL, b.win = 0.05,
                        sg.order, sg.length) {
   
   require(signal)
@@ -97,8 +97,8 @@ preprocess <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL,
   for(i in 1:length(inib)) {
     if(!is.na(inib[i])) {
       
-      valid[t > (t[i]-0.05) & t <= (t[i]+0.05)] <- F
-      label[t > (t[i]-0.05) & t <= (t[i]+0.05)] <- 0
+      valid[t > (t[i]-b.win) & t <= (t[i]+b.win)] <- F
+      label[t > (t[i]-b.win) & t <= (t[i]+b.win)] <- 0
       
     }  
   }

### 2020-06-22 manuscript/internship_report.Rmd
update report

@@
 When applied to benchmark data, model comparisons for gazeHMM with different events did not match theoretical expectations.
-Compared to other algorithms, gazeHMM improved the event classification which was assessed by the similarity to human coding.
+Compared to other algorithms, gazeHMM improved the event classification which was assessed by the similarity to human coding, especially for dynamic stimuli.
 The classification of smooth pursuits was found to be inadequate and is thus only recommended for exploration.
@@
 
-keywords          : "keywords"
-wordcount         : "X"
+keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
+wordcount         : "12808"
 
@@
 However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012].
-NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019, @Pekkanen2017].
+NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].
 In sum, HMMs seem to be a promising method for classifying eye movements.
@@
 
-grViz("
-  digraph basic_hmm {
-
-    graph [overlap = true, fontsize = 10, fontname = 'Times-Roman']
-
-    node [shape = square, fixedsize = true, width = 0.5, penwidth = 1.5]
-    Y1 [label = '<I>V@_{@@1-1}</I>']
-    Y2 [label = '<I>V@_{@@1-2}</I>']
-    Y3 [label = '<I>V@_{@@1-3}</I>']
-    Y4 [label = '<I>V@_{@@1-4}</I>']
-    Y5 [label = '<I>V@_{@@1-5}</I>']
-    Y6 [label = '<I>V@_{@@1-6}</I>']
-
-    node [shape = circle, fixedsize = true, width = 0.5]
-    S1 [label = '<I>S@_{@@1-1}</I>', color = '@@2-1']
-    S2 [label = '<I>S@_{@@1-2}</I>', color = '@@2-2']
-    S3 [label = '<I>S@_{@@1-3}</I>', color = '@@2-3']
-    S4 [label = '<I>S@_{@@1-4}</I>', color = '@@2-4']
-    S5 [label = '<I>S@_{@@1-5}</I>', color = '@@2-5']
-    S6 [label = '<I>S@_{@@1-6}</I>', color = '@@2-6']
-
-    edge [minlen = 2, arrowsize = 0.7]
-    S1->{S2 Y1}; S2->{S3 Y2}; S3->{S4 Y3}; S4->{S5 Y4}; S5->{S6 Y5}; S6->Y6;
-
-    {rank = same; S1 S2 S3 S4 S5 S6}
-    {rank = same; Y1 Y2 Y3 Y4 Y5 Y6}
-
-  }
-
-  [1]: 1:6
-  [2]: rep(c('#F8766D','#00BFC4'), each = 3)
-", width = 300, height = 300)
+knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"),auto_pdf = T)
 
@@
 However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018].
-This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics.
-Belonging to the most basic metrics, which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [similar to relative or change in angle; @Larsson2013] in gazeHMM.
+This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics.
+Because they belong to the most basic metrics which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] in gazeHMM.
 Theoretically, these three metrics should separate eye movement events clearly.
@@
 In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction.
-PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013].
-They are specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
+PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013].
+They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
 Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
@@
 Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
-To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the forward difference (see Appendix B).
+To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle as the backward difference and the first discrete derivative as the forward difference (see Appendix B).
 
@@
 In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model [@Visser2019].
-In gazeHMM, the response model consists of three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal.
+In gazeHMM, the response model has three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal.
 The response variables are treated as conditionally independent on the states.
-I acknowledge that conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
+Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
 This step was merely taken to keep the HMM simple and identifyable.
 Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them).
-However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation) and due to noise above zero.
+However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation).
 Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean.
@@
 Thus, the uniform distribution should distinguish fixations from the other events.
-The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, no time-varying covariates were included.
+The initial state model and the transition model use multinomial distributions (with a category for each state).
+For all three submodels of the HMM, no time-varying covariates were included.
+Since the implementation of gazeHMM allows for regression models for each submodel and state, this implies that only intercepts for each parameter are estimated.
 
@@
 ## Optimization and Classification
-The paramters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997].
-The EM is generally suitable to estimate likelihoods with missing variables.
+The parameters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997].
+The EM algorithm is generally suitable to estimate likelihoods with missing variables.
 For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data (velocity, acceleration, and sample-to-sample angle) and the expected hidden states [eye movement events; @Visser2019].
@@
 In the study, the HMM repeatedly generated data with a set of parameters (true parameter values).
-The same model was then be applied to estimate the parameters from the generated data (estimated parameter values).
+The same model was then applied to estimate the parameters from the generated data (estimated parameter values).
 I compared the true with the estimated parameter values to assess whether a parameter was recovered by the model.
 Additionally, I compared the true states of the HMM with the estimated states to judge how accurately the model recovered the states that generated the data.
+
+## Starting Values
+The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
+To generate random starting values for the estimation of shape, scale, and concentration parameters, I used gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated.
+This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
+Mean parameters of the von Mises distribution always started at their true values.
 
@@
 In the first part, I varied the parameters of the HMM.
-For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were varied respectively.
+For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were varied, respectively.
 For each parameter, the HMM generated 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps.
@@
 To simulate noise, I replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
-This procedure ensured that velocity and acceleration values remained positive and were drawn from moderately skewed distributions with modes equal to the original values.
+This procedure ensured that velocity and acceleration values remained positive and were taken from moderately skewed distributions with modes equal to the original values.
 To angle, I added white noise from a von Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
@@
 The length of the missing data interval $l \in [1,200]$ samples varied in equidistant steps between the data sets.
-
-## Starting Values
-The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
-To generate random starting values for the estimation of shape, scale, and concentration parameters, I used gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated.
-This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
-Mean parameters of the von Mises distribution always started at their true values.
 
@@
                            rep((max(state.varied)+2):(max(state.varied)+7), max(state.varied))), nrow(x))) %>%
-    mutate_at(vars(par.varied, state.varied, par.est, state.est), as.factor)
+    mutate_at(vars(par.varied, state.varied, par.est, state.est), as.factor) %>%
+    mutate(state.varied = fct_recode(state.varied, "State 1" = "1", "State 2" = "2", "State 3" = "3", "State 4" = "4"),
+           state.est = fct_recode(state.est, "State 1" = "1", "State 2" = "2", "State 3" = "3", "State 4" = "4"))
 
@@
                            rep((k+2):(k+7), k)), nrow(x))) %>%
-    mutate_at(vars(cond, par.est, state.est), as.factor)
+    mutate_at(vars(cond, par.est, state.est), as.factor) %>%
+    mutate(state.est = fct_recode(state.est, "State 1" = "1", "State 2" = "2", "State 3" = "3", "State 4" = "4"))
 
@@
     mutate(par = rep(1:(x+1)^2, each = D),
-           from = rep(rep(1:(x+1), each = D), x+1),
-           to = rep(1:(x+1), each = D*(x+1)))
+           From = rep(rep(1:(x+1), each = D), x+1),
+           To = rep(1:(x+1), each = D*(x+1)))
 
   p <- ggplot(data, aes(x = true, y = est)) + 
-    facet_grid(rows = vars(from), cols = vars(to), labeller = label_both) +
+    facet_grid(rows = vars(From), cols = vars(To), labeller = label_both) +
     geom_point() + geom_smooth(method = "lm") +
@@
     pivot_longer(names(x), names_to = "par.est", values_to = "accuracy") %>%
-    mutate_at(vars(par.varied, state.varied), as.factor)
+    mutate_at(vars(par.varied, state.varied), as.factor) %>%
+    mutate(state.varied = fct_recode(state.varied, "State 1" = "1", "State 2" = "2", "State 3" = "3", "State 4" = "4"))
 
@@
   p <- ggplot(data = data.long, aes(x = int, y = accuracy, color = cond)) +
-    facet_grid(cols = vars(k), labeller = label_both) + 
+    facet_grid(cols = vars(k), labeller = partial(label_both, sep = " = ")) + 
     geom_point()
@@
 #### Two States
-In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation and accuracy between the true and estimated parameters and state sequence.
+In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation of estimated parameters and accuracy of estimated state sequences.
 Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM.
-The RMdSPD was below 0.1 for all estimated and manipulated parameters, that is, the median deviation never exceeded 10% of the true parameters [^2].
+The RMdSPD was below 0.1 for all estimated and manipulated parameters, indicating good recovery [^2].
 The regressions between manipulated true and estimated parameters are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
@@
 The deviation seemed to increase slighlty with parameter magnitude.
+Thus, parameter change was capture well and the estimation almost unbiased.
 Considering accuracy, Figure \@ref(fig:plot-acc-1-2) displays Cohen's kappa between true and estimated hidden state sequences.
-With two exceptions, kappa values were almost one, indicating nearly perfect agreement.
+With two exceptions, kappa values were almost one, suggesting nearly perfect agreement.
 In sum, the HMM with two states recovered parameters and hidden states very well and outliers only occured rarely.
@@
 
-[^2]: Note that the intial state probability $\rho_i$ has also been estimated.
-Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to an RMdSPD of one).
+[^2]: Note that the intial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
+Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
 Therefore, I decided not to include it in the analysis.
@@
 Top facet labels show to and right facet labels show from which state the HMM is moving.
+Dashed lines refer to perfect recovery.
 
@@
 
-(ref:plot-regresp-1-2) Regression lines between true and estimated response parameters of the two-state HMM in part one. Top facet labels indicate response parameters.
+(ref:plot-regresp-1-2) Regression lines between true and estimated response parameters of the two-state HMM in part one.
+Top facet labels indicate response parameters.
+Dashed lines refer to perfect recovery.
 
@@
 When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently below 0.5.
-Varying $\kappa$ in states two and three led to RMdSPDs below 0.5 in the respective states.
-Otherwise, RMdSPDs were consistently lower than 0.1, indicating a median deviation between true and estimated parameters below 10% of the true parameter.
+Varying $\kappa$ in states two and three led to RMdSPDs below 0.5 in the respective states, which can be interpreted as moderate recovery.
+Otherwise, RMdSPDs were consistently lower than 0.1, indicating good recovery.
 Inspecting the regressions between manipulated true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one).
@@
 In contrast to the two-state HMM, larger deviatons and more outliers were observed.
-Cohen's kappa values for the three-state HMM are presented in Figure \@ref(fig:plot-acc-1-3).
-For most estimated models, the kappas between true and estimated state sequences were above 0.95, indicating almost perfect agreement.
-However, for some models, I observed kappas clustered around zero or -0.33, suggesting that state labels switched.
+Cohen's kappa values are presented in Figure \@ref(fig:plot-acc-1-3).
+For most estimated models, the kappas between true and estimated state sequences were above 0.95, meaning almost perfect agreement.
+However, for some models, I observed kappas clustered around zero or -0.33, which suggests that state labels were switched.
 To summarize, the three-state HMM had a slightly worse parameter recovery and more outliers than the two-state model but shows a similar good accuracy.
@@
 
-(ref:plot-regresp-1-3) Regression lines between true and estimated response parameters of the three-state HMM in part one. Top facet labels indicate response parameters.
+(ref:plot-regresp-1-3) Regression lines between true and estimated response parameters of the three-state HMM in part one.
+Top facet labels indicate response parameters.
+Dashed lines refer to perfect recovery.
 
@@
 #### Four States
-The RMdSPDs for the four-state HMM is illustrated in Figure \@ref(fig:plot-rmdspd-1-4).
-For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were below 0.5.
-Estimated kappa parameters in smooth pursuits were also often below 0.5 when paramteres in states two, three, and four were varied.
-Otherwise, RMdSPDs were below 0.1, indicating a median deviation below 10% of the true parameter.
-Looking at the regressions between true and estimated paramters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) show strong and unbiased relationships.
+The RMdSPDs for the four-state HMM is shown in Figure \@ref(fig:plot-rmdspd-1-4).
+For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were below 0.5, suggesting moderate recovery.
+Estimated kappa parameters in smooth pursuits were also often below 0.5 when parameters in states two, three, and four were varied.
+Otherwise, RMdSPDs were below 0.1, indicating good recovery.
+Looking at the regressions between true and estimated paramters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) illustrate strong and unbiased relationships.
 However, there were larger deviations and more outliers than in the previous models, especially for states one and four.
 Accuracy measured by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
-Here, some kappa values clustered around 0.25 and zero, which again can be interpreted as the result of label switching.
+Here, some kappa values clustered around 0.25 and zero, which, again, can be interpreted as the result of label switching.
 Summarizing, the parameter recovery for the four-state HMM was slightly worse with even more outliers compared to the three-state model.
 Especially the recovery of transition parameters in states corresponding to fixations and smooth pursuits decreased.
-Overall, simulations in part one demonstrated that the HMM recovered parameters at least moderately well when parameters were manipulated (all RMdSPDs below 0.5).
-The HMM estimated state sequences very accurately.
-Adding states to the model decreased the accuracy and recovery slightly, leading to more outliers, especially when smooth pursuits were added.
+Overall, simulations in part one demonstrated that the HMM recovered parameters very well when parameters were manipulated.
+Deviations from true parameters were mostly small.
+In the four-state model, estimated transition probabilities for state one and four deviated moderately.
+Moreover, the HMM estimated state sequences very accurately.
+Again, the four-state model showed lower accuracy.
 
@@
 
-(ref:plot-regresp-1-4) Regression lines between true and estimated response parameters of the four-state HMM in part one. Top facet labels indicate response parameters.
+(ref:plot-regresp-1-4) Regression lines between true and estimated response parameters of the four-state HMM in part one.
+Top facet labels indicate response parameters.
+Dashed lines refer to perfect recovery.
 
@@
 In the second part, I varied the sample size of the HMM and added noise to the generated data.
-For the two-state HMM, the RMdSPDs were above 0.5 for $\beta_{vel}$ and $\beta_{acc}$ in both states (see Figure \@ref(fig:plot-rmdspd-2-2)).
-The other estimated parameters showed RMdSPDs close to or below 0.1.
+For the two-state HMM, the RMdSPDs were above 0.5 for $\beta_{vel}$ and $\beta_{acc}$ in both states (see Figure \@ref(fig:plot-rmdspd-2-2)), suggesting bad recovery.
+The other estimated parameters showed RMdSPDs close to or below 0.1, which means they were recovered well.
 Increasing the sample size seemed to improve RMdSPDs for most parameters slightly.
@@
 Accuracy measured by Cohen's kappa was almost perfect with kappa values very close to one (see Figure \@ref(fig:plot-acc-2), left plot).
-To conclude, adding noise did not affect the parameter recover or accuracy of the two-state HMM, whereas increasing the sample size improved the recovery slightly.
+To conclude, adding noise only affected the recovery of scale parameters but not the accuracy of the two-state HMM.
+Increasing the sample size improved the recovery slightly.
 
@@
 #### Three States
-The RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)).
+The RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)), indicating bad recovery.
 Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5.
@@
 RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
-For states one and four, values for most parameters (including all transition probabilites) were above 0.5.
+For states one and four, values for most parameters (including all transition probabilites) were above 0.5, suggesting bad recovery.
 Similarly, RMdSPDs for $\beta_{vel}$ and $\beta_{acc}$ in states two and three were above 0.5.
@@
 In general, the HMM recovered parameters well despite noise being added to the data.
-However, when smooth pursuits were added to the model, the parameter recovery for fixation and smooth pursuit states substantially decreased.
-When PSOs or smooth pursuits were included, scale parameters of gamma distributions were badly recovered.
-Increasing the samples in the HMM slightly improved the recovery of most parameters.
-The accuracy of the model was slightly lowered when including more states, but it was neither affected by the noise variability $\tau_{noise}$ nor the sample size.
+However, in the four-state model, the parameter recovery for states one and four substantially decreased.
+In the three- and four-state models, scale parameters of gamma distributions were badly recovered.
+Increasing the sample size in the HMM slightly improved the recovery of most parameters.
+The accuracy of the model was slightly lowered when more states were included, but it was neither affected by the noise variability $\tau_{noise}$ nor the sample size.
 
@@
 The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM.
-For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure \@ref(fig:plot-rmdspd-3-2)).
+For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure \@ref(fig:plot-rmdspd-3-2)), suggesting good recovery.
 Cohen's kappa values were slightly lower than 1, indicating almost perfect accuracy (see Figure \@ref(fig:plot-acc-3)).
-For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5 (see Figure \@ref(fig:plot-rmdspd-3-3)).
+For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5 (see Figure \@ref(fig:plot-rmdspd-3-3)), which means they were moderately well recovered.
 As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33 (see Figure \@ref(fig:plot-acc-3)).
@@
 State sequences were accurately estimated and more states being included in the model slightly decreased the accuracy.
-Increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor accuracy.
+Increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor the accuracy.
 
@@
 In this part, I explored post-hoc whether clusters of low Cohen's kappa values were due to label switching.
-According to Visser and Speekenbrink [-@Visser2019], label switching occurs when exchanging the order of the states leads to equally likely models.
+According to @Visser2019, label switching occurs when exchanging the order of the states leads to equally likely models.
 Thus, the model is accurate but has oppositely estimated states compared to the true states.
@@
 For the four-state model, kappas started around 0.85 and decreased to 0.5 for five missing intervals, suggesting moderate to substantial accuracy.
-As in previous parts, a several kappa values clustered around zero and -0.33 for the three-state model and around 0.6, 0.25, and zero for the four-state model.
+As in previous parts, several kappa values clustered around zero and -0.33 for the three-state model and around 0.6, 0.25, and zero for the four-state model.
 In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model from an nearly perfect to a moderate extent.
 With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences.
-The most critical decrease in recovery occured when noise was added to the data generated by models including smooth pursuit.
+The most critical decrease in recovery occured when noise was added to the data generated by models including four states.
 In the noise condition, scale parameters of gamma distributions were often badly recovered.
@@
 ### Starting Values and Model Fitting
-I applied gazeHMM on two benchmark data sets to test my hypotheses.
-Note that, for the @Ehinger2019 data set, I only fitted models to the Eyelink data to reduce the complexity of the analyses.
+To test my hypotheses, I applied gazeHMM on two benchmark data sets.
+Note that, for the @Ehinger2019 data set, I only fitted models to the Eyelink data to reduce the complexity of the analysis.
 For both data sets, I used the same set of starting values (initial state model: $\rho=1/k$; transition model: $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$; response model: see Table \@ref(tab:tab-starting-values)) to estimate the parameters of the HMM.
 In contrast to the simulation study, generating random starting value often led to bad model fits and label switching between states.
-To improve the fitting of the gamma distributions, velocity and acceleration signals were scaled down by factor 100 (so were the starting values for their gamma distributions).
+To improve the fitting of the gamma distributions, velocity and acceleration signals were scaled down by factor 100 [^7] (so were the starting values for their gamma distributions).
 The algorithm was applied separately for every participant and every condition, task, or block.
 For the data set by @Andersson2017, the algorithm was successfully fitted for every participant and condition.
-An example for event classification by gazeHMM is displayed in Figure \@(fig:plot-example-classification) and examples for parameter estimates are presented in Tables \@ref(tab:tab-example-trans) and \@ref(tab:tab-example-resp).
+Examples of event classification and parameter estimates from gazeHMM are presented in Appendix A.
+
+[^7]: Scaling down by factor 100 differs from the simulation study (scaling down by 10).
+The algorithm allows the user to manually specify this factor and in this case, factor 100 led to better model fits than factor 10.
+Thus, I chose it exploratorily.
 
@@
 To examine whether HMMs are accurately describing eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
-Schwarz weights are the result of transforming a set of BIC values.
+Schwarz weights are the result of transforming a set of BIC values [@Wagenmakers2004].
 They can be interpreted as the probability for a model to have generated the data it was fitted to.
@@
 Figure \@ref(fig:plot-schwarz-weights-img) shows the Schwarz weights for different subjects and models in the image condition.
-For all subjects, the five-state model displays the highest weight, suggesting that it hast most likely generated the data.
-In the moving dots condition, the five-state model has the highest weights for most subjects, but the one-, three, and four-state models exhibit the highest weights for two subjects each (see Figure \@ref(fig:plot-schwarz-weights-dots)).
-The video condition has the same pattern as the image condition, as the five-state model consistently has the highest weights (see Figure \@ref(fig:plot-schwarz-weights-vid)).
+For all subjects, the five-state model displayed the highest weight, suggesting that it most likely generated the data.
+In the moving dots condition, the five-state model had the highest weights for most subjects, but the one-, three, and four-state models exhibited the highest weights for two subjects each (see Figure \@ref(fig:plot-schwarz-weights-dots)).
+The video condition had the same pattern as the image condition, as the five-state model consistently had the highest weights (see Figure \@ref(fig:plot-schwarz-weights-vid)).
 In contrast to my hypothesis, applying different HMMs to the @Andersson2017 data suggests that the five-state model has most likely generated the data.
 Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
-The model comparison for the @Ehinger2019 data set yielded similar results (see Figures \@ref(fig:plot-schwarz-weights-task4) and \@ref(fig:plot-schwarz-weights-task5)).
+For the @Ehinger2019 data, I predicted a one-state model to be preferred. Here, the model comparison for the yielded similar results (see Appendix A), contradicting my hypothesis.  
 A recent model recovery study showed that the BIC tended to prefer overly complex HMMs when they were misspecified [e.g., the conditional independence assumption was violated; @Pohle2017].
 Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2000] performed better in choosing the correct data generating model.
-Therefore, I computed the weighted ICL criterion for the HMMs fitted to the @Andersson2017 data set.
-Using the ICL as the model selection criterion yielded very similar results to the BIC (see Figures \@ref(fig:plot-icl-weights-img), \@ref(fig:plot-icl-weights-dots), and \@ref(fig:plot-icl-weights-vid)).
+Therefore, I computed the weighted ICL criterion for the HMMs fitted to the @Andersson2017 data set as an exploratory analysis.
+Using the ICL as the model selection criterion yielded very similar results to the BIC (see Appendix A).
 The preference for the five-state model was even more consistent across conditions and subjects.
@@
 The high RMSD values were mostly due to a large amount of very short fixations.
-For saccades, gazeHMM has a relatively low RMSD of the algorithms in the image and video condition and a relatively high RMSD in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)).
+For saccades, gazeHMM has a relatively low RMSD in the image and video condition and a relatively high RMSD in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)).
 Here, gazeHMM overestimated the duration of saccades, as indicated by high mean and SD durations.
@@
 Compared to the other algorithms, the agreement of gazeHMM was the lowest in the image condition but the highest in the moving dots and video condition.
-For saccades, gazeHMM showed kappa values that correspond to a moderate to substantial agreement were relatively high compared to the other algorithms.
+For saccades, gazeHMM showed kappa values that correspond to a moderate to substantial agreement and were relatively high compared to the other algorithms.
 However, gazeHMM never reached the highest agreement.
@@
 It was relatively low in the moving dots condition and mediocre in the image and video conditions compared to NH and LNS.
-The agreement for smooth puruits was moderate in the moving dots condition but slight in the image and video conditions. No other algorithm in the study was designed to detect smooth pursuits.  
+The agreement for smooth puruits was moderate in the moving dots condition, fair in the video condition, and slight in the image condition.
+No other algorithm in the study was designed to detect smooth pursuits.
 In sum, gazeHMM revealed a relatively high sample-to-sample agreement to human coders for saccades.
@@
 Looking at the confusion matrix in Table \@ref(tab:tab-conf-4), it can be seen that gazeHMM confused fixations and smooth pursuit to a large extend.
-In the image condition, it overclassified smooth pursuit events, while they were underclassified in the dynamic conditions.
+In the static image condition, it overclassified smooth pursuit events, while they were underclassified in the dynamic conditions.
 
@@
 Including only three states could prevent gazeHMM from overclassifying smooth pursuits in the image data and yield better validation results.
-Therefore, I decided to apply gazeHMM with three states to the image dat and repeat the comparison to other algorithms exploratorily.
+Therefore, I decided to apply gazeHMM with three states to the image data and explore the comparison to other algorithms.
 Table \@ref(tab:tab-rmsd-3) shows the RMSD between gazeHMM and human coders only for the image condition.
-For fixations, gazeHMM shows a comparably low RMSD but it was relatively high for saccades and PSOs.
+For fixations, gazeHMM shows a comparably low RMSD, but it was relatively high for saccades and PSOs.
 The sample-to-sample agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kappa-3) (only for the image condition).
@@
 In all three conditions, the disagreement was slightly lower for gazeHMM than for all the other algorithms.
-Compared to gazeHMM with four states, the RMSD decreased while the sample-to-sample agreement increased for fixations.
+Compared to gazeHMM with four states, the RMSD for fixations decreased while the sample-to-sample agreement increased.
 For saccades and PSOs, the opposite pattern was found.
@@
 Overall, applying gazeHMM with three states to static data seems more appropriate, since the increase in fixation classification outweighs the decrease in classification of saccades and PSOs.
-There was even some support for my second hypothesis since gazeHMM had lower overall disagreement to human coders than all other algorithms.
+There was support for my second hypothesis since gazeHMM had lower overall disagreement to human coders than all other algorithms.
 
@@
 At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
-Despite these differences, they recommend comparisons with human coding to demonstrate the performance of new algorithms and finding errors in their design.
+Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and finding errors in their design.
 

### 2020-06-22 manuscript/report_appendix.Rmd
update report

@@
----
-suppress-bibliography: true
----
-
 # Additional Figures and Tables
@@
 
-tab.exp.tr <- matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[5:20], 2), 4, 4, byrow = T)
+tab.exp.tr <- cbind(1:4, matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[5:20], 2), 4, 4, byrow = T))
 
 apa_table(tab.exp.tr,
-          col.names = as.character(1:4),
-          col_spanners = list("To state" = c(1,4)),
-          row.names = TRUE,
-          added_stub_head = "From\nstate",
+          col.names = c("From\nstate", as.character(1:4)),
+          col_spanners = list("To state" = c(2,5)),
           caption = "(ref:tab-example-trans)",
@@
 
-tab.exp.resp <- matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[21:44], 2), 4, 6, byrow = T)
+tab.exp.resp <- cbind(1:4, matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[21:44], 2), 4, 6, byrow = T))
 
@@
 apa_table(tab.exp.resp,
-          col.names = c("Shape (vel)", "Scale (vel)", "Shape (acc)", "Scale (acc)", "Mean", "Con"),
-          row.names = TRUE,
+          col.names = c("State", "Shape (vel)", "Scale (vel)", "Shape (acc)", "Scale (acc)", "Mean", "Con"),
           format.args = list(na_string = "-"),
-          added_stub_head = "State",
           caption = "(ref:tab-example-resp)",

### 2020-06-23 manuscript/internship_report.Rmd
change abstract

@@
 Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere weaknesses.
-I developed gazeHMM, an algorithm that improves on previous HMM algorithms.
+Therefore, I developed gazeHMM, an HMM algorithm that has no critical parameters to be set by users, does not require human coded data as input, and classifies fixations, saccades, PSOs, and smooth pursuits.
 The development was guided by the question of whether HMMs are useful at describing eye movements and whether they improve the event classification.
-A simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences with only few exceptions.
-When applied to benchmark data, model comparisons for gazeHMM with different events did not match theoretical expectations.
-Compared to other algorithms, gazeHMM improved the event classification which was assessed by the similarity to human coding, especially for dynamic stimuli.
-The classification of smooth pursuits was found to be inadequate and is thus only recommended for exploration.
-Otherwise, I consider gazeHMM as appropriate for practical applications.
-In sum, gazeHMM improves classification but is not accurately describing eye movements, yet.
-I identify several explanations and discuss potential remedies for this result.
+I evaluated gazeHMM in a simulation study and on benchmark data.
+The simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences.
+Critical exceptions for good recovery were adding a smooth pursuit like state to gazeHMM and noisy data.
+For benchmark data, model comparisons with gazeHMM yielded preferred models with more states than expected.
+I assessed the classification performance of gazeHMM compared to other algorithms by the agreement to human event coding.
+Here, gazeHMM improved the event classification but did not outperform all other algorithms in most cases.
+Both the simulation study and benchmark application showed poor classification performance for smooth pursuits.
+Thus, I advice to classify smooth pursuits only for exploration and recommend gazeHMM with three events for application.
+Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.
 

### 2020-06-23 manuscript/report_appendix.Rmd
change abstract

@@
 (ref:tab-example-trans) Transition probabability estimates for example data from Andersson et al. (2017)
-(ref:tab-example-trans-note) Rows indicate from which state and columns into which state the model transitions.
 
@@
 
-tab.exp.tr <- cbind(1:4, matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[5:20], 2), 4, 4, byrow = T))
+tab.exp.tr <- cbind(as.character(1:4), matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[5:20], 3), 4, 4, byrow = T))
 
 apa_table(tab.exp.tr,
-          col.names = c("From\nstate", as.character(1:4)),
-          col_spanners = list("To state" = c(2,5)),
-          caption = "(ref:tab-example-trans)",
-          note = "(ref:tab-example-trans-note)")
+          col.names = c("From", as.character(1:4)),
+          col_spanners = list("To" = c(2,5)),
+          caption = "(ref:tab-example-trans)")
 
@@
 (ref:tab-example-resp) Response parameter estimates for example data from Andersson et al. (2017)
-(ref:tab-example-resp-note) Con = concentration parameter of the von-Mises distribution.
 
@@
 
-tab.exp.resp <- cbind(1:4, matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[21:44], 2), 4, 6, byrow = T))
+tab.exp.resp <- cbind(as.character(1:4), matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[21:44], 2), 4, 6, byrow = T))
 
-tab.exp.resp[1, c(5,6)] <- NA
+tab.exp.resp[1, c(6,7)] <- NA
 
@@
           format.args = list(na_string = "-"),
-          caption = "(ref:tab-example-resp)",
-          note = "(ref:tab-example-resp-note)")
+          caption = "(ref:tab-example-resp)")
 

### 2020-06-23 manuscript/internship_report.Rmd
minor changes results

@@
 @Houpt2018 compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm.
-Similarly, @Nystrom2010 use a Savitzky-Golay (SG) filter @Savitzky1964 to estimate velocity and acceleration from a gaze signal.
+Similarly, @Nystrom2010 use a Savitzky-Golay (SG) filter [@Savitzky1964] to estimate velocity and acceleration from a gaze signal.
 Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods.
@@
 To assess state recovery, I computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set.
+Cohen's kappa estimates the agreement between two classifiers accounting for the agreement due to chance.
 Higher kappa values were interpreted as better model accuracy.
@@
 For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were below 0.5, suggesting moderate recovery.
-Estimated kappa parameters in smooth pursuits were also often below 0.5 when parameters in states two, three, and four were varied.
+Estimated kappa parameters in state four were also often below 0.5 when parameters in states two, three, and four were varied.
 Otherwise, RMdSPDs were below 0.1, indicating good recovery.
@@
 In the four-state model, estimated transition probabilities for state one and four deviated moderately.
-Moreover, the HMM estimated state sequences very accurately.
-Again, the four-state model showed lower accuracy.
+Moreover, the HMM estimated state sequences very accurately with decreasing accuracy for the four-state model.
 
@@
 For the three-state HMM, adding noise led to a slightly worse recovery and accuracy overall.
-However, scale parameters of gamma distributions in all three states were only badly recovered.
+Scale parameters of gamma distributions in all three states were only badly recovered.
 
@@
 The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero (see Figure \@ref(fig:plot-acc-3)).
-In conclusion, the parameter recovery of the model worked very well.
-State sequences were accurately estimated and more states being included in the model slightly decreased the accuracy.
-Increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor the accuracy.
+Taken together, increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor the accuracy.
 
@@
 Cohen's kappa values linearly decreased with the length of missing data intervals and the decrease became linearly steeper when more intervals were included.
-For two and three state models, kappas started at almost one and decreased to 0.6 for five missing intervals, indicating substantial to almost perfect accuracy.
-For the four-state model, kappas started around 0.85 and decreased to 0.5 for five missing intervals, suggesting moderate to substantial accuracy.
+For two and three state models, kappas started at almost one (almost perfect) and decreased to 0.6 (substantial accuracy) for five missing intervals.
+For the four-state model, kappas started around 0.85 (almost perfect) and decreased to 0.5 (substantial accuracy) for five missing intervals.
 As in previous parts, several kappa values clustered around zero and -0.33 for the three-state model and around 0.6, 0.25, and zero for the four-state model.
-In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model from an nearly perfect to a moderate extent.
+In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model from a nearly perfect to a moderate extent.
 With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences.
@@
 For both data sets, I used the same set of starting values (initial state model: $\rho=1/k$; transition model: $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$; response model: see Table \@ref(tab:tab-starting-values)) to estimate the parameters of the HMM.
-In contrast to the simulation study, generating random starting value often led to bad model fits and label switching between states.
+In contrast to the simulation study, generating random starting values often led to bad model fits and label switching between states.
 To improve the fitting of the gamma distributions, velocity and acceleration signals were scaled down by factor 100 [^7] (so were the starting values for their gamma distributions).
@@
 
-To examine whether HMMs are accurately describing eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
+To examine whether gazeHMM accurately describes eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
 Schwarz weights are the result of transforming a set of BIC values [@Wagenmakers2004].
@@
 
+ref.acc[ref.acc == 0] <- NA
+
 A2017.acc <- lapply(2:4, function(k) {
@@
 #### Event Durations
-The hypothesis that HMMs improve the classification performance of eye movement events was investigated by comparing gazeHMM against other algorithms.
+The hypothesis that gazeHMM improves the classification performance of eye movement events was investigated by comparing it against other algorithms.
 Therefore, I applied gazeHMM with four events to the @Andersson2017 data set.
-Therefore, I calculated the RMSD [^3] of event durations as described in the original article by Andersson et al.
+I calculated the RMSD [^3] of event durations as described in the original article by Andersson et al.
 for all algorithms included in the study plus gazeHMM: First, the mean and standard deviation of event durations as well as the number of events were normalized to [0,1] for each coder and algorithm.
@@
 Here, gazeHMM underestimated the number of PSOs and their duration compared to the human coders.
-No other algorithm in the study was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)).
+No other algorithm in the study was designed to parse smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)).
 In sum, gazeHMM did not outperform all other algorithms regarding the RMSD of event durations compared to human coders.
@@
 [^3]: In my proposal, I declared to compute the RMSE instead of RMSD.
-However, since I conducted an in sample comparison, the term deviation is more appropriate.
-Andersson et al.
-(2017) also referred to it as RMSD.
+However, since I conducted an in-sample comparison and it was used by @Andersson2017, the term *deviation* is more appropriate (i.e., RMSD).
 
 #### Sample-to-sample Agreement
-Moreover, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of sample-to-sample agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
-Cohen's kappa indicates the agreement between to classifiers accounting for the agreement due to chance.
-Absolute values were interpreted according to @Landis1977.
+Additionally, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of sample-to-sample agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
+Absolute kappa values were interpreted according to @Landis1977.
 For fixations, absolute kappa values in all three conditions indicated a slight to fair agreement between gazeHMM and human coders.
@@
 For PSOs, absolute kappa values showed slight to fair agreement agreement to human coders.
-It was relatively low in the moving dots condition and mediocre in the image and video conditions compared to NH and LNS.
+It was relatively low in the moving dots condition and second-rate in the image and video conditions compared to NH and LNS.
 The agreement for smooth puruits was moderate in the moving dots condition, fair in the video condition, and slight in the image condition.
@@
 In sum, gazeHMM revealed a relatively high sample-to-sample agreement to human coders for saccades.
-For PSOs and smooth pursuits, I found mediocre agreement.
+For PSOs and smooth pursuits, I found fair to moderate agreement.
 The performance for fixations was absolutely and relatively low.
@@
           font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-kappa-4)",
@@
 Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events had a higher overall disagreement than all but two of the other algorithms.
-In the moving dots and video conditions, the disagreement was considerably lower than for all the other algorithms.
+In the moving dots and video conditions, the disagreement was considerably lower than for all other algorithms.
 Notably, gazeHMM's disagreement was similar across conditions (around 45%).
@@
 Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms in terms of RMSD and sample-to-sample agreement.
-The two criteria rather indicated a mediocre performance of gazeHMM.
+The two criteria rather indicated a moderate performance of gazeHMM.
 However, the overall disagreement was lower for gazeHMM than for all other algorithms in the dynamic conditions but the highest in the static image condition.
@@
           col.names = c("Algorithm", "Fixations", "Saccades", "PSOs"),
+          format.args = list(na_string = "-", drop0trailing = T),
           caption = "(ref:tab-kappa-3)",
@@
 I treat this pattern as moderate evidence for my second hypothesis that gazeHMM improves the classification of eye movements.
-Taken together, I only found some evidence for one part of my research question, that gazeHMM improves classifying gaze data but not describing them.
+Taken together, I only found some evidence for one part of my research question, that gazeHMM improved classifying gaze data but not describing them.
 Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle.
@@
 The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).
-The model comparison on the benchmark data clearly provided evidence against gazeHMM being an accurate generative model for eye movement data.
+The model comparison on the benchmark data suggested that gazeHMM is not the best generative model for eye movement data.
 Even using a model selection criterion that identified correct models succesfully in previous studies [@Pohle2017], did not yield the expected pattern.
@@
 Comparing gazeHMM to other algorithms on benchmark data showed that gazeHMM moderately improved the event classification in agreement with human coders.
-However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yieled different results.
+However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yielded different results.
 The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events.
@@
 Then, the maximum averaged probability could be used to classify the samples into events.
-This approach could lead to a more robust classification because averaging might reduce the inaccuracy of each competing model.
+This approach could lead to a more robust classification because it reduces the overconfidence of each competing model and easily adapts to new data [analogous to Bayesian model averaging; @Hinne2020].
 However, the model comparison for gazeHMM often showed extreme weights for a five-state model, which would lead to a very limited influence of the other models in the averaged probabilities.
@@
 This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades.
-A drawback of the extension would be higher computational costs.
+Drawback of this extension are higher computational costs and difficulties with including covariates [@Zucchini2016].
 Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM.

### 2020-06-24 test/test.R
remove test and data

@@ -1,151 +1,0 @@
-### gazeHMM test ###
-
-# Author: Malte Lüken
-# Date: 27.04.2020
-
-
-library(tidyverse)
-library(depmixS4)
-source("algorithm/gazeHMM.R")
-source("algorithm/plotting_functions.R")
-
-
-# Test on static data -----------------------------------------------------
-
-# Eprep data set
-
-load("data/Eprep")
-
-dim <- c(473, 296) # screen dimensions
-res <- c(1680, 1050) # screen resolution
-dist <- 600 # distance to screen
-fr <- 500 # frame rate
-
-set.seed(123)
-
-Eprep.fit.3 <- Eprep %>%
-  dplyr::filter(block == sample(unique(Eprep$block), 1)) %>%
-  group_by(subject) %>%
-  summarise(fit = list(try(gazeHMM(x = xp, y = yp, t = (time - time[1])/1e3, 
-                                              unit = "px", res = res, dim = dim, fr = fr, dist = dist, 
-                                              nstates = as.integer(3), random.respstart = F)))) %>%
-  pull(fit)
-
-
-# Inspect parameters
-
-lapply(Eprep.fit.3, function(x) {summary(x$model)})
-
-
-# Plot classification result
-
-for (i in 1:length(Eprep.fit.3)) {
-  
-  set.seed(i*25)
-  
-  tp <- sample(Eprep.fit.3[[i]]$samples$t[Eprep.fit.3[[i]]$samples$t > 0.5], 1)
-  
-  int <- c(tp - 0.5, tp) # 0.5 second interval
-  
-  plot_samples_time(Eprep.fit.3[[i]], from = int[1], to = int[2]) + 
-    ggsave(filename = paste("Eprep_3comp_subj_", i, "_overtime.png", sep = ""), 
-           width = 12, height = 6)
-  
-  plot_samples_xy(Eprep.fit.3[[i]], from = int[1], to = int[2]) + 
-    ggsave(filename = paste("Eprep_3comp_subj_", i, "_xy.png", sep = ""), 
-           width = 12, height = 6)
-  
-}
-
-
-# Plot bivariate classification 
-
-for (i in 1:length(Eprep.fit.3)) {
-  
-  ggplot(Eprep.fit.3[[i]]$samples, aes(vel, acc, color = as.factor(state))) + 
-    geom_point() + scale_color_discrete(name = "state", labels = c("Fix", "Sac", "PSO")) +
[…]

### 2020-06-24 manuscript/internship_report.Rmd
fix typos

@@
 
-dir <- "~/Uni/Psychologie Master/Internship/GazeHMM"
-
 library(papaja)
@@
 
-```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)"}
-
-knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"),auto_pdf = T)
+```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)", fig.height=3, fig.width=5}
+
+knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 300)
 
@@
 The models will differ in their assumed number of events in the data.
-The model comparison will result in a preferred model which scores best on the evaluation criterium.
+The model comparison will result in a preferred model which scores best on the evaluation criterion.
 The preferred model is expected to match the theoretically assumed number of present events in the benchmark data.
@@
 Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018].
-Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best.
+Here, my goal was to find those metrics that separate the gaze data samples belonging to different events the best.
 However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018].
@@
 
-## Fitering and Smoothing
+## Filtering and Smoothing
 As with metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics.
@@
 Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration).
-@Houpt2018 compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm.
-Similarly, @Nystrom2010 use a Savitzky-Golay (SG) filter [@Savitzky1964] to estimate velocity and acceleration from a gaze signal.
-Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods.
-They follow a similar trend, but the SG velocity signal is less noisy and the acceleration signal is consequently lower than the Laplace acceleration signal.
+@Houpt2018 computed the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm.
+Similarly, @Nystrom2010 used a Savitzky-Golay (SG) filter [@Savitzky1964] to estimate velocity and acceleration from a gaze signal.
+Figure \@ref(fig:filter-comparison) displays an example of velocity and acceleration signals obtained by both methods.
+They follow a similar trend, but the SG velocity signal was less noisy and the acceleration signal was consequently lower than the Laplace acceleration signal.
 Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
@@
 It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
-Thus, users can choose wether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth puruits.
+Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.
 In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model [@Visser2019].
@@
 Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
-This step was merely taken to keep the HMM simple and identifyable.
+This step was merely taken to keep the HMM simple and identifiable.
 Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them).
@@
 In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose.
-It has to be noted that the gamma was choosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
+It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
 To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution.
@@
 The initial state model and the transition model use multinomial distributions (with a category for each state).
-For all three submodels of the HMM, no time-varying covariates were included.
+For all three submodels of the HMM, no time-varying covariates are included.
 Since the implementation of gazeHMM allows for regression models for each submodel and state, this implies that only intercepts for each parameter are estimated.
@@
 However, most of the missing data in eye movement classification are due to blinks.
-Therefore, the user can indicate in gazHMM which samples should be labeled as blinks (other missing samples are treated as noise).
+Therefore, the user can indicate in gazeHMM which samples should be labeled as blinks (other missing samples are treated as noise).
 Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs.
@@
 The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.
-When evaluating the likelihood of the HMM given parameters and the joint density of response distributions for missing data, the model integrates over all possible values.
+When evaluating the likelihood of the HMM, given the parameters and joint density of response distributions, for missing data, the model integrates over all possible values.
 The resulting density for missing samples is therefore set to one [@Visser2019].
@@
 The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability.
-Parameters of the response distributions (except for the uniform distribution) were optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein steplengths [@Barzilai1988].
+Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988].
 
@@
 After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence.
-I implemented this rountine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
-Moreover, gazeHMM did not explicitly control the duration of events in the HMM which ocassionally led to unreasonably short events.
+I implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
+Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events.
 Thus, the postprocessing routine heuristically compensates for such violations.
 This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events.
-Samples were relabeled as the state of the previous event.
-Finally, samples initially indicated as missing were labeled as noise (including blinks) and event metrics were computed (e.g., fixation durations).
+Samples are relabeled as the state of the previous event.
+Finally, samples initially indicated as missing are labeled as noise (including blinks) and event metrics are computed (e.g., fixation duration).
 
@@
 I did not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
-For the transition probabilities, I only simultaneously varied the probabilities for staying in the same state (diagonals of the tranistion matrix) to reduce the complexity of the simulation.
-The left over probability mass was split evenly between the probabilites for switching to a different state (per row of the transition matrix).
+For the transition probabilities, I only simultaneously varied the probabilities for staying in the same state (diagonals of the transition matrix) to reduce the complexity of the simulation.
+The left over probability mass was split evenly between the probabilities for switching to a different state (per row of the transition matrix).
 Moreover, I did not modify the mean parameters of the von Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
@@
 
-load(paste(dir, "simulation/part3_expl.Rdata", sep = "/"))
+load(here("simulation/part3_expl.Rdata"))
 
@@
 With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely.
-The deviation seemed to increase slighlty with parameter magnitude.
+The deviation seemed to increase slightly with parameter magnitude.
 Thus, parameter change was capture well and the estimation almost unbiased.
@@
 With two exceptions, kappa values were almost one, suggesting nearly perfect agreement.
-In sum, the HMM with two states recovered parameters and hidden states very well and outliers only occured rarely.
+In sum, the HMM with two states recovered parameters and hidden states very well and outliers only occurred rarely.
 
@@
 
-[^2]: Note that the intial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
+[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
 Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
@@
 Again, the deviation seemed to increase with true parameter magnitude (reverse for kappas).
-In contrast to the two-state HMM, larger deviatons and more outliers were observed.
+In contrast to the two-state HMM, larger deviations and more outliers were observed.
 Cohen's kappa values are presented in Figure \@ref(fig:plot-acc-1-3).
@@
 Otherwise, RMdSPDs were below 0.1, indicating good recovery.
-Looking at the regressions between true and estimated paramters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) illustrate strong and unbiased relationships.
+Looking at the regressions between true and estimated parameters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) illustrate strong and unbiased relationships.
 However, there were larger deviations and more outliers than in the previous models, especially for states one and four.
@@
 RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
-For states one and four, values for most parameters (including all transition probabilites) were above 0.5, suggesting bad recovery.
+For states one and four, values for most parameters (including all transition probabilities) were above 0.5, suggesting bad recovery.
 Similarly, RMdSPDs for $\beta_{vel}$ and $\beta_{acc}$ in states two and three were above 0.5.
@@
 Thus, the model is accurate but has oppositely estimated states compared to the true states.
-In this case, an HMM with three states that is perfectly accurate but has one label swtiched would have a Cohen's kappa of 0.
-To test if label switching occured, I manually switched one or two labels post-hoc (see Figure \@ref(fig:plot-acc-3-expl)).
+In this case, an HMM with three states that is perfectly accurate but has one label switched would have a Cohen's kappa of 0.
+To test if label switching occurred, I manually switched one or two labels post-hoc (see Figure \@ref(fig:plot-acc-3-expl)).
 It can be seen that this approach resolved low accuracy clusters for three and four states.
@@
 With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences.
-The most critical decrease in recovery occured when noise was added to the data generated by models including four states.
+The most critical decrease in recovery occurred when noise was added to the data generated by models including four states.
 In the noise condition, scale parameters of gamma distributions were often badly recovered.
 Higher sample sizes slightly improved parameter recovery, but neither variation in starting values nor missing data affected it.
-In contrast, the accuracy of the model was linearly descreasing with more data missing but not influenced by manipulating parameters, noise, or starting value variation.
+In contrast, the accuracy of the model was linearly decreasing with more data missing but not influenced by manipulating parameters, noise, or starting value variation.
 Adding more states to the HMM generally decreased the parameter recovery and the accuracy.
@@
 It was relatively low in the moving dots condition and second-rate in the image and video conditions compared to NH and LNS.
-The agreement for smooth puruits was moderate in the moving dots condition, fair in the video condition, and slight in the image condition.
+The agreement for smooth pursuits was moderate in the moving dots condition, fair in the video condition, and slight in the image condition.
 No other algorithm in the study was designed to detect smooth pursuits.
@@
 An HMM classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
-A simulation study revealed that the generative model of gazeHMM recovered parameters and hidden state sequence well with a few exceptions.
+A simulation study revealed that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions.
 Adding smooth pursuits to the model and noise to the generated data were the most critical factors for decreasing recovery.
@@
 I applied gazeHMM with different numbers of states to benchmark data and compared the model fit, showing that a five-state HMM had consistently most likely generated the data.
-Thus, the model comparison provided evidence against my first hypothesis that the events in gazeHMM will match theoretical expectations.
+Thus, the model comparison provided evidence against my first hypothesis that the number of events in gazeHMM matches theoretical expectations.
 When comparing gazeHMM with four events against other algorithms and human coders, it did not show the best performance for the duration and number of classified events nor the sample-to-sample agreement across all conditions.
@@
 The model comparison on the benchmark data suggested that gazeHMM is not the best generative model for eye movement data.
-Even using a model selection criterion that identified correct models succesfully in previous studies [@Pohle2017], did not yield the expected pattern.
+Even using a model selection criterion that identified correct models successfully in previous studies [@Pohle2017], did not yield the expected pattern.
 There are several explanations for this result:
@@
 Since the subevents are usually not interesting for users of classification algorithms, the ability of HMMs to classify might limit their ability to describe eye movements here.
-*Model selection criteria are generally not appropriate for comparing HMMs with differnt numbers of states*.
+*Model selection criteria are generally not appropriate for comparing HMMs with different numbers of states*.
 This argument has been discussed in the field of ecology [see @Li2017], where studies found that selection criteria preferred models with more states than expected [similar to the result of this study; e.g., @Langrock2015].
@@
 However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yielded different results.
-The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events.
+The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events (incl.
+blinks).
 Nevertheless, Cohen's kappa values of 0.72 (saccades - video - four states) or 0.68 (fixations - image - three states) indicate substantial agreement to human coders, especially in light of the maximum reference of ~0.8 from the simulation study.
 At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
-Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and finding errors in their design.
+Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and to find errors in their design.
 
@@
 ## Future Directions
-Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for an example, see @Balakrishnan2009].
+Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for examples, see @Balakrishnan2009].
 Potential problem of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters).
@@
 For smooth pursuits, the classification is not optimal and thus not recommended, yet.
-On one hand, the algorithm has some advantages over concurrent event classfication algorithms, not relying on human-labeled input being the most important one.
+On one hand, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one.
 On the other hand, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events.
-These difficulties highlight challenges for HMMs in general, but also specifically for eye movements.
-Revisting my research question, I conclude that gazeHMM is suitable for classifying eye movement events but still lacks the ability to accurately describe them.
+These difficulties highlight challenges for HMMs in general but also specifically for eye movements.
+Revisiting my research question, I conclude that gazeHMM is suitable for classifying eye movement events but still lacks the ability to accurately describe them.
 

### 2020-06-28 validation/validation_Andersson2017.R
adapt paths for reproducibility

@@ -69,17 +69,15 @@ onestate_HMM <- function(data, respstart, sf = c(10, 10), random.respstart = T, 
 
 # Load data
 
-dir <- "~/Uni/Psychologie Master/Internship/Data/Test data/Andersson et al. (2017)/annotated_data/data used in the article"
-
 A2017 <- list()
 
 for(i in c("dots", "img", "video")) {
   
-  filenames.MN <- list.files(path = paste(dir, "/", i, "/", sep = ""), pattern = "_MN")
-  filenames.RA <- list.files(path = paste(dir, "/", i, "/", sep = ""), pattern = "_RA")
+  filenames.MN <- list.files(path = paste(here("validation/data"), "/", i, "/", sep = ""), pattern = "_MN")
+  filenames.RA <- list.files(path = paste(here("validation/data"), "/", i, "/", sep = ""), pattern = "_RA")
   
-  data.MN <- lapply(paste(dir, "/", i, "/", filenames.MN, sep = ""), function(x) readMat(x)$ETdata[[1]])
-  data.RA <- lapply(paste(dir, "/", i, "/", filenames.RA, sep = ""), function(x) readMat(x)$ETdata[[1]])
+  data.MN <- lapply(paste(here("validation/data"), "/", i, "/", filenames.MN, sep = ""), function(x) readMat(x)$ETdata[[1]])
+  data.RA <- lapply(paste(here("validation/data"), "/", i, "/", filenames.RA, sep = ""), function(x) readMat(x)$ETdata[[1]])
   
   A2017[[i]] <- lapply(1:length(data.MN), function(i, x, y) {
 
@@ -93,7 +91,7 @@ for(i in c("dots", "img", "video")) {
   }, x = data.MN, y = data.RA)
 }
 
-save(A2017, file = "validation/Andersson2017_raw.Rdata")
+save(A2017, file = here("validation/Andersson2017_raw.Rdata"))
 
 
 # Classify data
@@ -152,213 +150,4 @@ A2017.fit <- lapply(1:length(A2017), function(stim) parLapply(clust, 1:length(A2
 
 stopCluster(clust)
 
-save("A2017.fit", file = "validation/Andersson2017_fitted.Rdata")
-
-
-# Compare models with different states
-
-schwarz.weights <- function(bic, na.rm = T) {
-  
-  d.bic <- bic - min(bic, na.rm = na.rm) # eq 2
-  
[…]

### 2020-06-28 validation/validation_Ehinger2019.R

@@ -48,23 +48,16 @@ onestate_HMM <- function(data, respstart, sf = c(10, 10)) {
 
 # Load Eyelink data
 
-dir <- "~/Uni/Psychologie Master/Internship/Data/Test data/Ehinger et al. (2019)/"
-
-
 # # Convert .EDF files into .asc files
 # 
-# setwd("~/Uni/Psychologie Master/Internship/Data/Test data/Ehinger et al. (2019)/")
-# 
-# Sys.getenv("C:/Programme (x86)/SR Research/EyeLink/EDF_Access_API/Example", names = T)
-# 
-# edf.filenames <- list.files(path = getwd(), pattern = ".EDF")
+# edf.filenames <- list.files(path = here("validation/data"), pattern = ".EDF")
 # 
 # converted <- edf2asc(edf.filenames)
 
   
-filenames.el <- list.files(path = dir, pattern = ".asc")
+filenames.el <- list.files(path = here("validation/data"), pattern = ".asc")
   
-E2019 <- lapply(paste(dir, "/", filenames.el, sep = ""), function(x) { 
+E2019 <- lapply(paste(here("validation/data"), "/", filenames.el, sep = ""), function(x) { 
   
   data <- read.asc(x) 
   
@@ -127,16 +120,4 @@ E2019.fit <- parLapply(clust, E2019, function(subj) lapply(subj, function(task) 
 
 stopCluster(clust)
 
-save("E2019.fit", file = "validation/Ehinger2019_fitted.Rdata")
-
-
-# Compare models with different states
-
-schwarz.weights <- function(bic, na.rm = T) {
-  
-  d.bic <- bic - min(bic, na.rm = na.rm) # eq 2
-  
-  exp(-0.5 * d.bic)/sum(exp(-0.5 * d.bic), na.rm = na.rm) # eq 4
-  
-}
-
+save("E2019.fit", file = here("validation/Ehinger2019_fitted.Rdata"))

### 2020-06-28 manuscript/internship_report.Rmd
minor fixes

@@
 author: 
-  - name          : "Malte Valentin Lueken"
+  - name          : "Malte Valentin Lüken"
 
@@
 Both the simulation study and benchmark application showed poor classification performance for smooth pursuits.
-Thus, I advice to classify smooth pursuits only for exploration and recommend gazeHMM with three events for application.
+Thus, I advice to classify smooth pursuits only for exploration and recommend gazeHMM with fixations, saccades, and PSOs for application.
 Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.
@@
 
-floatsintext      : no
+floatsintext      : yes
 figurelist        : yes
@@
 On this basis, several classification algorithms using HMMs have been developed.
-One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named identification by HMM [I-HMM]).
+One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]).
 Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM.
@@
 plotList[[4]] <- plotList[[4]] + xlab("Sample") +
-  geom_segment(aes(x = 1, y = 180, xend = 4, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 5, y = 180, xend = 9, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 10, y = 180, xend = 27, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 28, y = 180, xend = 29, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 30, y = 180, xend = 40, yend = 180), size = 1.2)
+  geom_segment(aes(x = 1, y = 180, xend = 7, yend = 180), size = 1.2) +
+  geom_segment(aes(x = 8, y = 180, xend = 16, yend = 180), size = 1.2) +
+  geom_segment(aes(x = 17, y = 180, xend = 40, yend = 180), size = 1.2)
 
@@
 
-```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)", fig.height=3, fig.width=5}
-
-knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 300)
+```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)", fig.height=2, fig.width=4}
+
+knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 200)
 
@@
 They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
+Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes.
+Those samples in between typically follow the same direction.
+Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees.
 Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
+Other algorithms focus exclusively on classifying microsaccades (e.g., @Mihali2017), but as proposed earlier, they are not in the scope of gazeHMM.
 
@@
 Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.
-In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model [@Visser2019].
+In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model.
+The initial state model contains probabilites for the first state of the hidden sequence.
+The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM.
+The response model encompasses distributions describing the response variables for every state in the model.
 In gazeHMM, the response model has three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal.
@@
 ## Data Analysis [^6]
-For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values[^1]: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
+For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
+Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\psi_{true} \ne 0$.
+This was not the case for some mean parameters of the von Mises distributions.
+In those cases, I used $\psi_{true} = 2\pi$ instead.
 I treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter.
@@
 [^6]: Note that this section deviates from my proposed analysis plan. However, since I preregistered the changed analysis plan, I treated the results of the simulation as confirmatory.
-[^1]: This measure is only appropriate when $\psi_{true} \ne 0$. This was not the case for some mean parameters of the von Mises distributions. In those cases, I used $\psi_{true} = 2\pi$ instead.
-
-(ref:sim-ranges) Approximate ranges of response variables used to generate parameter values
+
+(ref:sim-ranges) Approximate Ranges of Response Variables Used to Generate Parameter Values in the Simulation
 (ref:sim-ranges-note) Units are deg/s (velocity), deg/s^2^ (acceleration), and radians (angle).
@@
 
-(ref:sim-parameters-trans) Intervals and default parameter values for the response models of the HMM in the simulation
+(ref:sim-parameters-trans) Intervals and Default Parameter Values for the Transition Model in the Simulation
 (ref:sim-parameters-trans-note) The transition probability for staying in the same state is denoted by $a_{i=j}$ and the probability for switching to a different state by $a_{i\neq j}$.
@@
 
-(ref:sim-parameters-resp) Intervals and default parameter values for the response models of the HMM in the simulation
+(ref:sim-parameters-resp) Intervals and Default Parameter Values for the Response Model in the Simulation
 (ref:sim-parameters-resp-note) Shape parameters are denoted by $\alpha$, scale parameters by $\beta$, mean parameters by $\mu$, and concentration parameters by $\kappa$.
@@
     scale_x_discrete(name = "Estimated parameter", labels = names.pars.est) +
-    scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2)) +
+    scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2, 3, 4, 5)) +
     geom_hline(yintercept = 0.1, linetype = "dashed") +
@@
 Thus, the model is accurate but has oppositely estimated states compared to the true states.
-In this case, an HMM with three states that is perfectly accurate but has one label switched would have a Cohen's kappa of 0.
+In this case, an HMM with three states that is perfectly accurate but has two labels switched would have a Cohen's kappa of 0.
 To test if label switching occurred, I manually switched one or two labels post-hoc (see Figure \@ref(fig:plot-acc-3-expl)).
@@
 
-(ref:tab-starting-values) Starting values for response models in fitting gazeHMM to benchmark data sets
+(ref:tab-starting-values) Starting Values for the Response Model for Fitting gazeHMM to Benchmark Data
 (ref:tab-starting-values-note) Starting values for velocity and acceleration signals are shown before scaling down by factor 100.
@@
 To examine whether gazeHMM accurately describes eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
+The starting values for the fifth state were chosen to represent distributions for microsaccade samples (velocity and sample-to-sample angle as for smooth pursuits, acceleration as for saccades and PSOs).
 Schwarz weights are the result of transforming a set of BIC values [@Wagenmakers2004].
@@
 The video condition had the same pattern as the image condition, as the five-state model consistently had the highest weights (see Figure \@ref(fig:plot-schwarz-weights-vid)).
+In all three conditions, response parameters of the fifth state were mostly similar to those of the smooth pursuit state.
 In contrast to my hypothesis, applying different HMMs to the @Andersson2017 data suggests that the five-state model has most likely generated the data.
 Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
-For the @Ehinger2019 data, I predicted a one-state model to be preferred. Here, the model comparison for the yielded similar results (see Appendix A), contradicting my hypothesis.  
+For the @Ehinger2019 data, I predicted a one-state model to be preferred.
+Here, the model comparison for the yielded similar results (see Appendix A), contradicting my hypothesis.
 A recent model recovery study showed that the BIC tended to prefer overly complex HMMs when they were misspecified [e.g., the conditional independence assumption was violated; @Pohle2017].
@@
 
-ref.acc <- matrix(c(0.92, 0.81, 0.83, 0.95, 0.91, 0.94, 0.88, 0.82, 0.83, 0.00, 0.00, 0.00,
-                    0.92, 0.84, 0.82, 0.95, 0.91, 0.94, 0.88, 0.80, 0.81, 0.00, 0.00, 0.00,
-                    0.38, 0.06, 0.11, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.00, 0.00, 0.00, 0.64, 0.66, 0.67, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.36, 0.00, 0.03, 0.45, 0.26, 0.38, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.63, 0.03, 0.14, 0.58, 0.46, 0.59, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.38, 0.00, 0.03, 0.54, 0.30, 0.52, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.67, 0.03, 0.13, 0.69, 0.60, 0.71, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.67, 0.03, 0.13, 0.75, 0.63, 0.76, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.52, 0.00, 0.01, 0.67, 0.60, 0.68, 0.24, 0.20, 0.25, 0.00, 0.00, 0.00,
-                    0.67, 0.03, 0.14, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
-                    0.00, 0.00, 0.00, 0.81, 0.75, 0.81, 0.64, 0.59, 0.63, 0.00, 0.00, 0.00), ncol = 12, byrow = T)
-
-ref.acc[ref.acc == 0] <- NA
+ref.acc <- matrix(c(0.92, 0.81, 0.83, 0.95, 0.91, 0.94, 0.88, 0.82, 0.83, NA, NA, NA,
+                    0.92, 0.84, 0.82, 0.95, 0.91, 0.94, 0.88, 0.80, 0.81, NA, NA, NA,
+                    0.38, 0.06, 0.11, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.00, 0.00, 0.00, 0.64, 0.66, 0.67, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.36, 0.00, 0.03, 0.45, 0.26, 0.38, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.63, 0.03, 0.14, 0.58, 0.46, 0.59, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.38, 0.00, 0.03, 0.54, 0.30, 0.52, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.67, 0.03, 0.13, 0.69, 0.60, 0.71, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.67, 0.03, 0.13, 0.75, 0.63, 0.76, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.52, 0.00, 0.01, 0.67, 0.60, 0.68, 0.24, 0.20, 0.25, NA, NA, NA,
+                    0.67, 0.03, 0.14, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, NA, NA, NA,
+                    0.00, 0.00, 0.00, 0.81, 0.75, 0.81, 0.64, 0.59, 0.63, NA, NA, NA), ncol = 12, byrow = T)
 
@@
 The duration of fixations and smooth pursuits was largely underestimated and the number of these two events largely overestimated.
-For saccades and PSOs, gazeHMM estimated relatively accurate durations and event numbers, but didn't reach the best algorithms.
-
-(ref:tab-rmsd-4-fix) Mean and standard deviations of fixation durations, number of fixations, and RMSD between fixation durations classified by algorithms and human coders
+For saccades and PSOs, gazeHMM estimated relatively accurate durations and event numbers, but did not reach the best algorithms.
+
+(ref:tab-rmsd-4-fix) Fixation Duration Descriptives and RMSD Between Algorithms and Human Coders
 (ref:tab-rmsd-4-fix-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
@@
 
-(ref:tab-rmsd-4-sac) Mean and standard deviations of saccade durations, number of saccades, and RMSD between saccade durations classified by algorithms and human coders
+(ref:tab-rmsd-4-sac) Saccade Duration Descriptives and RMSD Between Algorithms and Human Coders
 (ref:tab-rmsd-4-sac-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
@@
 
-(ref:tab-rmsd-4-pso) Mean and standard deviations of PSO durations, number of PSOs, and RMSD between PSO durations classified by algorithms and human coders
+(ref:tab-rmsd-4-pso) PSO Duration Descriptives and RMSD Between Algorithms and Human Coders
 (ref:tab-rmsd-4-pso-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
@@
 
-(ref:tab-rmsd-4-sp) Mean and standard deviations of smooth pursuit durations, number of smooth pursuits, and RMSD between smooth pursuit durations classified by algorithms and human coders
+(ref:tab-rmsd-4-sp) Smooth Pursuit Duration Descriptives and RMSD Between Algorithms and Human Coders
 (ref:tab-rmsd-4-sp-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
@@
 
-(ref:tab-kappa-4) Cohen's kappa between human coders and algorithms for different conditions and events
+(ref:tab-kappa-4) Cohen's Kappa Between Human Coders and Algorithms for Different Conditions and Events
 (ref:tab-kappa-4-note) Negative values were set to zero. gazeHMM classified four events. Table adapted from @Andersson2017.
@@
 At last, I computed the overall disagreement [^4] as the percentage of all samples taken together that were classified differently by gazeHMM than the human coders.
-Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events had a higher overall disagreement than all but two of the other algorithms.
-In the moving dots and video conditions, the disagreement was considerably lower than for all other algorithms.
+Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events had a higher overall disagreement by between 10 and 20% than all but two of the other algorithms.
+In the moving dots and video conditions, the disagreement was between 20 and 30% lower than for all other algorithms.
 Notably, gazeHMM's disagreement was similar across conditions (around 45%).
@@
 
-(ref:tab-conf-4) Confusion matrix between gazeHMM (rows) and human coders (columns) for different events and conditions
+(ref:tab-conf-4) Confusion Matrix Between gazeHMM (Rows) and Human Coders (Columns) for Different Conditions
 (ref:tab-conf-4-note) gazeHMM classified four events and blinks.
@@
 Figure \@ref(fig:plot-disag) illustrates the overall disagreement between gazeHMM with three events and the human coders compared to the other algorithms.
-In all three conditions, the disagreement was slightly lower for gazeHMM than for all the other algorithms.
+In all three conditions, the disagreement was slightly lower for gazeHMM than for all the other algorithms (5-10% for images and moving dots, around 20% for videos).
 Compared to gazeHMM with four states, the RMSD for fixations decreased while the sample-to-sample agreement increased.
@@
 Overall, applying gazeHMM with three states to static data seems more appropriate, since the increase in fixation classification outweighs the decrease in classification of saccades and PSOs.
-There was support for my second hypothesis since gazeHMM had lower overall disagreement to human coders than all other algorithms.
-
-(ref:tab-rmsd-3) Mean and standard deviation of event durations, number of events, and RMSD between event durations classified by algorithms and human coders for the image condition
+There was support for my second hypothesis since gazeHMM had lower overall disagreement to human coders, ranging between 5 and 20% across conditions, than all other algorithms.
+
+(ref:tab-rmsd-3) Event Duration Descriptives and RMSD between Algorithms and Human Coders for the Image Condition
 (ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
@@
 
-(ref:tab-kappa-3) Cohen's kappa between human coders and algorithms for different events in the image condition
+(ref:tab-kappa-3) Cohen's Kappa Between Human Coders and Algorithms for Different Events in the Image Condition
 (ref:tab-kappa-3-note) Negative values were set to zero. gazeHMM-3 classified three and gazeHMM-4 classified four events. Table adapted from @Andersson2017.

### 2020-06-28 manuscript/report_appendix.Rmd
minor fixes

@@
 
-(ref:tab-example-trans) Transition probabability estimates for example data from Andersson et al. (2017)
+(ref:tab-example-trans) Transition Probabability Estimates for Example Data from Andersson et al. (2017)
 
@@
 
-(ref:tab-example-resp) Response parameter estimates for example data from Andersson et al. (2017)
+(ref:tab-example-resp) Response Parameter Estimates for Example Data from Andersson et al. (2017)
 

### 2020-06-29 algorithm/gazeHMM.R
add blink window default value

@@ -4,7 +4,7 @@
 # Date: 24.04.2020
 
 
-gazeHMM <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL, b.win,
+gazeHMM <- function(x, y, t, unit = "px", res, dim, dist, fr, blink = NULL, b.win = 0.05,
                     sg.order = 3, sg.length = 5, 
                     nstates, respstart, trstart, instart, sf = c(10, 10), 
                     random.respstart = T, start.seed = NULL,

### 2020-06-29 validation/validation_Ehinger2019.R

@@ -8,6 +8,7 @@ library(depmixS4)
 library(parallel)
 library(FDBeye)
 library(eyelinker)
+library(here)
 
 source("algorithm/gazeHMM.R")
 source("algorithm/model_helper_functions.R")
@@ -97,19 +98,20 @@ E2019.fit <- parLapply(clust, E2019, function(subj) lapply(subj, function(task) 
   
   df$time <- (df$time - df$time[1])/1e3
   
-  trstart <- matrix(0.1/(k-1), k, k)
-  diag(trstart) <- 0.9
-  
   fit <- list()
   
   fit[["1"]] <- try(onestate_HMM(preprocess(x = df$xp, y = df$yp, t = df$time, unit = "px",
                                             res = res, dim = dim, dist = dist, fr = fr, blink = is.na(df$ps),
-                                            sg.order = 3, sg.length = 5, sf = c(100, 100))))
+                                            sg.order = 3, sg.length = 5)))
   
   for (k in 2:5) {
-
+    
+    trstart <- matrix(0.1/(k-1), k, k)
+    diag(trstart) <- 0.9
+    
     fit[[as.character(k)]] <- try(gazeHMM(x = df$xp, y = df$yp, t = df$time, unit = "px",
-                                          res = res, dim = dim, dist = dist, fr = fr, blink = is.na(df$ps),
+                                          res = res, dim = dim, dist = dist, fr = fr, sf = c(100, 100), 
+                                          blink = is.na(df$ps),
                                           nstates = k, trstart = trstart, 
                                           random.respstart = F))
 

### 2020-06-29 manuscript/internship_report.Rmd
last fixes

@@
 keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
-wordcount         : "12808"
+wordcount         : "12962"
 
@@
 The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and angle (named NSLR-HMM).
-Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named Bayesian microsaccade detection [BMD]).
+Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]).
 Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).
@@
 The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events.
-Therefore, the authors suggest using it as an exploratory tool followed by further event classification.
+Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].
 HMMs can also be used as a generative model to make inferences about the underlying processes of eye movements, a property that has been rarely used in the context of classification [cf.
@@
 
-knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 200)
+knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 400)
 
@@
 Therefore, its classification method is oriented to four major goals: (a) it does not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it covers the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it confirms rather than explores the presence of events in the data.
-Besides implementing the algorithm in the statistical programming language R @RCoreTeam2020, I will develop its concrete technical design exploratorily.
+Besides implementing the algorithm in the statistical programming language R [@RCoreTeam2020], I will develop its concrete technical design exploratorily.
 The first step will be choosing an appropriate metric that allows separating the gaze data into events.
@@
 Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
-Other algorithms focus exclusively on classifying microsaccades (e.g., @Mihali2017), but as proposed earlier, they are not in the scope of gazeHMM.
+Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as proposed earlier, they are not in the scope of gazeHMM.
 
@@
 ## The Generative Model
-The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM).
+The generative model underlying gazeHMM is a multivariate HMM.
 It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
@@
 (ref:sim-ranges) Approximate Ranges of Response Variables Used to Generate Parameter Values in the Simulation
-(ref:sim-ranges-note) Units are deg/s (velocity), deg/s^2^ (acceleration), and radians (angle).
+(ref:sim-ranges-note) Units are deg/s (velocity), deg/s (acceleration), and radians (angle).
 ~ indicates that the distribution has a peak at this value.
@@
   p <- ggplot(data = data.long, aes(x = par.est, y = RMdSPD, color = cond)) +
-    geom_point(position = position_dodge(0.25)) + facet_grid(cols = vars(state.est)) +
+    geom_point(position = position_dodge(0.25)) + facet_wrap(vars(state.est)) +
     scale_x_discrete(name = "Estimated parameter", labels = names.pars.est) +
-    scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2, 3, 4, 5)) +
+    scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2, 2.5, 3, 4, 5)) +
     geom_hline(yintercept = 0.1, linetype = "dashed") +
@@
 
-print(plots.rmsd.234[[1]][[1]] + theme(axis.text = element_text(size = 10)))
+print(plots.rmsd.234[[1]][[1]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-print(plots.rmsd.234[[1]][[2]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[1]][[2]] + theme(axis.text = element_text(size = 6)))
 
@@
 
-print(plots.rmsd.234[[1]][[3]] + theme(axis.text = element_text(size = 6)))
+print(plots.rmsd.234[[1]][[3]] + theme(axis.text = element_text(size = 8)))
 
@@
 As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33 (see Figure \@ref(fig:plot-acc-3)).
-Regarding the four state-HMM, RMdSPS of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1 (see Figure \@ref(fig:plot-rmdspd-3-4)).
+Regarding the four state-HMM, RMdSPDs of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1 (see Figure \@ref(fig:plot-rmdspd-3-4)).
 The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero (see Figure \@ref(fig:plot-acc-3)).
@@
 
-print(plots.rmsd.234[[2]][[1]] + theme(axis.text = element_text(size = 10)))
+print(plots.rmsd.234[[2]][[1]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-print(plots.acc.234[[2]] + theme(axis.text = element_text(size = 8)))
+print(plots.acc.234[[2]])
 
@@
 
-print(plots.rmsd.234[[2]][[2]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[2]][[2]] + theme(axis.text = element_text(size = 6)))
 
@@
 
-print(plots.rmsd.234[[2]][[3]] + theme(axis.text = element_text(size = 6)))
+print(plots.rmsd.234[[2]][[3]] + theme(axis.text = element_text(size = 8)))
 
@@
 Colours indicate the variation in starting values used to estimate parameters.
-Solid vertical lines symbolize medians and hinges the first and third quartile.
+Solid lines symbolize medians and hinges the first and third quartile.
 Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
@@
 
-print(plots.acc.expl[[1]] + theme(axis.text = element_text(size = 8)))
+print(plots.acc.expl[[1]])
 
@@
 
-print(plots.rmsd.234[[3]][[1]] + theme(axis.text = element_text(size = 10)))
+print(plots.rmsd.234[[3]][[1]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-print(plots.rmsd.234[[3]][[2]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[3]][[2]] + theme(axis.text = element_text(size = 6)))
 
@@
 
-print(plots.rmsd.234[[3]][[3]] + theme(axis.text = element_text(size = 6)))
+print(plots.rmsd.234[[3]][[3]] + theme(axis.text = element_text(size = 8)))
 
@@
 
-(ref:tab-rmsd-3) Event Duration Descriptives and RMSD between Algorithms and Human Coders for the Image Condition
+(ref:tab-rmsd-3) Event Duration Descriptives and RMSD Between Algorithms and Human Coders for the Image Condition
 (ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
@@
 Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for examples, see @Balakrishnan2009].
-Potential problem of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters).
+Potential problems of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters).
 Another option to model the correlation could be to include one of the response variables as a covariate of the other.

### 2020-07-13 manuscript/paper_BRM.Rmd
shorten results sections

--- 2020-06-29 manuscript/internship_report.Rmd
@@
 author: 
-  - name          : "Malte Valentin Lüken"
+  - name          : "Malte Lueken"
+    affiliation   : "1"
+    corresponding : yes
+    address       : "Postal address"
+    email         : "malte_lueken@arcor.de"
+  - name          : "Simon Kucharsky"
+    affiliation   : "1"
+  - name          : "Ingmar Visser"
+    affiliation   : "1"
+
+affiliation:
+  - id            : "1"
+    institution   : "Department of Psychology, University of Amsterdam. Amsterdam, The Netherlands"
 
 authornote: |
-  Affiliation: University of Amsterdam, Graduate School of Psychology, Department of Methodology and Statistics.
-
-  This manuscript constitutes the internship report as part of the Research Master's Psychology.
-
-  Student number: 12750166. Email address: malte.luken@student.uva.nl.
-
-  Specialization: Methodology and Statistics. Supervisor: dhr. dr. Ingmar Visser.
-
-  I would like to thank Simon Kucharsky and Ingmar Visser for their helpful advice on developing and implementing this research project.
+  This is an author note.
 
@@
 bibliography      : "report.bib"
-appendix          : 
- - "report_appendix.Rmd"
- - "report_technical_appendix.Rmd"
 
@@
 # Introduction
-For how long did humans seek to open a window into each other's minds, revealing their hidden cognitive processes?
-While many methods aiming to achieve this have been developed in the last decades, eye-tracking as one of them has become more and more reliable, accurate, and widely applied in cognitive psychology [@Duchowski2017].
-Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017].
-Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015].
-Commonly distinguished events are fixations, saccades, smooth pursuit, and post-saccadic oscillations (PSOs).
-Classifying eye-tracking data reduces their complexity and is the first step towards cognitive interpretation [@Andersson2017; @Salvucci2000].
-The classification is normally done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017].
-@Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most classification algorithms consist of four steps: First, the raw gaze position is transformed into a metric of position and/or time.
-Second, a filtering or smoothing procedure is applied to the data to separate the gaze signal from noise and artifacts [@Spakov2012].
-Third, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
-Fourth, the algorithm decides which candidates to accept, relabel, or merge [@Hessels2017; @Komogortsev2010].
-A frequently implemented method uses probabilistic models to classify events.
-Instead of discrete classification, these models assign probabilities for belonging to an event to each sample.
-Probabilistic models have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
-324].
-One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs).
-Figure \@ref(fig:plot-hmm-structure) illustrates the structure of HMMs: Parallel to the gaze signal evolves a sequence of distinct states that cannot be directly observed.
-Each gaze sample depends on its corresponding state.
-Each state depends on the previous but not on earlier states of the sequence [@Cappe2005; @Visser2019].
-In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as events and gaze data are dependent time series.
-On this basis, several classification algorithms using HMMs have been developed.
-One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]).
-Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM.
-Recently, @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR).
-The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and angle (named NSLR-HMM).
-Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]).
-Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).
-Several studies have tested the performance of HMM algorithms against other classification methods: I-HMM has been deemed as robust against noise, behaviorally accurate, and showing a high sample-to-sample agreement with a human coder [@Andersson2017; @Komogortsev2010; @Salvucci2000].
-However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012].
-NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].
-In sum, HMMs seem to be a promising method for classifying eye movements.
-Still, the existing HMM algorithms each have their weaknesses that could be improved upon (and that could explain their inferior performance in comparison to other methods).
-First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
-Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability.
-It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals.
-Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017].
-The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events.
-Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].
-HMMs can also be used as a generative model to make inferences about the underlying processes of eye movements, a property that has been rarely used in the context of classification [cf.
-@Mihali2017].
-For example, an HMM with four events could be compared to one with three events to see whether an event is present in the data.
-I propose to develop an algorithm that uses HMMs to classify eye movement data. It will address the weaknesses of previous HMM algorithms and utilize the generative property of HMMs.
 
@@
 
-(ref:plot-algorithm-workflow) Example workflow for eye movement event classification algorithms: (a) the raw gaze signal is transformed into a velocity signal (in deg/s); (b) the velocity signal is smoothed and filtered; (c) samples are labeled as events (indicated by colors), and (d) relabeled.
-Sequences of samples belonging to the same event are merged (indicated by black segments).
-
-```{r plot-algorithm-workflow, fig.cap="(ref:plot-algorithm-workflow)", fig.height = 8}
-
-set.seed(126)
-
-velRaw <- sample(c(rgamma(25, 3, 1/6), rgamma(7, 25, 1/5), rgamma(8, 15, 1/5)), 40)
-
-sample <- 1:length(velRaw)
-
-velSmooth <- as.numeric(smooth(velRaw, kind = "3R"))
-
-eventPre <- ifelse(velSmooth > 40, "s", "f")
-
-eventPost <- ifelse(velSmooth < 60, "f", eventPre)
-
-df <- data.frame(sample, velRaw, velSmooth, eventPre, eventPost)
-
-# Create plot
-
-p1 <- ggplot(df, aes(sample, velRaw)) + geom_point()
-
-p2 <- ggplot(df, aes(sample, velSmooth)) + geom_point()
-
-p3 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPre))
-
-p4 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPost))
-
-plotList <- list(p1, p2, p3, p4)
-
-for(p in 1:length(plotList)) {
-
-  plotList[[p]] <- plotList[[p]] + scale_y_continuous(name = "Velocity\n(deg/s)", 
-                                                      limits = c(0, 200), breaks = c(0, 100, 200)) + 
-    xlab("") + geom_line(color = "grey") + ggtitle(letters[p]) + 
-    theme(legend.position = "none")
-
-}
-
-plotList[[4]] <- plotList[[4]] + xlab("Sample") +
-  geom_segment(aes(x = 1, y = 180, xend = 7, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 8, y = 180, xend = 16, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 17, y = 180, xend = 40, yend = 180), size = 1.2)
-
-grid.arrange(plotList[[1]], plotList[[2]], plotList[[3]], plotList[[4]], ncol = 1)
-
-```
-
-(ref:plot-hmm-structure) Example structure of HMMs: The observed velocity signal *V* depends on hidden states *S* (displayed for six samples).
-Each state depends on the previous state.
-Colors indicate different states.
-Figure inspired by Visser and Speekenbrink [-@Visser2019, p.
-133].
-
-```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)", fig.height=2, fig.width=4}
-
-knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 400)
-
-```
-
-## Research Question
-Can HMMs improve describing and classifying eye-tracking data?
-This is the general research question of my proposal and can be specified as following: Are HMMs useful generative models for recorded eye movements that improve classifying them into events?
-This research question will guide the development of the algorithm and leads to two hypotheses: First, I predict that HMMs are useful generative models for eye movements.
-This hypothesis will be supported if the algorithm yields events that match theoretical expectations for benchmark data.
-Other results will be treated as evidence against the hypothesis.
-Second, I hypothesize that HMMs will improve the classification performance of eye movements.
-I will treat it as support for this hypothesis if the algorithm outperforms algorithms using other methods when applied to benchmark data.
-I will treat other results as evidence against the hypothesis.
-
-# Operationalization
-The proposed algorithm aims to improve the drawbacks of previous HMM algorithms.
-Therefore, its classification method is oriented to four major goals: (a) it does not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it covers the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it confirms rather than explores the presence of events in the data.
-Besides implementing the algorithm in the statistical programming language R [@RCoreTeam2020], I will develop its concrete technical design exploratorily.
-The first step will be choosing an appropriate metric that allows separating the gaze data into events.
-@Zemblys2018 provide an overview of different gaze data metrics and compare how well they predict human-labeled events.
-The next step will choosing a filtering/smoothing method to reduce the noise in the gaze data [for a comparison of different gaze data filters, see @Spakov2012].
-The criteria for choosing a metric and a filter will be its accuracy and not relying on user parameter settings.
-The generative part of the algorithm will allow users to build an HMM which matches their theoretical assumptions about the gaze data.
-The algorithm will take a preprocessed gaze data metric (e.g., filtered velocity) as input and optimize the model parameters.
-The output of the model will be probabilities for hidden states.
-The states can be interpreted as event labels for each sample.
-For instance, the model could assign to a sample a probability of 0.6 for belonging to a fixation and 0.4 for belonging to a saccade.
-These probabilities can be used to compute the hidden state sequence (e.g., by choosing events with the highest probability).
-Moreover, the model will output the parameters of the distribution of labeled samples for each event.
-It will allow generating samples from these distributions to simulate gaze data.
-Before testing the hypotheses, I will conduct a simulation study to estimate the parameter recovery of the generative model: Artificial eye movement data will be generated by the model.
-The parameters of the model will be varied across sensible ranges.
-Then, the model will be applied to the data.
-Finally, the estimated results of the model will be compared to the true parameters that have been used to generate the data.
-The first hypothesis will be tested by comparing different generative models on benchmark data.
-The models will differ in their assumed number of events in the data.
-The model comparison will result in a preferred model which scores best on the evaluation criterion.
-The preferred model is expected to match the theoretically assumed number of present events in the benchmark data.
-To test the second hypothesis, the proposed algorithm will be compared against other algorithms that have also been applied to benchmark data.
-The performance of the proposed algorithm will be assessed by metrics that are commonly used in classification (e.g., Cohen's Kappa).
-
-## Sample Characteristics
-The hypotheses will be tested on two benchmark data sets that have been published in previous studies.
-The first set was published in @Andersson2017 and contains eye-tracking data from 17 subjects.
-Three types of stimuli were presented on a screen ($1024 \times 768$ pixels) to each subject: A static image, a linearly moving dot, and a video.
-Subjects were instructed to look freely at the images and follow the movements of the dots and the objects in the videos.
-Their eye movements were recorded with a Hi-Speed 1250 tower-mounted eye-tracker (SensoMotoric Instruments) and a sampling rate of 500 Hz.
-Only data from the right eye were used.
-The data were labeled by two human experts in eye movement research as fixations, saccades, smooth pursuits, PSOs, or blinks [@Andersson2017, p.
-621].
-The second data set, published by @Ehinger2019, consists of eye-tracking data from 15 subjects who performed six blocks of 10 different experimental tasks.
-Here, I will only consider data from Tasks 4 and 5 because they are qualitatively different from the first data set.
-In Task 4, subjects were instructed to fixate a central target on the screen ($1920 \times 1080$ pixels) for 20 s.
-In Task 5, subjects also fixated a central target but were instructed to blink each time they heard a beep.
-Seven beeps with a duration of 100 ms and a 1.5 s interval in between were presented.
-For both tasks, eye-movements were recorded by two eye-trackers simultaneously (EyeLink 1000, SR Research; and Pupil Labs glasses, Pupil Labs) with a sampling rate of 250-500 Hz and 240 Hz (interpolated), respectively [@Ehinger2019, pp.
-3-11].
-
-## Data Analysis
-I will treat the parameter recovery of the generative model as satisfactory when it yields a 95% CI covering the true parameters.
-The median rate of correct classifications should be a minimum of 90%.
-The first hypothesis will be tested using a model comparison approach: For both data sets, models with one, two, three, four, and five eye movement events plus a noise event will be compared.
-The noise event will represent blinks.
-For the first data set and static images, I expect the model with three events (excl.
-noise) to be the preferred model since smooth pursuit movements are not assumed to be present.
-For moving dots and video stimuli, I expect the model with four events (excl.
-noise) to be preferred, since smooth pursuits are expected in the data.
-For the second data set, I expect a model with one event (excl.
-noise) to be preferred because only fixations should be present in the data from both tasks.
-For all comparisons, the preferred model will be chosen according to the highest Schwarz weight [@Wagenmakers2004].
-I will test the second hypothesis using the metrics that have been used in the original study of the first data set: @Andersson2017 compared ten algorithms to human-labeled data by the root mean squared error (RMSE) of event distribution descriptives, the sample-to-sample agreement indicated by Cohen's Kappa, and the disagreement ratio across all samples indicated by the confusion matrix.
-I expect the algorithm with a generative model assuming four events (excl.
-noise) to achieve either a lower RMSE, higher Cohen's Kappa, or lower disagreement ratio than all the other algorithms which are evaluated in the study.
-
-## Intended Results
-Preferred generative models that match the theoretical assumptions on the number of events in the data would support the notion that HMMs describe gaze data appropriately.
-If a generative model with fewer events than expected is preferred, this would mean that the model is not able to distinguish the events well enough.
-If more events than expected are preferred, this would indicate that the model is describing noise or an unexpected event is present in the data.
-Outperforming other algorithms in agreement to human-coded data would demonstrate that the proposed algorithm classifies events similar to human coders.
-However, @Hooge2018a observed considerable differences in how humans code gaze data.
-They concluded that human coders cannot be seen as a gold standard for event classification.
-The proposed algorithm would be the first allowing the user to conduct model comparisons as part of the classification.
-This comparison could be used to investigate the data more thoroughly on a qualitative (e.g., assumed present events) or quantitative (e.g., event sample distributions) level.
-In contrast to many other classification algorithms, the proposed algorithm would not only classify gaze data but also explain why it is classified this way by using a generative model.
-This outstanding property of the algorithm would guarantee its competitiveness against other popular algorithms, especially the recently developed machine learning approaches which do not make generative assumptions [e.g., @Bellet2019].
-It could spark a shift from pure classification output towards understanding eye movements on a more fundamental level.
-
 ```{r load data, include=FALSE}
@@
 
-# Algorithm Development
-As part of answering my research questions, I developed an algorithm named gazeHMM to classify gaze data into discrete eye movement events.
-The following section describes the development of gazeHMM and its final version that has been used to obtain results.
-Technical details can be found in Appendix B.
-
-## Eye Movement Metrics
-Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018].
-Here, my goal was to find those metrics that separate the gaze data samples belonging to different events the best.
-However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018].
-This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics.
-Because they belong to the most basic metrics which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] in gazeHMM.
-Theoretically, these three metrics should separate eye movement events clearly.
-Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
-Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017].
-In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction.
-PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013].
-They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
-Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes.
-Those samples in between typically follow the same direction.
-Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees.
-Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
-Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as proposed earlier, they are not in the scope of gazeHMM.
-
-## Filtering and Smoothing
-As with metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics.
-Their purpose is to remove noise or artifacts from the gaze data that could distort the classification [@Duchowski2017].
-Filtering (smoothing) methods that are applied before computing the eye movement metrics target the recorded gaze position.
-While removing noise, filters might also erase differences in the metrics between samples.
-For instance, tremor movements during fixations could be filtered out and in consequence, the samples would follow the same direction.
-This could complicate separating fixations from smooth pursuits.
-The same problem applies to PSOs and saccades where filtering could erase oscillations (see Figure \@ref(fig:filter-angle)).
-
-(ref:filter-angle) Example data from @Andersson2017 displayed as sample-to-sample angle (in rad) over time (in s).
-Line colors indicate whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue).
-For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3.
-The full line segment marks where potential oscillations are filtered out.
-The dashed line segment illustrates where potentially random change in angle is filtered out.
-
-```{r filter-angle, fig.cap="(ref:filter-angle)"}
-
-filter.ex <- A2017$video[[3]] %>%
-  mutate(x = px2va(x, dim[1], res[1], dist),
-         y = px2va(y, dim[2], res[2], dist),
-         angle = calc_theta(x, y),
-         x.filt = as.numeric(signal::filter(signal::butter(3, 0.3), x)),
-         y.filt = as.numeric(signal::filter(signal::butter(3, 0.3), y)),
-         angle.filt = calc_theta(x.filt, y.filt),
-         t = (t-t[1])/1e6) %>%
-  dplyr::filter(t > 0.5, t < 0.75)
-
-ggplot(filter.ex) + 
-  geom_path(aes(x = t, y = force_neg_pi_pi(angle)), color = "blue") +
-  geom_path(aes(x = t, y = force_neg_pi_pi(angle.filt)), color = "red") +
-  annotate(geom = "segment", x = 0.525, xend = 0.54, y = -3, yend = -3, color = "black", size = 0.8) +
-  annotate(geom = "segment", x = 0.635, xend = 0.67, y = -3, yend = -3, color = "black", size = 0.8, linetype = 6) +
-  scale_y_circular(name = "Sample-to-sample angle (in rad)", units = "radians", limits = c(-pi, pi)) +
-  scale_x_continuous(name = "Time (in s)")
-
-```
-
-Instead of filtering before computing the metrics, I decided to combine both in one step.
-Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration).
-@Houpt2018 computed the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm.
-Similarly, @Nystrom2010 used a Savitzky-Golay (SG) filter [@Savitzky1964] to estimate velocity and acceleration from a gaze signal.
-Figure \@ref(fig:filter-comparison) displays an example of velocity and acceleration signals obtained by both methods.
-They follow a similar trend, but the SG velocity signal was less noisy and the acceleration signal was consequently lower than the Laplace acceleration signal.
-Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
-To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle as the backward difference and the first discrete derivative as the forward difference (see Appendix B).
-
-(ref:filter-comparison) Example data from @Andersson2017 displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s).
-Line colors indicate which filter has been used to derive the signal.
-Savitzky-Golay filters were applied with order three and length five (corresponding to 10 ms).
-
-```{r filter-comparison, fig.cap="(ref:filter-comparison)"}
-
-filter.comp <- A2017$video[[3]] %>%
-  mutate(x = px2va(x, dim[1], res[1], dist),
-         y = px2va(y, dim[2], res[2], dist),
-         vel.x.sg = signal::sgolayfilt(x, m = 1),
-         vel.y.sg = signal::sgolayfilt(y, m = 1),
-         vel.sg = sqrt(vel.x.sg^2 + vel.y.sg^2)*fr,
-         acc.x.sg = signal::sgolayfilt(x, m = 2),
-         acc.y.sg = signal::sgolayfilt(y, m = 2),
-         acc.sg = sqrt(acc.x.sg^2 + acc.y.sg^2)*fr^2,
-         vel.x.sb = lead(x) - lag(x),
-         vel.y.sb = lead(y) - lag(y),
-         vel.sb = sqrt(vel.x.sb^2 + vel.y.sb^2)*fr,
-         acc.x.lp = lead(x) - 2*x + lag(x),
-         acc.y.lp = lead(y) - 2*y + lag(y),
-         acc.lp = sqrt(acc.x.lp^2 + acc.y.lp^2)*fr^2,
-         t = (t-t[1])/1e6) %>%
-  pivot_longer(c("vel.sg", "vel.sb", "acc.sg", "acc.lp"), names_to = c("metric", "filter"), names_sep = "[:punct:]", ) %>%
-  mutate_at(c("metric", "filter"), as.factor) %>%
-  dplyr::filter(t > 0.5, t < 1.5)
-
-filter.comp$metric <- factor(filter.comp$metric, labels = c("Acceleration", "Velocity"))
-
-ggplot(filter.comp, aes(x = t, y = value, color = filter)) + 
-  geom_path() + facet_wrap(vars(metric), nrow = 2, scales = "free_y", strip.position = "left") + 
-  scale_x_continuous(name = "Time (in s)") +
-  scale_y_continuous(name = "") +
-  scale_color_discrete(name = "Filter", labels = c("Laplace", "Sobel", "Savitzky-Golay")) +
-  theme(strip.placement = "outside")
-
-```
-
-## The Generative Model
-The generative model underlying gazeHMM is a multivariate HMM.
-It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
-Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.
-In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model.
-The initial state model contains probabilites for the first state of the hidden sequence.
-The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM.
-The response model encompasses distributions describing the response variables for every state in the model.
-In gazeHMM, the response model has three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal.
-The response variables are treated as conditionally independent on the states.
-Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
-This step was merely taken to keep the HMM simple and identifiable.
-Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them).
-However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation).
-Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean.
-Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties.
-In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose.
-It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
-To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution.
-Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions.
-Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution.
-Thus, the uniform distribution should distinguish fixations from the other events.
-The initial state model and the transition model use multinomial distributions (with a category for each state).
-For all three submodels of the HMM, no time-varying covariates are included.
-Since the implementation of gazeHMM allows for regression models for each submodel and state, this implies that only intercepts for each parameter are estimated.
-
-## Missing Data and Blinks
-HMMs can estimate parameters and hidden states despite missing data.
-They either treat data as missing at random or as missing depending on states [@Visser2019].
-In gazeHMM, I assume data to be missing at random.
-However, most of the missing data in eye movement classification are due to blinks.
-Therefore, the user can indicate in gazeHMM which samples should be labeled as blinks (other missing samples are treated as noise).
-Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs.
-Since these samples could distort the classification of saccades in the HMM, I decided to remove them heuristically.
-Before classifying the samples, gazeHMM sets all samples within 50 ms before and after blink samples as missing.
-The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.
-When evaluating the likelihood of the HMM, given the parameters and joint density of response distributions, for missing data, the model integrates over all possible values.
-The resulting density for missing samples is therefore set to one [@Visser2019].
-
-## Optimization and Classification
-The parameters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997].
-The EM algorithm is generally suitable to estimate likelihoods with missing variables.
-For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data (velocity, acceleration, and sample-to-sample angle) and the expected hidden states [eye movement events; @Visser2019].
-The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability.
-Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988].
-
-## Postprocessing
-After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence.
-I implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
-Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events.
-Thus, the postprocessing routine heuristically compensates for such violations.
-This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events.
-Samples are relabeled as the state of the previous event.
-Finally, samples initially indicated as missing are labeled as noise (including blinks) and event metrics are computed (e.g., fixation duration).
-
-## Implementation
-The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
-The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
-
-\newpage
-
 # Simulation Study
-To assess how well the HMM recovers parameters and state sequences, I conducted a simulation study.
+To assess how well the HMM recovers parameters and state sequences, we conducted a simulation study.
 The design and analysis of the study were preregistered on the Open Science Framework (https://doi.org/10.17605/OSF.IO/VDJGP).
-The major part of this section is a direct copy of the preregistration.
-When appropriate, additional explanations were added and tenses adapted to make the design easier to read and understand.
-In the study, the HMM repeatedly generated data with a set of parameters (true parameter values).
-The same model was then applied to estimate the parameters from the generated data (estimated parameter values).
-I compared the true with the estimated parameter values to assess whether a parameter was recovered by the model.
-Additionally, I compared the true states of the HMM with the estimated states to judge how accurately the model recovered the states that generated the data.
+This section is majorly copied from the preregistration (with adapted tenses).
+The study was divided in four parts.
+Here, we only report the first two parts, the other two parts, which address starting values and missing data, can be found in the supplementary material (URL).
+The HMM repeatedly generated data with a set of parameters (true parameter values).
+The same model was applied to estimate the parameters from the generated data (estimated parameter values).
+We compared the true with the estimated parameter values to assess whether a parameter was recovered by the model.
+Additionally, we contrasted the true states of the HMM with the estimated states to judge how accurately the model recovered the states that generated the data.
 
@@
 The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
-To generate random starting values for the estimation of shape, scale, and concentration parameters, I used gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated.
+Random starting values for the estimation of shape, scale, and concentration parameters were generated by gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$, with $\psi_{true}$ being the true value of the parameter to be estimated.
 This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
@@
 ## Design
-### Parameter Variation
-The simulation study was divided into four parts.
-In the first part, I varied the parameters of the HMM.
-For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were varied, respectively.
+In the first part, we varied the parameters of the HMM.
+For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were manipulated, respectively.
 For each parameter, the HMM generated 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps.
 This resulted in $100 \times (10+15+20) = 4500$ recoveries.
-Only one parameter was varied at once, the other parameters were set to their default values.
+Only one parameter alternated at once, the other parameters were set to their default values.
 All parameters of the HMM were estimated freely (i.e., there were no fixed parameters in the model).
-I did not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
-For the transition probabilities, I only simultaneously varied the probabilities for staying in the same state (diagonals of the transition matrix) to reduce the complexity of the simulation.
+We did not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
+For the transition probabilities, we only simultaneously changed the probabilities for staying in the same state (diagonals of the transition matrix) to reduce the complexity of the simulation.
 The left over probability mass was split evenly between the probabilities for switching to a different state (per row of the transition matrix).
-Moreover, I did not modify the mean parameters of the von Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
-I defined approximate ranges for each response variable and chose true parameter intervals and default values so that they produced samples that roughly corresponded to these ranges.
-Table \@ref(tab:sim-ranges) shows the assumed ranges for each event and Tables \@ref(tab:sim-parameters-trans) and \@ref(tab:sim-parameters-resp) show the intervals and default values for each parameter in the simulation.
+Moreover, we did not modify the mean parameters of the von Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
+We defined approximate ranges for each response variable (see supplementary material) and chose true parameter intervals and default values so that they produced samples that roughly corresponded to these ranges.
+Tables \@ref(tab:sim-parameters-trans) and \@ref(tab:sim-parameters-resp) show the intervals and default values for each parameter in the simulation.
 Parameters were scaled down by factor 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
-I set the intervals for shape parameters of the gamma distributions for all events to [1,5] to examine how skewness influenced the recovery (shape values above five approach a symmetric distribution).
+We set the intervals for shape parameters of the gamma distributions for all events to [1,5] to examine how skewness influenced the recovery (shape values above five approach a symmetric distribution).
 The scale parameters were set so that the respective distribution approximately matched the assumed ranges.
@@
 An example of simulated data from the HMM with default parameters is visualized in Figure \@ref(fig:plot-sim-ex).
-
-### Sample Size and Noise Variation
-In the second part, I varied the sample size of the generated data and the amount of noise added to it.
+In the second part, we manipulated the sample size of the generated data and the amount of noise added to it.
 The model parameters were set to their default values.
-For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, I generated 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
+For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we generated 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
 These samples sizes roughly corresponded to small, medium, and large eye-tracking data sets for a single participant and trial.
-To simulate noise, I replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
+To simulate noise, we replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
 This procedure ensured that velocity and acceleration values remained positive and were taken from moderately skewed distributions with modes equal to the original values.
-To angle, I added white noise from a von Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
-$\tau_{noise}$ and $\kappa_{noise}$ were varied simultaneously in equidistant steps in their intervals.
-
-### Variation of Starting Values
-In the third part, I increased the variation in the starting values used for parameter estimation.
-The model parameters were set to their default values.
-For the shape, scale, and concentration parameters, I simultaneously increased the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta_{start} = (\psi_{true}/2)\tau_{start}$ with $\tau_{start} \in \{1, 2, 3\}$, 100 data sets with $N = 2500$ samples were generated each ($100 \times 3 \times 3 = 900$ recoveries).
-
-### Missing Data
-In the last part, I set intervals of the generated data to be missing.
-The model parameters were set to their default values.
-For $k \in \{2, 3, 4\}$ states and $m \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples were generated ($100 \times 3 \times 3 = 900$ recoveries).
-The length of the missing data interval $l \in [1,200]$ samples varied in equidistant steps between the data sets.
-
-## Data Analysis [^6]
-For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
+To angle, we added white noise from a von Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
+$\tau_{noise}$ and $\kappa_{noise}$ were increased simultaneously in equidistant steps in their intervals.
+
+## Data Analysis
+For each parameter, we calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
 Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\psi_{true} \ne 0$.
 This was not the case for some mean parameters of the von Mises distributions.
-In those cases, I used $\psi_{true} = 2\pi$ instead.
-I treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter.
-By taking the median, I reduced the influence of potential outliers in the estimation and using proportions enabled me to compare RMdSPD values across parameters and data sets.
-Additionally, I applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in part one.
+In those cases, we used $\psi_{true} = 2\pi$ instead.
+We treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter.
+By taking the median, we reduced the influence of potential outliers in the estimation and using proportions enabled us to compare RMdSPD values across parameters and data sets.
+Additionally, we applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in part one.
 Regression slopes closer to one indicated that the model better captured parameter change.
 Regression intercepts different from zero reflected a bias in parameter estimation.
-To assess state recovery, I computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set.
+To assess state recovery, we computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set.
 Cohen's kappa estimates the agreement between two classifiers accounting for the agreement due to chance.
 Higher kappa values were interpreted as better model accuracy.
-I adopted the ranges proposed by @Landis1977 to interpret kappa values.
+We adopted the ranges proposed by @Landis1977 to interpret kappa values.
 Models that could not be fitted were excluded from the recovery.
-
-[^6]: Note that this section deviates from my proposed analysis plan. However, since I preregistered the changed analysis plan, I treated the results of the simulation as confirmatory.
-
-(ref:sim-ranges) Approximate Ranges of Response Variables Used to Generate Parameter Values in the Simulation
-(ref:sim-ranges-note) Units are deg/s (velocity), deg/s (acceleration), and radians (angle).
-~ indicates that the distribution has a peak at this value.
-Velocity ranges are based on event velocities reported in @Larsson2013.
-Since I would also like to test the algorithm on extreme data distributions, I extended the ranges beyond those found in typical eye movement data.
-
-```{r sim-ranges}
-
-apa_table(data.frame("Event" = rep(c("Fixation", "Saccade", "PSO", "Smooth pursuit"), each = 3),
-                     "Resp variable" = rep(c("Velocity", "Acceleration", "Angle"), 4),
-                     "Range" = c("0-50", "0-50", "uniform", "50-1000", "50-500", "~0", "20-100", "10-90", paste("~", expression(pi), sep = ""), "20-100", "0-30", "~0")), 
-          caption = "(ref:sim-ranges)",
-          note = "(ref:sim-ranges-note)")
-
-```
 
@@
 
-\newpage
-
-# Results [^5]
-## Simulation Study
+## Results
 ### Parameter Variation
-#### Two States
 In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation of estimated parameters and accuracy of estimated state sequences.
@@
 With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely.
-The deviation seemed to increase slightly with parameter magnitude.
-Thus, parameter change was capture well and the estimation almost unbiased.
+Thus, parameter change was captured well and the estimation almost unbiased.
 Considering accuracy, Figure \@ref(fig:plot-acc-1-2) displays Cohen's kappa between true and estimated hidden state sequences.
 With two exceptions, kappa values were almost one, suggesting nearly perfect agreement.
-In sum, the HMM with two states recovered parameters and hidden states very well and outliers only occurred rarely.
-
-[^5]: The design and analysis plan of the simulation study have not been included in my proposal.
-However, since I preregistered both the design and analysis, I treated the results as confirmatory.
-In general, the evaluation of an algorithm can hardly be proposed in detail before the algorithm has been developed.
-Therefore, I chose to report a general results section.
-Nevertheless, I indicated whenever results were obtained through exploratory analyses that were not planned beforehand.
 
@@
 
-#### Three States
-For the simulation with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
-When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently below 0.5.
-Varying $\kappa$ in states two and three led to RMdSPDs below 0.5 in the respective states, which can be interpreted as moderate recovery.
+For the HMM with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
+When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently between 0.1 and 0.5.
+Varying $\kappa$ in states two and three led to RMdSPDs between 0.1 and 0.5 in the respective states, which we interpreted as moderate recovery.
 Otherwise, RMdSPDs were consistently lower than 0.1, indicating good recovery.
 Inspecting the regressions between manipulated true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one).
-Again, the deviation seemed to increase with true parameter magnitude (reverse for kappas).
 In contrast to the two-state HMM, larger deviations and more outliers were observed.
 Cohen's kappa values are presented in Figure \@ref(fig:plot-acc-1-3).
-For most estimated models, the kappas between true and estimated state sequences were above 0.95, meaning almost perfect agreement.
-However, for some models, I observed kappas clustered around zero or -0.33, which suggests that state labels were switched.
-To summarize, the three-state HMM had a slightly worse parameter recovery and more outliers than the two-state model but shows a similar good accuracy.
+For most estimated models, the kappa values between true and estimated state sequences were above 0.95, meaning almost perfect agreement.
+However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched.
 
@@
 
-#### Four States
 The RMdSPDs for the four-state HMM is shown in Figure \@ref(fig:plot-rmdspd-1-4).
-For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were below 0.5, suggesting moderate recovery.
-Estimated kappa parameters in state four were also often below 0.5 when parameters in states two, three, and four were varied.
+For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were between 0.1 and 0.5, suggesting moderate recovery.
+Estimated kappa parameters in state four were also often moderately recovered when parameters in states two, three, and four were varied.
 Otherwise, RMdSPDs were below 0.1, indicating good recovery.
@@
 However, there were larger deviations and more outliers than in the previous models, especially for states one and four.
-Accuracy measured by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
+Cohen's kappa ranged majorly between 0.6 and 0.9, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
 Here, some kappa values clustered around 0.25 and zero, which, again, can be interpreted as the result of label switching.
-Summarizing, the parameter recovery for the four-state HMM was slightly worse with even more outliers compared to the three-state model.
-Especially the recovery of transition parameters in states corresponding to fixations and smooth pursuits decreased.
-Overall, simulations in part one demonstrated that the HMM recovered parameters very well when parameters were manipulated.
+Overall, simulations in part one demonstrated that the HMM recovered parameters very well when they were manipulated.
 Deviations from true parameters were mostly small.
@@
 ### Sample Size and Noise Variation
-#### Two States
-In the second part, I varied the sample size of the HMM and added noise to the generated data.
+In the second part, we varied the sample size of the HMM and added noise to the generated data.
 For the two-state HMM, the RMdSPDs were above 0.5 for $\beta_{vel}$ and $\beta_{acc}$ in both states (see Figure \@ref(fig:plot-rmdspd-2-2)), suggesting bad recovery.
@@
 Accuracy measured by Cohen's kappa was almost perfect with kappa values very close to one (see Figure \@ref(fig:plot-acc-2), left plot).
-To conclude, adding noise only affected the recovery of scale parameters but not the accuracy of the two-state HMM.
-Increasing the sample size improved the recovery slightly.
 
@@
 
-#### Three States
-The RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)), indicating bad recovery.
+Regarding three states, the RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)), indicating bad recovery.
 Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5.
-For most parameters across all three states, higher sample sizes had lower RMdSPDs.
-The accuracy of the estimated models was almost perfect with most kappa values above 0.95 (see Figure \@ref(fig:plot-acc-2), middle plot).
+For most parameters across all three states, models with higher sample sizes had lower RMdSPDs.
+The state recovery of the estimated models was almost perfect with most kappa values above 0.95 (see Figure \@ref(fig:plot-acc-2), middle plot).
 Several outliers clustered around kappas of zero and -0.33, signaling label switching.
-For the three-state HMM, adding noise led to a slightly worse recovery and accuracy overall.
-Scale parameters of gamma distributions in all three states were only badly recovered.
 
@@
 
-#### Four States
 RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
 For states one and four, values for most parameters (including all transition probabilities) were above 0.5, suggesting bad recovery.
-Similarly, RMdSPDs for $\beta_{vel}$ and $\beta_{acc}$ in states two and three were above 0.5.
+Similarly, $\beta_{vel}$ and $\beta_{acc}$ in states two and three showed bad recovery.
 For states two and three, higher samples sizes showed slightly lower RMdSPDs.
@@
 Multiple kappa values clustered around 0.25 or zero, which can be explained by label switching.
-In summary, adding noise to data generated by the four-state HMM heavily decreased the recovery for most parameters in states corresponding to fixations and smooth pursuits.
-For saccade and PSO states, only scale parameters of gamma distributions were badly recovered, but increasing the sample size slightly improved the recovery.
 In general, the HMM recovered parameters well despite noise being added to the data.
@@
 Increasing the sample size in the HMM slightly improved the recovery of most parameters.
-The accuracy of the model was slightly lowered when more states were included, but it was neither affected by the noise variability $\tau_{noise}$ nor the sample size.
+The state recovery of the model was slightly lowered when more states were included, but it was neither affected by the noise variability $\tau_{noise}$ nor the sample size.
 
@@
 
-### Variation of Starting Values
-The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM.
-For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure \@ref(fig:plot-rmdspd-3-2)), suggesting good recovery.
-Cohen's kappa values were slightly lower than 1, indicating almost perfect accuracy (see Figure \@ref(fig:plot-acc-3)).
-For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5 (see Figure \@ref(fig:plot-rmdspd-3-3)), which means they were moderately well recovered.
-As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33 (see Figure \@ref(fig:plot-acc-3)).
-Regarding the four state-HMM, RMdSPDs of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1 (see Figure \@ref(fig:plot-rmdspd-3-4)).
-The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero (see Figure \@ref(fig:plot-acc-3)).
-Taken together, increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor the accuracy.
-
-(ref:plot-rmdspd-3-2) RMdSPD between true and estimated parameters of the two-state HMM in part three of the simulation.
-Colours indicate the variation in starting values used to estimate parameters.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-3-2, fig.cap="(ref:plot-rmdspd-3-2)"}
-
-print(plots.rmsd.234[[2]][[1]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-acc-3) Boxplots displaying Cohen's kappa depending on the number of states in the HMM in part three.
-Colours indicate the variation in starting values used to estimate parameters.
-Solid vertical lines symbolize medians and hinges the first and third quartile.
-Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
-Crosses represent outliers.
-
-```{r plot-acc-3, fig.cap="(ref:plot-acc-3)"}
-
-print(plots.acc.234[[2]])
-
-```
-
-(ref:plot-rmdspd-3-3) RMdSPD between true and estimated parameters of the three-state HMM in part three of the simulation.
-Colours indicate the variation in starting values used to estimate parameters.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-3-3, fig.cap="(ref:plot-rmdspd-3-3)"}
-
-print(plots.rmsd.234[[2]][[2]] + theme(axis.text = element_text(size = 6)))
-
-```
-
-(ref:plot-rmdspd-3-4) RMdSPD between true and estimated parameters of the four-state HMM in part three of the simulation.
-Colours indicate the variation in starting values used to estimate parameters.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-3-4, fig.cap="(ref:plot-rmdspd-3-4)"}
-
-print(plots.rmsd.234[[2]][[3]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-In this part, I explored post-hoc whether clusters of low Cohen's kappa values were due to label switching.
-According to @Visser2019, label switching occurs when exchanging the order of the states leads to equally likely models.
-Thus, the model is accurate but has oppositely estimated states compared to the true states.
-In this case, an HMM with three states that is perfectly accurate but has two labels switched would have a Cohen's kappa of 0.
-To test if label switching occurred, I manually switched one or two labels post-hoc (see Figure \@ref(fig:plot-acc-3-expl)).
-It can be seen that this approach resolved low accuracy clusters for three and four states.
-
-(ref:plot-acc-3-expl) Boxplots displaying Cohen's kappa depending on the number of states in the HMM in part three after state labels were switched post-hoc (exploratory analysis).
-Colours indicate the variation in starting values used to estimate parameters.
-Solid lines symbolize medians and hinges the first and third quartile.
-Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
-Crosses represent outliers.
-
-```{r plot-acc-3-expl, fig.cap="(ref:plot-acc-3-expl)"}
-
-print(plots.acc.expl[[1]])
-
-```
-
-### Missing Data
-In the last part, data intervals of varying length were set to be missing.
-Regarding parameter recovery, RMdSPDs were almost exactly mirroring those in the previous part for two, three, and four states (see Figures \@ref(fig:plot-rmdspd-4-2), \@ref(fig:plot-rmdspd-4-3), and \@ref(fig:plot-rmdspd-4-4), respectively).
-Looking at accuracies, Figure \@ref(fig:plot-acc-4) illustrates an interaction between the length of and amount of missing data intervals.
-Cohen's kappa values linearly decreased with the length of missing data intervals and the decrease became linearly steeper when more intervals were included.
-For two and three state models, kappas started at almost one (almost perfect) and decreased to 0.6 (substantial accuracy) for five missing intervals.
-For the four-state model, kappas started around 0.85 (almost perfect) and decreased to 0.5 (substantial accuracy) for five missing intervals.
-As in previous parts, several kappa values clustered around zero and -0.33 for the three-state model and around 0.6, 0.25, and zero for the four-state model.
-In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model from a nearly perfect to a moderate extent.
-With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences.
-The most critical decrease in recovery occurred when noise was added to the data generated by models including four states.
-In the noise condition, scale parameters of gamma distributions were often badly recovered.
-Higher sample sizes slightly improved parameter recovery, but neither variation in starting values nor missing data affected it.
-In contrast, the accuracy of the model was linearly decreasing with more data missing but not influenced by manipulating parameters, noise, or starting value variation.
-Adding more states to the HMM generally decreased the parameter recovery and the accuracy.
-
-(ref:plot-rmdspd-4-2) RMdSPD between true and estimated parameters of the two-state HMM in part four of the simulation.
-Colours indicate the number of missing data intervals.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-4-2, fig.cap="(ref:plot-rmdspd-4-2)"}
-
-print(plots.rmsd.234[[3]][[1]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-rmdspd-4-3) RMdSPD between true and estimated parameters of the three-state HMM in part four of the simulation.
-Colours indicate the number of missing data intervals.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-4-3, fig.cap="(ref:plot-rmdspd-4-3)"}
-
-print(plots.rmsd.234[[3]][[2]] + theme(axis.text = element_text(size = 6)))
-
-```
-
-(ref:plot-rmdspd-4-4) RMdSPD between true and estimated parameters of the four-state HMM in part four of the simulation.
-Colours indicate the number of missing data intervals.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-4-4, fig.cap="(ref:plot-rmdspd-4-4)"}
-
-print(plots.rmsd.234[[3]][[3]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-acc-4) Cohen's kappa depending on the length of missing data intervals. Colours indicate the number of missing data intervals. Top facet labels indicate the number of states in the HMM.
-
-```{r plot-acc-4, fig.cap="(ref:plot-acc-4)"}
-
-print(plots.acc.int.24[[2]])
-
-```
-
-## Validation
-### Starting Values and Model Fitting
-To test my hypotheses, I applied gazeHMM on two benchmark data sets.
-Note that, for the @Ehinger2019 data set, I only fitted models to the Eyelink data to reduce the complexity of the analysis.
-For both data sets, I used the same set of starting values (initial state model: $\rho=1/k$; transition model: $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$; response model: see Table \@ref(tab:tab-starting-values)) to estimate the parameters of the HMM.
+# Validation Study
+To validate gazeHMM, we applied the algortihm on two benchmark data sets.
+As starting values, we used $\rho=1/k$ for the initial state model as well as $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$ for the transition model.
+The values for the response model are displayed in Table \@ref(tab:tab-starting-values).
 In contrast to the simulation study, generating random starting values often led to bad model fits and label switching between states.
 To improve the fitting of the gamma distributions, velocity and acceleration signals were scaled down by factor 100 [^7] (so were the starting values for their gamma distributions).
-The algorithm was applied separately for every participant and every condition, task, or block.
-For the data set by @Andersson2017, the algorithm was successfully fitted for every participant and condition.
-Examples of event classification and parameter estimates from gazeHMM are presented in Appendix A.
 
@@
 ```
+
+## Data Sets
+
+We chose two data sets for validation: One was published in a study by @Andersson2017 and has been widely used for validation purposes [e.g., @Pekkanen2017].
+It contains eye-tracking data from three conditions: A static condition, where subjects had to look freely at images, and two dynamic conditions, where they had to follow a constantly moving dot or objects in a video.
+The data was sampled with 500 Hz and two human coders (MN and RA) labeled them as belonging to six different eye movement events: Fixation, saccade, PSO, smooth pursuit, blink, or other.
+@Andersson2017 used the data to compare 10 different classification algorithms.
+We used their results to compare these 10 algorithms and the two human coders with gazeHMM.
+The second data set was published in Ehinger2019 and has to our knowledge not yet been used for validation.
+Here, we only use tasks four and five out of 10 tasks, because these are qualitatively different to the first data set.
+In task four, subjects were instructed to fixate a central target for 20 s.
+Task 5 was set up similarly, but subjects had to blink when they heared one out of seven beeps (with a beep duration of 100 ms and 1.5 s intervals in between).
+Eye movements were recorded with 250-500 Hz.
+For simplicity, we only used data obtained by the EyeLink eye-tracker.
 
@@
 
-# Discussion
-In this report, I presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events.
-The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von Mises and a uniform distribution.
-An HMM classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
-A simulation study revealed that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions.
-Adding smooth pursuits to the model and noise to the generated data were the most critical factors for decreasing recovery.
-In contrast, higher sample sizes yielded a better parameter recovery.
-The variation in starting values did not affect the recovery but the amount of missing data decreased the recovery of hidden states.
-I applied gazeHMM with different numbers of states to benchmark data and compared the model fit, showing that a five-state HMM had consistently most likely generated the data.
-Thus, the model comparison provided evidence against my first hypothesis that the number of events in gazeHMM matches theoretical expectations.
-When comparing gazeHMM with four events against other algorithms and human coders, it did not show the best performance for the duration and number of classified events nor the sample-to-sample agreement across all conditions.
-However, it had a lower overall disagreement for dynamic data but higher disagreement for static data.
-I argued that four states might not be appropriate for static data and additionally applied gazeHMM with three states to them.
-That led to substantially better classifications for fixations, but slightly worse performance for saccades and PSOs.
-With three states, gazeHMM had a lower overall disagreement than all other algorithms.
-I treat this pattern as moderate evidence for my second hypothesis that gazeHMM improves the classification of eye movements.
-Taken together, I only found some evidence for one part of my research question, that gazeHMM improved classifying gaze data but not describing them.
-Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle.
-The overlap can cause the HMM to confuse parameters and hidden states.
-The decrease in parameter recovery (especially for scale parameters) due to noise shows that the overlap is enhanced by more dispersion in the data.
-The scale parameters might be particularly vulnerable to extreme data points.
-The decrease in state recovery for long and many intervals of missing data can be explained by the tendency of HMMs to stay in the same state when no information is available from the data.
-However, the decrease was rather small, leading to the conclusion that the HMM can deal well with missing data.
-Despite these drawbacks, the recovery of the generative model in gazeHMM seems very promising.
-The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).
-The model comparison on the benchmark data suggested that gazeHMM is not the best generative model for eye movement data.
-Even using a model selection criterion that identified correct models successfully in previous studies [@Pohle2017], did not yield the expected pattern.
-There are several explanations for this result:
-*There were more eye movement events present in the data than I expected*.
-Eye movement events can be divided into subevents.
-For example, fixations consist of drift and tremor movements [@Duchowski2017] and PSOs encompass dynamic, static, and glissadic over- and undershoots [@Larsson2013].
-A study on a recently developed HMM algorithm supports this explanation: @Houpt2018 applied the unsupervised BP-AR-HMM algorithm to the @Andersson2017 data set and classified more distinct states than the human coders.
-Some of the states classified by BP-AR-HMM matched the same event coded by humans.
-Since the subevents are usually not interesting for users of classification algorithms, the ability of HMMs to classify might limit their ability to describe eye movements here.
-*Model selection criteria are generally not appropriate for comparing HMMs with different numbers of states*.
-This argument has been discussed in the field of ecology [see @Li2017], where studies found that selection criteria preferred models with more states than expected [similar to the result of this study; e.g., @Langrock2015].
-@Li2017 explain this bias with the simplicity of the submodels in HMMs: Initial state, transition, and response models for each state are usually relatively simple.
-When they do not describe the processes in the respective states accurately, the selection criteria compensate for that by preferring a model with more states.
-Thus, there are not more latent states present in the data, but the submodels of the HMM are misspecified or too simple.
-Correcting for model misspecifications, led to better model recovery in studies on animal movements [@Langrock2015; @Li2017].
-However, @Pohle2017 showed in simulations that the ICL identified the correct model despite several misspecifications.
-It has to be noted that the study by @Pohle2017 only used data generating models with two states.
-*The submodels of gazeHMM were misspecified*.
-@Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models.
-Both situations could apply to gazeHMM and eye movement data.
-Outliers occur frequently in eye-tracking data due to measurement error.
-Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task the data is obtained from [@Langrock2015].
-Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf.
-@Li2017].
-This aspect could indeed oversimplify the complex nature of eye movement data.
-Comparing gazeHMM to other algorithms on benchmark data showed that gazeHMM moderately improved the event classification in agreement with human coders.
-However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yielded different results.
-The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events (incl.
-blinks).
-Nevertheless, Cohen's kappa values of 0.72 (saccades - video - four states) or 0.68 (fixations - image - three states) indicate substantial agreement to human coders, especially in light of the maximum reference of ~0.8 from the simulation study.
-At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
-Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and to find errors in their design.
-
-## Advantages of gazeHMM
-In view of the four proposed goals that gazeHMM should fulfill, I can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they only have minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
-Their default values should be appropriate for most applications.
-A major advantage of gazeHMM is that it does not require human labeled data as input.
-Instead, it estimates all parameters and hidden states from the data.
-Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018a], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
-This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data.
-Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit.
-Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2018], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
-However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events.
-Therefore, I recommend using gazeHMM with four events only for exploratory purposes.
-Without smooth pursuits, I consider gazeHMM's classification as appropriate for application.
-Lastly, its implementation in R using depmixS4 [@Visser2010] should make gazeHMM a tool that is easy to use and customize for individual needs.
-
-## Future Directions
-Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for examples, see @Balakrishnan2009].
-Potential problems of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters).
-Another option to model the correlation could be to include one of the response variables as a covariate of the other.
-Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred.
-This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.
-Third, one solution to diverging results when comparing gazeHMM with different events could be model averaging: Instead of using the maximum posterior state probability of each sample from the preferred model, the probabilities could be weighted according to a model selection criterion (e.g., Schwarz weight) and averaged.
-Then, the maximum averaged probability could be used to classify the samples into events.
-This approach could lead to a more robust classification because it reduces the overconfidence of each competing model and easily adapts to new data [analogous to Bayesian model averaging; @Hinne2020].
-However, the model comparison for gazeHMM often showed extreme weights for a five-state model, which would lead to a very limited influence of the other models in the averaged probabilities.
-Fourth, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data.
-As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements.
-As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
-Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and auto-correlation functions [see @Zucchini2016].
-Fifth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonably short saccades, gazeHMM could explicitly model the duration of events.
-This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2006].
-Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs.
-This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades.
-Drawback of this extension are higher computational costs and difficulties with including covariates [@Zucchini2016].
-Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM.
-This could potentially be achieved by using different response distributions or parameter optimization methods.
-Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.
-
-## Conclusion
-Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events.
-For smooth pursuits, the classification is not optimal and thus not recommended, yet.
-On one hand, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one.
-On the other hand, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events.
-These difficulties highlight challenges for HMMs in general but also specifically for eye movements.
-Revisiting my research question, I conclude that gazeHMM is suitable for classifying eye movement events but still lacks the ability to accurately describe them.
-
-\newpage
-
 # References

### 2020-07-14 manuscript/paper_BRM.Rmd
shorten validation section

@@
 [^7]: Scaling down by factor 100 differs from the simulation study (scaling down by 10).
-The algorithm allows the user to manually specify this factor and in this case, factor 100 led to better model fits than factor 10.
-Thus, I chose it exploratorily.
+The algorithm allows the user to manually specify this factor, and in this case, factor 100 led to better model fits than factor 10.
 
@@
 ## Data Sets
-
 We chose two data sets for validation: One was published in a study by @Andersson2017 and has been widely used for validation purposes [e.g., @Pekkanen2017].
@@
 We used their results to compare these 10 algorithms and the two human coders with gazeHMM.
-The second data set was published in Ehinger2019 and has to our knowledge not yet been used for validation.
-Here, we only use tasks four and five out of 10 tasks, because these are qualitatively different to the first data set.
+The second data set was published in @Ehinger2019 and has to our knowledge not yet been used for validation.
+Here, we only use tasks four and five out of 10 tasks because these are qualitatively different to the first data set.
 In task four, subjects were instructed to fixate a central target for 20 s.
@@
 Eye movements were recorded with 250-500 Hz.
-For simplicity, we only used data obtained by the EyeLink eye-tracker.
-
+For simplicity, we only used data obtained by the EyeLink (SR Research Ltd., Ontario, Canada) eye-tracker.
+
+## Data Analysis
+A successful validation of gazeHMM was determined by using two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets.
+The fits of the generative models were compared using Schwarz weights [@Wagenmakers2004], a weighted version of the BIC.
+They can be interpreted as the probability of a model having generated the data compared to the competing models.
+For the static condition in the @Andersson2017 data set, we expected the generative model with three states (fixation, saccade, and PSO), and for the dynamic conditions the model with four states (incl.
+smooth pursuit) to display the highest Schwarz weight.
+Regarding the @Ehinger2019 data set, we assumed that the one-state model (only fixation) would show the highest weights for both tasks.
+The algorithm was applied separately to every subject, condition, or task.
+For the @Andersson2017 data set, all generative models were successfully fitted, whereas for the @Ehinger2019 data set, it was only 780 out of 900 models (87%, 60 models per task).
+Second, we compared gazeHMM to other algorithms and human coders.
+We applied our algorithm with a three-state generative model to the static condition in the @Andersson2017 data set, and with a four-state model to the dynamic conditions.
+For comparison criteria, we followed @Andersson2017: We calculated the RMSD of event durations and counts between all algorithms and the two human coders.
+Cohen's kappa was calculated for each event as the binary agreement between gazeHMM and the human coders.
+Lastly, the overall disagreement indicated which samples were classified differently by gazeHMM across all events.
+
+## Results
 ### Model Comparison
@@
 
+  df$subject <- 1:nrow(df)
+
   return(df)
-})
+}) %>% reduce(rbind) %>% mutate(condition = rep(c("Moving dots", "Image", "Video"), c(11, 14, 9)))
 
@@
 
-bic.plot <- lapply(A2017.bic, function(x) {
-
-  data.long <- x %>% 
-    mutate(subject = 1:nrow(x)) %>%
-    pivot_longer(names(x), names_to = "model", values_to = "weight") %>%
-    mutate_at(c("subject", "model"), as.factor)
-
-  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
+bic.plot <- A2017.bic %>%
+  pivot_longer(names(A2017.bic)[1:5], names_to = "model", values_to = "weight") %>%
+  mutate_at(c("subject", "model", "condition"), as.factor) %>%
+  ggplot(aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
+  facet_grid(cols = vars(condition)) + 
   scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
@@
 
-  return(p)
-})
-
-```
-
-To examine whether gazeHMM accurately describes eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
-The starting values for the fifth state were chosen to represent distributions for microsaccade samples (velocity and sample-to-sample angle as for smooth pursuits, acceleration as for saccades and PSOs).
-Schwarz weights are the result of transforming a set of BIC values [@Wagenmakers2004].
-They can be interpreted as the probability for a model to have generated the data it was fitted to.
-For the @Andersson2017 data set, I expected the highest weights for the three-state models in the image condition.
-For the moving dots and video conditions, I predicted the highest weights for the four-state model.
-Figure \@ref(fig:plot-schwarz-weights-img) shows the Schwarz weights for different subjects and models in the image condition.
-For all subjects, the five-state model displayed the highest weight, suggesting that it most likely generated the data.
-In the moving dots condition, the five-state model had the highest weights for most subjects, but the one-, three, and four-state models exhibited the highest weights for two subjects each (see Figure \@ref(fig:plot-schwarz-weights-dots)).
-The video condition had the same pattern as the image condition, as the five-state model consistently had the highest weights (see Figure \@ref(fig:plot-schwarz-weights-vid)).
-In all three conditions, response parameters of the fifth state were mostly similar to those of the smooth pursuit state.
-In contrast to my hypothesis, applying different HMMs to the @Andersson2017 data suggests that the five-state model has most likely generated the data.
-Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
-For the @Ehinger2019 data, I predicted a one-state model to be preferred.
-Here, the model comparison for the yielded similar results (see Appendix A), contradicting my hypothesis.
+```
+
+Examining the Schwarz weights displayed in \@ref(fig:plot-schwarz-weights), we observed that the five-state generative model showed the highest weights in all three conditions.
+Only in the moving dots condition, two subjects each displayed the highest weights for one- and three-state models.
+In sum, we concluded that the five-state generative model has most likely generated the @Andersson2017 data, opposing our expectations.
+Because the @Ehinger2019 data set showed a similar pattern, we included the results for this data in the supplementary material.
 A recent model recovery study showed that the BIC tended to prefer overly complex HMMs when they were misspecified [e.g., the conditional independence assumption was violated; @Pohle2017].
-Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2000] performed better in choosing the correct data generating model.
-Therefore, I computed the weighted ICL criterion for the HMMs fitted to the @Andersson2017 data set as an exploratory analysis.
-Using the ICL as the model selection criterion yielded very similar results to the BIC (see Appendix A).
-The preference for the five-state model was even more consistent across conditions and subjects.
-
-(ref:plot-schwarz-weights-img) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the image condition of the @Andersson2017 data set.
-Higher weights indicate better model fit.
-
-```{r plot-schwarz-weights-img, fig.cap="(ref:plot-schwarz-weights-img)"}
-
-print(bic.plot[[2]])
-
-```
-
-(ref:plot-schwarz-weights-dots) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the moving dots condition of the @Andersson2017 data set.
-Higher weights indicate better model fit.
-
-```{r plot-schwarz-weights-dots, fig.cap="(ref:plot-schwarz-weights-dots)"}
-
-print(bic.plot[[1]])
-
-```
-
-(ref:plot-schwarz-weights-vid) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the video condition of the @Andersson2017 data set.
-Higher weights indicate better model fit.
-
-```{r plot-schwarz-weights-vid, fig.cap="(ref:plot-schwarz-weights-vid)"}
-
-print(bic.plot[[3]])
+Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2000] performed better at choosing the correct data generating model.
+Therefore, we post hoc computed the weighted ICL criterion (analogous to Schwarz weights) for the models fitted to the @Andersson2017 data set.
+Using the ICL as the model selection criterion yielded very similar results to the BIC (see supplementary material).
+The preference for the five-state generative model was even more consistent across conditions and subjects.
+
+(ref:plot-schwarz-weights) Schwarz weights displayed for each subject and generative models with different numbers of states.
+Top facet labels indicate the condition in the @Andersson2017 data set.
+Higher weights indicate a better model fit.
+
+```{r plot-schwarz-weights, fig.cap="(ref:plot-schwarz-weights)"}
+
+print(bic.plot)
 
@@
       colnames(tabmat) <- c("Mean", "SD", "Events", "RMSD")
-      rownames(tabmat)[1:3] <- c("coderMN", "coderRA", "gazeHMM")
+      rownames(tabmat)[1:3] <- c("coderMN", "coderRA", paste0("gazeHMM-", x+1))
 
@@
 })
+
+rmsd.table[[3]][1:3] <- lapply(1:3, function(x) {
+
+  df <- rbind(rmsd.table[[3]][[x]][1:2,], rmsd.table[[2]][[x]][3,], rmsd.table[[3]][[x]][3:nrow(rmsd.table[[3]][[x]]),])
+
+  df[c(1:3, 5:nrow(df)),2:5] <- rmsd.table[[2]][[x]][,2:5]
+
+  df[3,6:13] <- NA
+  df[4,2:5] <- NA
+
+  return(df)
+})
+
+rmsd.table[[3]][[4]][3,2:5] <- NA
 
@@
 
+# Compute Cohen's kappa between human coders
+
+A2017.cod.acc <- lapply(A2017, function(x) {
+  df <- lapply(x, function(y) {
+
+    return(y[,6:7])
+  }) %>% reduce(rbind)
+
+  events <- lapply(1:4, function(e) {
+
+    MN <- df[,1] == e
+    RA <- df[,2] == e
+
+    mat <- matrix(c(sum(MN & RA, na.rm = T), sum(MN & !RA, na.rm = T), 
+                    sum(!MN & RA, na.rm = T), sum(!MN & !RA, na.rm = T)), 2, 2)
+
+    return(cohen.kappa(mat)$kappa)
+  })
+})
+
 # Create tables for Cohen's kappa and confusion matrices
@@
 
-#### Event Durations
-The hypothesis that gazeHMM improves the classification performance of eye movement events was investigated by comparing it against other algorithms.
-Therefore, I applied gazeHMM with four events to the @Andersson2017 data set.
-I calculated the RMSD [^3] of event durations as described in the original article by Andersson et al.
-for all algorithms included in the study plus gazeHMM: First, the mean and standard deviation of event durations as well as the number of events were normalized to [0,1] for each coder and algorithm.
-Second, the sum of the root square deviation between each algorithm and the average of the two human coders is computed.
-This metric can only be relatively interpreted [@Andersson2017].
-Table \@ref(tab:tab-rmsd-4-fix) shows that, in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations.
-The high RMSD values were mostly due to a large amount of very short fixations.
-For saccades, gazeHMM has a relatively low RMSD in the image and video condition and a relatively high RMSD in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)).
-Here, gazeHMM overestimated the duration of saccades, as indicated by high mean and SD durations.
-For PSOs, gazeHMM had the highest RMSD in the image and video condition (see Table \@ref(tab:tab-rmsd-4-pso)).
-In the moving dots condition, its RMSD was lower than for NH [@Nystrom2010] but higher than for LNS [@Larsson2013].
-Here, gazeHMM underestimated the number of PSOs and their duration compared to the human coders.
-No other algorithm in the study was designed to parse smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)).
-In sum, gazeHMM did not outperform all other algorithms regarding the RMSD of event durations compared to human coders.
-The duration of fixations and smooth pursuits was largely underestimated and the number of these two events largely overestimated.
-For saccades and PSOs, gazeHMM estimated relatively accurate durations and event numbers, but did not reach the best algorithms.
+As displayed in \@ref(tab:tab-rmsd-4-fix), gazeHMM showed a relatively low RMSD for fixations in the static condition compared to the other algorithms that were applied to the @Andersson2017 data set.
+The lower RMSD for fixations indicated more similar classification to the human coders in terms of their mean and SD duration as well as the number of classified fixations.
+Oppositely, for fixations in the dynamic conditions, the RMSD of gazeHMM was the highest among the compared algorithms, suggesting substantial differences to the human coders.
+It is likely that these occured because gazeHMM classified a much larger number of fixations with very short durations.
+For saccades, gazeHMM had high RMSDs for the image and moving dots conditions, but for the video condition the RMSD was comparably low (see Table \@ref(tab:tab-rmsd-4-sac)).
+Only two other algorithms classified PSOs [NH and LNS; @Nystrom2010; @Larsson2013], and gazeHMM showed consistently higher RMSDs than LNS (see Table \@ref(tab:tab-rmsd-4-pso)).
+Compared to NH, the RMSD was higher in the image and video conditions but lower in the moving dots condition.
+No other algorithm parsed smooth pursuits, but the RMSD for gazeHMM was higher than among the human coders (see Table \@ref(tab:tab-rmsd-4-sp)).
+Again, it classified a much larger number of smooth pursuits with short durations.
 
 (ref:tab-rmsd-4-fix) Fixation Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-fix-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-fix-note) Durations are displayed in seconds. gazeHMM-3 classified three and gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
 
@@
 (ref:tab-rmsd-4-sac) Saccade Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-sac-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-sac-note) Durations are displayed in seconds. gazeHMM-3 classified three and gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
 
@@
 (ref:tab-rmsd-4-pso) PSO Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-pso-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-pso-note) Durations are displayed in seconds. gazeHMM-3 classified three and gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
 
@@
 
-(ref:tab-rmsd-4-sp) Smooth Pursuit Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-sp-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-sp) Smooth Pursuit Duration Descriptives and RMSD Between gazeHMM and Human Coders
+(ref:tab-rmsd-4-sp-note) Durations are displayed in seconds. gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
 
@@
 
-[^3]: In my proposal, I declared to compute the RMSE instead of RMSD.
-However, since I conducted an in-sample comparison and it was used by @Andersson2017, the term *deviation* is more appropriate (i.e., RMSD).
-
-#### Sample-to-sample Agreement
+Table \@ref(tab:tab-kappa-4) contains the sample-to-sample agreement between the algorithms and human coders measured by Cohen's kappa.
+For fixations, gazeHMM shows one of the highest agreements for static, and *the* highest agreements for dynamic data.
+The absolute agreement is substantial for static and slight to fair [@Landis1977].
+For saccades, the relative agreement for gazeHMM is low for the image, moderate for the moving dots, and high for the video condition.
+In absolute terms, the agreement is moderate to substantial.
+Concerning PSOs, gazeHMM shows higher agreement than NH in the image and video conditions but consistently lower agreement compared to LNS.
+Absolutely, the agreement is slight to fair.
+Lastly, the
+
 Additionally, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of sample-to-sample agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
@@
 (ref:tab-kappa-4) Cohen's Kappa Between Human Coders and Algorithms for Different Conditions and Events
-(ref:tab-kappa-4-note) Negative values were set to zero. gazeHMM classified four events. Table adapted from @Andersson2017.
+(ref:tab-kappa-4-note) Negative values were set to zero. gazeHMM-3 classified three and gazeHMM-4 classified four events. Table adapted from @Andersson2017.
 
@@
 
-tab.kappa.4 <- cbind(c("coderMN", "coderRA", "gazeHMM", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
-                     as.data.frame(rbind(ref.acc[1:2,], kappa.table[[3]], ref.acc[3:nrow(ref.acc),])))
+tab.kappa.4 <- cbind(c("coderMN", "coderRA", "gazeHMM-3", "gazeHMM-4", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
+                     as.data.frame(rbind(ref.acc[1:2,], c(kappa.table[[2]], NA, NA, NA), kappa.table[[3]], ref.acc[3:nrow(ref.acc),])))
+
+tab.kappa.4[4, c(2, 5, 8, 11)] <- NA
+tab.kappa.4[3, -c(1, 2, 5, 8, 11)] <- NA
 
@@
 
-(ref:tab-rmsd-3) Event Duration Descriptives and RMSD Between Algorithms and Human Coders for the Image Condition
-(ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
-
-```{r tab-rmsd-3}
-
-tab.rmsd.3 <- lapply(A2017.rmsd[[2]][[2]], function(x) {
-
-  df <- as.data.frame(t(as.matrix(x)))
-
-  rownames(df)[3] <- "gazeHMM"
-
-  out <- cbind(algorithm = rownames(df), df)
-
-  return(out)
-}) %>% reduce(left_join, by = "algorithm")
-
-apa_table(tab.rmsd.3,
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Fixations" = c(2, 5), "Saccades" = c(6, 9), "PSOs" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-rmsd-3)",
-          note = "(ref:tab-rmsd-3-note)")
-
-```
-
 (ref:tab-kappa-3) Cohen's Kappa Between Human Coders and Algorithms for Different Events in the Image Condition

### 2020-07-20 manuscript/paper_BRM.Rmd
add introduction and algorithm sections

@@
 ---
-
-# Introduction
 
@@
 ```
+
+# Introduction
+Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017].
+Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015].
+Commonly distinguished events are fixations, saccades, smooth pursuits, and post-saccadic oscillations (PSOs).
+Classifying eye-tracking data reduces their complexity and is the first step towards cognitive interpretation [@Salvucci2000].
+The classification is normally done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017].
+@Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
+Probabilistic models are frequently implemented to classify events.
+Instead of discrete classification, these models assign probabilities for belonging to an event to each sample.
+Probabilistic models have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
+324].
+One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs).
+They estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal.
+Each gaze sample depends on its corresponding state.
+Each state depends on the previous but not on earlier states of the sequence [@Zucchini2016].
+In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous).
+On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]).
+Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM.
+@Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR).
+The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and relative angle (named NSLR-HMM).
+Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]).
+Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).
+Several studies have tested the performance of HMM algorithms against other classification methods: I-HMM has been deemed as robust against noise, behaviorally accurate, and showing a high sample-to-sample agreement to human coders [@Andersson2017; @Komogortsev2010; @Salvucci2000].
+However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012].
+NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].
+In sum, HMMs seem to be a promising method for classifying eye movements.
+Still, the existing HMM algorithms each have their weaknesses that could be improved upon (and that could explain their inferior performance in comparison to other methods).
+First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
+Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability.
+It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals (contrary to the authors, we argue that this is a disadvantage).
+Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017].
+The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events.
+Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].
+HMMs have another property that has been rarely used in the context of eye movement classification: They can serve as a generative model for gaze data.
+Only @Mihali2017 used this property by building an explicit generative model for drift and microsaccades.
+For this article, we developed a novel algorithm for classifying gaze data named gazeHMM that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms.
+The development was guided by four goals: (a) it should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it should confirm rather than explore the presence of events in the data.
+
+# The Algorithm
+## Preprocessing
+Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events.
+Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018 and @Andersson2017].
+However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007].
+This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics [@Andersson2017].
+In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.
+Theoretically, these three metrics should separate eye movement events clearly.
+Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
+Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017].
+In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction.
+PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013].
+They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
+Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes.
+Those samples in between typically follow the same direction.
+Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees.
+Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
+Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, they were not in the scope of gazeHMM.
+The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010].
+In contrast, the absolute angle between two samples is calculated as the finite backward difference and the sample-to-sample angle as the finite forward difference.
+
+## The Generative Model
+We denote the three eye movement metrics by *X*, *Y*, and *Z*.
+Each sample was generated by a hidden state variable *S*.
+Given *S*, the HMM treats *X*, *Y*, and *Z* as conditionally independent.
+Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
+This step was merely taken to keep the HMM simple and identifiable.
+In gazeHMM, *S* can take one of two, three, or four hidden states.
+The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
+Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.
+HMMs can be described by three submodels: An initial state model, a transition model, and a response model.
+The initial state model contains probabilites for the first state of the hidden sequence $\rho_i = P(S_1=i)$, with *i* denoting the hidden states.
+In gazeHMM, the initial states are modeled by a multinomial distribution.
+The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM.
+Typically, probabilities to transition from state *i* to *j* $a_{ij} = P(S_{t+1}=j|S_{t}=i)$ are expressed in matrix form:
+$$\mathbf{A} = \left[\begin{array}
+{rrr}
+a_{11} & ... & a_{1j} \\
+\vdots & \ddots & \vdots \\
+a_{i1} & ... & a_{ij}
+\end{array}\right].
+$$
+Again, the transition probabilities for each state are modeled by multinomial distributions.  
+The response model encompasses distributions describing the response variables for every state in the model.
+Previous algorithms have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them).
+However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation).
+Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean.
+Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties.
+In gazeHMM, we use gamma distributions with a shape and scale parametrization for this purpose:
+$$X \sim Gamma(\alpha,\beta),$$
+$$Y \sim Gamma(\alpha,\beta).$$
+It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.   
+To model the sample-to-sample angle, we pursued a novel approach in gazeHMM: A mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution: 
+$$Z \sim vonMises(\mu, \kappa)+U(0, 2\pi).$$
+Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions.
+Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution.
+Thus, the uniform distribution should distinguish fixations from the other events.
 

### 2020-07-21 manuscript/paper_BRM.Rmd
complete algorithm section

@@
 In contrast, the absolute angle between two samples is calculated as the finite backward difference and the sample-to-sample angle as the finite forward difference.
+Most of the missing data in eye movement classification are due to blinks.
+Therefore, the user can indicate in gazeHMM which samples should be labeled as blinks (other missing samples are treated as noise).
+Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs.
+Since these samples could distort the classification of saccades in the HMM, gazeHMM removes them heuristically.
+Before classifying the samples, it sets all samples within 50 ms before and after blink samples as missing.
+The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.
 
@@
 We denote the three eye movement metrics by *X*, *Y*, and *Z*.
-Each sample was generated by a hidden state variable *S*.
+Each metric was generated by a hidden state variable *S*.
 Given *S*, the HMM treats *X*, *Y*, and *Z* as conditionally independent.
@@
 The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM.
-Typically, probabilities to transition from state *i* to *j* $a_{ij} = P(S_{t+1}=j|S_{t}=i)$ are expressed in matrix form:
+Typically, probabilities to transition from state *i* to *j* $a_{ij} = P(S_{t+1}=j|S_{t}=i)$ are expressed in matrix form [@Visser2011]:
 $$\mathbf{A} = \left[\begin{array}
@@
 Thus, the uniform distribution should distinguish fixations from the other events.
+All three submodels taken together, the joint likelihood of the observed data and hidden states can be expressed as:
+$$L(X, Y, Z, S|\mathbf{\lambda})=\rho_{S_{1}}f_{S_{1}}(X_1)f_{S_{1}}(Y_1)f_{S_{1}}(Z_1)\prod_{t=1}^{T-1}a_{S_{t}S_{t+1}}f_{S_{t+1}}(X_{t+1})f_{S_{t+1}}(Y_{t+1})f_{S_{t+1}}(Z_{t+1}),$$
+with $\mathbf{\lambda}$ denoting the vector containing the initial state and transtition probabilities as well as the response parameters.
+By summing over all possible state sequences, the likelihood of the data given the HMM parameters becomes [@Visser2011]:
+$$L(X, Y, Z|\mathbf{\lambda})=\sum_{\text{all S}}\rho_{S_{1}}f_{S_{1}}(X_1)f_{S_{1}}(Y_1)f_{S_{1}}(Z_1)\prod_{t=1}^{T-1}a_{S_{t}S_{t+1}}f_{S_{t+1}}(X_{t+1})f_{S_{t+1}}(Y_{t+1})f_{S_{t+1}}(Z_{t+1}).$$
+The parameters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997].
+The EM algorithm is generally suitable to estimate likelihoods with missing variables.
+For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [eye movement events; @Visser2019].
+When evaluating the likelihood for missing data, gazeHMM integrates over all possible values, which results in a probability density of one.
+The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability.
+Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988].
+
+## Postprocessing
+After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence.
+We implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
+Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events.
+Thus, the postprocessing routine heuristically compensates for such violations.
+This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events.
+Samples are relabeled as the state of the previous event.
+Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).
+The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
+The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
 
@@
 
-ref.acc <- matrix(c(0.92, 0.81, 0.83, 0.95, 0.91, 0.94, 0.88, 0.82, 0.83, NA, NA, NA,
-                    0.92, 0.84, 0.82, 0.95, 0.91, 0.94, 0.88, 0.80, 0.81, NA, NA, NA,
+ref.acc <- matrix(c(0.92, 0.81, 0.83, 0.95, 0.91, 0.94, 0.88, 0.82, 0.83, 0.51, 0.85, 0.83,
+                    0.92, 0.84, 0.82, 0.95, 0.91, 0.94, 0.88, 0.80, 0.81, 0.74, 0.83, 0.83,
                     0.38, 0.06, 0.11, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, NA, NA, NA,
@@
 
+# stimuliTypes <- c("dots", "img", "video")
+# 
+# kappaMat <- list()
+# 
+# for(i in 1:6) {
+#   
+#   kappaMat[[i]] = matrix(0, 2, 3)
+#   
+#   for(j in 1:length(stimuliTypes)) {
+#     
+#     events2 <- A2017[[stimuliTypes[j]]]
+#     
+#     numTot <- 0
+#     
+#     for (q in 1:length(events2)) {
+#       
+#       numTot <- numTot + nrow(events2[[q]])
+#       
+#     }
+#    
+#     event_vector <- matrix(0, numTot, 2)
+#     
+#     idx_start <- 1
+#     idx_stop <- 0
+#     
+#     for (q in 1:length(events2)) {
+#       
+#       idx_stop <- idx_start -1 + nrow(events2[[q]])
+#       
+#       for (r in 1:2) {
+#        
+#         event_vector[idx_start:idx_stop,r] <- events2[[q]][,r+5]
+#          
+#       }
+#       
+#       idx_start <- idx_start + nrow(events2[[q]])
+#     }
+#     
+#     event_vector <- rbind(event_vector, event_vector)
+#     comp_vector <- matrix(0, nrow(event_vector), ncol(event_vector))
+#     
+#     for (z in 1:2) {
+#       
+#       idx_start <- ((z-1) * nrow(event_vector)/2) + 1
+#       idx_stop <- z*(nrow(event_vector)/2)
+#       
+#       comp_vector[idx_start:idx_stop,1] <- event_vector[idx_start:idx_stop, z]
+#       
+#     }
+#     
+#     for (k in 1:length(events2)) {
+#       
+#       evMat <- event_vector == i
+#       cpMat <- event_vector == i
+#       
+#       X <- matrix(0, 2, 2)
+#       
+#       for (l in 1:2) {
+#        
+#         X[1,1] <- sum(evMat[,1] & cpMat[,1])
+#         X[1,2] <- sum(evMat[,1] & !cpMat[,1])
+#         X[2,1] <- sum(!evMat[,1] & cpMat[,1])
+#         X[2,2] <- sum(!evMat[,1] & !cpMat[,1])
+#         
+#         kappaMat[[i]][l,j] <- cohen.kappa(X)$kappa
+#       }
+#     }
+#   }
+# }
+
 A2017.cod.acc <- lapply(A2017, function(x) {
@@
 
-  events <- lapply(1:4, function(e) {
+  df <- df %>% dplyr::filter(label_MN < 4 & label_RA < 4)
+
+  events <- lapply(1:3, function(e) {
+
+    # df <- df %>% dplyr::filter(label_MN == e | label_RA == e)
 
@@
 })
+
+table(A2017.cod.acc$dots[[1]])
 

### 2020-07-23 manuscript/paper_BRM.Rmd
recompute validation results with full data

@@
 
-# Calculate event descriptives for human coders
+# Calculate event descriptives for algorithms and human coders
 
@@
 
-A2017.events <- lapply(A2017, function(stim) {
-  lapply(1:4, function(e) {
-    out <- lapply(stim, function(df) {
-
-      if(all(is.nan(df$t) || df$t == 0)) {
-
-        df$t <- seq(0, length(df$t)-1)/fr
-
+alg.names <- names(A2017[[1]][[1]])[-c(1:5)]
+
+alg.names <- c(alg.names[1:2], "gazeHMM-3", "gazeHMM-4", alg.names[3:length(alg.names)])
+
+A2017.events <- lapply(1:4, function(e) {
+  lapply(1:length(A2017), function(stim) {
+    lapply(alg.names, function(alg) {
+      if(alg == "gazeHMM-3") {
+
+        dur <- sapply(A2017.fit[[stim]], function(df){
+
+          return(df[[3]]$events[[e]]$dur)
+        }) %>% reduce(append)
+      } else if(alg == "gazeHMM-4") {
+
+        dur <- sapply(A2017.fit[[stim]], function(df){
+
+          return(df[[4]]$events[[e]]$dur)
+        }) %>% reduce(append)
       } else {
 
-        df$t <- (df$t - df$t[1])/1e6
-
-        df <- df[df$t >= 0,]
-
-      }
-
-      counter_MN <- 1
-      counter_RA <- 1
-
-      number_MN <- numeric(nrow(df))
-      number_RA <- numeric(nrow(df))
-
-      number_MN[1] <- 1
-      number_RA[1] <- 1
+        dur <- sapply(A2017[[stim]], function(df) {
+
+          counter <- 1
+
+          number <- numeric(nrow(df))
+
+          number[1] <- 1
 
           for (i in 2:nrow(df)) {
-        if(df$label_MN[i] != df$label_MN[i-1]) counter_MN <- counter_MN + 1
-        if(df$label_RA[i] != df$label_RA[i-1]) counter_RA <- counter_RA + 1
-
-        number_MN[i] <- counter_MN
-        number_RA[i] <- counter_RA
-
-      }
-
-      dur_MN <- numeric(max(number_MN))
-      event_MN <- numeric(max(number_MN))
-
-      for (n in unique(number_MN)) {
+            if(df[[alg]][i] != df[[alg]][i-1]) counter <- counter + 1
+
+            number[i] <- counter
+
+          }
+
+          dur <- numeric(max(number))
+          event <- numeric(max(number))
+
+          for (n in unique(number)) {
             if(n > 1) {
-          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - max(df$t[number_MN == (n-1)], na.rm = T)
+              dur[n] <- max(df$t[number == n], na.rm = T) - max(df$t[number == (n-1)], na.rm = T)
             } else {
-          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - min(df$t[number_MN == n], na.rm = T)
-        }
-
-        event_MN[n] <- max(df$label_MN[number_MN == n], na.rm = T)
-      }
-
-      dur_RA <- numeric(max(number_RA))
-      event_RA <- numeric(max(number_RA))
-
-      for (n in unique(number_RA)) {
-        if(n > 1) {
-          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - max(df$t[number_RA == (n-1)], na.rm = T)
-        } else {
-          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - min(df$t[number_RA == n], na.rm = T)
-        }
-
-        event_RA[n] <- max(df$label_RA[number_RA == n], na.rm = T)
-      }
-
-      return(list(dur_MN[event_MN == e], dur_RA[event_RA == e]))
-    })
-
-    df <- reduce(out, cbind)
-
-    dur_MN <- unlist(df[1,])
-    dur_RA <- unlist(df[2,])
-
-    return(data.frame(MN = c(mean(dur_MN), sd(dur_MN), length(dur_MN)),
-                      RA = c(mean(dur_RA), sd(dur_RA), length(dur_RA))))
+              dur[n] <- max(df$t[number == n], na.rm = T) - min(df$t[number == n], na.rm = T)
+            }
+
+            event[n] <- max(df[[alg]][number == n], na.rm = T)
+          }
+
+          return(dur[event == e])
+        }) %>% reduce(append)
+      }
+
+      return(data.frame(mean(dur, na.rm = T), sd(dur, na.rm = T), length(dur)))
+    }) %>% reduce(rbind)
   })
@@
 
-ref <- list(dot = list(fix = list(CDT = c(60, 127, 165),
-                                  EM = c(NA, NA, NA),
-                                  IDT = c(323, 146, 8),
-                                  IKF = c(217, 184, 72),
-                                  IMST = c(268, 140, 12),
-                                  IHMM = c(214, 286, 67),
-                                  IVT = c(203, 282, 71),
-                                  NH = c(380, 333, 30),
-                                  BIT = c(189, 113, 67),
-                                  LNS = c(NA, NA, NA)),
-                       sac = list(CDT = c(NA, NA, NA),
-                                  EM = c(17, 14, 93),
-                                  IDT = c(32, 14, 10),
-                                  IKF = c(60, 26, 29),
-                                  IMST = c(13, 5, 18),
-                                  IHMM = c(41, 17, 27),
-                                  IVT = c(36, 14, 28),
-                                  NH = c(43, 16, 42),
-                                  BIT = c(NA, NA, NA),
-                                  LNS = c(26, 11, 53)),
-                       pso = list(NH = c(24, 12, 17),
-                                  LNS = c(20, 9, 31))),
-            img = list(fix = list(CDT = c(397, 559, 251),
-                                  EM = c(NA, NA, NA),
-                                  IDT = c(399, 328, 242),
-                                  IKF = c(174, 239, 513),
-                                  IMST = c(304, 293, 333),
-                                  IHMM = c(133, 216, 701),
-                                  IVT = c(114, 204, 827),
-                                  NH = c(258, 299, 292),
-                                  BIT = c(209, 136, 423),
-                                  LNS = c(NA, NA, NA)),
-                       sac = list(CDT = c(NA, NA, NA),
-                                  EM = c(25, 22, 787),
-                                  IDT = c(25, 15, 258),
-                                  IKF = c(62, 37, 353),
-                                  IMST = c(17, 10, 335),
-                                  IHMM = c(48, 26, 368),
-                                  IVT = c(41, 22, 373),
-                                  NH = c(50, 20, 344),
-                                  BIT = c(NA, NA, NA),
-                                  LNS = c(29, 12, 390)),
-                       pso = list(NH = c(28, 13, 237),
-                                  LNS = c(25, 9, 319))),
-            vid = list(fix = list(CDT = c(213, 297, 211),
-                                  EM = c(NA, NA, NA),
-                                  IDT = c(554, 454, 48),
-                                  IKF = c(258, 296, 169),
-                                  IMST = c(526, 825, 71),
-                                  IHMM = c(234, 319, 194),
-                                  IVT = c(202, 306, 227),
-                                  NH = c(429, 336, 83),
-                                  BIT = c(248, 215, 170),
-                                  LNS = c(NA, NA, NA)),
-                       sac = list(CDT = c(NA, NA, NA),
-                                  EM = c(20, 16, 252),
-                                  IDT = c(24, 53, 41),
-                                  IKF = c(55, 20, 107),
-                                  IMST = c(18, 10, 76),
-                                  IHMM = c(42, 18, 109),
-                                  IVT = c(36, 16, 112),
-                                  NH = c(44, 18, 104),
-                                  BIT = c(NA, NA, NA),
-                                  LNS = c(28, 12, 122)),
-                       pso = list(NH = c(28, 13, 78),
-                                  LNS = c(24, 10, 87))))
-
-A2017.rmsd <- lapply(2:4, function(k) {
-  lapply(1:length(A2017.fit), function(x) { 
-    out <- lapply(1:k, function(y) {
-      out <- lapply(A2017.fit[[x]], function(z) {
-
-        if(class(z[[k]]) == "gazeHMM") z[[k]]$events[[y]]
-
-      })
-
-      df <- try(reduce(out, rbind))
-
-      alg <- try(c(mean(df$dur, na.rm = T), sd(df$dur, na.rm = T), nrow(df)))
-
-      if(y == 4) {
-        M <- try(cbind(A2017.events[[x]][[y]], alg))
-      } else {
-        M <- try(cbind(A2017.events[[x]][[y]], alg, as.data.frame(ref[[x]][[y]])))
-        M[1:2,4:ncol(M)] <- M[1:2,4:ncol(M)]/1e3
-      }
-
-      M.norm <- try(apply(M, 1, function(x) {(x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))}))
-
-      if(y == 4) {
-        rmsd <- try(sum(sqrt((M.norm[3,] - colSums(M.norm[1:2,]/2))^2)))
-      } else {
-        rmsd <- try(apply(M.norm[3:nrow(M.norm),], 1, function(r) {sum(sqrt((r - colSums(M.norm[1:2,]/2))^2))}))
-      }
-
-      coder <- try(sum(sqrt((M.norm[1,] - M.norm[2,])^2)))
-
-      rmsd <- c(coder, coder, rmsd)
-
-      return(rbind(M, rmsd))
-    })
-  })
-})
-
-# Create tables for RMSDs
-
-rmsd.table <- lapply(1:length(A2017.rmsd), function(x) {
-  out <- lapply(1:length(A2017.rmsd[[x]]), function(y) {
-    out <- lapply(1:length(A2017.rmsd[[x]][[y]]), function(z) {
-
-      tabmat <- t(apply(A2017.rmsd[[x]][[y]][[z]], 2, function(d) {as.character(round(d, 3))}))
-
-      # tabmat[3,] <- paste("**", tabmat[3,], "**", sep = "")
-
-      colnames(tabmat) <- c("Mean", "SD", "Events", "RMSD")
-      rownames(tabmat)[1:3] <- c("coderMN", "coderRA", paste0("gazeHMM-", x+1))
-
-      return(cbind(z, rownames(tabmat), tabmat))
-    })
-
-    cbind(y, reduce(out, rbind))
-  })
-
-  df <- as.data.frame(reduce(out, rbind), stringsAsFactors = F) %>%
-    mutate(Condition = factor(y, labels = c("moving dots", "image", "video")))
-
-  lapply(1:(x+1), function(x) {
-
-    df <- df %>% dplyr::filter(z == x) %>%
-      dplyr::select(-c("z", "y")) %>%
-      rename(Algorithm = V3) %>%
-      pivot_wider(names_from = "Condition", names_glue = "{Condition}_{.value}", names_sort = T, values_from = c("Mean", "SD", "Events", "RMSD"))
-
-    df %>% dplyr::select(c("Algorithm", "image_Mean", "image_SD", "image_Events", "image_RMSD",
-                           "moving dots_Mean", "moving dots_SD", "moving dots_Events", "moving dots_RMSD",
-                           "video_Mean", "video_SD", "video_Events", "video_RMSD"))
-  })
-})
-
-rmsd.table[[3]][1:3] <- lapply(1:3, function(x) {
-
-  df <- rbind(rmsd.table[[3]][[x]][1:2,], rmsd.table[[2]][[x]][3,], rmsd.table[[3]][[x]][3:nrow(rmsd.table[[3]][[x]]),])
-
-  df[c(1:3, 5:nrow(df)),2:5] <- rmsd.table[[2]][[x]][,2:5]
-
-  df[3,6:13] <- NA
-  df[4,2:5] <- NA
-
-  return(df)
-})
-
-rmsd.table[[3]][[4]][3,2:5] <- NA
+A2017.rmsd <- lapply(A2017.events, function(event) {
+  rmsd <- lapply(event, function(stim) {
+
+    M.norm <- apply(stim, 2, function(x) {(x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))})
+
+    H <- M.norm[1:2,]
+    A <- M.norm[-c(1:2),]
+
+    rmsd.alg <- apply(A, 1, function(x) {sum(sqrt((x - rowSums(H/2))^2))})
+    rmsd.cod <- sum(sqrt((H[1,] - H[2,])^2))
+
+    return(as.data.frame(cbind(stim, rmsd = c(rmsd.cod, rmsd.cod, rmsd.alg))))
+  }) %>% reduce(cbind)
+
+  out <- cbind(alg.names, rmsd, stringsAsFactors = FALSE)[-7,]
+
+  names(out) <- c("Algorithm", rep(c("moving_dots", "image", "video"), each = 4))
+
+  out[3, -c(1,6:9)] <- NA
+  out[4, 6:9] <- NA
+
+  return(out[,order(names(out))])
+})
+
+# A2017.rmsd <- lapply(2:4, function(k) {
+#   lapply(1:length(A2017.fit), function(x) { 
+#     out <- lapply(1:k, function(y) {
+#       out <- lapply(A2017.fit[[x]], function(z) {
+#         
+#         if(class(z[[k]]) == "gazeHMM") z[[k]]$events[[y]]
+#         
+#       })
+#       
+#       df <- try(reduce(out, rbind))
+#       
+#       alg <- try(c(mean(df$dur, na.rm = T), sd(df$dur, na.rm = T), nrow(df)))
+#       
+#       if(y == 4) {
+#         M <- try(cbind(A2017.events[[x]][[y]], alg))
+#       } else {
+#         M <- try(cbind(A2017.events[[x]][[y]], alg, as.data.frame(ref[[x]][[y]])))
+#         M[1:2,4:ncol(M)] <- M[1:2,4:ncol(M)]/1e3
+#       }
+#       
+#       M.norm <- try(apply(M, 1, function(x) {(x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))}))
+#       
+#       if(y == 4) {
+#         rmsd <- try(sum(sqrt((M.norm[3,] - colSums(M.norm[1:2,]/2))^2)))
+#       } else {
+#         rmsd <- try(apply(M.norm[3:nrow(M.norm),], 1, function(r) {sum(sqrt((r - colSums(M.norm[1:2,]/2))^2))}))
+#       }
+#       
+#       coder <- try(sum(sqrt((M.norm[1,] - M.norm[2,])^2)))
+#       
+#       rmsd <- c(coder, coder, rmsd)
+#       
+#       return(rbind(M, rmsd))
+#     })
+#   })
+# })
 
@@
 
-ref.acc <- matrix(c(0.92, 0.81, 0.83, 0.95, 0.91, 0.94, 0.88, 0.82, 0.83, 0.51, 0.85, 0.83,
-                    0.92, 0.84, 0.82, 0.95, 0.91, 0.94, 0.88, 0.80, 0.81, 0.74, 0.83, 0.83,
-                    0.38, 0.06, 0.11, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.00, 0.00, 0.00, 0.64, 0.66, 0.67, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.36, 0.00, 0.03, 0.45, 0.26, 0.38, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.63, 0.03, 0.14, 0.58, 0.46, 0.59, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.38, 0.00, 0.03, 0.54, 0.30, 0.52, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.67, 0.03, 0.13, 0.69, 0.60, 0.71, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.67, 0.03, 0.13, 0.75, 0.63, 0.76, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.52, 0.00, 0.01, 0.67, 0.60, 0.68, 0.24, 0.20, 0.25, NA, NA, NA,
-                    0.67, 0.03, 0.14, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.00, 0.00, 0.00, 0.81, 0.75, 0.81, 0.64, 0.59, 0.63, NA, NA, NA), ncol = 12, byrow = T)
-
-A2017.acc <- lapply(2:4, function(k) {
-  lapply(1:length(A2017.fit), function(x) { 
-    out <- lapply(1:length(A2017.fit[[x]]), function(y) {
-
-      if(class(A2017.fit[[x]][[y]][[as.character(k)]]) == "gazeHMM") {
-        cod.MN <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_MN))
-        cod.RA <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_RA))
-
-        return(rbind(cod.MN, cod.RA))
-      }
-    })
-
-    df <- reduce(out, rbind)
-
-    df[df[,1] == 0,1] <- 5
-
-    events <- lapply(1:k, function(e) {
-
-      alg <- df[,1] == e
-      cod <- df[,2] == e
+A2017.kappa <- cbind(alg.names, lapply(1:4, function(event) {
+  lapply(1:length(A2017), function(stim) {
+
+    gazeHMM.3 <- lapply(A2017.fit[[stim]], function(subj) {
+
+      subj[[3]]$samples$label
+
+    }) %>% reduce(append)
+
+    gazeHMM.4 <- lapply(A2017.fit[[stim]], function(subj) {
+
+      subj[[4]]$samples$label
+
+    }) %>% reduce(append)
+
+    df <- cbind(reduce(A2017[[stim]], rbind)[,c(6,7)], gazeHMM.3, gazeHMM.4, reduce(A2017[[stim]], rbind)[,c(8:18)])
+
+    kappa.alg <- sapply(names(df)[-c(1,2)], function(alg) {
+
+      alg <- rep(df[[alg]], 2) == event
+      cod <- c(df[,1], df[,2]) == event
 
@@
 
-      over <- mat[2,1]/sum(mat[,1])
-      under <- mat[1,2]/sum(mat[,2])
-
-      return(list(kappa = cohen.kappa(mat)$kappa, 
-                  over = over,
-                  under = under))
-    })
-
-    ratio <- mean(df[,1] != df[,2], na.rm = T)
-
-    conmat <- table(df[,1], df[,2])
-
-    return(list(events = events, ratio =  ratio, conmat = conmat))
-  })
-})
+      return(cohen.kappa(mat)$kappa)
+    })
+
+    kappa.cod <- sapply(1:2, function(cod) {
+
+      cod1 <- df[[cod]] == event
+      cod2 <- df[,c(1,2)][-cod] == event
+
+      mat <- matrix(c(sum(cod1 & cod2, na.rm = T), sum(cod1 & !cod2, na.rm = T), 
+                      sum(!cod1 & cod2, na.rm = T), sum(!cod1 & !cod2, na.rm = T)), 2, 2)
+
+      return(cohen.kappa(mat)$kappa)
+    })
+
+    return(c(kappa.cod, kappa.alg))
+  }) %>% reduce(cbind)
+}) %>% reduce(cbind))
+
+# A2017.acc <- lapply(2:4, function(k) {
+#   lapply(1:length(A2017.fit), function(x) { 
+#     out <- lapply(1:length(A2017.fit[[x]]), function(y) {
+#       
+#       if(class(A2017.fit[[x]][[y]][[as.character(k)]]) == "gazeHMM") {
+#         cod.MN <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_MN))
+#         cod.RA <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_RA))
+#         
+#         return(rbind(cod.MN, cod.RA))
+#       }
+#     })
+#     
+#     df <- reduce(out, rbind)
+#     
+#     df[df[,1] == 0,1] <- 5
+#     
+#     events <- lapply(1:k, function(e) {
+#       
+#       alg <- df[,1] == e
+#       cod <- df[,2] == e
+#       
+#       mat <- matrix(c(sum(alg & cod, na.rm = T), sum(alg & !cod, na.rm = T), 
+#                       sum(!alg & cod, na.rm = T), sum(!alg & !cod, na.rm = T)), 2, 2)
+#       
+#       over <- mat[2,1]/sum(mat[,1])
+#       under <- mat[1,2]/sum(mat[,2])
+#       
+#       return(list(kappa = cohen.kappa(mat)$kappa, 
+#                   over = over,
+#                   under = under))
+#     })
+#     
+#     ratio <- mean(df[,1] != df[,2], na.rm = T)
+#     
+#     conmat <- table(df[,1], df[,2])
+#     
+#     return(list(events = events, ratio =  ratio, conmat = conmat))
+#   })
+# })
 
@@
 
-kappa.table <- lapply(A2017.acc, function(x) {
-  out <- lapply(x, function(y){
-    out <- lapply(y$events, function(z){
-      return(round(z$kappa, 3))
-    })
-
-    return(reduce(out, cbind))
-  })
-
-  tabmat <- reduce(out, rbind)
-
-  tabmat <- rbind(tabmat[2,], tabmat[1,], tabmat[3,])
-
-  return(as.vector(tabmat))
-})
-
-conf.table <- lapply(A2017.acc, function(x) {
-  out <- lapply(x, function(y){
-
-    tabmat <- apply(y$conmat, 2, as.character)
-
-    # diag(tabmat)[1:(nrow(tabmat)-1)] <- paste("**", diag(tabmat)[1:(nrow(tabmat)-1)], "**", sep = "")
-
-    # tabmat[nrow(tabmat), 5] <- paste("**", tabmat[nrow(tabmat), 5], "**", sep = "")
-
-    colnames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks", "Other")
-    rownames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks", "Other")[1:nrow(tabmat)]
-    rownames(tabmat)[nrow(tabmat)] <- "Blinks"
-
-    return(as.data.frame(tabmat))
-  })
-})
+# kappa.table <- lapply(A2017.acc, function(x) {
+#   out <- lapply(x, function(y){
+#     out <- lapply(y$events, function(z){
+#       return(round(z$kappa, 3))
+#     })
+#     
+#     return(reduce(out, cbind))
+#   })
+#   
+#   tabmat <- reduce(out, rbind)
+#   
+#   tabmat <- rbind(tabmat[2,], tabmat[1,], tabmat[3,])
+#   
+#   return(as.vector(tabmat))
+# })
+
+A2017.conf <- cbind.data.frame(alg.names, lapply(1:length(A2017), function(stim) {
+
+  gazeHMM.3 <- lapply(A2017.fit[[stim]], function(subj) {
+
+    subj[[3]]$samples$label
+
+  }) %>% reduce(append)
+
+  gazeHMM.4 <- lapply(A2017.fit[[stim]], function(subj) {
+
+    subj[[4]]$samples$label
+
+  }) %>% reduce(append)
+
+  df <- cbind(reduce(A2017[[stim]], rbind)[,c(6,7)], gazeHMM.3, gazeHMM.4, reduce(A2017[[stim]], rbind)[,c(8:18)])
+
+  conf.alg <- sapply(names(df)[-c(1,2)], function(alg) {
+
+    if(alg == "gazeHMM.3" || alg == "gazeHMM.4") {df[[alg]][df[[alg]] == 0] <- 5}
+
+    alg <- rep(df[[alg]], 2)
+    cod <- c(df[,1], df[,2])
+
+    return(mean(alg != cod)*100)
+  })
+
+  conf.cod <- sapply(1:2, function(cod) {
+
+    cod1 <- df[[cod]]
+    cod2 <- df[,c(1,2)][-cod]
+
+    return(mean(cod1 != cod2)*100)
+  })
+
+  return(c(conf.cod, conf.alg))
+}) %>% reduce(cbind))
 
@@
 
-ref.disag <- list(c(7, 7, A2017.acc[[3]][[2]]$ratio*100, A2017.acc[[2]][[2]]$ratio*100, 23, 92, 20, 24, 20, 20, 19, 32, 31, 84),
-                  c(11, 11, A2017.acc[[3]][[1]]$ratio*100, A2017.acc[[2]][[1]]$ratio*100, 89, 96, 86, 85, 86, 84, 84, 93, 89, 93),
-                  c(19, 19, A2017.acc[[3]][[3]]$ratio*100, A2017.acc[[2]][[3]]$ratio*100, 64, 95, 61, 62, 61, 59, 59, 70, 67, 92))
-
-disag.df <- data.frame(Image = ref.disag[[1]], Dots = ref.disag[[2]], Video = ref.disag[[3]],
-                       Algorithm = c("coderMN", "coderRA", "gazeHMM-4", "gazeHMM-3", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
-                       stringsAsFactors = F)
+names(A2017.conf) <- c("Algorithm", "Dots", "Image", "Video")
+
+disag.df <- A2017.conf %>% pivot_longer(names(A2017.conf)[2:4], names_to = "Condition", values_to = "Ratio")
 
@@
 
-disag.df <- disag.df %>% pivot_longer(c("Image", "Dots", "Video"), names_to = "Condition", values_to = "Ratio")
+disag.df$Ratio[disag.df$Algorithm == "gazeHMM-3" & disag.df$Condition != "Image"] <- NA
+disag.df$Ratio[disag.df$Algorithm == "gazeHMM-4" & disag.df$Condition == "Image"] <- NA
 
@@
 
-apa_table(rmsd.table[[3]][[1]],
+apa_table(A2017.rmsd[[1]],
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
@@
 
-apa_table(rmsd.table[[3]][[2]],
+apa_table(A2017.rmsd[[2]],
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
@@
 
-apa_table(rmsd.table[[3]][[3]],
+apa_table(A2017.rmsd[[3]],
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
@@
 
-apa_table(rmsd.table[[3]][[4]],
+apa_table(A2017.rmsd[[4]],
           col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),

### 2020-07-24 manuscript/paper_BRM.Rmd
change validation results layout

@@
   return(df)
-}) %>% reduce(rbind) %>% mutate(condition = rep(c("Moving dots", "Image", "Video"), c(11, 14, 9)))
+}) %>% reduce(rbind) %>% mutate(condition = rep(c("Moving dots", "Image", "Video"), c(9, 14, 9)))
 
@@
 
-alg.names <- c(alg.names[1:2], "gazeHMM-3", "gazeHMM-4", alg.names[3:length(alg.names)])
+alg.names <- c(alg.names[1:2], "gazeHMM-3", "gazeHMM-4", alg.names[3:length(alg.names)])[-7]
 
@@
 
-      return(data.frame(mean(dur, na.rm = T), sd(dur, na.rm = T), length(dur)))
+      return(data.frame(mean(dur, na.rm = T)*1e3, sd(dur, na.rm = T)*1e3, length(dur)))
     }) %>% reduce(rbind)
@@
 
-  out <- cbind(alg.names, rmsd, stringsAsFactors = FALSE)[-7,]
+  out <- cbind(alg.names, rmsd, stringsAsFactors = FALSE)
 
@@
 
+  out[out == 0] <- NA
+
   return(out[,order(names(out))])
@@
 
-# A2017.rmsd <- lapply(2:4, function(k) {
-#   lapply(1:length(A2017.fit), function(x) { 
-#     out <- lapply(1:k, function(y) {
-#       out <- lapply(A2017.fit[[x]], function(z) {
-#         
-#         if(class(z[[k]]) == "gazeHMM") z[[k]]$events[[y]]
-#         
-#       })
-#       
-#       df <- try(reduce(out, rbind))
-#       
-#       alg <- try(c(mean(df$dur, na.rm = T), sd(df$dur, na.rm = T), nrow(df)))
-#       
-#       if(y == 4) {
-#         M <- try(cbind(A2017.events[[x]][[y]], alg))
-#       } else {
-#         M <- try(cbind(A2017.events[[x]][[y]], alg, as.data.frame(ref[[x]][[y]])))
-#         M[1:2,4:ncol(M)] <- M[1:2,4:ncol(M)]/1e3
-#       }
-#       
-#       M.norm <- try(apply(M, 1, function(x) {(x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))}))
-#       
-#       if(y == 4) {
-#         rmsd <- try(sum(sqrt((M.norm[3,] - colSums(M.norm[1:2,]/2))^2)))
-#       } else {
-#         rmsd <- try(apply(M.norm[3:nrow(M.norm),], 1, function(r) {sum(sqrt((r - colSums(M.norm[1:2,]/2))^2))}))
-#       }
-#       
-#       coder <- try(sum(sqrt((M.norm[1,] - M.norm[2,])^2)))
-#       
-#       rmsd <- c(coder, coder, rmsd)
-#       
-#       return(rbind(M, rmsd))
-#     })
-#   })
-# })
-
 ```
@@
 
-# Compute Cohen's kappa and confusion matrices
-
-A2017.kappa <- cbind(alg.names, lapply(1:4, function(event) {
-  lapply(1:length(A2017), function(stim) {
+# Compute Cohen's kappa
+
+A2017.kappa <- cbind.data.frame(alg.names, lapply(1:4, function(event) {
+  out <- lapply(1:length(A2017), function(stim) {
 
@@
 
-    df <- cbind(reduce(A2017[[stim]], rbind)[,c(6,7)], gazeHMM.3, gazeHMM.4, reduce(A2017[[stim]], rbind)[,c(8:18)])
+    df <- cbind(reduce(A2017[[stim]], rbind)[,c(6,7)], gazeHMM.3, gazeHMM.4, reduce(A2017[[stim]], rbind)[,c(8:18)])[-7]
 
@@
   }) %>% reduce(cbind)
-}) %>% reduce(cbind))
-
-# A2017.acc <- lapply(2:4, function(k) {
-#   lapply(1:length(A2017.fit), function(x) { 
-#     out <- lapply(1:length(A2017.fit[[x]]), function(y) {
-#       
-#       if(class(A2017.fit[[x]][[y]][[as.character(k)]]) == "gazeHMM") {
-#         cod.MN <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_MN))
-#         cod.RA <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_RA))
-#         
-#         return(rbind(cod.MN, cod.RA))
-#       }
-#     })
-#     
-#     df <- reduce(out, rbind)
-#     
-#     df[df[,1] == 0,1] <- 5
-#     
-#     events <- lapply(1:k, function(e) {
-#       
-#       alg <- df[,1] == e
-#       cod <- df[,2] == e
-#       
-#       mat <- matrix(c(sum(alg & cod, na.rm = T), sum(alg & !cod, na.rm = T), 
-#                       sum(!alg & cod, na.rm = T), sum(!alg & !cod, na.rm = T)), 2, 2)
-#       
-#       over <- mat[2,1]/sum(mat[,1])
-#       under <- mat[1,2]/sum(mat[,2])
-#       
-#       return(list(kappa = cohen.kappa(mat)$kappa, 
-#                   over = over,
-#                   under = under))
-#     })
-#     
-#     ratio <- mean(df[,1] != df[,2], na.rm = T)
-#     
-#     conmat <- table(df[,1], df[,2])
-#     
-#     return(list(events = events, ratio =  ratio, conmat = conmat))
-#   })
-# })
-
-# Compute Cohen's kappa between human coders
-
-# stimuliTypes <- c("dots", "img", "video")
-# 
-# kappaMat <- list()
-# 
-# for(i in 1:6) {
-#   
-#   kappaMat[[i]] = matrix(0, 2, 3)
-#   
-#   for(j in 1:length(stimuliTypes)) {
-#     
-#     events2 <- A2017[[stimuliTypes[j]]]
-#     
-#     numTot <- 0
-#     
-#     for (q in 1:length(events2)) {
-#       
-#       numTot <- numTot + nrow(events2[[q]])
-#       
-#     }
-#    
-#     event_vector <- matrix(0, numTot, 2)
-#     
-#     idx_start <- 1
-#     idx_stop <- 0
-#     
-#     for (q in 1:length(events2)) {
-#       
-#       idx_stop <- idx_start -1 + nrow(events2[[q]])
-#       
-#       for (r in 1:2) {
-#        
-#         event_vector[idx_start:idx_stop,r] <- events2[[q]][,r+5]
-#          
-#       }
-#       
-#       idx_start <- idx_start + nrow(events2[[q]])
-#     }
-#     
-#     event_vector <- rbind(event_vector, event_vector)
-#     comp_vector <- matrix(0, nrow(event_vector), ncol(event_vector))
-#     
-#     for (z in 1:2) {
-#       
-#       idx_start <- ((z-1) * nrow(event_vector)/2) + 1
-#       idx_stop <- z*(nrow(event_vector)/2)
-#       
-#       comp_vector[idx_start:idx_stop,1] <- event_vector[idx_start:idx_stop, z]
-#       
-#     }
-#     
-#     for (k in 1:length(events2)) {
-#       
-#       evMat <- event_vector == i
-#       cpMat <- event_vector == i
-#       
-#       X <- matrix(0, 2, 2)
-#       
-#       for (l in 1:2) {
-#        
-#         X[1,1] <- sum(evMat[,1] & cpMat[,1])
-#         X[1,2] <- sum(evMat[,1] & !cpMat[,1])
-#         X[2,1] <- sum(!evMat[,1] & cpMat[,1])
-#         X[2,2] <- sum(!evMat[,1] & !cpMat[,1])
-#         
-#         kappaMat[[i]][l,j] <- cohen.kappa(X)$kappa
-#       }
-#     }
-#   }
-# }
-
-A2017.cod.acc <- lapply(A2017, function(x) {
-  df <- lapply(x, function(y) {
-
-    return(y[,6:7])
-  }) %>% reduce(rbind)
-
-  df <- df %>% dplyr::filter(label_MN < 4 & label_RA < 4)
-
-  events <- lapply(1:3, function(e) {
-
-    # df <- df %>% dplyr::filter(label_MN == e | label_RA == e)
-
-    MN <- df[,1] == e
-    RA <- df[,2] == e
-
-    mat <- matrix(c(sum(MN & RA, na.rm = T), sum(MN & !RA, na.rm = T),
-                    sum(!MN & RA, na.rm = T), sum(!MN & !RA, na.rm = T)), 2, 2)
-
-    return(cohen.kappa(mat)$kappa)
-  })
-})
-
-table(A2017.cod.acc$dots[[1]])
-
-# Create tables for Cohen's kappa and confusion matrices
-
-# kappa.table <- lapply(A2017.acc, function(x) {
-#   out <- lapply(x, function(y){
-#     out <- lapply(y$events, function(z){
-#       return(round(z$kappa, 3))
-#     })
-#     
-#     return(reduce(out, cbind))
-#   })
-#   
-#   tabmat <- reduce(out, rbind)
-#   
-#   tabmat <- rbind(tabmat[2,], tabmat[1,], tabmat[3,])
-#   
-#   return(as.vector(tabmat))
-# })
-
-A2017.conf <- cbind.data.frame(alg.names, lapply(1:length(A2017), function(stim) {
+
+  return(cbind(out[,2], out[,1], out[,3]))
+}) %>% reduce(cbind), stringsAsFactors = F)
+
+# Compute disagreement
+
+A2017.disag <- cbind.data.frame(alg.names, lapply(1:length(A2017), function(stim) {
 
@@
 
-  df <- cbind(reduce(A2017[[stim]], rbind)[,c(6,7)], gazeHMM.3, gazeHMM.4, reduce(A2017[[stim]], rbind)[,c(8:18)])
-
-  conf.alg <- sapply(names(df)[-c(1,2)], function(alg) {
+  df <- cbind(reduce(A2017[[stim]], rbind)[,c(6,7)], gazeHMM.3, gazeHMM.4, reduce(A2017[[stim]], rbind)[,c(8:18)])[-7]
+
+  disag.alg <- sapply(names(df)[-c(1,2)], function(alg) {
 
@@
 
-  conf.cod <- sapply(1:2, function(cod) {
+  disag.cod <- sapply(1:2, function(cod) {
 
@@
 
-  return(c(conf.cod, conf.alg))
-}) %>% reduce(cbind))
+  return(c(disag.cod, disag.alg))
+}) %>% reduce(cbind), stringsAsFactors = F)
 
@@
 
-names(A2017.conf) <- c("Algorithm", "Dots", "Image", "Video")
-
-disag.df <- A2017.conf %>% pivot_longer(names(A2017.conf)[2:4], names_to = "Condition", values_to = "Ratio")
+names(A2017.disag) <- c("Algorithm", "Dots", "Image", "Video")
+
+disag.df <- A2017.disag %>% pivot_longer(names(A2017.disag)[2:4], names_to = "Condition", values_to = "Ratio")
 
@@
   scale_x_discrete(name = "") + coord_flip()
+
+# Compute confusion matrix
+
+A2017.conf <- as.data.frame(lapply(1:length(A2017), function(stim) {
+
+  gazeHMM.3 <- lapply(A2017.fit[[stim]], function(subj) {
+
+    subj[[3]]$samples$label
+
+  }) %>% reduce(append)
+
+  gazeHMM.4 <- lapply(A2017.fit[[stim]], function(subj) {
+
+    subj[[4]]$samples$label
+
+  }) %>% reduce(append)
+
+  df <- reduce(A2017[[stim]], rbind)[,c(6,7)]
+
+  cod <- c(df[,1], df[,2])
+
+  conf.tab <- matrix(0, 6, 5)
+
+  if(stim == 2) {
+
+    alg <- rep(gazeHMM.3, 2)
+
+    alg[alg == 0] <- 5
+
+    conf.tab[,c(1:3, 5)] <- table(cod, alg)/rowSums(table(cod, alg))
+
+  } else if(stim == 1) {
+
+    alg <- rep(gazeHMM.4, 2)
+
+    alg[alg == 0] <- 5
+
+    conf.tab[c(1:4,6),] <- table(cod, alg)/rowSums(table(cod, alg))
+
+  } else {
+
+    alg <- rep(gazeHMM.4, 2)
+
+    alg[alg == 0] <- 5
+
+    conf.tab <- table(cod, alg)/rowSums(table(cod, alg))
+
+  }
+
+  return(t(conf.tab))
+}) %>% reduce(rbind))  %>% cbind(Event = rep(c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks"), 3),
+                                 Condition = rep(c("Moving dots", "Image", "Video"), each = 5), .) %>%
+  arrange(Condition) %>%
+  dplyr::select(-Condition)
+
+names(A2017.conf) <- c("Event", "Fixation", "Saccade", "PSO", "Pursuit", "Blink", "Other")
 
@@
           font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
+          format.args = list(na_string = "-"),
           caption = "(ref:tab-rmsd-4-fix)",
@@
           font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
+          format.args = list(na_string = "-"),
           caption = "(ref:tab-rmsd-4-sac)",
@@
           font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
+          format.args = list(na_string = "-"),
           caption = "(ref:tab-rmsd-4-pso)",
@@
           font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
+          format.args = list(na_string = "-"),
           caption = "(ref:tab-rmsd-4-sp)",
@@
 
-tab.kappa.4 <- cbind(c("coderMN", "coderRA", "gazeHMM-3", "gazeHMM-4", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
-                     as.data.frame(rbind(ref.acc[1:2,], c(kappa.table[[2]], NA, NA, NA), kappa.table[[3]], ref.acc[3:nrow(ref.acc),])))
-
-tab.kappa.4[4, c(2, 5, 8, 11)] <- NA
-tab.kappa.4[3, -c(1, 2, 5, 8, 11)] <- NA
-
-apa_table(tab.kappa.4,
-          col.names = c("Algorithm", rep(c("Image", "Dots", "Video"), 4)),
-          col_spanners = list("Fixations" = c(2, 4), "Saccades" = c(5, 7), "PSOs" = c(8, 10), "Smooth pursuit" = c(11, 13)),
+A2017.kappa[3, -c(1, seq(2, 13, 3))] <- 0
+A2017.kappa[4, seq(2, 13, 3)] <- 0
+
+names(A2017.kappa) <- c("Algorithm", rep(c("Image", "Dots", "Video"), 4))
+
+apa_table(A2017.kappa,
+          col_spanners = list("Fixations" = c(2, 4), "Saccades" = c(5, 7), "PSOs" = c(8, 10), "Smooth pursuits" = c(11, 13)),
+          row.names = FALSE,
           font_size = "tiny",
@@
 
-tab.conf.4 <- cbind(Event = rep(c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks"), 3), reduce(conf.table[[3]], rbind)) %>%
-  mutate(Condition = rep(c("Moving dots", "Image", "Video"), each = 5)) %>%
-  arrange(Condition) %>%
-  dplyr::select(-Condition)
-
-apa_table(tab.conf.4,
+apa_table(A2017.conf,
           stub_indents = list("Image" = c(1:5), "Moving dots" = c(6:10), "Video" = c(11:15)),
+          row.names = FALSE,
           midrules = c(6, 12),
@@
 
-### gazeHMM With Three States
-Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data seems theoretically more appropriate.
-Including only three states could prevent gazeHMM from overclassifying smooth pursuits in the image data and yield better validation results.
-Therefore, I decided to apply gazeHMM with three states to the image data and explore the comparison to other algorithms.
-Table \@ref(tab:tab-rmsd-3) shows the RMSD between gazeHMM and human coders only for the image condition.
-For fixations, gazeHMM shows a comparably low RMSD, but it was relatively high for saccades and PSOs.
-The sample-to-sample agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kappa-3) (only for the image condition).
-For fixations, the absolute kappa indicated substantial agreement and was relatively high compared to the other algorithms.
-The absolute kappa for saccades corresponded to moderate agreement.
-Compared to the other algorithms, it was rather low.
-For PSOs, the absolute agreement was fair, but relatively low compared to the other algorithms.
-Figure \@ref(fig:plot-disag) illustrates the overall disagreement between gazeHMM with three events and the human coders compared to the other algorithms.
-In all three conditions, the disagreement was slightly lower for gazeHMM than for all the other algorithms (5-10% for images and moving dots, around 20% for videos).
-Compared to gazeHMM with four states, the RMSD for fixations decreased while the sample-to-sample agreement increased.
-For saccades and PSOs, the opposite pattern was found.
-The overall disagreement substantially decreased.
-Overall, applying gazeHMM with three states to static data seems more appropriate, since the increase in fixation classification outweighs the decrease in classification of saccades and PSOs.
-There was support for my second hypothesis since gazeHMM had lower overall disagreement to human coders, ranging between 5 and 20% across conditions, than all other algorithms.
-
-(ref:tab-kappa-3) Cohen's Kappa Between Human Coders and Algorithms for Different Events in the Image Condition
-(ref:tab-kappa-3-note) Negative values were set to zero. gazeHMM-3 classified three and gazeHMM-4 classified four events. Table adapted from @Andersson2017.
-
-```{r tab-kappa-3}
-
-tab.kappa.3 <- cbind(c("coderMN", "coderRA", "gazeHMM-4", "gazeHMM-3", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
-                     as.data.frame(rbind(ref.acc[1:2, 1:9], kappa.table[[3]][1:9], kappa.table[[2]], ref.acc[3:nrow(ref.acc), 1:9]))) %>%
-  dplyr::select(c(1, seq(2, 10, 3)))
-
-apa_table(tab.kappa.3,
-          col.names = c("Algorithm", "Fixations", "Saccades", "PSOs"),
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-kappa-3)",
-          note = "(ref:tab-kappa-3-note)")
-
-```
-
 \newpage

### 2020-07-28 manuscript/paper_BRM.Rmd
complete validation results section

@@
 library(signal)
-library(flexcircmix)
 library(CircStats)
@@
 library(here)
-library(grid)
-library(gridExtra)
-library(DiagrammeR)
-library(DiagrammeRsvg)
-library(magrittr)
-library(rsvg)
 source(here("algorithm/preprocessing_helper_functions.R"))
@@
 We used their results to compare these 10 algorithms and the two human coders with gazeHMM.
+We used the original data from the study but removed two recordings from the moving dots condition because they contained majorly samples labeled as "other" or blinks.
+Moreover, the recordings could not be matched to the results obtained by @Andersson2017.
 The second data set was published in @Ehinger2019 and has to our knowledge not yet been used for validation.
@@
 We applied our algorithm with a three-state generative model to the static condition in the @Andersson2017 data set, and with a four-state model to the dynamic conditions.
-For comparison criteria, we followed @Andersson2017: We calculated the RMSD of event durations and counts between all algorithms and the two human coders.
-Cohen's kappa was calculated for each event as the binary agreement between gazeHMM and the human coders.
-Lastly, the overall disagreement indicated which samples were classified differently by gazeHMM across all events.
+For comparison criteria, we followed @Andersson2017: We calculated the RMSD of event durations and counts between all algorithms and the average of the two human coders.
+Our results differ slightly from the original study because we excluded two recordings (leading to less events) and calculated the event durations as $Dur_{e} = max(\mathbf{t}_{e})-max(\mathbf{t}_{e-1})$, where $\mathbf{t_{e}}$ is the vector of sample time stamps for the event *e*.
+Cohen's kappa was calculated for each event as the binary agreement between gazeHMM and the average of the human coders.
+Lastly, the overall disagreement indicated which samples were classified differently by gazeHMM compared to the average of the human coders across all events.
 
@@
 Examining the Schwarz weights displayed in \@ref(fig:plot-schwarz-weights), we observed that the five-state generative model showed the highest weights in all three conditions.
-Only in the moving dots condition, two subjects each displayed the highest weights for one- and three-state models.
+Only in the moving dots condition, two subjects displayed the highest weights for the four-, and one subject for the three-state model.
 In sum, we concluded that the five-state generative model has most likely generated the @Andersson2017 data, opposing our expectations.
@@
 
-      return(data.frame(mean(dur, na.rm = T)*1e3, sd(dur, na.rm = T)*1e3, length(dur)))
+      mdur <- ifelse(is.nan(mean(dur, na.rm = T)), NA, round(mean(dur, na.rm = T)*1e3))
+      sddur <- ifelse(is.nan(sd(dur, na.rm = T)), NA, round(sd(dur, na.rm = T)*1e3))
+      nrev <- ifelse(length(dur) == 0, NA, length(dur))
+
+      return(data.frame(mdur, sddur, as.numeric(nrev)))
     }) %>% reduce(rbind)
@@
 
-  out[out == 0] <- NA
+  out <- out[!apply(out[,-1], 1, function(x) {all(is.na(x))}),]
 
@@
 
-As displayed in \@ref(tab:tab-rmsd-4-fix), gazeHMM showed a relatively low RMSD for fixations in the static condition compared to the other algorithms that were applied to the @Andersson2017 data set.
+As displayed in Figure \@ref(tab:tab-rmsd-4-fix), gazeHMM showed a relatively low RMSD for fixations in the static condition compared to the other algorithms that were applied to the @Andersson2017 data set.
 The lower RMSD for fixations indicated more similar classification to the human coders in terms of their mean and SD duration as well as the number of classified fixations.
-Oppositely, for fixations in the dynamic conditions, the RMSD of gazeHMM was the highest among the compared algorithms, suggesting substantial differences to the human coders.
+Oppositely, for fixations in the dynamic conditions, the RMSD of gazeHMM was one of the highest among the compared algorithms, suggesting substantial differences to the human coders.
 It is likely that these occured because gazeHMM classified a much larger number of fixations with very short durations.
-For saccades, gazeHMM had high RMSDs for the image and moving dots conditions, but for the video condition the RMSD was comparably low (see Table \@ref(tab:tab-rmsd-4-sac)).
-Only two other algorithms classified PSOs [NH and LNS; @Nystrom2010; @Larsson2013], and gazeHMM showed consistently higher RMSDs than LNS (see Table \@ref(tab:tab-rmsd-4-pso)).
-Compared to NH, the RMSD was higher in the image and video conditions but lower in the moving dots condition.
+For saccades, gazeHMM had a relatively high RMSD for the static condition, but it had the lowest RMSD for the moving dots condition, and a moderate value for the video condition (see Table \@ref(tab:tab-rmsd-4-sac)).
+The deviation was mostly because gazeHMM classified a higher number of saccades than the human coders.
+Only two other algorithms classified PSOs [NH and LNS; @Nystrom2010; @Larsson2013].
+Here, gazeHMM showed a consistently higher RMSD than LNS and lower RMSD than NH (see Table \@ref(tab:tab-rmsd-4-pso)).
+Our algorithm classified shorter and more PSOs than the human coders.
 No other algorithm parsed smooth pursuits, but the RMSD for gazeHMM was higher than among the human coders (see Table \@ref(tab:tab-rmsd-4-sp)).
@@
 (ref:tab-rmsd-4-fix) Fixation Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-fix-note) Durations are displayed in seconds. gazeHMM-3 classified three and gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-fix-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
 
@@
           font_size = "tiny",
-          format.args = list(na_string = "-"),
+          format.args = list(na_string = "-", drop0trailing = TRUE),
           caption = "(ref:tab-rmsd-4-fix)",
@@
 (ref:tab-rmsd-4-sac) Saccade Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-sac-note) Durations are displayed in seconds. gazeHMM-3 classified three and gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-sac-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
 
@@
           font_size = "tiny",
-          format.args = list(na_string = "-"),
+          format.args = list(na_string = "-", drop0trailing = TRUE),
           caption = "(ref:tab-rmsd-4-sac)",
@@
 (ref:tab-rmsd-4-pso) PSO Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-pso-note) Durations are displayed in seconds. gazeHMM-3 classified three and gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-pso-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
 
@@
           font_size = "tiny",
-          format.args = list(na_string = "-"),
+          format.args = list(na_string = "-", drop0trailing = TRUE),
           caption = "(ref:tab-rmsd-4-pso)",
@@
 (ref:tab-rmsd-4-sp) Smooth Pursuit Duration Descriptives and RMSD Between gazeHMM and Human Coders
-(ref:tab-rmsd-4-sp-note) Durations are displayed in seconds. gazeHMM-4 classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
+(ref:tab-rmsd-4-sp-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
 
@@
           font_size = "tiny",
-          format.args = list(na_string = "-"),
+          format.args = list(na_string = "-", drop0trailing = TRUE),
           caption = "(ref:tab-rmsd-4-sp)",
@@
 Table \@ref(tab:tab-kappa-4) contains the sample-to-sample agreement between the algorithms and human coders measured by Cohen's kappa.
-For fixations, gazeHMM shows one of the highest agreements for static, and *the* highest agreements for dynamic data.
-The absolute agreement is substantial for static and slight to fair [@Landis1977].
-For saccades, the relative agreement for gazeHMM is low for the image, moderate for the moving dots, and high for the video condition.
-In absolute terms, the agreement is moderate to substantial.
-Concerning PSOs, gazeHMM shows higher agreement than NH in the image and video conditions but consistently lower agreement compared to LNS.
-Absolutely, the agreement is slight to fair.
-Lastly, the
-
-Additionally, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of sample-to-sample agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
-Absolute kappa values were interpreted according to @Landis1977.
-For fixations, absolute kappa values in all three conditions indicated a slight to fair agreement between gazeHMM and human coders.
-Compared to the other algorithms, the agreement of gazeHMM was the lowest in the image condition but the highest in the moving dots and video condition.
-For saccades, gazeHMM showed kappa values that correspond to a moderate to substantial agreement and were relatively high compared to the other algorithms.
-However, gazeHMM never reached the highest agreement.
-For PSOs, absolute kappa values showed slight to fair agreement agreement to human coders.
-It was relatively low in the moving dots condition and second-rate in the image and video conditions compared to NH and LNS.
-The agreement for smooth pursuits was moderate in the moving dots condition, fair in the video condition, and slight in the image condition.
-No other algorithm in the study was designed to detect smooth pursuits.
-In sum, gazeHMM revealed a relatively high sample-to-sample agreement to human coders for saccades.
-For PSOs and smooth pursuits, I found fair to moderate agreement.
-The performance for fixations was absolutely and relatively low.
+For fixations, gazeHMM showed one of the highest agreements for static, and *the* highest agreements for dynamic data.
+The absolute agreement was substantial for the static and slight to fair for the dynamic conditions [@Landis1977].
+For saccades, gazeHMM had the lowest agreement for the static condition and moderate agreement for the dynamic conditions.
+In absolute terms, the agreement was fair to moderate.
+Concerning PSOs, gazeHMM showed higher agreement than NH in the image and video conditions but consistently lower agreement compared to LNS.
+Absolutely, the agreement was slight (image) to moderate (video).
+Lastly, the agreement for smooth pursuit was lower compared to the human coders and fair in absolute values.
 
 (ref:tab-kappa-4) Cohen's Kappa Between Human Coders and Algorithms for Different Conditions and Events
-(ref:tab-kappa-4-note) Negative values were set to zero. gazeHMM-3 classified three and gazeHMM-4 classified four events. Table adapted from @Andersson2017.
+(ref:tab-kappa-4-note) gazeHMM-3 classified three and gazeHMM-4 classified four events. Table design adapted from @Andersson2017.
 
@@
 
-#### Disagreement and Confusion
-At last, I computed the overall disagreement [^4] as the percentage of all samples taken together that were classified differently by gazeHMM than the human coders.
-Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events had a higher overall disagreement by between 10 and 20% than all but two of the other algorithms.
-In the moving dots and video conditions, the disagreement was between 20 and 30% lower than for all other algorithms.
-Notably, gazeHMM's disagreement was similar across conditions (around 45%).
-In summary, gazeHMM had a lower overall disagreement than all other algorithms in the moving dots and video conditions but higher disagreement in the image condition.
-
-[^4]: Even though I proposed to compute the *disagreement ratio* and @Andersson2017 also used that term, the comparison technically involves percentages and not ratios.
-Thus, I decided to use only the term *disagreement* instead.
-
-(ref:plot-disag) Disagreement between algorithms and human coders for different conditions (in %). gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.
+Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for all three conditions.
+For the static condition, the difference to most other algorithms was only slight ($\text{Md} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$), but for the dynamic conditions, the differences were substantial (video: $\text{Md} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$) and large (dots: $\text{Md} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$).
+To explore which events gazeHMM classified differently than the average of human coder, we looked at the confusion matrix between the two (see Table \@ref(tab:tab-conf-4)).
+It can be seen that gazeHMM classified many fixations as smooth pursuits and vice versa.
+Moreover, it confused many PSOs with saccades.
+The heuristic to detect blinks seemed to work successfully since gazeHMM classified most blink samples in agreement with human coding and only a minor part was mistaken for saccades.
+
+(ref:plot-disag) Disagreement between algorithms and human coders for different conditions (in %). gazeHMM-3 classified three and gazeHMM-4 classified four events.
 
@@
 (ref:tab-conf-4-note) gazeHMM classified four events and blinks.
+Values indicate proportions of samples where gazeHMM and human coders agree divided by the total number of samples classified by the human coders for each event (i.e., columns sum to one).
 
@@
 
-Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms in terms of RMSD and sample-to-sample agreement.
-The two criteria rather indicated a moderate performance of gazeHMM.
-However, the overall disagreement was lower for gazeHMM than for all other algorithms in the dynamic conditions but the highest in the static image condition.
-Looking at the confusion matrix in Table \@ref(tab:tab-conf-4), it can be seen that gazeHMM confused fixations and smooth pursuit to a large extend.
-In the static image condition, it overclassified smooth pursuit events, while they were underclassified in the dynamic conditions.
-
 \newpage

### 2020-07-29 manuscript/paper_BRM.Rmd
adjust discussion section

@@
 @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-Probabilistic models are frequently implemented to classify events.
+Probabilistic models are frequently implemented to classify eye movement events.
 Instead of discrete classification, these models assign probabilities for belonging to an event to each sample.
@@
 The heuristic to detect blinks seemed to work successfully since gazeHMM classified most blink samples in agreement with human coding and only a minor part was mistaken for saccades.
+Inspecting an example of gaze data classified by gazeHMM compared to human coding leads to a similar notion: Figure \@ref(fig:plot-example-classificaiton) illustrates that gazeHMM is rapidly switching between classifying fixations and smooth pursuits, whereas the human coder treats the same data as one large smooth pursuit event.
+In the example, the human coder also delayed the start of the PSO compared to gazeHMM.
 
@@
 
+(ref:plot-example-classification) Example data from @Andersson2017 displayed as x-, and y-coordinates, velocity (in deg/s), acceleration (in deg/s$^2$), and sample-to-sample (relative) angle (in rad).
+Colors indicate classification by gazeHMM and symbols by human coder MN.
+Note that the human coder classified all fixations by gazeHMM as smooth pursuits and delayed the separation between the saccade and PSO.
+
+```{r plot-example-classification, fig.cap="(ref:plot-example-classification)", fig.height=7.5}
+
+cbind(A2017.fit[[3]][[3]][[4]]$samples, coderMN = as.factor(A2017[[3]][[3]]$coderMN)) %>% 
+  gather(c("x", "y", "vel", "acc", "angle"), key = metric, value = value, factor_key = T) %>% 
+  mutate_at(vars(label, state), as.factor) %>%
+  dplyr::filter(t > 4.0, t <= 4.5) %>% 
+  rename(gazeHMM = label, Time = t) %>% 
+  mutate(metric = fct_recode(metric, "x-coordinate" = "x", "y-coordinate" = "y", "Velocity" = "vel",
+                             "Acceleration" = "acc", "Rel. angle" = "angle"),
+         coderMN = fct_recode(coderMN, "Fix" = "1", "Sacc" = "2", "PSO" = "3", "SP" = "4", "Blink" = "5", "Other" = "6"),
+         gazeHMM = fct_recode(gazeHMM, "Noise" = "0", "Fix" = "1", "Sacc" = "2", "PSO" = "3", "SP" = "4")) %>%
+  ggplot(aes(x = Time, y = value)) + 
+  facet_wrap(vars(metric), ncol = 1, scales = "free", strip.position = "left") +
+  geom_path() + xlab("Time in s") + ylab("") +
+  geom_point(aes(color = gazeHMM, shape = coderMN), size = 0.8) + 
+  theme(strip.placement = "outside",
+        axis.text = element_text(size = 8))
+
+```
+
+# Discussion
+In this report, we presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events.
+The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von Mises and a uniform distribution.
+An HMM serves as the generative model of the algorithm and classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
+We showed in a simulation study that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions.
+Adding a fourth event (i.e., smooth pursuit) to the model and noise to the generated data were the most critical factors for decreasing recovery.
+Furthermore, we applied gazeHMM with different numbers of states to benchmark data by @Andersson2017 and compared the model fit.
+The model comparison revealed that a five-state HMM had consistently most likely generated the data.
+This result opposed our expectation that a three-state model would be preferred for static and a four-state model for dynamic data.
+When comparing gazeHMM against other algorithms, gazeHMM showed mostly good agreement to human coding.
+On one hand, it outperformed the other algorithms in the overall disagreement to human coding.
+On the other hand, gazeHMM confused a lot of fixations with smooth pursuits, which led to a rapid switching between the two events.
+It also tended to mistake PSO samples as belonging to saccades.
+Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle.
+The overlap can cause the HMM to confuse parameters and hidden states.
+The decrease in parameter recovery (especially for scale parameters) due to noise shows that the overlap is enhanced by more dispersion in the data.
+The scale parameters might be particularly vulnerable to extreme data points.
+Despite these drawbacks, the recovery of the generative model in gazeHMM seems very promising.
+The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).
+The model comparison on the benchmark data suggested that the generative model in gazeHMM is not  optimally specified for eye movement data, yet. There are several explanations for this result:  
+*There were more eye movement events present in the data than we expected*.
+Eye movement events can be divided into subevents.
+For example, fixations consist of drift and tremor movements [@Duchowski2017] and PSOs encompass dynamic, static, and glissadic over- and undershoots [@Larsson2013].
+A study on a recently developed HMM algorithm supports this explanation: @Houpt2018 applied the unsupervised BP-AR-HMM algorithm to the @Andersson2017 data set and classified more distinct states than the human coders.
+Some of the states classified by BP-AR-HMM matched the same event coded by humans.
+Since the subevents are usually not interesting for users of classification algorithms, the ability of HMMs to classify might limit their ability to generate eye movements.
+*Model selection criteria are generally not appropriate for comparing HMMs with different numbers of states*.
+This argument has been discussed in the field of ecology [see @Li2017], where studies found that selection criteria preferred models with more states than expected [similar to the result of this study; e.g., @Langrock2015].
+@Li2017 explain this bias with the simplicity of the submodels in HMMs: Initial state, transition, and response models for each state are usually relatively simple.
+When they do not describe the processes in the respective states accurately, the selection criteria compensate for that by preferring a model with more states.
+Thus, there are not more latent states present in the data, but the submodels of the HMM are misspecified or too simple.
+Correcting for model misspecifications led to better model recovery in studies on animal movements [@Langrock2015; @Li2017].
+However, @Pohle2017 showed in simulations that the ICL identified the correct model despite several misspecifications.
+It has to be noted that the study by @Pohle2017 only used data generating models with two states.
+*The submodels of gazeHMM were misspecified*.
+@Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models.
+Both situations could apply to gazeHMM and eye movement data.
+Outliers occur frequently in eye-tracking data due to measurement error.
+Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task the data is obtained from [@Langrock2015].
+Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf.
+@Li2017].
+This aspect could indeed oversimplify the complex nature of eye movement data.
+Comparing gazeHMM to other algorithms on benchmark data showed that gazeHMM showed good agreement with human coders.
+However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yielded different results.
+The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events (incl.
+blinks).
+Nevertheless, Cohen's kappa values of 0.67 (fixations - image - three states) or 0.62 (saccades - moving dots) indicate substantial agreement to human coders, especially in light of the maximum references from the simulation study.
+At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
+Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and to find errors in their design.
+
+## Advantages of gazeHMM
+In view of the four proposed goals that gazeHMM should fulfill, we can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they only have minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
+Their default values should be appropriate for most applications.
+A major advantage of gazeHMM is that it does not require human labeled data as input.
+Instead, it estimates all parameters and hidden states from the data.
+Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018a], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
+This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data.
+Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit.
+Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2018], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
+However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events.
+Therefore, we recommend using gazeHMM with four events only for exploratory purposes.
+Without smooth pursuits, we consider gazeHMM's classification as appropriate for application.
+Lastly, its implementation in R using depmixS4 [@Visser2010] should make gazeHMM a tool that is easy to use and customize for individual needs.
+
+## Future Directions
+Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for examples, see @Balakrishnan2009].
+Potential problems of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters).
+Another option to model the correlation could be to include one of the response variables as a covariate of the other.
+Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred.
+This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.
+Third, one solution to diverging results when comparing gazeHMM with different events could be model averaging: Instead of using the maximum posterior state probability of each sample from the preferred model, the probabilities could be weighted according to a model selection criterion (e.g., Schwarz weight) and averaged.
+Then, the maximum averaged probability could be used to classify the samples into events.
+This approach could lead to a more robust classification because it reduces the overconfidence of each competing model and easily adapts to new data [analogous to Bayesian model averaging; @Hinne2020].
+However, the model comparison for gazeHMM often showed extreme weights for a five-state model, which would lead to a very limited influence of the other models in the averaged probabilities.
+Fourth, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data.
+As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements.
+As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
+Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and auto-correlation functions [see @Zucchini2016].
+Fifth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonably short saccades, gazeHMM could explicitly model the duration of events.
+This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2006].
+Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs.
+This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades.
+Drawbacks of this extension are higher computational costs and difficulties with including covariates [@Zucchini2016].
+Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM.
+This could potentially be achieved by using different response distributions or parameter optimization methods.
+Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.
+
+## Conclusion
+Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events.
+For smooth pursuits, the classification is not optimal and thus not recommended, yet.
+On one side, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one.
+On the other side, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events.
+These difficulties highlight challenges for HMMs in general but also specifically for eye movements.
+
 \newpage

### 2020-08-03 manuscript/paper_BRM.Rmd
add general writing improvements

@@
 @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-Probabilistic models are frequently implemented to classify eye movement events.
+Gaze data has been frequently classified by probabilistic models.
 Instead of discrete classification, these models assign probabilities for belonging to an event to each sample.
-Probabilistic models have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
+They have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
 324].
@@
 Only @Mihali2017 used this property by building an explicit generative model for drift and microsaccades.
-For this article, we developed a novel algorithm for classifying gaze data named gazeHMM that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms.
-The development was guided by four goals: (a) it should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it should confirm rather than explore the presence of events in the data.
+For this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms.
+The development was guided by four goals: (a) the algorithm should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it should confirm rather than explore the presence of events in the data.
 
@@
 Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events.
-Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018 and @Andersson2017].
+Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018, and @Andersson2017].
 However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007].
@@
 Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
-Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017].
+Due to tremor, we assume that the angle between samples should not follow any direction but a random walk [@Duchowski2017; @Larsson2015].
 In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction.
@@
 The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010].
-In contrast, the absolute angle between two samples is calculated as the finite backward difference and the sample-to-sample angle as the finite forward difference.
+The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}) -\arctan(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}),$$ with $\alpha(t) + 2\pi$ for $\alpha(t) < 0$.
 Most of the missing data in eye movement classification are due to blinks.
-Therefore, the user can indicate in gazeHMM which samples should be labeled as blinks (other missing samples are treated as noise).
+Therefore, the user can indicate which samples should be labeled as blinks (other missing samples are treated as noise).
 Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs.
@@
 HMMs can be described by three submodels: An initial state model, a transition model, and a response model.
-The initial state model contains probabilites for the first state of the hidden sequence $\rho_i = P(S_1=i)$, with *i* denoting the hidden states.
+The initial state model contains probabilites for the first state of the hidden sequence $\rho_i = P(S_1=i)$, with *i* denoting the hidden state.
 In gazeHMM, the initial states are modeled by a multinomial distribution.
 The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM.
-Typically, probabilities to transition from state *i* to *j* $a_{ij} = P(S_{t+1}=j|S_{t}=i)$ are expressed in matrix form [@Visser2011]:
+Typically, probabilities to transition from state *i* to *j*, $a_{ij} = P(S_{t+1}=j|S_{t}=i)$, are expressed in matrix form [@Visser2011]:
 $$\mathbf{A} = \left[\begin{array}
@@
 Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988].
+The implementation in depmixS4 allows to include time varying covariates for each parameter in the HMM.
+In gazeHMM, no such covariates were included and thus, only intercepts were estimated for each parameter.
 
@@
 Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).
-The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
+The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, CircStats [@Agostinelli2018] for the von Mises distribution, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
 The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
@@
 The study was divided in four parts.
-Here, we only report the first two parts, the other two parts, which address starting values and missing data, can be found in the supplementary material (URL).
+Here, we only report the first two parts, the other two parts, which address starting values and missing data, can be found in the supplementary material (www.github.com/maltelueken/gazeHMM_validation).
 The HMM repeatedly generated data with a set of parameters (true parameter values).
@@
 Since the concentration parameters of the von Mises distribution are the inverse of standard deviations, they were varied on the inverse scale.
-An example of simulated data from the HMM with default parameters is visualized in Figure \@ref(fig:plot-sim-ex).
 In the second part, we manipulated the sample size of the generated data and the amount of noise added to it.
@@
 ### Parameter Variation
-In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation of estimated parameters and accuracy of estimated state sequences.
+In the first part of the simulation, we examined how varying the parameters in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
 Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM.
 The RMdSPD was below 0.1 for all estimated and manipulated parameters, indicating good recovery [^2].
-The regressions between manipulated true and estimated parameters are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
+The regression lines between manipulated true and estimated parameters are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
 With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely.
@@
 Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
-Therefore, I decided not to include it in the analysis.
+Therefore, we decided not to include it in the analysis.
 
@@
 Otherwise, RMdSPDs were consistently lower than 0.1, indicating good recovery.
-Inspecting the regressions between manipulated true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one).
+Inspecting the regression lines between true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one).
 In contrast to the two-state HMM, larger deviations and more outliers were observed.
@@
 For most estimated models, the kappa values between true and estimated state sequences were above 0.95, meaning almost perfect agreement.
-However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched.
+However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched (see supplementary material).
 
@@
 For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were between 0.1 and 0.5, suggesting moderate recovery.
-Estimated kappa parameters in state four were also often moderately recovered when parameters in states two, three, and four were varied.
+Also, estimated kappa parameters in state four were often moderately recovered when parameters in states two, three, and four were varied.
 Otherwise, RMdSPDs were below 0.1, indicating good recovery.
-Looking at the regressions between true and estimated parameters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) illustrate strong and unbiased relationships.
+Looking at the regression lines between true and estimated parameters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) exhibit strong and unbiased relationships.
 However, there were larger deviations and more outliers than in the previous models, especially for states one and four.
@@
 Similarly, $\beta_{vel}$ and $\beta_{acc}$ in states two and three showed bad recovery.
-For states two and three, higher samples sizes showed slightly lower RMdSPDs.
+For states two and three, higher sample sizes showed slightly lower RMdSPDs.
 As in the previous part, most Cohen's kappa values ranged between 0.6 and 0.9, meaning substantial to almost perfect agreement between true and estimated states (Figure \@ref(fig:plot-acc-2), right plot).
@@
 @Andersson2017 used the data to compare 10 different classification algorithms.
-We used their results to compare these 10 algorithms and the two human coders with gazeHMM.
+We adopted their results to compare these 10 algorithms and the two human coders with gazeHMM.
 We used the original data from the study but removed two recordings from the moving dots condition because they contained majorly samples labeled as "other" or blinks.
@@
 The second data set was published in @Ehinger2019 and has to our knowledge not yet been used for validation.
-Here, we only use tasks four and five out of 10 tasks because these are qualitatively different to the first data set.
+Here, we only took tasks four and five out of 10 tasks because these are qualitatively different to the first data set.
 In task four, subjects were instructed to fixate a central target for 20 s.
@@
 ## Data Analysis
-A successful validation of gazeHMM was determined by using two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets.
+A successful validation of gazeHMM was determined by two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets.
 The fits of the generative models were compared using Schwarz weights [@Wagenmakers2004], a weighted version of the BIC.
@@
 For comparison criteria, we followed @Andersson2017: We calculated the RMSD of event durations and counts between all algorithms and the average of the two human coders.
-Our results differ slightly from the original study because we excluded two recordings (leading to less events) and calculated the event durations as $Dur_{e} = max(\mathbf{t}_{e})-max(\mathbf{t}_{e-1})$, where $\mathbf{t_{e}}$ is the vector of sample time stamps for the event *e*.
-Cohen's kappa was calculated for each event as the binary agreement between gazeHMM and the average of the human coders.
-Lastly, the overall disagreement indicated which samples were classified differently by gazeHMM compared to the average of the human coders across all events.
+Our results differ slightly from the original study because we excluded two recordings (leading to less events) and calculated the event durations as $Dur(e) = max(\mathbf{t}_{e})-max(\mathbf{t}_{e-1})$, where $\mathbf{t_{e}}$ is the vector of sample time stamps for the event *e*.
+Cohen's kappa was calculated for each event as the binary agreement between the algorithms and the average of the human coders.
+Lastly, the overall disagreement indicated which samples were classified differently by the algorithms compared to the average of the human coders across all events.
+The human coders were compared directly to each other.
 
@@
 
-Examining the Schwarz weights displayed in \@ref(fig:plot-schwarz-weights), we observed that the five-state generative model showed the highest weights in all three conditions.
+Examining the Schwarz weights displayed in Figure \@ref(fig:plot-schwarz-weights), we observed that the five-state generative model showed the highest weights in all three conditions.
 Only in the moving dots condition, two subjects displayed the highest weights for the four-, and one subject for the three-state model.
@@
 
-As displayed in Figure \@ref(tab:tab-rmsd-4-fix), gazeHMM showed a relatively low RMSD for fixations in the static condition compared to the other algorithms that were applied to the @Andersson2017 data set.
+As displayed in Table \@ref(tab:tab-rmsd-4-fix), gazeHMM showed a relatively low RMSD for fixations in the static condition compared to the other algorithms that were applied to the @Andersson2017 data set.
 The lower RMSD for fixations indicated more similar classification to the human coders in terms of their mean and SD duration as well as the number of classified fixations.
 Oppositely, for fixations in the dynamic conditions, the RMSD of gazeHMM was one of the highest among the compared algorithms, suggesting substantial differences to the human coders.
-It is likely that these occured because gazeHMM classified a much larger number of fixations with very short durations.
+It can be seen that gazeHMM classified a much larger number of fixations with very short durations.
 For saccades, gazeHMM had a relatively high RMSD for the static condition, but it had the lowest RMSD for the moving dots condition, and a moderate value for the video condition (see Table \@ref(tab:tab-rmsd-4-sac)).
@@
 Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for all three conditions.
-For the static condition, the difference to most other algorithms was only slight ($\text{Md} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$), but for the dynamic conditions, the differences were substantial (video: $\text{Md} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$) and large (dots: $\text{Md} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$).
+For the static condition, the difference to most other algorithms was only slight ($\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$%), but for the dynamic conditions, the differences were substantial (video: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$%) and large (dots: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$%).
 To explore which events gazeHMM classified differently than the average of human coder, we looked at the confusion matrix between the two (see Table \@ref(tab:tab-conf-4)).
@@
 The heuristic to detect blinks seemed to work successfully since gazeHMM classified most blink samples in agreement with human coding and only a minor part was mistaken for saccades.
-Inspecting an example of gaze data classified by gazeHMM compared to human coding leads to a similar notion: Figure \@ref(fig:plot-example-classificaiton) illustrates that gazeHMM is rapidly switching between classifying fixations and smooth pursuits, whereas the human coder treats the same data as one large smooth pursuit event.
+Inspecting an example of gaze data classified by gazeHMM compared to human coding leads to a similar notion: Figure \@ref(fig:plot-example-classification) illustrates that gazeHMM is rapidly switching between classifying fixations and smooth pursuits, whereas the human coder treats the same data as one large smooth pursuit event.
 In the example, the human coder also delayed the start of the PSO compared to gazeHMM.
@@
 Outliers occur frequently in eye-tracking data due to measurement error.
-Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task the data is obtained from [@Langrock2015].
+Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task from which the data are obtained from [@Langrock2015].
 Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf.
@@
 blinks).
-Nevertheless, Cohen's kappa values of 0.67 (fixations - image - three states) or 0.62 (saccades - moving dots) indicate substantial agreement to human coders, especially in light of the maximum references from the simulation study.
+Nevertheless, Cohen's kappa values of 0.67 (fixations - image) or 0.62 (saccades - moving dots) indicate substantial agreement to human coders, especially in light of the maximum references from the simulation study.
 At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
@@
 As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
-Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and auto-correlation functions [see @Zucchini2016].
+Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and autocorrelation functions [see @Zucchini2016].
 Fifth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonably short saccades, gazeHMM could explicitly model the duration of events.

### 2020-08-06 manuscript/paper_BRM.Rmd
add abstract

@@
   Eye-tracking allows researchers to infer cognitive processes from eye movements that are classified into distinct events.
-Parsing the events is typically done by algorithms that transform, filter, classify, and merge raw data samples.
-Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere weaknesses.
-Therefore, I developed gazeHMM, an HMM algorithm that has no critical parameters to be set by users, does not require human coded data as input, and classifies fixations, saccades, PSOs, and smooth pursuits.
-The development was guided by the question of whether HMMs are useful at describing eye movements and whether they improve the event classification.
-I evaluated gazeHMM in a simulation study and on benchmark data.
+Parsing the events is typically done by algorithms.
+Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere several weaknesses.
+To adress these weaknesses, we developed gazeHMM, an algorithm that uses an HMM as a generative model, has no critical parameters to be set by users, and does not require human coded data as input.
+The algorithm classifies gaze data into fixations, saccades, and optionally PSOs and smooth pursuits.
+We evaluated the performance of gazeHMM in a simulation study and on two benchmark data sets.
 The simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences.
-Critical exceptions for good recovery were adding a smooth pursuit like state to gazeHMM and noisy data.
-For benchmark data, model comparisons with gazeHMM yielded preferred models with more states than expected.
-I assessed the classification performance of gazeHMM compared to other algorithms by the agreement to human event coding.
-Here, gazeHMM improved the event classification but did not outperform all other algorithms in most cases.
-Both the simulation study and benchmark application showed poor classification performance for smooth pursuits.
-Thus, I advice to classify smooth pursuits only for exploration and recommend gazeHMM with fixations, saccades, and PSOs for application.
+Critical exceptions for good recovery were including the smooth pursuit state in the generative model and noisy data.
+The performance on benchmark data was assessed by model comparison and agreement of the classification to human coding compared to other algorithms.
+The model comparison yielded preferred generative models for the benchmark data with more events than we expected.
+Nevertheless, gazeHMM showed relatively high similarity to human coding and outperformed the other algorithms in some agreement criteria.
+For dynamic stimuli, our algorithm tended to switch rapidly between fixations and smooth pursuits.
+Thus, we recommend to use gazeHMM for classifiying smooth pursuits only in explorative settings.
 Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.
@@
 keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
-wordcount         : "12962"
+wordcount         : "8861"
 
@@
 Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events.
-Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018, and @Andersson2017].
+Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017].
 However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007].
@@
 The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010].
-The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}) -\arctan(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}),$$ with $\alpha(t) + 2\pi$ for $\alpha(t) < 0$.
+The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}) -\arctan(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$.
 Most of the missing data in eye movement classification are due to blinks.
@@
 The EM algorithm is generally suitable to estimate likelihoods with missing variables.
-For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [eye movement events; @Visser2019].
+For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [eye movement events; @Visser2011].
 When evaluating the likelihood for missing data, gazeHMM integrates over all possible values, which results in a probability density of one.
@@
 Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).
-The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, CircStats [@Agostinelli2018] for the von Mises distribution, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
+The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, CircStats [@Lund2018] for the von Mises distribution, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
 The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
@@
 The values for the response model are displayed in Table \@ref(tab:tab-starting-values).
+For a fifth eye movement event, we chose starting values that would enable the HMM to split any other event into two subevents (e.g., fixations into drift and microsaccades).
 In contrast to the simulation study, generating random starting values often led to bad model fits and label switching between states.
@@
 (ref:tab-starting-values-note) Starting values for velocity and acceleration signals are shown before scaling down by factor 100.
-Values for event 5 were chosen so that they approximately match plausible distributions for microsaccades.
 
@@
 We adopted their results to compare these 10 algorithms and the two human coders with gazeHMM.
-We used the original data from the study but removed two recordings from the moving dots condition because they contained majorly samples labeled as "other" or blinks.
+We used the original data from the study but removed two recordings from the moving dots condition because they majorly contained samples labeled as "other" or blinks.
 Moreover, the recordings could not be matched to the results obtained by @Andersson2017.
@@
 Nevertheless, Cohen's kappa values of 0.67 (fixations - image) or 0.62 (saccades - moving dots) indicate substantial agreement to human coders, especially in light of the maximum references from the simulation study.
-At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
+At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018 observed substantial differences between coders and within coders over time.
 Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and to find errors in their design.
@@
 Instead, it estimates all parameters and hidden states from the data.
-Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018a], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
+Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
 This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data.
@@
 
+# Open Practices and Data Availability Statement
+The data and materials for the validation study [@Andersson2017; @Larsson2013; @Larsson2015] are available on GitHub at https://github.com/richardandersson/EyeMovementDetectorEvaluation.
+The code to obtain the results for the simulation and validation study is available at https://github.com/maltelueken/gazeHMM_validation and the algorithm can be accessed at https://github.com/maltelueken/gazeHMM.
+The simulation study was preregistered (https://doi.org/10.17605/OSF.IO/VDJGP), whereas the validation study was not.
+
 \newpage

### 2020-08-26 manuscript/paper_BRM.Rmd
Adapt to minor comments by Simon

@@
 author: 
-  - name          : "Malte Lueken"
+  - name          : "Malte Lüken"
     affiliation   : "1"
@@
     email         : "malte_lueken@arcor.de"
-  - name          : "Simon Kucharsky"
+  - name          : "Šimon Kucharský"
     affiliation   : "1"
@@
 Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015].
-Commonly distinguished events are fixations, saccades, smooth pursuits, and post-saccadic oscillations (PSOs).
-Classifying eye-tracking data reduces their complexity and is the first step towards cognitive interpretation [@Salvucci2000].
-The classification is normally done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017].
+Basic events are fixations, saccades, smooth pursuits, and post-saccadic oscillations (PSOs).
+Classifying raw eye-tracking data into these events reduces their complexity and is usually the first step towards cognitive interpretation [@Salvucci2000].
+The classification is typically done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017].
 @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
 Gaze data has been frequently classified by probabilistic models.
-Instead of discrete classification, these models assign probabilities for belonging to an event to each sample.
+Compared to discrete (hard) classification, these models compute the probability of each event for every sample of the raw eye-tracking data.
 They have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
@@
 One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs).
-They estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal.
+HMMs estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal.
 Each gaze sample depends on its corresponding state.
@@
 In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous).
-On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]).
+Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need training labeled data as it would be the case for supervised machine learning.
+On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a fixed threshold approach (named "Identification by HMM" [I-HMM]).
 Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM.
@@
 NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].
-In sum, HMMs seem to be a promising method for classifying eye movements.
-Still, the existing HMM algorithms each have their weaknesses that could be improved upon (and that could explain their inferior performance in comparison to other methods).
-First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
+In sum, HMMs seem to be a promising method for classifying eye movements. However, the existing HMM algorithms each have their weaknesses that could be improved upon.  
+First, I-HMM relies on setting an appropriate threshold to determine the initial classification, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
 Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability.
@@
 HMMs have another property that has been rarely used in the context of eye movement classification: They can serve as a generative model for gaze data.
-Only @Mihali2017 used this property by building an explicit generative model for drift and microsaccades.
+To our knowledge, only @Mihali2017 used this property by building an explicit generative model for drift and microsaccades.
 For this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms.
@@
 However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007].
-This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics [@Andersson2017].
+This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017].
 In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.
@@
 Due to tremor, we assume that the angle between samples should not follow any direction but a random walk [@Duchowski2017; @Larsson2015].
-In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction.
+In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction.
 PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013].
@@
 Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
-Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, they were not in the scope of gazeHMM.
+Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.
 The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010].
-The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}) -\arctan(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$.
+The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan\left(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}\right) -\arctan\left(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}\right),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$, and is therefore bound between 0 and $2\pi$.
 Most of the missing data in eye movement classification are due to blinks.
@@
 To model the sample-to-sample angle, we pursued a novel approach in gazeHMM: A mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution: 
-$$Z \sim vonMises(\mu, \kappa)+U(0, 2\pi).$$
+$$\begin{aligned}
+Z \mid fixation &\sim U(0, 2\pi) \\
+Z \mid saccade &\sim von\ Mises(\mu, \kappa) \\
+Z \mid PSO &\sim von\ Mises(\mu, \kappa) \\
+Z \mid smooth\ pursuit &\sim von\ Mises(\mu, \kappa)
+\end{aligned}.$$
 Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions.
@@
 Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).
-The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, CircStats [@Lund2018] for the von Mises distribution, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
+The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages \texttt{signal} [@Ligges2015] to compute velocity and acceleration signals, \texttt{depmixS4} [@Visser2010] for the HMM, \texttt{CircStats} [@Lund2018] for the von Mises distribution, and \texttt{BB} [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
 The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
@@
 ## Starting Values
-The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
-Random starting values for the estimation of shape, scale, and concentration parameters were generated by gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$, with $\psi_{true}$ being the true value of the parameter to be estimated.
+The HMM always started with a uniform distribution for the initial state and state transition probabilities.
+Random starting values for the estimation of shape, scale, and concentration parameters were generated by gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start;i}=\theta_i/2$, with $\theta_i$ being the true value of the parameter to be estimated in simulation $i\in(1,\dots,I)$.
 This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
@@
 ## Data Analysis
-For each parameter, we calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
-Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\psi_{true} \ne 0$.
+For each parameter, we calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values:
+$$
+\begin{aligned}
+\text{RMdSPD} & = \sqrt{\text{Median}(\epsilon_1^2,\dots,\epsilon_I^2)} \\ 
+\epsilon_i^2 & = \left(\frac{\hat{\theta}_i - \theta_i}{\theta_i}\right)^2, 
+\end{aligned}
+$$
+where $\theta_i$ is the true parameter value and $\hat{\theta}_i$ is the estimated parameter value for simulation $i\in(1,\dots,I)$, respectively.
+Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\theta_i \ne 0$.
 This was not the case for some mean parameters of the von Mises distributions.
-In those cases, we used $\psi_{true} = 2\pi$ instead.
+In those cases, we used $\theta_i = 2\pi$ instead.
 We treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter.
@@
 Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 (ref:plot-regtr-1-2) Regression lines between true and estimated transition probabilities for the two-state HMM in part one.
-Top facet labels show to and right facet labels show from which state the HMM is moving.
+Right facet labels show *from* and top facet labels show *to* which state the HMM is moving.
 Dashed lines refer to perfect recovery.
@@
 (ref:plot-regtr-1-3) Regression lines between true and estimated transition probabilities for the three-state HMM in part one.
-Top facet labels show to and right facet labels show from which state the HMM is moving.
+Right facet labels show *from* and top facet labels show *to* which state the HMM is moving.
 Dashed lines refer to perfect recovery.
@@
 Otherwise, RMdSPDs were below 0.1, indicating good recovery.
-Looking at the regression lines between true and estimated parameters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) exhibit strong and unbiased relationships.
+Looking at Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4), the regression lines between true and estimated parameters exhibit strong and unbiased relationships.
 However, there were larger deviations and more outliers than in the previous models, especially for states one and four.
-Cohen's kappa ranged majorly between 0.6 and 0.9, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
-Here, some kappa values clustered around 0.25 and zero, which, again, can be interpreted as the result of label switching.
+Cohen's kappa ranged mostly between 0.6 and 0.9, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
+Here, some kappa values clustered around 0.25 and zero, which, again, suggests label switching.
 Overall, simulations in part one demonstrated that the HMM recovered parameters very well when they were manipulated.
@@
 (ref:plot-regtr-1-4) Regression lines between true and estimated transition probabilities for the four-state HMM in part one.
-Top facet labels show to and right facet labels show from which state the HMM is moving.
+Right facet labels show *from* and top facet labels show *to* which state the HMM is moving.
 Dashed lines refer to perfect recovery.
@@
 Top facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 A successful validation of gazeHMM was determined by two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets.
-The fits of the generative models were compared using Schwarz weights [@Wagenmakers2004], a weighted version of the BIC.
+The fits of the generative models were compared using Schwarz weights [@Wagenmakers2004], a conversion of the BIC [@Schwarz1978] into model weights.
 They can be interpreted as the probability of a model having generated the data compared to the competing models.
@@
 As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements.
-As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
+A candidate for such a covariate might be a periodic function of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
 Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and autocorrelation functions [see @Zucchini2016].
@@
 This could potentially be achieved by using different response distributions or parameter optimization methods.
-Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.
+Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain, efficient sampling plans, accounting for label switching, and computational efficiency, naming only a few.
 

### 2020-09-13 manuscript/paper_BRM.Rmd
change intro to focus more on generative models

@@
     corresponding : yes
-    address       : "Postal address"
+    address       : "Speelmanstraat 7 1063ZC Amsterdam, The Netherlands"
     email         : "malte_lueken@arcor.de"
@@
 authornote: |
-  This is an author note.
+  We would like to thank Daan van Renswoude and Maartje Raijmakers for their helpful comments on earlier versions of this paper.
+Furthermore, we would like to acknowledge feedback by Karel Veldkamp and Phil Norberts in the early stage of this project.
 
@@
 @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-Gaze data has been frequently classified by probabilistic models.
-Compared to discrete (hard) classification, these models compute the probability of each event for every sample of the raw eye-tracking data.
-They have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
-324].
-One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs).
+Most of these classification algorithms use non-parametric methods to differentiate between eye movement events.
+For instance, the classic "Velocity-Threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stample1993].
+More recently, supervised statistical learning techniques have been successfully used to parse eye movement data.
+For example, @Bellet2019 trained a convolutional neural network on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders.
+However, these supervised methods have several disadvantages: First, they require labeled data (usually labeled by humans) as input.
+Second, they often rely on hyperparameters which are difficult to theoretically justify.
+Third, their parameters are usually difficult to interpret: Most supervised methods operate in a black-box fashion, which implies that they don't make any explicit assumptions about the data generating process.
+Oppositely, several algorithms using unsupervised techniques have been developed.
+They typically don't need to be trained on labeled data and are thus more flexible in classifying data from different individuals, tasks, or eye-trackers.
+Besides differentiating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data.
+To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf.
+@Mihali2017, @Wadehn2019].
+However, they come with various advantages and we will summarize the two most important for eye movement classification here [for an overview, see @Epstein2008].
+First, generative models allow to not only predict (classify) but also explain eye-tracking phenomena.
+For instance, they can inhere parameters that explain why oscillations only occur after but not before saccades.
+When the goal is to understand eye movements and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods.
+Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions.
+For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017; maybe others].
+One class of generative models that are used in eye movement classification are hidden Markov models (HMMs).
 HMMs estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal.
@@
 Each state depends on the previous but not on earlier states of the sequence [@Zucchini2016].
-In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous).
-Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need training labeled data as it would be the case for supervised machine learning.
+In the context of eye movements, HMMs are suitable models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous).
+Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need training labeled data as it would be the case for supervised methods.
 On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a fixed threshold approach (named "Identification by HMM" [I-HMM]).
@@
 Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].
-HMMs have another property that has been rarely used in the context of eye movement classification: They can serve as a generative model for gaze data.
-To our knowledge, only @Mihali2017 used this property by building an explicit generative model for drift and microsaccades.
 For this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms.

### 2020-09-14 manuscript/paper_BRM.Rmd
add general eye movement event classification section to introduction

@@
 @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-Most of these classification algorithms use non-parametric methods to differentiate between eye movement events.
+On one hand, many of these classification algorithms use non-parametric methods to differentiate between eye movement events.
 For instance, the classic "Velocity-Threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stample1993].
-More recently, supervised statistical learning techniques have been successfully used to parse eye movement data.
-For example, @Bellet2019 trained a convolutional neural network on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders.
+On the other hand, many parametric methods have been developed recently.
+Some of them require human labeled training data as input and can therefore be termed as supervised.
+For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders.
 However, these supervised methods have several disadvantages: First, they require labeled data (usually labeled by humans) as input.
@@
 Third, their parameters are usually difficult to interpret: Most supervised methods operate in a black-box fashion, which implies that they don't make any explicit assumptions about the data generating process.
-Oppositely, several algorithms using unsupervised techniques have been developed.
-They typically don't need to be trained on labeled data and are thus more flexible in classifying data from different individuals, tasks, or eye-trackers.
-Besides differentiating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data.
+Beyond that, there are several algorithms using unsupervised techniques.
+They typically don't need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers.
+Besides discriminating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data.
 To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf.
 @Mihali2017, @Wadehn2019].
-However, they come with various advantages and we will summarize the two most important for eye movement classification here [for an overview, see @Epstein2008].
+However, they come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].
 First, generative models allow to not only predict (classify) but also explain eye-tracking phenomena.
@@
 When the goal is to understand eye movements and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods.
-Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions.
-For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017; maybe others].
+Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions [@Epstein2008].
+For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017].
 One class of generative models that are used in eye movement classification are hidden Markov models (HMMs).
@@
 However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012].
-NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].
+NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by the CNN algorithm by @Bellet2019 [@Pekkanen2017].
 In sum, HMMs seem to be a promising method for classifying eye movements. However, the existing HMM algorithms each have their weaknesses that could be improved upon.  
 First, I-HMM relies on setting an appropriate threshold to determine the initial classification, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
-Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability.
-It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals (contrary to the authors, we argue that this is a disadvantage).
+Second, the current implementation of NSLR-HMM requires human-coded data, which narrows its applicability.
+It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals.
 Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017].
@@
 The development was guided by four goals: (a) the algorithm should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it should confirm rather than explore the presence of events in the data.
+The following section will describe gazeHMM and the underlying HMM in detail.
+Then, we will present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding.
+Finally, we will discuss these results and propose directions in which gazeHMM and other HMM algorithms could be improved.
 
 # The Algorithm
+As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most eye movement event classification algorithms consist of three steps: During *preprocessing*, the raw gaze position is transformed into a metric of position and/or time.
+Often, a filtering or smoothing procedure is applied to the data before or after the tranformation to separate the gaze signal from noise and artifacts [@Spakov2012].
+It follows the *classification*, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
+Lastly, as part of the *postprocessing*[^1], the algorithm decides which candidates to accept, relabel, or merge [@Hessels2017; @Komogortsev2010].
+
+[^1]: @Hessels2017 called step two the *search rule* and step three the *classifcation rule*.
+For non-parametric methods, this distinction might be accurate.
+However, for parametric methods, we think calling step two *classification* is more appropriate since the probabilistic classification is done here.
+Step three usually consist of some heuristic relabeling and correcting for classification errors.
+
 ## Preprocessing

### 2020-09-16 manuscript/paper_BRM.Rmd
add gamma distribution justification

@@
 Second, they often rely on hyperparameters which are difficult to theoretically justify.
-Third, their parameters are usually difficult to interpret: Most supervised methods operate in a black-box fashion, which implies that they don't make any explicit assumptions about the data generating process.
+Third, their parameters are usually difficult to interpret: Most supervised methods operate in a black-box fashion, which implies that they do not make any explicit assumptions about the data generating process.
 Beyond that, there are several algorithms using unsupervised techniques.
-They typically don't need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers.
+They typically do not need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers.
 Besides discriminating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data.
@@
 In the context of eye movements, HMMs are suitable models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous).
-Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need training labeled data as it would be the case for supervised methods.
+Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need labeled training data as it would be the case for supervised methods.
 On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a fixed threshold approach (named "Identification by HMM" [I-HMM]).
@@
 Second, the current implementation of NSLR-HMM requires human-coded data, which narrows its applicability.
-It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals.
+It also inheres fixed parameters that prevent the algorithm to adapt to individual- or task-specific signals.
 Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017].
@@
 # The Algorithm
-As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most eye movement event classification algorithms consist of three steps: During *preprocessing*, the raw gaze position is transformed into a metric of position and/or time.
-Often, a filtering or smoothing procedure is applied to the data before or after the tranformation to separate the gaze signal from noise and artifacts [@Spakov2012].
+As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most eye movement event classification algorithms consist of three steps [cf.
+@Hessels2017]: During *preprocessing*, the raw gaze position is transformed into a metric of position and/or time.
+Often, a filtering or smoothing procedure is applied to the data, before or after the tranformation, to separate the gaze signal from noise and artifacts [@Spakov2012].
 It follows the *classification*, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
@@
 $$Y \sim Gamma(\alpha,\beta).$$
-It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.   
+When we developed gazeHMM, the gamma distribution appeared to fit eye movement data well, but we also note that it might not necessarily be the best fitting distribution for every type of eye movement data.
+We assume that the best fitting distribution will depend on the task, eye-tracker, and individual (see also the discussion).
 To model the sample-to-sample angle, we pursued a novel approach in gazeHMM: A mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution: 

### 2020-09-20 manuscript/paper_BRM.Rmd
add some references and remove two-state sim

@@
 library(here)
+library(grid)
+library(gridExtra)
 source(here("algorithm/preprocessing_helper_functions.R"))
@@
 On one hand, many of these classification algorithms use non-parametric methods to differentiate between eye movement events.
-For instance, the classic "Velocity-Threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stample1993].
+For instance, the classic "Velocity-threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stampe1993].
 On the other hand, many parametric methods have been developed recently.
@@
 For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders.
-However, these supervised methods have several disadvantages: First, they require labeled data (usually labeled by humans) as input.
+However, these supervised methods have several disadvantages: First, they require a sufficient amount of labeled data (usually labeled by humans) as input, which can become costly and time consuming [@Zemblys2019].
 Second, they often rely on hyperparameters which are difficult to theoretically justify.
-Third, their parameters are usually difficult to interpret: Most supervised methods operate in a black-box fashion, which implies that they do not make any explicit assumptions about the data generating process.
+Third, most supervised methods operate in a black-box fashion, which implies that they do not make any explicit assumptions about the data generating process [cf.
+@Zembyls2019].
 Beyond that, there are several algorithms using unsupervised techniques.
-They typically do not need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers.
+They typically do not need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers [e.g., @Houpt2018; @Hessels2017].
 Besides discriminating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data.
 To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf.
-@Mihali2017, @Wadehn2019].
+@Mihali2017; @Wadehn2019].
 However, they come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].
@@
 For instance, they can inhere parameters that explain why oscillations only occur after but not before saccades.
-When the goal is to understand eye movements and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods.
+When the goal is to understand eye movement events and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods.
 Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions [@Epstein2008].
@@
 However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012].
-NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by the CNN algorithm by @Bellet2019 [@Pekkanen2017].
+NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by the CNN algorithm in @Bellet2019 [@Pekkanen2017].
 In sum, HMMs seem to be a promising method for classifying eye movements. However, the existing HMM algorithms each have their weaknesses that could be improved upon.  
@@
 For non-parametric methods, this distinction might be accurate.
-However, for parametric methods, we think calling step two *classification* is more appropriate since the probabilistic classification is done here.
+However, for parametric methods, we think calling step two "classification" is more appropriate since the probabilistic classification is done here.
 Step three usually consist of some heuristic relabeling and correcting for classification errors.
+
+(ref:plot-algorithm-workflow) Example work flow for eye movement event classification algorithms: (a) the raw gaze signal in x (dark red) and y (dark blue) coordinates; (b) the raw gaze signal is filtered and transformed into a velocity signal; (c) samples are classified as events (indicated by colors), and (d) relabeled.
+Sequences of samples belonging to the same event are merged (indicated by black segments).
+Data from @Andersson2017.
+
+```{r plot-algorithm-workflow, fig.cap="(ref:plot-algorithm-workflow)", fig.height = 8}
+
+toPlot <- A2017[[3]][[3]] %>%
+  mutate(x.va = px2va(x, dim[1], res[1], dist),
+         y.va = px2va(y, dim[2], res[2], dist),
+         x.vel = sgolayfilt(x.va, m = 1),
+         y.vel = sgolayfilt(y.va, m = 1),
+         vel = sqrt(x.vel^2 + y.vel ^2)*fr) %>%
+  dplyr::filter(t > 4.3 & t < 4.5) %>%
+  mutate(error = vel > 80,
+         event.pre = as.factor(ifelse(error, 2, coderMN)),
+         event.post = as.factor(coderMN))
+
+# Create plot
+
+p1 <- ggplot(toPlot, aes(x = t, x.va)) + 
+  geom_line(aes(y = x.va)) + 
+  geom_line(aes(y = y.va)) + 
+  geom_point(aes(y = x.va), color = "darkred", size = 0.9) + 
+  geom_point(aes(y = y.va), color = "darkblue", size = 0.9) +
+  scale_y_continuous(name = "Gaze position (deg)", limits = c(-5, 10), breaks = pretty(c(-5, 10)))
+
+p2 <- ggplot(toPlot, aes(x = t, y = vel)) + 
+  geom_line() + 
+  geom_point(size = 0.9) +
+  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
+
+p3 <- ggplot(toPlot, aes(x = t, y = vel)) + 
+  geom_line() + 
+  geom_point(aes(color = event.pre), size = 0.9) +
+  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
+
+p4 <- ggplot(toPlot, aes(x = t, y = vel)) + 
+  geom_line() + 
+  geom_point(aes(color = event.post), size = 0.9) +
+  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
+
+plotList <- list(p1, p2, p3, p4)
+
+for(p in 1:length(plotList)) {
+
+  plotList[[p]] <- plotList[[p]] + 
+    xlab("") + geom_line(color = "black") + ggtitle(letters[p]) + 
+    theme(legend.position = "none")
+
+}
+
+plotList[[4]] <- plotList[[4]] + xlab("Time (s)") +
+  geom_segment(aes(x = first(t[event.post == 4]), y = 250, xend = first(t[event.post == 2])-(1/fr), yend = 250), size = 1.2) +
+  geom_segment(aes(x = first(t[event.post == 2]), y = 250, xend = first(t[event.post == 3])-(1/fr), yend = 250), size = 1.2) +
+  geom_segment(aes(x = first(t[event.post == 3]), y = 250, xend = last(t[event.post == 3]), yend = 250), size = 1.2) +
+  geom_segment(aes(x = last(t[event.post == 3])+(1/fr), y = 250, xend = last(t[event.post == 4]), yend = 250), size = 1.2)
+
+grid.arrange(plotList[[1]], plotList[[2]], plotList[[3]], plotList[[4]], ncol = 1)
+
+```
 
@@
 Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
-Due to tremor, we assume that the angle between samples should not follow any direction but a random walk [@Duchowski2017; @Larsson2015].
+Due to tremor, we assume that the angle between samples should not follow any direction but a uniformly random walk [@Larsson2015].
 In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction.
@@
 When we developed gazeHMM, the gamma distribution appeared to fit eye movement data well, but we also note that it might not necessarily be the best fitting distribution for every type of eye movement data.
-We assume that the best fitting distribution will depend on the task, eye-tracker, and individual (see also the discussion).
+We assume that the best fitting distribution will depend on the task, eye-tracker, and individual (see discussion).
 To model the sample-to-sample angle, we pursued a novel approach in gazeHMM: A mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution: 
@@
 Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions.
-Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution.
+Because we assume fixations to change their direction similar to a uniformly random walk [@Larsson2013; @Larsson2015], their sample-to-sample angle can be modeled by a uniform distribution.
 Thus, the uniform distribution should distinguish fixations from the other events.
@@
 After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence.
-We implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
+We implemented this routine because constraining the probabilities for nonsaccade events to turn into PSOs to zero often caused PSOs not to appear in the state sequence at all.
 Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events.
 Thus, the postprocessing routine heuristically compensates for such violations.
-This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events.
+This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow nonsaccade events.
 Samples are relabeled as the state of the previous event.
@@
 ### Parameter Variation
-In the first part of the simulation, we examined how varying the parameters in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
-Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM.
-The RMdSPD was below 0.1 for all estimated and manipulated parameters, indicating good recovery [^2].
-The regression lines between manipulated true and estimated parameters are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
-With one outlier at parameter $\alpha_{acc;1}$, the estimated parameters matched the true parameters very closely.
-Thus, parameter change was captured well and the estimation almost unbiased.
-Considering accuracy, Figure \@ref(fig:plot-acc-1-2) displays Cohen's kappa between true and estimated hidden state sequences.
-With two exceptions, kappa values were almost one, suggesting nearly perfect agreement.
-
-[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
-Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
-Therefore, we decided not to include it in the analysis.
-
-(ref:plot-rmdspd-1-2) RMdSPD between true and estimated parameters of the two-state HMM in part one of the simulation.
-Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
-Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
-$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
-
-```{r plot-rmdspd-1-2, fig.cap="(ref:plot-rmdspd-1-2)"}
-
-print(plots.rmsd.1[[1]])
-
-```
-
-(ref:plot-regtr-1-2) Regression lines between true and estimated transition probabilities for the two-state HMM in part one.
-Right facet labels show *from* and top facet labels show *to* which state the HMM is moving.
-Dashed lines refer to perfect recovery.
-
-```{r plot-regtr-1-2, fig.cap="(ref:plot-regtr-1-2)"}
-
-print(plots.lm.tr[[1]])
-
-```
-
-(ref:plot-regresp-1-2) Regression lines between true and estimated response parameters of the two-state HMM in part one.
-Top facet labels indicate response parameters.
-Dashed lines refer to perfect recovery.
-
-```{r plot-regresp-1-2, fig.cap="(ref:plot-regresp-1-2)"}
-
-print(plots.lm.resp[[1]] + theme(axis.text = element_text(size = 10)))
-
-```
-
-(ref:plot-acc-1-2) Boxplots displaying Cohen's kappa depending on which parameter of the two-state HMM has been manipulated in part one.
-Top facet labels indicate for which state parameters have been manipulated.
-Black solid lines symbolize medians.
-Crosses represent outliers (distance to first/third quartile higher than 1.5 times the inter-quartile range [IQR]).
-
-```{r plot-acc-1-2, fig.cap="(ref:plot-acc-1-2)"}
-
-print(plots.acc.1[[1]])
-
-```
-
+In the first part of the simulation, we examined how varying the parameters[^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
+For the two-state HMM, the recovery of parameters and states was nearly perfect (all RMdSPDs < 0.1, intercepts and slopes of regression lines almost zero and one, respectively, and Cohen's kappa close to 1).
+We therefore chose to include the respective figures in the supplementary material.
 For the HMM with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
@@
 However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched (see supplementary material).
+[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
+Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
+Therefore, we decided not to include it in the analysis.
 
@@
 
-A2017.kappa[3, -c(1, seq(2, 13, 3))] <- 0
-A2017.kappa[4, seq(2, 13, 3)] <- 0
+A2017.kappa[3, -c(1, seq(2, 13, 3))] <- NA
+A2017.kappa[4, seq(2, 13, 3)] <- NA
 
@@
 However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yielded different results.
-The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events (incl.
-blinks).
+The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events [incl.
+blinks; see also the *accuracy paradox*, @Zemblys2019].
 Nevertheless, Cohen's kappa values of 0.67 (fixations - image) or 0.62 (saccades - moving dots) indicate substantial agreement to human coders, especially in light of the maximum references from the simulation study.

### 2020-10-02 manuscript/paper_BRM.Rmd
Fix minor issues

@@
 To adress these weaknesses, we developed gazeHMM, an algorithm that uses an HMM as a generative model, has no critical parameters to be set by users, and does not require human coded data as input.
-The algorithm classifies gaze data into fixations, saccades, and optionally PSOs and smooth pursuits.
+The algorithm classifies gaze data into fixations, saccades, and optionally postsaccadic oscillations and smooth pursuits.
 We evaluated the performance of gazeHMM in a simulation study and on two benchmark data sets.
@@
 For dynamic stimuli, our algorithm tended to switch rapidly between fixations and smooth pursuits.
-Thus, we recommend to use gazeHMM for classifiying smooth pursuits only in explorative settings.
+Thus, we recommend using gazeHMM for classifiying smooth pursuits only in explorative settings.
 Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.
@@
 @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-On one hand, many of these classification algorithms use non-parametric methods to differentiate between eye movement events.
+To motivate our decision to add another algorithm to this already rich selection of classification tools, we first need to distinguish between the different classes of methods they use.
+On one hand, many classification algorithms use non-parametric methods to differentiate between eye movement events.
 For instance, the classic "Velocity-threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stampe1993].
@@
 For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders.
-However, these supervised methods have several disadvantages: First, they require a sufficient amount of labeled data (usually labeled by humans) as input, which can become costly and time consuming [@Zemblys2019].
+Despite their accuracy, supervised methods have several disadvantages: First, they require a sufficient amount of labeled data (usually labeled by humans) as input, which can become costly and time consuming [@Zemblys2019].
 Second, they often rely on hyperparameters which are difficult to theoretically justify.
 Third, most supervised methods operate in a black-box fashion, which implies that they do not make any explicit assumptions about the data generating process [cf.
-@Zembyls2019].
+@Zemblys2019].
 Beyond that, there are several algorithms using unsupervised techniques.
@@
 To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf.
-@Mihali2017; @Wadehn2019].
-However, they come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].
-First, generative models allow to not only predict (classify) but also explain eye-tracking phenomena.
-For instance, they can inhere parameters that explain why oscillations only occur after but not before saccades.
+@Mihali2017; @Wadehn2020].
+They come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].
+First, generative models allow to not only predict (classify) but also help to explain eye-tracking phenomena.
+For instance, they can inhere parameters that indicate that oscillations only occur after but not before saccades.
 When the goal is to understand eye movement events and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods.
@@
 NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by the CNN algorithm in @Bellet2019 [@Pekkanen2017].
-In sum, HMMs seem to be a promising method for classifying eye movements. However, the existing HMM algorithms each have their weaknesses that could be improved upon.  
+In sum, HMMs seem to be a promising method for classifying eye movements. Nevertheless, the existing HMM algorithms each have their weaknesses that could be improved upon.  
 First, I-HMM relies on setting an appropriate threshold to determine the initial classification, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
@@
 For this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms.
-The development was guided by four goals: (a) the algorithm should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it should confirm rather than explore the presence of events in the data.
-The following section will describe gazeHMM and the underlying HMM in detail.
+The development was guided by four goals: (a) the algorithm should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuits, and PSOs; (d) it should confirm rather than explore the presence of events in the data.
+The following section will describe gazeHMM and the underlying generative model in detail.
 Then, we will present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding.
@@
   geom_line(aes(y = y.va)) + 
-  geom_point(aes(y = x.va), color = "darkred", size = 0.9) + 
-  geom_point(aes(y = y.va), color = "darkblue", size = 0.9) +
+  geom_point(aes(y = x.va), color = "darkred", size = 1) + 
+  geom_point(aes(y = y.va), color = "darkblue", size = 1) +
   scale_y_continuous(name = "Gaze position (deg)", limits = c(-5, 10), breaks = pretty(c(-5, 10)))
@@
   geom_line() + 
-  geom_point(size = 0.9) +
+  geom_point(size = 1) +
   scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
@@
   geom_line() + 
-  geom_point(aes(color = event.pre), size = 0.9) +
+  geom_point(aes(color = event.pre), size = 1) +
   scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
@@
   geom_line() + 
-  geom_point(aes(color = event.post), size = 0.9) +
+  geom_point(aes(color = event.post), size = 1) +
   scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
@@
 Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events.
-Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017].
-However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007].
+Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017], but most of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007].
 This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017].
 In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.
-Theoretically, these three metrics should separate eye movement events clearly.
+Theoretically, these three metrics should clearly separate eye movement events.
 Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
@@
 The EM algorithm is generally suitable to estimate likelihoods with missing variables.
-For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [eye movement events; @Visser2011].
+For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [i.e., eye movement events; @Visser2011].
 When evaluating the likelihood for missing data, gazeHMM integrates over all possible values, which results in a probability density of one.
@@
 Thus, the postprocessing routine heuristically compensates for such violations.
-This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow nonsaccade events.
+This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold (here: 10 ms), and PSOs that follow nonsaccade events.
 Samples are relabeled as the state of the previous event.
@@
 ### Parameter Variation
-In the first part of the simulation, we examined how varying the parameters[^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
+In the first part of the simulation, we examined how varying the parameters [^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
 For the two-state HMM, the recovery of parameters and states was nearly perfect (all RMdSPDs < 0.1, intercepts and slopes of regression lines almost zero and one, respectively, and Cohen's kappa close to 1).
@@
 Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
-Therefore, we decided not to include it in the analysis.
+Therefore, we decided to exclude it from the analysis.
 
@@
 
-Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for all three conditions.
+Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for the dynamic conditions.
 For the static condition, the difference to most other algorithms was only slight ($\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$%), but for the dynamic conditions, the differences were substantial (video: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$%) and large (dots: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$%).
@@
 An HMM serves as the generative model of the algorithm and classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
-We showed in a simulation study that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions.
-Adding a fourth event (i.e., smooth pursuit) to the model and noise to the generated data were the most critical factors for decreasing recovery.
+We showed in a simulation study that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions: Adding a fourth event (i.e., smooth pursuit) to the model and noise to the generated data were the most critical factors for decreasing recovery.
 Furthermore, we applied gazeHMM with different numbers of states to benchmark data by @Andersson2017 and compared the model fit.
@@
 When comparing gazeHMM against other algorithms, gazeHMM showed mostly good agreement to human coding.
-On one hand, it outperformed the other algorithms in the overall disagreement to human coding.
+On one hand, it outperformed the other algorithms in the overall disagreement to human coding for dynamic data.
 On the other hand, gazeHMM confused a lot of fixations with smooth pursuits, which led to a rapid switching between the two events.
@@
 @Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models.
-Both situations could apply to gazeHMM and eye movement data.
-Outliers occur frequently in eye-tracking data due to measurement error.
+Both situations could apply to gazeHMM and eye movement data: Outliers occur frequently in eye-tracking data due to measurement error.
 Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task from which the data are obtained from [@Langrock2015].
@@
 Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events.
-For smooth pursuits, the classification is not optimal and thus not recommended, yet.
+For smooth pursuits, the classification is not optimal and thus not yet recommended.
 On one side, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one.

### 2020-10-03 manuscript/paper_BRM.Rmd
apply minor changes intro

@@
 @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-To motivate our decision to add another algorithm to this already rich selection of classification tools, we first need to distinguish between the different classes of methods they use.
+To motivate our decision to add another algorithm to this array of classification tools, we first need to distinguish between the different classes of methods they use.
 On one hand, many classification algorithms use non-parametric methods to differentiate between eye movement events.
-For instance, the classic "Velocity-threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stampe1993].
+A classic example is the "Velocity-threshold" algorithm [@Stampe1993], which classifies samples with a velocity above a fixed threshold as saccades [see also @Nystrom2010; @Larsson2013; @Larsson2015].
 On the other hand, many parametric methods have been developed recently.
-Some of them require human labeled training data as input and can therefore be termed as supervised.
-For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders.
-Despite their accuracy, supervised methods have several disadvantages: First, they require a sufficient amount of labeled data (usually labeled by humans) as input, which can become costly and time consuming [@Zemblys2019].
-Second, they often rely on hyperparameters which are difficult to theoretically justify.
-Third, most supervised methods operate in a black-box fashion, which implies that they do not make any explicit assumptions about the data generating process [cf.
-@Zemblys2019].
-Beyond that, there are several algorithms using unsupervised techniques.
-They typically do not need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers [e.g., @Houpt2018; @Hessels2017].
+Some of them require human labeled training data as input and can therefore be termed as *supervised* [@Hastie2017].
+For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders [see also @Zemblys2018; @Zemblys2019; @Startsev2019].
+Despite their accuracy, the requirement of labeled training data is a disadvantage of supervised methods because the labeling process can easily become costly and time consuming [@Zemblys2019].
+In contrast, *unsupervised* classification algorithms do not require labeled training input.
+Instead they learn parameters from the characteristics of the data themselves [@Hastie2017].
+In consequence, they are also more flexible in classifying data from different individuals, tasks, or eye-trackers [e.g., @Houpt2018; @Hessels2017].
 Besides discriminating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data.
@@
 @Mihali2017; @Wadehn2020].
-They come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].
-First, generative models allow to not only predict (classify) but also help to explain eye-tracking phenomena.
-For instance, they can inhere parameters that indicate that oscillations only occur after but not before saccades.
+Classifiers with generative assumptions have the advantage that their parameters can be easily interpreted in terms of the underlying theory.
+In the context of eye movements, they can also help explain or confirm observed phenomena: For instance, their parameters can indicate that oscillations only occur after but not before saccades.
 When the goal is to understand eye movement events and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods.
-Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions [@Epstein2008].
-For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017].
-One class of generative models that are used in eye movement classification are hidden Markov models (HMMs).
-HMMs estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal.
+Moreover, generative models can challenge common theoretical assumptions and bring up new research questions [@Epstein2008].
+For example, they might suggest that oscillations also occur before saccadic eye movements [as mentioned in @Nystrom2010] or that the assumption of discrete eye movements does not hold [as discussed in @Andersson2017].
+One class of generative models that are used in eye movement classification are HMMs.
+They estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal.
 Each gaze sample depends on its corresponding state.
 Each state depends on the previous but not on earlier states of the sequence [@Zucchini2016].
-In the context of eye movements, HMMs are suitable models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous).
-Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need labeled training data as it would be the case for supervised methods.
+Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need labeled training data.
+They are suitable models for eye movement classification because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous).
 On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a fixed threshold approach (named "Identification by HMM" [I-HMM]).
@@
 Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees.
-Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
+Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades, they follow the same direction [@Larsson2013; @Leigh2015].
 Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.
@@
 ### Parameter Variation
-In the first part of the simulation, we examined how varying the parameters [^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
+In the first part of the simulation, we examined how varying the parameters[^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
 For the two-state HMM, the recovery of parameters and states was nearly perfect (all RMdSPDs < 0.1, intercepts and slopes of regression lines almost zero and one, respectively, and Cohen's kappa close to 1).
@@
 However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched (see supplementary material).
+
 [^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.

### 2020-10-04 manuscript/paper_BRM.Rmd
add supplementary materials

@@
 
-bibliography      : "report.bib"
+bibliography      : "references.bib"
 
@@
 Colours indicate different sizes of generated data.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
 Top facet labels denote to which state estimated parameters belong.
@@
 
-print(plots.rmsd.234[[1]][[1]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[1]][[1]] + coord_flip())
 
@@
 (ref:plot-rmdspd-2-3) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
 Top facet labels denote to which state estimated parameters belong.
@@
 
-print(plots.rmsd.234[[1]][[2]] + theme(axis.text = element_text(size = 6)))
+print(plots.rmsd.234[[1]][[2]] + coord_flip())
 
@@
 (ref:plot-rmdspd-2-4) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
 Top facet labels denote to which state estimated parameters belong.
@@
 
-print(plots.rmsd.234[[1]][[3]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[1]][[3]] + coord_flip())
 
@@
 
-# Compute Schwarz weights
+# Compute Schwarz weights (from Wagenmakers & Farrell, 2004)
 

### 2020-10-04 manuscript/supplementary_material_simulation.Rmd
add supplementary materials

--- 2020-06-29 manuscript/internship_report.Rmd
@@
 ---
-title             : "Keeping an Eye on Hidden Markov Models in Gaze Data Classification"
+title             : "Keeping an Eye on Hidden Markov Models in Gaze Data Classification: Supplementary Material Simulation Study"
 shorttitle        : "HMMs in Gaze Classification"
 
-author: 
-  - name          : "Malte Valentin Lüken"
-
-authornote: |
-  Affiliation: University of Amsterdam, Graduate School of Psychology, Department of Methodology and Statistics.
-
-  This manuscript constitutes the internship report as part of the Research Master's Psychology.
-
-  Student number: 12750166. Email address: malte.luken@student.uva.nl.
-
-  Specialization: Methodology and Statistics. Supervisor: dhr. dr. Ingmar Visser.
-
-  I would like to thank Simon Kucharsky and Ingmar Visser for their helpful advice on developing and implementing this research project.
-
-abstract: |
-  Eye-tracking allows researchers to infer cognitive processes from eye movements that are classified into distinct events.
-Parsing the events is typically done by algorithms that transform, filter, classify, and merge raw data samples.
-Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere weaknesses.
-Therefore, I developed gazeHMM, an HMM algorithm that has no critical parameters to be set by users, does not require human coded data as input, and classifies fixations, saccades, PSOs, and smooth pursuits.
-The development was guided by the question of whether HMMs are useful at describing eye movements and whether they improve the event classification.
-I evaluated gazeHMM in a simulation study and on benchmark data.
-The simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences.
-Critical exceptions for good recovery were adding a smooth pursuit like state to gazeHMM and noisy data.
-For benchmark data, model comparisons with gazeHMM yielded preferred models with more states than expected.
-I assessed the classification performance of gazeHMM compared to other algorithms by the agreement to human event coding.
-Here, gazeHMM improved the event classification but did not outperform all other algorithms in most cases.
-Both the simulation study and benchmark application showed poor classification performance for smooth pursuits.
-Thus, I advice to classify smooth pursuits only for exploration and recommend gazeHMM with fixations, saccades, and PSOs for application.
-Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.
-
-keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
-wordcount         : "12962"
-
-bibliography      : "report.bib"
-appendix          : 
- - "report_appendix.Rmd"
- - "report_technical_appendix.Rmd"
-
-floatsintext      : yes
-figurelist        : yes
-tablelist         : yes
+floatsintext      : no
+figurelist        : no
+tablelist         : no
 footnotelist      : no
@@
 
+bibliography      : "references.bib"
 documentclass     : "apa6"
@@
 
-# Introduction
-For how long did humans seek to open a window into each other's minds, revealing their hidden cognitive processes?
-While many methods aiming to achieve this have been developed in the last decades, eye-tracking as one of them has become more and more reliable, accurate, and widely applied in cognitive psychology [@Duchowski2017].
-Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017].
-Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015].
-Commonly distinguished events are fixations, saccades, smooth pursuit, and post-saccadic oscillations (PSOs).
-Classifying eye-tracking data reduces their complexity and is the first step towards cognitive interpretation [@Andersson2017; @Salvucci2000].
-The classification is normally done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017].
-@Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].
-As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most classification algorithms consist of four steps: First, the raw gaze position is transformed into a metric of position and/or time.
-Second, a filtering or smoothing procedure is applied to the data to separate the gaze signal from noise and artifacts [@Spakov2012].
-Third, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
-Fourth, the algorithm decides which candidates to accept, relabel, or merge [@Hessels2017; @Komogortsev2010].
-A frequently implemented method uses probabilistic models to classify events.
-Instead of discrete classification, these models assign probabilities for belonging to an event to each sample.
-Probabilistic models have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p.
-324].
-One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs).
-Figure \@ref(fig:plot-hmm-structure) illustrates the structure of HMMs: Parallel to the gaze signal evolves a sequence of distinct states that cannot be directly observed.
-Each gaze sample depends on its corresponding state.
-Each state depends on the previous but not on earlier states of the sequence [@Cappe2005; @Visser2019].
-In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as events and gaze data are dependent time series.
-On this basis, several classification algorithms using HMMs have been developed.
-One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]).
-Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM.
-Recently, @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR).
-The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and angle (named NSLR-HMM).
-Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]).
-Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).
-Several studies have tested the performance of HMM algorithms against other classification methods: I-HMM has been deemed as robust against noise, behaviorally accurate, and showing a high sample-to-sample agreement with a human coder [@Andersson2017; @Komogortsev2010; @Salvucci2000].
-However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012].
-NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].
-In sum, HMMs seem to be a promising method for classifying eye movements.
-Still, the existing HMM algorithms each have their weaknesses that could be improved upon (and that could explain their inferior performance in comparison to other methods).
-First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
-Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability.
-It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals.
-Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017].
-The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events.
-Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].
-HMMs can also be used as a generative model to make inferences about the underlying processes of eye movements, a property that has been rarely used in the context of classification [cf.
-@Mihali2017].
-For example, an HMM with four events could be compared to one with three events to see whether an event is present in the data.
-I propose to develop an algorithm that uses HMMs to classify eye movement data. It will address the weaknesses of previous HMM algorithms and utilize the generative property of HMMs.
-
-```{r setup, include=FALSE}
+```{r packages, include=FALSE}
 
@@
 library(tidyverse)
-library(signal)
-library(flexcircmix)
-library(CircStats)
-library(depmixS4)
+library(here)
 library(psych)
-library(here)
-library(grid)
-library(gridExtra)
-library(DiagrammeR)
-library(DiagrammeRsvg)
-library(magrittr)
-library(rsvg)
-source(here("algorithm/preprocessing_helper_functions.R"))
-source(here("algorithm/model_helper_functions.R"))
-
 ```
@@
 
-(ref:plot-algorithm-workflow) Example workflow for eye movement event classification algorithms: (a) the raw gaze signal is transformed into a velocity signal (in deg/s); (b) the velocity signal is smoothed and filtered; (c) samples are labeled as events (indicated by colors), and (d) relabeled.
-Sequences of samples belonging to the same event are merged (indicated by black segments).
-
-```{r plot-algorithm-workflow, fig.cap="(ref:plot-algorithm-workflow)", fig.height = 8}
-
-set.seed(126)
-
-velRaw <- sample(c(rgamma(25, 3, 1/6), rgamma(7, 25, 1/5), rgamma(8, 15, 1/5)), 40)
-
-sample <- 1:length(velRaw)
-
-velSmooth <- as.numeric(smooth(velRaw, kind = "3R"))
-
-eventPre <- ifelse(velSmooth > 40, "s", "f")
-
-eventPost <- ifelse(velSmooth < 60, "f", eventPre)
-
-df <- data.frame(sample, velRaw, velSmooth, eventPre, eventPost)
-
-# Create plot
-
-p1 <- ggplot(df, aes(sample, velRaw)) + geom_point()
-
-p2 <- ggplot(df, aes(sample, velSmooth)) + geom_point()
-
-p3 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPre))
-
-p4 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPost))
-
-plotList <- list(p1, p2, p3, p4)
-
-for(p in 1:length(plotList)) {
-
-  plotList[[p]] <- plotList[[p]] + scale_y_continuous(name = "Velocity\n(deg/s)", 
-                                                      limits = c(0, 200), breaks = c(0, 100, 200)) + 
-    xlab("") + geom_line(color = "grey") + ggtitle(letters[p]) + 
-    theme(legend.position = "none")
-
-}
-
-plotList[[4]] <- plotList[[4]] + xlab("Sample") +
-  geom_segment(aes(x = 1, y = 180, xend = 7, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 8, y = 180, xend = 16, yend = 180), size = 1.2) +
-  geom_segment(aes(x = 17, y = 180, xend = 40, yend = 180), size = 1.2)
-
-grid.arrange(plotList[[1]], plotList[[2]], plotList[[3]], plotList[[4]], ncol = 1)
-
-```
-
-(ref:plot-hmm-structure) Example structure of HMMs: The observed velocity signal *V* depends on hidden states *S* (displayed for six samples).
-Each state depends on the previous state.
-Colors indicate different states.
-Figure inspired by Visser and Speekenbrink [-@Visser2019, p.
-133].
-
-```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)", fig.height=2, fig.width=4}
-
-knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 400)
-
-```
-
-## Research Question
-Can HMMs improve describing and classifying eye-tracking data?
-This is the general research question of my proposal and can be specified as following: Are HMMs useful generative models for recorded eye movements that improve classifying them into events?
-This research question will guide the development of the algorithm and leads to two hypotheses: First, I predict that HMMs are useful generative models for eye movements.
-This hypothesis will be supported if the algorithm yields events that match theoretical expectations for benchmark data.
-Other results will be treated as evidence against the hypothesis.
-Second, I hypothesize that HMMs will improve the classification performance of eye movements.
-I will treat it as support for this hypothesis if the algorithm outperforms algorithms using other methods when applied to benchmark data.
-I will treat other results as evidence against the hypothesis.
-
-# Operationalization
-The proposed algorithm aims to improve the drawbacks of previous HMM algorithms.
-Therefore, its classification method is oriented to four major goals: (a) it does not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it covers the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it confirms rather than explores the presence of events in the data.
-Besides implementing the algorithm in the statistical programming language R [@RCoreTeam2020], I will develop its concrete technical design exploratorily.
-The first step will be choosing an appropriate metric that allows separating the gaze data into events.
-@Zemblys2018 provide an overview of different gaze data metrics and compare how well they predict human-labeled events.
-The next step will choosing a filtering/smoothing method to reduce the noise in the gaze data [for a comparison of different gaze data filters, see @Spakov2012].
-The criteria for choosing a metric and a filter will be its accuracy and not relying on user parameter settings.
-The generative part of the algorithm will allow users to build an HMM which matches their theoretical assumptions about the gaze data.
-The algorithm will take a preprocessed gaze data metric (e.g., filtered velocity) as input and optimize the model parameters.
-The output of the model will be probabilities for hidden states.
-The states can be interpreted as event labels for each sample.
-For instance, the model could assign to a sample a probability of 0.6 for belonging to a fixation and 0.4 for belonging to a saccade.
-These probabilities can be used to compute the hidden state sequence (e.g., by choosing events with the highest probability).
-Moreover, the model will output the parameters of the distribution of labeled samples for each event.
-It will allow generating samples from these distributions to simulate gaze data.
-Before testing the hypotheses, I will conduct a simulation study to estimate the parameter recovery of the generative model: Artificial eye movement data will be generated by the model.
-The parameters of the model will be varied across sensible ranges.
-Then, the model will be applied to the data.
-Finally, the estimated results of the model will be compared to the true parameters that have been used to generate the data.
-The first hypothesis will be tested by comparing different generative models on benchmark data.
-The models will differ in their assumed number of events in the data.
-The model comparison will result in a preferred model which scores best on the evaluation criterion.
-The preferred model is expected to match the theoretically assumed number of present events in the benchmark data.
-To test the second hypothesis, the proposed algorithm will be compared against other algorithms that have also been applied to benchmark data.
-The performance of the proposed algorithm will be assessed by metrics that are commonly used in classification (e.g., Cohen's Kappa).
-
-## Sample Characteristics
-The hypotheses will be tested on two benchmark data sets that have been published in previous studies.
-The first set was published in @Andersson2017 and contains eye-tracking data from 17 subjects.
-Three types of stimuli were presented on a screen ($1024 \times 768$ pixels) to each subject: A static image, a linearly moving dot, and a video.
-Subjects were instructed to look freely at the images and follow the movements of the dots and the objects in the videos.
-Their eye movements were recorded with a Hi-Speed 1250 tower-mounted eye-tracker (SensoMotoric Instruments) and a sampling rate of 500 Hz.
-Only data from the right eye were used.
-The data were labeled by two human experts in eye movement research as fixations, saccades, smooth pursuits, PSOs, or blinks [@Andersson2017, p.
-621].
-The second data set, published by @Ehinger2019, consists of eye-tracking data from 15 subjects who performed six blocks of 10 different experimental tasks.
-Here, I will only consider data from Tasks 4 and 5 because they are qualitatively different from the first data set.
-In Task 4, subjects were instructed to fixate a central target on the screen ($1920 \times 1080$ pixels) for 20 s.
-In Task 5, subjects also fixated a central target but were instructed to blink each time they heard a beep.
-Seven beeps with a duration of 100 ms and a 1.5 s interval in between were presented.
-For both tasks, eye-movements were recorded by two eye-trackers simultaneously (EyeLink 1000, SR Research; and Pupil Labs glasses, Pupil Labs) with a sampling rate of 250-500 Hz and 240 Hz (interpolated), respectively [@Ehinger2019, pp.
-3-11].
-
-## Data Analysis
-I will treat the parameter recovery of the generative model as satisfactory when it yields a 95% CI covering the true parameters.
-The median rate of correct classifications should be a minimum of 90%.
-The first hypothesis will be tested using a model comparison approach: For both data sets, models with one, two, three, four, and five eye movement events plus a noise event will be compared.
-The noise event will represent blinks.
-For the first data set and static images, I expect the model with three events (excl.
-noise) to be the preferred model since smooth pursuit movements are not assumed to be present.
-For moving dots and video stimuli, I expect the model with four events (excl.
-noise) to be preferred, since smooth pursuits are expected in the data.
-For the second data set, I expect a model with one event (excl.
-noise) to be preferred because only fixations should be present in the data from both tasks.
-For all comparisons, the preferred model will be chosen according to the highest Schwarz weight [@Wagenmakers2004].
-I will test the second hypothesis using the metrics that have been used in the original study of the first data set: @Andersson2017 compared ten algorithms to human-labeled data by the root mean squared error (RMSE) of event distribution descriptives, the sample-to-sample agreement indicated by Cohen's Kappa, and the disagreement ratio across all samples indicated by the confusion matrix.
-I expect the algorithm with a generative model assuming four events (excl.
-noise) to achieve either a lower RMSE, higher Cohen's Kappa, or lower disagreement ratio than all the other algorithms which are evaluated in the study.
-
-## Intended Results
-Preferred generative models that match the theoretical assumptions on the number of events in the data would support the notion that HMMs describe gaze data appropriately.
-If a generative model with fewer events than expected is preferred, this would mean that the model is not able to distinguish the events well enough.
-If more events than expected are preferred, this would indicate that the model is describing noise or an unexpected event is present in the data.
-Outperforming other algorithms in agreement to human-coded data would demonstrate that the proposed algorithm classifies events similar to human coders.
-However, @Hooge2018a observed considerable differences in how humans code gaze data.
-They concluded that human coders cannot be seen as a gold standard for event classification.
-The proposed algorithm would be the first allowing the user to conduct model comparisons as part of the classification.
-This comparison could be used to investigate the data more thoroughly on a qualitative (e.g., assumed present events) or quantitative (e.g., event sample distributions) level.
-In contrast to many other classification algorithms, the proposed algorithm would not only classify gaze data but also explain why it is classified this way by using a generative model.
-This outstanding property of the algorithm would guarantee its competitiveness against other popular algorithms, especially the recently developed machine learning approaches which do not make generative assumptions [e.g., @Bellet2019].
-It could spark a shift from pure classification output towards understanding eye movements on a more fundamental level.
-
-```{r load data, include=FALSE}
-
-load(here("validation/Andersson2017_raw.Rdata"))
-
-res <- c(1024, 768)
-dim <- c(380, 300)
-dist <- 670
-fr <- 500
-
-```
-
-# Algorithm Development
-As part of answering my research questions, I developed an algorithm named gazeHMM to classify gaze data into discrete eye movement events.
-The following section describes the development of gazeHMM and its final version that has been used to obtain results.
-Technical details can be found in Appendix B.
-
-## Eye Movement Metrics
-Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018].
-Here, my goal was to find those metrics that separate the gaze data samples belonging to different events the best.
-However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018].
-This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics.
-Because they belong to the most basic metrics which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] in gazeHMM.
-Theoretically, these three metrics should separate eye movement events clearly.
-Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
-Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017].
-In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction.
-PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013].
-They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
-Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes.
-Those samples in between typically follow the same direction.
-Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees.
-Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].
-Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as proposed earlier, they are not in the scope of gazeHMM.
-
-## Filtering and Smoothing
-As with metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics.
-Their purpose is to remove noise or artifacts from the gaze data that could distort the classification [@Duchowski2017].
-Filtering (smoothing) methods that are applied before computing the eye movement metrics target the recorded gaze position.
-While removing noise, filters might also erase differences in the metrics between samples.
-For instance, tremor movements during fixations could be filtered out and in consequence, the samples would follow the same direction.
-This could complicate separating fixations from smooth pursuits.
-The same problem applies to PSOs and saccades where filtering could erase oscillations (see Figure \@ref(fig:filter-angle)).
-
-(ref:filter-angle) Example data from @Andersson2017 displayed as sample-to-sample angle (in rad) over time (in s).
-Line colors indicate whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue).
-For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3.
-The full line segment marks where potential oscillations are filtered out.
-The dashed line segment illustrates where potentially random change in angle is filtered out.
-
-```{r filter-angle, fig.cap="(ref:filter-angle)"}
-
-filter.ex <- A2017$video[[3]] %>%
-  mutate(x = px2va(x, dim[1], res[1], dist),
-         y = px2va(y, dim[2], res[2], dist),
-         angle = calc_theta(x, y),
-         x.filt = as.numeric(signal::filter(signal::butter(3, 0.3), x)),
-         y.filt = as.numeric(signal::filter(signal::butter(3, 0.3), y)),
-         angle.filt = calc_theta(x.filt, y.filt),
-         t = (t-t[1])/1e6) %>%
-  dplyr::filter(t > 0.5, t < 0.75)
-
-ggplot(filter.ex) + 
-  geom_path(aes(x = t, y = force_neg_pi_pi(angle)), color = "blue") +
-  geom_path(aes(x = t, y = force_neg_pi_pi(angle.filt)), color = "red") +
-  annotate(geom = "segment", x = 0.525, xend = 0.54, y = -3, yend = -3, color = "black", size = 0.8) +
-  annotate(geom = "segment", x = 0.635, xend = 0.67, y = -3, yend = -3, color = "black", size = 0.8, linetype = 6) +
-  scale_y_circular(name = "Sample-to-sample angle (in rad)", units = "radians", limits = c(-pi, pi)) +
-  scale_x_continuous(name = "Time (in s)")
-
-```
-
-Instead of filtering before computing the metrics, I decided to combine both in one step.
-Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration).
-@Houpt2018 computed the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm.
-Similarly, @Nystrom2010 used a Savitzky-Golay (SG) filter [@Savitzky1964] to estimate velocity and acceleration from a gaze signal.
-Figure \@ref(fig:filter-comparison) displays an example of velocity and acceleration signals obtained by both methods.
-They follow a similar trend, but the SG velocity signal was less noisy and the acceleration signal was consequently lower than the Laplace acceleration signal.
-Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events.
-To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle as the backward difference and the first discrete derivative as the forward difference (see Appendix B).
-
-(ref:filter-comparison) Example data from @Andersson2017 displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s).
-Line colors indicate which filter has been used to derive the signal.
-Savitzky-Golay filters were applied with order three and length five (corresponding to 10 ms).
-
-```{r filter-comparison, fig.cap="(ref:filter-comparison)"}
-
-filter.comp <- A2017$video[[3]] %>%
-  mutate(x = px2va(x, dim[1], res[1], dist),
-         y = px2va(y, dim[2], res[2], dist),
-         vel.x.sg = signal::sgolayfilt(x, m = 1),
-         vel.y.sg = signal::sgolayfilt(y, m = 1),
-         vel.sg = sqrt(vel.x.sg^2 + vel.y.sg^2)*fr,
-         acc.x.sg = signal::sgolayfilt(x, m = 2),
-         acc.y.sg = signal::sgolayfilt(y, m = 2),
-         acc.sg = sqrt(acc.x.sg^2 + acc.y.sg^2)*fr^2,
-         vel.x.sb = lead(x) - lag(x),
-         vel.y.sb = lead(y) - lag(y),
-         vel.sb = sqrt(vel.x.sb^2 + vel.y.sb^2)*fr,
-         acc.x.lp = lead(x) - 2*x + lag(x),
-         acc.y.lp = lead(y) - 2*y + lag(y),
-         acc.lp = sqrt(acc.x.lp^2 + acc.y.lp^2)*fr^2,
-         t = (t-t[1])/1e6) %>%
-  pivot_longer(c("vel.sg", "vel.sb", "acc.sg", "acc.lp"), names_to = c("metric", "filter"), names_sep = "[:punct:]", ) %>%
-  mutate_at(c("metric", "filter"), as.factor) %>%
-  dplyr::filter(t > 0.5, t < 1.5)
-
-filter.comp$metric <- factor(filter.comp$metric, labels = c("Acceleration", "Velocity"))
-
-ggplot(filter.comp, aes(x = t, y = value, color = filter)) + 
-  geom_path() + facet_wrap(vars(metric), nrow = 2, scales = "free_y", strip.position = "left") + 
-  scale_x_continuous(name = "Time (in s)") +
-  scale_y_continuous(name = "") +
-  scale_color_discrete(name = "Filter", labels = c("Laplace", "Sobel", "Savitzky-Golay")) +
-  theme(strip.placement = "outside")
-
-```
-
-## The Generative Model
-The generative model underlying gazeHMM is a multivariate HMM.
-It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits.
-Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.
-In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model.
-The initial state model contains probabilites for the first state of the hidden sequence.
-The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM.
-The response model encompasses distributions describing the response variables for every state in the model.
-In gazeHMM, the response model has three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal.
-The response variables are treated as conditionally independent on the states.
-Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated).
-This step was merely taken to keep the HMM simple and identifiable.
-Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them).
-However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation).
-Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean.
-Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties.
-In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose.
-It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.
-To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution.
-Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions.
-Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution.
-Thus, the uniform distribution should distinguish fixations from the other events.
-The initial state model and the transition model use multinomial distributions (with a category for each state).
-For all three submodels of the HMM, no time-varying covariates are included.
-Since the implementation of gazeHMM allows for regression models for each submodel and state, this implies that only intercepts for each parameter are estimated.
-
-## Missing Data and Blinks
-HMMs can estimate parameters and hidden states despite missing data.
-They either treat data as missing at random or as missing depending on states [@Visser2019].
-In gazeHMM, I assume data to be missing at random.
-However, most of the missing data in eye movement classification are due to blinks.
-Therefore, the user can indicate in gazeHMM which samples should be labeled as blinks (other missing samples are treated as noise).
-Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs.
-Since these samples could distort the classification of saccades in the HMM, I decided to remove them heuristically.
-Before classifying the samples, gazeHMM sets all samples within 50 ms before and after blink samples as missing.
-The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.
-When evaluating the likelihood of the HMM, given the parameters and joint density of response distributions, for missing data, the model integrates over all possible values.
-The resulting density for missing samples is therefore set to one [@Visser2019].
-
-## Optimization and Classification
-The parameters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997].
-The EM algorithm is generally suitable to estimate likelihoods with missing variables.
-For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data (velocity, acceleration, and sample-to-sample angle) and the expected hidden states [eye movement events; @Visser2019].
-The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability.
-Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988].
-
-## Postprocessing
-After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence.
-I implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all.
-Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events.
-Thus, the postprocessing routine heuristically compensates for such violations.
-This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events.
-Samples are relabeled as the state of the previous event.
-Finally, samples initially indicated as missing are labeled as noise (including blinks) and event metrics are computed (e.g., fixation duration).
-
-## Implementation
-The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
-The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
-
-\newpage
-
-# Simulation Study
-To assess how well the HMM recovers parameters and state sequences, I conducted a simulation study.
-The design and analysis of the study were preregistered on the Open Science Framework (https://doi.org/10.17605/OSF.IO/VDJGP).
-The major part of this section is a direct copy of the preregistration.
-When appropriate, additional explanations were added and tenses adapted to make the design easier to read and understand.
-In the study, the HMM repeatedly generated data with a set of parameters (true parameter values).
-The same model was then applied to estimate the parameters from the generated data (estimated parameter values).
-I compared the true with the estimated parameter values to assess whether a parameter was recovered by the model.
-Additionally, I compared the true states of the HMM with the estimated states to judge how accurately the model recovered the states that generated the data.
-
-## Starting Values
-The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities.
-To generate random starting values for the estimation of shape, scale, and concentration parameters, I used gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated.
-This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
-Mean parameters of the von Mises distribution always started at their true values.
-
-## Design
-### Parameter Variation
-The simulation study was divided into four parts.
-In the first part, I varied the parameters of the HMM.
-For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were varied, respectively.
-For each parameter, the HMM generated 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps.
-This resulted in $100 \times (10+15+20) = 4500$ recoveries.
-Only one parameter was varied at once, the other parameters were set to their default values.
-All parameters of the HMM were estimated freely (i.e., there were no fixed parameters in the model).
-I did not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification.
-For the transition probabilities, I only simultaneously varied the probabilities for staying in the same state (diagonals of the transition matrix) to reduce the complexity of the simulation.
-The left over probability mass was split evenly between the probabilities for switching to a different state (per row of the transition matrix).
-Moreover, I did not modify the mean parameters of the von Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
-I defined approximate ranges for each response variable and chose true parameter intervals and default values so that they produced samples that roughly corresponded to these ranges.
-Table \@ref(tab:sim-ranges) shows the assumed ranges for each event and Tables \@ref(tab:sim-parameters-trans) and \@ref(tab:sim-parameters-resp) show the intervals and default values for each parameter in the simulation.
-Parameters were scaled down by factor 10 (compared to the reported ranges) to improve fitting of the gamma distributions.
-I set the intervals for shape parameters of the gamma distributions for all events to [1,5] to examine how skewness influenced the recovery (shape values above five approach a symmetric distribution).
-The scale parameters were set so that the respective distribution approximately matched the assumed ranges.
-Since the concentration parameters of the von Mises distribution are the inverse of standard deviations, they were varied on the inverse scale.
-An example of simulated data from the HMM with default parameters is visualized in Figure \@ref(fig:plot-sim-ex).
-
-### Sample Size and Noise Variation
-In the second part, I varied the sample size of the generated data and the amount of noise added to it.
-The model parameters were set to their default values.
-For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, I generated 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
-These samples sizes roughly corresponded to small, medium, and large eye-tracking data sets for a single participant and trial.
-To simulate noise, I replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
-This procedure ensured that velocity and acceleration values remained positive and were taken from moderately skewed distributions with modes equal to the original values.
-To angle, I added white noise from a von Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets.
-$\tau_{noise}$ and $\kappa_{noise}$ were varied simultaneously in equidistant steps in their intervals.
-
-### Variation of Starting Values
-In the third part, I increased the variation in the starting values used for parameter estimation.
-The model parameters were set to their default values.
-For the shape, scale, and concentration parameters, I simultaneously increased the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta_{start} = (\psi_{true}/2)\tau_{start}$ with $\tau_{start} \in \{1, 2, 3\}$, 100 data sets with $N = 2500$ samples were generated each ($100 \times 3 \times 3 = 900$ recoveries).
-
-### Missing Data
-In the last part, I set intervals of the generated data to be missing.
-The model parameters were set to their default values.
-For $k \in \{2, 3, 4\}$ states and $m \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples were generated ($100 \times 3 \times 3 = 900$ recoveries).
-The length of the missing data interval $l \in [1,200]$ samples varied in equidistant steps between the data sets.
-
-## Data Analysis [^6]
-For each parameter, I calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
-Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\psi_{true} \ne 0$.
-This was not the case for some mean parameters of the von Mises distributions.
-In those cases, I used $\psi_{true} = 2\pi$ instead.
-I treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter.
-By taking the median, I reduced the influence of potential outliers in the estimation and using proportions enabled me to compare RMdSPD values across parameters and data sets.
-Additionally, I applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in part one.
-Regression slopes closer to one indicated that the model better captured parameter change.
-Regression intercepts different from zero reflected a bias in parameter estimation.
-To assess state recovery, I computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set.
-Cohen's kappa estimates the agreement between two classifiers accounting for the agreement due to chance.
-Higher kappa values were interpreted as better model accuracy.
-I adopted the ranges proposed by @Landis1977 to interpret kappa values.
-Models that could not be fitted were excluded from the recovery.
-
-[^6]: Note that this section deviates from my proposed analysis plan. However, since I preregistered the changed analysis plan, I treated the results of the simulation as confirmatory.
+# Approximate Ranges for Response Variables
+The approximate ranges that were used to define true parameter ranges and default values are displayed in Table \@ref(tab:sim-ranges).
+Since we would also like to test the algorithm on extreme data distributions, we extended the ranges beyond those found in typical eye movement data.
 
@@
 Velocity ranges are based on event velocities reported in @Larsson2013.
-Since I would also like to test the algorithm on extreme data distributions, I extended the ranges beyond those found in typical eye movement data.
 
@@
 
-(ref:sim-parameters-trans) Intervals and Default Parameter Values for the Transition Model in the Simulation
-(ref:sim-parameters-trans-note) The transition probability for staying in the same state is denoted by $a_{i=j}$ and the probability for switching to a different state by $a_{i\neq j}$.
-The number of states in the model is denoted by *k*.
-
-```{r sim-parameters-trans}
-
-sim.pars.trans <- as.data.frame(matrix(c("Interval", "-", "[.01,.99]", "1-a(i=j)/(k-1)",
-                                           "Default", "1/k", "0.9", "0.1/(k-1)"), nrow = 2, byrow = T))
-
-apa_table(sim.pars.trans,
-          col.names = c("", "Initial\nstate\nprob.", "Trans. prob.\nsame state", "Trans. prob.\nother state"),
-          caption = "(ref:sim-parameters-trans)",
-          note = "(ref:sim-parameters-trans-note)")
-
-```
-
-(ref:sim-parameters-resp) Intervals and Default Parameter Values for the Response Model in the Simulation
-(ref:sim-parameters-resp-note) Shape parameters are denoted by $\alpha$, scale parameters by $\beta$, mean parameters by $\mu$, and concentration parameters by $\kappa$.
-The default values for the uniform distribution in state one were min = 0 and max = $2\pi$.
-
-```{r sim-parameters-resp}
-
-sim.pars.resp <- as.data.frame(t(matrix(c("[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
-                                  "3", "0.35", "3", "0.25", "-", "-",
-                                  "[1,5]", "[5,15]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
-                                  "3", "10", "3", "3", "0", "1",
-                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
-                                  "3", "1", "3", "3", paste(expression(pi), sep = ""), "1",
-                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]",
-                                  "3", "1", "3", "0.15", "0", "1"), nrow = 6, byrow = F)))
-
-apa_table(cbind(rep(c("Interval", "Default"), 4), sim.pars.resp),
-          col.names = c("", "Shape", "Scale", "Shape", "Scale", "Mean", "Concentration"),
-          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
-          stub_indents = list("State 1" = c(1, 2), "State 2" = c(3, 4), "State 3" = c(5, 6), "State 4" = c(7, 8)),
-          caption = "(ref:sim-parameters-resp)",
-          note = "(ref:sim-parameters-resp-note)")
-
-```
+# Starting Values and Missing Data
+In the third part of the simulation study, we increased the variation in the starting values used for parameter estimation.
+The model parameters were set to their default values.
+For the shape, scale, and concentration parameters, we simultaneously increased the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta_{start} = (\psi_{true}/2)\tau_{start}$ with $\tau_{start} \in \{1, 2, 3\}$, 100 data sets with $N = 2500$ samples were generated each ($100 \times 3 \times 3 = 900$ recoveries).
+In the last part, we set intervals of the generated data to be missing.
+The model parameters were set to their default values.
+For $k \in \{2, 3, 4\}$ states and $m \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples were generated ($100 \times 3 \times 3 = 900$ recoveries).
+The length of the missing data interval $l \in [1,200]$ samples varied in equidistant steps between the data sets.
 
@@
 
-\newpage
-
-# Results [^5]
-## Simulation Study
-### Parameter Variation
-#### Two States
-In the first part of the simulation, I examined how varying the parameters in the HMM affects the deviation of estimated parameters and accuracy of estimated state sequences.
+# Results
+## Addendum Part One
 Figure \@ref(fig:plot-rmdspd-1-2) displays the RMdSPD between true and estimated parameters depending on which parameter has been manipulated in the HMM.
-The RMdSPD was below 0.1 for all estimated and manipulated parameters, indicating good recovery [^2].
+The RMdSPD was below 0.1 for all estimated and manipulated parameters, indicating good recovery.
 The regressions between manipulated true and estimated parameters are shown in Figures \@ref(fig:plot-regtr-1-2) and \@ref(fig:plot-regresp-1-2).
@@
 
-[^5]: The design and analysis plan of the simulation study have not been included in my proposal.
-However, since I preregistered both the design and analysis, I treated the results as confirmatory.
-In general, the evaluation of an algorithm can hardly be proposed in detail before the algorithm has been developed.
-Therefore, I chose to report a general results section.
-Nevertheless, I indicated whenever results were obtained through exploratory analyses that were not planned beforehand.
-
-[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
-Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
-Therefore, I decided not to include it in the analysis.
-
-(ref:plot-rmdspd-1-2) RMdSPD between true and estimated parameters of the two-state HMM in part one of the simulation.
-Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
-Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
+(ref:plot-rmdspd-1-2) RMdSPD between true and estimated parameters of the two-state HMM in part two of the simulation.
+Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Top facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 
-#### Three States
-For the simulation with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
-When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently below 0.5.
-Varying $\kappa$ in states two and three led to RMdSPDs below 0.5 in the respective states, which can be interpreted as moderate recovery.
-Otherwise, RMdSPDs were consistently lower than 0.1, indicating good recovery.
-Inspecting the regressions between manipulated true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one).
-Again, the deviation seemed to increase with true parameter magnitude (reverse for kappas).
-In contrast to the two-state HMM, larger deviations and more outliers were observed.
-Cohen's kappa values are presented in Figure \@ref(fig:plot-acc-1-3).
-For most estimated models, the kappas between true and estimated state sequences were above 0.95, meaning almost perfect agreement.
-However, for some models, I observed kappas clustered around zero or -0.33, which suggests that state labels were switched.
-To summarize, the three-state HMM had a slightly worse parameter recovery and more outliers than the two-state model but shows a similar good accuracy.
-
-(ref:plot-rmdspd-1-3) RMdSPD between true and estimated parameters of the three-state HMM in part one of the simulation.
-Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
-Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-1-3, fig.cap="(ref:plot-rmdspd-1-3)"}
-
-print(plots.rmsd.1[[2]]+ theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-regtr-1-3) Regression lines between true and estimated transition probabilities for the three-state HMM in part one.
-Top facet labels show to and right facet labels show from which state the HMM is moving.
-Dashed lines refer to perfect recovery.
-
-```{r plot-regtr-1-3, fig.cap="(ref:plot-regtr-1-3)"}
-
-print(plots.lm.tr[[2]])
-
-```
-
-(ref:plot-regresp-1-3) Regression lines between true and estimated response parameters of the three-state HMM in part one.
-Top facet labels indicate response parameters.
-Dashed lines refer to perfect recovery.
-
-```{r plot-regresp-1-3, fig.cap="(ref:plot-regresp-1-3)", fig.height=7}
-
-print(plots.lm.resp[[2]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-acc-1-3) Boxplots displaying Cohen's kappa depending on which parameter of the three-state HMM has been manipulated in part one.
-Top facet labels indicate for which state parameters have been manipulated.
-Black solid lines symbolize medians and hinges the first and third quartile.
-Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
-Crosses represent outliers.
-
-```{r plot-acc-1-3, fig.cap="(ref:plot-acc-1-3)"}
-
-print(plots.acc.1[[2]])
-
-```
-
-#### Four States
-The RMdSPDs for the four-state HMM is shown in Figure \@ref(fig:plot-rmdspd-1-4).
-For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were below 0.5, suggesting moderate recovery.
-Estimated kappa parameters in state four were also often below 0.5 when parameters in states two, three, and four were varied.
-Otherwise, RMdSPDs were below 0.1, indicating good recovery.
-Looking at the regressions between true and estimated parameters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) illustrate strong and unbiased relationships.
-However, there were larger deviations and more outliers than in the previous models, especially for states one and four.
-Accuracy measured by Cohen's kappa ranged between 0.6 and 0.9 for the majority of models, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
-Here, some kappa values clustered around 0.25 and zero, which, again, can be interpreted as the result of label switching.
-Summarizing, the parameter recovery for the four-state HMM was slightly worse with even more outliers compared to the three-state model.
-Especially the recovery of transition parameters in states corresponding to fixations and smooth pursuits decreased.
-Overall, simulations in part one demonstrated that the HMM recovered parameters very well when parameters were manipulated.
-Deviations from true parameters were mostly small.
-In the four-state model, estimated transition probabilities for state one and four deviated moderately.
-Moreover, the HMM estimated state sequences very accurately with decreasing accuracy for the four-state model.
-
-(ref:plot-rmdspd-1-4) RMdSPD between true and estimated parameters of the four-state HMM in part one of the simulation.
-Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
-Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-1-4, fig.cap="(ref:plot-rmdspd-1-4)"}
-
-print(plots.rmsd.1[[3]] + theme(axis.text = element_text(size = 6)))
-
-```
-
-(ref:plot-regtr-1-4) Regression lines between true and estimated transition probabilities for the four-state HMM in part one.
-Top facet labels show to and right facet labels show from which state the HMM is moving.
-Dashed lines refer to perfect recovery.
-
-```{r plot-regtr-1-4, fig.cap="(ref:plot-regtr-1-4)", fig.height=8}
-
-print(plots.lm.tr[[3]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-regresp-1-4) Regression lines between true and estimated response parameters of the four-state HMM in part one.
-Top facet labels indicate response parameters.
-Dashed lines refer to perfect recovery.
-
-```{r plot-regresp-1-4, fig.cap="(ref:plot-regresp-1-4)", fig.height=8}
-
-print(plots.lm.resp[[3]] + theme(axis.text = element_text(size = 6)))
-
-```
-
-(ref:plot-acc-1-4) Boxplots displaying Cohen's kappa depending on which parameter of the four-state HMM has been manipulated in part one.
-Top facet labels indicate for which state parameters have been manipulated.
-Black solid lines symbolize medians and hinges the first and third quartile.
-Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
-Crosses represent outliers.
-
-```{r plot-acc-1-4, fig.cap="(ref:plot-acc-1-4)"}
-
-print(plots.acc.1[[3]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-### Sample Size and Noise Variation
-#### Two States
-In the second part, I varied the sample size of the HMM and added noise to the generated data.
-For the two-state HMM, the RMdSPDs were above 0.5 for $\beta_{vel}$ and $\beta_{acc}$ in both states (see Figure \@ref(fig:plot-rmdspd-2-2)), suggesting bad recovery.
-The other estimated parameters showed RMdSPDs close to or below 0.1, which means they were recovered well.
-Increasing the sample size seemed to improve RMdSPDs for most parameters slightly.
-For $\beta_{vel}$ and $\beta_{acc}$ in both states, models with 2500 samples had the lowest RMdSPDs.
-Accuracy measured by Cohen's kappa was almost perfect with kappa values very close to one (see Figure \@ref(fig:plot-acc-2), left plot).
-To conclude, adding noise only affected the recovery of scale parameters but not the accuracy of the two-state HMM.
-Increasing the sample size improved the recovery slightly.
-
-(ref:plot-rmdspd-2-2) RMdSPD between true and estimated parameters of the two-state HMM in part two of the simulation.
-Colours indicate different sizes of generated data.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-2-2, fig.cap="(ref:plot-rmdspd-2-2)"}
-
-print(plots.rmsd.234[[1]][[1]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-acc-2) Cohen's kappa depending on the variation of noise added to the data generated by the HMM.
-Colours indicate different sizes of generated data.
-Top facet labels indicate the number of states in the HMM.
-
-```{r plot-acc-2, fig.cap="(ref:plot-acc-2)"}
-
-print(plots.acc.int.24[[1]])
-
-```
-
-#### Three States
-The RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)), indicating bad recovery.
-Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5.
-For most parameters across all three states, higher sample sizes had lower RMdSPDs.
-The accuracy of the estimated models was almost perfect with most kappa values above 0.95 (see Figure \@ref(fig:plot-acc-2), middle plot).
-Several outliers clustered around kappas of zero and -0.33, signaling label switching.
-For the three-state HMM, adding noise led to a slightly worse recovery and accuracy overall.
-Scale parameters of gamma distributions in all three states were only badly recovered.
-
-(ref:plot-rmdspd-2-3) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-2-3, fig.cap="(ref:plot-rmdspd-2-3)"}
-
-print(plots.rmsd.234[[1]][[2]] + theme(axis.text = element_text(size = 6)))
-
-```
-
-#### Four States
-RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
-For states one and four, values for most parameters (including all transition probabilities) were above 0.5, suggesting bad recovery.
-Similarly, RMdSPDs for $\beta_{vel}$ and $\beta_{acc}$ in states two and three were above 0.5.
-For states two and three, higher samples sizes showed slightly lower RMdSPDs.
-As in the previous part, most Cohen's kappa values ranged between 0.6 and 0.9, meaning substantial to almost perfect agreement between true and estimated states (Figure \@ref(fig:plot-acc-2), right plot).
-Multiple kappa values clustered around 0.25 or zero, which can be explained by label switching.
-In summary, adding noise to data generated by the four-state HMM heavily decreased the recovery for most parameters in states corresponding to fixations and smooth pursuits.
-For saccade and PSO states, only scale parameters of gamma distributions were badly recovered, but increasing the sample size slightly improved the recovery.
-In general, the HMM recovered parameters well despite noise being added to the data.
-However, in the four-state model, the parameter recovery for states one and four substantially decreased.
-In the three- and four-state models, scale parameters of gamma distributions were badly recovered.
-Increasing the sample size in the HMM slightly improved the recovery of most parameters.
-The accuracy of the model was slightly lowered when more states were included, but it was neither affected by the noise variability $\tau_{noise}$ nor the sample size.
-
-(ref:plot-rmdspd-2-4) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-
-```{r plot-rmdspd-2-4, fig.cap="(ref:plot-rmdspd-2-4)"}
-
-print(plots.rmsd.234[[1]][[3]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-### Variation of Starting Values
+## Part Three
 The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM.
@@
 Colours indicate the variation in starting values used to estimate parameters.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
 Top facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 
-print(plots.rmsd.234[[2]][[1]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[2]][[1]] + coord_flip())
 
@@
 
-print(plots.rmsd.234[[2]][[2]] + theme(axis.text = element_text(size = 6)))
+print(plots.rmsd.234[[2]][[2]] + coord_flip())
 
@@
 
-print(plots.rmsd.234[[2]][[3]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-In this part, I explored post-hoc whether clusters of low Cohen's kappa values were due to label switching.
+print(plots.rmsd.234[[2]][[3]] + coord_flip())
+
+```
+
+## Part Four
+In this part, we explored post-hoc whether clusters of low Cohen's kappa values were due to label switching.
 According to @Visser2019, label switching occurs when exchanging the order of the states leads to equally likely models.
@@
 In this case, an HMM with three states that is perfectly accurate but has two labels switched would have a Cohen's kappa of 0.
-To test if label switching occurred, I manually switched one or two labels post-hoc (see Figure \@ref(fig:plot-acc-3-expl)).
+To test if label switching occurred, we manually switched one or two labels post-hoc (see Figure \@ref(fig:plot-acc-3-expl)).
 It can be seen that this approach resolved low accuracy clusters for three and four states.
@@
 
-### Missing Data
 In the last part, data intervals of varying length were set to be missing.
@@
 In total, missing data did not affect the parameter recovery but linearly decreased the accuracy of the model from a nearly perfect to a moderate extent.
-With a few exceptions, the simulation study revealed that the HMM recovered parameters well and accurately estimated the true state sequences.
-The most critical decrease in recovery occurred when noise was added to the data generated by models including four states.
-In the noise condition, scale parameters of gamma distributions were often badly recovered.
-Higher sample sizes slightly improved parameter recovery, but neither variation in starting values nor missing data affected it.
-In contrast, the accuracy of the model was linearly decreasing with more data missing but not influenced by manipulating parameters, noise, or starting value variation.
-Adding more states to the HMM generally decreased the parameter recovery and the accuracy.
-
-(ref:plot-rmdspd-4-2) RMdSPD between true and estimated parameters of the two-state HMM in part four of the simulation.
+
+(ref:plot-rmdspd-4-2) RRMdSPD between true and estimated parameters of the two-state HMM in part four of the simulation.
 Colours indicate the number of missing data intervals.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
+Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
+Right facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 
-print(plots.rmsd.234[[3]][[1]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[3]][[1]] + coord_flip())
 
@@
 Colours indicate the number of missing data intervals.
-Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed.
+Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
 Top facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 
-print(plots.rmsd.234[[3]][[2]] + theme(axis.text = element_text(size = 6)))
+print(plots.rmsd.234[[3]][[2]] + coord_flip())
 
@@
 
-print(plots.rmsd.234[[3]][[3]] + theme(axis.text = element_text(size = 8)))
+print(plots.rmsd.234[[3]][[3]] + coord_flip())
 
@@
 ```
-
-## Validation
-### Starting Values and Model Fitting
-To test my hypotheses, I applied gazeHMM on two benchmark data sets.
-Note that, for the @Ehinger2019 data set, I only fitted models to the Eyelink data to reduce the complexity of the analysis.
-For both data sets, I used the same set of starting values (initial state model: $\rho=1/k$; transition model: $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$; response model: see Table \@ref(tab:tab-starting-values)) to estimate the parameters of the HMM.
-In contrast to the simulation study, generating random starting values often led to bad model fits and label switching between states.
-To improve the fitting of the gamma distributions, velocity and acceleration signals were scaled down by factor 100 [^7] (so were the starting values for their gamma distributions).
-The algorithm was applied separately for every participant and every condition, task, or block.
-For the data set by @Andersson2017, the algorithm was successfully fitted for every participant and condition.
-Examples of event classification and parameter estimates from gazeHMM are presented in Appendix A.
-
-[^7]: Scaling down by factor 100 differs from the simulation study (scaling down by 10).
-The algorithm allows the user to manually specify this factor and in this case, factor 100 led to better model fits than factor 10.
-Thus, I chose it exploratorily.
-
-(ref:tab-starting-values) Starting Values for the Response Model for Fitting gazeHMM to Benchmark Data
-(ref:tab-starting-values-note) Starting values for velocity and acceleration signals are shown before scaling down by factor 100.
-Values for event 5 were chosen so that they approximately match plausible distributions for microsaccades.
-
-```{r tab-starting-values}
-
-tab.start <- as.data.frame(matrix(c(10, 10, 10, 10, NA, NA,
-                      50, 50, 50, 50, 0, 10,
-                      50, 50, 50, 50, pi, 10,
-                      20, 20, 20, 20, 0, 10,
-                      20, 20, 50, 50, 0, 10), nrow = 5, byrow = T))
-
-colnames(tab.start) <- c("Shape", "Scale", "Shape", "Scale", "Mean", "Concentration")
-rownames(tab.start) <- c("Fixation", "Saccade", "PSO", "Pursuit", "Event 5")
-
-apa_table(as.data.frame(tab.start),
-          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-starting-values)",
-          note = "(ref:tab-starting-values-note)")
-
-```
-
-### Model Comparison
-
-```{r load validation results, include=FALSE}
-
-load(here("validation/Andersson2017_fitted.Rdata"))
-load(here("validation/Andersson2017_raw.Rdata"))
-
-```
-
-```{r calculate Schwarz weights, include=FALSE}
-
-# Compute Schwarz weights
-
-schwarz.weights <- function(bic, na.rm = T) {
-
-  d.bic <- bic - min(bic, na.rm = na.rm) # eq 2
-
-  exp(-0.5 * d.bic)/sum(exp(-0.5 * d.bic), na.rm = na.rm) # eq 4
-
-}
-
-A2017.bic <- lapply(A2017.fit, function(stim) {
-  out <- lapply(stim, function(subj) {
-    out <- lapply(1:length(subj), function(mod) {
-
-      if(mod == 1) {
-
-        bic <- try(-2*subj[[mod]][["LL"]] + 6*log(subj[[mod]][["N"]]))
-
-      } else {
-
-        bic <- try(BIC(subj[[mod]]$model))
-
-      }
-
-      return(ifelse(is.numeric(bic), bic, NA))
-    })
-
-    return(schwarz.weights(unlist(out)))
-  })
-
-  df <- as.data.frame(reduce(out, rbind))
-
-  names(df) <- paste("model_", 1:length(stim[[1]]), sep = "")
-
-  return(df)
-})
-
-# Create Schwarz weight plots
-
-bic.plot <- lapply(A2017.bic, function(x) {
-
-  data.long <- x %>% 
-    mutate(subject = 1:nrow(x)) %>%
-    pivot_longer(names(x), names_to = "model", values_to = "weight") %>%
-    mutate_at(c("subject", "model"), as.factor)
-
-  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
-    scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
-    scale_y_discrete(name = "Subject") +
-    scale_fill_distiller(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")
-
-  return(p)
-})
-
-```
-
-To examine whether gazeHMM accurately describes eye movements, I applied gazeHMM with one, two, three, four, and five states on two benchmark data sets and compared the Schwarz weights of the HMMs.
-The starting values for the fifth state were chosen to represent distributions for microsaccade samples (velocity and sample-to-sample angle as for smooth pursuits, acceleration as for saccades and PSOs).
-Schwarz weights are the result of transforming a set of BIC values [@Wagenmakers2004].
-They can be interpreted as the probability for a model to have generated the data it was fitted to.
-For the @Andersson2017 data set, I expected the highest weights for the three-state models in the image condition.
-For the moving dots and video conditions, I predicted the highest weights for the four-state model.
-Figure \@ref(fig:plot-schwarz-weights-img) shows the Schwarz weights for different subjects and models in the image condition.
-For all subjects, the five-state model displayed the highest weight, suggesting that it most likely generated the data.
-In the moving dots condition, the five-state model had the highest weights for most subjects, but the one-, three, and four-state models exhibited the highest weights for two subjects each (see Figure \@ref(fig:plot-schwarz-weights-dots)).
-The video condition had the same pattern as the image condition, as the five-state model consistently had the highest weights (see Figure \@ref(fig:plot-schwarz-weights-vid)).
-In all three conditions, response parameters of the fifth state were mostly similar to those of the smooth pursuit state.
-In contrast to my hypothesis, applying different HMMs to the @Andersson2017 data suggests that the five-state model has most likely generated the data.
-Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.
-For the @Ehinger2019 data, I predicted a one-state model to be preferred.
-Here, the model comparison for the yielded similar results (see Appendix A), contradicting my hypothesis.
-A recent model recovery study showed that the BIC tended to prefer overly complex HMMs when they were misspecified [e.g., the conditional independence assumption was violated; @Pohle2017].
-Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2000] performed better in choosing the correct data generating model.
-Therefore, I computed the weighted ICL criterion for the HMMs fitted to the @Andersson2017 data set as an exploratory analysis.
-Using the ICL as the model selection criterion yielded very similar results to the BIC (see Appendix A).
-The preference for the five-state model was even more consistent across conditions and subjects.
-
-(ref:plot-schwarz-weights-img) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the image condition of the @Andersson2017 data set.
-Higher weights indicate better model fit.
-
-```{r plot-schwarz-weights-img, fig.cap="(ref:plot-schwarz-weights-img)"}
-
-print(bic.plot[[2]])
-
-```
-
-(ref:plot-schwarz-weights-dots) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the moving dots condition of the @Andersson2017 data set.
-Higher weights indicate better model fit.
-
-```{r plot-schwarz-weights-dots, fig.cap="(ref:plot-schwarz-weights-dots)"}
-
-print(bic.plot[[1]])
-
-```
-
-(ref:plot-schwarz-weights-vid) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the video condition of the @Andersson2017 data set.
-Higher weights indicate better model fit.
-
-```{r plot-schwarz-weights-vid, fig.cap="(ref:plot-schwarz-weights-vid)"}
-
-print(bic.plot[[3]])
-
-```
-
-### Comparison to Other Algorithms
-
-```{r calculate duration RMSD, include=FALSE}
-
-# Calculate event descriptives for human coders
-
-fr <- 500
-
-A2017.events <- lapply(A2017, function(stim) {
-  lapply(1:4, function(e) {
-    out <- lapply(stim, function(df) {
-
-      if(all(is.nan(df$t) || df$t == 0)) {
-
-        df$t <- seq(0, length(df$t)-1)/fr
-
-      } else {
-
-        df$t <- (df$t - df$t[1])/1e6
-
-        df <- df[df$t >= 0,]
-
-      }
-
-      counter_MN <- 1
-      counter_RA <- 1
-
-      number_MN <- numeric(nrow(df))
-      number_RA <- numeric(nrow(df))
-
-      number_MN[1] <- 1
-      number_RA[1] <- 1
-
-      for (i in 2:nrow(df)) {
-        if(df$label_MN[i] != df$label_MN[i-1]) counter_MN <- counter_MN + 1
-        if(df$label_RA[i] != df$label_RA[i-1]) counter_RA <- counter_RA + 1
-
-        number_MN[i] <- counter_MN
-        number_RA[i] <- counter_RA
-
-      }
-
-      dur_MN <- numeric(max(number_MN))
-      event_MN <- numeric(max(number_MN))
-
-      for (n in unique(number_MN)) {
-        if(n > 1) {
-          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - max(df$t[number_MN == (n-1)], na.rm = T)
-        } else {
-          dur_MN[n] <- max(df$t[number_MN == n], na.rm = T) - min(df$t[number_MN == n], na.rm = T)
-        }
-
-        event_MN[n] <- max(df$label_MN[number_MN == n], na.rm = T)
-      }
-
-      dur_RA <- numeric(max(number_RA))
-      event_RA <- numeric(max(number_RA))
-
-      for (n in unique(number_RA)) {
-        if(n > 1) {
-          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - max(df$t[number_RA == (n-1)], na.rm = T)
-        } else {
-          dur_RA[n] <- max(df$t[number_RA == n], na.rm = T) - min(df$t[number_RA == n], na.rm = T)
-        }
-
-        event_RA[n] <- max(df$label_RA[number_RA == n], na.rm = T)
-      }
-
-      return(list(dur_MN[event_MN == e], dur_RA[event_RA == e]))
-    })
-
-    df <- reduce(out, cbind)
-
-    dur_MN <- unlist(df[1,])
-    dur_RA <- unlist(df[2,])
-
-    return(data.frame(MN = c(mean(dur_MN), sd(dur_MN), length(dur_MN)),
-                      RA = c(mean(dur_RA), sd(dur_RA), length(dur_RA))))
-  })
-})
-
-# Calculate RMSD between event distribution descriptives
-
-ref <- list(dot = list(fix = list(CDT = c(60, 127, 165),
-                                  EM = c(NA, NA, NA),
-                                  IDT = c(323, 146, 8),
-                                  IKF = c(217, 184, 72),
-                                  IMST = c(268, 140, 12),
-                                  IHMM = c(214, 286, 67),
-                                  IVT = c(203, 282, 71),
-                                  NH = c(380, 333, 30),
-                                  BIT = c(189, 113, 67),
-                                  LNS = c(NA, NA, NA)),
-                       sac = list(CDT = c(NA, NA, NA),
-                                  EM = c(17, 14, 93),
-                                  IDT = c(32, 14, 10),
-                                  IKF = c(60, 26, 29),
-                                  IMST = c(13, 5, 18),
-                                  IHMM = c(41, 17, 27),
-                                  IVT = c(36, 14, 28),
-                                  NH = c(43, 16, 42),
-                                  BIT = c(NA, NA, NA),
-                                  LNS = c(26, 11, 53)),
-                       pso = list(NH = c(24, 12, 17),
-                                  LNS = c(20, 9, 31))),
-            img = list(fix = list(CDT = c(397, 559, 251),
-                                  EM = c(NA, NA, NA),
-                                  IDT = c(399, 328, 242),
-                                  IKF = c(174, 239, 513),
-                                  IMST = c(304, 293, 333),
-                                  IHMM = c(133, 216, 701),
-                                  IVT = c(114, 204, 827),
-                                  NH = c(258, 299, 292),
-                                  BIT = c(209, 136, 423),
-                                  LNS = c(NA, NA, NA)),
-                       sac = list(CDT = c(NA, NA, NA),
-                                  EM = c(25, 22, 787),
-                                  IDT = c(25, 15, 258),
-                                  IKF = c(62, 37, 353),
-                                  IMST = c(17, 10, 335),
-                                  IHMM = c(48, 26, 368),
-                                  IVT = c(41, 22, 373),
-                                  NH = c(50, 20, 344),
-                                  BIT = c(NA, NA, NA),
-                                  LNS = c(29, 12, 390)),
-                       pso = list(NH = c(28, 13, 237),
-                                  LNS = c(25, 9, 319))),
-            vid = list(fix = list(CDT = c(213, 297, 211),
-                                  EM = c(NA, NA, NA),
-                                  IDT = c(554, 454, 48),
-                                  IKF = c(258, 296, 169),
-                                  IMST = c(526, 825, 71),
-                                  IHMM = c(234, 319, 194),
-                                  IVT = c(202, 306, 227),
-                                  NH = c(429, 336, 83),
-                                  BIT = c(248, 215, 170),
-                                  LNS = c(NA, NA, NA)),
-                       sac = list(CDT = c(NA, NA, NA),
-                                  EM = c(20, 16, 252),
-                                  IDT = c(24, 53, 41),
-                                  IKF = c(55, 20, 107),
-                                  IMST = c(18, 10, 76),
-                                  IHMM = c(42, 18, 109),
-                                  IVT = c(36, 16, 112),
-                                  NH = c(44, 18, 104),
-                                  BIT = c(NA, NA, NA),
-                                  LNS = c(28, 12, 122)),
-                       pso = list(NH = c(28, 13, 78),
-                                  LNS = c(24, 10, 87))))
-
-A2017.rmsd <- lapply(2:4, function(k) {
-  lapply(1:length(A2017.fit), function(x) { 
-    out <- lapply(1:k, function(y) {
-      out <- lapply(A2017.fit[[x]], function(z) {
-
-        if(class(z[[k]]) == "gazeHMM") z[[k]]$events[[y]]
-
-      })
-
-      df <- try(reduce(out, rbind))
-
-      alg <- try(c(mean(df$dur, na.rm = T), sd(df$dur, na.rm = T), nrow(df)))
-
-      if(y == 4) {
-        M <- try(cbind(A2017.events[[x]][[y]], alg))
-      } else {
-        M <- try(cbind(A2017.events[[x]][[y]], alg, as.data.frame(ref[[x]][[y]])))
-        M[1:2,4:ncol(M)] <- M[1:2,4:ncol(M)]/1e3
-      }
-
-      M.norm <- try(apply(M, 1, function(x) {(x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))}))
-
-      if(y == 4) {
-        rmsd <- try(sum(sqrt((M.norm[3,] - colSums(M.norm[1:2,]/2))^2)))
-      } else {
-        rmsd <- try(apply(M.norm[3:nrow(M.norm),], 1, function(r) {sum(sqrt((r - colSums(M.norm[1:2,]/2))^2))}))
-      }
-
-      coder <- try(sum(sqrt((M.norm[1,] - M.norm[2,])^2)))
-
-      rmsd <- c(coder, coder, rmsd)
-
-      return(rbind(M, rmsd))
-    })
-  })
-})
-
-# Create tables for RMSDs
-
-rmsd.table <- lapply(1:length(A2017.rmsd), function(x) {
-  out <- lapply(1:length(A2017.rmsd[[x]]), function(y) {
-    out <- lapply(1:length(A2017.rmsd[[x]][[y]]), function(z) {
-
-      tabmat <- t(apply(A2017.rmsd[[x]][[y]][[z]], 2, function(d) {as.character(round(d, 3))}))
-
-      # tabmat[3,] <- paste("**", tabmat[3,], "**", sep = "")
-
-      colnames(tabmat) <- c("Mean", "SD", "Events", "RMSD")
-      rownames(tabmat)[1:3] <- c("coderMN", "coderRA", "gazeHMM")
-
-      return(cbind(z, rownames(tabmat), tabmat))
-    })
-
-    cbind(y, reduce(out, rbind))
-  })
-
-  df <- as.data.frame(reduce(out, rbind), stringsAsFactors = F) %>%
-    mutate(Condition = factor(y, labels = c("moving dots", "image", "video")))
-
-  lapply(1:(x+1), function(x) {
-
-    df <- df %>% dplyr::filter(z == x) %>%
-      dplyr::select(-c("z", "y")) %>%
-      rename(Algorithm = V3) %>%
-      pivot_wider(names_from = "Condition", names_glue = "{Condition}_{.value}", names_sort = T, values_from = c("Mean", "SD", "Events", "RMSD"))
-
-    df %>% dplyr::select(c("Algorithm", "image_Mean", "image_SD", "image_Events", "image_RMSD",
-                           "moving dots_Mean", "moving dots_SD", "moving dots_Events", "moving dots_RMSD",
-                           "video_Mean", "video_SD", "video_Events", "video_RMSD"))
-  })
-})
-
-```
-
-```{r calculate Cohen s kappa and confusion matrices, include= FALSE}
-
-# Compute Cohen's kappa and confusion matrices
-
-ref.acc <- matrix(c(0.92, 0.81, 0.83, 0.95, 0.91, 0.94, 0.88, 0.82, 0.83, NA, NA, NA,
-                    0.92, 0.84, 0.82, 0.95, 0.91, 0.94, 0.88, 0.80, 0.81, NA, NA, NA,
-                    0.38, 0.06, 0.11, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.00, 0.00, 0.00, 0.64, 0.66, 0.67, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.36, 0.00, 0.03, 0.45, 0.26, 0.38, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.63, 0.03, 0.14, 0.58, 0.46, 0.59, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.38, 0.00, 0.03, 0.54, 0.30, 0.52, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.67, 0.03, 0.13, 0.69, 0.60, 0.71, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.67, 0.03, 0.13, 0.75, 0.63, 0.76, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.52, 0.00, 0.01, 0.67, 0.60, 0.68, 0.24, 0.20, 0.25, NA, NA, NA,
-                    0.67, 0.03, 0.14, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, NA, NA, NA,
-                    0.00, 0.00, 0.00, 0.81, 0.75, 0.81, 0.64, 0.59, 0.63, NA, NA, NA), ncol = 12, byrow = T)
-
-A2017.acc <- lapply(2:4, function(k) {
-  lapply(1:length(A2017.fit), function(x) { 
-    out <- lapply(1:length(A2017.fit[[x]]), function(y) {
-
-      if(class(A2017.fit[[x]][[y]][[as.character(k)]]) == "gazeHMM") {
-        cod.MN <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_MN))
-        cod.RA <- as.data.frame(cbind(A2017.fit[[x]][[y]][[as.character(k)]]$samples$label, A2017[[x]][[y]]$label_RA))
-
-        return(rbind(cod.MN, cod.RA))
-      }
-    })
-
-    df <- reduce(out, rbind)
-
-    df[df[,1] == 0,1] <- 5
-
-    events <- lapply(1:k, function(e) {
-
-      alg <- df[,1] == e
-      cod <- df[,2] == e
-
-      mat <- matrix(c(sum(alg & cod, na.rm = T), sum(alg & !cod, na.rm = T), 
-                      sum(!alg & cod, na.rm = T), sum(!alg & !cod, na.rm = T)), 2, 2)
-
-      over <- mat[2,1]/sum(mat[,1])
-      under <- mat[1,2]/sum(mat[,2])
-
-      return(list(kappa = cohen.kappa(mat)$kappa, 
-                  over = over,
-                  under = under))
-    })
-
-    ratio <- mean(df[,1] != df[,2], na.rm = T)
-
-    conmat <- table(df[,1], df[,2])
-
-    return(list(events = events, ratio =  ratio, conmat = conmat))
-  })
-})
-
-# Create tables for Cohen's kappa and confusion matrices
-
-kappa.table <- lapply(A2017.acc, function(x) {
-  out <- lapply(x, function(y){
-    out <- lapply(y$events, function(z){
-      return(round(z$kappa, 3))
-    })
-
-    return(reduce(out, cbind))
-  })
-
-  tabmat <- reduce(out, rbind)
-
-  tabmat <- rbind(tabmat[2,], tabmat[1,], tabmat[3,])
-
-  return(as.vector(tabmat))
-})
-
-conf.table <- lapply(A2017.acc, function(x) {
-  out <- lapply(x, function(y){
-
-    tabmat <- apply(y$conmat, 2, as.character)
-
-    # diag(tabmat)[1:(nrow(tabmat)-1)] <- paste("**", diag(tabmat)[1:(nrow(tabmat)-1)], "**", sep = "")
-
-    # tabmat[nrow(tabmat), 5] <- paste("**", tabmat[nrow(tabmat), 5], "**", sep = "")
-
-    colnames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks", "Other")
-    rownames(tabmat) <- c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks", "Other")[1:nrow(tabmat)]
-    rownames(tabmat)[nrow(tabmat)] <- "Blinks"
-
-    return(as.data.frame(tabmat))
-  })
-})
-
-# Create figure for disagreement ratio
-
-ref.disag <- list(c(7, 7, A2017.acc[[3]][[2]]$ratio*100, A2017.acc[[2]][[2]]$ratio*100, 23, 92, 20, 24, 20, 20, 19, 32, 31, 84),
-                  c(11, 11, A2017.acc[[3]][[1]]$ratio*100, A2017.acc[[2]][[1]]$ratio*100, 89, 96, 86, 85, 86, 84, 84, 93, 89, 93),
-                  c(19, 19, A2017.acc[[3]][[3]]$ratio*100, A2017.acc[[2]][[3]]$ratio*100, 64, 95, 61, 62, 61, 59, 59, 70, 67, 92))
-
-disag.df <- data.frame(Image = ref.disag[[1]], Dots = ref.disag[[2]], Video = ref.disag[[3]],
-                       Algorithm = c("coderMN", "coderRA", "gazeHMM-4", "gazeHMM-3", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
-                       stringsAsFactors = F)
-
-disag.df$Algorithm <- factor(disag.df$Algorithm, levels = unique(disag.df$Algorithm))
-
-disag.df <- disag.df %>% pivot_longer(c("Image", "Dots", "Video"), names_to = "Condition", values_to = "Ratio")
-
-disag.plot <- ggplot(disag.df, aes(x = Algorithm, y = Ratio, fill = Condition)) + 
-  geom_col(position = "dodge") + 
-  scale_y_continuous(name = "Disagreement (in %)", limits = c(0, 100), breaks = c(0, 0.25, 0.5, 0.75, 1)*100) +
-  scale_x_discrete(name = "") + coord_flip()
-
-```
-
-#### Event Durations
-The hypothesis that gazeHMM improves the classification performance of eye movement events was investigated by comparing it against other algorithms.
-Therefore, I applied gazeHMM with four events to the @Andersson2017 data set.
-I calculated the RMSD [^3] of event durations as described in the original article by Andersson et al.
-for all algorithms included in the study plus gazeHMM: First, the mean and standard deviation of event durations as well as the number of events were normalized to [0,1] for each coder and algorithm.
-Second, the sum of the root square deviation between each algorithm and the average of the two human coders is computed.
-This metric can only be relatively interpreted [@Andersson2017].
-Table \@ref(tab:tab-rmsd-4-fix) shows that, in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations.
-The high RMSD values were mostly due to a large amount of very short fixations.
-For saccades, gazeHMM has a relatively low RMSD in the image and video condition and a relatively high RMSD in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)).
-Here, gazeHMM overestimated the duration of saccades, as indicated by high mean and SD durations.
-For PSOs, gazeHMM had the highest RMSD in the image and video condition (see Table \@ref(tab:tab-rmsd-4-pso)).
-In the moving dots condition, its RMSD was lower than for NH [@Nystrom2010] but higher than for LNS [@Larsson2013].
-Here, gazeHMM underestimated the number of PSOs and their duration compared to the human coders.
-No other algorithm in the study was designed to parse smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)).
-In sum, gazeHMM did not outperform all other algorithms regarding the RMSD of event durations compared to human coders.
-The duration of fixations and smooth pursuits was largely underestimated and the number of these two events largely overestimated.
-For saccades and PSOs, gazeHMM estimated relatively accurate durations and event numbers, but did not reach the best algorithms.
-
-(ref:tab-rmsd-4-fix) Fixation Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-fix-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
-
-```{r tab-rmsd-4-fix}
-
-apa_table(rmsd.table[[3]][[1]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-rmsd-4-fix)",
-          note = "(ref:tab-rmsd-4-fix-note)")
-
-```
-
-(ref:tab-rmsd-4-sac) Saccade Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-sac-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
-
-```{r tab-rmsd-4-sac}
-
-apa_table(rmsd.table[[3]][[2]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-rmsd-4-sac)",
-          note = "(ref:tab-rmsd-4-sac-note)")
-
-```
-
-(ref:tab-rmsd-4-pso) PSO Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-pso-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
-
-```{r tab-rmsd-4-pso}
-
-apa_table(rmsd.table[[3]][[3]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-rmsd-4-pso)",
-          note = "(ref:tab-rmsd-4-pso-note)")
-
-```
-
-(ref:tab-rmsd-4-sp) Smooth Pursuit Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-sp-note) Durations are displayed in seconds. gazeHMM classified four events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
-
-```{r tab-rmsd-4-sp}
-
-apa_table(rmsd.table[[3]][[4]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-rmsd-4-sp)",
-          note = "(ref:tab-rmsd-4-sp-note)")
-
-```
-
-[^3]: In my proposal, I declared to compute the RMSE instead of RMSD.
-However, since I conducted an in-sample comparison and it was used by @Andersson2017, the term *deviation* is more appropriate (i.e., RMSD).
-
-#### Sample-to-sample Agreement
-Additionally, I compared gazeHMM to the other algorithms by using Cohen's kappa as a measure of sample-to-sample agreement between algorithms and human coders (see Table \@ref(tab:tab-kappa-4)).
-Absolute kappa values were interpreted according to @Landis1977.
-For fixations, absolute kappa values in all three conditions indicated a slight to fair agreement between gazeHMM and human coders.
-Compared to the other algorithms, the agreement of gazeHMM was the lowest in the image condition but the highest in the moving dots and video condition.
-For saccades, gazeHMM showed kappa values that correspond to a moderate to substantial agreement and were relatively high compared to the other algorithms.
-However, gazeHMM never reached the highest agreement.
-For PSOs, absolute kappa values showed slight to fair agreement agreement to human coders.
-It was relatively low in the moving dots condition and second-rate in the image and video conditions compared to NH and LNS.
-The agreement for smooth pursuits was moderate in the moving dots condition, fair in the video condition, and slight in the image condition.
-No other algorithm in the study was designed to detect smooth pursuits.
-In sum, gazeHMM revealed a relatively high sample-to-sample agreement to human coders for saccades.
-For PSOs and smooth pursuits, I found fair to moderate agreement.
-The performance for fixations was absolutely and relatively low.
-
-(ref:tab-kappa-4) Cohen's Kappa Between Human Coders and Algorithms for Different Conditions and Events
-(ref:tab-kappa-4-note) Negative values were set to zero. gazeHMM classified four events. Table adapted from @Andersson2017.
-
-```{r tab-kappa-4}
-
-tab.kappa.4 <- cbind(c("coderMN", "coderRA", "gazeHMM", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
-                     as.data.frame(rbind(ref.acc[1:2,], kappa.table[[3]], ref.acc[3:nrow(ref.acc),])))
-
-apa_table(tab.kappa.4,
-          col.names = c("Algorithm", rep(c("Image", "Dots", "Video"), 4)),
-          col_spanners = list("Fixations" = c(2, 4), "Saccades" = c(5, 7), "PSOs" = c(8, 10), "Smooth pursuit" = c(11, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-kappa-4)",
-          note = "(ref:tab-kappa-4-note)")
-
-```
-
-#### Disagreement and Confusion
-At last, I computed the overall disagreement [^4] as the percentage of all samples taken together that were classified differently by gazeHMM than the human coders.
-Figure \@ref(fig:plot-disag) shows that in the image condition, gazeHMM with four events had a higher overall disagreement by between 10 and 20% than all but two of the other algorithms.
-In the moving dots and video conditions, the disagreement was between 20 and 30% lower than for all other algorithms.
-Notably, gazeHMM's disagreement was similar across conditions (around 45%).
-In summary, gazeHMM had a lower overall disagreement than all other algorithms in the moving dots and video conditions but higher disagreement in the image condition.
-
-[^4]: Even though I proposed to compute the *disagreement ratio* and @Andersson2017 also used that term, the comparison technically involves percentages and not ratios.
-Thus, I decided to use only the term *disagreement* instead.
-
-(ref:plot-disag) Disagreement between algorithms and human coders for different conditions (in %). gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.
-
-```{r plot-disag, fig.cap="(ref:plot-disag)"}
-
-print(disag.plot)
-
-```
-
-(ref:tab-conf-4) Confusion Matrix Between gazeHMM (Rows) and Human Coders (Columns) for Different Conditions
-(ref:tab-conf-4-note) gazeHMM classified four events and blinks.
-
-```{r tab-conf-4}
-
-tab.conf.4 <- cbind(Event = rep(c("Fixations", "Saccades", "PSOs", "Pursuits", "Blinks"), 3), reduce(conf.table[[3]], rbind)) %>%
-  mutate(Condition = rep(c("Moving dots", "Image", "Video"), each = 5)) %>%
-  arrange(Condition) %>%
-  dplyr::select(-Condition)
-
-apa_table(tab.conf.4,
-          stub_indents = list("Image" = c(1:5), "Moving dots" = c(6:10), "Video" = c(11:15)),
-          midrules = c(6, 12),
-          caption = "(ref:tab-conf-4)",
-          note = "(ref:tab-conf-4-note)")
-
-```
-
-Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms in terms of RMSD and sample-to-sample agreement.
-The two criteria rather indicated a moderate performance of gazeHMM.
-However, the overall disagreement was lower for gazeHMM than for all other algorithms in the dynamic conditions but the highest in the static image condition.
-Looking at the confusion matrix in Table \@ref(tab:tab-conf-4), it can be seen that gazeHMM confused fixations and smooth pursuit to a large extend.
-In the static image condition, it overclassified smooth pursuit events, while they were underclassified in the dynamic conditions.
-
-### gazeHMM With Three States
-Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data seems theoretically more appropriate.
-Including only three states could prevent gazeHMM from overclassifying smooth pursuits in the image data and yield better validation results.
-Therefore, I decided to apply gazeHMM with three states to the image data and explore the comparison to other algorithms.
-Table \@ref(tab:tab-rmsd-3) shows the RMSD between gazeHMM and human coders only for the image condition.
-For fixations, gazeHMM shows a comparably low RMSD, but it was relatively high for saccades and PSOs.
-The sample-to-sample agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kappa-3) (only for the image condition).
-For fixations, the absolute kappa indicated substantial agreement and was relatively high compared to the other algorithms.
-The absolute kappa for saccades corresponded to moderate agreement.
-Compared to the other algorithms, it was rather low.
-For PSOs, the absolute agreement was fair, but relatively low compared to the other algorithms.
-Figure \@ref(fig:plot-disag) illustrates the overall disagreement between gazeHMM with three events and the human coders compared to the other algorithms.
-In all three conditions, the disagreement was slightly lower for gazeHMM than for all the other algorithms (5-10% for images and moving dots, around 20% for videos).
-Compared to gazeHMM with four states, the RMSD for fixations decreased while the sample-to-sample agreement increased.
-For saccades and PSOs, the opposite pattern was found.
-The overall disagreement substantially decreased.
-Overall, applying gazeHMM with three states to static data seems more appropriate, since the increase in fixation classification outweighs the decrease in classification of saccades and PSOs.
-There was support for my second hypothesis since gazeHMM had lower overall disagreement to human coders, ranging between 5 and 20% across conditions, than all other algorithms.
-
-(ref:tab-rmsd-3) Event Duration Descriptives and RMSD Between Algorithms and Human Coders for the Image Condition
-(ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.
-
-```{r tab-rmsd-3}
-
-tab.rmsd.3 <- lapply(A2017.rmsd[[2]][[2]], function(x) {
-
-  df <- as.data.frame(t(as.matrix(x)))
-
-  rownames(df)[3] <- "gazeHMM"
-
-  out <- cbind(algorithm = rownames(df), df)
-
-  return(out)
-}) %>% reduce(left_join, by = "algorithm")
-
-apa_table(tab.rmsd.3,
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Fixations" = c(2, 5), "Saccades" = c(6, 9), "PSOs" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-rmsd-3)",
-          note = "(ref:tab-rmsd-3-note)")
-
-```
-
-(ref:tab-kappa-3) Cohen's Kappa Between Human Coders and Algorithms for Different Events in the Image Condition
-(ref:tab-kappa-3-note) Negative values were set to zero. gazeHMM-3 classified three and gazeHMM-4 classified four events. Table adapted from @Andersson2017.
-
-```{r tab-kappa-3}
-
-tab.kappa.3 <- cbind(c("coderMN", "coderRA", "gazeHMM-4", "gazeHMM-3", "CDT", "EM", "IDT", "IKF", "IMST", "IHMM", "IVT", "NH", "BIT", "LNS"),
-                     as.data.frame(rbind(ref.acc[1:2, 1:9], kappa.table[[3]][1:9], kappa.table[[2]], ref.acc[3:nrow(ref.acc), 1:9]))) %>%
-  dplyr::select(c(1, seq(2, 10, 3)))
-
-apa_table(tab.kappa.3,
-          col.names = c("Algorithm", "Fixations", "Saccades", "PSOs"),
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-kappa-3)",
-          note = "(ref:tab-kappa-3-note)")
-
-```
-
-\newpage
-
-# Discussion
-In this report, I presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events.
-The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von Mises and a uniform distribution.
-An HMM classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
-A simulation study revealed that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions.
-Adding smooth pursuits to the model and noise to the generated data were the most critical factors for decreasing recovery.
-In contrast, higher sample sizes yielded a better parameter recovery.
-The variation in starting values did not affect the recovery but the amount of missing data decreased the recovery of hidden states.
-I applied gazeHMM with different numbers of states to benchmark data and compared the model fit, showing that a five-state HMM had consistently most likely generated the data.
-Thus, the model comparison provided evidence against my first hypothesis that the number of events in gazeHMM matches theoretical expectations.
-When comparing gazeHMM with four events against other algorithms and human coders, it did not show the best performance for the duration and number of classified events nor the sample-to-sample agreement across all conditions.
-However, it had a lower overall disagreement for dynamic data but higher disagreement for static data.
-I argued that four states might not be appropriate for static data and additionally applied gazeHMM with three states to them.
-That led to substantially better classifications for fixations, but slightly worse performance for saccades and PSOs.
-With three states, gazeHMM had a lower overall disagreement than all other algorithms.
-I treat this pattern as moderate evidence for my second hypothesis that gazeHMM improves the classification of eye movements.
-Taken together, I only found some evidence for one part of my research question, that gazeHMM improved classifying gaze data but not describing them.
-Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle.
-The overlap can cause the HMM to confuse parameters and hidden states.
-The decrease in parameter recovery (especially for scale parameters) due to noise shows that the overlap is enhanced by more dispersion in the data.
-The scale parameters might be particularly vulnerable to extreme data points.
-The decrease in state recovery for long and many intervals of missing data can be explained by the tendency of HMMs to stay in the same state when no information is available from the data.
-However, the decrease was rather small, leading to the conclusion that the HMM can deal well with missing data.
-Despite these drawbacks, the recovery of the generative model in gazeHMM seems very promising.
-The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).
-The model comparison on the benchmark data suggested that gazeHMM is not the best generative model for eye movement data.
-Even using a model selection criterion that identified correct models successfully in previous studies [@Pohle2017], did not yield the expected pattern.
-There are several explanations for this result:
-*There were more eye movement events present in the data than I expected*.
-Eye movement events can be divided into subevents.
-For example, fixations consist of drift and tremor movements [@Duchowski2017] and PSOs encompass dynamic, static, and glissadic over- and undershoots [@Larsson2013].
-A study on a recently developed HMM algorithm supports this explanation: @Houpt2018 applied the unsupervised BP-AR-HMM algorithm to the @Andersson2017 data set and classified more distinct states than the human coders.
-Some of the states classified by BP-AR-HMM matched the same event coded by humans.
-Since the subevents are usually not interesting for users of classification algorithms, the ability of HMMs to classify might limit their ability to describe eye movements here.
-*Model selection criteria are generally not appropriate for comparing HMMs with different numbers of states*.
-This argument has been discussed in the field of ecology [see @Li2017], where studies found that selection criteria preferred models with more states than expected [similar to the result of this study; e.g., @Langrock2015].
-@Li2017 explain this bias with the simplicity of the submodels in HMMs: Initial state, transition, and response models for each state are usually relatively simple.
-When they do not describe the processes in the respective states accurately, the selection criteria compensate for that by preferring a model with more states.
-Thus, there are not more latent states present in the data, but the submodels of the HMM are misspecified or too simple.
-Correcting for model misspecifications, led to better model recovery in studies on animal movements [@Langrock2015; @Li2017].
-However, @Pohle2017 showed in simulations that the ICL identified the correct model despite several misspecifications.
-It has to be noted that the study by @Pohle2017 only used data generating models with two states.
-*The submodels of gazeHMM were misspecified*.
-@Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models.
-Both situations could apply to gazeHMM and eye movement data.
-Outliers occur frequently in eye-tracking data due to measurement error.
-Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task the data is obtained from [@Langrock2015].
-Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf.
-@Li2017].
-This aspect could indeed oversimplify the complex nature of eye movement data.
-Comparing gazeHMM to other algorithms on benchmark data showed that gazeHMM moderately improved the event classification in agreement with human coders.
-However, the evaluation criteria (RMSD of event durations, sample-to-sample agreement, and overall disagreement) yielded different results.
-The fact that gazeHMM outperformed all other algorithms regarding the overall disagreement can be because it is the only algorithm classifying five events (incl.
-blinks).
-Nevertheless, Cohen's kappa values of 0.72 (saccades - video - four states) or 0.68 (fixations - image - three states) indicate substantial agreement to human coders, especially in light of the maximum reference of ~0.8 from the simulation study.
-At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018a observed substantial differences between coders and within coders over time.
-Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and to find errors in their design.
-
-## Advantages of gazeHMM
-In view of the four proposed goals that gazeHMM should fulfill, I can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they only have minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
-Their default values should be appropriate for most applications.
-A major advantage of gazeHMM is that it does not require human labeled data as input.
-Instead, it estimates all parameters and hidden states from the data.
-Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018a], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
-This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data.
-Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit.
-Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2018], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
-However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events.
-Therefore, I recommend using gazeHMM with four events only for exploratory purposes.
-Without smooth pursuits, I consider gazeHMM's classification as appropriate for application.
-Lastly, its implementation in R using depmixS4 [@Visser2010] should make gazeHMM a tool that is easy to use and customize for individual needs.
-
-## Future Directions
-Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for examples, see @Balakrishnan2009].
-Potential problems of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters).
-Another option to model the correlation could be to include one of the response variables as a covariate of the other.
-Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred.
-This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.
-Third, one solution to diverging results when comparing gazeHMM with different events could be model averaging: Instead of using the maximum posterior state probability of each sample from the preferred model, the probabilities could be weighted according to a model selection criterion (e.g., Schwarz weight) and averaged.
-Then, the maximum averaged probability could be used to classify the samples into events.
-This approach could lead to a more robust classification because it reduces the overconfidence of each competing model and easily adapts to new data [analogous to Bayesian model averaging; @Hinne2020].
-However, the model comparison for gazeHMM often showed extreme weights for a five-state model, which would lead to a very limited influence of the other models in the averaged probabilities.
-Fourth, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data.
-As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements.
-As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.
-Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and auto-correlation functions [see @Zucchini2016].
-Fifth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonably short saccades, gazeHMM could explicitly model the duration of events.
-This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2006].
-Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs.
-This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades.
-Drawback of this extension are higher computational costs and difficulties with including covariates [@Zucchini2016].
-Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM.
-This could potentially be achieved by using different response distributions or parameter optimization methods.
-Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.
-
-## Conclusion
-Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events.
-For smooth pursuits, the classification is not optimal and thus not recommended, yet.
-On one hand, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one.
-On the other hand, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events.
-These difficulties highlight challenges for HMMs in general but also specifically for eye movements.
-Revisiting my research question, I conclude that gazeHMM is suitable for classifying eye movement events but still lacks the ability to accurately describe them.
 

### 2020-10-04 manuscript/supplementary_material_validation.Rmd
add supplementary materials

--- 2020-06-28 manuscript/report_appendix.Rmd
@@
-# Additional Figures and Tables
+---
+title             : "Keeping an Eye on Hidden Markov Models in Gaze Data Classification: Supplementary Material Validation Study"
+shorttitle        : "HMMs in Gaze Classification"
 
-(ref:plot-sim-ex) Simulated data from the HMM with the default parameters of the simulation study.
-Displayed are acceleration (in deg/s$^2$, multiplied by 10000), sample-to-sample (relative) angle (in rad), and velocity (in deg/s, multiplied by 10) over time (in s).
-Acceleration and velocity were multiplied back to their natural scale.
-Note that PSOs (State 3, blue dots) do not only occur after saccades because the HMM parameters were unconstrained.
+floatsintext      : no
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+linenumbers       : no
+mask              : no
+draft             : no
 
-```{r plot-sim-ex, fig.cap="(ref:plot-sim-ex)"}
+bibliography      : "references.bib"
+documentclass     : "apa6"
+classoption       : "man"
+output            : papaja::apa6_pdf
+---
 
-source(here::here("simulation/model_simulation.R"))
+```{r packages, include=FALSE}
 
-resp.sim.true <- list(fix = list(vel = c(3, 0.35), acc = c(3, 0.15)), 
-                     sac = list(vel = c(3, 10), acc = c(3, 3), angle = c(0, 1)),
-                     pso = list(vel = c(3, 1), acc = c(3, 3), angle = c(pi, 1)),
-                     sp = list(vel = c(3, 1), acc = c(3, 0.15), angle = c(0, 1)))
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = "asis")
 
-tr.sim.true <- matrix(0.1/3, 4, 4)
+library(papaja)
+library(tidyverse)
+library(here)
+library(psych)
+library(depmixS4)
+source(here("algorithm/model_helper_functions.R"))
+```
 
-diag(tr.sim.true) <- 0.9
+```{r set graphics theme, include=FALSE}
 
-sim.model <- simulate(HMM_simulate(n = 500, nstates = 4, trueresp = resp.sim.true, truetr = tr.sim.true,
-                          truein = rep(1/4, 4)), seed = 123)
-
-data.frame(State = as.factor(sim.model@states), Velocity = sim.model@response[[1]][[1]]@y*10,
-                           Acceleration = sim.model@response[[1]][[2]]@y*1e4, Angle = sim.model@response[[1]][[3]]@y) %>% 
-  mutate(Time = 1:500*0.002) %>%
-  pivot_longer(c("Velocity", "Acceleration", "Angle"), names_to = "Metric", values_to = "Value") %>%
-  mutate(Metric = ifelse(Metric == "Angle", "Rel. angle", Metric)) %>%
-  ggplot(aes(x = Time, y = Value)) + 
-  facet_grid(rows = vars(Metric), scales = "free_y", switch =  "y") + 
-  geom_path() + geom_point(aes(color = State), size = 0.8) + 
-  scale_y_continuous(name = "") + scale_x_continuous(name = "Time (in s)") +
-  theme(strip.placement = "outside")
+theme_set(theme_apa())
 
@@
 load(here("validation/Ehinger2019_fitted.Rdata"))
+
+# From Wagenmakers & Farrell (2004)
+
+schwarz.weights <- function(bic, na.rm = T) {
+
+  d.bic <- bic - min(bic, na.rm = na.rm) # eq 2
+
+  exp(-0.5 * d.bic)/sum(exp(-0.5 * d.bic), na.rm = na.rm) # eq 4
+
+}
 
@@
 (ref:plot-schwarz-weights-task4) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to task 4 of the Ehinger et al.
-(2019) data set.
+Models were applied to task 4 of the @Ehinger2019 data set.
 Higher weights indicate better model fit.
@@
 (ref:plot-schwarz-weights-task5) Schwarz weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to task 5 of the Ehinger et al.
-(2019) data set.
+Models were applied to task 5 of the @Ehinger2019 data set.
 Higher weights indicate better model fit.
@@
 ```{r calculate integrated complete likelihood criterion, include=FALSE}
+
+load(here("validation/Andersson2017_fitted.Rdata"))
 
@@
 (ref:plot-icl-weights-img) ICL weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the image condition of the Andersson et al.
-(2017) data set.
+Models were applied to the image condition of the @Andersson2017 data set.
 Higher weights indicate better model fit.
@@
 (ref:plot-icl-weights-dots) ICL weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the moving dots condition of the Andersson et al.
-(2017) data set.
+Models were applied to the moving dots condition of the @Andersson2017 data set.
 Higher weights indicate better model fit.
@@
 (ref:plot-icl-weights-vid) ICL weights displayed for each subject and HMMs with different numbers of states.
-Models were applied to the video condition of the Andersson et al.
-(2017) data set.
+Models were applied to the video condition of the @Andersson2017 data set.
 Higher weights indicate better model fit.
@@
 
-(ref:plot-example-classification) Example data from Andersson et al.
-(2017) displayed as x-, and y-coordinates, velocity (in deg/s), acceleration (in deg/s), and sample-to-sample (relative) angle (in rad).
-Colors indicate classification by gazeHMM and symbols by human coder MN.
-Note that the human coder classified all fixations from gazeHMM as smooth pursuits and delayed the separation between the saccade and PSOs.
+\newpage
 
-```{r plot-example-classification, fig.cap="(ref:plot-example-classification)", fig.height=7.5}
+# References
 
-cbind(A2017.fit[[3]][[3]][[4]]$samples, label_MN = as.factor(A2017[[3]][[3]]$label_MN)) %>% 
-  gather(c("x", "y", "vel", "acc", "angle"), key = metric, value = value, factor_key = T) %>% 
-  mutate_at(vars(label, state), as.factor) %>%
-  dplyr::filter(t > 4.0, t <= 4.5) %>% 
-  rename(coderMN = label_MN, gazeHMM = label, Time = t) %>% 
-  mutate(metric = fct_recode(metric, "x-coordinate" = "x", "y-coordinate" = "y", "Velocity" = "vel",
-                             "Acceleration" = "acc", "Rel. angle" = "angle"),
-         coderMN = fct_recode(coderMN, "Fix" = "1", "Sacc" = "2", "PSO" = "3", "SP" = "4", "Blink" = "5", "Other" = "6"),
-         gazeHMM = fct_recode(gazeHMM, "Noise" = "0", "Fix" = "1", "Sacc" = "2", "PSO" = "3", "SP" = "4")) %>%
-  ggplot(aes(x = Time, y = value)) + 
-  facet_wrap(vars(metric), ncol = 1, scales = "free", strip.position = "left") +
-  geom_path() + xlab("Time in s") + ylab("") +
-  geom_point(aes(color = gazeHMM, shape = coderMN), size = 0.8) + 
-  theme(strip.placement = "outside",
-        axis.text = element_text(size = 8))
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
 
-```
-
-(ref:tab-example-trans) Transition Probabability Estimates for Example Data from Andersson et al. (2017)
-
-```{r tab-example-trans}
-
-tab.exp.tr <- cbind(as.character(1:4), matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[5:20], 3), 4, 4, byrow = T))
-
-apa_table(tab.exp.tr,
-          col.names = c("From", as.character(1:4)),
-          col_spanners = list("To" = c(2,5)),
-          caption = "(ref:tab-example-trans)")
-
-```
-
-(ref:tab-example-resp) Response Parameter Estimates for Example Data from Andersson et al. (2017)
-
-```{r tab-example-resp}
-
-tab.exp.resp <- cbind(as.character(1:4), matrix(round(backtrans(getpars(A2017.fit[[3]][[3]][[4]]$model))[21:44], 2), 4, 6, byrow = T))
-
-tab.exp.resp[1, c(6,7)] <- NA
-
-apa_table(tab.exp.resp,
-          col.names = c("State", "Shape (vel)", "Scale (vel)", "Shape (acc)", "Scale (acc)", "Mean", "Con"),
-          format.args = list(na_string = "-"),
-          caption = "(ref:tab-example-resp)")
-
-```
+<div id = "refs"></div>
+\endgroup

### 2020-10-04 simulation/parameter_recovery_analysis.R
remove old simulation results stuff

@@ -1,672 +1,0 @@
-### Parameter recovery analysis ###
-
-# Author: Malte Lüken
-# Date: 11.05.2020
-
-
-library(tidyverse)
-library(depmixS4)
-
-
-# Load data
-
-for (part in 1:4) {
-  
-  load(file = paste("simulation/part", part, ".Rdata", sep = ""))
-  
-}
-
-linkinv <- function(eta,base) {
-  linv <- function(eta,base) {
-    pp <- numeric(length(eta))
-    if(any(is.infinite(eta)) || any(eta > log(.Machine$double.xmax)) || any(eta < log(.Machine$double.xmin))) {
-      pp[which(is.infinite(eta))] <- 1
-      pp[which(eta > log(.Machine$double.xmax))] <- 1 # change this to something better!
-    } else {
-      expb <- exp(eta)
-      sumb <- sum(expb)
-      pp[base] <- 1/sumb
-      pp[-base] <- expb[-base]/sumb
-    }
-    return(pp)
-  }
-  if(is.matrix(eta)) {
-    if(ncol(eta)==1) {
-      pp <- as.matrix(apply(eta,1,linv,base=base)) # fixes problem with column matrix eta
-    } else pp <- t(apply(eta,1,linv,base=base)) 	
-  } else {
-    pp <- linv(eta,base)
-  }
-  return(pp)
-}
-linkfun <- function(p, base) {
-  lfun <- function(p, base) {
-    p <- p/sum(p)
-    beta <- numeric(length(p))
-    if (any(p == 1)) 
-      beta[which(p == 1)] = Inf
-    else beta[-base] <- log(p[-base]/p[base])
-    return(beta)
-  }
-  if (is.matrix(p)) {
-    beta <- t(apply(p, 1, lfun, base = base))
-  }
-  else {
-    beta <- lfun(p, base)
-  }
-  return(beta)
-}
-
-
-# Function to transform parameters to normal scale
-
-backtrans <- function(x) {
-  
-  out <- x
-  
-  nms <- names(x)
-  
-  out[str_detect(nms, "(Intercept)")] <- as.vector(apply(matrix(x[str_detect(nms, "(Intercept)")],
-                                                        ncol = sqrt(length(x[str_detect(nms, "(Intercept)")])),
-                                                        byrow = T), 1, linkinv, base = 1))
-  
[…]

### 2020-10-04 manuscript/paper_BRM.Rmd
some figure layout changes

@@
 authornote: |
-  We would like to thank Daan van Renswoude and Maartje Raijmakers for their helpful comments on earlier versions of this paper.
+  We would like to thank Daan van Renswoude, Maartje Raijmakers, and Maximilian Maier for their helpful comments on earlier versions of this paper.
 Furthermore, we would like to acknowledge feedback by Karel Veldkamp and Phil Norberts in the early stage of this project.
@@
 keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
-wordcount         : "8861"
+wordcount         : "9605"
 
@@
 
-print(plots.rmsd.234[[1]][[1]] + coord_flip())
+print(plots.rmsd.234[[1]][[1]] + coord_flip() + theme(axis.text.x = element_text(size = 10)))
 
@@
 
-print(plots.rmsd.234[[1]][[2]] + coord_flip())
+print(plots.rmsd.234[[1]][[2]] + coord_flip() + theme(axis.text.x = element_text(size = 8)))
 
@@
 
-print(plots.rmsd.234[[1]][[3]] + coord_flip())
+print(plots.rmsd.234[[1]][[3]] + coord_flip() + theme(axis.text.x = element_text(size = 9)))
 

### 2020-11-01 manuscript/paper_BRM.Rmd
minor fixes according to Simon's and Ingmar's comments + color palette

@@
 library(gridExtra)
+library(viridis)
 source(here("algorithm/preprocessing_helper_functions.R"))
@@
 Despite their accuracy, the requirement of labeled training data is a disadvantage of supervised methods because the labeling process can easily become costly and time consuming [@Zemblys2019].
+Supervised methods also (implicitly) treat human labeled training data as a gold standard, an assumption that may be unwarranted [see discussion in @Hooge2018].
 In contrast, *unsupervised* classification algorithms do not require labeled training input.
@@
 Moreover, generative models can challenge common theoretical assumptions and bring up new research questions [@Epstein2008].
-For example, they might suggest that oscillations also occur before saccadic eye movements [as mentioned in @Nystrom2010] or that the assumption of discrete eye movements does not hold [as discussed in @Andersson2017].
+For example, they might suggest that oscillations also occur before saccadic eye movements [as mentioned in @Nystrom2010] or that the assumption that eye movements are discrete events (e.g., saccades and PSOs cannot overlap) does not hold [as discussed in @Andersson2017].
 One class of generative models that are used in eye movement classification are HMMs.
@@
 @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR).
-The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and relative angle (named NSLR-HMM).
+The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and change in angle (named NSLR-HMM).
 Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]).
@@
 NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by the CNN algorithm in @Bellet2019 [@Pekkanen2017].
-In sum, HMMs seem to be a promising method for classifying eye movements. Nevertheless, the existing HMM algorithms each have their weaknesses that could be improved upon.  
+In sum, HMMs seem to be a promising method for classifying eye movements. Nevertheless, the existing HMM algorithms each have at least one aspect in which they could be improved.  
 First, I-HMM relies on setting an appropriate threshold to determine the initial classification, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008].
@@
 Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].
-For this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms.
+In this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the aforementioned aspects of previous algorithms.
 The development was guided by four goals: (a) the algorithm should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuits, and PSOs; (d) it should confirm rather than explore the presence of events in the data.
-The following section will describe gazeHMM and the underlying generative model in detail.
-Then, we will present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding.
-Finally, we will discuss these results and propose directions in which gazeHMM and other HMM algorithms could be improved.
-
-# The Algorithm
+The following section describes gazeHMM and the underlying generative model in detail.
+Then, we present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding.
+Finally, we discuss these results and propose directions in which gazeHMM and other HMM algorithms could be improved.
+
+# Developing gazeHMM
 As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most eye movement event classification algorithms consist of three steps [cf.
@@
 Often, a filtering or smoothing procedure is applied to the data, before or after the tranformation, to separate the gaze signal from noise and artifacts [@Spakov2012].
-It follows the *classification*, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
+Then follows the *classification*, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
 Lastly, as part of the *postprocessing*[^1], the algorithm decides which candidates to accept, relabel, or merge [@Hessels2017; @Komogortsev2010].
@@
 However, for parametric methods, we think calling step two "classification" is more appropriate since the probabilistic classification is done here.
-Step three usually consist of some heuristic relabeling and correcting for classification errors.
+Step three usually consists of some heuristic relabeling and correcting for classification errors.
 
@@
   geom_line(aes(y = y.va)) + 
-  geom_point(aes(y = x.va), color = "darkred", size = 1) + 
-  geom_point(aes(y = y.va), color = "darkblue", size = 1) +
+  geom_point(aes(y = x.va), size = 1) + 
+  geom_point(aes(y = y.va), size = 1) +
   scale_y_continuous(name = "Gaze position (deg)", limits = c(-5, 10), breaks = pretty(c(-5, 10)))
@@
     xlab("") + geom_line(color = "black") + ggtitle(letters[p]) + 
+    scale_color_viridis_d() +
     theme(legend.position = "none")
@@
 Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees.
-Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades, they follow the same direction [@Larsson2013; @Leigh2015].
+Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades, they follow a similar direction [@Larsson2013; @Leigh2015].
 Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.
@@
 However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation).
-Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean.
-Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties.
+Second, the distributions of both signals appear to be positively skewed conditionally on the states, and third, to have variances increasing with their mean.
+Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that has these three properties.
 In gazeHMM, we use gamma distributions with a shape and scale parametrization for this purpose:
@@
 We assume that the best fitting distribution will depend on the task, eye-tracker, and individual (see discussion).
+We emphasize that gazeHMM does not critically depend on the choice of distribution and other distributions than the gamma can be readily included in the model.
 To model the sample-to-sample angle, we pursued a novel approach in gazeHMM: A mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution: 
@@
 For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we generated 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
-These samples sizes roughly corresponded to small, medium, and large eye-tracking data sets for a single participant and trial.
+These samples sizes roughly corresponded to small, medium, and large eye-tracking data sets for a single participant and trial (e.g., with a frequency of 500 Hz, the sample sizes would correspond to recorded data with lengths of 1 s, 5 s, and 20 s, respectively).
 To simulate noise, we replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
@@
     scale_y_discrete(name = "Estimated parameter", labels = names.pars.est) +
-    scale_fill_distiller(breaks = c(0, 0.1, 0.5, 1), palette = "Spectral")
+    scale_fill_viridis_c(breaks = c(0, 0.1, 0.5, 1))
 
@@
     scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2, 2.5, 3, 4, 5)) +
+    scale_color_viridis_d() +
     geom_hline(yintercept = 0.1, linetype = "dashed") +
@@
 
-  p <- p + scale_color_discrete(name = label.cond, labels = names.cond)
+  p <- p + scale_color_viridis_d(name = label.cond, labels = names.cond)
 
@@
     scale_x_discrete(name = "k (number of states)")  + 
-    scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1))
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1)) +
+    scale_color_viridis_d()
 
@@
 
-  p <- p + scale_color_discrete(name = label.cond, labels = names.cond) +
+  p <- p + scale_color_viridis_d(name = label.cond, labels = names.cond) +
     scale_x_continuous(name = x.name) + 
@@
 We used the original data from the study but removed two recordings from the moving dots condition because they majorly contained samples labeled as "other" or blinks.
-Moreover, the recordings could not be matched to the results obtained by @Andersson2017.
+Moreover, the recordings could not be matched to the results obtained by @Andersson2017 [^4].
 The second data set was published in @Ehinger2019 and has to our knowledge not yet been used for validation.
@@
 For simplicity, we only used data obtained by the EyeLink (SR Research Ltd., Ontario, Canada) eye-tracker.
+
+[^4]: Two recordings from the moving dots conditions were substantially longer than the other recordings in the condition and contained more samples than were classified by the algorithms in the study by @Andersson2017.
+Since no sample indices were available in the data set, we could not match samples from the two recordings to the labels assigned by the algorithms and therefore decided to remove them from the analysis.
+We do not expect the conclusions of our analyses to depend on these two data sets.
 
@@
 Regarding the @Ehinger2019 data set, we assumed that the one-state model (only fixation) would show the highest weights for both tasks.
-The algorithm was applied separately to every subject, condition, or task.
+The algorithm was applied separately to every subject, condition/task.
 For the @Andersson2017 data set, all generative models were successfully fitted, whereas for the @Ehinger2019 data set, it was only 780 out of 900 models (87%, 60 models per task).
@@
   scale_y_discrete(name = "Subject") +
-  scale_fill_distiller(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")
+  scale_fill_viridis_c(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1))
 
@@
 
+disag.df$Ratio[disag.df$Algorithm == "gazeHMM-4" & disag.df$Condition == "Image"] <- disag.df$Ratio[disag.df$Algorithm == "gazeHMM-3" & disag.df$Condition == "Image"]
+disag.df$Algorithm[disag.df$Algorithm == "gazeHMM-4"] <- "gazeHMM-3/4"
+
+disag.df <- disag.df %>%
+  dplyr::filter(Algorithm != "gazeHMM-3") %>%
+  group_by(Algorithm) %>%
+  mutate(MRatio = mean(Ratio)) %>%
+  ungroup() %>%
+  arrange(MRatio)
+
 disag.df$Algorithm <- factor(disag.df$Algorithm, levels = unique(disag.df$Algorithm))
-
-disag.df$Ratio[disag.df$Algorithm == "gazeHMM-3" & disag.df$Condition != "Image"] <- NA
-disag.df$Ratio[disag.df$Algorithm == "gazeHMM-4" & disag.df$Condition == "Image"] <- NA
 
@@
   scale_y_continuous(name = "Disagreement (in %)", limits = c(0, 100), breaks = c(0, 0.25, 0.5, 0.75, 1)*100) +
-  scale_x_discrete(name = "") + coord_flip()
+  scale_x_discrete(name = "") + 
+  scale_fill_viridis_d() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
@@
 
-(ref:plot-disag) Disagreement between algorithms and human coders for different conditions (in %). gazeHMM-3 classified three and gazeHMM-4 classified four events.
+(ref:plot-disag) Disagreement between algorithms and human coders for different conditions (in %).
+gazeHMM-3/4 classified three events for image data and four events for moving dots/video data.
+Algorithms are displayed in order according to mean disagreement over conditions (least/left to highest/right).
 
@@
   geom_point(aes(color = gazeHMM, shape = coderMN), size = 0.8) + 
+  scale_color_viridis_d() +
   theme(strip.placement = "outside",
@@
 ## Conclusion
-Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events.
+In the previous sections, we developed and tested a generative, HMM-based algorithm called gazeHMM.
+Both a simulation and validation study showed that gazeHMM is a suitable algorithm for classifying eye movement events.
 For smooth pursuits, the classification is not optimal and thus not yet recommended.
-On one side, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one.
+On one side, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled training data being the most important one.
 On the other side, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events.

### 2020-11-01 manuscript/supplementary_material_simulation.Rmd
minor fixes according to Simon's and Ingmar's comments + color palette

@@
 library(psych)
+library(viridis)
 ```
@@
     scale_x_continuous(name = x.name) + 
-    scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1))
+    scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1)) +
+    scale_color_viridis_d()
 

### 2020-11-01 manuscript/supplementary_material_validation.Rmd
minor fixes according to Simon's and Ingmar's comments + color palette

@@
 library(depmixS4)
+library(viridis)
 source(here("algorithm/model_helper_functions.R"))
@@
   scale_y_discrete(name = "Subject") +
-  scale_fill_distiller(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), 
-                       palette = "Spectral")
+  scale_fill_viridis_c(name = "Schwarz\nweight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1))
 
@@
     scale_y_discrete(name = "Subject") +
-    scale_fill_distiller(name = "ICL weight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), palette = "Spectral")
+    scale_fill_viridis_c(name = "ICL weight", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1))
 

### 2020-11-10 manuscript/paper_BRM.Rmd
add noise intervals and final adjustments

@@
 ---
-title             : "Keeping an Eye on Hidden Markov Models in Gaze Data Classification"
-shorttitle        : "HMMs in Gaze Classification"
+title             : "Classifying Gaze Data With an Unsupervised Generative Hidden Markov Model"
+shorttitle        : "Classifying Gaze Data with an HMM"
 
@@
     corresponding : yes
-    address       : "Speelmanstraat 7 1063ZC Amsterdam, The Netherlands"
+    address       : "Postbus 15906, 1001 NK Amsterdam, The Netherlands"
     email         : "malte_lueken@arcor.de"
@@
 Parsing the events is typically done by algorithms.
-Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere several weaknesses.
-To adress these weaknesses, we developed gazeHMM, an algorithm that uses an HMM as a generative model, has no critical parameters to be set by users, and does not require human coded data as input.
+Previous algorithms have successfully used hidden Markov models (HMMs) for classification but can still be improved in several aspects.
+To address these aspects, we developed gazeHMM, an algorithm that uses an HMM as a generative model, has no critical parameters to be set by users, and does not require human coded data as input.
 The algorithm classifies gaze data into fixations, saccades, and optionally postsaccadic oscillations and smooth pursuits.
@@
 The simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences.
-Critical exceptions for good recovery were including the smooth pursuit state in the generative model and noisy data.
-The performance on benchmark data was assessed by model comparison and agreement of the classification to human coding compared to other algorithms.
-The model comparison yielded preferred generative models for the benchmark data with more events than we expected.
-Nevertheless, gazeHMM showed relatively high similarity to human coding and outperformed the other algorithms in some agreement criteria.
+Critical exceptions for parameter recovery were including the smooth pursuit state in the generative model and even low noise levels.
+We assessed the performance on benchmark data by model comparing generative models with different events and agreement of the classification to human coding compared to other algorithms.
+The model comparison preferred more events in the generative models than we expected.
+Nevertheless, gazeHMM showed relatively high similarity to human coding and outperformed other algorithms in this aspect.
 For dynamic stimuli, our algorithm tended to switch rapidly between fixations and smooth pursuits.
-Thus, we recommend using gazeHMM for classifiying smooth pursuits only in explorative settings.
-Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.
+Thus, we recommend using gazeHMM for parsing smooth pursuits only in explorative settings.
+Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to classify smooth pursuits more accurately.
 
 keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
-wordcount         : "9605"
+wordcount         : "10095"
 
@@
 On the other hand, many parametric methods have been developed recently.
-Some of them require human labeled training data as input and can therefore be termed as *supervised* [@Hastie2017].
+Some of them require human-labeled training data as input and can therefore be termed as *supervised* [@Hastie2017].
 For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders [see also @Zemblys2018; @Zemblys2019; @Startsev2019].
-Despite their accuracy, the requirement of labeled training data is a disadvantage of supervised methods because the labeling process can easily become costly and time consuming [@Zemblys2019].
-Supervised methods also (implicitly) treat human labeled training data as a gold standard, an assumption that may be unwarranted [see discussion in @Hooge2018].
+Despite their accuracy, the requirement of labeled training data is a disadvantage of supervised methods because the labeling process can easily become costly and time-consuming [@Zemblys2019].
+Supervised methods also (implicitly) treat human-labeled training data as a gold standard, an assumption that may be unwarranted [see discussion in @Hooge2018].
 In contrast, *unsupervised* classification algorithms do not require labeled training input.
-Instead they learn parameters from the characteristics of the data themselves [@Hastie2017].
+Instead, they learn parameters from the characteristics of the data themselves [@Hastie2017].
 In consequence, they are also more flexible in classifying data from different individuals, tasks, or eye-trackers [e.g., @Houpt2018; @Hessels2017].
@@
 Classifiers with generative assumptions have the advantage that their parameters can be easily interpreted in terms of the underlying theory.
-In the context of eye movements, they can also help explain or confirm observed phenomena: For instance, their parameters can indicate that oscillations only occur after but not before saccades.
-When the goal is to understand eye movement events and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods.
+In the context of eye movements, they can also help to explain or confirm observed phenomena: For instance, their parameters can indicate that oscillations only occur after but not before saccades.
+When the goal is to understand eye movement events and improve their classification based on this understanding, this aspect is an advantage over non-parametric or supervised methods.
 Moreover, generative models can challenge common theoretical assumptions and bring up new research questions [@Epstein2008].
@@
 The following section describes gazeHMM and the underlying generative model in detail.
-Then, we present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding.
+Then, we present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms concerning the agreement to human coding.
 Finally, we discuss these results and propose directions in which gazeHMM and other HMM algorithms could be improved.
@@
 @Hessels2017]: During *preprocessing*, the raw gaze position is transformed into a metric of position and/or time.
-Often, a filtering or smoothing procedure is applied to the data, before or after the tranformation, to separate the gaze signal from noise and artifacts [@Spakov2012].
+Often, a filtering or smoothing procedure is applied to the data, before or after the transformation, to separate the gaze signal from noise and artifacts [@Spakov2012].
 Then follows the *classification*, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events.
@@
 
-[^1]: @Hessels2017 called step two the *search rule* and step three the *classifcation rule*.
+[^1]: @Hessels2017 called step two the *search rule* and step three the *classification rule*.
 For non-parametric methods, this distinction might be accurate.
@@
 
-(ref:plot-algorithm-workflow) Example work flow for eye movement event classification algorithms: (a) the raw gaze signal in x (dark red) and y (dark blue) coordinates; (b) the raw gaze signal is filtered and transformed into a velocity signal; (c) samples are classified as events (indicated by colors), and (d) relabeled.
+(ref:plot-algorithm-workflow) Example workflow for eye movement event classification algorithms: (a) the raw gaze signal in x (dark red) and y (dark blue) coordinates; (b) the raw gaze signal is filtered and transformed into a velocity signal; (c) samples are classified as events (indicated by colors), and (d) relabeled.
 Sequences of samples belonging to the same event are merged (indicated by black segments).
@@
   geom_line(aes(y = y.va)) + 
-  geom_point(aes(y = x.va), size = 1) + 
-  geom_point(aes(y = y.va), size = 1) +
+  geom_point(aes(y = x.va), size = 2) + 
+  geom_point(aes(y = y.va), size = 2) +
   scale_y_continuous(name = "Gaze position (deg)", limits = c(-5, 10), breaks = pretty(c(-5, 10)))
@@
   geom_line() + 
-  geom_point(size = 1) +
+  geom_point(size = 2) +
   scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
@@
   geom_line() + 
-  geom_point(aes(color = event.pre), size = 1) +
+  geom_point(aes(color = event.pre), size = 2) +
   scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
@@
   geom_line() + 
-  geom_point(aes(color = event.post), size = 1) +
+  geom_point(aes(color = event.post), size = 2) +
   scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
@@
 In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.
-Theoretically, these three metrics should clearly separate eye movement events.
+Theoretically, these three metrics should separate eye movement events.
 Fixations typically inherit samples with low velocity and acceleration [@Larsson2013].
@@
 In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction.
-PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013].
+PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low-velocity events [@Larsson2013].
 They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017].
@@
 HMMs can be described by three submodels: An initial state model, a transition model, and a response model.
-The initial state model contains probabilites for the first state of the hidden sequence $\rho_i = P(S_1=i)$, with *i* denoting the hidden state.
+The initial state model contains probabilities for the first state of the hidden sequence $\rho_i = P(S_1=i)$, with *i* denoting the hidden state.
 In gazeHMM, the initial states are modeled by a multinomial distribution.
-The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM.
+The evolution of the sequence is in turn described by the transition model, which comprises the probabilities for transitioning between different states in the HMM.
 Typically, probabilities to transition from state *i* to *j*, $a_{ij} = P(S_{t+1}=j|S_{t}=i)$, are expressed in matrix form [@Visser2011]:
@@
 Thus, the uniform distribution should distinguish fixations from the other events.
-All three submodels taken together, the joint likelihood of the observed data and hidden states can be expressed as:
+Taking all three submodels together, the joint likelihood of the observed data and hidden states can be expressed as:
 $$L(X, Y, Z, S|\mathbf{\lambda})=\rho_{S_{1}}f_{S_{1}}(X_1)f_{S_{1}}(Y_1)f_{S_{1}}(Z_1)\prod_{t=1}^{T-1}a_{S_{t}S_{t+1}}f_{S_{t+1}}(X_{t+1})f_{S_{t+1}}(Y_{t+1})f_{S_{t+1}}(Z_{t+1}),$$
-with $\mathbf{\lambda}$ denoting the vector containing the initial state and transtition probabilities as well as the response parameters.
+with $\mathbf{\lambda}$ denoting the vector containing the initial state and transition probabilities as well as the response parameters.
 By summing over all possible state sequences, the likelihood of the data given the HMM parameters becomes [@Visser2011]:
@@
 For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [i.e., eye movement events; @Visser2011].
-When evaluating the likelihood for missing data, gazeHMM integrates over all possible values, which results in a probability density of one.
+When evaluating the likelihood of missing data, gazeHMM integrates over all possible values, which results in a probability density of one.
 The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability.
 Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988].
-The implementation in depmixS4 allows to include time varying covariates for each parameter in the HMM.
+The implementation in depmixS4 allows to include time-varying covariates for each parameter in the HMM.
 In gazeHMM, no such covariates were included and thus, only intercepts were estimated for each parameter.
@@
 The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages \texttt{signal} [@Ligges2015] to compute velocity and acceleration signals, \texttt{depmixS4} [@Visser2010] for the HMM, \texttt{CircStats} [@Lund2018] for the von Mises distribution, and \texttt{BB} [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization.
-The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).
+The algorithm is available on GitHub (<https://github.com/maltelueken/gazeHMM>).
 
@@
 To assess how well the HMM recovers parameters and state sequences, we conducted a simulation study.
-The design and analysis of the study were preregistered on the Open Science Framework (https://doi.org/10.17605/OSF.IO/VDJGP).
+The design and analysis of the study were preregistered on the Open Science Framework (<https://doi.org/10.17605/OSF.IO/VDJGP>).
 This section is majorly copied from the preregistration (with adapted tenses).
 The study was divided in four parts.
-Here, we only report the first two parts, the other two parts, which address starting values and missing data, can be found in the supplementary material (www.github.com/maltelueken/gazeHMM_validation).
+Here, we only report the first two parts, the other two parts, which address starting values and missing data, can be found in the supplementary material (<https://github.com/maltelueken/gazeHMM_validation>).
 The HMM repeatedly generated data with a set of parameters (true parameter values).
@@
 This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values.
-Mean parameters of the von Mises distribution always started at their true values.
+The mean parameters of the von Mises distribution always started at their true values.
 
@@
 For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were manipulated, respectively.
-For each parameter, the HMM generated 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps.
+For each parameter, the HMM generated 100 data sets with $N = 2500$ samples, and the parameter varied in a specified interval in equidistant steps.
 This resulted in $100 \times (10+15+20) = 4500$ recoveries.
@@
 For the transition probabilities, we only simultaneously changed the probabilities for staying in the same state (diagonals of the transition matrix) to reduce the complexity of the simulation.
-The left over probability mass was split evenly between the probabilities for switching to a different state (per row of the transition matrix).
+The leftover probability mass was split evenly between the probabilities for switching to a different state (per row of the transition matrix).
 Moreover, we did not modify the mean parameters of the von Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.
@@
 For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we generated 100 data sets ($100 \times 3 \times 3 = 900$ recoveries).
-These samples sizes roughly corresponded to small, medium, and large eye-tracking data sets for a single participant and trial (e.g., with a frequency of 500 Hz, the sample sizes would correspond to recorded data with lengths of 1 s, 5 s, and 20 s, respectively).
+These sample sizes roughly match small, medium, and large eye-tracking data sets for a single participant and trial (e.g., with a frequency of 500 Hz, the sample sizes would correspond to recorded data with lengths of 1 s, 5 s, and 20 s, respectively).
 To simulate noise, we replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets.
@@
   p <- ggplot(data = data.long, aes(x = par.varied, y = par.est, fill = RMdSPD)) +
-    geom_tile() + facet_grid(cols = vars(state.varied), rows = vars(state.est)) +
+    geom_tile(color = "black") + facet_grid(cols = vars(state.varied), rows = vars(state.est)) +
     scale_x_discrete(name = "Varied parameter", labels = names.pars.varied) +
@@
   p <- ggplot(data = data.long, aes(x = par.est, y = RMdSPD, color = cond)) +
-    geom_point(position = position_dodge(0.25)) + facet_wrap(vars(state.est)) +
+    geom_point(position = position_dodge(0.25)) + facet_grid(rows = vars(state.est)) +
     scale_x_discrete(name = "Estimated parameter", labels = names.pars.est) +
@@
     facet_grid(rows = vars(From), cols = vars(To), labeller = label_both) +
-    geom_point() + geom_smooth(method = "lm") +
+    geom_point() + geom_smooth(method = "lm", color = "#FDE725FF") +
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
@@
     facet_wrap(vars(par), scales = "free", labeller = label_bquote(.(parnames[[par]]))) +
-    geom_point() + geom_smooth(method = "lm") +
+    geom_point() + geom_smooth(method = "lm", color = "#FDE725FF") +
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
@@
 
+  noise.scale <- paste(1:5, round(1/seq(0.1, 10, length.out = 5), 2), sep = "\n")
+
   data.long <- x %>%
@@
     label.cond <- "N"
-    x.name <- bquote(tau["noise"])
+    x.name <- "Noise"
   } else {
@@
   p <- p + scale_color_viridis_d(name = label.cond, labels = names.cond) +
-    scale_x_continuous(name = x.name) + 
+    scale_x_continuous(name = x.name, labels = noise.scale) + 
     scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1))
@@
 ### Parameter Variation
-In the first part of the simulation, we examined how varying the parameters[^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences.
+In the first part of the simulation, we examined how varying the parameters[^2] in the HMM affected the deviation of estimated parameters and the accuracy of estimated state sequences.
 For the two-state HMM, the recovery of parameters and states was nearly perfect (all RMdSPDs < 0.1, intercepts and slopes of regression lines almost zero and one, respectively, and Cohen's kappa close to 1).
-We therefore chose to include the respective figures in the supplementary material.
+Therefore, we chose to include the respective figures in the supplementary material.
 For the HMM with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3).
@@
 Dashed lines refer to perfect recovery.
+$\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+Parameter subscripts indicate states and eye movement metrics.
 
@@
 Crosses represent outliers.
+$a_{i=j}$ is the probability to stay in the same state, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
-Top facet labels denote to which state estimated parameters belong.
-$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+Right facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 
-print(plots.rmsd.234[[1]][[1]] + coord_flip() + theme(axis.text.x = element_text(size = 10)))
+print(plots.rmsd.234[[1]][[1]] + theme(axis.text.x = element_text(size = 10)))
 
@@
 (ref:plot-acc-2) Cohen's kappa depending on the variation of noise added to the data generated by the HMM.
+The upper labels on the x-axis indicate $\tau_{noise}$ and the lower labels $\kappa_{noise}$.
 Colours indicate different sizes of generated data.
 Top facet labels indicate the number of states in the HMM.
+$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
 
@@
 
-print(plots.rmsd.234[[1]][[2]] + coord_flip() + theme(axis.text.x = element_text(size = 8)))
+print(plots.rmsd.234[[1]][[2]] + theme(axis.text.x = element_text(size = 10)))
 
@@
 In general, the HMM recovered parameters well despite noise being added to the data.
-However, in the four-state model, the parameter recovery for states one and four substantially decreased.
-In the three- and four-state models, scale parameters of gamma distributions were badly recovered.
+However, in the four-state model, the parameter recovery for states one and four substantially decreased (even for low amounts of noise, see supplementary material).
+In the three- and four-state models, scale parameters of gamma distributions were badly recovered (also even for low noise levels, see supplementary material).
 Increasing the sample size in the HMM slightly improved the recovery of most parameters.
-The state recovery of the model was slightly lowered when more states were included, but it was neither affected by the noise variability $\tau_{noise}$ nor the sample size.
+The state recovery of the model was slightly lowered when more states were included, but it was neither affected by the noise level nor the sample size.
 
@@
 
-print(plots.rmsd.234[[1]][[3]] + coord_flip() + theme(axis.text.x = element_text(size = 9)))
+print(plots.rmsd.234[[1]][[3]] + theme(axis.text.y = element_text(size = 8)))
 
@@
 # Validation Study
-To validate gazeHMM, we applied the algortihm on two benchmark data sets.
+To validate gazeHMM, we applied the algorithm on two benchmark data sets.
 As starting values, we used $\rho=1/k$ for the initial state model as well as $a_{i=j}=0.9$ and $a_{i\neq j}=0.1/k$ for the transition model.
@@
 It contains eye-tracking data from three conditions: A static condition, where subjects had to look freely at images, and two dynamic conditions, where they had to follow a constantly moving dot or objects in a video.
-The data was sampled with 500 Hz and two human coders (MN and RA) labeled them as belonging to six different eye movement events: Fixation, saccade, PSO, smooth pursuit, blink, or other.
+The data were sampled with 500 Hz and two human coders (MN and RA) labeled them as belonging to six different eye movement events: Fixation, saccade, PSO, smooth pursuit, blink, or other.
 @Andersson2017 used the data to compare 10 different classification algorithms.
@@
 The second data set was published in @Ehinger2019 and has to our knowledge not yet been used for validation.
-Here, we only took tasks four and five out of 10 tasks because these are qualitatively different to the first data set.
+Here, we only took tasks four and five out of 10 tasks because these are qualitatively different from the first data set.
 In task four, subjects were instructed to fixate a central target for 20 s.
-Task 5 was set up similarly, but subjects had to blink when they heared one out of seven beeps (with a beep duration of 100 ms and 1.5 s intervals in between).
+Task 5 was set up similarly, but subjects had to blink when they heard one out of seven beeps (with a beep duration of 100 ms and 1.5 s intervals in between).
 Eye movements were recorded with 250-500 Hz.
@@
 ## Data Analysis
-A successful validation of gazeHMM was determined by two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets.
+Successful validation of gazeHMM was determined by two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets.
 The fits of the generative models were compared using Schwarz weights [@Wagenmakers2004], a conversion of the BIC [@Schwarz1978] into model weights.
@@
 The algorithm was applied separately to every subject, condition/task.
-For the @Andersson2017 data set, all generative models were successfully fitted, whereas for the @Ehinger2019 data set, it was only 780 out of 900 models (87%, 60 models per task).
+For the @Andersson2017 data set, all generative models were successfully fitted, whereas, for the @Ehinger2019 data set, it was only 780 out of 900 models (87%, 60 models per task).
 Second, we compared gazeHMM to other algorithms and human coders.
@@
   mutate_at(c("subject", "model", "condition"), as.factor) %>%
-  ggplot(aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
+  ggplot(aes(x = model, y = subject)) + geom_tile(aes(fill = weight), color = "black") +
   facet_grid(cols = vars(condition)) + 
@@
 A recent model recovery study showed that the BIC tended to prefer overly complex HMMs when they were misspecified [e.g., the conditional independence assumption was violated; @Pohle2017].
-Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2000] performed better at choosing the correct data generating model.
+Instead, the integrated completed likelihood (ICL) criterion [@Biernacki2000] performed better at choosing the correct data-generating model.
 Therefore, we post hoc computed the weighted ICL criterion (analogous to Schwarz weights) for the models fitted to the @Andersson2017 data set.
@@
 disag.plot <- ggplot(disag.df, aes(x = Algorithm, y = Ratio, fill = Condition)) + 
-  geom_col(position = "dodge") + 
+  geom_col(position = "dodge", color = "black") + 
   scale_y_continuous(name = "Disagreement (in %)", limits = c(0, 100), breaks = c(0, 0.25, 0.5, 0.75, 1)*100) +
@@
 It can be seen that gazeHMM classified a much larger number of fixations with very short durations.
-For saccades, gazeHMM had a relatively high RMSD for the static condition, but it had the lowest RMSD for the moving dots condition, and a moderate value for the video condition (see Table \@ref(tab:tab-rmsd-4-sac)).
+For saccades, gazeHMM had a relatively high RMSD for the static condition but the lowest RMSD for the moving dots condition, and a moderate value for the video condition (see Table \@ref(tab:tab-rmsd-4-sac)).
 The deviation was mostly because gazeHMM classified a higher number of saccades than the human coders.
@@
 Our algorithm classified shorter and more PSOs than the human coders.
-No other algorithm parsed smooth pursuits, but the RMSD for gazeHMM was higher than among the human coders (see Table \@ref(tab:tab-rmsd-4-sp)).
+No other algorithm parsed smooth pursuits, but the RMSD for gazeHMM was higher than among human coders (see Table \@ref(tab:tab-rmsd-4-sp)).
 Again, it classified a much larger number of smooth pursuits with short durations.
@@
 Table \@ref(tab:tab-kappa-4) contains the sample-to-sample agreement between the algorithms and human coders measured by Cohen's kappa.
-For fixations, gazeHMM showed one of the highest agreements for static, and *the* highest agreements for dynamic data.
+For fixations, gazeHMM showed one of the highest agreements for static and *the* highest agreements for dynamic data.
 The absolute agreement was substantial for the static and slight to fair for the dynamic conditions [@Landis1977].
@@
 Concerning PSOs, gazeHMM showed higher agreement than NH in the image and video conditions but consistently lower agreement compared to LNS.
-Absolutely, the agreement was slight (image) to moderate (video).
+The absolute agreement was slight (image) to moderate (video).
 Lastly, the agreement for smooth pursuit was lower compared to the human coders and fair in absolute values.
@@
 Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for the dynamic conditions.
-For the static condition, the difference to most other algorithms was only slight ($\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$%), but for the dynamic conditions, the differences were substantial (video: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$%) and large (dots: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$%).
-To explore which events gazeHMM classified differently than the average of human coder, we looked at the confusion matrix between the two (see Table \@ref(tab:tab-conf-4)).
+For the static condition, we interpreted the difference to most other algorithms as slight ($\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$%), but for the dynamic conditions, as substantial (video: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$%) and large (dots: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$%).
+To explore which events gazeHMM classified differently than the average human coder, we looked at the confusion matrix between the two (see Table \@ref(tab:tab-conf-4)).
 It can be seen that gazeHMM classified many fixations as smooth pursuits and vice versa.
@@
   mutate_at(vars(label, state), as.factor) %>%
-  dplyr::filter(t > 4.0, t <= 4.5) %>% 
+  dplyr::filter(t > 4.2, t <= 4.5) %>% 
   rename(gazeHMM = label, Time = t) %>% 
@@
   geom_path() + xlab("Time in s") + ylab("") +
-  geom_point(aes(color = gazeHMM, shape = coderMN), size = 0.8) + 
+  geom_point(aes(color = gazeHMM, shape = coderMN), size = 2) + 
   scale_color_viridis_d() +
@@
 An HMM serves as the generative model of the algorithm and classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits.
-We showed in a simulation study that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions: Adding a fourth event (i.e., smooth pursuit) to the model and noise to the generated data were the most critical factors for decreasing recovery.
+We showed in a simulation study that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions: Adding a fourth event (i.e., smooth pursuit) to the model and introducing even small amounts of noise to the generated data were the most critical factors for decreasing recovery.
 Furthermore, we applied gazeHMM with different numbers of states to benchmark data by @Andersson2017 and compared the model fit.
@@
 When comparing gazeHMM against other algorithms, gazeHMM showed mostly good agreement to human coding.
-On one hand, it outperformed the other algorithms in the overall disagreement to human coding for dynamic data.
-On the other hand, gazeHMM confused a lot of fixations with smooth pursuits, which led to a rapid switching between the two events.
+On one hand, it outperformed the other algorithms in the overall disagreement with human coding for dynamic data.
+On the other hand, gazeHMM confused a lot of fixations with smooth pursuits, which led to rapid switching between the two events.
 It also tended to mistake PSO samples as belonging to saccades.
@@
 Thus, there are not more latent states present in the data, but the submodels of the HMM are misspecified or too simple.
-Correcting for model misspecifications led to better model recovery in studies on animal movements [@Langrock2015; @Li2017].
+Correcting for model misspecifications led to a better model recovery in studies on animal movements [@Langrock2015; @Li2017].
 However, @Pohle2017 showed in simulations that the ICL identified the correct model despite several misspecifications.
@@
 Both situations could apply to gazeHMM and eye movement data: Outliers occur frequently in eye-tracking data due to measurement error.
-Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task from which the data are obtained from [@Langrock2015].
+Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task from which the data are obtained [@Langrock2015].
 Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf.
@@
 Nevertheless, Cohen's kappa values of 0.67 (fixations - image) or 0.62 (saccades - moving dots) indicate substantial agreement to human coders, especially in light of the maximum references from the simulation study.
-At this point it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018 observed substantial differences between coders and within coders over time.
+At this point, it is important to mention that human coding should not be considered a gold standard in event classification: @Hooge2018 observed substantial differences between coders and within coders over time.
 Despite these differences, they recommend comparisons to human coding to demonstrate the performance of new algorithms and to find errors in their design.
@@
 ## Advantages of gazeHMM
-In view of the four proposed goals that gazeHMM should fulfill, we can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they only have minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
+Given the four proposed goals that gazeHMM should fulfill, we can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they only have a minor influence on the classification and are merely included to compensate for some drawbacks of the generative model.
 Their default values should be appropriate for most applications.
-A major advantage of gazeHMM is that it does not require human labeled data as input.
+A major advantage of gazeHMM is that it does not require human-labeled data as input.
 Instead, it estimates all parameters and hidden states from the data.
 Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019].
-This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data.
+This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human-labeled data.
 Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit.
-Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2018], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
+Whereas most algorithms only parse fixations and saccades [@Andersson2017], few classify PSOs [e.g., @Zemblys2018], and even less categorize smooth pursuits [e.g., @Pekkanen2017].
 However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events.
@@
 Another option to model the correlation could be to include one of the response variables as a covariate of the other.
-Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred.
-This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.
+Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with fewer states being preferred.
+This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data but would also undermine the advantages of using a parametric model.
 Third, one solution to diverging results when comparing gazeHMM with different events could be model averaging: Instead of using the maximum posterior state probability of each sample from the preferred model, the probabilities could be weighted according to a model selection criterion (e.g., Schwarz weight) and averaged.
@@
 Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs.
-This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades.
+This extension of the HMM is also called the hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades.
 Drawbacks of this extension are higher computational costs and difficulties with including covariates [@Zucchini2016].
@@
 This could potentially be achieved by using different response distributions or parameter optimization methods.
-Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain, efficient sampling plans, accounting for label switching, and computational efficiency, naming only a few.
+Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help to avoid convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain, efficient sampling plans, accounting for label switching, and computational efficiency, naming only a few.
 
@@
 
+# Declarations
+## Funding
+The authors did not receive support from any organization for the submitted work.
+
+## Conflicts of Interest
+The authors declare that they have no financial or non-financial interests regarding the content of this article.
+
+## Ethics Approval
+Not applicable.
+
+## Consent to Participate
+Not applicable.
+
+## Consent to Publication
+Not applicable.
+
+## Availability of Data and Materials
+See Open Practices and Data Availability Statement.
+
+## Code Availability
+See Open Practices and Data Availability Statement.
+
 # Open Practices and Data Availability Statement
-The data and materials for the validation study [@Andersson2017; @Larsson2013; @Larsson2015] are available on GitHub at https://github.com/richardandersson/EyeMovementDetectorEvaluation.
-The code to obtain the results for the simulation and validation study is available at https://github.com/maltelueken/gazeHMM_validation and the algorithm can be accessed at https://github.com/maltelueken/gazeHMM.
-The simulation study was preregistered (https://doi.org/10.17605/OSF.IO/VDJGP), whereas the validation study was not.
+The data and materials for the validation study [@Andersson2017; @Ehinger2019; @Larsson2013; @Larsson2015] are available on GitHub at <https://github.com/richardandersson/EyeMovementDetectorEvaluation> and on Figshare at <https://doi.org/10.6084/m9.figshare.c.4379810.v1>.
+The code to obtain the results for the simulation and validation study is available at <https://github.com/maltelueken/gazeHMM_validation> and the algorithm can be accessed at <https://github.com/maltelueken/gazeHMM>.
+The simulation study was preregistered (<https://doi.org/10.17605/OSF.IO/VDJGP>), whereas the validation study was not.
 

### 2020-11-16 manuscript/paper_BRM.Rmd
change abstract and add interim discussion

@@
 ---
-title             : "Classifying Gaze Data With an Unsupervised Generative Hidden Markov Model"
-shorttitle        : "Classifying Gaze Data with an HMM"
+title             : "Classifying Eye Movement Events With an Unsupervised Generative Hidden Markov Model"
+shorttitle        : "Classifying Eye Movement Events"
 
@@
 Furthermore, we would like to acknowledge feedback by Karel Veldkamp and Phil Norberts in the early stage of this project.
+Šimon Kucharský was supported by the NWO (Nederlandse Organisatie voor Wetenschappelijk Onderzoek) grant no.
+406.10.559.
 
@@
 The algorithm classifies gaze data into fixations, saccades, and optionally postsaccadic oscillations and smooth pursuits.
-We evaluated the performance of gazeHMM in a simulation study and on two benchmark data sets.
-The simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences.
-Critical exceptions for parameter recovery were including the smooth pursuit state in the generative model and even low noise levels.
-We assessed the performance on benchmark data by model comparing generative models with different events and agreement of the classification to human coding compared to other algorithms.
-The model comparison preferred more events in the generative models than we expected.
-Nevertheless, gazeHMM showed relatively high similarity to human coding and outperformed other algorithms in this aspect.
-For dynamic stimuli, our algorithm tended to switch rapidly between fixations and smooth pursuits.
-Thus, we recommend using gazeHMM for parsing smooth pursuits only in explorative settings.
+We evaluated gazeHMM’s performance in a simulation study, showing that it successfully recovered HMM parameters and hidden states.
+Parameters were less well recovered when we included a smooth pursuit state and/or added even small noise to simulated data.
+We applied generative models with different numbers of events to benchmark data.
+Comparing them indicated that HMMs with more events than expected had most likely generated the data.
+We also applied the full algorithm to benchmark data and assessed its similarity to human coding.
+For static stimuli, gazeHMM showed high similarity and outperformed other algorithms in this regard.
+For dynamic stimuli, gazeHMM tended to rapidly switch between fixations and smooth pursuits but still displayed higher similarity than other algorithms.
+Concluding that gazeHMM can be used in practice, we recommend parsing smooth pursuits only for exploratory purposes.
 Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to classify smooth pursuits more accurately.
 
-keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
-wordcount         : "10095"
+keywords          : "eye-tracking; event detection; parameter recovery; dependent mixture models"
+wordcount         : "10328"
 
@@
 
+header-includes   :
+  - \usepackage{placeins}
 floatsintext      : yes
@@
 documentclass     : "apa6"
-classoption       : "man"
+classoption       : "man, noextraspace"
 output            : papaja::apa6_pdf
@@
 For non-parametric methods, this distinction might be accurate.
-However, for parametric methods, we think calling step two "classification" is more appropriate since the probabilistic classification is done here.
+However, for parametric methods, calling step two "classification" is more appropriate since the probabilistic classification is done here.
 Step three usually consists of some heuristic relabeling and correcting for classification errors.
@@
 Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017], but most of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007].
-This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017].
-In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.
+This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's characteristics [e.g., sampling frequency; @Andersson2017].
+In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require additional parameter settings.
 Theoretically, these three metrics should separate eye movement events.
@@
 In gazeHMM, we use gamma distributions with a shape and scale parametrization for this purpose:
-$$X \sim Gamma(\alpha,\beta),$$
-$$Y \sim Gamma(\alpha,\beta).$$
+$$\begin{aligned}
+(X \mid S = i) &\sim \text{Gamma}(\alpha_{xi}, \beta_{xi}) \\
+(Y \mid S = i) &\sim \text{Gamma}(\alpha_{yi}, \beta_{yi}),
+\end{aligned}$$
+with *i* denoting the hidden state.
 When we developed gazeHMM, the gamma distribution appeared to fit eye movement data well, but we also note that it might not necessarily be the best fitting distribution for every type of eye movement data.
@@
 $$\begin{aligned}
-Z \mid fixation &\sim U(0, 2\pi) \\
-Z \mid saccade &\sim von\ Mises(\mu, \kappa) \\
-Z \mid PSO &\sim von\ Mises(\mu, \kappa) \\
-Z \mid smooth\ pursuit &\sim von\ Mises(\mu, \kappa)
-\end{aligned}.$$
+(Z \mid S = 1) &\sim U(0, 2\pi) \\
+(Z \mid S = 2) &\sim \text{von Mises}(\mu_{\text{1}}, \kappa_{\text{1}}) \\
+(Z \mid S = 3) &\sim \text{von Mises}(\mu_{\text{2}}, \kappa_{\text{2}}) \\
+(Z \mid S = 4) &\sim \text{von Mises}(\mu_{\text{3}}, \kappa_{\text{3}}).
+\end{aligned}$$
 Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions.
@@
 The study was divided in four parts.
-Here, we only report the first two parts, the other two parts, which address starting values and missing data, can be found in the supplementary material (<https://github.com/maltelueken/gazeHMM_validation>).
+Here, we only report the first two parts, which investigate the influence of parameter variation and adding noise to generated data on recovery.
+The other two parts, which address starting values and missing data, can be found in the supplementary material (<https://github.com/maltelueken/gazeHMM_validation>).
 The HMM repeatedly generated data with a set of parameters (true parameter values).
@@
 
+```{r sim-parameters-trans}
+
+sim.pars.trans <- as.data.frame(matrix(c("Interval", "-", "[.01,.99]", "1-a(i=j)/(k-1)",
+                                           "Default", "1/k", "0.9", "0.1/(k-1)"), nrow = 2, byrow = T))
+
+apa_table(sim.pars.trans,
+          col.names = c("", "Initial\nstate\nprob.", "Trans. prob.\nsame state", "Trans. prob.\nother state"),
+          caption = "(ref:sim-parameters-trans)",
+          note = "(ref:sim-parameters-trans-note)")
+
+```
+
+(ref:sim-parameters-resp) Intervals and Default Parameter Values for the Response Model in the Simulation
+(ref:sim-parameters-resp-note) Shape parameters are denoted by $\alpha$, scale parameters by $\beta$, mean parameters by $\mu$, and concentration parameters by $\kappa$.
+The default values for the uniform distribution in state one were min = 0 and max = $2\pi$.
+
+```{r sim-parameters-resp}
+
+sim.pars.resp <- as.data.frame(t(matrix(c("[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
+                                  "3", "0.35", "3", "0.25", "-", "-",
+                                  "[1,5]", "[5,15]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
+                                  "3", "10", "3", "3", "0", "1",
+                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
+                                  "3", "1", "3", "3", paste(expression(pi), sep = ""), "1",
+                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]",
+                                  "3", "1", "3", "0.15", "0", "1"), nrow = 6, byrow = F)))
+
+apa_table(cbind(rep(c("Interval", "Default"), 4), sim.pars.resp),
+          col.names = c("", "Shape", "Scale", "Shape", "Scale", "Mean", "Concentration"),
+          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
+          stub_indents = list("State 1" = c(1, 2), "State 2" = c(3, 4), "State 3" = c(5, 6), "State 4" = c(7, 8)),
+          caption = "(ref:sim-parameters-resp)",
+          note = "(ref:sim-parameters-resp-note)")
+
+```
+
+\FloatBarrier
+
 ## Data Analysis
@@
 
-```{r sim-parameters-trans}
-
-sim.pars.trans <- as.data.frame(matrix(c("Interval", "-", "[.01,.99]", "1-a(i=j)/(k-1)",
-                                           "Default", "1/k", "0.9", "0.1/(k-1)"), nrow = 2, byrow = T))
-
-apa_table(sim.pars.trans,
-          col.names = c("", "Initial\nstate\nprob.", "Trans. prob.\nsame state", "Trans. prob.\nother state"),
-          caption = "(ref:sim-parameters-trans)",
-          note = "(ref:sim-parameters-trans-note)")
-
-```
-
-(ref:sim-parameters-resp) Intervals and Default Parameter Values for the Response Model in the Simulation
-(ref:sim-parameters-resp-note) Shape parameters are denoted by $\alpha$, scale parameters by $\beta$, mean parameters by $\mu$, and concentration parameters by $\kappa$.
-The default values for the uniform distribution in state one were min = 0 and max = $2\pi$.
-
-```{r sim-parameters-resp}
-
-sim.pars.resp <- as.data.frame(t(matrix(c("[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
-                                  "3", "0.35", "3", "0.25", "-", "-",
-                                  "[1,5]", "[5,15]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
-                                  "3", "10", "3", "3", "0", "1",
-                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
-                                  "3", "1", "3", "3", paste(expression(pi), sep = ""), "1",
-                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]",
-                                  "3", "1", "3", "0.15", "0", "1"), nrow = 6, byrow = F)))
-
-apa_table(cbind(rep(c("Interval", "Default"), 4), sim.pars.resp),
-          col.names = c("", "Shape", "Scale", "Shape", "Scale", "Mean", "Concentration"),
-          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
-          stub_indents = list("State 1" = c(1, 2), "State 2" = c(3, 4), "State 3" = c(5, 6), "State 4" = c(7, 8)),
-          caption = "(ref:sim-parameters-resp)",
-          note = "(ref:sim-parameters-resp-note)")
-
-```
-
 ```{r load sim results and helper functions, include=FALSE}
@@
 For most estimated models, the kappa values between true and estimated state sequences were above 0.95, meaning almost perfect agreement.
-However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched (see supplementary material).
-
-[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
-Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
-Therefore, we decided to exclude it from the analysis.
-
-(ref:plot-rmdspd-1-3) RMdSPD between true and estimated parameters of the three-state HMM in part one of the simulation.
-Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
-Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
-$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
-
-```{r plot-rmdspd-1-3, fig.cap="(ref:plot-rmdspd-1-3)"}
-
-print(plots.rmsd.1[[2]]+ theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-regtr-1-3) Regression lines between true and estimated transition probabilities for the three-state HMM in part one.
-Right facet labels show *from* and top facet labels show *to* which state the HMM is moving.
-Dashed lines refer to perfect recovery.
-
-```{r plot-regtr-1-3, fig.cap="(ref:plot-regtr-1-3)"}
-
-print(plots.lm.tr[[2]])
-
-```
-
-(ref:plot-regresp-1-3) Regression lines between true and estimated response parameters of the three-state HMM in part one.
-Top facet labels indicate response parameters.
-Dashed lines refer to perfect recovery.
-$\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
-Parameter subscripts indicate states and eye movement metrics.
-
-```{r plot-regresp-1-3, fig.cap="(ref:plot-regresp-1-3)", fig.height=7}
-
-print(plots.lm.resp[[2]] + theme(axis.text = element_text(size = 8)))
-
-```
-
-(ref:plot-acc-1-3) Boxplots displaying Cohen's kappa depending on which parameter of the three-state HMM has been manipulated in part one.
-Top facet labels indicate for which state parameters have been manipulated.
-Black solid lines symbolize medians and hinges the first and third quartile.
-Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
-Crosses represent outliers.
-$a_{i=j}$ is the probability to stay in the same state, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
-
-```{r plot-acc-1-3, fig.cap="(ref:plot-acc-1-3)"}
-
-print(plots.acc.1[[2]])
-
-```
-
+However, for some models, we observed kappas clustered around zero or -0.33, which is far from the majority of model accuracies.
+An exploratory examination of these clusters suggests that state labels were switched (see supplementary material).
 The RMdSPDs for the four-state HMM is shown in Figure \@ref(fig:plot-rmdspd-1-4).
@@
 Cohen's kappa ranged mostly between 0.6 and 0.9, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)).
-Here, some kappa values clustered around 0.25 and zero, which, again, suggests label switching.
-Overall, simulations in part one demonstrated that the HMM recovered parameters very well when they were manipulated.
-Deviations from true parameters were mostly small.
-In the four-state model, estimated transition probabilities for state one and four deviated moderately.
-Moreover, the HMM estimated state sequences very accurately with decreasing accuracy for the four-state model.
+Here, some outlying kappa values clustered around 0.25 and zero.
+
+[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$.
+Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$).
+Therefore, we decided to exclude it from the analysis.
+
+(ref:plot-rmdspd-1-3) RMdSPD between true and estimated parameters of the three-state HMM in part one of the simulation.
+Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed.
+Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+
+```{r plot-rmdspd-1-3, fig.cap="(ref:plot-rmdspd-1-3)"}
+
+print(plots.rmsd.1[[2]]+ theme(axis.text = element_text(size = 8)))
+
+```
+
+(ref:plot-regtr-1-3) Regression lines between true and estimated transition probabilities for the three-state HMM in part one.
+Right facet labels show *from* and top facet labels show *to* which state the HMM is moving.
+Dashed lines refer to perfect recovery.
+
+```{r plot-regtr-1-3, fig.cap="(ref:plot-regtr-1-3)"}
+
+print(plots.lm.tr[[2]])
+
+```
+
+(ref:plot-regresp-1-3) Regression lines between true and estimated response parameters of the three-state HMM in part one.
+Top facet labels indicate response parameters.
+Dashed lines refer to perfect recovery.
+$\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+Parameter subscripts indicate states and eye movement metrics.
+
+```{r plot-regresp-1-3, fig.cap="(ref:plot-regresp-1-3)", fig.height=7}
+
+print(plots.lm.resp[[2]] + theme(axis.text = element_text(size = 8)))
+
+```
+
+(ref:plot-acc-1-3) Boxplots displaying Cohen's kappa depending on which parameter of the three-state HMM has been manipulated in part one.
+Top facet labels indicate for which state parameters have been manipulated.
+Black solid lines symbolize medians and hinges the first and third quartile.
+Whiskers range from hinges to lowest/highest value within 1.5 times the IQR.
+Crosses represent outliers.
+$a_{i=j}$ is the probability to stay in the same state, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+
+```{r plot-acc-1-3, fig.cap="(ref:plot-acc-1-3)"}
+
+print(plots.acc.1[[2]])
+
+```
 
@@
 
-```{r plot-regresp-1-4, fig.cap="(ref:plot-regresp-1-4)", fig.height=8}
+```{r plot-regresp-1-4, fig.cap="(ref:plot-regresp-1-4)", fig.height=7}
 
@@
 ```
+
+\FloatBarrier
 
@@
 Accuracy measured by Cohen's kappa was almost perfect with kappa values very close to one (see Figure \@ref(fig:plot-acc-2), left plot).
+For three states, the RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)), indicating bad recovery.
+Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5.
+For most parameters across all three states, models with higher sample sizes had lower RMdSPDs.
+The state recovery of the estimated models was almost perfect with most kappa values above 0.95 (see Figure \@ref(fig:plot-acc-2), middle plot).
+Several outliers clustered around kappa values of zero and -0.33.
+RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
+For states one and four, values for most parameters (including all transition probabilities) were above 0.5, suggesting bad recovery.
+Similarly, $\beta_{vel}$ and $\beta_{acc}$ in states two and three showed bad recovery.
+For states two and three, higher sample sizes showed slightly lower RMdSPDs.
+As in the previous part, most Cohen's kappa values ranged between 0.6 and 0.9, meaning substantial to almost perfect agreement between true and estimated states (Figure \@ref(fig:plot-acc-2), right plot).
+Multiple outliers clustered around 0.25 or zero.
 
@@
 
-Regarding three states, the RMdSPDs for the $\beta_{vel}$ and $\beta_{acc}$ were above 0.5 in all three states (see Figure \@ref(fig:plot-rmdspd-2-3)), indicating bad recovery.
-Again, the other estimated parameters were below or close to 0.1, only $a_{12}$ and $a_{31}$ with 500 samples were closer to 0.5.
-For most parameters across all three states, models with higher sample sizes had lower RMdSPDs.
-The state recovery of the estimated models was almost perfect with most kappa values above 0.95 (see Figure \@ref(fig:plot-acc-2), middle plot).
-Several outliers clustered around kappas of zero and -0.33, signaling label switching.
-
 (ref:plot-rmdspd-2-3) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
@@
 
-RMdSPDs regarding the four-state HMM are displayed in Figure \@ref(fig:plot-rmdspd-2-4).
-For states one and four, values for most parameters (including all transition probabilities) were above 0.5, suggesting bad recovery.
-Similarly, $\beta_{vel}$ and $\beta_{acc}$ in states two and three showed bad recovery.
-For states two and three, higher sample sizes showed slightly lower RMdSPDs.
-As in the previous part, most Cohen's kappa values ranged between 0.6 and 0.9, meaning substantial to almost perfect agreement between true and estimated states (Figure \@ref(fig:plot-acc-2), right plot).
-Multiple kappa values clustered around 0.25 or zero, which can be explained by label switching.
-In general, the HMM recovered parameters well despite noise being added to the data.
+(ref:plot-rmdspd-2-4) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
+Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
+Right facet labels denote to which state estimated parameters belong.
+$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+
+```{r plot-rmdspd-2-4, fig.cap="(ref:plot-rmdspd-2-4)"}
+
+print(plots.rmsd.234[[1]][[3]] + theme(axis.text.y = element_text(size = 8)))
+
+```
+
+\FloatBarrier
+
+## Discussion
+In the simulation study, we assessed the recovery of parameters and hidden states in the generative model of gazeHMM.
+Simulations in part one demonstrated that the HMM recovered parameters very well when they were manipulated.
+Deviations from true parameters were mostly small.
+In the four-state model, estimated transition probabilities for state one and four deviated moderately.
+Moreover, the HMM estimated state sequences very accurately with decreasing accuracy for the four-state model.
+In the second part, noise was added to the generated data and the sample size was varied.
+Despite noise, the generative model was still able to recover most parameters well.
 However, in the four-state model, the parameter recovery for states one and four substantially decreased (even for low amounts of noise, see supplementary material).
@@
 The state recovery of the model was slightly lowered when more states were included, but it was neither affected by the noise level nor the sample size.
-
-(ref:plot-rmdspd-2-4) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation.
-Labels on the y-axis indicate for which estimated parameter the RMdSPD is displayed.
-Right facet labels denote to which state estimated parameters belong.
-$\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
-
-```{r plot-rmdspd-2-4, fig.cap="(ref:plot-rmdspd-2-4)"}
-
-print(plots.rmsd.234[[1]][[3]] + theme(axis.text.y = element_text(size = 8)))
-
-```
+In the third part (included in the supplementary material), we showed that the variation in starting values used to fit the HMM did not influence parameter and state recovery.
+Missing data (in part four, also in the supplementary material) did not affect the parameter recovery but linearly decreased the recovery of hidden states.
+In all four parts, we observed clusters of outlying accuracy values.
+In part three, we exploratorily examined these clusters and reasoned that they can be attributed to label switching (i.e., flipping one or two state labels resolved the outlying clusters).
+In general, the generative model recovers parameters and hidden states well and, thus, we conclude that it can be used in our classification algorithm.
+However, the recovery decreases when a fourth state (i.e., smooth pursuit) is added to the model and, especially with four states, many parameters in the HMM are vulnerable to noise.
+In the next sections, we will see how noise that is present in real eye movement data affects the performance of gazeHMM.
 
@@
 
+\FloatBarrier
+
 ### Comparison to Other Algorithms
@@
 Again, it classified a much larger number of smooth pursuits with short durations.
-
-(ref:tab-rmsd-4-fix) Fixation Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-fix-note) Durations are displayed in milliseconds.
-gazeHMM-3 classified three and gazeHMM-4 classified four events.
-RMSD = root mean square deviation.
-Table design adapted from @Andersson2017.
-
-```{r tab-rmsd-4-fix}
-
-apa_table(A2017.rmsd[[1]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = TRUE),
-          caption = "(ref:tab-rmsd-4-fix)",
-          note = "(ref:tab-rmsd-4-fix-note)")
-
-```
-
-(ref:tab-rmsd-4-sac) Saccade Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-sac-note) Durations are displayed in milliseconds.
-gazeHMM-3 classified three and gazeHMM-4 classified four events.
-RMSD = root mean square deviation.
-Table design adapted from @Andersson2017.
-
-```{r tab-rmsd-4-sac}
-
-apa_table(A2017.rmsd[[2]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = TRUE),
-          caption = "(ref:tab-rmsd-4-sac)",
-          note = "(ref:tab-rmsd-4-sac-note)")
-
-```
-
-(ref:tab-rmsd-4-pso) PSO Duration Descriptives and RMSD Between Algorithms and Human Coders
-(ref:tab-rmsd-4-pso-note) Durations are displayed in milliseconds.
-gazeHMM-3 classified three and gazeHMM-4 classified four events.
-RMSD = root mean square deviation.
-Table design adapted from @Andersson2017.
-
-```{r tab-rmsd-4-pso}
-
-apa_table(A2017.rmsd[[3]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = TRUE),
-          caption = "(ref:tab-rmsd-4-pso)",
-          note = "(ref:tab-rmsd-4-pso-note)")
-
-```
-
-(ref:tab-rmsd-4-sp) Smooth Pursuit Duration Descriptives and RMSD Between gazeHMM and Human Coders
-(ref:tab-rmsd-4-sp-note) Durations are displayed in milliseconds.
-gazeHMM-3 classified three and gazeHMM-4 classified four events.
-RMSD = root mean square deviation.
-Table design adapted from @Andersson2017.
-
-```{r tab-rmsd-4-sp}
-
-apa_table(A2017.rmsd[[4]],
-          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
-          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = TRUE),
-          caption = "(ref:tab-rmsd-4-sp)",
-          note = "(ref:tab-rmsd-4-sp-note)")
-
-```
-
 Table \@ref(tab:tab-kappa-4) contains the sample-to-sample agreement between the algorithms and human coders measured by Cohen's kappa.
@@
 Lastly, the agreement for smooth pursuit was lower compared to the human coders and fair in absolute values.
-
-(ref:tab-kappa-4) Cohen's Kappa Between Human Coders and Algorithms for Different Conditions and Events
-(ref:tab-kappa-4-note) gazeHMM-3 classified three and gazeHMM-4 classified four events. Table design adapted from @Andersson2017.
-
-```{r tab-kappa-4}
-
-A2017.kappa[3, -c(1, seq(2, 13, 3))] <- NA
-A2017.kappa[4, seq(2, 13, 3)] <- NA
-
-names(A2017.kappa) <- c("Algorithm", rep(c("Image", "Dots", "Video"), 4))
-
-apa_table(A2017.kappa,
-          col_spanners = list("Fixations" = c(2, 4), "Saccades" = c(5, 7), "PSOs" = c(8, 10), "Smooth pursuits" = c(11, 13)),
-          row.names = FALSE,
-          font_size = "tiny",
-          format.args = list(na_string = "-", drop0trailing = T),
-          caption = "(ref:tab-kappa-4)",
-          note = "(ref:tab-kappa-4-note)")
-
-```
-
 Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for the dynamic conditions.
@@
 In the example, the human coder also delayed the start of the PSO compared to gazeHMM.
+
+(ref:tab-rmsd-4-fix) Fixation Duration Descriptives and RMSD Between Algorithms and Human Coders
+(ref:tab-rmsd-4-fix-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
+
+```{r tab-rmsd-4-fix}
+
+apa_table(A2017.rmsd[[1]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = TRUE),
+          caption = "(ref:tab-rmsd-4-fix)",
+          note = "(ref:tab-rmsd-4-fix-note)")
+
+```
+
+(ref:tab-rmsd-4-sac) Saccade Duration Descriptives and RMSD Between Algorithms and Human Coders
+(ref:tab-rmsd-4-sac-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
+
+```{r tab-rmsd-4-sac}
+
+apa_table(A2017.rmsd[[2]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = TRUE),
+          caption = "(ref:tab-rmsd-4-sac)",
+          note = "(ref:tab-rmsd-4-sac-note)")
+
+```
+
+(ref:tab-rmsd-4-pso) PSO Duration Descriptives and RMSD Between Algorithms and Human Coders
+(ref:tab-rmsd-4-pso-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
+
+```{r tab-rmsd-4-pso}
+
+apa_table(A2017.rmsd[[3]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = TRUE),
+          caption = "(ref:tab-rmsd-4-pso)",
+          note = "(ref:tab-rmsd-4-pso-note)")
+
+```
+
+(ref:tab-rmsd-4-sp) Smooth Pursuit Duration Descriptives and RMSD Between gazeHMM and Human Coders
+(ref:tab-rmsd-4-sp-note) Durations are displayed in milliseconds.
+gazeHMM-3 classified three and gazeHMM-4 classified four events.
+RMSD = root mean square deviation.
+Table design adapted from @Andersson2017.
+
+```{r tab-rmsd-4-sp}
+
+apa_table(A2017.rmsd[[4]],
+          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
+          col_spanners = list("Image" = c(2, 5), "Moving dots" = c(6, 9), "Video" = c(10, 13)),
+          font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = TRUE),
+          caption = "(ref:tab-rmsd-4-sp)",
+          note = "(ref:tab-rmsd-4-sp-note)")
+
+```
+
+(ref:tab-kappa-4) Cohen's Kappa Between Human Coders and Algorithms for Different Conditions and Events
+(ref:tab-kappa-4-note) gazeHMM-3 classified three and gazeHMM-4 classified four events. Table design adapted from @Andersson2017.
+
+```{r tab-kappa-4}
+
+A2017.kappa[3, -c(1, seq(2, 13, 3))] <- NA
+A2017.kappa[4, seq(2, 13, 3)] <- NA
+
+names(A2017.kappa) <- c("Algorithm", rep(c("Image", "Dots", "Video"), 4))
+
+apa_table(A2017.kappa,
+          col_spanners = list("Fixations" = c(2, 4), "Saccades" = c(5, 7), "PSOs" = c(8, 10), "Smooth pursuits" = c(11, 13)),
+          row.names = FALSE,
+          font_size = "tiny",
+          format.args = list(na_string = "-", drop0trailing = T),
+          caption = "(ref:tab-kappa-4)",
+          note = "(ref:tab-kappa-4-note)")
+
+```
 
@@
 
-# Discussion
+\FloatBarrier
+
+# General Discussion
 In this report, we presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events.
@@
 ## Funding
-The authors did not receive support from any organization for the submitted work.
+Šimon Kucharský was supported by the NWO (Nederlandse Organisatie voor Wetenschappelijk Onderzoek) grant no. 406.10.559.
 

### 2020-11-16 manuscript/supplementary_material_simulation.Rmd
change abstract and add interim discussion

@@
     scale_y_continuous(name = "Estimated response parameter")
+
+  return(p)
+})
+
+# Plot parameters over noise interval part 2
+
+par.noise.data <- lapply(get("estimates.2"), function(x) {
+  lapply(x, function(y) {
+    true <- lapply(y, function(z) {
+
+      true <- backtrans(z$pars.true)
+
+      return(true)
+    }) %>% reduce(cbind)
+
+    est <- lapply(y, function(z) {
+
+      true <- backtrans(z$pars.true)
+      est <- try(backtrans(z$pars.est))#ifelse(class(z$pars.est) == "try-error", rep(NA, length(true)), backtrans(z$pars.est))
+
+      if(class(est) == "try-error") est <- rep(NA, length(true))
+
+      return(est)
+    }) %>% reduce(cbind)
+
+    return(list(as.data.frame(true), as.data.frame(est)))
+  })
+})
+
+plot.par.noise <- lapply(1:length(par.noise.data), function(x) {
+  out <- lapply(par.noise.data[[x]], function(y) {
+
+    if(nrow(y[[1]]) == 18) {
+
+    k <- 2
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]))
+
+    } else if(nrow(y[[1]]) == 30) {
+
+    k <- 3
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]))
+
+    } else {
+
+    k <-4
+
+    trnames <- list(bquote(rho["i"]), bquote(a["i1"]), bquote(a["i2"]), bquote(a["i3"]), bquote(a["i4"]))
+
+    }
+
+    names.pars.est <<- append(trnames, list(bquote(alpha["vel"]), bquote(beta["vel"]), bquote(alpha["acc"]),
+                                         bquote(beta["acc"]), bquote(mu), bquote(kappa)))
+
+    true <- as.data.frame(t(y[[1]])) %>% pivot_longer(everything(), names_to = "par", values_to = "true.value")
+    est <- as.data.frame(t(y[[2]])) %>% pivot_longer(everything(), names_to = "par", values_to = "est.value")
+
+    par <- as.factor(rep(c(rep(1, k), 
+                 rep(2:(k+1), each = k), 
+                 rep((k+2):(k+7), k)), D))
+    par.state <- as.factor(rep(c(rep(1:k, k+1), rep(1:k, each = 6)), D))
+    noise <- rep(seq(1, 5, length.out = D), each = length(par)/D)
+
+    minpos <- 1 + k + 5
+    maxpos <- 1 + k + 6
+
+    est$est.value[par %in% c(minpos, maxpos) & par.state == 1] <- NA
+
+    diff <- ifelse(true$true.value == 0, (est$est.value - true$true.value)/2*pi, (est$est.value - true$true.value)/true$true.value)
+
+    df <- data.frame(par = par, diff = diff, state = par.state, noise = noise)
+
+    return(df)
+  }) %>% reduce(rbind)
+
+  N <- as.factor(rep(c("500", "2500", "10000"), each = nrow(out)/3))
+  out$N <- N
+  levels(out$state) <- paste0("State ", levels(out$state))
+
+  noise.scale <- paste(1:5, round(1/seq(0.1, 10, length.out = 5), 2), sep = "\n")
+
+  #return(out)
+
+  p <- ggplot(data = na.omit(out), aes(x = noise, y = abs(diff), color = par)) + geom_smooth() +
+    facet_wrap(vars(state)) +
+    scale_color_viridis_d(name = "Estimated\nparameter", labels = names.pars.est) +
+    scale_x_continuous(name = "Noise", labels = noise.scale) +
+    scale_y_continuous(name = "Absolute proportional deviation", limits = c(0, 5))
 
@@
 
-  p <- ggplot(data = data.long, aes(x = k, y = accuracy, color = cond)) +
-    geom_boxplot(outlier.shape = 4) + 
+  p <- ggplot(data = data.long, aes(x = k, y = accuracy, fill = cond)) +
+    geom_boxplot(color = "black", outlier.shape = 4) + 
     scale_x_discrete(name = "k (number of states)")  + 
@@
 
-  p <- p + scale_color_viridis_d(name = label.cond, labels = names.cond)
+  p <- p + scale_fill_viridis_d(name = label.cond, labels = names.cond)
 
@@
     scale_y_continuous(name = "Cohen's kappa", breaks = c(-1, -0.33, 0, 0.25, 0.5, 0.75, 1)) +
-    scale_color_viridis_d()
+    scale_color_viridis_d(name = label.cond, labels = names.cond)
 
@@
 
+## Addendum Part Two
+This section describes how we exploratorily examined the recovery of parameters depending on the amount of noise added to the data.
+Figure \@ref(fig:plot-noise-2-2) shows the trends in absolute proportional deviation between estimated and true parameters (i.e., $\text{abs}((\hat{\theta}_i - \theta_i)/\theta_i)$) for the two-state HMM.
+We observed that the that the deviation is low for most parameters and does not change with increased noise.
+Exceptions were the scale parameters of the gamma distributions in both states, which started around 0.6 and almost linearly increased (with a plateau at the end) until 2.5 absolute proportional deviation when the noise level rose.
+In light of the previously adopted scale for RMdSPD, we conclude that these parameters are badly recovered across all noise levels (i.e., they are higher than 0.5).
+Conversely, for the parameters of the von Mises distribution in state two, the deviation exponentially declined with increasing noise, starting from 1 and ending at almost 0.
+The exponential decline occurs because we manipulated the noise for values generated by the von Mises distribution on the inverse scale.
+Thus, for high noise the von Mises parameters are badly recovered, whereas for lower levels of noise they are well recovered.
+Figure \@ref(fig:plot-noise-2-3) shows a similar pattern for the three-state HMM.
+In contrast to the previous model, the mean parameter of the von Mises distribution in state two exhibited increased deviation for lower noise levels.
+Moreover, the deviations for transition probabilities from state two to state three increased slightly with higher noise.
+The trends for the HMM with four states are shown in Figure \@ref(fig:plot-noise-2-4).
+States two and three behave similarly as in the previous model, except that the probabilities to transition to state one are initially high and decrease with higher levels of noise.
+For state one, the trends are almost indistinguishable from each other and do not visibly change with the noise level.
+State 4 still shows an increasing trend for the scale parameters and decreasing trends for transition probabilities towards states one and three.
+In conclusion, this exploratory analysis showed that the scale parameters of the gamma distributions were already badly disturbed by small amounts of noise.
+Parameters of the von Mises distributions were susceptible to deviations through large noise levels but recovered well for lower noise levels.
+
+(ref:plot-noise-2-2) Absoluted proportional deviation between estimated and true parameters of the two-state HMM depending on noise level in part two of the simulation.
+The upper labels on the x-axis indicate $\tau_{noise}$ and the lower labels $\kappa_{noise}$.
+Colours indicate for which estimated parameter and top facet labels for which state the deviation is displayed.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+
+```{r plot-noise-2-2, fig.cap="(ref:plot-noise-2-2)"}
+
+print(plot.par.noise[[1]])
+
+```
+
+(ref:plot-noise-2-3) Absoluted proportional deviation between estimated and true parameters of the three-state HMM depending on noise level in part two of the simulation.
+The upper labels on the x-axis indicate $\tau_{noise}$ and the lower labels $\kappa_{noise}$.
+Colours indicate for which estimated parameter and top facet labels for which state the deviation is displayed.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+
+```{r plot-noise-2-3, fig.cap="(ref:plot-noise-2-3)"}
+
+print(plot.par.noise[[2]])
+
+```
+
+(ref:plot-noise-2-4) Absoluted proportional deviation between estimated and true parameters of the four-state HMM depending on noise level in part two of the simulation.
+The upper labels on the x-axis indicate $\tau_{noise}$ and the lower labels $\kappa_{noise}$.
+Colours indicate for which estimated parameter and top facet labels for which state the deviation is displayed.
+$\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.
+
+```{r plot-noise-2-4, fig.cap="(ref:plot-noise-2-4)"}
+
+print(plot.par.noise[[3]])
+
+```
+
 ## Part Three
@@
 
-print(plots.rmsd.234[[2]][[1]] + coord_flip())
+print(plots.rmsd.234[[2]][[1]])
 
@@
 
-print(plots.rmsd.234[[2]][[2]] + coord_flip())
+print(plots.rmsd.234[[2]][[2]])
 
@@
 
-print(plots.rmsd.234[[2]][[3]] + coord_flip())
+print(plots.rmsd.234[[2]][[3]])
 
@@
 
-print(plots.rmsd.234[[3]][[1]] + coord_flip())
+print(plots.rmsd.234[[3]][[1]])
 
@@
 
-print(plots.rmsd.234[[3]][[2]] + coord_flip())
+print(plots.rmsd.234[[3]][[2]])
 
@@
 
-print(plots.rmsd.234[[3]][[3]] + coord_flip())
+print(plots.rmsd.234[[3]][[3]])
 

### 2020-11-18 manuscript/preprint_Luken_Kucharsky_Visser_Classifying_Eye_Movement_Events.Rmd
upload preprint

--- 2020-11-16 manuscript/paper_BRM.Rmd
@@
 authornote: |
+ *Please note that this is a preprint and has not been certified by peer review*.
+
  We would like to thank Daan van Renswoude, Maartje Raijmakers, and Maximilian Maier for their helpful comments on earlier versions of this paper.
@@
 keywords          : "eye-tracking; event detection; parameter recovery; dependent mixture models"
-wordcount         : "10328"
+wordcount         : "10279"
 
@@
 ```
+
+\raggedbottom
 
@@
 
-```{r plot-algorithm-workflow, fig.cap="(ref:plot-algorithm-workflow)", fig.height = 8}
+```{r plot-algorithm-workflow, fig.cap="(ref:plot-algorithm-workflow)", fig.height = 7}
 
@@
   geom_point(size = 2) +
-  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
+  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 300), breaks = c(0, 100, 200, 300))
 
@@
   geom_point(aes(color = event.pre), size = 2) +
-  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
+  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 300), breaks = c(0, 100, 200, 300))
 
@@
   geom_point(aes(color = event.post), size = 2) +
-  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))
+  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 300), breaks = c(0, 100, 200, 300))
 
@@
 
+(ref:sim-parameters-trans) Intervals and Default Parameter Values for the Transition Model in the Simulation
+(ref:sim-parameters-trans-note) The initial state probability is denoted by $\rho_i$.
+The transition probability for staying in the same state is denoted by $a_{i=j}$ and the probability for switching to a different state by $a_{i\neq j}$.
+The number of states in the model is denoted by *k*.
+
 ```{r sim-parameters-trans}
 
-sim.pars.trans <- as.data.frame(matrix(c("Interval", "-", "[.01,.99]", "1-a(i=j)/(k-1)",
+sim.pars.trans <- as.data.frame(matrix(c("Interval", "-", "[0.01,0.99]", "(1-a(i=j))/(k-1)",
                                            "Default", "1/k", "0.9", "0.1/(k-1)"), nrow = 2, byrow = T))
@@
 apa_table(sim.pars.trans,
-          col.names = c("", "Initial\nstate\nprob.", "Trans. prob.\nsame state", "Trans. prob.\nother state"),
+          align = "c",
+          col.names = c("", "$\\rho_i$", "$a_{i=j}$", "$a_{i \\neq j}$"),
           caption = "(ref:sim-parameters-trans)",
@@
 apa_table(cbind(rep(c("Interval", "Default"), 4), sim.pars.resp),
-          col.names = c("", "Shape", "Scale", "Shape", "Scale", "Mean", "Concentration"),
+          align = "c",
+          col.names = c("", "$\\alpha$", "$\\beta$", "$\\alpha$", "$\\beta$", "$\\mu$", "$\\kappa$"),
           col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
@@
 Models that could not be fitted were excluded from the recovery.
-
-(ref:sim-parameters-trans) Intervals and Default Parameter Values for the Transition Model in the Simulation
-(ref:sim-parameters-trans-note) The transition probability for staying in the same state is denoted by $a_{i=j}$ and the probability for switching to a different state by $a_{i\neq j}$.
-The number of states in the model is denoted by *k*.
 
@@
 (ref:tab-starting-values-note) Starting values for velocity and acceleration signals are shown before scaling down by factor 100.
+Shape parameters are denoted by $\alpha$, scale parameters by $\beta$, mean parameters by $\mu$, and concentration parameters by $\kappa$.
 
@@
 
-colnames(tab.start) <- c("Shape", "Scale", "Shape", "Scale", "Mean", "Concentration")
+#colnames(tab.start) <- c("$\\alpha$", "$\\beta$", "$\\alpha$", "$\\beta$", "$\\mu$", "$\\kappa$")
 rownames(tab.start) <- c("Fixation", "Saccade", "PSO", "Pursuit", "Event 5")
@@
 apa_table(as.data.frame(tab.start),
+          align = "c",
+          col.names = c("", "$\\alpha$", "$\\beta$", "$\\alpha$", "$\\beta$", "$\\mu$", "$\\kappa$"),
           col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
@@
 
-(ref:plot-example-classification) Example data from @Andersson2017 displayed as x-, and y-coordinates, velocity (in deg/s), acceleration (in deg/s$^2$), and sample-to-sample (relative) angle (in rad).
+(ref:plot-example-classification) Example data from @Andersson2017 displayed as x-, and y-coordinates (in deg), velocity (in deg/s), acceleration (in deg/s$^2$), and sample-to-sample (relative) angle (in rad).
 Colors indicate classification by gazeHMM and symbols by human coder MN.
@@
 
-```{r plot-example-classification, fig.cap="(ref:plot-example-classification)", fig.height=7.5}
+```{r plot-example-classification, fig.cap="(ref:plot-example-classification)", fig.height=7}
 
@@
 
-# Declarations
-## Funding
-Šimon Kucharský was supported by the NWO (Nederlandse Organisatie voor Wetenschappelijk Onderzoek) grant no. 406.10.559.
-
-## Conflicts of Interest
-The authors declare that they have no financial or non-financial interests regarding the content of this article.
-
-## Ethics Approval
-Not applicable.
-
-## Consent to Participate
-Not applicable.
-
-## Consent to Publication
-Not applicable.
-
-## Availability of Data and Materials
-See Open Practices and Data Availability Statement.
-
-## Code Availability
-See Open Practices and Data Availability Statement.
-
 # Open Practices and Data Availability Statement

### 2022-01-20 manuscript/supplementary_material_validation.Rmd

@@
 
-floatsintext      : no
+header-includes   :
+  - \usepackage{placeins}
+floatsintext      : yes
 figurelist        : no
@@
   mutate_at(c("Subject", "Block", "model"), as.factor) %>%
-  ggplot(aes(x = model, y = Subject)) + geom_tile(aes(fill = weight)) +
+  ggplot(aes(x = model, y = Subject)) + geom_tile(aes(fill = weight), color = "black") +
   facet_wrap(vars(Block), ncol = 2, labeller = label_both) +
@@
 
-  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight)) +
+  p <- ggplot(data.long, aes(x = model, y = subject)) + geom_tile(aes(fill = weight), color = "black") +
     scale_x_discrete(name = "Number of states", labels = as.character(1:5)) +
@@
 
+\FloatBarrier
 \newpage

### 2022-01-20 manuscript/scarf.R
add scarf plot

@@ -1,0 +1,87 @@
+library(tidyverse)
+library(here)
+library(patchwork)
+theme_set(theme_classic(base_size = 12.5))
+
+load(here("validation", "Andersson2017_raw.Rdata"))
+load(here("validation", "Andersson2017_fitted.Rdata"))
+
+merged <- bind_cols(A2017.fit[[3]][[3]][[4]]$samples, 
+                coderMN = A2017[[3]][[3]]$coderMN,
+                coderRA = A2017[[3]][[3]]$coderRA)
+
+rm(A2017, A2017.fit)
+
+df <- merged %>%
+  mutate(
+    gazeHMM = case_when(label == 0 ~ "Noise",
+                        label == 1 ~ "Fixation",
+                        label == 2 ~ "Saccade",
+                        label == 3 ~ "PSO",
+                        label == 4 ~ "Smooth pursuit"),
+    coderMN = case_when(coderMN == 1 ~ "Fixation",
+                        coderMN == 2 ~ "Saccade",
+                        coderMN == 3 ~ "PSO",
+                        coderMN == 4 ~ "Smooth pursuit",
+                        coderMN == 5 ~ "Blink",
+                        coderMN == 6 ~ "Other"),
+    coderRA = case_when(coderRA == 1 ~ "Fixation",
+                        coderRA == 2 ~ "Saccade",
+                        coderRA == 3 ~ "PSO",
+                        coderRA == 4 ~ "Smooth pursuit",
+                        coderRA == 5 ~ "Blink",
+                        coderRA == 6 ~ "Other")
+  ) %>%
+  mutate(
+    gazeHMM = factor(gazeHMM, levels = c("Fixation", "Saccade", "PSO", "Smooth pursuit", "Blink", "Noise", "Other")),
+    coderMN = factor(coderMN, levels = c("Fixation", "Saccade", "PSO", "Smooth pursuit", "Blink", "Noise", "Other")),
+    coderRA = factor(coderRA, levels = c("Fixation", "Saccade", "PSO", "Smooth pursuit", "Blink", "Noise", "Other"))
+  ) %>%
+  filter(t > 4.3, t <= 4.5)
+
+
+df_scarf <- df %>%
+  pivot_longer(cols = c("coderMN", "coderRA", "gazeHMM"), names_to = "method", values_to = "classification") %>%
+  mutate(
+    method = factor(method, levels = c("coderRA", "coderMN", "gazeHMM"))
+  )
+
+
[…]
